### 2014-10-20 paper/appendix.tex
paper and sim source files

\section{Modification to vertical structure by self-gravity}\label{vertsg_mod}
We describe a simple procedure to set up the vertical structure of a locally
isothermal, self-gravitating disc. We imagine setting up a non-self-gravitating disc,
then slowly switch on the vertical force due to self-gravity. We expect the midplane
density to increase at the expense of gas density higher in the atmosphere. It is assumed
that the temperature profile remains unchanged.

Vertical hydrostatic equilibrium between
gas pressure, stellar gravity and self-gravity reads
\begin{align}
c_\mathrm{iso}^2(R)\frac{\p\ln{\rho}}{\p z} = -\frac{\p\Phi_*}{\p z} - \frac{\p\Phi}{\p z}. 
\end{align}
Assuming a smooth radial density profile, we use the plane-parallel atmosphere
approximation for the disc potential, i.e.
\begin{align}
\frac{\p^2\Phi}{\p z^2} = 4\pi G \rho. 
\end{align}
Next, we write the density field as
\begin{align}
\rho(R,z) = \rho_N(R, z) \times \beta(z; R)
\end{align}
where $\rho_N$ is the density field corresponding to the non-self-gravitating disc:
\begin{align}
&c_\mathrm{iso}^2(R)\frac{\p\ln{\rho_N}}{\p z} = -\frac{\p\Phi_*}{\p z},\\
&\rho_N = \frac{\Sigma(R)}{\sqrt{2\pi}H(R)}\exp{(-z^2/2H^2)} \notag\\
&\phantom{\rho_N}\equiv \rho_{N0}(R)\exp{(-z^2/2H^2)},
\end{align}
where $\rho_{N0}=\rho_N(R,z=0)$ is the midplane density. The explicit
expression for $\rho_N$ above assumes a thin disc. The function
$\beta$ describes the modification to the local density in order to be
consistent with self-gravity. By construction, its governing equation
is 
\begin{align}
c_\mathrm{iso}^2(R)\frac{\p^2\ln{\beta}}{\p z^2} = -4\pi G\rho_N\beta. 
\end{align}
Let 
\begin{align}
& \chi  = \ln\beta - z^2/2H^2 ,\\
& \xi= \left(\frac{4\pi G \rho_{N0} }{c_\mathrm{iso}^2}\right)^{1/2} z,
\end{align}
then the governing equation can be written in dimensionless form
\begin{align}
&\frac{\p^2 \chi}{\p\xi^2} = -K -\exp{\chi}\label{vertsg_eqn},\\
& K \equiv \frac{c_\mathrm{iso}^2}{4\pi G \rho_{N0} H^2}. \notag
\end{align}
Note that $K(R)$ is proportional to the local Keplerian Toomre
parameter. Eq. \ref{vertsg_eqn} can be further reduced to a first
order differential equation, but this is unnecessary because we
pursue a numerical solution at the end. Appropriate 
boundary conditions are
\begin{align}
&\chi(z=0)  =  \ln\beta_0, \\
&\left.\frac{\p \chi}{\p \xi} \right|_{z=0} = 0.
\end{align}
$\beta_0(R)$ is the midplane density enhancement. To determine its value, we impose
the surface density before and after modification by self-gravity to
remain the same. Then we require 
\begin{align}\label{beta0_eqn}
&F(\beta_0) \equiv \sqrt{\frac{2K}{\pi}}\int_0^{ n /\sqrt{K} } 
\exp{\chi(\xi;\beta_0)}d\xi - \erf{\left(\frac{n}{\sqrt{2}}\right)} \notag\\
&\phantom{F(\beta_0)} = 0,  
\end{align}
where $n$ is the number of scale-heights of the non-self-gravitating
disc we originally considered. At a given cylindrical radius $R$, we
solve Eq. \ref{beta0_eqn} using Newton-Raphson iteration. Each
iteration involves integrating the governing ODE for $\chi$
(Eq. \ref{vertsg_eqn}). At the end of the iteration, we have $\beta(z;
R)$ and the midplane enhancement $\beta_0(R)$.

We comment that the procedure outlined above can be extended to
polytropic discs. In this case, there is an additional unknown --- 
the new disc thickness after adjustment by self-gravity and an
additional constraint --- the density should vanish at the new disc
surface.

### 2014-10-20 paper/discussion.tex
paper and sim source files

\section{Additional results analysis}\label{additional}
In this section we examine some secondary quantities derived
from the hydrodynamic simulations above. To keep this discussion
concise, we will use selected simulations from above for illustration.

\subsection{Three-dimensionality}
A simple measure of three-dimensionality of the flow is to compare vertical 
to horizontal motion. Since we are interested in non-axisymmetric
perturbations to the gap edge, we first Fourier transform the
meridional momentum densities

\begin{align}
  (v_{Rm}, v_{zm} ) \equiv\int_0^{2\pi}\rho\times(u_R,
  u_z)\exp{(-\ii m\phi)}d\phi. 
\end{align}
We define the three dimensionality as $\Theta_m(z/H)$, where 
\begin{align}
  \Theta_m^2 \equiv \frac{\avg{|v_{zm}|^2}}{\avg{|v_{zm}|^2}
    +\avg{|v_{Rm}|^2}},  
\end{align}
and $\avg{\cdot}$ denotes a radial average. 
Admittedly, this is a crude measure, and exact values of $\Theta_m$
varies somewhat with details of the average. However, we have
experimented with different averaging domains and 
found the features described below are robust.

The top panel in Fig. \ref{compare_vprofiles_3d008} shows $\Theta_m$
for Cases 1--3 at $t=40P_0$. The radial average is taken over
$r-r_p\in[3,7]r_h$. These are all vortex modes (see
Fig. \ref{vortex8_polar_dens} and
Fig. \ref{vortex4_vortex10_overall}). The flow becomes increasingly
three-dimensional away from the midplane but 
$\Theta_m = O(10^{-1})$ is small. In an averaged sense the flow is
mostly horizontal. At the end of the  
simulation for Case 1, an azimuthally extended vortex dominates the
flow, for which we measured $\Theta_1\sim
0.2$---0.3. Thus, although vertical motion can become an appreciable
fraction of horizontal motion, the former never dominates.

$\Theta_m$ for Cases 4---7 are shown in in bottom panel of
Fig. \ref{compare_vprofiles_3d008}. The radial average is performed
over $r-r_p\in[2,6]r_h$ 
because the global spirals in Cases 6---7 significantly protrude the
gap edge. The snapshot is taken at $t=50P_0$ for Cases 4---5, at
$t=40P_0$ for Case 6 and at $t=30P_0$, so that the
vortices and spirals have comparable over-densities at the
gap edge. It also reflects the fact that spiral modes are more
unstable than vortex modes and develop earlier  
\citep{lin11b}. $\Theta_m \sim 0.2$ is again not particularly
large, but the spiral modes are distinctly more three-dimensional than
vortex modes. This is likely due to additional vertical
acceleration provided by the strong self-gravity in those cases.

%
% can't do detail comparision between cases because instability have
% different growth rates. different 'stages' in instability

% Case 6 show 3 edge disturbances but this is likely to be a mixture
% of spiral and vortex modes, rather than the m=3 spiral mode. This is
% because 
% we clearly see m=2 spiral in Case 7 (Q_0=1.5), and to see higher m spiral
% modes one should increase the strength of self-gravity (e.g. lin11b
% observe m=3 spirals with Q_0=1.2) 
%

\begin{figure}
  \centering
  \includegraphics[scale=.425,clip=true,trim=0.2cm 1.7cm 0cm
    0cm]{figures/compare_vprofiles_3d008}  
  \includegraphics[scale=.425,clip=true,trim=0.2cm 0cm 0cm
    0.2cm]{figures/compare_vprofiles_3d010}  
  \caption{Three-dimensionality of the non-axisymmetric
    flow near the outer gap edge. Top: Cases 1---3 (vortex modes). 
    Bottom: Cases 4---5 (vortex modes) and Cases 6---7 (spiral
    modes). The azimuthal wavenumber $m$ is chosen to match the number
    of vortices or large-scale spirals observed. 
    \label{compare_vprofiles_3d008}}
\end{figure}

\subsection{Vortensity field}\label{vortensity}
A fundamental distinction between the linear vortex and edge mode
instability is their association with local vortensity minimum and
maximum, respectively. In this section we compare vortensity fields of 
discs with vortex modes (Case 2) and edge modes (Case 7). 
%We use Case 2 
%and Case 7 as examples, respectively. 
More specifically, we examine the relative
perturbation to the vertical component of vortensity,  
\begin{align}
  \Delta\eta_z\equiv \frac{\eta_z - \eta_z(t=0)}{\eta_z(t=0)},
\end{align}
where
\begin{align}
  \eta_z \equiv \frac{\bm{\hat{z}}\cdot\nabla\times\bm{u}}{\rho}. 
\end{align}

Fig. \ref{vortex4_vortex1_vortxy} compares $\Delta\eta_z$ in the
midplane when vortices and spirals develop. For planetary gaps, 
vortensity maxima and minima are both located near the gap edges with 
characteristic separation of the local scale-height. The vortensity
ring at the inner gap edge ($r-r_p\simeq - 2r_h$) remain
well-defined. The vortex instability is associated with the local
vortensity minimum near the outer gap edge --- seen as localised closed
contour lines centred about $r - r_p \sim 4r_h$. The vortensity ring at
$r-r_p\sim +2r_h$ becomes distorted as a \emph{consequence} of
large-scale vortex formation just exterior to it. By contrast, the
edge-spiral mode is associated with the local vortensity
maximum. Their development inherently disrupts the vortensity
rings. This is seen in the right panel as the outer ring is broken up.

\begin{figure}
  \centering
  \includegraphics[scale=.425,clip=true,trim=0cm 0cm 1.9cm 
    0.9cm]{figures/vortex4_vortxy_008}\includegraphics[scale=.425,clip=true,trim=2.3cm    
    0.0cm 0cm 
    0.9cm]{figures/vortex1_vortxy_006}
  \caption{Relative perturbation to the vertical component of
    vortensity at the midplane in a disc with the vortex instability
    (left, Case 2 at $t=40P_0$) and the 
    spiral instability (right, Case 7 at $t=30P_0$). Negative
    perturbations in the region $r-r_p\in[2,6]r_h$ are outlined by white lines. 
    Dotted horizontal lines indicate azimuthal cuts taken in
    Fig. \ref{vortex4_vortex1_vortRZ}.   
    \label{vortex4_vortex1_vortxy}}
\end{figure}

The vertical structures also differ. Fig. \ref{vortex4_vortex1_vortRZ}
compares $\Delta\eta_z$ at azimuths coinciding with 
a vortex or the edge disturbance of the spiral mode. Both
instabilities involve $\Delta\eta_z<0$. It is clear that 
the spiral mode has stronger vertical dependence. Its 
region of $\Delta\eta_z<0$ becomes thinner away from
the midplane. In the vortex case this region remains 
about the same width and  $\Delta\eta_z$ is approximately uniform
within it. 
%This is qualitatively
%consistent with increased (downards) vertical motion. 
While $\mathrm{min}(\Delta\eta_z)$ is of comparable
magnitude, the vortensity ring at $r-r_p=2r_h$is much weaker and
thinner in the spiral case ($\Delta\eta_z$ being a factor $\sim 4$
smaller than the vortex case).

\begin{figure}
  \centering
  \includegraphics[scale=.47,clip=true,trim=0cm 1.32cm .0cm
    .6cm]{figures/vortex4_vortRZ_008}\\\includegraphics[scale=.47,clip=true,trim=0cm 
    0.cm .0cm
    .6cm]{figures/vortex1_vortRZ_006}
  \caption{Relative vertical vortensity perturbation
    associated with the vortex instability (top) and spiral
    instability (bottom). The slices are taken at azimuths shown by
    white dotted lines in Fig. \ref{vortex4_vortex1_vortxy}. Negative
    perturbations in the region $r-r_p\in[2,5.5]r_h$ are outlined by
    white lines. \label{vortex4_vortex1_vortRZ}} 
\end{figure}

\subsection{Disc-planet torques}
The presence of non-axisymmetric disturbances at gap edges is
expected to significantly affect disc-planet torques. It has been 
confirmed in 2D simulations that both vortex and spiral modes lead to
oscillatory torques of either sign \citep{li05,lin11b}. 
%Limited 
%numerical resolution the current 3D calculations prevent accurate torque measurements, 
It this section we measure the disc-on-planet torques in several of the above simulations
to confirm the main features found in 2D.

We calculate the specific torque acting on the planet due to a mass
element as 
\begin{align}
d\bm{T}(\bm{r}) &=
\frac{\bm{r}_p\times\bm{r}G\rho(\bm{r})d^3\bm{r}}{d_p^3}f(\bm{r},\bm{r}_p),\\
f(\bm{r},\bm{r}_p)&\equiv 1
-\exp{\left(-\frac{1}{2}\left|\frac{\bm{r}-\bm{r}_p}{\epsilon_c
      r_h}\right|^2\right)}.  
\end{align} 
The tapering function $f$ reduce contributions from close to the
planet, thereby reducing numerical artifacts arising from this region
because of the diverging potential and limited resolution. We set  
the parameter $\epsilon_c=1$ so that tapering does not significantly
reduce contributions from the instabilities, since they develop
at $\gtrsim 2r_h$ away from the planet's orbital radius.

Fig. \ref{torque3} shows the disc-on-planet torques in Case 3 and Case 5, which develop
the $m=5$ and $m=6$ vortex modes, respectively. These plots are qualitatively similar to
2D simulations \citep[e.g.][]{li05}. The torques oscillate on orbital 
time-scales and its instantaneous values can be of either sign. However, 
upon averaging over the simulation we find the total torques are negative in both cases. This
means inwards migration is still favoured. 
%The planet is expected to interact

We extended Case 5 to $t=135P_0$ and find the vortices 
have similar over-densities as at $t=50P_0$. However, 
Fig. \ref{torque3} show the torque oscillation amplitudes decrease
towards the end of Case 5 compared to $t\in[40,80]P_0$.  
At $t=50P_0$ the vortices are located in $r-r_p\in[3.5,5.5]r_h$ but by
$t=135P_0$ they are located in $r-r_p\in[4,6]r_h$. 
Given that $t\in[40,80]P_0$ is only $20P_0$ to $50P_0$ after the planet  
potential has been fully introduced, gap-formation is probably 
ongoing during this time. We expect torque amplitudes to be larger
during  gap-formation since the vortices lie closer to the planet. 
%On the
%other hand, gap-widening in the presence of vortices was also seen in  
%2D simulations on long timescales \citep{lin11a}  
%So there may also be
%contributions from radial redistribution of material due to the
%vortices. 
% This might be
%brought about by the of the density bump at the outer gap edge
%\citep{meheut12b}

%The original
%density bump at the outer gap edge ($r-r_p\simeq 4r_h$) shifts outward
%between these times. 
%This is unlike in the fixed-orbit 2D simulation by \cite{li05}, where
%the oscillation amplitude does not decrease over several hundred
%orbits. 
%The different setups (especially self-gravity and numerical resolution) 
%prevent a detail comparison 
roclinic instability

\begin{figure}
  \centering
  \includegraphics[scale=.425,clip=true,trim=0.2cm 1.cm 0cm
    0cm]{figures/vortex10_torque3_tqex}  
  \includegraphics[scale=.425,clip=true,trim=0.2cm 0cm 0cm 0.2cm]{figures/vortex2b_torque3_tqex} 
  \caption{Instantaneous disc-on-planet torques in simulations where the vortex mode develops. 
    Top: Case 3. Bottom: Case 5. Note that Case 5 has been extended to $t=135P_0$. \label{torque3}}
\end{figure}

Next we examine disc-planet torques in the presence of the spiral
modes. The top panel of Fig. \ref{torque3_spiral} shows the
instantaneous disc-on-planet torques. Contributions from the
inner disc ($r < r_p$) and outer disc ($r > r_p$) are plotted
separately for comparison with Fig. 18b in \cite{lin11b}, which is 
similar to the present plot. Large oscillations in the outer torque
due to edge mode spirals cause the total torque to be positive or
negative at a given instant.

Unlike the vortex modes, Fig. \ref{torque3_spiral} shows that spiral
modes can lead to a positive running-time averaged torques (bottom panel). 
The average torques become more positive with time after spiral modes develop, 
and with increasing self-gravity (which increases the instability strength). 
%The torque is more positive with
%increasing strength of the spiral instability.
This was also observed in high-resolution 2D simulations in
\cite{lin11b}. There it was suggested that the creation of large   
`voids' in between spiral arms decreases the time-averaged density in  
the planet-induced wakes, thereby reducing associated torque
magnitudes. Since the outer planetary wake normally provide a negative
torque, the spiral modes make the total torque more positive. 
The similarity between 2D and 3D results indicate that
outwards migration induced by spiral modes, which was seen in 2D by
\cite{lin11b,lin12b}, will also operate in 3D.

%It was later shown that the spiral instability can supply material to
%go on horseshoe turns ahead of the planet, which also provide a
%positive torque \citep{lin12b}. This is because the spiral modes 
%are associated with local vortensity maximum which are located just
%inside the gap edge in the co-orbital region.

%This is because the
%spirals are associated with local vortensity maximum, which is just
%inside the outer gap edge. Thus over-densities can be supplied by the
%instability to go on horseshoe orbits

%Since spiral modes are associated with disturbances at the local
%vortensity maximum just inside the outer gap edge, its influence on
%disc-planet torques is expected to be more si

\begin{figure}
  \centering
  \includegraphics[scale=.425,clip=true,trim=0.2cm 1.825cm 0cm
    0cm]{figures/vortex1_torque2_tqex}  
  \includegraphics[scale=.425,clip=true,trim=0.2cm 0cm 0cm 0.2cm]{figures/torque3_tqex} 
  \caption{Disc-on-planet torques in the presence of spiral modes
    associated with the outer gap edge. Top: total torque (solid),
    torque from the inner disc (dotted) and from the outer disc
    (dashed) in Case 7. Bottom: time-averaged torques in Case 6
    (solid) and Case 7 (dotted).
\label{torque3_spiral}}
\end{figure}

### 2014-10-20 paper/model.tex
paper and sim source files

\section{Governing equations and numerical models}\label{model} 
The system of interest is a self-gravitating, inviscid fluid disc
orbting a central star of mass $M_*$. We will mainly consider two-dimensional
(2D, or razor-thin) discs in favour of numerical
resolution, but have also carried out some three-dimensional (3D) disc
simulations. Hence, for generality we describe the system in
3D. We use both cylindrical $(R,\phi,z)$ and spherical polar
coordinates $(r,\theta,\phi)$ centred on the star. 
The governing fluid equations in 3D are  
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) =
  0,\label{cont_eq}\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bv = -\frac{1}{\rho}\nabla
  p -\nabla \Phi_\mathrm{tot}\label{mom_eq},\\ 
  & \nabla^2\Phi_d = 4 \pi G \rho \label{poisson}, 
\end{align}
where $\rho$ is the mass density, $\bm{v}=(v_r,v_\theta,v_\phi)$ the
fluid velocity, $p$ is the pressure and the total potential
$\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
central star, 
\begin{align}
  \Phi_*(r) = -\frac{GM_*}{r}, 
\end{align}
where $G$ is the gravitational constant,  and the disc potential
$\Phi_d$. We further assume a locally 
isothermal equation of state 
\begin{align}
  p = c_s^2(R)\rho,
\end{align}
where the sound-speed is given by $c_s=H\Omega_k$, where $H=hR$ is the
disc scale-height with constant aspect-ratio $h$ chosen below, and
$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency.

We have purposefully adopted the simplest model that was found to
develop one-arm spirals in discs with radial structure. The indirect
potential associated with the non-inertial reference frame is
neglected to avoid complications arising from the motion of the
central star. For the adopted disc masses $M_d\lesssim 0.1 M_*$, this
effect is not expected to be significant. The locally isothermal
equation of state allows us to apply standard linear density wave
theory for results interpretation. 
%This allows us to consider
%an isolated disc.

The 2D disc equations are obtained by replacing $\rho$ with the surface
mass density $\Sigma$ and $p$ with the vertically-integrated pressure
$\Pi$ in the continuity and momenta equations, and the 2D fluid
velocity $\bm{v}=(v_R,v_\phi)$ is evaluated at the midplane, as are
the forces from the potential. In the Poisson equation, $\rho$ is
replaced by $\Sigma\delta(z)$, where $\delta(z)$ is the
delta function.

\subsection{Disc model and initial conditions}
%We consider a smooth disc which contains a surface density bump
%between $R\in[R_{d1},R_{d2}]$, mimicking mass built-up in a dead
%zone. 
We adopt a modified power-law disc with surface
density profile 
\begin{align}
  \Sigma(R) = \Sigma_\mathrm{ref} \left(\frac{R}{R_0}\right)^{-s}\times B(R;
  R_{d1}, R_{d2}, \epsilon, \delta R_d), 
\end{align}
where $R_0$ is a reference cylindrical radius and $s$ is the power-law
index describing the smooth disc, and $\Sigma_\mathrm{ref}$ is a
surface density scale chosen by specifying $Q_\mathrm{out}$,
the Keplerian Toomre parameter at $R=R_{d2}$,
\begin{align}
  Q_\mathrm{out} = \left.\frac{c_s\Omega_k}{\pi G
    \Sigma}\right|_{R=R_{d2}}. 
\end{align}

The bump function
$B(R)$ represents a surface density boost between
$R\in[R_{d1},R_{d2}]$ by a factor $\epsilon^{-1}$ ($\epsilon < 1$),
and $\delta R_d$ is the transition width between the bump and the
smooth disc. We choose 
\begin{align}
  &B(R) = f_1(R)\times f_2(R),\\
  &f_1(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 +
    \tanh\left(\frac{R-R_{d1}}{\Delta_1}\right)\right]  + \epsilon,\\
  &f_2(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 -
    \tanh\left(\frac{R-R_{d2}}{\Delta_2}\right)\right]  + \epsilon,
\end{align}
where $\Delta_1 = \delta R_d \times H(R_{d1})$ and similarly for
$R_{d2}$. The bump function mimicks mass built-up in a dead zone.

In 3D, we assume vertical hydrostatic balance
\begin{align}
  0 = \frac{1}{\rho}\frac{\p p}{\p z} + \frac{\p\Phi_*}{\p z} + \frac{\p
    \Phi_d}{\p z},  
\end{align}
which gives the mass density as 
\begin{align}
  \rho = \frac{\Sigma}{\sqrt{2\pi}H}Z(R,z),
\end{align}
where $Z(R,z)$ describes vertical stratification. In practice, we
numerically solve for $Z(R,z)$ by neglecting radial self-gravity
force compared to vertical self-gravity, which reduces the equations
for vertical hydrostatic equilibrium to ordinary differential
equations.  This procedure is described in \cite{lin12b}.

Our fiducial parameter values are: $s=2$, $R_{d1}=R_0$, $R_{d2}=2R_0$,
$\epsilon=0.1$, $\delta R_d=5$, $h=0.05$ and
$Q_\mathrm{out}=2$. Initially there is no vertical or radial velocity
($v_R = v_r = v_\theta = 0$). The azimuthal velocity is initialized to
satisfy centrifugal balance with pressure and gravity,
\begin{align}
  \frac{v_\phi^2}{r} = \frac{1}{\rho}\frac{\p p}{\p r} + \frac{\p
    \Phi_\mathrm{tot}}{\p r}
\end{align}
and similarly in 2D. The angular velocity is $\Omega = v_\phi/R$.

\section{Numerical methods}
%We will mostly employ 2D simulations in favour of resolution, but have
%also run some cases in 3D. We describe here the numerical setup used
%in each code. 
%introduce notation, terminlogy for each code

We adopt computational units such that $G=M_*=R_0=1$. Time is measured
in the Keplerian orbital period at the reference radius, $P_0\equiv 
2\pi/\Omega_k(R_0)$.

\subsection{FARGO}
Our primary code is FARGO with self-gravity \citep{baruteau08}. This
is a popular, simple finite-difference code for 2D discs. `FARGO' refers
to its azimuthal transport algorithm, which circumvents time-step
contraint set by the inner disc boundary. 
The 2D Poisson
equation is solved in integral form using Fast Fouier Transform. This
requires a softening length $\epsilon_g H$, where $\epsilon_g$ is an
input parameter.

The 2D disc occupies
$R\in[R_\mathrm{min},R_\mathrm{max}],\,\phi\in[0,2\pi]$ and is
divided into $(N_R,N_\phi)$ grids, logarithimically spaced in radius and
uniform spaced in azimuth. At radial boundaries we set the
hydrodynamic variables to their initial values.

%poisson kernal integration, softening 
%where used?
%BC
\subsection{ZEUS-MP}
ZEUS-MP  is a general-purpose finite difference
code \citep{hayes06}. We use the code in 3D spherical geometry, covering
$r\in[r_\mathrm{min},r_\mathrm{max}]$, $\theta\in[\theta_\mathrm{min},\pi/2]$,
$\phi\in[0,2\pi]$. The vertical domain is chosen to cover $n_H$
scale-heights, i.e. $\tan{(\pi/2 - \theta_\mathrm{min})}/h=n_H$. 
The grid is logarithmically spaced in radius and uniformly spaced in the angular
coordinates. We assume reflectional symmetry across the midplane, and
apply relfective boundary conditions at radial boundaries and the
upper disc boundary.

ZEUS-MP solves the 3D Poisson equation using a conjugate gradient
method. To suppliment boundary conditions for the linear solver, we
expand the boundary potential using in spherical harmonics $Y_{lm}$ 
as described in \cite{boss80}. The expansion is truncated at
$\lmax,\mmax$. This code was used in \cite{lin12b} for
self-gravitating disc-planet simulations.

%used in lin 2012 
\subsection{PLUTO} 
PLUTO is a general-purpose Godunov code \citep{mignone07}. The grid
setup is the same as that adopted in ZEUS-MP above. We configure the
code similarly to that used in \cite{lin14}: piecewise linear
reconstruction, a Roe solver and second order Runge-Kutta time
integration. We also enable the FARGO algorithm for azimuthal
transport.

We solve the 3D Poisson equation throughout the domain using spherical
harmonic expansion \citep{boss80}, as used for the boundary potential
in ZEUS-MP. This version of PLUTO was used in \cite{lin14b} for
self-gravitating disc-planet simulations, producing similar results to
that of ZEUS-MP and FARGO.

%used in conference proceeding

\section{Fiducial simulations}
Here we describe one numerical simulation from each code to give an
overall picture of disc evolution. In the next section, we explore the
effect of disc parameters using 2D simulations.

The disc evolution is quantified using mode amplitudes and angular
momenta as follows. We list the 2D definitions with obvious 3D
generalisations. The hydrodynamic variables are written as 
\begin{align}
  f(R,\phi) &= \sum_{m=-\infty}^{\infty}f_m(R)\exp{\ii m \phi} \notag\\
  &= f_0 + 2 \real\left[\sum_{m=1}^\infty f_m \exp{(\ii
      m\phi)}\right], 
\end{align}
where the $f_m$ may be obtained from Fourier transform in $\phi$. The
mode amplitude is 
\begin{align}
  A_m = \frac{|C_m|}{M_{d0}}
\end{align}
where $M_{d0}$ is the initial disc mass and 
\begin{align}
  C_m = 2\pi\int_{R_1}^{R_2} \Sigma_m RdR, 
\end{align}
with integration limits chosen later.

The total disc angular momentum in $R\in[R_1,R_2]$ is
\begin{align}
  J &= \int_{R_1}^{R_2}\int_0^{2\pi} \Sigma Rv_\phi RdRd\phi \notag\\
  &= 2\pi\int_{R_1}^{R_2}R\Sigma_0 v_{\phi0} R dR \notag\\ 
  &\phantom{=}+
  \sum_{m=1}^\infty2\pi\int_{R_1}^{R_2} 2R\real\left[\Sigma_m v_{\phi
      m}^*\right] RdR, \notag\\
  &= \sum_{m=0}^\infty 2\pi \int_{R_1}^{R_2} j_m R dR = \sum_{m=0}^\infty J_m. 
\end{align}
where $^*$ denotes complex conjugation, and we idenfity $J_m$ as the angular momentum 
associated with the $m^\mathrm{th}$ mode, and $j_m$ its density.

\subsection{WKB perturbations}
We will use results from linear density wave theory to set up
perturbations. 
Here, we follow conventions in \cite{shu91}. Recall the
dispersion relation for tightly-wound density waves of the form
$\exp{(\ii \omega t - m \phi + kR)}$ in a razor-thin
isothermal disc, 
\begin{align}\label{dispersion}
  (\omega - m\Omega)^2 = \kappa^2 + k^2c_s^2 - 2\pi G \Sigma |k|, 
\end{align}
where $\omega$ is the mode frequency, $\kappa^2 =
R^{-3}\p_R(R^4\Omega^2)$ is the square of the epicycle frequency, and
$k$ is a real wavenumber.

If we specify the pattern speed $\Omega_p \equiv \omega/m$,
Eq. \ref{dispersion} can be solved for $|k|$,
\begin{align}\label{wavenumber}
  \frac{|k|}{k_T} = \frac{2}{Q^2}\left[1 \pm \sqrt{1 -
      Q^2(1-\nu^2)}\right], 
\end{align}
where 
\begin{align}
  k_T \equiv \frac{\kappa^2}{2\pi G \Sigma}
\end{align}
is a characteristic wavenumber, 
\begin{align}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}
\end{align}
is the usual Toomre parameter, and
\begin{align}
\nu \equiv \frac{(\omega - m\Omega)}{\kappa}
\end{align}
is a dimensionless frequency. 
We use Eq. \ref{wavenumber} to set trailing long waves as
perturbations, i.e. $k/k_T = -2Q^{-2}\left[1 - \sqrt{1 -
    Q^2(1-\nu^2)}\right]$.  
We only apply the perturbations where $Q^{-2}(1-\nu^2)<1$.

The amplitude of the surface density and velocity perturbations are
related by
\begin{align}
  &\left(\omega - m\Omega\right)\delta\Sigma + k\Sigma\delta v_R= 0,\\
  &\frac{\kappa^2}{2\Omega} \delta v_R + \ii (\omega - m\Omega)\delta
  v_\phi=0. 
\end{align}
We set $\delta\Sigma = 10^{-3}\Sigma$ and $(\delta v_R, \delta
v_\phi)$ accordingly.

%Having chosen the wave frequency and set the wavenumber according to
%the above,

### 2014-10-20 paper/paper.tex
paper and sim source files

\documentclass[useAMS,usenatbib]{mn2e}
\usepackage{epsfig,rotate,graphicx}
\usepackage[fleqn]{amsmath}
\usepackage{subfigure}
\usepackage{lscape}
\usepackage{bm}

\newcommand{\p}{\partial}
\newcommand{\mnras}{MNRAS}
\newcommand{\apj}{ApJ}
\newcommand{\aap}{A\&A}
\newcommand{\apjl}{ApJL}
\newcommand{\araa}{ARAA}
\newcommand{\dd}{\delta}
\newcommand{\adot}{\dot{a}}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\gtrsim}{\;\raisebox{-.8ex}{$\buildrel{\textstyle>}\over\sim$}\;}
\newcommand{\lesssim}{\; \raisebox{-.8ex}{$\buildrel{\textstyle<}\over\sim$}\;}
\newcommand{\anrev}{{\it ARA\&A, }}
\newcommand{\apjs}{{\it ApJS, }}
\newcommand{\icar}{{\it Icarus, }}
\newcommand{\mnr}{{\it MNRAS, }}
\newcommand{\nat}{{\it Nature, }}
\newcommand{\sci}{{\it Sci, }}
\newcommand{\ana}{{\it A\&A, }}
\newcommand{\anas}{{\it A\&AS, }}
\newcommand{\aaps}{{\it A\&AS, }}
\newcommand{\anar}{{\it A\&AR, }}
\newcommand{\prd}{{\it Phys. Rev. D }}
\newcommand{\qjl}{{\it QJRAS, }}
\newcommand{\sbar}{\bar{\sigma}}
\newcommand{\avg}[1]{\langle #1 \rangle_\phi}
\newcommand{\bu}{\bm{u}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\mmax}{m_\mathrm{max}}
\newcommand{\trel}{t_\mathrm{rel}}
\newcommand{\zeus}{{\tt ZEUS-MP }}
\newcommand{\ii}{\mathrm{i}}
\newcommand{\bv}{\bm{v}}
\newcommand{\rin}{r_\mathrm{in}}
\newcommand{\rout}{r_\mathrm{out}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\tbeta}{\tilde{\beta}}
\newcommand{\teta}{\tilde{\eta}}
\newcommand{\tcool}{t_\mathrm{cool}}
\DeclareMathOperator{\erf}{erf}
\DeclareMathOperator{\real}{Re}
\DeclareMathOperator{\imag}{Im}

\usepackage{array,booktabs,tabularx}
\newcolumntype{R}{>{\centering\arraybackslash}X} % right justified tabularx columns

\title[One-arm spirals in self-gravitating discs]{One-arm spirals in
  self-gravitating dead zones in protoplanetary discs}

\author[Lin]{Min-Kai Lin$^{1}$
  \thanks{ minkailin@email.arizona.edu} \\ 
  $^1$Department of Astronomy and Steward Observatory, University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA 
}

\begin{document}

\maketitle
\begin{abstract}

\end{abstract}

\begin{keywords}
  planetary systems: formation --- planetary systems:
  protoplanetary discs
\end{keywords}

\section{Introduction}\label{intro}

\input{model}
%\input{results}
%\input{discussion}
%\input{summary}

%\section*{Acknowledgments}

\bibliographystyle{mn2e}
\bibliography{ref}

\appendix
%\input{appendix}
%\input{appendix2}

\end{document}

### 2014-10-20 paper/results.tex
paper and sim source files

\section{Growth of non-axisymmetric modes without the influence of the
  planet}\label{linear1}
In this section, the planet is introduced at $t=20P_0$ and 
its potential switched on over $10P_0$. At $t=30P_0$ we switch off the
planet potential and azimuthally average the surface density, energy
and velocity fields. At this point the planet has carved a partial
gap. We then perturb the surface density for $r>r_p$ and continue to 
evolve the disc. We impose  sinusoidal perturbations with 
azimuthal wavenumbers $m\in[1,10]$ and random amplitudes within $\pm 0.01$
 times the local surface density.% {\bf what are PERTMIN and PERTMAX values?}  
This procedure allows us to analyse the growth of 
non-axisymmetric modes associated with the gap, but without
complications from non-axisymmetry arising directly from disc-planet
interaction.

Note that these `planet-off' simulations are not linear stability
calculations because since the cooling term in our energy equation
restores the initial temperature profile corresponding to $H/r=0.05$,
rather than the heated gap edge. Linear stability calculations and
adiabatic cases will be further discussed in
section~\ref{adiabatic_section}.

Simulations here employ a resolution of $(N_r,N_{\phi})=(1024,2048)$
with an open boundry condition at the outer disk edge $r_\mathrm{out}=25r_\mathrm{in}$. We consider 
%for a time up to $200P_0$ allowing high resolution and detection of high
%$m$ structures.
cases of $\tilde{\beta}=0.1,1,10$ corresponding to fast, moderate,
and slowly cooled discs. % {\bf boundary condition for these sims?}

\subsection{Gap structure}
%Since stability of the gap is of interest 
We first examine the gap structure formed by planet-disc
interaction as a function of the cooling time. The azimuthally-averaged 
gap profiles are shown in Fig. \ref{intial1D} for varying
$\tilde\beta$. Gaps formed with lower $\tilde\beta$ (faster cooling)
are deeper with steeper gradients at the gap edges. Faster cooling rates also 
increase and decrease the surface density maxima and 
minima increase, respectively. However, a clean gap does not form
in this short time period.

%For all cases the gap width is $\approx 5r_{hill}$
%which is set by torque balances in the disk \citep{crida06}.

Larger $\tilde\beta$ values result in higher disc aspect ratios $h=H/r$,
i.e. higher temperatures. Heating mostly occur at the gap edges
due to planet-induced spiral shocks. Decreasing the cooling time
implies that this heat is retained in the disc. In the inviscid limit the gap
opening condition is $r_h\gtrsim H$ or $q\gtrsim 3h^3$
\citep{crida06}. 
% which is a value directly correlated to temperature in our model. 
%For a planet to carve a gap in the surrounding material it has been
%shown that $q>H^3$ is needed \citep{crida06}. 
This indicates that for hotter discs (higher
$h$), it becomes more difficult for a planet of fixed $q$ to open a
gap. This explains the shallower gaps in surface density when
$\tilde{\beta}$ is increased.

%This is seen in our simulations as changes in $\tilde{\beta}$
%result in changes of $h$ by less than $20\%$ yet gap depth almost
%doubles.

The important consequence of a heated gap edge is that the
generalized vortensity profiles become smoother with increasing
cooling times, with the extrema becoming less pronounced. Previous locally
isothermal disc-planet simulations show the RWI associated with PV
minima \citep{li05,lin10}. We can therefore expect the RWI to be associated with
minima in the generalized vortensity (corresponding to local surface
density maxima) in the non-isothermal case. Because the extrema are
less sharp, the RWI is expected to be weaker and the gap to be more
stable with longer cooling times.

%Charactereristicly
%for such disk-planet Rossby vortices the generalized vortensity
%minima correspond to density maxima in the disk. 
%Since generalized vortensity extrema are also
% correspondingly less pronounced and density gradients become smaller
% we expect that the stability of gaps to increase with larger
% $\tilde{\beta}$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
    2cm 0cm 0cm]{figures/compare_sigma}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
    2cm 0cm 1cm]{figures/compare_aspectratio}
\includegraphics[width=\linewidth,clip=true,trim=0.5cm
    0.5cm 0cm 1cm]{figures/compare_gvortensity}
  \caption{Gap profiles at $t=30P_0$ for the intial partial gap opened
    before instability emerges for fast (solid), moderate
    (dashed), and slow cooling (dashed-dot). The relative surface density
    pertibation (top), disc aspectratio (middle) and generalized
    vortensity pertibation (bottom) are shown. \label{intial1D}}  
\end{figure}

%\begin{tabularx}{0.4\textwidth}{l*{10}{R}} \toprule
%  \multicolumn{11}{c}{$\tilde{\beta}=0.1$} \\ \midrule
%  m                    & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10  \\ 
 % $\gamma10^2/\Omega(r_o)$ & 6.56 & 6.82 & 6.73 & 5.78 & 6.00 & 6.38 & 5.97 & 5.62 & 4.61 & 3.36   \\ \bottomrule
%\end{tabularx}

%\begin{tabularx}{0.4\textwidth}{l*{5}{R}} \toprule
%  \multicolumn{6}{c}{$\tilde{\beta}=1.0$} \\ \midrule
%  m                    & 1 & 2 & 3 & 4 & 5  \\ 
%  $\gamma10^2/\Omega(r_o)$ & 1.27 & 1.28 & 1.35 & 1.01 & 0.61   \\ \bottomrule
%\end{tabularx}

\begin{table}
  \centering
  \caption{Dominant mode and growth rates for
    $\tilde{\beta}=0.1,1.0,10.0$ (fast, moderate, and slow cooling)
    values during `planet-off' simulations \label{modetable}} 
  \hfill
  \begin{minipage}{0.3\linewidth}
    \begin{tabularx}{\textwidth}{l R} 
      \multicolumn{2}{c}{$\tilde{\beta}=0.1$} \\ 
      \toprule
      $m$ & $10^2\gamma/\Omega(r_p)$ \\
      \midrule
      6 & 7.3 \\
      7 & 7.8 \\
      8 & 7.9 \\
      9 & 7.9 \\
      10 & 6.8 \\ 
      \bottomrule
    \end{tabularx}
  \end{minipage}
  \hfill
  \begin{minipage}{0.3\linewidth}
    \begin{tabularx}{\textwidth}{l R} 
      \multicolumn{2}{c}{$\tilde{\beta}=1.0$} \\ 
      \toprule
      $m$ & $10^2\gamma/\Omega(r_p)$ \\
      \midrule
      3 & 2.0 \\
      4 & 2.2 \\
      5 & 2.3 \\
      6 & 1.6 \\
      7 & 1.1 \\ 
      \bottomrule
    \end{tabularx}
  \end{minipage}
  \hfill
  \begin{minipage}{0.3\linewidth}
    \begin{tabularx}{\textwidth}{l R} 
      \multicolumn{2}{c}{$\tilde{\beta}=10.0$} \\ 
      \toprule
      $m$ & $10^2\gamma/\Omega(r_p)$ \\
      \midrule
      1 & 1.1 \\
      2 & 1.6 \\
      3 & 1.7 \\
      4 & 1.2 \\
      5 & 0.1 \\ 
      \bottomrule
    \end{tabularx}
  \end{minipage}
  \hfill
\end{table}

{\bf
\subsection{Axisymmetric stability}
Check that the gap profiles are stable against axisymmetric
stability. Can use analytical criteria given in section 3.1 of Li et
al (2000, ApJ, 533, 1023). No plot needed, just statements. 
}

The intial planet-disk interaction form bumps and grooves in the gap profiles
 which can potentially be unstable due to axisymmetric instabilities. The
 generalized local axisymmetric stability condition is the Solberg-Hoiland
 criterion,
\begin{equation}
 \kappa^2+N^2 \geq 0 $ where $ N^2=\frac{1}{\Sigma} \frac{\partial P}{\partial r} \left(\frac{1}{\Sigma} \frac{\partial \Sigma}{\partial r}-\frac{1}{\gamma P} \frac{\partial P}{\partial r}  \right)
\end{equation}
where $N$ is the Brunt-V\"ais\"al\"a frequency for the disk due to entropy
 variation which is non-zero for non-adiabatic models such as being considered.
 At  $t=30P_0$ the outer gap edge of $r=2.5r_h$,  where the RWI is excited the
 criterion value, is found to be approximately $0.002$ for all $\tilde\beta$
 which is the local minimum in the gap region which indicates local
 axisymmetric stability of gap. The Solberg-Hoiland criteria is similary
 satisfied for the entire disk and for all times within the simulation length
 with little change of values. Thus for all cases of $\tilde\beta$ the
 planet-induced gaps formed are stable to axisymmertic instabilities.

\subsection{Non-axisymmetric instability}\label{linear}
%{\bf present `planet-off' simulations. one `fourier mode v.s. time'
%  plot to show growth of linear instability. do an adiabatic case (or
%  extremely long cooling time) to see the effect of heated gap edge
%  (i.e. temp doesn't go back to t=0 value too quickly)
%  compare growth rate and dominant
%  m as function of beta (table). 2D figs to contrast (also used to
%  show it's the minimum in generalized pv that goes unstable.    
%  result: increasing cooling time makes the gap
%  more stable, and favors lower m. note: the `basic state' should be
%  the system at t=30 after azimuthal average. linear results should
%  have small perturbations. 
%}

%After the azimuthal averaging, the small pertibation excites rapidly
%growing instabilities by the RWI.

In all cases we observe growth of non-axisymmetric structures after the
planet potential is switched off and the system subject to random
perturbations. An example is shown in Fig. \ref{linearmodes} for 
$\tilde{\beta}=10$. We charcterize these
modes with an azimuthal wavenumber $m$ and growth rate $\gamma(m)$ as defined by
Eq.~\ref{fouriertransform}---\ref{growth} with radial range of $[2r_h,5r_h]$. Table \ref{modetable}
lists the growth rates for the three cooling
times, measured during linear growth.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=1.2cm
    0cm 0cm 0cm]{figures/linear_stability}
  \caption{Evolution of azimuthal Fourier modes of disc surface
    density, non-dimenionlized by the axisymmetric background keplerian mode
 $\Sigma_0(t=0)$ for the
    `planet-off' simulations with $\tilde{\beta}=10$. Colours correspond
    to different $m$ values. The $m=3$ component is the fastest growing
    mode during linear growth and has a corresponding
    $\gamma=0.017\Omega(r_p)$.\label{linearmodes}}
% {\bf is the $\Sigma_0$ at t=0?} }  
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Table \ref{modetable} show that as
$\tilde{\beta}$ was increased from $ 0.1\rightarrow10$ the dominant
azimuthal Fourier mode decreased from $ m=9\rightarrow3$ and the
respective growth rate decreased from $ \gamma/\Omega(r_p)=0.079
\rightarrow 0.017$. However, despite two orders of magnitude increase in the
cooling time, the instability remains dynamical with growth time
$\lesssim 10P_0$. Snapshots of the instability in $r-\phi$ plane for
the different $\tilde\beta$ are shown in Fig \ref{2Dlinear}.

These `planet-off' simulations show that gap edges become more stable with
longer cooling times. This is expected because larger $\tilde{\beta}$
result in hotter gap profiles at $t=30P_0$ with less pronounced
generalized vortensity minima. Stabilization with increased
cooling time is therefore due to a smoother basic state for the
instability, as it is more difficult for the planet to open a gap if
the disc is allowed to heat up.

% which result in hotter gaps since the heating
%due
%This is as expected from the
%intial gap profiles, because 
% and the gap forming criteria discussed in the
%previous section.

\begin{figure}
  \centering
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity50}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity67}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity75}
  }  
  \caption{Generalized vortensity perturbation (relative to $t=0$) for
    cases of $\tilde{\beta}=0.1,1,10$ (left,middle,right) during
    the growth of non-axisymmetric modes. The planet potential has
    been switched off.  The number of vortices
    decrease as $\tilde{\beta}$ increases. Note that snapshots are
    taken later because it takes longer for the vortices to grow and
    become visible with increasing cooling time. 
    %    correspondingly increase to capture the longer growth phases of
    %    the instabilities. 
    \label{2Dlinear} {\bf I'm curious to know if plots clearer if
      perturbations are measured relative to t=30? remember t=30 is
      our reference state for instability. But we'll probably keep
      these for consistency (later non-linear plots are all measure
      with respect to t=0).}} 
\end{figure}

\subsubsection{Nearly adiabatic discs}
\label{adiabatic_section}
%{\bf a case with extrmely long cooling time (or adiabatic run) in
%  order to look at effect of heated gap edge.  prelim result: not
%  important, at t=30, gap edge only heats to $H/r\simeq0.06$ even for
%  purely adiabatic disk. change in $h$ not important for linear
%  perturbations (but important for setting up the basic state). 
%}
The above `planet-off' simulations are not formally linear
stability calculations, because the cooling time is shorter
than the instability growth time, 
%
%as we do not capture the effects of a heated
%disk on the instability growth. 
$t_c<\gamma^{-1}$.  
Thus the disk cools back to its intial value of $h=0.05$
before or during the instability growth, implying we do not have a
steady basic state to formulate a standard linear stability
problem. %This means the disc temperture of $h=0.05$ in the the above cases

In order to perform a proper linear stability analysis and capture the
effect of a heated gap edge during instability growth, we ran a simulation  with
$\tilde{\beta}=100$, corresponding to an almost adiabtaic disc.  
%To compare our 'planet-off' analysis with a true
%linear stability analysis 
In this simulation the cooling rate is slow enough that the gap 
temperature profile (e.g. middle panel of Fig. \ref{intial1D}) changes
only marginally over the instability growth timescale. %heat is
                                %retained/maintained.

We find very similar gap profiles and mode growth rates for
$\tilde{\beta}=100$ as with $\tilde{\beta}=10$. The disc heats up to
values $h\simeq0.06$ in the nearly adiabatic case. This is close to
the original temperature of $h=0.05$, so linear growth rates are not expected
to change significantly %according to the linear theory presented in
\citep{li00}.

According to \cite{li00}, increasing $h$ increases linear growth rates
of the RWI because it is pressure-driven. However, in the case 
of disc-planet interaction, increasing $h$ has a stabilizing effect
through the gap profile because it results in smoother gap
edges. The fact that we observe smaller growth rates as $h$ is
increased indicate that for planetary gaps, the importance of $h$ on
the \emph{linear} RWI is through the setting the gap profile (i.e. basic
state for the instability).

%results in shallower gaps
%which is expected to be

%the difference in growth rates between this and the
%original disc temperature of $h=0.05$ changes almost insignificantly
%as shown by linear theory \citep{li00}. 
%Although linear
%theory predicts higher growth rates with increasing $h$

%This indicate that a
%continuously hot gap edge is not as important for development of
%vortices as the heating effects on the formation of the intial gap
%state.

{\bf
If there's time, do low resolution planet off simulations but keep the planet on
until $t=40P_0$ or $50P_0$. Look at stability properties when the gaps
are generally deeper (including the nearly adiabatic case). Although
with the planet instability may already appear by $t=40P_0$, the
azimtuhal average performed prior to perturbations will erase it. 
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Non-linear evolution of
  gap-edge vortices with finite cooling time}\label{nonlinear} 
%{\bf main fig: vortex amplitude v.s. time for diff beta. enough data
%  to plot vortex lifetime v.s. beta? table: 
%  averaged quantities over quasi-steady state: aspect-ratio (to
%  compare with fu at al), rossby number (vortex strength
%  v.s. cooling?), maybe alpha visc. vortex size: visible difference?  
%  only inviscid cases. describe evolution of one case. main
%  conclusion: longer vortex lifetime with increasing cooling time (up
%  to some optimal timescale). vortex death: induced-shock and/or
%  smoothing the gap edge. describe simulation setup, resolution?
%  should mention that results consistent with lower-resolution prelim
%  runs. maybe torques? 
%}

We now consider long term simulations of gap-edge vortices for
$\tilde{\beta}=0.1,0.5,1,5,10$. The planet potential is kept on
throughout.  
% up to a total of
%$2000P_0$. 
%Planets were left free to interact dynamiclly with the disk
%and vortices after $t=30P_0$ as apposed to previous section.  
We employ a grid with $(N_r,N_{\phi})=(512,1024)$ 
in order for these simulations to be computationally feasible. We also use a larger
disc with $r_{\mathrm{out}}=45r_{\mathrm{in}}$ to minimize boundary
effects on vortex evolution. We apply an open boundary at
$r_\mathrm{out}$.

Preliminary simulations with same numerical scheme but lower resolution of
 $(N_r,N_{\phi})=(256,512)$ where also done for the above $\tilde\beta$ values
with the inclusions of $\tilde{\beta}=0.25,0.75,2.5,7.5$ as well.
Results and vortex behaviour are convergent with
higher resolution simulations as discussed in this section.
 {\bf mention lower resolution preliminary runs -
  similar results to those presented here (?). want to demonstrate convergence}

%An open boundry
%condition was used in the outer edge of
% allowing significant room for
%linblad resonances of the vortices.

\subsection{Generic evolution}

The linear growth of the RWI and vortex-formation is followed by 
vortex merging. Merging timescales are independent of $\tilde\beta$ value,
 and by $70P_o$ only one vortex remains as the growth of these modes are now
 supported by the continuous driving of the planet.
%i.e. only the $m=1$ mode remains. 
{\bf is vortex merging affected by $\tilde{\beta}$? e.g. timescale
  taken to reach one vortex as a function of cooling}
The evolution of the amplitude of the $m=1$ surface density mode
 over the radial range of $[2r_h,10r_h]$ is shown in Fig~\ref{lifetimeplot}
 for different $\tilde\beta$.
% {\bf should state (also for planet-off runs)
 % what the radial range is used for averaging the Fourier amplitudes}

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
    0cm 0cm 1cm]{figures/longterm_stability}
  \caption{Log plot of the $m=1$ mode non-dimenionlized by the $m=0$
    background mode for longterm non-linear simulations corrsponding
    to vortex lifetime. The longest living mode correpsonds to
    $\tilde\beta=5.0$ (the red line). \label{lifetimeplot}} 
\end{figure}

In all cases the system remains in a quasi-steady state for
$\gtrsim800P_0$ with a single vortex circulating 
the outer gap edge at the local Keplerian  
frequency. Fig. \ref{Vortex2D} shows a typical plot of the relative 
surface density perturbation in this state. During this stage, the 
vortex grows in intensity. As the vortex forms it has a
 value of $Ro\approx-0.1$ for all $\tilde\beta$
 indicating anti-cyclonic
motion. As the $m=1$ amplitude grows the relative spin of the vortex
continuously grows along with it, reaching values of
$Ro\approx-0.4$ characteristicly for all cooling rates. Relative surface
 density pertibations at the center of vortices can reach values of 7 times the
 intial density for all cases of $\tilde\beta$ in this quasi-steady state and
 can be found as high as 9 times for the $\tilde\beta=1.0$ case.
 The gap profiles apply a surface density boost in the vortex region as the
 vortices are located at the gap edge maxima
 but the values associated with this boost have relative density pertibations
 of under 1 and thus are negligibe compared to the max amplitudes of the vortices.
 Vortex growth is attributed to
the continuous generation of vorticity by the planet-disc interactions. 
This is supported by the evidence that in long term simulations where the
planet is removed such as in section \ref{nonlinearplanetoff} the amplitude of
 the vortices are not seen to contiunously grow without the planet influence.

{\bf for what cooling are these values quoted for? do
  these values depend on cooling?}
 {\bf is the evidence for this the fact that growth is not observed in
  planet-off simulations?}

{\bf note that there
  are two origins for the surface density boost relative to $t=0$: 1) gap
  formation, which increases the surface density at the location of
  the instability (since they happen at surface density max., and 2)
  vortex instability). what is the surface density boost relative to
  the gap profile?}

Interestingly, in most cases we observe a sudden drop in 
the $m=1$ amplitude within the simulation time, which occurs over
timescales of $50P_0$, i.e. much shorter than the duration of the quasi-steady
state. {\bf is this dissipation timescale independent of cooling? No but will wait for simulation to finish before commenting}.  
%Eventually
%the $m=1$ amplitudes are also characterized by 
After these characteristic drops, the instability is never seen to
reform at such large overdensities and surface density bumps are never larger
than that due to the planetary wakes. We designate the time of these
amplitude drops as the lifetime 
of the quasi-steady vortex.
The instability is not seen to emerge again after the intial vortex dissipation
 due to the smoothening out of the gap edge during dissipation by the
 transfer of density from the once tightly held vortex region to the surrounding
 disk and gap.  
{\bf question: does instability
  actually happen again? if it does, but just weaker, that may be
  because the first vortex has smoothed out the gap edge, so
  instabilities following that would be weaker.}

\begin{figure}
  \includegraphics[width=\linewidth,height=\linewidth]{figures/vortex2D}
  \caption{Relative surface density perturbation for the
    $\tilde\beta=0.1$ case during quasi-steady state with a single
    vortex. The plot for other values of the cooling time $\tilde{\beta}$ are
    similar.% {\bf is this plot truncated? in the text it says the
      %domain goes to $r=45$?}
    %The large scale non-axisyemetric $m=1$ overdensity can be
    %seen. 
    \label{Vortex2D}} 
\end{figure}

\subsection{Non-linear instability without planet influence} \label{nonlinearplanetoff}
Non-linear observations were also made for the `planet-off' simulations.
After the linear growth phase of the vortices, vortex merging takes hold but
 now on timescales of up to $150P_0$ as the vortices grow as linear
 instabilities discussed previously. The vortex merging time is depedendent on
 the growth rates of the modes and how long they take to saturate. Similar
 quasi-steady vortices form but the long-term behaviour is significantly
 different without the planet as seen in Fig.\ref{planetofflifetimeplot}.
 The quasi-steady vortices formed have significantly lower density pertibations
 and now slowly decrease in amplitute in time as apposed to previously
 where their growth was supported by planet interactions. The vortices simply
 dissipate viscously and no longer undergo the quick dissipation that
 characterized the dissipation as earlier. The dissipation rate is dependent
 on the cooling rate with viscous dissipation decreasing with larger
$\tilde\beta$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
    0cm 0cm 1cm]{figures/longterm_planetoff}
  \caption{Similar plot as Fig. \ref{lifetimeplot} but with `planet-off' simulations. Vortices dissipate viscously without planet driving influence} \label{planetofflifetimeplot}
\end{figure}

{\bf
\subsection{Additional analysis on vortex dissipation}
Conjecture: vortex death due to vortex shocking the system and
dissipating its energy. discuss one case. here we want to explain or
suggest why vortices die} 
%Rossby values have dynamics that are coupled with the lifetime and
%$m=1$ amplitude of the vortices.

As the vortex dies off as described above its associated Rossby number
of quickly approaches $0$. Gradients of surface density are seen to
spike up in wake-like features around the vortex during dissipation. 
{\bf show
  this. e.g. vortex before, during and after death. maybe show
  snapshots of surface density perturbation (vortex) compared with
  $\nabla\Sigma/\Sigma$ (vortex shocks). demonstrate vortex shape
  changes and shocks develop prior ot death. state whether this is
  generic for all coolings}.  
In addition, we also observed large increases in the Mach number near
the vortex during dissipation, which indicate susceptibility to
shocks. {\bf show mach number plot, or quote values. relevant mach
  number is fluid flow speed with respect to 
  vortex bulk motion around the disc. does the typical mach number
  near the vortex just before its death depend on cooling?} 
Thus, the sudden drop in vortex amplitude may  
be a result of the development of associated spiral shocks dissipating
its energy \citep[cf.][who 
  suggested damping due to planet-induced shocks]{fu14}.

{\bf note that fu et al suggested vortex damping by the planet
  shocks. here, we are suggesting vortex dissipates energy by shocking
  the disk. our evidence is that we observe vortex growth before it
  dissipates, and spiral shocks associated with the vortex always
  develop before vortex death. once it starts shocking the disc it
  loses energy. if the planet-induced shocks were responsible, death
  should be more gradual. can we think of other evidences that it's
  not the planet shock that's responsible?                    
}

Vortex growth during the quasi-steady state also acts to smooth out
the gap edge. 
%Of note is also that the large overdensity the vortex
%creates during it's quasi-steady state distorts the background
%disk. 
{\bf show this (smoothening of gap edge due to vortex growth),
  e.g. general vortensity evolution or surface density 
  evolution}. 
This is because the vortex has an associated viscosity {\bf quote a
  typical value of alpha viscosity associated with vortex}. This acts
against gap-opening by the planet, so that the condition for the RWI
becomes less favourable as the vortex grows. {\bf self-limiting
  behaviour?}

\subsection{Vortex lifetimes as a function of cooling rate}

{\bf describe dependency on
cooling. Do we have enough data to have a `lifetime v.s. beta'
  plot? here we want to explain why vortex lives longer with longer
  cooling, but very long cooling results in shorter lifetimes. 
  Consider the following quantities as a function of cooling:
  initial strength of the single vortex; growth rate of single vortex
  strength from its formation to just before death; ampiltude of
  vortex just before death; disc temperature in the vortex region.

  we expect the initial $m=1$ vortex (after merging) to be weaker with
  increasing cooling time because the instability is weaker. (see
  meheut et al 2013, mnras, 430, 1988 who show that the initial vortex
  amplitude after linear growth is correlated with the linear growth
  rate).

  suppose the vortex shocks when it reaches a certain amplitude (which
  probably increases with longer cooling rate because it's more
  difficult to shock a hotter disc). as we
  increase the cooling time, the initial $m=1$ vortex is weaker, it
  also grows more slowly because gap-opening is opposed by a hotter
  disc. the above implies that it takes longer for the vortex reach
  the shock-inducing amplitude and dissipate.

  but the above is not consistent with non-monotonic behaviour, which
  predicts the $\tilde{\beta}=10$ vortex to survive longest. this is
  because we expect the initial vortex to be weakest (check) which
  means it should take the longest to grow and shock. check
  whether there's anything special about this case. one possibility is
  that vortex growth during the quasi-steady state is actually faster
  because of the raised temperature (li et al 2000 show increased
  temperature favours RWI - possibiliy evidenced here by a stronger
  vortex before death than other cases)

  perhaps have a case of $\tilde{\beta}=7.5$ as further evidence of
  non-monotonic behaviour
}

Fig. \ref{lifetimeplot} shows that increasing $\tilde{\beta}$
increases the lifetime of the vortex up to a critical 
value $\tilde{\beta}=5.0$ for which the vortex lasted beyond the
simulation time of $2000P_0$. This is consistent with vortex
dissipation when it induce shocks. Our `planet-off' simulations
indicate weaker vortices are formed initially with increased cooling
times. After merging, the growth of the single vortex, mediated by disc-planet
interaction, is expected to be slower with increased cooling time
because gap-opening becomes more difficult in a hotter
disc. Furthermore, the increased sound-speed suggest the vortex should
reach larger amplitudes in order to induce shocks. 
These expectations
imply that it takes longer for the vortex to induce shocks and
dissipate, and hence longer lifetimes with increasing
$\tilde{\beta}$.

However, Fig. \ref{lifetimeplot} shows that increasing the cooling
time further to $\tilde{\beta}=10$ actually results in a
\emph{shorter} vortex lifetime. Notice for $\tilde{\beta}=10$ the vortex growth 
during the quasi-steady state is faster than for
$\tilde{\beta}=5$. This may be attributed to the higher temperature
reached in the former case ($h\simeq 0.08$). {\bf temperature for
  $\tilde{\beta}=5$?}, since \cite{li00} showed that the RWI grows
faster with increased disc temperature. At long cooling times, this
effect may dominate over the weaker gap-opening effect at higher $h$.

Thus, unlike our `planet-off' simulations we find that there is
a cooling time $t_c$ which maximises the vortex lifetime. In the long term
 'planet-off' simulations we find that the vortices dissipate very gradually and visciously. {\bf to make
this comparison we need to discuss vortex lifetimes for planet-off
simulations as well} 
This
non-monotonic behaviour was also reported by \cite{fu14}, who
performed isothermal disc-planet simluations with different fixed
aspectratios.   
%Cooling rates with
%$\tilde\beta>5.0$ showed lifetimes that were well below the max
%lifetime. 
By analyzing the region
around the vortex {\bf define this region} we find the
$\tilde\beta=5.0$ case, with the longest vortex lifetime, has
$h\approx0.06$. This value of $h$ was also found to maximize vortex
lifetimes in the isothermal simulations of \cite{fu14}.

%be the optimum lifetime for long-term simulations of
%isothermal disks with varying intial temperature profiles $h$.

%Competing effects of gap stability with $t_{\mathrm{coo}l}$ result in
%the the non-montonic form of the lifetime. As the $\tilde\beta$
%increases so does the stability of the gap edge as seen in
%section~\ref{linear}, however as $\tilde\beta$ increases so does the
%$c_{\mathrm{iso}}$ which is known from theory to effect the growth of
%instability \citep{li00}. Thus increasing $\tilde\beta$ increases the
%sound speed $c_{\mathrm{iso}}$ and likewise the gap instability up to
%a certain point in which the effects of the more stable gap edge
%becomes a significant effect.

%The correlation of growth rate and sound
%speed is non-negligible as in previous `planet-off' case as the disk
%now can reach values of $h\simeq0.08$ for very slow cooling and the
%importance of the effect can be imagined to accumulate over the now
%considerably longer timescale. These effects dictate when then the
%vortex reaches the critical value in which it shocks the system that
%subsequently causes disspation and hence the lifetime fo the vortex.

### 2014-10-20 sims/fargoHOLE/src_m1/Force.c

@@ -1,0 +1,213 @@
+#include "mp.h"
+
+extern boolean OpenInner, NonReflecting;
+extern Pair DiskOnPrimaryAcceleration;
+
+Force *AllocateForce (dimfxy)
+int dimfxy;
+{
+  int i;
+  Force *force;
+  real *globalforce;
+  force  = (Force *) prs_malloc (sizeof(Force));
+  globalforce = (real *) prs_malloc (sizeof(real) * 4 * dimfxy);
+  for (i = 0; i < 4*dimfxy; i++)
+    globalforce[i] = 0.;
+  force->GlobalForce = globalforce;
+  return force;
+}
+
+void FreeForce (force)
+     Force *force;
+{
+  free (force->GlobalForce);
+}
+
+void ComputeForce (force, Rho, x, y, rsmoothing, mass, dimfxy)
+     Force *force;
+     PolarGrid *Rho;
+     real x, y, rsmoothing, mass;
+     int dimfxy;
+{
+  int i, j, l, ns, k;
+  real xc, yc, cellmass, dx, dy, distance, d2, dist2, rh, a;
+  real InvDist3, hill_cut, hillcutfactor;
+  real *fxi, *fxo, *fyi, *fyo;
+  real *localforce, *globalforce;
+  real *dens, *abs, *ord;
+  fxi = (real *) prs_malloc (sizeof(real) * dimfxy);
+  fxo = (real *) prs_malloc (sizeof(real) * dimfxy);
+  fyi = (real *) prs_malloc (sizeof(real) * dimfxy);
+  fyo = (real *) prs_malloc (sizeof(real) * dimfxy);
+  localforce = (real *) prs_malloc (sizeof(real) * 4 * dimfxy);
+  globalforce = force->GlobalForce;
+  ns = Rho->Nsec;
+  dens = Rho->Field;
+  abs = CellAbscissa->Field;
+  ord = CellOrdinate->Field;
+  a = sqrt(x*x+y*y);
+  rh = pow(mass/3., 1./3.)*a+1e-15;
+  for ( k = 0; k < dimfxy; k++ ) {
+    fxi[k] = 0.;
+    fxo[k] = 0.;
+    fyi[k] = 0.;
+    fyo[k] = 0.;
+  }
+  for ( k = 0; k < 4*dimfxy; k++ ) {
+    localforce[k] = 0.;
+    globalforce[k] = 0.;
+  }
[…]

### 2014-10-20 sims/fargoHOLE/src_m1/Init.c

@@ -1,0 +1,131 @@
+#include "mp.h"
+
+extern boolean Restart;
+extern int     NbRestart;
+
+void ReadfromFile (array, fileprefix, filenumber)
+     PolarGrid *array;
+     char *fileprefix;
+     int filenumber;
+{
+  int nr, ns, c, foo=0;
+  real *field;
+  char name[256];
+  FILE *input;
+  /* Simultaneous read access to the same file have been observed to
+     give wrong results. */
+  /* A sequential reading is imposed below. */
+  /* If current CPU has a predecessor, wait for a message from him */
+  if (CPU_Rank > 0) MPI_Recv (&foo, 1, MPI_INT, CPU_Prev, 10, MPI_COMM_WORLD, &stat);
+  sprintf (name, "%s%s%d.dat", OUTPUTDIR, fileprefix, filenumber);
+  input = fopen (name, "r");
+  if (input == NULL) {
+    fprintf (stderr, "WARNING ! Can't read %s. Restarting with t=0 settings.\n", name); 
+    if (CPU_Rank < CPU_Highest) MPI_Send (&foo, 1, MPI_INT, CPU_Next, 10, MPI_COMM_WORLD);
+    return;
+  }
+  field = array->Field;
+  nr = array->Nrad;
+  ns = array->Nsec;
+  for (c = 0; c < IMIN; c++) {
+    fread (field, sizeof(real), ns, input); 
+    /* Can't read at once in order not to overflow 'field' */
+  }
+  fread (field, sizeof(real), nr*ns, input);
+  fclose (input);
+  /* Next CPU is waiting. Tell it to start now by sending the message
+     that it expects */
+  if (CPU_Rank < CPU_Highest) MPI_Send (&foo, 1, MPI_INT, CPU_Next, 10, MPI_COMM_WORLD);
+  MPI_Barrier (MPI_COMM_WORLD);	/* previous CPUs do not touch anything
+				   meanwhile */
+}
+
+void InitLabel (array, sys)
+     PolarGrid *array;
+     PlanetarySystem *sys;
+{
+  int nr, ns, i, j, l;
+  real xp, yp, rp;
+  real x, y, angle, distance, rhill;
+  real *field;
+  field = array->Field;
+  nr = array->Nrad;
+  ns = array->Nsec;
+  xp = sys->x[0];
+  yp = sys->y[0];
+  rp = sqrt ( xp*xp + yp*yp );
[…]

### 2014-10-20 sims/fargoHOLE/src_m1/Pframeforce.c

@@ -1,0 +1,357 @@
+#include "mp.h"
+
+extern boolean AllowAccretion, Corotating, Indirect_Term, Cooling;
+extern Pair DiskOnPrimaryAcceleration;
+static Pair IndirectTerm;
+static real q0[MAX1D], q1[MAX1D], PlanetMasses[MAX1D];
+static real vt_int[MAX1D], vt_cent[MAX1D];
+
+void ComputeIndirectTerm () {
+  IndirectTerm.x = -DiskOnPrimaryAcceleration.x;
+  IndirectTerm.y = -DiskOnPrimaryAcceleration.y; 
+  if (Indirect_Term == NO) {
+    IndirectTerm.x = 0.0;
+    IndirectTerm.y = 0.0;
+  }
+}
+/* Below : work in non-rotating frame */
+/* centered on the primary */
+void FillForcesArrays (sys, Rho, Energy)
+     PlanetarySystem *sys;
+     PolarGrid *Rho, *Energy;
+{
+  int i, j, l, nr, ns, k, NbPlanets;
+  real x, y, angle, distance, distancesmooth, iplanet;
+  real xplanet, yplanet, RRoche,smooth, mplanet;
+  real PlanetDistance, *Pot, pot, smoothing;
+  real InvPlanetDistance3, InvDistance;
+  real xbin, ybin, mbin, distbin, Invdistbin3;
+  Pot= Potential->Field;
+  nr = Potential->Nrad;
+  ns = Potential->Nsec;
+  NbPlanets = sys->nb;
+
+  /* Indirect term star on gas here */
+  ComputeIndirectTerm ();
+
+#pragma omp parallel for
+  for (i = 0; i < (nr+1)*ns; i++) Pot[i] = 0.0;
+  /* -- Gravitational potential from planet on gas -- */
+  for (k = 0; k < NbPlanets; k++) {
+    xplanet = sys->x[k];
+    yplanet = sys->y[k];
+    mplanet = sys->mass[k]*MassTaper;
+    PlanetDistance = sqrt(xplanet*xplanet+yplanet*yplanet);
+    InvPlanetDistance3 =  1.0/PlanetDistance/PlanetDistance/PlanetDistance;
+    RRoche = PlanetDistance*pow((1.0/3.0*mplanet),1.0/3.0);
+    if (RocheSmoothing)
[…]

### 2014-10-20 sims/fargoHOLE/src_m1/SideEuler.c

@@ -1,0 +1,537 @@
+#include "mp.h"
+
+extern boolean OpenInner, NonReflecting, OuterSourceMass, Evanescent;
+extern boolean SelfGravity, SGZeroMode, Adiabatic;
+extern Pair DiskOnPrimaryAcceleration;
+extern int dimfxy;
+real Hp0, Hg0, Ht0;
+
+real GasTotalMass (array)
+     PolarGrid *array;
+{
+  int i, j, ns;
+  real *density, total = 0.0, fulltotal=0.0;
+  ns = array->Nsec;
+  density = array->Field;
+  if (FakeSequential && (CPU_Rank > 0)) 
+    MPI_Recv (&total, 1, MPI_DOUBLE, CPU_Rank-1, 0, MPI_COMM_WORLD, &stat);
+  for (i = Zero_or_active; i < Max_or_active; i++) {
+    for (j = 0; j < ns; j++) {
+      total += Surf[i]*density[j+i*ns];
+    }
+  }
+  if (FakeSequential) {
+    if (CPU_Rank < CPU_Number-1)
+      MPI_Send (&total, 1, MPI_DOUBLE, CPU_Rank+1, 0, MPI_COMM_WORLD);
+  }
+  else
+    MPI_Allreduce (&total, &fulltotal, 1, MPI_DOUBLE, MPI_SUM, MPI_COMM_WORLD);
+  if (FakeSequential) {
+    MPI_Bcast (&total, 1, MPI_DOUBLE, CPU_Number-1, MPI_COMM_WORLD);
+    fulltotal = total;
+  }
+  return fulltotal;
+}
+
+real GasMomentum (Density, Vtheta)
+     PolarGrid *Density, *Vtheta;
+{
+  int i,j,ns;
+  real *density, *vtheta, total = 0.0, fulltotal=0.0;
+  ns = Density->Nsec;
+  density = Density->Field;
+  vtheta = Vtheta->Field;
+  if (FakeSequential && (CPU_Rank > 0)) 
+    MPI_Recv (&total, 1, MPI_DOUBLE, CPU_Rank-1, 2, MPI_COMM_WORLD, &stat);
+  for (i = Zero_or_active; i < Max_or_active; i++) {
+    for (j = 1; j < ns; j++) {
+      total += Surf[i]*(density[j+i*ns]+density[j-1+i*ns])*Rmed[i]*(vtheta[j+i*ns]+OmegaFrame*Rmed[i]);
[…]

### 2014-10-20 sims/fargoHOLE/src_m1/global.h

@@ -1,0 +1,45 @@
+int CPU_Rank;
+int CPU_Number;
+boolean CPU_Master;
+int CPU_Next, CPU_Prev, CPU_Highest;
+/* ------------------------------------- */
+/* Variables specific to fftw mesh split */
+/* ------------------------------------- */
+int CPU_Friend, CPU_NoFriend;
+real *dens_friend;
+real *SGP_buffft_Accr_friend, *SGP_buffft_Acct_friend;
+real *ffttohydro_transfer, *ffttohydro_transfer_friend;
+int local_Nx, local_i_start, local_i_start_friend, total_local_size_friend, local_Nx_friend;
+int local_Ny_after_transpose, local_j_start_after_transpose;
+int total_local_size, ifront, Zero_or_active_friend;
+int transfer_size, transfer_size_friend;
+int hydro_totalsize, active_hydro_totalsize, active_hydro_totalsize_friend;  
+/* ------------------------------------- */
+int IMIN;
+int IMAX;
+int Zero_or_active;
+int Max_or_active;
+int One_or_active;
+int MaxMO_or_active;		/* MO: Minus One */
+int GLOBALNRAD;
+real Rinf[MAX1D], Rsup[MAX1D], Rmed[MAX1D], Surf[MAX1D];
+real InvRmed[MAX1D], InvSurf[MAX1D], InvDiffRmed[MAX1D];
+real InvDiffRsup[MAX1D], InvRinf[MAX1D], Radii[MAX1D], GlobalRmed[MAX1D];
+real SigmaMed[MAX1D], SigmaInf[MAX1D], MassTaper, HeatTaper;
+real EnergyMed[MAX1D], QplusMed[MAX1D], CoolingTimeMed[MAX1D];
+real OmegaFrame, PhysicalTime, PhysicalTimeInitial;
+int TimeStep;
+real HillRadius, mdcp, mdcp0, exces_mdcp;
+real GLOBAL_bufarray[MAX1D];
+boolean Merge, AdvecteLabel, FakeSequential, MonitorIntegral, debug, OnlyInit;
+boolean	GotoNextOutput, StoreSigma, StoreEnergy, ViscosityAlpha, RocheSmoothing;
+boolean CentrifugalBalance, ExcludeHill, SloppyCFL;
+MPI_Status stat;
+PolarGrid *CellAbscissa, *CellOrdinate;
[…]

### 2014-10-20 sims/fargoHOLE/src_m1/global_ex.h

@@ -1,0 +1,50 @@
+/********************************/
+/*                              */
+/* This file is created         */
+/* automatically during         */
+/* compilation. Do not edit.    */
+/* See perl script              */
+/* "varparser.pl" for details   */
+/*                              */
+/********************************/
+extern int CPU_Rank;
+extern int CPU_Number;
+extern boolean CPU_Master;
+extern int CPU_Next, CPU_Prev, CPU_Highest;
+extern int CPU_Friend, CPU_NoFriend;
+extern real *dens_friend;
+extern real *SGP_buffft_Accr_friend, *SGP_buffft_Acct_friend;
+extern real *ffttohydro_transfer, *ffttohydro_transfer_friend;
+extern int local_Nx, local_i_start, local_i_start_friend, total_local_size_friend, local_Nx_friend;
+extern int local_Ny_after_transpose, local_j_start_after_transpose;
+extern int total_local_size, ifront, Zero_or_active_friend;
+extern int transfer_size, transfer_size_friend;
+extern int hydro_totalsize, active_hydro_totalsize, active_hydro_totalsize_friend;  
+extern int IMIN;
+extern int IMAX;
+extern int Zero_or_active;
+extern int Max_or_active;
+extern int One_or_active;
+extern int MaxMO_or_active;		/* MO: Minus One */
+extern int GLOBALNRAD;
+extern real Rinf[MAX1D], Rsup[MAX1D], Rmed[MAX1D], Surf[MAX1D];
+extern real InvRmed[MAX1D], InvSurf[MAX1D], InvDiffRmed[MAX1D];
+extern real InvDiffRsup[MAX1D], InvRinf[MAX1D], Radii[MAX1D], GlobalRmed[MAX1D];
+extern real SigmaMed[MAX1D], SigmaInf[MAX1D], MassTaper, HeatTaper;
+extern real EnergyMed[MAX1D], QplusMed[MAX1D], CoolingTimeMed[MAX1D];
+extern real OmegaFrame, PhysicalTime, PhysicalTimeInitial;
+extern int TimeStep;
+extern real HillRadius, mdcp, mdcp0, exces_mdcp;
+extern real GLOBAL_bufarray[MAX1D];
+extern boolean Merge, AdvecteLabel, FakeSequential, MonitorIntegral, debug, OnlyInit;
[…]

### 2014-10-20 sims/fargoHOLE/src_m1/main.c

@@ -1,0 +1,226 @@
+#include "mp.h"
+
+boolean         TimeToWrite, Restart = NO, OpenInner = NO;
+int             TimeStep = 0, begin_i = 0, NbRestart = 0, verbose = NO;
+int             dimfxy = 11;
+static int      InnerOutputCounter=0, StillWriteOneOutput;
+extern real     LostMass;
+extern boolean  Corotating;
+extern boolean  SelfGravity, SGZeroMode, Adiabatic;
+real            ScalingFactor = 1.0;
+
+int
+main(argc, argv)
+     int argc;
+     char *argv[];
+{
+  PolarGrid   *gas_density;
+  PolarGrid   *gas_v_rad; 
+  PolarGrid   *gas_v_theta; 
+  PolarGrid   *gas_energy; 
+  PolarGrid   *gas_label;
+  int          i;
+  real         foostep = 0., dummy=0.;
+  boolean      disable = NO, TimeInfo = NO, Profiling = NO;
+  boolean      Stockholm = NO, updatevelocities = NO;
+  TimeProcess  t_Hydro;
+  char         ParameterFile[256];
+  PlanetarySystem *sys;
+  Force *force;
+  
+  MPI_Init (&argc, &argv);
+  MPI_Comm_rank (MPI_COMM_WORLD, &CPU_Rank);
+  MPI_Comm_size (MPI_COMM_WORLD, &CPU_Number);
+  CPU_Master = (CPU_Rank == 0 ? 1 : 0);
+  setfpe ();  /* Control behavior for floating point
+		 exceptions trapping (default is not to do anything) */
+  if (argc == 1) PrintUsage (argv[0]);
+  strcpy (ParameterFile, "");
+  for (i = 1; i < argc; i++) {
+    if (*(argv[i]) == '-') {
+      if (strspn (argv[i], "-secndovtpfamzib0123456789") != strlen (argv[i]))
+	PrintUsage (argv[0]);
+      if (strchr (argv[i], 'n'))
+	disable = YES;
+      if (strchr (argv[i], 'e'))
+	Stockholm = YES;
+      if (strchr (argv[i], 'v'))
+	verbose = YES;
+      if (strchr (argv[i], 't'))
+	TimeInfo = YES;
+      if (strchr (argv[i], 'c'))
+	SloppyCFL = YES;
+      if (strchr (argv[i], 'p'))
+	Profiling = YES;
+      if (strchr (argv[i], 'd'))
+	debug = YES;
+      if (strchr (argv[i], 'b'))
[…]

### 2014-10-20 sims/fargoHOLE/src_m1/param_noex.h

@@ -1,0 +1,90 @@
+/********************************/
+/*                              */
+/* This file is created         */
+/* automatically during         */
+/* compilation. Do not edit.    */
+/* See perl script              */
+/* "varparser.pl" for details   */
+/*                              */
+/********************************/
+real	DT;
+real	SIGMA0;
+int	NINTERM;
+int	NTOT;
+char	OUTPUTDIR[512];
+char	OPENINNERBOUNDARY[512];
+char	ADVLABEL[512];
+char	TRANSPORT[512];
+char	PLANETCONFIG[512];
+real	MASSTAPER;
+char	GRIDSPACING[512];
+int	NRAD;
+int	NSEC;
+real	RMIN;
+real	RMAX;
+real	THICKNESSSMOOTHING;
+real	ROCHESMOOTHING;
+real	ASPECTRATIO;
+real	VISCOSITY;
+real	ALPHAVISCOSITY;
+real	SIGMASLOPE;
+real	RELEASERADIUS;
+real	RELEASEDATE;
+real	OMEGAFRAME;
+char	DISK[512];
+char	FRAME[512];
+char	OUTERSOURCEMASS[512];
+char	WRITEDENSITY[512];
+char	WRITEVELOCITY[512];
+char	WRITEENERGY[512];
+char	WRITETEMPERATURE[512];
+char	WRITEDIVV[512];
+char	WRITEQPLUS[512];
+char	INDIRECTTERM[512];
+char	EXCLUDEHILL[512];
+real	IMPOSEDDISKDRIFT;
+real	FLARINGINDEX;
+real	ECCENTRICITY;
+real	CAVITYRADIUS;
+real	CAVITYRATIO;
+real	CAVITYWIDTH;
+real	TRANSITIONRADIUS;
+real	TRANSITIONRATIO;
+real	TRANSITIONWIDTH;
+real	LAMBDADOUBLING;
+char	SELFGRAVITY[512];
+char	CICPLANET[512];
+char	FORCEDCIRCULAR[512];
+char	ZMPLUS[512];
+char	ADIABATIC[512];
+real	ADIABATICINDEX;
+char	COOLING[512];
+real	COOLINGTIME0;
+real	TRACEWIDTH;
+real	OUTTRUNC;
+real	INTRUNC;
+real	FIN;
+real	SGSMOOTH;
+real	DENFLOOR;
[…]

### 2014-10-20 sims/fargoHOLE/src_m1/sg.h

@@ -1,0 +1,39 @@
+/* ------------------------------------- */
+/*       Global variables used for       */ 
+/*   computation of disk self-gravity    */
+/* ------------------------------------- */
+
+int SG_initcounter;      /* Initialization counter */
+
+/* ================================= */
+/* Self-gravity on a polar grid */
+/* ================================= */
+
+real SGP_eps;             /* Smoothing parameter of potential */
+real SGP_rstep;
+real SGP_tstep;
+real SG_aniso_coeff;   /* to mimic non-axisymmetric effects of self-gravity */
+/* with an anisotropic pressure */
+
+/* The plans for 2D and 1D fft computation */
+
+rfftwnd_mpi_plan SGP_fftplan_forward;
+rfftwnd_mpi_plan SGP_fftplan_backward;
+rfftw_plan SGP_fft1Dplan_forward;
+rfftw_plan SGP_fft1Dplan_backward;
+
+/* Arrays and fft-arrays, since (mpi)fft computes in-place transforms... */
+fftw_real *SGP_Sr, *SGP_St;
+fftw_real *SGP_Kr, *SGP_Kt;
+fftw_complex *SGP_Accr, *SGP_Acct;
+real *SG_Accr, *SG_Acct;
+
+/* Buffer arrays needed for the fft computation */
+real *SGP_buffft_Sr, *SGP_buffft_St;
+real *SGP_buffft_Kr, *SGP_buffft_Kt;
+real *SGP_buffft_Accr, *SGP_buffft_Acct;
+
+/* Global axisymetric arrays used in initialization or for zero mode of SG */
+real *GLOBAL_Axidens;
+real *GLOBAL_AxiSGAccr;
+real *GLOBAL_AxiSGAccr0;

### 2014-10-20 sims/sgPLUTO/Src/Chombo/LevelPluto.H

@@ -1,0 +1,164 @@
+#ifdef CH_LANG_CC
+/*
+ *      _______              __
+ *     / ___/ /  ___  __ _  / /  ___
+ *    / /__/ _ \/ _ \/  V \/ _ \/ _ \
+ *    \___/_//_/\___/_/_/_/_.__/\___/
+ *    Please refer to Copyright.txt, in Chombo's root directory.
+ */
+#endif
+
+#ifndef LEVEL_GODUNOV_H
+#define LEVEL_GODUNOV_H
+
+#include "FluxBox.H"
+#include "DisjointBoxLayout.H"
+#include "ExtrapFillPatch.H"
+#include "LevelFluxRegister.H"
+
+#include "PatchPluto.H"
+#include "NamespaceHeader.H"
+
+/// Level Pluto - really a hyperbolic level integrator
+/**
+ */
+class LevelPluto
+{
+public:
+  /// Default constructor
+  /**
+     Object requires define() to be called before all other functions.
+   */
+  LevelPluto();
+
+  /// Destructor
+  /**
+     Destroys all objects created by define(). Passed in data references
+     of define() are left alone.
+   */
+  ~LevelPluto();
+
+  /// Actual constructor.
+  /**
+     Inside the routine, we cast away const-ness on the data members
+     for the assignment.  The arguments passed in are maintained const
+     (coding standards).  a_refine is the refinement ratio between
+     this level and the next coarser level.  For the coarsest level, an
+     empty DisjointBoxLayout is passed in for coarserDisjointBoxLayout.
+   */
+  void define(const DisjointBoxLayout&  a_thisDisjointBoxLayout,
+              const DisjointBoxLayout&  a_coarserDisjointBoxLayout,
+              const ProblemDomain&      a_domain,
+              const int&                a_refineCoarse,
+              const int&                a_level,
+              const Real&               a_dx,
+              const PatchPluto*  const  a_patchPlutoFactory,
+              const bool&               a_hasCoarser,
+              const bool&               a_hasFiner);
+
+  /// Take one timestep for this grid level.
+  /**
+     For the coarsest level, UcoarseOld and UCoarseNew are empty
+     LevelData<FArrayBox> objects.  Also, either UCoarseOld or
+     UCoarseNew might be empty to indicate that t(nf) = t(nc) the
+     one grid is at the current time and interpolation is not required
+     for boundary condition generation.  U must be defined on the same
[…]

### 2014-10-20 sims/sgPLUTO/Src/Chombo/PatchCTU.1D.cpp

@@ -1,0 +1,305 @@
+#include <cstdio>
+#include <string>
+using std::string;
+
+#include "PatchPluto.H"
+#include "LoHiSide.H"
+
+/* *********************************************************** */
+void PatchPluto::updateSolution(FArrayBox&  a_U,
+                                FArrayBox&  Utmp,
+                                FArrayBox&  split_tags,
+                                BaseFab<unsigned char>& a_Flags,
+                                FluxBox&    a_F,
+                                Time_Step   *Dts,
+                                const Box&  UBox,
+                                Grid *grid) 
+/*
+ *
+ *
+ *
+ *
+ ************************************************************* */
+{
+  CH_assert(isDefined());
+  CH_assert(UBox == m_currentBox);
+
+  int nv, ii, jj, kk, nxf, nxb, indf;
+  double ***splitcells;
+  static Data d;
+  static State_1D state;
+  static double **u;
+  Riemann_Solver *Riemann;
+  Index indx;
+  Riemann = rsolver;
+
+  nxf = grid[IDIR].np_int + 1;
+  nxb = grid[IDIR].lbeg   - 1;
+
+  jj = kk = 0; g_j = &jj; g_k = &kk;
+
+/* -----------------------------------------------------------------
+               Check algorithm compatibilities
+   ----------------------------------------------------------------- */
+
+  #if PARABOLIC_FLUX == YES && ARTIFICIAL_VISCOSITY == NO
+   print1 ("! Parabolic Terms are not implemented in 1D CTU at the moment\n");
+   QUIT_PLUTO(1);
+  #endif
+ 
+  #if !(GEOMETRY == CARTESIAN || GEOMETRY == CYLINDRICAL)
+   print1 ("! CTU only works in cartesian or cylindrical coordinates\n");
+   QUIT_PLUTO(1);
+  #endif     
+
+  if (NX1_TOT > NMAX_POINT || NX2_TOT > NMAX_POINT || NX3_TOT > NMAX_POINT){
+    print ("! PatchUnsplit: need to re-allocate matrix\n");
+    QUIT_PLUTO(1);
+  }
+
+/* -----------------------------------------------------------------
+                          Allocate memory
[…]

### 2014-10-20 sims/sgPLUTO/Src/Chombo/PatchGrid.H

@@ -1,0 +1,122 @@
+#ifndef PATCH_GRID_H
+#define PATCH_GRID_H
+
+extern "C" {
+#include "pluto.h"
+}
+
+#include "MayDay.H"
+#include "PiecewiseLinearFillPatch.H"
+#include "FineInterp.H"
+
+class PatchGrid
+{
+public:
+
+ PatchGrid(){;}
+
+  ~PatchGrid(){
+
+    if (PGrid[0].x != NULL) { 
+      FreeGrid(PGrid);  
+    } 
+  }
+
+ struct GRID* getGrid()
+ {
+   return PGrid;
+ }
+
+protected:
+
+ struct GRID PGrid[3];
+
+private:
+
+
+};
+
+class PiecewiseLinearFillPluto : public PiecewiseLinearFillPatch
+{
+public:
+
+  PiecewiseLinearFillPluto(){m_is_defined = false;}
+
+ ~PiecewiseLinearFillPluto(){;}
+
+  int get_geometry ()
+  {
+   #if GEOMETRY == CARTESIAN
+     return 1;
+   #endif
+
+   #if GEOMETRY == CYLINDRICAL
+    #if DIMENSIONS == 2
+     return 2;
+    #else
+     MayDay::Error("Cylindrical geometry supported in Pluto/Chombo only in 2 dimensions");
+    #endif
+   #endif
+
+   #if GEOMETRY == SPHERICAL
+//    MayDay::Error("Spherical geometry not supported in Pluto/Chombo");
+    #if DIMENSIONS >= 2
+     #if LOGR == NO
+      return 3;
+     #else
+      return 4;
+     #endif
+    #else
+     MayDay::Error("Spherical geometry supported in Pluto/Chombo only in 2 dimensions");
+    #endif
+   #endif
+
+   #if GEOMETRY == POLAR
+     MayDay::Error("Only cartesian and 2D cylindrical/spherical geometries are supported in Pluto/Chombo");
+   #endif
+  }
+
+};
+
+class FineInterpPluto : public FineInterp
+{
+public:
+
+  FineInterpPluto(){is_defined = false;}
+
+ ~FineInterpPluto(){;}
+
+  int get_geometry ()
+  {
+   #if GEOMETRY == CARTESIAN
+     return 1;
+   #endif
+
+   #if GEOMETRY == CYLINDRICAL
+    #if DIMENSIONS == 2
+     return 2;
+    #else
[…]

### 2014-10-20 sims/sgPLUTO/Src/Chombo/PatchRK.3D.cpp

@@ -1,0 +1,405 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*!
+  \file
+  \brief Implements a RK-midpoint time stepping routine.
+
+  The RK-midpoing advance the system of conservation law by first
+  evolving the solution at the half-time level and then doing a full
+  step using midpoint quadrature rule:
+  \f[ \left\{\begin{array}{lcl}
+       U^{n+\HALF} &=& U^n + \frac{\Delta t}{2} R^n \\ \noalign{\medskip}
+       U^{n+1}     &=& U^n + \Delta t R^{n+\HALF} 
+      \end{array}\right.
+  \f]
+  
+  \authors A. Mignone (mignone@ph.unito.it)\n
+           C. Zanni   (zanni@oato.inaf.it)
+  \date   Sep 20, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+
+#include <cstdio>
+#include <string>
+using std::string;
+
+#include "PatchPluto.H"
+#include "LoHiSide.H"
+
+/* ********************************************************************* */
+void PatchPluto::updateSolution(FArrayBox&  a_U,
+                                FArrayBox&  a_Utmp,
+                                FArrayBox&  split_tags,
+                                BaseFab<unsigned char>& a_Flags,
+                                FluxBox&    a_F,
+                                Time_Step   *Dts,
+                                const Box&  UBox,
+                                Grid *grid)
+/*
+ *
+ *
+ *
+ *
+ *********************************************************************** */
+{
+  CH_assert(isDefined());
+  CH_assert(UBox == m_currentBox);
+
+  int nv, in;
+  int nxf, nyf, nzf, indf;
+  int nxb, nyb, nzb;
+  int *i, *j, *k;
+  int ii, jj, kk;
+
+  double ***UU[NVAR];
+ #ifdef SKIP_SPLIT_CELLS
+  double ***splitcells;
+ #endif
+  double *inv_dl, dl2;
+  static Data d;
[…]

### 2014-10-20 sims/sgPLUTO/Src/Chombo/PatchTools.cpp

@@ -1,0 +1,570 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*!
+  \file
+  \brief Contains some implementation of the PatchPluto class.
+
+   PatchTools.cpp contains a number of function used by PatchUnsplit.
+   Most of them are simple wrappers to other functions employed by the 
+   static version of the code.
+
+  \authors C. Zanni   (zanni@oato.inaf.it)\n
+           A. Mignone (mignone@ph.unito.it)
+  \date   Sep 20, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+
+#include <cstdio>
+#include <string>
+using std::string;
+
+#include "PatchPluto.H"
+#include "LoHiSide.H"
+
+/* ********************************************************************* */
+void PatchPluto::saveFluxes (const State_1D *state, int beg, int end, 
+                             Grid *grid)
+/*! 
+ *  Rebuild fluxes in a way suitable for AMR operation
+ *  by adding pressure and multiplying by area.
+ *
+ *********************************************************************** */
+{
+  int  i, nv;
+  double **f, *p, r;
+
+  f = state->flux;
+  p = state->press;
+
+  for (i = beg; i <= end; i++) {
+    f[i][MXn] += p[i];
+  }    
+  #if GEOMETRY == CYLINDRICAL
+   if (g_dir == IDIR){
+     for (i = beg; i <= end; i++) {
+     for (nv = 0; nv < NVAR; nv++) {
+       f[i][nv] *= grid[IDIR].A[i];
+     }}
+   }else{
+     r = fabs(grid[IDIR].x[*g_i]);
+     for (i = beg; i <= end; i++) {
+     for (nv = 0; nv < NVAR; nv++) {
+       f[i][nv] *= r;
+     }}
+   }
+  #endif
+  #if GEOMETRY == SPHERICAL
+   if (g_dir == IDIR){
+     for (i = beg; i <= end; i++) {
+     for (nv = 0; nv < NVAR; nv++) {
+       f[i][nv] *= grid[IDIR].A[i];
+#if CH_SPACEDIM > 1
+       f[i][nv] *= grid[JDIR].dV[*g_j]/m_dx;
+#endif
+     }}
+   }else{
+     for (i = beg; i <= end; i++) {
+     for (nv = 0; nv < NVAR; nv++) {
+       f[i][nv] *= grid[JDIR].A[i]*grid[IDIR].x[*g_i];
[…]

### 2014-10-20 sims/sgPLUTO/Src/Chombo/TagCells.cpp

@@ -1,0 +1,264 @@
+#include <cstdio>
+#include <string>
+using std::string;
+
+#include "PatchPluto.H"
+#include "LoHiSide.H"
+
+static void computeRefVar(double ***UU[], double ***q, double, RBox *Ubox);
+
+#if (EOS != ISOTHERMAL) && (AMR_EN_SWITCH == NO)
+ #define REF_VAR  ENG
+#else
+ #define REF_VAR  RHO
+#endif
+
+// #define REF_VAR -1  /* means user-defined */
+
+#define REF_CRIT 2   /* 1 == first derivative, 2 == second derivative */
+
+/* ************************************************************************* */
+void PatchPluto::computeRefGradient(FArrayBox& gFab, FArrayBox& UFab, const Box& b)
+/*!
+ * Compute numerical gradient of the solution in order to tag 
+ * zones for refinement.
+ * The gradient is computed by standard finite differences using
+ *
+ * - REF_CRIT equal to 1 --> compute (normalized) gradient using 1st 
+ *                           derivative of the solution;
+ * - REF_CRIT equal to 2 --> compute (normalized) gradient using 2nd 
+ *                           derivative of the solution (default);
+ *                           This approach is based on Lohner (1987).
+ *
+ * Zones will be flagged for refinement whenever grad[k][j][i] exceeds 
+ * the threshold value specified by the 'Refine_thresh' parameter read in
+ * pluto.ini.
+ *
+ * Derivatives are computed using the conserved variable U[REF_VAR] 
+ * where REF_VAR is taken to be energy density (default).
+ * However, by setting REF_VAR = -1, you can provide your own 
+ * physical variable through the function computeRefVar().
+ * 
+ * \authors C. Zanni   (zanni@oato.inaf.it)\n
+ *          A. Mignone (mignone@ph.unito.it)
+ * \date    Oct 11, 2012
+ *************************************************************************** */
+{
+  CH_assert(m_isDefined);
+
[…]

### 2014-10-20 sims/sgPLUTO/Src/Cooling/MINEq/jacobian.c

@@ -1,0 +1,180 @@
+#include "pluto.h"
+#include "cooling_defs.h"
+
+/* ******************************************************** */
+void Jacobian (real *v, real *rhs, real **dfdy)
+/*
+ *
+ *   Compute the jacobian J(k,l) = dfdy
+ *
+ *   k = row index
+ *   l = column index
+ *
+ *    J(0,0)   J(0,1) ...  J(0, n-1)
+ *    J(1,0)   J(1,1) .... J(1, n-1)
+ *      .         .           .
+ *      .         .           .
+ *      .         .           .
+ *    J(n-1,0)  ....      J(n-1, n-1)
+ *
+ *
+ *   or, 
+ *
+ *   +-----------------------+
+ *   +              |        |
+ *   +              |        |
+ *   +              |        |
+ *   +    dX'/dX    | dX'/dp |
+ *   +     (JXX)    |  (JXp) |
+ *   +              |        |
+ *   +              |        |
+ *   +--------------+--------+
+ *   +   dp'/dX     | dp'/dp | 
+ *   +    (JpX)     |  Jpp   |
+ *   +-----------------------+
+ *
+ *
+ *
+ ********************************************************** */
+{
+
+  int    k, l, nv, n;
+  real   dmu_dX[NIONS], N, mu;
+  real   T, dLdX, dCdX, dRdX, eps, scrh;
+  real   vp[NVAR], vm[NVAR], rhs_m[NVAR], rhs_p[NVAR];
+  real   dfdp[NIONS];
+  real   *L, *La, *Lb, *Lc;
+  real   *C, *Ca, *Cb, *Cc;
+  real   *R, *Ra, *Rb, *Rc;
+  real   *X, *dnel_dX, *de;
+  real   Unit_Time, E_cost;
+  static real **delta;
+
+  n = NIONS + 1;
+
+  Unit_Time = g_unitLength/g_unitVelocity;
+
+  E_cost    = g_unitLength/g_unitDensity/
+              (g_unitVelocity*g_unitVelocity*g_unitVelocity);
+
+  if (delta == NULL) {
+    delta = ARRAY_2D(NIONS, NIONS, double);
+    for (k = 0; k < NIONS; k++){
[…]

### 2014-10-20 sims/sgPLUTO/Src/Cooling/MINEq/maxrate.c

@@ -1,0 +1,56 @@
+#include "pluto.h"
+#include "cooling_defs.h"
+
+/* ********************************************************** */
+real GetMaxRate (real *v0, real *k1, real T0)
+/*
+ *
+ *  PURPOSE:
+ *
+ *    return an estimate of the maximum rate (dimension 1/time) 
+ *    in the chemical network. This will serve as a
+ *    "stiffness" detector in the main ode integrator.
+ *   
+ *    For integration to be carried explicitly all the time,
+ *    return a small value (1.e-12).
+ *
+ ************************************************************ */
+{
+  int  nv;
+  real   ev1[NIONS+1];
+  real   scrh, maxrate;
+  static real **J1;
+
+  if (J1 == NULL)  {
+    J1 = ARRAY_2D(NIONS+1, NIONS+1, double);
+  }
+
+  maxrate = 0.0;
+  for (nv = 0; nv < NIONS - Fe_IONS; nv++) {  /* -- exclude Iron -- */
+
+  /* ---------------------------------------------- 
+       if the initial concentration is close to 0,
+       do not take Crate as a stiffness indicator
+     ---------------------------------------------- */
+
+    if (v0[NFLX + nv] < 1.e-6) continue;
+
+    maxrate = MAX(maxrate, CoolCoeffs.Crate[nv]);
+  } 
+  maxrate *= g_unitLength/g_unitVelocity;
+
+/*
+  Numerical_Jacobian    (v0, J1);
+  lmax = Decompose(J1[0], NIONS + 1, ev1);
+  Radiat(v0, k1);
+  scrh = maxrate/fabs(lmax) - 1.0;
+  if (fabs(scrh) > 5.e-2){
+     printf ("Crate are not the max eig, %12.6e  %12.6e\n", 
+              maxrate, fabs(lmax));
+     QUIT_PLUTO(1);
+  }
+  return(fabs(lmax));
+*/
+
+  return (maxrate);
+}

### 2014-10-20 sims/sgPLUTO/Src/Cooling/SNEq/maxrate.c

@@ -1,0 +1,20 @@
+#include "pluto.h"
+
+/* ********************************************************** */
+real GetMaxRate (real *v0, real *k1, real T0)
+/*
+ *
+ *  PURPOSE:
+ *
+ *    return an estimate of the maximum rate (dimension 1/time) 
+ *    in the chemical network. This will serve as a
+ *    "stiffness" detector in the main ode integrator.
+ *   
+ *    For integration to be carried explicitly all the time,
+ *    return a small value (1.e-12).
+ *
+ ************************************************************ */
+{
+  return (1.e-12);
+}
+

### 2014-10-20 sims/sgPLUTO/Src/Fargo/fargo.h

@@ -1,0 +1,84 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*!
+  \file
+  \brief FARGO-MHD module header file.
+
+  Contains contains basic definitions and declarations used by the
+  FARGO-MHD module.
+
+  \authors A. Mignone (mignone@ph.unito.it)\n
+           G. Muscianisi (g.muscianisi@cineca.it)
+  \date    Sep 12, 2012
+  \todo   
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+
+#define FARGO_ORDER          3    /**<  Set the order of interpolation during 
+     the linear transport step. Either 2 or 3. */
+        
+/*! Set how often (in number of steps) the total azimuthal 
+    velocity should be averaged.                           */
+#define FARGO_NSTEP_AVERAGE  10   
+
+/*! Average background velocity is computed by averaging the orbital 
+    velocity (YES) or by prescribing the velocity analytically exactly in 
+    the user-supplied function FARGO_SetVelocity  */
+#ifndef SHEARINGBOX
+ #define FARGO_AVERAGE_VELOCITY  YES 
+#else                             
+ #define FARGO_AVERAGE_VELOCITY  NO  
+#endif                            
+                                 
+
+/* ----------------------------------------------------------------
+    Note: when both FARGO and SHEARINGBOX modules are used, *ALL* 
+          source terms are accounted for by the body_force function 
+   ---------------------------------------------------------------- */
+
+/* -- SBEG and SEND are the initial and final 
+      indices in the direction of the orbital velocity -- */
+
+#if GEOMETRY != SPHERICAL
+ #define SDIR   JDIR  /* Orbital direction. */
+ #define SBEG   JBEG  /* Starting index in the orbital dir. */
+ #define SEND   JEND  /* Final index in the orbital dir. */
+ #define NS     NX2
+ #define NS_TOT NX2_TOT
+#else
+ #define SDIR   KDIR
+ #define SBEG   KBEG
+ #define SEND   KEND
+ #define NS     NX3
+ #define NS_TOT NX3_TOT
+#endif
+
+/* ------------------------------------------------------------------
+             Handy macros for dimensional loops
[…]

### 2014-10-20 sims/sgPLUTO/Src/Fargo/fargo_velocity.c

@@ -1,0 +1,294 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*! 
+  \file  
+  \brief Functions for computing/retrieving the mean aziumthal velocity.
+
+  Collects various functions for computing, adding, subtracting and
+  retrieving the azimuthally-averaged velocity.
+  Depending on the macro ::FARGO_AVERAGE_VELOCITY (YES/NO) this is
+  done either numerically or analytically.
+
+  \authors A. Mignone (mignone@ph.unito.it)\n
+           G. Muscianisi (g.muscianisi@cineca.it)
+         
+  \date   Sept 12, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+#include "pluto.h"
+
+/*! Defines a 2D array containing the azimuthally-averaged velocity.
+    When the average is done along the X2 direction (Cartesian and polar
+    geometries) the array should have size (NX3_TOT, NX1_TOT).
+    When the average is performed along the X3 direction (spherical geometry)
+    the array has dimensions (NX2_TOT, NX1_TOT)                           */
+static double **wA;  
+/*static double **wAx1, **wAx2, **wAx3;*/  /* -- average orbital speed at
+                                            x1, x2 and x3 faces -- */
+                                            
+/* ********************************************************************* */
+void FARGO_ComputeVelocity(const Data *d, Grid *grid)
+/*!
+ * Compute the mean orbital velocity as a 2D array by averaging
+ * for each pair (x,z)/(r,z)/(r,theta)  (in Cartesian/polar/spherical)
+ * along the orbital coordinate y/phi/phi.
+ *
+ *********************************************************************** */
+#if GEOMETRY != SPHERICAL
+ #define s j   /* -- orbital direction index -- */
+#else
+ #define s k
+#endif
+{
+  int i,j,k;
+  static int first_call = 1;
+  double *x1, *x2, *x3, th, w;
+  double ***vphi;
+
+  x1 = grid[IDIR].x;
+  x2 = grid[JDIR].x;
+  x3 = grid[KDIR].x;
+
[…]

### 2014-10-20 sims/sgPLUTO/Src/HD/eigenv.c

@@ -1,0 +1,654 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*!
+  \file
+  \brief Wave-speeds and characteristic decomposition for the HD equations.
+
+  This file contains various functions containing Jacobian-related information
+  such as characteristic signal speeds, eigenvalues and eigenvector
+  decomposition for the HD module.
+
+  The function MaxSignalSpeed() computes the maximum and minimum
+  characteristic signal velocity for the HD equations.
+
+  The function Eigenvalues() computes the 3 characteristic waves.
+
+  The function PrimEigenvectors() calculates left and right eigenvectors
+  and the corresponding eigenvalues for the \e primitive form the the
+  HD equations with an adiabatic or isothermal EoS.
+
+  The function ConsEigenvectors() provides the characteristic decomposition
+  of the convervative HD equations.
+
+  The function PrimToChar()  compute the matrix-vector multiplcation
+  between the L matrix (containing primitive left eigenvectors)
+  and the vector v. The result containing the characteristic
+  variables is stored in the vector w = L.v
+
+  \author A. Mignone (mignone@ph.unito.it)
+  \date   Oct 30, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+
+#include "pluto.h"
+
+/* ********************************************************************* */
+void MaxSignalSpeed (double **v, double *cs2, double *cmin, double *cmax,
+                     int beg, int end)
+/*!
+ * Compute the maximum and minimum characteristic velocities for the
+ * HD equation, cmin= v - cs, cmax = v + cs
+ *
+ * \param [in]  v     1-D array of primitive variables
+ * \param [out]  cs2  1-D array containing the square of the sound speed
+ * \param [out] cmin  1-D array containing the leftmost characteristic
+ * \param [out] cmin  1-D array containing the rightmost characteristic
[…]

### 2014-10-20 sims/sgPLUTO/Src/HD/fluxes.c

@@ -1,0 +1,51 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*! 
+ \file
+ \brief Compute the hydro (HD) flux.                                             
+
+  Compute the flux of the conservative HD equations in the direction 
+  given by g_dir.
+  This function defines the component of the hyperbolic flux tensor 
+  of the standard HD equations.\n
+  In what follows:
+  - VXn, MXn are the velocity, momentum components in the direction given
+    by g_dir (normal, "n")
+  - VXt, MXt and VXb, MXb are the transverse components
+    (tangent "t" and bi-tangent "b").
+
+ \author A. Mignone (mignone@ph.unito.it)
+ \date   Aug 16, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+#include "pluto.h"
+
+/* ********************************************************************* */
+void Flux (double **u, double **w, double *a2, double **fx, double *p, 
+           int beg, int end)
+/*!
+ * \param [in]    u    1D array of conserved quantities
+ * \param [in]    w    1D array of primitive quantities
+ * \param [in]   a2    1D array of sound speeds
+ * \param [out]  fx    1D array of fluxes (total pressure excluded)
+ * \param [out]   p    1D array of pressure values
+ * \param [in]   beg   initial index of computation 
+ * \param [in]   end   final   index of computation
+ *
+ * \return  This function has no return value.
+ *********************************************************************** */
+{
+  int   nv, ii;
+
+  for (ii = beg; ii <= end; ii++) {
+    fx[ii][RHO] = u[ii][MXn];
+    EXPAND(fx[ii][MX1] = u[ii][MX1]*w[ii][VXn]; ,
+           fx[ii][MX2] = u[ii][MX2]*w[ii][VXn]; ,
+           fx[ii][MX3] = u[ii][MX3]*w[ii][VXn];)
+    #if EOS == IDEAL     
+     p[ii] = w[ii][PRS];
+     fx[ii][ENG] = (u[ii][ENG] + w[ii][PRS])*w[ii][VXn];
+    #elif EOS == ISOTHERMAL
+     p[ii] = a2[ii]*w[ii][RHO];
+    #endif
+  }
+}

### 2014-10-20 sims/sgPLUTO/Src/HD/mappers.c

@@ -1,0 +1,180 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*!
+  \file
+  \brief Convert between primitive and conservative variables.
+
+  The PrimToCons() converts an array of primitive quantities to 
+  an array of conservative variables for the HD equations.
+  
+  The ConsToPrim() converts an array of conservative quantities to 
+  an array of primitive quantities.
+  During the conversion, pressure is normally recovered from total 
+  energy unless zone has been tagged with FLAG_ENTROPY.
+  In this case we recover pressure from conserved entropy:
+  
+      if (FLAG_ENTROPY is TRUE)  --> p = p(S)
+      else                       --> p = p(E)
+  
+  \author A. Mignone (mignone@ph.unito.it)
+  \date   Oct 4, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+
+#include "pluto.h"
+
+/* ********************************************************************* */
+void PrimToCons (double **uprim, double **ucons, int ibeg, int iend)
+/*!
+ * Convert primitive variables to conservative variables. 
+ *
+ * \param [in]  uprim array of primitive variables
+ * \param [out] ucons array of conservative variables
+ * \param [in]  beg   starting index of computation
+ * \param [in]  end   final index of computation
+ *
+ *********************************************************************** */
+{
+  int  i, nv;
+  double *v, *u;
+  double rho, gmm1;
+
+  #if EOS == IDEAL
+   gmm1 = g_gamma - 1.0;
+  #endif
+  for (i = ibeg; i <= iend; i++) {
+  
+    v = uprim[i];
+    u = ucons[i];
+
+    u[RHO] = rho = v[RHO];
+    EXPAND (u[MX1] = rho*v[VX1];  ,
+            u[MX2] = rho*v[VX2];  ,
+            u[MX3] = rho*v[VX3];)
+
+    #if EOS == IDEAL
+     u[ENG] = EXPAND(v[VX1]*v[VX1], + v[VX2]*v[VX2], + v[VX3]*v[VX3]);
+     u[ENG] = 0.5*rho*u[ENG] + v[PRS]/gmm1;
+    #endif
+
+    for (nv = NFLX; nv < (NFLX + NSCL); nv++) u[nv] = rho*v[nv];
+  }
+}
+/* ********************************************************************* */
[…]

### 2014-10-20 sims/sgPLUTO/Src/HD/prim_eqn.c

@@ -1,0 +1,305 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*! 
+  \file  
+  \brief Compute the right hand side of the primitive HD equations.     
+  
+  Implements the right hand side of the quasi-linear form of the hydro
+  equations. 
+  In 1D this may be written as
+  \f[ 
+      \partial_t{\mathbf{V}} = - A\cdot\partial_x\mathbf{V} + \mathbf{S}
+  \f]
+  where \f$ A \f$ is the matrix of the primitive form of the equations,
+  \f$ S \f$ is the source term.
+
+  \b Reference
+ 
+  - "A solution adaptive upwind scheme for ideal MHD",
+    Powell et al., JCP (1999) 154, 284.
+ 
+  The function PrimRHS() implements the first term while PrimSource() 
+  implements the source term part.
+
+  \author A. Mignone (mignone@ph.unito.it)
+  \date   Jul 31, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+#include "pluto.h"
+
+/* *********************************************************************** */
+void PrimRHS (double *w, double *dw, double cs2, double h, double *Adw)
+/*!
+ * Compute the matrix-vector multiplication \f$ A(\mathbf{v})\cdot 
+ * d\mathbf{v} \f$ where A is the matrix of the quasi-linear form 
+ * of the HD equations.
+ *
+ * \param [in]  w    vector of primitive variables;
+ * \param [in]  dw   limited (linear) slopes;
+ * \param [in]  cs2  local sound speed;
+ * \param [in]  h    local enthalpy;
+ * \param [out] Adw  matrix-vector product.
+ * \return This function has no return value.
+ ************************************************************************** */
+{
+  int  nv;
+  real u;
+
+  u   = w[VXn];
+  
+  Adw[RHO] = u*dw[RHO] + w[RHO]*dw[VXn];
+  #if EOS == IDEAL
+   EXPAND(Adw[VXn] = u*dw[VXn] + dw[PRS]/w[RHO];  ,
+          Adw[VXt] = u*dw[VXt];                 ,
+          Adw[VXb] = u*dw[VXb];)
[…]

### 2014-10-20 sims/sgPLUTO/Src/HD/roe.c

@@ -1,0 +1,354 @@
+#include"pluto.h"
+
+#define ROE_AVERAGE       YES
+
+/* ********************************************************************** */
+void Roe_Solver (const State_1D *state, int beg, int end, 
+                 double *cmax, Grid *grid)
+/*
+ *
+ * NAME
+ *
+ *   Roe_Solver 
+ *
+ *
+ * PURPOSE
+ *
+ *   - Solve riemann problem for the Euler equations 
+ *     using the standard Roe method with local characteristic 
+ *     approach; it is identical to the Roe method.
+ *     Eigenvectors are identical to the ones given by Toro, 
+ *     and can also be derived from the maple script
+ *     "eigenv.maple", in Src/HD/.
+ *
+ *     when ROE_AVERAGE == YES use Roe average   
+ *     when ROE_AVERAGE == NO  use arithmetic average 
+ * 
+ *     Reference:    "Riemann Solver and Numerical Methods 
+ *                    for Fluid Dynamics"
+ *                    E.F. Toro
+ *                    --> see page 359-367
+ *   
+ *   - On input, it takes left and right primitive state
+ *     vectors state->vL and state->vR at zone edge i+1/2;
+ *     On output, return flux and pressure vectors at the
+ *     same interface.
+ *
+ *   - Also, compute maximum wave propagation speed (cmax) 
+ *     for  explicit time step computation
+ *  
+ *
+ * LAST_MODIFIED
+ *
+ *   June 5th 2006, by Andrea Mignone  (mignone@to.astro.it)
+ *
+ *
+ ************************************************************************ */
+{
+  int   nv, i, j, k, nn;
+  double  scrh;
+  double  um[NFLX], vel2;
+  double  a2, a, h;
+  double  dv[NFLX], eta[NFLX];
+  double  Rc[NFLX][NFLX], alambda[NFLX], lambda[NFLX];
+  double  delta, delta_inv, gmm1, gmm1_inv;
+#if ROE_AVERAGE == YES
+  double s, c, hl, hr;
+#endif
[…]

### 2014-10-20 sims/sgPLUTO/Src/MHD/CT/ct.c

@@ -1,0 +1,384 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*! 
+  \file  
+  \brief Update staggered magnetic field.
+
+  Magnetic field is updated using a discrete version of Stoke's theorem
+
+  \b References
+   - "A staggered mesh algorithm using high-order Godunov fluxes to 
+      ensure solenoidal magnetic field in MHD simulations"\n
+      Balsara \& Spicer, JCP (1999) 149, 270
+  
+  \author A. Mignone (mignone@ph.unito.it)
+  \date   Aug 16, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+#include "pluto.h"
+
+/* ********************************************************************* */
+void CT_Update(const Data *d, Grid *grid)
+/*!
+ * Update staggered magnetic field using discrete version of 
+ * Stoke's theorem.
+ *
+ * \param [in,out] d     pointer to PLUTO Data structure
+ * \param [in]     grid  pointer to an array of Grid structures
+ *
+ * \todo  dt may be set separately in a more concise manner.
+ *
+ * \author A. Mignone (mignone@ph.unito.it)
+ * \date   Aug 16, 2012
+ *********************************************************************** */
+              
+{
+  int  i, j, k, nv;
+  int  ibeg, jbeg, kbeg;
+  int  iend, jend, kend;
+  double rhs_x, rhs_y, rhs_z, dt;
+  double *dx, *dy, *dz, *A1, *A2, *dV1, *dV2;
+  double *r, *rp, *th, r_2;
+  double ***ex, ***ey, ***ez;
+  EMF *emf;
+  static Data_Arr bf0;
+  
+  if (bf0 == NULL){
+    bf0 = ARRAY_4D(DIMENSIONS, NX3_TOT, NX2_TOT, NX1_TOT, double);
+  }
+
+/* ---- check div.B ---- */
+
+  #if CHECK_DIVB_CONDITION == YES
+   if (g_intStage == 1) CT_CheckDivB (d->Vs, grid);
[…]

### 2014-10-20 sims/sgPLUTO/Src/MHD/CT/ct_emf_average.c

@@ -1,0 +1,470 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*! 
+ * \file  
+ * \brief  Collects different EMF averaging schemes.                     */
+/* ///////////////////////////////////////////////////////////////////// */
+#include "pluto.h"
+
+#define EX(k,j,i)  (vz[k][j][i]*By[k][j][i] - vy[k][j][i]*Bz[k][j][i])
+#define EY(k,j,i)  (vx[k][j][i]*Bz[k][j][i] - vz[k][j][i]*Bx[k][j][i])
+#define EZ(k,j,i)  (vy[k][j][i]*Bx[k][j][i] - vx[k][j][i]*By[k][j][i])
+
+/* ********************************************************************* */
+void CT_EMF_ArithmeticAverage (const EMF *Z1, const double w)
+/*!
+ * \brief Compute arithmetic average of EMF at cell edges.
+ * \details
+ *
+ *  Combine the four electric field values computed at zone faces 
+ *  as upwind Godunov fluxes into an edge-centered value.\n
+ *  The face-centered EMF should have been stored by previous calls
+ *  to CT_StoreEMF() during the one-dimensional sweeps.\n
+ *  This function employs a simple arithmetic averaging of the 
+ *  face-centered electric field.
+ *
+ * \b References:
+ *    - "A Staggered Mesh Algorithm Using High Order Godunov Fluxes to 
+ *       Ensure Solenoidal Magnetic Fields in Magnetohydrodynamic 
+ *       Simulations"\n
+ *       Balsara \& Spicer, JCP (1999) 149, 270.
+ *
+ * \param [in]      Z1    pointer to EMF structure
+ * \param [in]      w     weighting factor
+ *
+ * \return  This function has no return value.
+ *
+ *
+ * \author A. Mignone (mignone@ph.unito.it)
+ * \date   Aug 16, 31, 2012
[…]

### 2014-10-20 sims/sgPLUTO/Src/MHD/GLM/glm.h

@@ -1,0 +1,56 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*!
+  \file
+  \brief  Header file for GLM Divergence Cleaning
+
+  Contains function prototypes and global variable declaration
+  for the GLM formulation to control the divergence-free condition 
+  of magnetic field.
+
+  \b References
+     - "A Second-order unsplit Godunov scheme for cell-centered MHD:
+        The CTU-GLM scheme"\n
+        Mignone \& Tzeferacos, JCP (2010) 229, 2117
+
+     - "High-order conservative finite difference GLM-MHD scheme for
+        cell-centered MHD"\n
+        Mignone, Tzeferacos \& Bodo, JCP (2010) 229, 5896
+        
+  \authors A. Mignone (mignone@ph.unito.it)\n
+           P. Tzeferacos (petros.tzeferacos@ph.unito.it)
+  \date   Sep 28, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+#define GLM_MHD 
+
+#define EGLM   NO  /**< The EGLM macro may be turned to YES to enable
+                        the extended GLM formalism. 
+                        Although it breaks conservation of momentum and
+                        energy, it has proven to be more robust in treating
+                        low-beta plasma. */
+                        
+#define COMPUTE_DIVB  NO
+
+/* with chombo, COMPUTE_DIVB must be 
+   disabled or a segfault will occur */
+
+#ifdef CH_SPACEDIM    
+ #undef COMPUTE_DIVB  
+ #define COMPUTE_DIVB NO
+#endif
+
+   
+extern double glm_ch; /**< The propagation speed of divergence error. */
+
+    
+void  GLM_Solve (const State_1D *, double **, double **, int, int, Grid *);
+void  GLM_SolveNEW (const State_1D *state, int beg, int end, Grid *grid);
+void  GLM_Init      (const Data *, const Time_Step *, Grid *);
+void  GLM_Source (const Data_Arr, double, Grid *);
+void  EGLM_Source (const State_1D *, double, int, int, Grid *);
+
+#if COMPUTE_DIVB == YES
+ void GLM_ComputeDivB(const State_1D *state, Grid *grid);
+ double ***GLM_GetDivB(void);
+#endif
+

### 2014-10-20 sims/sgPLUTO/Src/MHD/Resistive/res.h

@@ -1,0 +1,20 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*!
+  \file
+  \brief Resistive MHD module header file.
+
+  Contains prototypes for the resistive MHD module.
+
+  \authors A. Mignone (mignone@ph.unito.it)\n
+           T. Matsakos
+  \date   Sep 13, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+
+void ResistiveFlux (Data_Arr, double **, double **, int, int, Grid *);
+
+void GetCurrent (Data_Arr, real **, Grid *);
+
+void ETA_Func (real *, real, real, real, real *);
+
+void ADD_OHM_HEAT (const Data *d, double, Grid *);

### 2014-10-20 sims/sgPLUTO/Src/MHD/Resistive/res_eta.c

@@ -1,0 +1,10 @@
+#include "pluto.h"
+
+void ETA_Func(real *v, real x1, real x2, real x3, real *eta)
+{
+
+eta[0] = 1.0;
+eta[1] = 1.0;
+eta[2] = 1.0;
+
+}

### 2014-10-20 sims/sgPLUTO/Src/MHD/Resistive/res_functions.c

@@ -1,0 +1,439 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*!
+  \file
+  \brief Compute the electric current.
+
+  Compute the curl of magnetic field at the cell interfaces in the 
+  direction given by ::g_dir.
+  It returns a 1D array containing two of the three components 
+  (J_x, J_y, J_z). For instance, during the IDIR sweep, we compute
+  (0, J_y, J_z) at interface (i+1/2,j,k).
+  We do NOT compute the x-component (for g_dir == IDIR) since any 
+  differential operator successively applied to J will not need it. 
+  For instance:
+ 
+            div(J)  = 0  (by definition)
+            curl(J) at i+1/2 does not need Jx;
+ 
+  The same argument applies to other directions by cyclic permutation of 
+  indexes.
+ 
+  \attention Do NOT use this function to compute terms like (curl V)^2 !!
+
+  \authors T. Matsakos\n
+           A. Mignone (mignone@ph.unito.it)\n
+           P. Tzeferacos (petros.tzeferacos@ph.unito.it)
+           
+  \date   Aug 16, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+#include "pluto.h"
+
+/* ********************************************************************* */
+void GetCurrent (Data_Arr V, double **curlB, Grid *grid)
+/*!
+ *
+ * \param [in] V       A 3D array of cell-centered primitive quantities
+ * \param [out] curlB  A 1D array containing two components of the current
+ *                     in a the direction set by ::g_dir
+ * \param [in] grid    a pointer to an array of Grid structures
+ *   
+ *********************************************************************** */
+{
+  int  i, j, k;
+  double dx_1, dy_1, dz_1, r_1, s_1;
+  double *inv_dx,  *inv_dy,  *inv_dz;
+  double *inv_dxi, *inv_dyi, *inv_dzi;
+  double *r, *th;
+  double ***Bx, ***By, ***Bz;
[…]

### 2014-10-20 sims/sgPLUTO/Src/MHD/ShearingBox/shearingbox.h

@@ -1,0 +1,89 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*!
+  \file
+  \brief Shearing-Box module header file
+
+  The Shearing-Box module header file contains basic macro definitions, 
+  function prototypes and declaration of global variables used by the 
+  sheraring-box module.
+  The variable ::sb_q and ::sb_Omega are the most important ones and 
+  \b must be defined and initialized in your init.c in order to 
+  configure your shearing-box problem.
+
+  Optionally, the order of interpolation (default is 2) at physical 
+  boundaries may be changed using the ::SB_ORDER macro.
+
+  The additional macros ::SB_SYMMETRIZE_HYDRO, ::SB_SYMMETRIZE_EY and
+  ::SB_SYMMETRIZE_EZ may be set to YES/NO to enable/disable enforcement
+  of conservation at the radial (x) boundaries.
+
+  \authors A. Mignone (mignone@ph.unito.it)\n
+           G. Muscianisi (g.muscianisi@cineca.it)
+  \date   Aug 16, 2012
+  \todo Check if sb_vy and sb_Ly are really needed as global variables.
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+
+/*! Sets the order of interpolation at physical boundaries (1, 2 or 3).*/
+#if INTERPOLATION == LINEAR || INTERPOLATION == WENO3 || INTERPOLATION == LimO3
+ #define SB_ORDER              2    
+#else
+ #define SB_ORDER              3    
+#endif                                 
+
+#ifndef FARGO
+ #define SB_SYMMETRIZE_HYDRO  YES  /**< Symmetrize the hydrodynamical fluxes 
+         at the left and right x-boundaries in order to enforce conservation of
+         hydrodynamic variables like density, momentum and energy
+        (no magnetic field). Default is YES.*/
+
+ #define SB_SYMMETRIZE_EY    (YES  && (DIMENSIONS == 3)) /**< Symmetrize the
+         y-component of the electric field at the left and right x-boundaries
+         to enforce conservation of magnetic field (only in 3D). */
+
+ #define SB_SYMMETRIZE_EZ     YES  /**< Symmetrize the z-component of electric
+         field at the left and right x-boundaries to enforce conservation of
+         magnetic field. */
+
[…]

### 2014-10-20 sims/sgPLUTO/Src/MHD/bckgrnd_field.c

@@ -1,0 +1,71 @@
+#include "pluto.h"
+
+#if BACKGROUND_FIELD == YES
+/* ******************************************************************** */
+double **GetBackgroundField (int beg, int end, int where, Grid *grid)
+/*
+ *
+ *
+ *
+ *
+ *
+ ********************************************************************* */
+{
+  int i;
+  double *x1, *x2, *x3;
+  static real **bck_fld;
+
+/* ----------------------------------------------------
+         Check for incompatibilities 
+   ---------------------------------------------------- */
+  
+  #if (TIME_STEPPING != RK2) && (TIME_STEPPING != RK3)
+   print1 ("! Background field splitting works with RK integrators ONLY \n");
+   QUIT_PLUTO(1);
+  #endif
+  #if MHD_FORMULATION != CONSTRAINED_TRANSPORT
+   print1 ("! Background field splitting works with CT ONLY \n");
+   QUIT_PLUTO(1);
+  #else
+   #if (CT_EMF_AVERAGE != ARITHMETIC) && (CT_EMF_AVERAGE != UCT_HLL)
+    print1 ("! Background field splitting works with ARITHMETIC or UCT_HLL averages only\n");
+    QUIT_PLUTO(1);
+   #endif
+  #endif
+  #if RESISTIVE_MHD != NO 
+    print1 ("! Background field splitting works with Ideal MHD only\n");
+    QUIT_PLUTO(1);
+  #endif  
+
+  if (bck_fld == NULL) {
+    bck_fld = ARRAY_2D(NMAX_POINT, NVAR, double);
+  }
+
+/* ----------------------------------
+     get pointers to coordinates 
+   ---------------------------------- */
+   
+  x1 = grid[IDIR].x;
+  x2 = grid[JDIR].x;
+  x3 = grid[KDIR].x;
+
+  if (g_dir == IDIR){
+    if (where == FACE_CENTER) x1 = grid[IDIR].xr;
+    for (i = beg; i <= end; i++){
+      BackgroundField (x1[i],x2[*g_j],x3[*g_k], bck_fld[i] + BX);
+    }
+  }else if (g_dir == JDIR){
+    if (where == FACE_CENTER) x2 = grid[JDIR].xr;
+    for (i = beg; i <= end; i++){
+      BackgroundField (x1[*g_i],x2[i],x3[*g_k], bck_fld[i] + BX);
+    }
+  }else{
[…]

### 2014-10-20 sims/sgPLUTO/Src/MHD/hll_speed.c

@@ -1,0 +1,171 @@
+#include"pluto.h"
+
+/*   ****  switches, see below for description  ****  */
+
+#define ROE_ESTIMATE    NO
+#define DAVIS_ESTIMATE  YES   
+
+/* ***************************************************************************** */
+void HLL_Speed (double **vL, double **vR, double *a2L, double *a2R, double **bgf, 
+                double *SL, double *SR, int beg, int end)
+/* 
+ * PURPOSE
+ *
+ *   Compute leftmost (SL) and rightmost (SR) speed for the Riemann fan
+ * 
+ *
+ * ARGUMENTS
+ *
+ *   vL, vR       1-D input array of left/right states of primitive vars; 
+ *   SL, SR       wave speed to be returned;
+ *   bgf          background field
+ *   beg, end     initial, final index for spatial loop
+ *
+ * SWITCHES
+ *
+ *    ROE_ESTIMATE (YES/NO), DAVIS_ESTIMATE (YES/NO)
+ *
+ *    These switches set how the wave speed estimates are
+ *    going to be computed. Only one can be set to 'YES', and
+ *    the rest of them must be set to 'NO'  
+ *
+ *     ROE_ESTIMATE:    b_m = \min(0, \min(u_R - c_R, u_L - c_L, u_{roe} - c_{roe}))     
+ *                      b_m = \min(0, \min(u_R + c_R, u_L + c_L, u_{roe} + c_{roe}))
+ * 
+ *                      where u_{roe} and c_{roe} are computed using Roe averages.
+ *  
+ *     DAVIS_ESTIMATE:  b_m = \min(0, \min(u_R - c_R, u_L - c_L))     
+ *                      b_m = \min(0, \min(u_R + c_R, u_L + c_L))  
+ *
+ *
+ *
+ * LAST_MODIFIED
+ *
+ *   August 24, 2011 by A. Mignone (mignone@ph.unito.it)
+ *              
+ *
+ ******************************************************************************** */
+{
+  int  i;
+  double scrh, s, c;
+  static double *sl_min, *sl_max;
+  static double *sr_min, *sr_max;
[…]

### 2014-10-20 sims/sgPLUTO/Src/MHD/mod_defs.h

@@ -1,0 +1,242 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*!
+  \file
+  \brief Set labels, indexes and prototypes for the MHD module.
+
+  Contains basic macro definitions, structure definitions and global
+  variable declarations used by the MHD module.
+
+  \author A. Mignone (mignone@ph.unito.it)
+  \date   Sep, 10, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+
+
+/* ****************************************************************
+     We make extra vector components, when not needed, point 
+     to the last element (255) of the array stored by startup.c.  
+   **************************************************************** */
+
+enum {
+
+ #if COMPONENTS == 1
+
+  RHO, MX1, BX1, 
+  #if EOS == IDEAL
+   ENG, PRS = ENG,
+  #endif
+  #if MHD_FORMULATION == DIV_CLEANING
+   PSI_GLM,
+  #endif
+  MX2 = 255, BX2 = 255, MX3 = 255, BX3 = 255,
+
+ #elif COMPONENTS == 2
+
+  RHO, MX1, MX2, BX1, BX2,
+  #if EOS == IDEAL
+   ENG, PRS = ENG,
+  #endif
+  #if MHD_FORMULATION == DIV_CLEANING
+   PSI_GLM,
+  #endif
+  MX3 = 255, BX3 = 255,
+
+ #elif COMPONENTS == 3
+
+  RHO, MX1, MX2, MX3, BX1, BX2, BX3, 
+  #if EOS == IDEAL
+   ENG, PRS = ENG,
+  #endif
+  #if MHD_FORMULATION == DIV_CLEANING
+   PSI_GLM,
+  #endif
+
+ #endif
+
+ VX1 = MX1, VX2 = MX2, VX3 = MX3, 
+
+/* -- backward compatibility -- */
+
+ DN = RHO,
+ #if EOS == IDEAL
+  PR = PRS, EN = ENG,
+ #endif
+ VX = VX1, VY = VX2, VZ = VX3,
+ MX = MX1, MY = MX2, MZ = MX3,
+ BX = BX1, BY = BX2, BZ = BX3
+
+};
+
+/* **************************************
+     add the PSI_GLM label if necessary 
+   ************************************** */
+
+/*
+#if MHD_FORMULATION == DIV_CLEANING 
+ #if EOS == IDEAL
[…]

### 2014-10-20 sims/sgPLUTO/Src/MHD/rhs.c

@@ -1,0 +1,1362 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*! 
+  \file  
+  \brief Compute the right hand side of the conservative 
+         HD/MHD equations.
+
+  This function constructs the one-dimensional right hand side of 
+  the conservative MHD or HD equations in the direction given by g_dir 
+  in different geometries.
+  The right hand side is computed as a two-point flux difference 
+  term plus a source term:
+  
+  \f[ \mathrm{RHS}_i = 
+      \frac{dt}{dV_i}\Big(A_{i+1/2}F_{i+1/2} - A_{i-1/2}F_{i-1/2}\Big)  
+      + dt S_i  \f]
+ 
+   where 
+ 
+   - \f$ A_{i\pm 1/2} \f$ : interface areas
+   - \f$ dV_i         \f$ : cell volume
+   - \f$ F_{i\pm 1/2} \f$ : interface fluxes
+   - \f$ dt           \f$ : time step
+   - \f$ S_i          \f$ : source term including geometrical terms and 
+                            body forces.
+  
+  See also \ref RHS_page
+  The right hand side is assembled through the following steps:
+ 
+   - If either one of FARGO, ROTATION or gravitational potential is 
+      used, fluxes are combined to enforce conservation of total angular
+      momentum and/or energy, see TotalFlux()
+   - initialize rhs with flux differences
+   - add geometrical source terms
+   - enforce conservation of total angular  momentum and/or energy;
+   - add gravity        
+
+  \author A. Mignone (mignone@ph.unito.it)
+  \date   Aug 16, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+#include "pluto.h"
+
+static void TotalFlux (const State_1D *, double *, int, int, Grid *);
+
+#ifdef CH_SPACEDIM   /*  implies Chombo is being used  */
+ #define USE_PR_GRADIENT  YES   
+#else
+ #ifdef FINITE_DIFFERENCE 
+  #define USE_PR_GRADIENT  NO   /* -- default for Finite Difference schemes -- */
+ #else
+  #define USE_PR_GRADIENT  YES   /* -- default, do not change!! -- */
+ #endif
+#endif
+
+#if defined FARGO && !defined SHEARINGBOX
+ #define IF_FARGO(a)  a
+#else 
[…]

### 2014-10-20 sims/sgPLUTO/Src/MHD/roe.bckfld.c

@@ -1,0 +1,435 @@
+#include"pluto.h"
+
+#define  sqrt_1_2  (0.70710678118654752440)
+
+/* **************************************************************************** */
+void ROE_SOLVER (const State_1D *state, int beg, int end, real *cmax, Grid *grid)
+/*
+ *
+ *
+ * NAME
+ *
+ *   ROE_SOLVER 
+ *
+ *
+ * PURPOSE
+ *
+ *   - Solve riemann problem for the MHD equations 
+ *     using the Roe method with arithmetic average; 
+ *     it is identical to the Roe method.
+ *
+ *     -----------------------------------------------------------
+ *       Reference paper:
+ *
+ *       "A solution adaptive upwind scheme for ideal MHD"
+ *
+ *        K. G. Powell, P. L. Roe, and T. J. Linde
+ *        Journal of Computational Physics, 154, 284-309, (1999).
+ *     -----------------------------------------------------------
+ *
+ *
+ *   - Also, compute maximum wave propagation speed (cmax) 
+ *     for  explicit time step computation
+ *  
+ *
+ * LAST_MODIFIED
+ *
+ *   April 4th 2006, by Andrea Mignone  (mignone@to.astro.it)
+ *
+ *
+ **************************************************************************** */
+ {
+  int    nv, i, j, k;
+  double gmm1;
+  real scrh0, scrh1, scrh2, scrh3, scrh4;
+  real rho, u, v, w, vel2, bx, by, bz, pr;
+  real b1x, b1y, b1z;
+  real a2, a, ca2, cf2, cs2;
+  real cs, ca, cf, b2, A2, At2;
+  real S;
+  real alpha_f, alpha_s, beta_y, beta_z;
+  real dw[NVAR], *vL, *vR, *uL, *uR;
+  double *SL, *SR;
+  real Rp[NVAR][NVAR], eta[NVAR], lambda[NVAR];
+  real Rc[NVAR][NVAR];
+
+  real tau, sqrt_rho;
+  real delta, delta_inv, gmm1_inv;
+
+  static real *br, *bl, *pR, *pL;
[…]

### 2014-10-20 sims/sgPLUTO/Src/MHD/set_solver.c

@@ -1,0 +1,82 @@
+#include"pluto.h"
+
+/* ****************************************************************** */
+Riemann_Solver *SetSolver (const char *solver)
+/*
+ *
+ *
+ *
+ *
+ *
+ *
+ ********************************************************************* */
+{
+
+  #ifdef FINITE_DIFFERENCE
+   return (&FD_Flux);
+  #endif
+
+/* ------------------------------------------------------
+       Set Pointers for SOLVERS
+   ------------------------------------------------------ */
+
+  #if EOS == IDEAL
+   if (!strcmp(solver, "tvdlf")) {
+
+     return (&LF_Solver);
+
+   }else if (!strcmp(solver, "roe")) {
+
+     return (&Roe_Solver);
+ 
+   }else if (!strcmp(solver, "hlle") ||
+             !strcmp(solver, "hll")) {
+
+     return (&HLL_Solver);
+
+   }else if (!strcmp(solver, "hlld")) {
+
+     return (&HLLD_Solver);
+ 
+   }else if (!strcmp(solver, "hllc")) {
+
+     return (&HLLC_Solver);
+
+   }
+  #elif EOS == ISOTHERMAL
+   if (!strcmp(solver, "tvdlf")) {
+
+     return (&LF_Solver);
+
+   }else if (!strcmp(solver, "roe")) {
+
+     return (&Roe_Solver);
+
+   }else if (!strcmp(solver, "hlle") ||
+             !strcmp(solver, "hll")) {
+
+     return (&HLL_Solver);
+
+   }else if (!strcmp(solver, "hlld")) {
+
+     return (&HLLD_Solver);
+
+   }
+  #elif EOS == BAROTROPIC
+   if (!strcmp(solver, "tvdlf")) {
+
+     return (&LF_Solver);
+
+   }else if (!strcmp(solver, "hlle") ||
+             !strcmp(solver, "hll")) {
+
+     return (&HLL_Solver);
+
+   }
+  #endif
+
+  print1 ("\n! SetSolver: '%s' is not available.\n", solver);
+  QUIT_PLUTO(1);
+
+}
+

### 2014-10-20 sims/sgPLUTO/Src/MHD/source.c

@@ -1,0 +1,275 @@
+#include "pluto.h"
+
+#if MHD_FORMULATION == EIGHT_WAVES
+/* *************************************************************************** */
+void ROE_DIVB_SOURCE (const State_1D *state, int is, int ie, Grid *grid)
+/* 
+ *
+ * PURPOSE
+ *
+ *   Include Powell div.B source term to momentum, induction
+ *   and energy equation for Roe and TVDLF solvers.
+ *
+ *********************************************************************** */
+{
+  int    i;
+  real btx, bty, btz, bx, by, bz, vx, vy, vz;
+  real r, s;
+  real *Ar, *Ath;
+  real *vm, **bgf;
+  real *src, *v;
+  static real *divB, *vp;
+  Grid   *GG;
+
+  if (divB == NULL){
+    divB = ARRAY_1D(NMAX_POINT, double);
+    vp   = ARRAY_1D(NMAX_POINT, double);
+  }
+
+/* ----------------------- ---------------------
+    compute magnetic field normal component 
+    interface value by arithmetic averaging
+   -------------------------------------------- */
+
+  for (i = is - 1; i <= ie; i++) {
+    vp[i] = 0.5*(state->vL[i][BXn] + state->vR[i][BXn]);
+  }
+  vm  = vp - 1;
+  GG  = grid + g_dir;
+  
+/* --------------------------------------------
+    Compute div.B contribution from the normal 
+    direction (1) in different geometries 
+   -------------------------------------------- */
+  
+  #if GEOMETRY == CARTESIAN
+
+   for (i = is; i <= ie; i++) {
+     divB[i] = (vp[i] - vm[i])/GG->dx[i];
+   }
+
+  #elif GEOMETRY == CYLINDRICAL
+
+   if (g_dir == IDIR){   /* -- r -- */
+     Ar  = grid[IDIR].A;
+     for (i = is; i <= ie; i++) {
+       divB[i] = (vp[i]*Ar[i] - vm[i]*Ar[i - 1])/GG->dV[i];
+     }
+   }else if (g_dir == JDIR){  /* -- z -- */
+     for (i = is; i <= ie; i++) {
+       divB[i] = (vp[i] - vm[i])/GG->dx[i];
[…]

### 2014-10-20 sims/sgPLUTO/Src/RHD/energy_solve.c

@@ -1,0 +1,101 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*!
+  \file
+  \brief Inversion scheme for RHD using total energy density.
+
+  Try to recover gas pressure from conserved variables {D, m, E}
+  using the algorithm outlined in Section 2 of Mignone, Plewa Bodo (2005).
+  Specifically, we solve Eq. (12) using a Newton-Raphson root finder.
+
+  \b References
+     - "The Piecewise Parabolic Method for Multidimensional Relativistic
+        Fluid Dynamics"\n
+        Mignone, Plewa \& Bodo, ApJS (2005) 160, 199.
+
+  \author A. Mignone (mignone@ph.unito.it)
+  \date   Oct 4, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+#include "pluto.h"
+
+#define MAX_ITER 20 
+/* ********************************************************************* */
+int EnergySolve (Map_param *par)
+/*!
+ *  Use Newton algorithm to recover pressure p0 from 
+ *  conservative quantities  D, m and E 
+ *
+ *********************************************************************** */
+{
+  int iter;
+  double p, D_1, alpha, alpha2, g2, g, m;
+  double tau, theta, h, dh_dp, dh_dtau, gmmr;
+  double yp, dyp, dp, scrh;
+  double D, E, m2;
+
+  D    = par->D;
+  E    = par->E;
+  m2   = par->m2;
+
+  gmmr = g_gamma/(g_gamma - 1.0);
+  m    = sqrt(m2);
+  p    = m - E;
+  p    = MAX(p, 1.e-18);
+  D_1  = 1.0/D;
+  for (iter = 0; iter < MAX_ITER; iter++) {
+
+    alpha  = E + p;
+    alpha2 = alpha*alpha;
+    g2     = 1.0 - m2/alpha2;
+
+    g2 = MAX(g2,1.e-9);
+    g2 = 1.0/g2;
+    g  = sqrt(g2);
+
+    tau   = g*D_1;
+    theta = p*tau;
+
+    #if EOS == IDEAL
+     h       = 1.0 + gmmr*theta;
+     dh_dp   = gmmr*tau;
+     dh_dtau = gmmr*p;
+    #elif EOS == TAUB
+     h       = 2.5*theta + sqrt(2.25*theta*theta + 1.0);
[…]

### 2014-10-20 sims/sgPLUTO/Src/RHD/eos.c

@@ -1,0 +1,101 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*! 
+  \file
+  \brief Collect various thermodynamic functions for the RHD module
+                    
+  \author A. Mignone (mignone@ph.unito.it)
+  \date   Aug 16, 2012   
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+#include"pluto.h"
+
+/* **************************************************************** */
+void SoundSpeed2 (double **v, double *cs2, double *h, int beg, int end,
+                  int pos, Grid *grid)
+/*!
+ * Define the square of the sound speed.
+ * 
+ * \param [in]    v   1D array of primitive quantities
+ * \param [out] cs2   1D array containing the square of the sound speed
+ * \param [in]    h   1D array of enthalpy values
+ * \param [in]  beg   initial index of computation 
+ * \param [in]  end   final   index of computation
+ * \param [in]  pos   an integer specifying the spatial position 
+ *                    inside the cell (only for spatially-dependent EOS)
+ * \param [in]  grid  pointer to Grid structure
+ *
+ * \return  This function has no return value.
+ *********************************************************************** */
+{
+  int     i;
+  double  a2, theta;
+
+  Enthalpy (v, h, beg, end);
+
+  for (i = beg; i <= end; i++) {
+    theta = v[i][PRS]/v[i][RHO];
+    #if EOS == IDEAL
+     a2 = g_gamma*theta/h[i];
+    #elif EOS == TAUB
+     a2 = theta/(3.0*h[i]) * (5.0*h[i] - 8.0*theta)/(h[i] - theta);
+    #endif
+    cs2[i] = a2;
+  }
+}
+
+/* ********************************************************************* */
+void Enthalpy (double **v, double *h, int beg, int end)
+/*!
+ * Compute the enthalpy.
+ *
+ * \param [in]    v   1D array of primitive quantities
+ * \param [in]    h   1D array of enthalpy values
+ * \param [in]  beg   initial index of computation 
+ * \param [in]  end   final   index of computation
+ *
+ * \return  This function has no return value.
+ *********************************************************************** */
+{
+  int i;
+  double theta, gmmr;
+
+  #if EOS == IDEAL
+   gmmr = g_gamma/(g_gamma - 1.0);
+  #endif
+  
[…]

### 2014-10-20 sims/sgPLUTO/Src/RHD/mod_defs.h

@@ -1,0 +1,127 @@
+/* ****************************************************************
+
+     Set label, indexes and basic prototyping for the relativistic 
+     Hydro (RHD) module.
+
+     We make extra vector components, when not needed, point 
+     to the last element (255) of the array stored by startup.c.  
+   **************************************************************** */
+
+enum {
+ #if COMPONENTS == 1
+  RHO, MX1, ENG, PRS = ENG,
+  MX2 = 255, MX3 = 255,
+ #elif COMPONENTS == 2
+  RHO, MX1, MX2, ENG, PRS = ENG,
+  MX3 = 255,
+ #elif COMPONENTS == 3
+  RHO, MX1, MX2, MX3, ENG, PRS = ENG,
+ #endif
+ VX1 = MX1, VX2 = MX2, VX3 = MX3, 
+
+/* -- backward compatibility -- */
+
+ DN = RHO, PR = PRS, EN = ENG,
+ VX = VX1, VY = VX2, VZ = VX3,
+ MX = MX1, MY = MX2, MZ = MX3
+
+};
+
+#define NFLX (2 + COMPONENTS)
+
+/* *****************************************
+    by turning SUBTRACT_DENSITY to YES, we 
+    let PLUTO evolve the total energy minus
+    the mass density contribution.
+   ***************************************** */
+
+/* #define SUBTRACT_DENSITY   YES  */
+
+/* *************************************************
+     Now define more convenient and user-friendly 
+     pointer labels for geometry setting      
+   ************************************************* */
+
+#if GEOMETRY == CYLINDRICAL 
+
+ #define iVR    VX
+ #define iVZ    VY
+ #define iVPHI  VZ
+
+ #define iMR    MX
+ #define iMZ    MY
+ #define iMPHI  MZ
+
+#endif
+
+#if GEOMETRY == POLAR 
+
+ #define iVR    VX
+ #define iVPHI  VY
+ #define iVZ    VZ
+
+ #define iMR    MX
+ #define iMPHI  MY
+ #define iMZ    MZ
+
+#endif
+
+#if GEOMETRY == SPHERICAL 
+
+ #define iVR    VX
+ #define iVTH   VY
+ #define iVPHI  VZ
+
+ #define iMR    MX
+ #define iMTH   MY
+ #define iMPHI  MZ
+
+#endif
+
+/* ********************************************************************* */
+/*! The Map_param structure is used to pass input/output arguments 
+    during the conversion from conservative to primitive variables 
[…]

### 2014-10-20 sims/sgPLUTO/Src/RHD/prim_eqn.c

@@ -1,0 +1,241 @@
+#include "pluto.h"
+
+/* *********************************************************************** */
+void PrimRHS (real *w, real *dw, real cs2, real h, real *Adw)
+/*
+ *
+ *  RHD matrix for the primitive equations
+ *
+ *
+ * -----------------------------------------------------------
+ *  Reference paper:
+ *
+ *   Mignone, Plewa & Bodo (2005) ApJ
+ * -----------------------------------------------------------
+ *
+ *********************************************************************** */
+{
+  int nv;
+  real rho, u1, u2, u3;
+  real d_2, g2, scrh;
+  real g, v1, v2, v3, gx2;
+
+  #if USE_FOUR_VELOCITY == YES
+
+   rho = w[RHO];
+   EXPAND(u1  = w[VXn];  ,
+          u2  = w[VXt];  ,
+          u3  = w[VXb];)
+
+   scrh = EXPAND(u1*u1, + u2*u2, + u3*u3);
+   g2   = 1.0 + scrh;
+   g    = sqrt(g2);
+   d_2  = g/(g2 - scrh*cs2);
+
+   /* get three-vel  */
+
+   EXPAND(v1 = u1/g; ,
+          v2 = u2/g; ,
+          v3 = u3/g;)
+
+   gx2 = 1.0/(1.0 - v1*v1);
+
+   scrh = EXPAND(v1*dw[VXn], + v2*dw[VXt], + v3*dw[VXb]);
+
+   Adw[PRS] =  d_2*(rho*h*cs2*(dw[VXn] - v1*scrh)
+                       + u1*(1.0 - cs2)*dw[PRS]);
+
+   Adw[RHO] = v1*dw[RHO] - (v1*dw[PRS] - Adw[PRS])/(h*cs2);
+ 
+   scrh = 1.0/(g*rho*h);
+   d_2  = u1*dw[PRS] - g*Adw[PRS];
+
+   EXPAND(Adw[VXn] = v1*dw[VXn] + scrh*(dw[PRS] + u1*d_2);  ,
+          Adw[VXt] = v1*dw[VXt] + scrh*u2*d_2;                ,
+          Adw[VXb] = v1*dw[VXb] + scrh*u3*d_2;)
+
+   for (nv = NFLX; nv < NVAR; nv++) Adw[nv] = v1*dw[nv];         
+
+  #else
+
+   rho = w[RHO];
+   EXPAND(v1 = w[VXn];  ,
+          v2 = w[VXt];  ,
+          v3 = w[VXb];)
[…]

### 2014-10-20 sims/sgPLUTO/Src/RHD/riemann.c

@@ -1,0 +1,568 @@
+#include "pluto.h"
+
+#define MAX_ITER   20
+#define small_p    1.e-12
+#define small_rho  1.e-12
+#define INTERPOLATE_RAREFACTION   YES
+
+#define accuracy   1.e-6
+
+static real GLORENTZ (real *U, int n);
+static void   SHOCK (real tau0, real u0, real p0, real g0, real V0, real h0,
+                     real p1, real *u1, real *dudp, real *zeta, int istate);
+static real RAREFACTION_SPEED (real *u, int side);
+
+static real qglob_r[NFLX], qglob_l[NFLX], gmmr;
+
+/* ********************************************************************  */
+void TwoShock_Solver (const State_1D *state, int beg, int end, 
+                      real *cmax, Grid *grid)
+/*
+ *
+ * NAME
+ *
+ *   RIEMANN
+ *
+ *
+ * PURPOSE
+ *
+ *   Solve riemann problem for the relativistic Euler equations 
+ *   using the two-shock approximation; return numerical
+ *   fluxes and source terms
+ * 
+ *   Reference:    "The Piecewise Parabolic Method for  
+ *                  Multidimensional Relativistic Fluid Dynamics"
+ *                  A. Mignone, T. Plewa and G. Bodo
+ *   
+ *  - On input, it takes left and right primitive state
+ *     vectors state->vL and state->vR at zone edge i+1/2;
+ *     On output, return flux and pressure vectors at the
+ *     same interface.
+ *
+ *  - Also, compute maximum wave propagation speed (cmax) 
+ *    for  explicit time step computation
+ *  
+ *
+ * LAST_MODIFIED
+ *
+ *   April 4th 2006, by Andrea Mignone  (mignone@to.astro.it)
+ *
+ ******************************************************************************* */
+{
+  int     iter, i, nv;
+  int     k, nfail = 0, izone_fail[1024];
+  real  Ustar[NFLX];
+  real  gL, gR, gL1, gR1;
+  real  uxR, uxR1, pR, j2R, wR, VR, VR1;
+  real  uxL, uxL1, pL, j2L, wL, VL, VL1;
[…]

### 2014-10-20 sims/sgPLUTO/Src/RHD/tvdlf.c

@@ -1,0 +1,93 @@
+#include "pluto.h"
+
+/* ********************************************************************* */
+void LF_Solver (const State_1D *state, int beg, int end, 
+                real *cmax, Grid *grid)
+/*
+ *
+ * NAME
+ *
+ *   LF_SOLVER
+ *
+ *
+ * PURPOSE
+ *
+ *   - Solve Riemann problem using the Lax-Friedrichs 
+ *     Rusanov solver:
+ *
+ *     F(i+1/2) = [FL + FR - c*(UR - UL)]*0.5
+ *
+ *     where c = max_speed[(UR + UL)*0.5]
+ *
+ *   - On input, it takes left and right primitive state
+ *     vectors state->vL and state->vR at zone edge i+1/2;
+ *     On output, return flux and pressure vectors at the
+ *     same interface.
+ *
+ *   - Also, compute maximum wave propagation speed (cmax) 
+ *     for  explicit time step computation
+ *  
+ *
+ *
+ * LAST_MODIFIED
+ *
+ *   April 4th 2006, by Andrea Mignone  (mignone@to.astro.it)
+ *              
+ *
+ **************************************************************************** */
+
+{
+  int    nv, i;
+  double *uL, *uR, cL, cR;
+  static double **fL, **fR, **vRL;
+  static double *cRL, *pR, *pL, *a2L, *a2R, *hL, *hR;
+  static double *cminL, *cmaxL;
+  static double *cminR, *cmaxR;
+
+  if (fR == NULL){
+    fR  = ARRAY_2D(NMAX_POINT, NFLX, double);
+    fL  = ARRAY_2D(NMAX_POINT, NFLX, double);
+    cRL = ARRAY_1D(NMAX_POINT, double);
+    pR  = ARRAY_1D(NMAX_POINT, double);
+    pL  = ARRAY_1D(NMAX_POINT, double);
+    cminL = ARRAY_1D(NMAX_POINT, double);
+    cmaxL = ARRAY_1D(NMAX_POINT, double);
+    cminR = ARRAY_1D(NMAX_POINT, double);
+    cmaxR = ARRAY_1D(NMAX_POINT, double);
+
+    a2R = ARRAY_1D(NMAX_POINT, double);
+    a2L = ARRAY_1D(NMAX_POINT, double);
+
+    hR  = ARRAY_1D(NMAX_POINT, double);
[…]

### 2014-10-20 sims/sgPLUTO/Src/RMHD/eos.c

@@ -1,0 +1,98 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*! 
+  \file
+  \brief Collect various thermodynamic functions for the relativistic
+         MHD module.
+                    
+  \author A. Mignone (mignone@ph.unito.it)
+  \date   Aug 16, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+#include "pluto.h"
+
+/* ********************************************************************* */
+void SoundSpeed2 (double **v, double *cs2, double *h, int beg, int end,
+                  int pos, Grid *grid)
+/*!
+ * Define the square of the sound speed.
+ * 
+ * \param [in]    u   1D array of primitive quantities
+ * \param [out] cs2   1D array containing the square of the sound speed
+ * \param [in]    h   1D array of enthalpy values
+ * \param [in]  beg   initial index of computation 
+ * \param [in]  end   final   index of computation
+ * \param [in]  pos   an integer specifying the spatial position 
+ *                    inside the cell (only for spatially-dependent EOS)
+ * \param [in]  grid  pointer to Grid structure
+ *
+ * \return  This function has no return value.
+ *********************************************************************** */
+{
+  int   i;
+  double  theta;
+
+  Enthalpy(v, h, beg, end);
+
+  for (i = beg; i <= end; i++) {
+    theta = v[i][PRS]/v[i][RHO];
+    #if EOS == IDEAL
+     cs2[i] = g_gamma*theta/h[i];
+    #elif EOS == TAUB
+     cs2[i] = theta/(3.0*h[i])*(5.0*h[i] - 8.0*theta)/(h[i] - theta);
+    #endif
+  }
+}
+
+/* ********************************************************************* */
+void Enthalpy (double **v, double *h, int beg, int end)
+/*!
+ * Compute the enthalpy.
+ *
+ * \param [in]    v   1D array of primitive quantities
+ * \param [in]    h   1D array of enthalpy values
+ * \param [in]  beg   initial index of computation 
+ * \param [in]  end   final   index of computation
+ *
+ * \return  This function has no return value.
+ *********************************************************************** */
+{
+  int   i;
+  double  theta, gmmr;
+
+  gmmr = g_gamma/(g_gamma - 1.0);
+  for (i = beg; i <= end; i++) {
[…]

### 2014-10-20 sims/sgPLUTO/Src/RMHD/hlld.c

@@ -1,0 +1,712 @@
+#include"pluto.h"
+#define MAX_ITER  20
+#define COUNT_FAILURES  NO
+
+typedef struct RIEMANN_STATE{
+  int    fail;
+  double vx, vy, vz;
+  double Bx, By, Bz;
+  double Kx, Ky, Kz, K2;
+  double w, sw, p, rho;
+  double u[NFLX], R[NVAR];
+  double S, Sa;
+/*
+  double fun1, fun2, fun3; 
+  double denL, denR;
+*/
+} Riemann_State;
+
+
+static double Sc, Bx;
+static double Fstar (Riemann_State *PaL, Riemann_State *PaR, double p);
+
+static int GET_RIEMANN_STATE (Riemann_State *Pv, double p, int);
+static void GET_ASTATE (Riemann_State *Pa,  double p);
+static void GET_CSTATE (Riemann_State *PaL, Riemann_State *PaR, double p, double *Uc);
+
+static double PTOT  (double *V);
+
+static void PRINT_STATES(double *vL, double *vR);
+static void PRINT_WHATSWRONG(Riemann_State *PaL, Riemann_State *PaR, 
+                      double phll, double *vL, double *vR);
+
+#define DEBUG  NO
+
+/* ********************************************************************* */
+void HLLD_Solver (const State_1D *state, int beg, int end, 
+                  real *cmax, Grid *grid)
+/*
+ *
+ * NAME
+ *
+ *   HLLD_SOLVER
+ *
+ *
+ * PURPOSE
+ *
+ *   Solve riemann problem for the relativistic MHD equations 
+ *   using th HLLD solver;
+ * 
+ *   Reference:  "A five wave Harte-Lax-van Leer Riemann solver 
+ *                for relativistic magnetohydrodynamics"
+ *                Mignone et al, MNRAS (2009) 393,1141.
+ *   
+ *
+ * LAST_MODIFIED
+ *
+ *   Feb 15/2010, by Andrea Mignone  (mignone@ph.unito.it)
+ *              
+ *
+ *
+ **************************************************************************** */
+{
+  int    nv, i, k;
+  int    switch_to_hll;
+  real   scrh;
[…]

### 2014-10-20 sims/sgPLUTO/Src/RMHD/mod_defs.h

@@ -1,0 +1,252 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*!
+  \file
+  \brief Module header file for relativistic MHD (RMHD).
+
+  Set label, indexes and basic prototyping for the relativistic 
+  MHD module.
+
+
+  \authors A. Mignone (mignone@ph.unito.it)\n
+           C. Zanni   (zanni@oato.inaf.it)\n
+  \date    Oct 5, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+
+#ifndef RESISTIVE_RMHD
+ #define RESISTIVE_RMHD   NO
+#endif
+
+/* Set variable name labels.
+  We make extra vector components, when not needed, point 
+  to the last element (255) of the array stored by startup.c.  */
+
+enum {
+
+ #if COMPONENTS == 1
+
+  RHO, MX1, BX1, ENG, PRS = ENG,
+  #if MHD_FORMULATION == DIV_CLEANING
+   PSI_GLM,
+  #endif
+  MX2 = 255, BX2 = 255, MX3 = 255, BX3 = 255,
+
+ #elif COMPONENTS == 2
+
+  RHO, MX1, MX2, BX1, BX2, ENG, PRS = ENG,
+  #if MHD_FORMULATION == DIV_CLEANING
+   PSI_GLM,
+  #endif
+  MX3 = 255, BX3 = 255,
+
+ #elif COMPONENTS == 3
+
+  RHO, MX1, MX2, MX3, BX1, BX2, BX3, ENG, PRS = ENG,
+  #if MHD_FORMULATION == DIV_CLEANING
+   PSI_GLM,
+  #endif
+
+ #endif
+
+ VX1 = MX1, VX2 = MX2, VX3 = MX3, 
+
+/* -- backward compatibility -- */
+
+ DN = RHO, PR = PRS, EN = ENG,
+ VX = VX1, VY = VX2, VZ = VX3,
+ MX = MX1, MY = MX2, MZ = MX3,
+ BX = BX1, BY = BX2, BZ = BX3
+
+};
+
+
+/* **************************************
+     add the PSI_GLM label if necessary 
+   ************************************** */
+/*
+#if MHD_FORMULATION == DIV_CLEANING 
+ #define PSI_GLM (2 + 2*COMPONENTS)
+ #define NFLX (2 + 2*COMPONENTS + 1)
+#else
[…]

### 2014-10-20 sims/sgPLUTO/Src/RMHD/source.c

@@ -1,0 +1,181 @@
+#include "pluto.h"
+
+#if MHD_FORMULATION == EIGHT_WAVES
+/* *************************************************************************  */
+void POWELL_DIVB_SOURCE(const State_1D *state, int is, int ie, Grid *grid)
+/*
+ *
+ *
+ ************************************************************************** */
+{
+  int    i;
+  real divB;
+  real g_2, vB, *vc;
+  Grid   *GG;
+
+  GG = grid + g_dir;
+  
+/* ------------------------------------------------------------
+              POWELL's divB monopole source term
+   ------------------------------------------------------------ */
+
+  for (i = is; i <= ie; i++) {
+
+    vc = state->vh[i];
+
+/*     Make (divB) contribution from the normal direction (1)     */
+
+    divB = (state->vR[i][BXn]     + state->vL[i][BXn])*GG->A[i] -
+           (state->vR[i - 1][BXn] + state->vL[i - 1][BXn])*GG->A[i - 1];
+/*
+    divB = state->bn[i]*GG->A[i] - state->bn[i - 1]*GG->A[i - 1];
+*/
+    divB /= 2.0*GG->dV[i];
+
+    vB  = EXPAND(vc[VX1]*vc[BX1], + vc[VX2]*vc[BX2], + vc[VX3]*vc[BX3]);
+    g_2 = EXPAND(vc[VX1]*vc[VX1], + vc[VX2]*vc[VX2], + vc[VX3]*vc[VX3]);
+    g_2 = 1.0 - g_2;
+
+    EXPAND (state->src[i][MX1] = -divB*(vc[BX1]*g_2 + vB*vc[VX1]);  ,
+            state->src[i][MX2] = -divB*(vc[BX2]*g_2 + vB*vc[VX2]);  ,
+            state->src[i][MX3] = -divB*(vc[BX3]*g_2 + vB*vc[VX3]);)
+
+    state->src[i][ENG] = -divB*vB;
+
+    EXPAND (state->src[i][BX1] = -divB*vc[VX1];  ,
+            state->src[i][BX2] = -divB*vc[VX2];  ,
+            state->src[i][BX3] = -divB*vc[VX3];)
+
+  }
+}
+
+/* *********************************************************************  */
+void HLL_DIVB_SOURCE (const State_1D *state, real **Uhll, int beg, 
[…]

### 2014-10-20 sims/sgPLUTO/Src/States/states_limo3.c

@@ -1,0 +1,261 @@
+#include "pluto.h"
+static double LIM_FUNC (double dvp, double dvm, double);
+
+/* ************************************************************* */
+void States (const State_1D *state, int beg, int end, Grid *grid)
+/* 
+ *
+ * PURPOSE
+ *
+ *   Provide a three-point stencil, third-order 
+ *   reconstruction algorithm based on the limiter 
+ *   function introduced  by:
+ *
+ *   "Compact third-order limiter functions for finite volume
+ *    methods", Cada & Torrilhon, JCP (2009) 228, 4118.
+ *
+ * LAST MODIFIED
+ *
+ *   Nov 3rd 2009
+ *   by A. Mignone (mignone@ph.unito.it)
+ *
+ **************************************************************** */
+{
+  int    k, nv, i;
+  double dmm;
+  double **v, *vp, *vm, *dvp, *dvm, *dx;
+  static double **dv;
+  double **L, **R, *lambda;
+  double dwp[NVAR], dwp_lim[NVAR];
+  double dwm[NVAR], dwm_lim[NVAR];
+  double dvpR, dvmR;
+
+  if (dv == NULL) {
+    dv = ARRAY_2D(NMAX_POINT, NVAR, double);
+  }
+
+  v  = state->v;
+  dx = grid[g_dir].dx;
+
+/* ----------------------------------------------------
+    compute slopes and left and right interface values 
+   ---------------------------------------------------- */
+
+  #if CHAR_LIMITING == NO  /* ----------------------------------------
+                                 Limiter on primitive variables
+                              ----------------------------------------  */
+   for (i = beg-1; i <= end; i++){
+   for (nv = NVAR; nv--; ){
+     dv[i][nv] = v[i+1][nv] - v[i][nv];
+   }}
+
+   for (i = beg; i <= end; i++){
+     dvp = dv[i];   vp = state->vp[i];
+     dvm = dv[i-1]; vm = state->vm[i];
+
+     #if SHOCK_FLATTENING == MULTID    
+      if (CheckZone (i, FLAG_MINMOD)) {
+        for (nv = NVAR; nv--;    ) {
+          dmm = MINMOD(dvp[nv], dvm[nv]);
[…]

### 2014-10-20 sims/sgPLUTO/Src/Thermal_Conduction/tc_flux.c

@@ -1,0 +1,178 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*!
+  \file
+  \brief Compute the thermal conduction flux.
+
+  Compute the thermal conduction flux along one row of computational 
+  zones for the HD and MHD  modules according to Spitzer (1962):
+  \f[
+     F_c = F_{\rm class}/(|F_{\rm class}| + q)
+  \f]
+  where \f$ F_{\rm class} \f$ is the classical thermal conduction flux, 
+  \f$ q \f$ is the saturated flux.
+  Since the first term is purely parabolic, it is discretized using standard
+  finite difference. 
+  The saturated flux is treated in an upwind manner following the guidelines
+  given in Balsara (2008) (see also Mignone et al. 2012)
+
+  The classical MHD flux further splits into 2 components, along and 
+  across the magnetic field lines.
+  This function also computes the inverse of the time
+  step and return its maximum over the current sweep.
+  
+  \b References
+     - "The PLUTO Code for Adaptive Mesh Computations in Astrophysical 
+        Fluid Dynamics" \n
+       Mignone et al, ApJS (2012) 198, 7M
+       
+  \authors A. Mignone (mignone@ph.unito.it)\n
+           T. Matsakos
+  \date   Sep 13, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+#include "pluto.h"
+
+/*! When set to YES, saturated flux is computed using an upwind 
+    selection rule.
+    When set to NO, staurated flux is treated in the same manner as 
+    the  conduction flux.                                          */ 
+#define HYPERBOLIC_SAT_FLUX YES 
+
+/* ********************************************************************* */
+void TC_Flux (double ***T, const State_1D *state, double **dcoeff, 
+             int beg, int end, Grid *grid)
+/*! 
+ * Compute the thermal conduction flux, state->par_flx.
+ *
+ * \param [in]     T       3D array containing the dimensionless 
[…]

### 2014-10-20 sims/sgPLUTO/Src/Time_Stepping/unsplit_ctu.c

@@ -1,0 +1,686 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*!
+  \file
+  \brief Advance equations using Corner Transport Upwind (CTU).
+
+  Implement the dimensionally unsplit, Corner Transport Upwind method 
+  (CTU) of Colella. It consists of
+ 
+  <b> A predictor step </b>: 
+  -# start from cell center values of \c V and compute time-centered 
+     interface states using either HANCOCK or CHARACTERISTIC_TRACING:
+     \f[
+       V^{n+\HALF}_{i,\pm} = V^n_{i,\pm} + \pd{V}{t}\frac{\Delta t}{2}
+     \f] 
+     Store the resulting states by converting \f$ V_\pm \f$ in \f$ U_\pm \f$
+     (= \c UP, \c UM).
+     Normal (n) and trasnverse (t) indexes for this step should span
+     \f$ n \in [n_{\rm beg} - 1, n_{\rm end} + 1] \f$, 
+     \f$ t \in [t_{\rm beg} - 1, t_{\rm end} + 1] \f$ 
+     for cell-centered schemes and
+     \f$ n \in [n_{\rm beg} - 2, n_{\rm end} + 2] \f$,  
+     \f$ t \in [t_{\rm beg} - 2, t_{\rm end} + 2] \f$ for staggered MHD. 
+  -# Solve Riemann problems between \f$ V_{i,+}\f$ and \f$ V_{i+1,-}\f$  
+     and store the flux difference in RHS.
+  -# Compute cell and time-centered values UH = U + dt/2*RHS with index
+     range \f$ n \in [n_{\rm beg}, n_{\rm end}]\f$, 
+           \f$ t \in [t_{\rm beg}, t_{\rm end}]\f$ for cell-centered schemes
+     and   \f$ n \in [n_{\rm beg}-1, n_{\rm end}+1]\f$, 
+           \f$ t \in [t_{\rm beg}-1, t_{\rm end}+1]\f$ for staggered MHD.
+   
+  <b> A corrector step </b>: 
+  -# correct left and right states \c UP and \c UM computed in the previous 
+     step by adding the transverse RHS contribution, e.g.
+     \f$ U^{n+\HALF}_{i\pm} += ({\cal RHS}^n_y + {\cal RHS}^n_z)\Delta t/2 \f$.
[…]

### 2014-10-20 sims/sgPLUTO/Src/Viscosity/viscosity.h

@@ -1,0 +1,8 @@
+/* -----------------------------------------------
+             Viscosity header
+   ----------------------------------------------- */
+
+void ViscousFlux (Data_Arr , real **, real **, double **, int, int, Grid *);
+void eta_visc_func(real *, real , real , real, real *, real *);
+
+

### 2014-10-20 sims/sgPLUTO/Src/cylsource.c

@@ -1,0 +1,710 @@
+#include"pluto.h"
+
+static void CONS_CYLSOURCE   (double *u, double *);
+static double ROS34_SOLVE (real *v0, real *k1, real *v4th, real dt, real tol);
+
+static void CYLSOURCE   (double *u, double *);
+static void CYLJACOBIAN (double *v, double **);
+
+static double rad;
+
+/* ***************************************************** */
+void CONS_CYLSOLVE (Data_Arr VV, double dt, Grid *grid)
+/*
+ *
+ *
+ ******************************************************* */
+{
+  int i,j,k,nv;
+  double u0[NVAR], u1[NVAR];
+  double k1[NVAR], k2[NVAR], k3[NVAR], k4[NVAR];
+  static double **u, **v;
+  static unsigned char *flag;
+
+  #if GEOMETRY != CARTESIAN
+   return;
+  #endif
+
+  if (u == NULL){
+    u = ARRAY_2D(2*NMAX_POINT,NVAR, double);
+    v = ARRAY_2D(2*NMAX_POINT,NVAR, double);
+    flag = ARRAY_1D(2*NMAX_POINT, unsigned char);
+  }
+
+  KTOT_LOOP(k){ 
+  JTOT_LOOP(j){
+
+  /* --------------------------------------
+      convert to conservative vars
+     -------------------------------------- */
+
+    ITOT_LOOP(i){
+      for (nv = NVAR; nv--;  ) v[i][nv] = VV[nv][k][j][i];
+    }
+    PrimToCons(v, u, 0, NX1_TOT-1);
+
+  /* --------------------------------------
+       Integrate
+     -------------------------------------- */
+
+    ITOT_LOOP(i){      
+      rad = grid[IDIR].x[i];
+      for (nv = 0; nv < NVAR; nv++) u0[nv] = u[i][nv];
+      CONS_CYLSOURCE(u0, k1);
+      for (nv = NVAR; nv--; ) u1[nv] = u0[nv] + 0.5*dt*k1[nv]; 
+
+      CONS_CYLSOURCE(u1, k2);
+      for (nv = NVAR; nv--; ){
+        u[i][nv] = u0[nv] + dt*k2[nv];
+      }
+
[…]

### 2014-10-20 sims/sgPLUTO/Src/fd_flux.c

@@ -1,0 +1,345 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*!
+  \file
+  \brief Compute finite difference fluxes.
+  
+  The function FD_Flux() computes a high-order approximation to 
+  the flux divergence using WENO or Monotonicity-preserving 
+  reconstruction in characteristic variables.
+  
+  It is called as a regular Riemann solver function except 
+  that the left and right states do not need to be defined 
+  since high-order reconstruction is done on the characteristic
+  projection of the cell-centered fluxes.
+
+  \authors A. Mignone (mignone@ph.unito.it)\n
+           P. Tzeferacos (petros.tzeferacos@ph.unito.it)
+  \date   Sep 14, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+#include "pluto.h"
+
+/* ********************************************************************* */
+void FD_Flux (const State_1D *state, int beg, int end, 
+               double *cmax, Grid *grid)
+/*!
+ * Compute interface flux by suitable high-order finite
+ * difference non-oscillatory interpolants.
+ *
+ * \param [in]  state   pointer to State_1D structure
+ * \param [in]    beg   initial index of computation
+ * \param [in]    end   final   index of computation
+ * \param [out]  cmax   array of maximum characteristic speeds
+ * \param [in]    grid  pointer to an array of Grid structures
+ *
+ * \b Reference
+ *    - "High-order conservative finite difference GLM-MHD schemes
+ *     for cell-centered MHD"\n
+ *      Mignone et al., JCP (2010), 229, 5896
+ *
+ * \return This function has no return value.    
+ * 
+ ****************************************************************** */
+{
+  int    nv, i, j, k, np_tot;
+  int    Kmax, S;
+  double **v, **flux, *press;
+  double fs, fp, fm, flx[NFLX], dx;
+
+  static double *Fp, *Fm, **F, *a2, **u;
[…]

### 2014-10-20 sims/sgPLUTO/Src/globals.h

@@ -1,0 +1,151 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*! 
+  \file  
+  \brief Collects global variables definitions.
+
+  This file contains definitions for all global variables 
+  (visible anywhere in the code) used by PLUTO. 
+  Global variables names, by convention, are prefixed with a "g_" 
+  unless they're used as constants throughout the code in which case 
+  they keep the full-capitalized notation typical of macros.
+  
+  For modules, global variables are prefixed with the initial letters
+  of the module name, e.g., \c sb_vy or \c glm_ch.
+  
+  In the following "local" means "for the local processor".
+  "Interior" means inside the computational domain.
+
+  \author A. Mignone (mignone@ph.unito.it)
+  \date   Sep 16, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+
+ int SZ;
+ int SZ_stagx;
+ int SZ_stagy;
+ int SZ_stagz;
+ int SZ_float;
+ int SZ_char;
+ int SZ_Float_Vect;
+ int SZ_rgb;
+ int SZ_short;
+
+int prank; /**< Processor rank. In serial mode it is defined to be 0. */
+
+long int IBEG; /**< Lower grid index of the computational domain in the 
+                    the X1 direction for the local processor. */
+long int IEND; /**< Upper grid index of the computational domain in the 
+                    the X1 direction for the local processor. */
+long int JBEG; /**< Lower grid index of the computational domain in the 
+                    the X2 direction for the local processor. */
+long int JEND; /**< Upper grid index of the computational domain in the 
+                    the X2 direction for the local processor. */
+long int KBEG; /**< Lower grid index of the computational domain in the 
+                    the X3 direction for the local processor. */
+long int KEND; /**< Upper grid index of the computational domain in the 
+                    the X3 direction for the local processor. */
+
+long int NX1; /**< Number of interior zones in the X1 directions (boundaries
+                   \e excluded) for the local processor. */
+long int NX2; /**< Number of interior zones in the X2 directions (boundaries
+                   \e excluded) for the local processor. */
+long int NX3; /**< Number of interior zones in the X3 directions (boundaries
+                   \e excluded) for the local processor. */
+
+long int NX1_TOT; /**< Total number of zones in the X1 direction (boundaries
+                       \e included) for the local processor.*/
[…]

### 2014-10-20 sims/sgPLUTO/Src/input_data.c

@@ -1,0 +1,481 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*! 
+  \file  
+  \brief Provide basic functionality for reading input data files.
+
+  Collects a number of functions for opening, reading and assigning 
+  initial conditions from user-supplied data file(s).
+  The geometry and dimensions of the input grid can be different from 
+  the actual grid employed by PLUTO, as long as the coordinate geometry
+  transformation has been implemented.
+  The input grid and data files should employ the same format and 
+  conventions employed by PLUTO. 
+  
+  - Gridfile: coordinates should be written using the PLUTO 4.0 grid format.
+  - Datafile: variables should be written in sequence in a single binary 
+              file using single or double precision. 
+              The file extension must be ".flt" or ".dbl" for the former and
+              the latter, respectively.
+     
+    Note that not all of the variables should be present and the input
+    array ::get_var specifies which one are to be searched for.
+
+  The InputDataSet() initialize the module and by assigning values to 
+  global variables such as size, geometry and dimensions of the input grid.
+  Data values are read through the function InputDataRead() while
+  InputDataInterpolate() can be finally used to map input data onto the
+  grid employed by PLUTO using bi- or tri-linear interpolation to fill the 
+  data array at the desired coordinate location.
+
+  \authors A. Mignone (mignone@ph.unito.it)\n
+           P. Tzeferacos 
+  \date   Aug 27, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+#include"pluto.h"
+#define ID_MAX_NVAR 256 
+
+static int id_nvar; /**< Number of variables to be read on input. */
+static int id_var_indx[ID_MAX_NVAR]; /**< The variable index. */
+static int id_nx1; /**< Size of input grid in the x1 direction. */
+static int id_nx2; /**< Size of input grid in the x2 direction. */
+static int id_nx3; /**< Size of input grid in the x3 direction. */
+
+static int id_geom;   /**< Geometry of the input grid. */
+
[…]

### 2014-10-20 sims/sgPLUTO/Src/prototypes.h

@@ -1,0 +1,268 @@
+/* ---------------------------------------------------------------------
+                    General function prototypes 
+   --------------------------------------------------------------------- */
+
+void   AdvectFlux (const State_1D *, int, int, Grid *);
+void   Analysis (const Data *, Grid *);
+#if EOS == BAROTROPIC
+ double BAROTROPIC_PR(double);
+#endif
+double BodyForcePotential(double, double, double);
+void   BodyForceVector(double *, double *, double, double, double);
+
+void  ChangeDumpVar ();
+void  CheckConsStates (double **, double **, double **, int, int);
+void  CheckPrimStates (double **, double **, double **, int, int);
+int   CheckNaN (double **, int, int, int);
+unsigned char CheckZone (int z, int bit);
+void  ComputeUserVar (const Data *, Grid *);
+float ***Convert_dbl2flt (double ***, int);
+
+void ConsToPrim3D(const Data *, int);
+void PrimToCons3D(const Data *, int);
+void CreateImage (char *);
+
+void EntropySwitch (const Data *, Grid *);
+
+#ifdef FINITE_DIFFERENCE  
+ Riemann_Solver FD_Flux;
+ Reconstruct MP5_Reconstruct, PPM_Reconstruct, LIMO3_Reconstruct,
+             WENOZ_Reconstruct, WENO3_Reconstruct;
+ void FD_GetMaxEigenvalues (const Data *d, State_1D *state, Grid *grid);
+#endif
+
+void FindShock (const Data *, Grid *);
+void FlagReset (const Data *d);
+void Flatten (const State_1D *, int, int, Grid *);
+void FreeGrid (Grid *);
+
+Image  *GetImage (char *);
+double *GetInverse_dl (const Grid *);
+int    GetNghost (Input *);
+double ***GetUserVar (char *);
+
+void Init (double *, double, double, double);
+void Initialize(int argc, char *argv[], Data *, Input *, Grid *, Cmd_Line *);
+
+void InternalBoundaryReset (const State_1D *, Time_Step *, int, int, Grid *);
+
+void InputDataSet (char *, int *);
+void InputDataRead (char *);
+void InputDataFree (void);
+
+int  IsLittleEndian (void);
+
+void MakeState (State_1D *);
+void MakeGeometry (Grid *);
+double Median (double a, double b, double c);
+
+void   ParabolicFlux(Data_Arr, const State_1D *, double **, int, int, Grid *);
[…]

### 2014-10-20 sims/sgPLUTO/Src/restart.c

@@ -1,0 +1,291 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*! 
+  \file  
+  \brief Restart PLUTO from binary or HDF5 double precision data files.
+ 
+  This file collects the necessary functions for restarting PLUTO 
+  from a double precision binary or HDF5 file in the static grid
+  version of the code.
+
+  \author A. Mignone (mignone@ph.unito.it)
+  \date   Aug 16, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+#include "pluto.h"
+
+/* *********************************************************************  */
+void Restart (Input *ini, int nrestart, int type, Grid *grid)
+/*!
+ * Read input binary / hdf5 data.
+ *
+ * \param [in] ini
+ * \param [in] nrestart
+ * \param [in] type specifies the output data type (type should be 
+ *             either DBL_OUTPUT or DBL_H5_OUTPUT).
+ * \param [in]      grid  pointer to an array of Grid structures
+ *
+ ***********************************************************************  */
+{
+  int     i,j,k;
+  int     nv, single_file, origin, nlines=0;
+  int     swap_endian=0;
+  char    fname[512], str[512];
+  double  dbl;
+  void   *Vpt;
+  Output *output;
+  FILE   *fbin;
+
+/* ----------------------------------------------------------
+    get the pointer to the output format specified by "type"
+   ---------------------------------------------------------- */
+
+  for (nv = 0; nv < MAX_OUTPUT_TYPES; nv++){
+    output = ini->output + nv;
+    if (output->type == type) break;
+  }
+
+/* -------------------------------------------------------
+    compare the endianity of the restart file (by reading
+    the corresponding entry in dbl.out or dbl.h5.out) 
+    with that of the current architecture.
+    Turn swap_endian to 1 if they're different.
+   ------------------------------------------------------- */
+
+  if (prank == 0){
+    if      (type == DBL_OUTPUT) {
+      fbin = fopen ("dbl.out", "r");
+    } else if (type == DBL_H5_OUTPUT) {
+      fbin = fopen ("dbl.h5.out", "r");
+    }
+    if (fbin == NULL){
[…]

### 2014-10-20 sims/sgPLUTO/Src/set_geometry.c

@@ -1,0 +1,345 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*!
+  \file
+  \brief Initialize geometry-dependent grid quantities.
+
+  Compute grid quantities (such as interface area, volume, geometrical
+  center, etc..) that depend on the geometry.
+
+  \author A. Mignone (mignone@ph.unito.it)
+  \date   Sep 23, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+#include "pluto.h"
+
+/* ********************************************************************* */
+void MakeGeometry (Grid *GXYZ)
+/*!
+ *
+ * \param [in,out] GXYZ  Pointer to an array of Grid structures;
+ *
+ *********************************************************************** */
+{
+  int     i, j, k, idim, ngh, ileft;
+  int     iright;
+  int     iend, jend, kend;
+  double  xiL, xiR, dxi, dvol;
+  double  x, dx, xr, xl;
+  double  y, dy, yr, yl;
+  struct  GRID *GG;
+
+  iend = GXYZ[0].lend + GXYZ[0].nghost;
+  jend = GXYZ[1].lend + GXYZ[1].nghost;
+  kend = GXYZ[2].lend + GXYZ[2].nghost;
+
+/*  --------------------------------------------------------------
+     Memory allocation. All values are defined at the cell center 
+     with the exception of the area element which is defined on a
+     staggered mesh and therefore starts at [-1].
+    ----------------------------------------------------------- */
+
+  for (idim = 0; idim < 3; idim++) {
+    (GXYZ + idim)->A       = ARRAY_1D (GXYZ[idim].np_tot+1, double)+1;
+    (GXYZ + idim)->xgc     = ARRAY_1D (GXYZ[idim].np_tot, double);
+    (GXYZ + idim)->dV      = ARRAY_1D (GXYZ[idim].np_tot, double);
[…]

### 2014-10-20 sims/sgPLUTO/Src/setup.c

@@ -1,0 +1,437 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*!
+  \file
+  \brief Read runtime information from pluto.ini.
+
+  Parse and read runtime information from the initialization file 
+  pluto.ini (default) and sets value of the input structure.
+
+  \authors A. Mignone (mignone@ph.unito.it)
+  \date    Oct 29, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+#include "pluto.h"
+
+static void GetOutputFrequency(Output *, char *);
+
+#define NOPT      32              /*  # of possible options in a menu */
+#define NLEN      128              /*  # default string length         */
+
+#define COMPARE(s1,s2,ii) \
+        for ( (ii) = 1 ; (ii) < NOPT && !(strcmp ( (s1), (s2)) == 0); (ii)++);
+
+/* ********************************************************************* */
+int Setup (Input *input, Cmd_Line *cmd_line, char *ini_file)
+/*!
+ * Open and parse the initialization file. 
+ * Assign values to the input structure.
+ *
+ * \param [out]  input     pointer to an Input structure
+ * \param [in]   cmd_line  pointer to a Cmd_Line structure (useful, e.g.,
+ *                         to resize the domain using the \c -xres option)
+ * \param [in]   ini_file  the name of the initialization file (default
+ *                         is "pluto.ini") specified with the \c -i option.
+ *
+ *********************************************************************** */
+{
+  int    idim, ip, ipos, itype, nlines;
+  char   *bound_opt[NOPT], str_var[512], *str;
+  char  *glabel[]     = {"X1-grid", "X2-grid","X3-grid"};
+  char  *bbeg_label[] = {"X1-beg", "X2-beg","X3-beg"};
+  char  *bend_label[] = {"X1-end", "X2-end","X3-end"};
+  double dbl_var, rx;
+  Output *output;
+  FILE *fp;
+
+  print1 ("> Reading %s (Setup) ...\n\n", ini_file);
+
+  for (itype = 0; itype < NOPT; itype++) {
+    bound_opt[itype] = "0000";
+  }
+
+/*  ---------------------------------------------------
+      available options are given as two set of names;
[…]

### 2014-10-20 sims/sgPLUTO/Src/show_config.c

@@ -1,0 +1,406 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*! 
+  \file  
+  \brief Print useful information about the current computations.
+
+  Detailed description goes here.         
+
+  \author A. Mignone (mignone@ph.unito.it)
+  \date   Aug 16, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+#include "pluto.h"
+static void CheckConfig();
+
+/* ********************************************************************* */
+void ShowConfig ()
+/*!
+ *  Write a summary of the selected options
+ *
+      ___  __   __  ____________ 
+     / _ \/ /  / / / /_  __/ __ \
+    / ___/ /__/ /_/ / / / / /_/ /
+   / /  /____/\____/ /_/  \____/ 
+   ============================== 
+                                  
+ *
+ * 
+ *                        
+ *
+ *********************************************************************** */
+{
+  FILE *fconf;
+  time_t time_now;
+  char  str1[128], str2[128], str3[128];
+
+  CheckConfig();
+
+  print1 ("\n");
+  print1("   ___  __   __  ____________   \n");
+  print1("  / _ \\/ /  / / / /_  __/ __ \\ \n");
+  print1(" / ___/ /__/ /_/ / / / / /_/ /  \n");
+  print1("/ /  /____/\\____/ /_/  \\____/   \n");
+  print1("=============================    v. %s  \n", PLUTO_VERSION);
+  print1 ("\n> System:\n\n");
+
+  if ( (fconf = fopen("sysconf.out","r")) != NULL){
+
+    while (fscanf (fconf, "%s %s %s\n", str1, str2, str3) != EOF) {
+
+      if (!strcmp(str1,"USER")) 
+        print1 ("  %s:             %s\n",str1, str3);
+      else if (!strcmp(str1,"WORKING_g_dir"))
+        print1 ("  %s:      %s\n",str1, str3);
[…]

### 2014-10-20 sims/sgPLUTO/Src/startup.c

@@ -1,0 +1,244 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*! 
+  \file  
+  \brief Loop on the computational cells to assign initial conditions.
+
+  This function is called anyway, even if restart from file is enabled.
+  This is useful to initialized a number of global variables and/or 
+  user-defined parameters by calling Init().
+
+  \author A. Mignone (mignone@ph.unito.it)
+  \date   Aug 16, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+#include "pluto.h"
+
+/* ********************************************************************* */
+void Startup (Data *d, Grid *G)
+/*! 
+ *
+ *
+ *
+ *
+ *********************************************************************** */
+{
+  int i, j, k;
+  int isub, jsub, ksub, nsub = 5;
+  int nv,  l_convert;
+  static double **ucons, **uprim;
+  double x1,  x2,  x3;
+  double x1p, x2p, x3p;
+  double x1s, x2s, x3s;
+  double dx1, dx2, dx3;
+  double us[256], u_av[256], b[3]; 
+  double scrh;
+  struct GRID *GX, *GY, *GZ;
+
+  GX = G;
+  GY = G + 1;
+  GZ = G + 2;
+
+  if (uprim == NULL){
+    uprim = ARRAY_2D(NMAX_POINT, NVAR, double);
+    ucons = ARRAY_2D(NMAX_POINT, NVAR, double); 
+  }
+
+/* ----  set labels  ---- */
+
+  EXPAND(MXn = VXn = VX1;  ,
+         MXt = VXt = VX2;  ,
+         MXb = VXb = VX3;)
+
+  #if PHYSICS == MHD || PHYSICS == RMHD
+   EXPAND(BXn = BX1;  ,
+          BXt = BX2;  ,
+          BXb = BX3;)
+  #endif
+
+/* ----------------------------------------------------------
+    If the -input-data-file command line argument is given, 
+    try to read initial conditions from external files.
+   ---------------------------------------------------------- */
+    
+  print1 ("> Assigning initial conditions (Startup) ...\n");
+
+/* --------------------------------------------------------------
[…]

### 2014-10-20 sims/sgPLUTO/Src/vec_pot_update.c

@@ -1,0 +1,207 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*!
+  \file 
+  \brief Function to handle vector potential.
+ 
+  Update the cell-centered vector potential by integrating
+  \f[
+     \pd{\vec{A}}{t} + \vec{E} = 0
+  \f] 
+  Depending on the time-stepping algorithm, we advance the solution
+  in time using the following increments
+  \f[
+   \begin{array}{ll}
+    \rm{RK2:}\quad &\DS A^{n+1} = A^n + \frac{\Delta t^n}{2} (E^n + E^*) 
+      \\ \noalign{\medskip}
+    \rm{RK3:}\quad &\DS A^{n+1} = A^n + \frac{\Delta t^n}{6} (E^n + E^* + 4E^{**}) 
+      \\ \noalign{\medskip}
+    \rm{CTU:}\quad &\DS A^{n+1} = A^n + \Delta t^nE^{n+\HALF}
+   \end{array}        
+  \f]
+  The cell-centered electric field \b E is computed by averaging values 
+  available at faces or edges, depending on the MHD formulation:
+  - staggered_mhd: we average the electric field values 
+                   available at cell edges and previously computed
+                   . In this
+                   case the function does the job only 
+                   if called from CT;
+  - Cell-cent. mhd: we average the face values using upwind fluxes. 
+                    In this case the function does the job only 
+                    if called from time stepping algorithms.
+                    The flux components of induction equation
+                    contain the emf: 
+                    \f$ F_{i+\HALF} = (   0,-E_z, E_y)_{i+\HALF}\f$,
+                    \f$ F_{j+\HALF} = ( E_z,   0,-E_x)_{j+\HALF}\f$,
+                    \f$ F_{k+\HALF} = (-E_y, E_x,   0)_{k+\HALF}\f$,
+                    
+ 
+  Note that the vector potential is NOT used in the rest
+  of the code and serves solely as a diagnostic tool.
+ 
+  \param [in,out] d   pointer to PLUTO Data structure
+  \param [in] vp      pointer to an EMF structure, used only with CT
+  \param [in] state   pointer to State_1D structure containing the 
[…]

### 2014-10-20 sims/sgPLUTO/Src/visc_flux.c

@@ -1,0 +1,45 @@
+#include "pluto.h"
+
+#define VISCOSITY_COEFF 0.1
+
+/* *********************************************************** */
+void VISC_FLUX (const State_1D *state, int beg, int end, Grid *grid)
+/* 
+ *
+ * PURPOSE
+ *
+ *   Add artificial viscosity term to the fluxes
+ *
+ *
+ ************************************************************* */
+{
+  int   i, nv;
+  real  scrh;
+  real  **flux, **v, **u;
+  double *uL, *uR, *vL, *vR;
+
+  flux = state->flux;
+
+ /* -- OLD version used cell-centered viscosity -- */
+/*
+  v    = state->v;
+  u    = state->u;
+  for (i = beg; i <= end; i++) {
+    scrh = VISCOSITY_COEFF*MAX(0.0, v[i][VXn] - v[i+1][VXn]);
+    for (nv = 0; nv < NFLX; nv++) {
+      flux[i][nv] += scrh*(u[i][nv] - u[i + 1][nv]);
+    }
+  }
+*/
+  for (i = beg; i <= end; i++) {
+    uL = state->uL[i]; uR = state->uR[i];
+    vL = state->vL[i]; vR = state->vR[i];
+    scrh = VISCOSITY_COEFF*(MAX(0.0, vL[VXn] - vR[VXn]));
+    for (nv = 0; nv < NFLX; nv++) {
+      flux[i][nv] += scrh*(uL[nv] - uR[nv]);
+    }
+  }
+
+}
+
+#undef VISCOSITY_COEFF

### 2014-10-20 sims/sgPLUTO/Src/write_img.c

@@ -1,0 +1,339 @@
+#include "pluto.h"
+#ifdef USE_PNG
+ #include <png.h>
+#endif
+
+static void GetSlice (double ***, Image *, Grid *);
+static int GET_SLICE_INDEX (int, double, Grid *);
+
+/* ****************************************************************** */
+void WritePPM (double ***Vdbl, char *var_name, char *filename, 
+                Grid *grid)
+
+/*
+ *
+ *
+ *
+ *
+ ******************************************************************** */
+{
+  int  ic, ir;
+  FILE *fl;
+  char header[512];
+  Image *ppm;
+
+  ppm = GetImage (var_name);
+  SetColorMap (ppm->r, ppm->g, ppm->b, ppm->colormap);
+  GetSlice (Vdbl, ppm, grid);
+  if (prank != 0) return;
+
+  sprintf (header,"P6\n%d %d\n255\n", ppm->ncol, ppm->nrow);
+  fl  = fopen (filename,"w");
+  fprintf(fl,"%s",header);
+  for (ir = 0; ir < ppm->nrow; ir++){
+    fwrite (ppm->rgb[ir], sizeof(RGB), ppm->ncol, fl);
+  } 
+  fclose(fl);
+}
+
+#ifdef USE_PNG
+/* ****************************************************************** */
+void WritePNG (double ***Vdbl, char *var_name, char *filename, 
+                Grid *grid)
+/*
+ *
+ *
+ *
+ *
+ ******************************************************************** */
+{
+  int ic, ir, i;
+  png_structp     png_ptr;
+  png_infop       info_ptr;
+  int backgroundcolour_;
+  int bit_depth_;
+  int colortype_;
+  int compressionlevel_;
+  int  indx;
+  double filegamma_;
+  Image *png;
+  unsigned char **image;
+  FILE   *fp;
+    
+  png = GetImage (var_name);
+  SetColorMap (png->r, png->g, png->b, png->colormap);
+  GetSlice (Vdbl, png, grid);
+  if (prank != 0) return;
+
+  image = (png_bytepp)malloc(png->nrow*sizeof(png_bytep));
+  for (ir = 0; ir < png->nrow; ir++) {
+    image[ir] = (png_bytep)malloc(6*png->ncol*sizeof(png_byte));
+  }
+
+  for(ic = 0; ic < png->ncol; ic++){
+  for(ir = 0; ir < png->nrow; ir++){
+    i = 6*ic;
[…]

### 2014-10-20 sims/sgPLUTO/runs/source/setup.py

@@ -1,0 +1,94 @@
+import os
+import os.path
+import sys
+
+# *****************************************************
+#   check options and compatibility between modules
+# *****************************************************
+
+get_arch = 0
+for x in sys.argv[1:]:
+
+  if (x == "--get-arch"):
+    print "Checking system architecture"
+    get_arch = 1
+    break
+
+  if (x == "--with-chombo" or x == "--with-chombo:"): 
+    print "Enabling Chombo support for AMR"
+    for y in sys.argv[1:]:
+      if (y == "--with-fd" or y == "--with-sb" or y == "--with-fargo"):
+        print "! Incompatible modules, ",x," + ",y
+        sys.exit(1)
+    break
+
+  elif (x == "--with-sb"): 
+    print "Enabling support for shearing box module"
+    for y in sys.argv[1:]:
+      if (y == "--with-fd"):
+        print "! Incompatible modules, ",x," + ",y
+        sys.exit(1)
+  
+  elif (x == "--with-fd"): 
+    print "Enabling support for finite difference module"
+
+  elif (x == "--with-fargo"): 
+    print "Enabling support for FARGO scheme"
+
+  elif (x == "--no-curses"):
+    print ""
+
+  elif (x == "--help" or x == "-help"):
+    print "Usage: python $sgPLUTO_DIR/setup.py [options]\n" 
+    print "Here [options] can be:\n"
+    print " --with-sb       enable the shearing box module."
+    print " --with-fd       enable the finite difference module."
+    print " --with-fargo    enable the FARGO-MHD module"
+    print " --with-chombo   enable support for adaptive mesh refinement."
+    print "                 (AMR) module using the Chombo library."
+    print " --no-curses     disable ncurses library and use a"
+    print "                 simpler text-based menu."
+    sys.exit(1)
+
+  else:
+    print "! Unrecognized option '",x,"'"
+    sys.exit(1)
+
+
+print 'loading...'
+
+# check if $sgPLUTO_DIR is defined
+
+try:
+  pluto_dir = os.environ["sgPLUTO_DIR"]
+except:
+  pluto_dir = os.getcwd()
+#  print "---------------------------------------------------------------"
+#  print " WARNING:"
+#  print "  "
+#  print " Your sgPLUTO_DIR shell variable does not seem to be defined. " 
[…]

### 2014-10-20 sims/sgPLUTO/runs/source/unsplit.c

@@ -1,0 +1,470 @@
+/* ///////////////////////////////////////////////////////////////////// */
+/*!
+  \file
+  \brief Advance equations using directionally-unsplit integrators 
+         (method of lines).
+
+  Main driver for RK unsplit integrations (DIMENSIONAL_SPLITTING == NO) 
+  and for finite difference methods (RK3).
+  Time stepping include Euler, RK2 and RK3.
+
+  \authors A. Mignone (mignone@ph.unito.it)\n
+           P. Tzeferacos (petros.tzeferacos@ph.unito.it)
+  \date   Oct 1, 2012
+*/
+/* ///////////////////////////////////////////////////////////////////// */
+#include "pluto.h"
+
+/* ********************************************************************* */
+int Unsplit (const Data *d, Riemann_Solver *Riemann, 
+             Time_Step *Dts, Grid *grid)
+/*!
+ * Advance the equations by a single time step using unsplit 
+ * integrators based on the method of lines.
+ *
+ * \param [in,out]      d  pointer to Data structure
+ * \param [in]    Riemann  pointer to a Riemann solver function
+ * \param [in,out]    Dts  pointer to time step structure
+ * \param [in]       grid  pointer to array of Grid structures
+ *    
+ *********************************************************************** */
+{
+  int  ii, jj, kk;
+  int  *i, *j, *k;
+  int  in, nv;
+  
+  static double  one_third = 1.0/3.0;
+  static State_1D state;
+  static Data_Arr UU, UU_1;
+  double dt;
+  double *inv_dl, dl2;
+  static double ***C_dt[NVAR], **dcoeff;
+  Index indx;
+
+/* ----------------------------------------------------
+                   Allocate memory 
+   ---------------------------------------------------- */
+
+  if (state.rhs == NULL){
+    MakeState (&state);
[…]

### 2014-10-20 sims/zeusGAP/src_dead/advx2.F

@@ -1,0 +1,369 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////                 A D V X 2                 \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c               University of Illinois at Urbana-Champaign
+c
+c=======================================================================
+c
+       subroutine advx2 (dlo,den
+     &                  ,eod,edn
+     &                  ,ero,ern
+     &                  ,abo,abn
+     &                  ,mflx,s1,s2,s3)
+c
+c    RAF, 2/17/97
+c
+c  PURPOSE: 
+c    Controls the update of density, energy, and momenta
+c    from the advection terms in the 2-direction.
+c
+c  INPUT:
+c    dlo         Mass            density prior to update
+c    eod         Specific energy density prior to update
+c
+c  OUTPUT:
+c    den         Mass            density    after update
+c    edn         Specific energy density    after update
+c
+c  I/O:
+c    s1,s2,s3    Momentum density components (get updated)
+c
+c  LOCAL:
+c    mflx        Mass flux in the 2-direction at zone faces 
+c
+c  EXTERNALS:
+c    BVALV1  , BVALV2  , BVALV3
+c    BVALD   , BVALE
+c
+c-----------------------------------------------------------------------
+      use real_prec
+      use param
+      use config
+      use root
+      use field
+      use bndry
+      use grid
+      use mpiyes
+      use mpipar
+c
+      implicit none
+c
+      integer  :: k1, k2, i, j, k
+      real(rl) :: p2
+c
+      real(rl) :: dlo(in,jn,kn), den(in,jn,kn), mflx(in,jn,kn),
+     .            s1 (in,jn,kn), s2 (in,jn,kn), s3  (in,jn,kn),
+     .            eod(in,jn,kn), edn(in,jn,kn),
+     .            ero(in,jn,kn), ern(in,jn,kn),
+     .            abo(in,jn,kn,nspec), abn(in,jn,kn,nspec)
+c
+      real(rl) :: atwid (ijkn)
+      real(rl) :: atwid1 (ijkn), atwid2 (ijkn), atwid3 (ijkn),
[…]

### 2014-10-20 sims/zeusGAP/src_dead/advx3.F

@@ -1,0 +1,305 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////                 A D V X 3                 \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c               University of Illinois at Urbana-Champaign
+c
+c=======================================================================
+c
+       subroutine advx3 (dlo,den
+     &                  ,eod,edn
+     &                  ,ero,ern
+     &                  ,abo,abn
+     &                  ,mflx,s1,s2,s3)
+c
+c    RAF, 2/19/97
+c
+c  PURPOSE: 
+c    Controls the update of density, energy, and momenta
+c    from the advection terms in the 3-direction.
+c
+c  INPUT:
+c    dlo         Mass            density prior to update
+c    eod         Specific energy density prior to update
+c    ero         Specific radiation energy density prior to update
+c
+c  OUTPUT:
+c    den         Mass            density    after update
+c    edn         Specific energy density    after update
+c    ern         Specific radiation energy density    after update
+c
+c  I/O:
+c    s1,s2,s3    Momentum density components (get updated)
+c
+c  LOCAL:
+c    mflx        Mass flux in the 3-direction at zone faces 
+c
+c  EXTERNALS:
+c    BVALV1  , BVALV2  , BVALV3
+c    BVALD   , BVALE   , BVALER
+c
+c-----------------------------------------------------------------------
+      use real_prec
+      use param
+      use config
+      use root
+      use field
+      use bndry
+      use grid
+      use mpiyes
+      use mpipar
+c
+      implicit none
+c
+      integer  :: k1,k2,i
+      integer  :: kbeg, kend, krange, kblocks, kskip, ktlb, kblk, kpage
+      integer  :: j, k
+      real(rl) :: p3
+c
+      real(rl) :: dlo(in,jn,kn), den(in,jn,kn), mflx(in,jn,kn),
+     .            s1 (in,jn,kn), s2 (in,jn,kn), s3  (in,jn,kn),
[…]

### 2014-10-20 sims/zeusGAP/src_dead/bndyflgs.F

@@ -1,0 +1,325 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////              B N D Y F L G S              \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c               University of Illinois at Urbana-Champaign
+c
+c=======================================================================
+c
+       subroutine bndyflgs
+c
+c    dac:zeus3d.bndyflgs <---------------- sets secondary boundary flags
+c                                                          october, 1990
+c
+c    written by: David Clarke
+c    modified 1: by RAF 2/13/96 for ZEUS-MP
+c
+c  PURPOSE:  This subroutine sets the secondary integer boundary flags
+c  ("niib2", "niib3", "niib23", etc.) which are used in addition to the
+c  primary flags ("niib", etc.) to account for the staggered grid.
+c  Thus, the following chart indicates the integer flag which sets the
+c  various variables at the various boundaries:
+c
+c      boundary            variable(s)         integer flag
+c
+c      inner (outer) i     d, e, v1, b1        niib   (noib  )
+c                          v2, b2              niib2  (noib2 )
+c                          v3, b3              niib3  (noib3 )
+c                          emf1, emf2, emf3    niib23 (noib23)
+c      inner (outer) j     d, e, v2, b2        nijb   (nojb  )
+c                          v3, b3              nijb3  (nojb3 )
+c                          v1, b1              nijb1  (nojb1 )
+c                          emf1, emf2, emf3    nijb31 (nojb31)
+c      inner (outer) k     d, e, v3, b3        nikb   (nokb  )
[…]

### 2014-10-20 sims/zeusGAP/src_dead/cgsolve.F

@@ -1,0 +1,520 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////               C G S O L V E               \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c                 University of California at San Diego
+c
+c=======================================================================
+      subroutine cgsolve(dd, ddp1,
+     .                       ddp2,
+     .                       ddp3,
+     .                   x,rhs,toler,error)
+c
+      use real_prec
+      use config
+      use param
+      use root
+      use grid
+      use impsoln
+      use mpiyes
+      use mpipar
+c
+      implicit none
+c
+      integer  :: isx, iex, isy, iey, isz, iez,
+     .            ierror
+c
+      real(rl) ::   dd(neqm,neqm,in,jn,kn)
+      real(rl) :: ddp1(neqm,neqm,in,jn,kn)
+      real(rl) :: ddp2(neqm,neqm,in,jn,kn)
+      real(rl) :: ddp3(neqm,neqm,in,jn,kn)
+c
+      real(rl) :: x   (neqm,in,jn,kn),
+     .            rhs (neqm,in,jn,kn)
+      real(rl) :: r   (neqm,in,jn,kn)
+      real(rl) :: p   (neqm,in,jn,kn)
+      real(rl) :: z   (neqm,in,jn,kn)
+c
+      real(rl) :: toler, error, glberror
+c
+      integer  :: i, jx, jy, jz
+c
+      integer  :: isxm1, isxp1, iexm1, iexp1, isym1, isyp1, ieym1, 
+     .            ieyp1, iszm1, iszp1, iezm1, iezp1
+c
+      real(rl) :: rnorm, beta_k, beta_ko, b_k
+      real(rl) :: denom, alpha_k
+c
+      real(rl) :: ls_dprd
+      real(rl) :: ls_dprd2
+c
+      error = 0.D0
+c
+c     set loop bounds
+c
+      isx = is
+      iex = ie
+      isy = js
+      iey = je
+      isz = ks
+      iez = ke
+      if(ldimen .eq. 1) then
+       isym1 = isy
+       isyp1 = isy
+       ieym1 = isy
+       ieyp1 = isy
+       iszm1 = isz
+       iszp1 = isz
+       iezm1 = isz
[…]

### 2014-10-20 sims/zeusGAP/src_dead/configure.F

@@ -1,0 +1,174 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////             C O N F I G U R E           \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c                 University of California at San Diego
+c
+c     PURPOSE:  Reads the code configuration namelists that debuted
+c               with ZEUS-MP Version 2.  These namelists serve the
+c               functions formerly performed by CPP macros in the
+c               discarded "zeusmp.def" file used to configure
+c               ZEUS-MP Version 1.
+c
+c     Written by:  John Hayes; way back in '03.
+c     Modified by: John Hayes; repeatedly since then.
+c     Modified 09/01/2006: moved file OPEN for zmp_inp after the MPI
+c     initialization and added "if(myid_w .eq. 0)" condition to the
+c     OPEN statement.
+c
+c=======================================================================
+c
+      subroutine configure
+c
+      use real_prec
+      use param
+      use config
+      use mpiyes
+      use mpipar
+c
+      implicit NONE
+c
+      integer ::  confi_buf(10)
+      logical ::  confl_buf(20)
+      real(rl)::  confr_buf(2)
+c
+      namelist /geomconf/  lgeom, ldimen
+      namelist /physconf/  lrad   , xhydro , xgrav, xmhd    , xgrvfft,
+     .                     xptmass, xtotnrg, xiso , xvgrid  , xsubav,
+     .                     xforce , xsphgrv, leos , nspec   , lopac
+      namelist /ioconf/    xascii , xhdf, xrestart, xtsl
+      namelist /preconf/   small_no, large_no
+      namelist /arrayconf/ izones, jzones, kzones, maxijk
+c
+c----------------------------------------------------------------------
+c     If parallel execution, start up MPI
+c----------------------------------------------------------------------
+c
+       call MPI_INIT( ierr )
+       call MPI_COMM_RANK( MPI_COMM_WORLD, myid_w  , ierr )
+       call MPI_COMM_SIZE( MPI_COMM_WORLD, nprocs_w, ierr )
+c
[…]

### 2014-10-20 sims/zeusGAP/src_dead/ct_1D.F

@@ -1,0 +1,284 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////                  C T _ 1 D                \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c               University of California at San Diego
+c
+c=======================================================================
+c
+       subroutine ct_1D
+c
+c    dac:zeus3d.ct <-------- updates B-field using constrained transport
+c    from jms:zeus2d.ct                                    october, 1989
+c
+c    written by: David Clarke
+c    modified 1: May, 1990 by David Clarke; reworked to call the new
+c                interpolation routines which need only be called once.
+c    modified 2: August, 1990 by David Clarke; moved magnetic fields to
+c                the face-centres (cospatial with the velocities).
+c                Implemented a method of characteristics for evaluating
+c                the emf's.  The transverse Lorentz accelerations are
+c                now applied to the velocities during this step.
+c    modified 3: November, 1990 by David Clarke; resurrected non-MoC
+c                algorithm for evaluating emfs.  Added EDITOR alias MOC
+c                which needs to be defined if MoC is to be used.
+c    modified 4: June, 1992 by David Clarke; reworked singularity
+c                formalism.
+c    modified 5: December, 1992 by David Clarke (as suggested by John
+c                Hawley); split velocity update from emf computation.
+c                Velocities are now Lorentz-accelerated with *old*
+c                magnetic field values. emf's are then estimated with
+c                Lorentz-updated velocities.
+c    modified 6: 3 March 1998, by Mordecai-Mark Mac Low; translated 
+c                into ZEUS-MP form.
+c
+c    modified 7: created "ct_1D" clone which assumes symmetry about the
+c                J and K axes, after ZEUS3D. (J. Hayes; October 2005)
+c
+c    modified 8: John Hayes, May 15, 2006; changed the call to LORENTZ_D
[…]

### 2014-10-20 sims/zeusGAP/src_dead/difco2.F

@@ -1,0 +1,111 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////               D I F C O 2                 \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c                 University of California at San Diego
+c
+c     PURPOSE: compute diffusion coefficients for 2-component of
+c              radiation flux
+c
+c     Written by: Robert Fiedler and John Hayes
+c
+c=======================================================================
+c
+c=======================================================================
+c
+      subroutine difco2(erold,dfc,ibeg,iend,jbeg,jend,kbeg,kend)
+c
+      use real_prec
+      use config
+      use param
+      use grid
+      use field
+      use root
+      use scratch
+      use radiation
+      use opac
+      use mpiyes
+      use mpipar
+      use cons
+c
+      implicit NONE
+c
+      integer  :: i, j, k, ibeg, iend, jbeg, jend, kbeg, kend
+c
+      real(rl) :: der1, der2, der3, lmda, qa, qb, grad_er, ros_mfp
+      real(rl) :: r(in,jn,kn), dfc(in,jn,kn), erold(in,jn,kn)
+      real(rl) :: onethird, rdmcm, rdmcp, flux1, flux2
+c
+c
+c=======================================================================
+c  Compute radiation energy gradient vector and R
+c=======================================================================
+c
+      if(ldimen .eq. 1) then
+       do k = kbeg, kend
+        do j = jbeg, jend
+         do i = ibeg, iend
+          dfc(i,j,k) = 0.0D0
+         enddo
+        enddo
+       enddo
+       return
+      endif ! ldimen = 1
+      do k = kbeg, kend
+       do j = jbeg, jend
+        do i = ibeg, iend
+         der1     = (erold(i  ,j-1,k) - erold(i-1,j-1,k  )) * dx1bi(i)
+     .            + (erold(i  ,j  ,k) - erold(i-1,j  ,k  )) * dx1bi(i)
+         der1     = der1 * 0.5 
+         der2     = (erold(i  ,j  ,k) - erold(i  ,j-1,k  )) * dx2bi(j)
+     .            * g2bi(i)
[…]

### 2014-10-20 sims/zeusGAP/src_dead/eos.F

@@ -1,0 +1,84 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////                    E O S                  \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c                 University of California at San Diego
+c
+c=======================================================================
+      subroutine eos(ibeg,iend,jbeg,jend,kbeg,kend)
+c
+      use real_prec
+      use config
+      use param
+      use field
+      use bndry
+      use grid
+      use root
+      use scratch
+      use cons
+      use eos_par
+      use mpiyes
+      use mpipar
+      use planet
+c
+      implicit NONE
+c
+      integer  :: i, j, k, ibeg, iend, jbeg, jend, 
+     .            kbeg, kend
+      real*8   :: cylind_rad, cs, rad, theta, custcs,
+     .            distp, x, y, z, omegap_sq, omegas_sq,
+     .            dcs
+      real*8   :: csq, bigH
+c
+      do k = kbeg, kend
+       do j = jbeg, jend
+        do i = ibeg, iend
+         if(leos .eq. 1) then
+          if(.not. xiso) then
+           if(.not. xtotnrg) then
+            p(i,j,k) = gamm1*e(i,j,k)
+           else
+            p(i,j,k) = gamm1*(
+     .                 e(i,j,k) - 0.125D0*d(i,j,k)*(
+     .                  (v1(i  ,j  ,k  )+v1(i+1,j  ,k  ))**2
+     .                 +(v2(i  ,j  ,k  )+v2(i  ,j+1,k  ))**2
+     .                 +(v3(i  ,j  ,k  )+v3(i  ,j  ,k+1))**2 )
+     .                        )
+           endif
+          else
+c
+c     Locally isothermal equation of state, now with arbitrary power-law
+c     With option to do Peplinski (2008)'s EOS, which heats up the gas
+c     near the planet. Enabled if parameter hp > 0. 
+             rad = x1b(i)
+             theta = x2b(j)
+             cylind_rad = rad*sin(theta)
+c             cs = smallh/sqrt(cylind_rad)
+             cs = sqrt(csq(cylind_rad))
+             if(hplanet.gt.0d0) then
+                x = rad*sin(theta)*cos(x3b(k))
+                y = rad*sin(theta)*sin(x3b(k))
[…]

### 2014-10-20 sims/zeusGAP/src_dead/gpbv.F

@@ -1,0 +1,871 @@
+c=======================================================================
+c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c               University of Illinois at Urbana-Champaign
+c
+      subroutine gpbv
+c     MKL comment: 
+c     This is the original boundary potential solver that comes with ZEUSMP2. 
+c     It assumes mass distribution is far from boundaries, which is inappropriate for disk
+c     problems. So we cannot use this routine. So we calculate boundary potential with spherical
+c     harmonics. 
+c
+c  Written by PSLi (12/4/99)
+c
+c  PURPOSE: Calculate the gravitational potential at the boundary
+c           surfaces contributed from monopole and quadrupole moments
+c           of a mass distribution.
+c           Boundary potentals are calculated if the flags niis(3),
+c           nois(3), nijs(3), nojs(3), niks(3), or noks(3) = 3,
+c           respectively.
+c
+c  EXTERNALS: [none]
+c
+c  LOCALS:
+c  dm(i,j,k)   is mass contained in zone i,j,k
+c  lqm         is the local quadrupole moment (qm)
+c  lcx,lcy,lcz coords of mass center at local processor (cx,cy,cz)
+c  ltm         total mass at local processor (tm)
+c-----------------------------------------------------------------------
+      use real_prec
+      use config
+      use param
+      use grid
+      use field
+      use bndry
+      use root
+      use scratch
+      use mpiyes
+      use mpipar
+c
+      implicit NONE
+c
+      real(rl) ::     errmax
+      integer  :: n
+c
+      real(rl) ::  dm(in,jn,kn),qm(6,nprocs_w),tm(nprocs_w),
+     .             cx(nprocs_w),
+     &             cy(nprocs_w),cz(nprocs_w),r1,dx,dy,dz,dx2,dy2,dz2,
[…]

### 2014-10-20 sims/zeusGAP/src_dead/grav3dcg.F

@@ -1,0 +1,718 @@
+c
+c                            Developed by
+c                Laboratory for Computational Astrophysics
+c                  University of California at San Diego
+c
+      subroutine grav3D_CG
+c
+c     3-D gravitational potential solver using John Hayes's
+c     CGSOLVE package for the covariant form of Poisson's equation
+c
+      use real_prec
+      use config
+      use param
+      use cons
+      use domain
+      use root
+      use field
+      use grid
+      use bndry
+      use impsoln
+      use gravmod
+      use mpiyes
+      use mpipar
+c
+      implicit NONE
+c
+      real(rl), dimension(:,:,:,:,:), allocatable, save ::
+     .          ad, a0, a1, a2
+      real(rl), dimension(:,:,:,:), allocatable, save ::
+     .          x, b
+c
+      real(rl), dimension(:,:), allocatable ::
+     .          gpiibc, gpoibc, gpijbc, gpojbc, gpikbc, gpokbc
+c
+      real(rl) :: toler, error, glbtoler, eps, glbeps, epstol, epsmax,
+     .            epsmin, glbemin, phip(in,jn,kn)
+c
+      integer :: i, j, k, prev_tot, imax, jmax, kmax
+c
+c-----------------------------------------------------------------------
+c     First timestep only: initialize everything required for CGSOLVE
+c-----------------------------------------------------------------------
+c
+      if(igcall .eq. 0) then
+c
+c-----------------------------------------------------------------------
+c      allocate matrix, solution, and RHS array storage
+c-----------------------------------------------------------------------
+c
+       allocate(ad(neqm,neqm,IN,JN,KN))
+       allocate(a0(neqm,neqm,IN,JN,KN))
+       allocate(a1(neqm,neqm,IN,JN,KN))
+       allocate(a2(neqm,neqm,IN,JN,KN))
+       allocate(x (     neqm,IN,JN,KN))
+       allocate(b (     neqm,IN,JN,KN))
+      endif ! igcall = 0
+c
+c-----------------------------------------------------------------------
+c     compute BC arrays
+c
+c     NOTE: see comments in grav2dcg.F regarding use of subroutine
[…]

### 2014-10-20 sims/zeusGAP/src_dead/grey_fld.F

@@ -1,0 +1,808 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////              G R E Y _ F L D              \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c                 University of California at San Diego
+c
+c=======================================================================
+c
+      subroutine grey_fld(dcopy, eold, eod , erold, erod)
+c
+c     call       grey_fld(w3dd , e   , w3de, w3dh , er)
+c
+c
+c     driver routine for computing an implicit solution to the
+c     radiation diffusion equation.  The structure of
+c     the N-R iteration and convergence control is taken from ZEUS-2D.
+c
+c     written by: John C. Hayes; June, 1997
+c     re-written for F90 by J. Hayes; May, 2003
+c
+      use real_prec
+      use config
+      use param
+      use field
+      use bndry
+      use grid
+      use root
+      use radiation
+      use cons
+      use opac
+      use opac_law
+      use mpiyes
+      use mpipar
+      use clockmod
+      use impsoln
+c
+      implicit NONE
+c
+c     matrix arrays
+c
+      real(rl) :: dd  (neqm,neqm,in,jn,kn)
+      real(rl) :: ddp1(neqm,neqm,in,jn,kn)
+      real(rl) :: ddp2(neqm,neqm,in,jn,kn)
+      real(rl) :: ddp3(neqm,neqm,in,jn,kn)
+      real(rl) :: rhs (neqm,     in,jn,kn)
+      real(rl) :: x   (neqm,     in,jn,kn)
+c
+c     work arrays
+c
+      real(rl) ::   fn1     (in,jn,kn)
+      real(rl) ::   fn2     (in,jn,kn)
+      real(rl) :: d_fn1_d_er(in,jn,kn)
+      real(rl) :: d_fn1_d_eg(in,jn,kn)
+      real(rl) :: d_fn2_d_er(in,jn,kn)
+      real(rl) :: d_fn2_d_eg(in,jn,kn)
+      real(rl) :: grdvcf    (in,jn,kn)
+c
+c     matrix solver control
+c
+      real(rl) ::    toler, error
+      real(rl) ::    lssave
+c
+c     N-R iteration control
+c
+      real(rl) :: dthydro, trad, epsemx, epsermx, q1, demx, dermx,
[…]

### 2014-10-20 sims/zeusGAP/src_dead/jacob.F

@@ -1,0 +1,378 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////                  J A C O B                \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c                 University of California at San Diego
+c
+c=======================================================================
+c
+      subroutine jacob(eold, erold, dd, rhs, fn1, fn2, 
+     .                 d_fn1_d_er, d_fn1_d_eg,
+     .                 d_fn2_d_er,
+     .                 d_fn2_d_eg, grdvcf, deldotv,deldotf)
+c
+c     completes the jacobian and the RHS for the Newton-Raphson
+c     solution to the implicit radiation energy equation
+c
+c     written by: John C. Hayes; March, 1997
+c
+c     modified for F90 by J. Hayes; May, 2003
+c
+      use real_prec
+      use config
+      use param
+      use field
+      use bndry
+      use grid
+      use root
+      use radiation
+      use opac
+      use cons
+      use mpiyes
+      use mpipar
+c
+      implicit NONE
+c
+c     loop indices and loop bounds
+c
+      integer  :: i, j, k
+      integer  :: jsm2, jep1, jep2, ksm2, kep1, kep2
+c
+c     matrix arrays
+c
+      real(rl) :: dd(neqm,neqm,in,jn,kn)
+c
+c     the right-hand side
+c
+      real(rl) :: rhs(neqm,in,jn,kn)
+c
+c     scalar work variables
+c
+      real(rl) :: d_f1p_d_e, d_f1_d_e,
+     .            d_f2p_d_e, d_f2_d_e,
+     .            d_f3p_d_e, d_f3_d_e,
+     .            cancel_term
+c
+c     work arrays
+c
+      real(rl) :: f1           (in,jn,kn)
+      real(rl) :: f2           (in,jn,kn)
+      real(rl) :: f3           (in,jn,kn)
+      real(rl) :: deldotf      (in,jn,kn)
+      real(rl) :: deldotv      (in,jn,kn)
+      real(rl) :: d_deldotf_d_e(in,jn,kn)
[…]

### 2014-10-20 sims/zeusGAP/src_dead/lorentz.F

@@ -1,0 +1,785 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////               L O R E N T Z               \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c               University of Illinois at Urbana-Champaign
+c
+c=======================================================================
+c
+       subroutine lorentz (ibeg,iend,jbeg,jend,kbeg,kend
+     &                   ,u1,u2,u3,w1,w2,w3)
+c
+c    dac:zeus3d.lorentz <----- MoC estimate of transverse lorentz forces
+c                                                          october, 1992
+c
+c    written by: David Clarke
+c    modified 1: Mordecai-Mark Mac Low, December 1997 - March 1998
+c                rewritten for ZEUS-MP 
+c    modified 2: John Hayes, years ago; Rewritten for f90
+c    modified 3: John Hayes, 09/01/2006; corrected metric factor
+c                in second loop for "st3"
+c
+c  PURPOSE:  Uses the Method of Characteristics (MoC) to compute the
+c  transverse components of the Lorentz force and then accelerate the
+c  velocities accordingly.  After a suggestion by John Hawley, these
+c  source terms are computed separately from the MoC estimate of the
+c  emfs (MOCEMFS) and before the magnetic field update.  This improves
+c  stability of multi-dimensional Alfven waves.
+c
+c  MOCEMFS solves the induction equation without the use of operator
+c  splitting.  Thus, the relevant characteristic velocities are v+/-va
+c  and includes the flow velocity v (from the advection term v grad(B))
+c  and the alfven velocity va (from the induction term va grad(B)).
+c  Conversely, Euler's equation is operator split, and the advection
+c  term (v grad(v)) is handled elsewhere (MOMX*).  Thus, the effective
+c  characteristic velocities when estimating the transverse Lorentz
+c  terms is simply +/-va (from the induction term va grad(v)).
[…]

### 2014-10-20 sims/zeusGAP/src_dead/lorentz_1D.F

@@ -1,0 +1,366 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////               L O R E N T Z               \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c               University of Illinois at Urbana-Champaign
+c
+c=======================================================================
+c
+       subroutine lorentz_1D(ibeg,iend,jbeg,jend,kbeg,kend
+     &                   ,u1,u2,u3,w1,w2,w3)
+c
+c    dac:zeus3d.lorentz <----- MoC estimate of transverse lorentz forces
+c                                                          october, 1992
+c
+c    written by: David Clarke
+c    modified 1: Mordecai-Mark Mac Low, December 1997 - March 1998
+c                rewritten for ZEUS-MP 
+c    modified 2: John Hayes, October 2005; created lorentz_1D clone
+c                which assumes symmetry about the J and K axes; after
+c                ZEUS3D.
+c
+c  PURPOSE:  Uses the Method of Characteristics (MoC) to compute the
+c  transverse components of the Lorentz force and then accelerate the
+c  velocities accordingly.  After a suggestion by John Hawley, these
+c  source terms are computed separately from the MoC estimate of the
+c  emfs (MOCEMFS) and before the magnetic field update.  This improves
+c  stability of multi-dimensional Alfven waves.
+c
+c  MOCEMFS solves the induction equation without the use of operator
+c  splitting.  Thus, the relevant characteristic velocities are v+/-va
+c  and includes the flow velocity v (from the advection term v grad(B))
+c  and the alfven velocity va (from the induction term va grad(B)).
+c  Conversely, Euler's equation is operator split, and the advection
+c  term (v grad(v)) is handled elsewhere (MOMX*).  Thus, the effective
+c  characteristic velocities when estimating the transverse Lorentz
+c  terms is simply +/-va (from the induction term va grad(v)).
+c
[…]

### 2014-10-20 sims/zeusGAP/src_dead/mapin.F

@@ -1,0 +1,943 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////                 M A P I N                 \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c                 University of California at San Diego
+c
+c     PURPOSE: maps restart file arrays into ZEUS-MP arrays
+c
+c     Written by: John Hayes
+c
+c=======================================================================
+      subroutine mapin
+c
+      use real_prec
+      use config
+      use param
+      use grid
+      use mpiyes
+      use mpipar
+      use field
+      use bndry
+      use root
+      use gravmod
+      use restart_arrays
+c
+      implicit NONE
+c
+      character*15 :: filename
+c
+      integer  :: i, j, k, index, l
+c
+c=======================================================================
+c     grid.h arrays
+c=======================================================================
+c
+c-----------------------------------------------------------------------
+c     integers
+c-----------------------------------------------------------------------
+c
+      is    = ntgrdvr(1)
+      ie    = ntgrdvr(2)
+      js    = ntgrdvr(3)
+      je    = ntgrdvr(4)
+      ks    = ntgrdvr(5)
+      ke    = ntgrdvr(6)
+      iga   = ntgrdvr(7)
+      jga   = ntgrdvr(8)
+      kga   = ntgrdvr(9)
+      igcon = ntgrdvr(10)
+      nx1z  = ntgrdvr(11)
+      nx2z  = ntgrdvr(12)
+      nx3z  = ntgrdvr(13)
+c
+c-----------------------------------------------------------------------
+c     Fixed coordinate arrays
+c-----------------------------------------------------------------------
+c
+      index = 0
+c
+      call get_grid_array(x1a,in,index)
+      call get_grid_array(x2a,jn,index)
+      call get_grid_array(x3a,kn,index)
+c
+      call get_grid_array(x1ai,in,index)
+      call get_grid_array(x2ai,jn,index)
+      call get_grid_array(x3ai,kn,index)
+c
+      call get_grid_array(dx1a,in,index)
+      call get_grid_array(dx2a,jn,index)
[…]

### 2014-10-20 sims/zeusGAP/src_dead/matprop.F

@@ -1,0 +1,108 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////               M A T P R O P               \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c                 University of California at San Diego
+c
+c=======================================================================
+c
+      subroutine matprop(e, er, d, gam, t, dtde, p, bb, dbbdt,
+     .                   kr, kp, sg, dkpdt, km, dkedt, dpde,
+     .                   ibeg, iend, jbeg, jend, 
+     .                   kbeg, kend)
+c
+c     This routine is a conglomeration of the old TEMP, EOS, PLANCK,
+c     ABSORP, and SCATT routines.
+c
+      use real_prec
+      use config
+      use param
+      use root
+      use cons
+      use grid
+      use opac_law
+      use mpiyes
+      use mpipar
+c
+      implicit NONE
+c
+      integer  :: i, ibeg, iend, j, jbeg, jend, k, kbeg, kend
+c
+      real(rl) :: e     (in,jn,kn), d   (in,jn,kn), t    (in,jn,kn), 
+     .            p     (in,jn,kn), bb  (in,jn,kn), kp   (in,jn,kn),
+     .            dtde  (in,jn,kn), dpde(in,jn,kn), dbbdt(in,jn,kn),
+     .            dkpdt (in,jn,kn), sg  (in,jn,kn), er   (in,jn,kn),
+     .            dkedt (in,jn,kn), km  (in,jn,kn), kr   (in,jn,kn)
+c
+      real(rl) :: gam, coef, temp, binv, gam1, power, stef
+      real(rl) :: t_max, rdmcm1, rdmcp1, rdmcm2, rdmcp2, rdmcm3, rdmcp3,
+     .            ros_mfp
+      real(rl) :: rmfp0i, hundred
+c
+      real(rl) :: so_eps , tau    , eriibn, timarg
+c
+      common /soui/ so_eps, tau, eriibn
+c
+      hundred = 100.0d0
+      gam1 = gam - 1.0
+      stef = clight * rad_con / (4.D0 * pi)
+      if(leos .eq. 1) then
+       do 10 k = kbeg, kend
+       do 10 j = jbeg, jend
[…]

### 2014-10-20 sims/zeusGAP/src_dead/minpack.F

@@ -1,0 +1,1752 @@
+      subroutine dogleg(n,r,lr,diag,qtb,delta,x,wa1,wa2)
+      integer n,lr
+      double precision delta
+      double precision r(lr),diag(n),qtb(n),x(n),wa1(n),wa2(n)
+c     **********
+c
+c     subroutine dogleg
+c
+c     given an m by n matrix a, an n by n nonsingular diagonal
+c     matrix d, an m-vector b, and a positive number delta, the
+c     problem is to determine the convex combination x of the
+c     gauss-newton and scaled gradient directions that minimizes
+c     (a*x - b) in the least squares sense, subject to the
+c     restriction that the euclidean norm of d*x be at most delta.
+c
+c     this subroutine completes the solution of the problem
+c     if it is provided with the necessary information from the
+c     qr factorization of a. that is, if a = q*r, where q has
+c     orthogonal columns and r is an upper triangular matrix,
+c     then dogleg expects the full upper triangle of r and
+c     the first n components of (q transpose)*b.
+c
+c     the subroutine statement is
+c
+c       subroutine dogleg(n,r,lr,diag,qtb,delta,x,wa1,wa2)
+c
+c     where
+c
+c       n is a positive integer input variable set to the order of r.
+c
+c       r is an input array of length lr which must contain the upper
+c         triangular matrix r stored by rows.
+c
+c       lr is a positive integer input variable not less than
+c         (n*(n+1))/2.
+c
+c       diag is an input array of length n which must contain the
+c         diagonal elements of the matrix d.
+c
+c       qtb is an input array of length n which must contain the first
+c         n elements of the vector (q transpose)*b.
+c
+c       delta is a positive input variable which specifies an upper
+c         bound on the euclidean norm of d*x.
+c
+c       x is an output array of length n which contains the desired
+c         convex combination of the gauss-newton direction and the
+c         scaled gradient direction.
+c
+c       wa1 and wa2 are work arrays of length n.
+c
+c     subprograms called
+c
+c       minpack-supplied ... dpmpar,enorm
+c
[…]

### 2014-10-20 sims/zeusGAP/src_dead/mnmx.F

@@ -1,0 +1,132 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////                  M N M X                  \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c               University of Illinois at Urbana-Champaign
+c
+c=======================================================================
+c
+       subroutine mnmx ( qty, i1, j1, k1, i2, j2, k2
+     1                 , qmin, imin, jmin, kmin
+     2                 , qmax, imax, jmax, kmax )
+c
+c    dac:zeus3d.mnmx <------------- finds extrema of a regular 3-d array
+c    from mln:zeus04.minmax; jms:zeus2d.minmax            february, 1990
+c
+c    written by: David Clarke
+c    modified 1: converted to ZEUS-MP by Mordecai-Mark Mac Low M
+c
+c  PURPOSE: This subroutine returns the maximum and minimum value of a
+c  rectangular 3-D array, along with the coordinates of the extrema.
+c
+c  INPUT VARIABLES:
+c    qty             the 3-D array to be searched for extrema.
+c    i1              inner i index.
+c    j1              inner j index.
+c    k1              inner k index.
+c    i2              outer i index.
+c    j2              outer j index.
+c    k2              outer k index.
+c
+c  OUTPUT VARIABLES:
+c    qmin            minimum value
+c    imin,jmin,kmin  coordinates of minimum value
+c    qmax            maximum value
+c    imax,jmax,kmax  coordinates of maximum value
+c
+c  LOCAL VARIABLES:
+c
+c  j-sweep
+c    qmaxj (qminj)   vector of maximum (minimum) values of "qty" from
+c                    each i-sweep.  This vector is filled during a
+c                    j-sweep.
+c    imaxj           i-index of each "qmaxj"
+c    iminj           i-index of each "qminj"
+c
+c  k-sweep
+c    qmaxk (qmink)   vector of maximum (minimum) values of "qmaxj"
+c                    ("qminj") from each j-sweep.
+c    imaxk           i-index of each "qmaxk"
+c    jmaxk           j-index of each "qmaxk"
[…]

### 2014-10-20 sims/zeusGAP/src_dead/mod_files.F

@@ -1,0 +1,577 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////            M O D _ F I L E S           \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c                  University of California at San Diego
+c
+c     PURPOSE: container for all ZEUS-MP F90 module files
+c
+c     Written by: John Hayes, 2003.
+c
+c=======================================================================
+      module actual_types
+c
+      logical*4   ::  logical4
+      integer*4   ::  integer4
+      integer*8   ::  integer8
+      real*4      ::  real4
+      real*8      ::  real8
+      complex*8   ::  complex4
+      complex*16  ::  complex8
+c
+      end module actual_types
+c
+      module real_prec
+c
+      use actual_types
+c
+      integer, parameter :: rl = kind(real8)
+      integer, parameter :: rl4= kind(real4)
+c
+      end module real_prec
+c
+      module param
+c
+      use real_prec
+c
+      integer :: in
+      integer :: jn
+      integer :: kn
+      integer :: ijkn
+      integer, parameter :: neqm = 1
+c
+      integer, parameter :: nbvar = 14
+c
+      real(rl), parameter :: pi = 3.14159265358979324
+      real(rl), parameter :: tiny = 1.0d-99
+      real(rl), parameter :: huge = 1.0d+99
+c
+      real(rl), parameter :: zro  = 0.0d0
+      real(rl), parameter :: one  = 1.0d0
+      real(rl), parameter :: two  = 2.0d0
+      real(rl), parameter :: haf  = 0.5d0
+c
+      integer, parameter :: nbuff = 128
+      integer, parameter :: mreq=300
+c
+      end module param
+c
+      module config
+c
+      use real_prec
+      use param
+c
+c======================================================================
+c     ZEUS-MP CONFIGURATION PARAMETERS -- Replacing the CPP macros
+c     contained in zeusmp.def
+c======================================================================
+c
+c----------------------------------------------------------------------
[…]

### 2014-10-20 sims/zeusGAP/src_dead/msendrec.F

@@ -1,0 +1,179 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////          M S E N D R E C _ B N D          \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c                 University of California at San Diego
+c
+c     PURPOSE: update matrix elements in ghost zones for CG solver
+c
+c     Written by: F. Douglas Swesty and John Hayes
+c
+c=======================================================================
+      subroutine msendrec_bnd(
+     .                  isx,iex,isy,iey,isz,iez,
+     .                  x)
+c
+      use real_prec
+      use config
+      use param
+      use mpiyes
+      use mpipar
+c
+      implicit none
+c
+      integer  :: isx, iex, isy, iey, isz, iez, i, j, k, n, nelm
+c
+      real(rl) :: x(neqm,neqm,in,jn,kn)
+c
+c-----------------------------------------------------------------------
+c
+      nelm = 1
+c
+c                               ! send south; receive north
+      if(ntiles(1) .gt. 1) then
+       if((periodic(1) .eqv. .true.) .or. (coords(1) .ne. 0)) then
+        nreq = nreq + 1
+        call mpi_isend(
+     .        x(1,1,isx,1,1),nelm,ilsm_slice,n1m,16300+nsub,
+     .        comm3d,req(nreq),ierr)
+       endif
+       if((periodic(1) .eqv. .true.) .or. (coords(1) .ne. ntiles(1)-1))
+     .  then
+        nreq = nreq + 1
+        call mpi_irecv(
+     &        x(1,1,iex+1,1,1),nelm,ilsm_slice,n1p,16300+nsub,
+     &        comm3d,req(nreq),ierr)
+       endif
+c
+c                               ! send north; receive south
+       if((periodic(1) .eqv. .true.) .or. (coords(1) .ne. ntiles(1)-1))
+     .  then
+        nreq = nreq + 1
+        call mpi_isend(
+     .        x(1,1,iex,1,1),nelm,ilsm_slice,n1p,16400+nsub,
+     &        comm3d,req(nreq),ierr)
+       endif
[…]

### 2014-10-20 sims/zeusGAP/src_dead/nudt.F

@@ -1,0 +1,477 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////                  N U D T                  \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c               University of Illinois at Urbana-Champaign
+c
+c=======================================================================
+c
+       subroutine nudt
+c
+c    mln:zeus3d.nudt <------------------------ mhd time step calculation
+c                                                              may, 1986
+c
+c    written by: Mike Norman
+c    modified 1: May, 1986 by David Clarke; adapted for mhd.
+c    modified 2: October, 1987 by Mike Norman; reworked for covariant
+c                formulation.
+c    modified 3: June, 1988 by Jim Stone; incorporated into ZEUS2D.
+c    modified 4: February, 1990 by David Clarke; incorporated into
+c                ZEUS3D.
+c    modified 5: September, 1990 by David Clarke; moved magnetic fields
+c                to face-centres.
+c    modified 6: March 5, 1996 by Robert Fiedler; completely rewritten
+c                for ZEUS-MP.
+c    modified 7: Aug. 23, 1996 by Robert Fiedler; minor change to 
+c                substep counter.
+c    modified 8: Dec. 19, 1996 by Robert Fiedler; added radiation 
+c                diffusion time step.
+c    modified 9: Feb. 14, 1997 by RAF; corrected for non-hydro and
+c                lower dimensionality options.
+c    modified 10: kluge for Garching T3E (search on M-MML) 26 Feb 98
+c    modified 11: December, 1999 by PSLi; add in dt calculation in case
+c                 of subcycle of artificial viscosity.
+c
+c  PURPOSE:  Computes the new timestep for explicit calculations from
+c  the values of the field variables updated by the source and transport
+c  steps.
+c
+c  In explicit calculations, the timestep is given by:
+c
+c     dt = courno * sqrt [ min ( dtcs**2 + dtv1**2 + dtv2**2 + dtv3**2
[…]

### 2014-10-20 sims/zeusGAP/src_dead/opac_d.F

@@ -1,0 +1,113 @@
+c======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////               O P A C _ D                 \\\\\\\\\c======================================================================
+c
+c
+      subroutine opac_d
+c
+c     Driver routine for computing opacities when radiation is included
+c
+c     Written by: John Hayes
+c     modified: 4-13-99; added entry for "kem" to OPACITY calls
+c     modified: 5-26-2003; rewritten for F90 code
+c
+c
+      use real_prec
+      use config
+      use param
+      use root
+      use field
+      use grid
+      use bndry
+      use scratch
+      use mpiyes
+      use mpipar
+      use radiation
+      use opac
+      use cons
+c
+      implicit NONE
+c
+      integer  :: i,j,k
+c
+      real(rl) :: temp
+c
+c----------------------------------------------------------------------
+c
+c     update ghost zones along 1-coordinate
+c
+       nreq = 0
+       nsub = nsub + 1
+       call bvald  (3,3,0,0,0,0,d)
+       call bvale  (3,3,0,0,0,0,e)
+c
+c     compute opacities in all interior zones
+c
+       call opacity(e, d, gamma, kapr, kap, sig, dkapdt, kem, dkemdt,
+     .              is, ie, js, je, ks, ke)
+c
+c     wait for communications to complete
+c
+       if (nreq .ne. 0) call MPI_WAITALL ( nreq, req, stat, ierr )
+      if(ldimen .eq. 1) then
+c
+c     compute opacities in 1-coordinate ghost zones and return
+c
+       call opacity(e, d, gamma, kapr, kap, sig, dkapdt, kem, dkemdt,
+     .              is-2, is-1, js, je, ks, ke)
+       call opacity(e, d, gamma, kapr, kap, sig, dkapdt, kem, dkemdt,
+     .              ie+1, ie+2, js, je, ks, ke)
+c
+       go to 999
+      endif ! ldimen = 1
+      if(ldimen .gt. 1) then
+c
+c     update ghost zones along 2-coordinate
+c
+       nreq = 0
+       nsub = nsub + 1
+       call bvald  (0,0,3,3,0,0,d)
+       call bvale  (0,0,3,3,0,0,e)
+c
[…]

### 2014-10-20 sims/zeusGAP/src_dead/opacity.F

@@ -1,0 +1,74 @@
+
+c=======================================================================
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////                  O P A C                  \\\\\\\\\c
+c=======================================================================
+c
+      subroutine opacity(e, d, gam, kr, kp, sg, dkpdt, km, dkedt,
+     .                   ibeg, iend, jbeg, jend, kbeg, kend)
+c
+      use real_prec
+      use config
+      use param
+      use cons
+      use grid
+      use opac_law
+      use mpiyes
+      use mpipar
+c
+      implicit NONE
+c
+      integer  :: i, ibeg, iend, j, jbeg, jend, k, kbeg, kend
+c
+      real(rl) :: e    (in,jn,kn), d   (in,jn,kn), t    (in,jn,kn), 
+     .            p    (in,jn,kn), kp   (in,jn,kn),
+     .            dtde (in,jn,kn), dpde(in,jn,kn), dbbdt(in,jn,kn),
+     .            dkpdt(in,jn,kn), sg  (in,jn,kn), er   (in,jn,kn),
+     .            kr   (in,jn,kn), km  (in,jn,kn), dkedt(in,jn,kn)
+c
+      real(rl) :: gam, coef, temp, binv, gam1, stef
+      real(rl) :: rmfp0i
+c
+      real(rl) :: kap_floor, kap_ceil, sp_xscn, hundred
+c
+      real(rl) :: so_eps , tau    , eriibn, timarg
+c
+      common /soui/ so_eps, tau, eriibn
+c
+      hundred = 100.0d0
+      stef = clight * rad_con / (4.D0 * pi)
+      gam1 = gam - 1.0D0
+c
+      binv = 1.D0 / boltz
+      do 20 k = kbeg, kend
+      do 20 j = jbeg, jend
+      do 20 i = ibeg, iend
+       t(i,j,k) = (gam1)*mmw*mh*binv*e(i,j,k)/d(i,j,k)
+20    continue
+c
+      rmfp0i  = 1.D0 / rmfp0
+      do 50 k = kbeg, kend
+      do 50 j = jbeg, jend
+      do 50 i = ibeg, iend
+         kr  (i,j,k) = rmfp0i * (d(i,j,k)/rho0)**xnu *
+     .                         (t(i,j,k)/ t_0)**(-powr)
+         sg   (i,j,k) = 0.D0
[…]

### 2014-10-20 sims/zeusGAP/src_dead/pdv_d.F

@@ -1,0 +1,322 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////                 P D V _ D                 \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c               University of Illinois at Urbana-Champaign
+c
+c=======================================================================
+c
+       subroutine pdv_d
+     1            (dcopy, eod)
+c
+c  PURPOSE
+c    Compute the compressional work term (PdV) in the material
+c    energy equation (for no radiation).
+c
+c  AUTHOR
+c    Robert A. Fiedler
+c
+c  LAST MODIFIED
+c    01/21/97
+c
+c  INPUT
+c    (none)
+c
+c  OUTPUT
+c    dcopy     A copy of the density.
+c    eod       e/d for the transport step.
+c
+c  We need just 1 "p" layer of updated boundary data for v1, v2, and v3,
+c  but none for d and e.  
+c
+c  Routine pdv also saves the density and e/d in (dcopy,eod) for use
+c  in the transport step.  This is why it is being called even when
+c  TOTAL_ENERGY is defined.
+c
+c  EXTERNALS:
+c    BVALV1  , BVALV2  , BVALV3
+c    PDV
+c.......................................................................
+c
+      use real_prec
+      use config
+      use param
+      use root
+      use field
+      use bndry
+      use grid
+      use mpiyes
+      use mpipar
+c
+      implicit NONE
+c
+      real(rl) :: dcopy(in,jn,kn), eod(in,jn,kn)
+c
+      integer  :: i, k1, k2, imax, jmax, kmax
+      integer  :: j, k
+c
+c-----------------------------------------------------------------------
+      if(ldimen .eq. 2) go to 222
+      if(ldimen .eq. 1) go to 111
+c-----------------------------------------------------------------------
+c
+c Divide the computational volume into three equal pieces.  We must
+c have at least 5 active zones in the 3-direction.
+c
+       k1 = int( real( ke - ks + 1 ) / 3.0 ) + ks
+       k2 = int( real( ke - ks + 1 ) / 3.0 ) + k1
+c
+c i boundaries
+c
+c    1) Post sends and receives.
+c
+       nreq = 0
[…]

### 2014-10-20 sims/zeusGAP/src_dead/rad_imp_dt.F

@@ -1,0 +1,120 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////           R A D _ I M P _ D T             \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c                 University of California at San Diego
+c
+c=======================================================================
+c
+       subroutine rad_imp_dt(erold,ernew) 
+c
+c  INPUT
+c    erold   radiation energy density at old time step
+c    ernew   updated radiation energy density
+c
+c  OUTPUT
+c    dtimrdi  proposed new timestep
+c
+c  EXTERNALS:
+c
+c  AUTHOR
+c    Robert Fiedler and John Hayes
+c
+      use real_prec
+      use config
+      use param
+      use field
+      use grid
+      use root
+      use radiation
+      use scratch
+      use cons
+      use mpiyes
+      use mpipar
+c
+      implicit NONE
+c
+      integer  :: i    , j    , k,
+     1            imax , jmax , kmax,
+     &            iamax, jamax, kamax
+c
+      real(rl) :: erold(in,jn,kn) , ernew(in,jn,kn)
+      real(rl) :: egold(in,jn,kn) , egnew(in,jn,kn)
+      real(rl) :: dtec, er_max, ader, ader_max, fac_dt,
+     .            eg_max, adeg, adeg_max, ade_max
+c
+      real(rl) :: delta_t_max
+      real(rl) :: inp(2), out(2)
+      real(rl) :: trnew, trold
+c
+      data fac_dt /1.0592537/
+c
+c-----------------------------------------------------------------------
+c
+c Find the minimum time step required by the Courant condition for
+c this tile.
+c
+c
+c-----------------------------------------------------------------------
+c
+c Find max new rad energy and max abs rad energy change for this tile.
+c
+      iamax = is
+      jamax = js
+      kamax = ks
+      ader_max = abs((ernew(is,js,ks)-erold(is,js,ks))/
+     .                (erold(is,js,ks)+tiny))
+      do k=ks,ke
+       do j=js,je
+        do i=is,ie
+         ader = abs((ernew(i,j,k) - erold(i,j,k))/
+     .              (erold(i,j,k)+tiny))
[…]

### 2014-10-20 sims/zeusGAP/src_dead/rad_solve.F

@@ -1,0 +1,43 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////              R A D S O L V E              \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c                 University of California at San Diego
+c
+c=======================================================================
+      subroutine rad_solve
+c
+      use real_prec
+      use config
+      use root
+      use field
+      use scratch
+      use mpiyes
+      use mpipar
+c
+      implicit NONE
+c
+c-----------------------------------------------------------------------
+c     radiation solver menu:
+c    
+c     LRAD = 0:  No radiation transport
+c     LRAD = 1:  Grey, flux-limited (non-equilibrium) diffusion
+c     LRAD > 1:  User-supplied options
+c-----------------------------------------------------------------------
+c
+      if(lrad .eq. 1) then
+       call grey_fld(w3dd, e, w3de, w3dh, er)
+       call rad_imp_dt(w3dh,er)
+       dtnri2 = dtnri2 / (1.26*1.26)
+      endif
+      if(lrad .gt. 1) then
+       if(myid .eq. 0) 
+     .    write(*,"('RAD_SOLVE: Unsupported value of LRAD')")
+       call mpi_finalize(ierr)
+       stop
+      endif
+c
+      return
+      end

### 2014-10-20 sims/zeusGAP/src_dead/sendrec.F

@@ -1,0 +1,176 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////           S E N D R E C _ B N D           \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c                 University of California at San Diego
+c
+c=======================================================================
+      subroutine sendrec_bnd(isx,iex,isy,iey,isz,iez,x)
+c
+      use real_prec
+      use config
+      use param
+      use mpiyes
+      use mpipar
+c
+      implicit none
+c
+      integer  :: isx, iex, isy, iey, isz, iez, i, j, k, n, nelm
+c
+      real(rl) :: x(neqm,in,jn,kn)
+c
+c-----------------------------------------------------------------------
+c
+      nelm = 1
+c
+c                               ! send south; receive north
+      if(ntiles(1) .gt. 1) then
+       if((periodic(1) .eqv. .true.) .or. (coords(1) .ne. 0))
+     .  then
+        nreq = nreq + 1
+        call mpi_isend(
+     .        x(1,isx,1,1),nelm,ils_slice,n1m,15700+nsub,
+     .        comm3d,req(nreq),ierr)
+       endif
+       if((periodic(1) .eqv. .true.) .or. (coords(1) .ne. ntiles(1)-1))
+     .  then
+        nreq = nreq + 1
+        call mpi_irecv(
+     &        x(1,iex+1,1,1),nelm,ils_slice,n1p,15700+nsub,
+     &        comm3d,req(nreq),ierr)
+       endif
+c
+c                               ! send north; receive south
+       if((periodic(1) .eqv. .true.) .or. (coords(1) .ne. ntiles(1)-1))
+     .  then
+        nreq = nreq + 1
+         call mpi_isend(
+     .        x(1,iex,1,1),nelm,ils_slice,n1p,15800+nsub,
+     &        comm3d,req(nreq),ierr)
+       endif
+       if((periodic(1) .eqv. .true.) .or. (coords(1) .ne. 0))
+     .  then
+        nreq = nreq + 1
+        call mpi_irecv(
+     &        x(1,isx-1,1,1),nelm,ils_slice,n1m,15800+nsub,
[…]

### 2014-10-20 sims/zeusGAP/src_dead/textdmp.F

@@ -1,0 +1,132 @@
+c=======================================================================
+c/////////////////////////  EXAMPLE USERDUMP  \\\\\\\\\\\\\\\\\\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c               University of Illinois at Urbana-Champaign
+c
+      subroutine textdmp
+c
+c  written by:   Robert Fiedler
+c
+c  PURPOSE:  Example USERDUMP routine to generate ASCII output
+c            files during a ZEUS-MP run.
+c
+c  USAGE:  To use this routine, include "-u textdmp" on the
+c          Make_zeusmp command line.
+c
+c  DESCRIPTION:  This example USERDUMP routine prints out the
+c                coordinates and the density, energy density,
+c                and velocity components for a 3-D
+c                hydrodynamics simulation.  It must be modified
+c                to list magnetic field components, 
+c                gravitational potential, radiation energy density, 
+c                etc.  
+c
+c-----------------------------------------------------------------------
+      use real_prec
+      use param
+      use config
+      use field
+      use grid
+      use bndry
+      use root
+      use cons
+      use scratch
+      use opac_law
+      use mpiyes
+      use mpipar
+c
+      implicit NONE
+c
+      integer :: i, j, k
+      real(rl) :: etotal, ekin, eint, etot_glb, dvb, totdvb,
+     .            glb_j, jtot, glb_ek(3),
+     .            ek(3), v1scr(in), v2scr(in), b2scr(in), b3scr(in),
+     .            v3scr(in), b1scr(in), dscr(in), pscr(in)
+c
+      real(rl) :: so_eps , tau    , eriibn, timarg
+c
+      common /soui/ so_eps, tau, eriibn
+c
+c\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\\///////////////////////////////////
+c=======================================================================
+c
+c Write out the coordinates and the density for each zone in the
+c computational domain to unit 12 (usrfile).
+c
+12    format(1p5e16.8)
+13    format(1p6e16.8)
+14    format(1p7e16.8)
+15    format(1p8e16.8)
+      open(12,file=usrfile)
+c
+c-----------------------------------------------------------------------
[…]

### 2014-10-20 sims/zeusGAP/src_dead/transprt.F

@@ -1,0 +1,308 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////              T R A N S P R T              \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c               University of Illinois at Urbana-Champaign
+c
+c=======================================================================
+c
+       subroutine transprt
+c
+c    mln:zeus3d.transprt <------------------ controls the transport step
+c                                                          october, 1987
+c
+c    written by: Mike Norman
+c    modified 1: June, 1988 by Jim Stone; incorporated into ZEUS2D
+c    modified 2: February, 1990 by David Clarke; incorporated into
+c                ZEUS3D
+c    modified 3: Feb. 15, 1996 by Robert Fiedler; completely
+c                rewritten for ZEUS-MP.
+c    modified 4: Dec. 20, 1996 by Robert Fiedler; added radiation.
+c    modified 5: Jan. 21, 1997 by Robert Fiedler; NO_TRANSPORT switch
+c    modified 6: Dec. 30, 1999 by PSLi; added update of momenta.
+c
+c  PURPOSE: This subroutine transports the field variables through the
+c  mesh in a directionally split manner.  In each succesive call to
+c  TRANSPRT, the order of the directions is permuted (resulting in
+c  XYZ...YXZ...YZX...ZYX...ZXY...XZY...XYZ...etc.).  This MAY be better
+c  than leaving the order the same each time (XYZ...XYZ...etc), and
+c  seems to be better than unsplit schemes (Hawley).  Momenta are
+c  computed from velocities in "avisc" and then transported.  Velocities
+c  are not updated until the end of the transport step.  
+c
+c  The magnetic field components are updated by CT which is a merger (as
+c  implemented by Jim Stone) of the method of characteristics and a
+c  variant of the constrained transport algorithm of Evans and Hawley.
+c
+c  Note that the order in which variables are transported is important
+c  (especially d).  
+c
+c  LOCAL VARIABLES:
+c
+c  EXTERNALS:
+c    CT
[…]

### 2014-10-20 sims/zeusGAP/src_dead/tranx3.F

@@ -1,0 +1,373 @@
+c=======================================================================
+c
+c    \\\\\\\\\\      B E G I N   S U B R O U T I N E      //////////
+c    //////////                T R A N X 3                \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c               University of Illinois at Urbana-Champaign
+c
+c=======================================================================
+c
+       subroutine tranx3 (ibeg,iend,jbeg,jend,kbeg,kend
+     &                   ,dlo,den
+     &                   ,eod,edn
+     &                   ,ero,ern
+     &                   ,abo,abn
+     &                   ,mflx,atwid,atwid2,dtwid,dd,mflux,etwid,deod)
+c
+c    dac:zeus3d.tranx3 <----- transports zone-centred variables along x3
+c    from jms:zeus2d.tranx2, mln:zeus04.tranz                  may, 1990
+c
+c    written by: David Clarke
+c    modified 1: June 1992, by David Clarke; added the total energy
+c                option originally designed by Byung-IL Jun.
+c    modified 2: Feb. 22, 1996 by Robert Fiedler; completely rewritten
+c                for ZEUS-MP.
+c    modified 3: Aug. 6,  1996 by Robert Fiedler; rearranged for
+c                maximum efficiency, and unrolled i-loops.
+c    modified 3: Dec. 18,  1996 by Robert Fiedler; added radiation.
+c    modified 4: by John Hayes; rewrote for F90 and removed CPP logic
+c    modified 5: by John Hayes; restored metric factors multiplying "xi"
+c    modified 6: by John Hayes; corrected transposed array subscripts in
+c                "dxo" array
+c
+c  PURPOSE:  Transports all zone centred variables in the 3-direction
+c  only.  Currently transported are:
+c
+c                      mass   density
+c                      energy density
+c
+c  The consistent advection algorithm, in which mass fluxes are used to
+c  construct the fluxes of all variables across the interfaces, is used
[…]

### 2014-10-20 sims/zeusGAP/src_dead/vsg.F

@@ -1,0 +1,815 @@
+      subroutine vsgiso(cylindrad, zaxis, beta)
+      use planet
+c     we solve for the approximate vertical structure of a self-gravitating, vertically isothermal disk    
+      implicit none  
+      real*8, intent(in) :: cylindrad, zaxis
+      real*8, intent(out) :: beta
+      real*8 :: beta0, bH, paramK, zprime, radius 
+      real*8, external :: bigK, get_varray, bigH
+      
+      common /share/ paramK, radius 
+      common /sharebeta/ beta0 
+      radius = cylindrad 
+c     Initialize a guess for beta0
+      beta0 = beta_trial
+c     get the K parameter for this radius. and re-scale the vertical axis (make it z')
+      
+      bH = bigH(cylindrad)
+      paramK    = bigK(cylindrad)
+      zprime = zaxis/(bH*sqrt(paramK))
+     
+      call get_beta0(beta0) ! Solve for beta_0
+      
+      beta = get_varray(zprime)*exp(0.5*(zaxis/bH)**2)
+           
+c     print*, 'got here', beta
+      return
+      end
+      real*8 function bigK(r)
+c     Get the radius-dependent paramter K in the ODE
+      use planet
+      implicit none
+      real*8, parameter :: bigG = 1d0, pi=2d0*acos(0d0)
+      real*8 :: r, ciso, omega_k
+      real*8, external :: surfden, csq
+      ciso = sqrt(csq(r))
+      omega_k = r**(-3d0/2d0)
+      bigK = ciso*omega_k/(pi*bigG*surfden(r))
+      bigK = bigK*sqrt(pi/8d0)
+      return
+      end function bigK
+      real*8 function surfden(r)
+!the density is sigma_0 * f(r) *exp(-z^2/2H^2) /( sqrt(2*pi)*H)
+!this function calculations the sigma_0 * f(r) \equiv sigma(r) 
+      use planet
+      use domain
+      implicit none
+      real*8 :: r, hin
+      real*8, external :: bigH, inner_hole
+   
+      surfden = den0*r**(-sig)*inner_hole(r)
+      
+      return 
+      end function surfden
+      subroutine get_beta0(beta_0)
+!! beta_0 is an initial condition, but we do not know how to set its value directly.
[…]

### 2014-10-20 sims/zeusGAP/src_dead/zeusmp.F

@@ -1,0 +1,591 @@
+c=======================================================================
+c
+c    \\\\\\\\\\        B E G I N   P R O G R A M          //////////
+c    //////////               Z E U S M P                 \\\\\\\\\c
+c                            Developed by
+c                Laboratory of Computational Astrophysics
+c               University of Illinois at Urbana-Champaign
+c
+c=======================================================================
+c
+      program zeusmp
+c
+c PURPOSE
+c   Main program for 3-D MPI version of ZEUS.
+c
+c AUTHOR
+c   Robert A. Fiedler
+c
+c LAST MODIFIED by JCH, for F90
+c   6/26/02.
+c.......................................................................
+c
+c DECLARATIONS
+c
+      use real_prec
+      use config
+      use param
+      use field
+      use grid
+      use root
+      use scratch
+      use mpiyes
+      use mpipar
+      use gravmod
+      use clockmod
+      use impsoln
+      use planet
+      use gboundary
+c
+      implicit NONE
+c
+      logical :: pert
+      real(rl4) :: cputime, wclock
+c
+      real(rl) :: zcs, etot, etot_glb
+      real(rl) :: cpuall
+c
+      integer :: i , j , k, nwrite
+      integer :: nx, ny, nz, snz, maxn
+c
+      ifsen(1) = 0
+      ifsen(2) = 0
+      ifsen(3) = 1
+      ifsen(4) = 1
+      ifsen(5) = 1
+c
+      myid_w    = 0
+      myid      = 0
+      nprocs_w  = 1
+      nprocs    = 1
+      coords(1) = 0
+      coords(2) = 0
+      coords(3) = 0
+      reorder   = .false.
+      totcgit = 0
+      ncgcall = 0
+c
+c  call CONFIGURE
+c
+      call configure
+c
+c  Master writes greeting.
+c
+      if (myid_w .eq. 0) then
+       call options
+      endif
+c
+c Set up the problem: read input deck and possibly a restart file.
+c
+      call mstart
+c
+c Write out initial data dumps.
+c
+      call dataio( ifsen(2), ifsen(3), ifsen(4), ifsen(5), ifsen(6))
+c
+c Create best calculating plans for FFTW to perform forward
+c and backward FFT.
+c 
+c
+c  Initialize cpu and wall clocks.  The quantities "cputime" and
+c  "wclock" are the CPU and wall clock times (in seconds) in the main 
+c  loop.
+c
+        wclock0 = 0.0
+        cputime0 = 0.0
[…]

### 2014-10-21 paper/model.tex
tex edits, first figures

@@
 contraint set by the inner disc boundary. 
-The 2D Poisson
-equation is solved in integral form using Fast Fouier Transform. This
-requires a softening length $\epsilon_g H$, where $\epsilon_g$ is an
-input parameter.
+The 2D Poisson equation is solved in integral form 
+\begin{align}\label{2d_grav}
+  \Phi_{d,z=0}(R,\phi) = - \int
+  \frac{G\Sigma(R^\prime,\phi^\prime)}{\sqrt{R^2+R^{\prime 2} -
+      2RR^\prime\cos{(\phi - \phi^\prime)} + \epsilon_g^2}} 
+\end{align}
+using Fast Fouier Transform (FFT), where $\epsilon_g$ is a softening
+length to prevent a numerical singularity and the integration is taken
+over the disc. The FFT approach requires
+$\epsilon_g\propto R$ \citep{baruteau08}.
 
@@
 
-\section{Fiducial simulations}
-Here we describe one numerical simulation from each code to give an
-overall picture of disc evolution. In the next section, we explore the
-effect of disc parameters using 2D simulations.
+\section{Diagnostics}
 
+\subsection{Evolution of non-axisymmetirc modes}
 The disc evolution is quantified using mode amplitudes and angular
-momenta as follows. We list the 2D definitions with obvious 3D
-generalisations. The hydrodynamic variables are written as 
+momenta as follows. We list the 2D definitions with obvious 3D generalisations. 
+A hydrodynamic variable $f$ (e.g. $\Sigma$) is written as 
 \begin{align}
@@
 \end{align}
-where the $f_m$ may be obtained from Fourier transform in $\phi$. The
-mode amplitude is 
+where the $f_m$ may be obtained from Fourier transform in $\phi$.
+
+The non-axisymmetric surface density can be visualised with
+\begin{align}
+  \Delta \Sigma = \frac{\Sigma - \Sigma_{00}}{\Sigma_{00}}, 
+\end{align}
+where $\Sigma_{00} = \Sigma_0(t=0)$, while an individual mode as
+\begin{align}
+  \Delta\Sigma_m = \frac{2}{\Sigma_{00}} \real\left[\Sigma_m \exp{(\ii
+      m\phi)}\right].
+\end{align}
+
+The global mode amplitude is 
 \begin{align}
@@
 
+The time evolution of the $m^\mathrm{th}$ mode can be characterized by
+assuming 
+$\Sigma_m\propto\exp{(\ii \sigma t)}$, where
+$\sigma = \omega - \ii\gamma$ is the complex frequency, $\omega$ is
+the real frequency and $\gamma$ is the growth rate.
+We estimate the growth rate as
+\begin{align}
+  \gamma = \frac{d\ln{A_m}}{dt},
+\end{align}
+and we estimate the real frequency by calculating the pattern speed
+$\Omega_p = \omega/m$,
+\begin{align}
+  \Omega_p = \frac{d\phi_\mathrm{max}}{dt},
+\end{align}
+where $\phi_\mathrm{max}$ is the azimuthal co-ordinate of
+$\mathrm{max}(\Delta\Sigma_m)$. 
+%\begin{align}
+%  \mathrm{max}\left\{2 \real\left[ \Sigma_m \exp{(\ii m
+%        \phi)}\right]\right\},
+
+%other methods possible too
+
+\subsection{Angular momentum decomposition}
 The total disc angular momentum in $R\in[R_1,R_2]$ is
@@
       m}^*\right] RdR, \notag\\
-  &= \sum_{m=0}^\infty 2\pi \int_{R_1}^{R_2} j_m R dR = \sum_{m=0}^\infty J_m. 
+  &= \sum_{m=0}^\infty \int_{R_1}^{R_2}  j_m(R) dR = \sum_{m=0}^\infty J_m. 
 \end{align}
 where $^*$ denotes complex conjugation, and we idenfity $J_m$ as the angular momentum 
-associated with the $m^\mathrm{th}$ mode, and $j_m$ its density.
+associated with the $m^\mathrm{th}$ mode, and $j_m$ the associated
+angular momentum per unit length.
 
-\subsection{WKB perturbations}
-We will use results from linear density wave theory to set up
-perturbations. 
-Here, we follow conventions in \cite{shu91}. Recall the
-dispersion relation for tightly-wound density waves of the form
-$\exp{(\ii \omega t - m \phi + kR)}$ in a razor-thin
-isothermal disc, 
+\subsection{Linear density waves}
+We will also use linear density wave theory to interpret
+simulation data. Here, we follow conventions in \cite{shu91} and quote
+results therein. The dispersion relation for tightly-wound density
+waves of the form 
+$\exp{[\ii(\omega t - m \phi + kR)]}$ in a razor-thin
+isothermal disc is 
 \begin{align}\label{dispersion}
@@
 \end{align}
-where $\omega$ is the mode frequency, $\kappa^2 =
-R^{-3}\p_R(R^4\Omega^2)$ is the square of the epicycle frequency, and
-$k$ is a real wavenumber.
+where $\kappa^2 = R^{-3}\p_R(R^4\Omega^2)$ is the square of the epicycle frequency, 
+and $k$ is a real wavenumber such that $|kR|\gg1$.
 
-If we specify the pattern speed $\Omega_p \equiv \omega/m$,
+Given the pattern speed $\Omega_p$,
 Eq. \ref{dispersion} can be solved for $|k|$, 
@@
 \end{align}
-is a dimensionless frequency. 
-We use Eq. \ref{wavenumber} to set trailing long waves as
-perturbations, i.e. $k/k_T = -2Q^{-2}\left[1 - \sqrt{1 -
-    Q^2(1-\nu^2)}\right]$.  
-We only apply the perturbations where $Q^{-2}(1-\nu^2)<1$.
+is a dimensionless frequency.  The upper (lower) signs in
+Eq. \ref{wavenumber} correspond to short (long) waves.
 
-The amplitude of the surface density and velocity perturbations are
-related by
+The \emph{co-rotation radius} $R_c$ is where the pattern speed matches
+the fluid rotation,
 \begin{align}
-  &\left(\omega - m\Omega\right)\delta\Sigma + k\Sigma\delta v_R= 0,\\
-  &\frac{\kappa^2}{2\Omega} \delta v_R + \ii (\omega - m\Omega)\delta
-  v_\phi=0. 
+  \Omega(R_c) = \Omega_p.
 \end{align}
-We set $\delta\Sigma = 10^{-3}\Sigma$ and $(\delta v_R, \delta
-v_\phi)$ accordingly.
+\emph{Lindblad resonances} $R_L$ occurs where
+\begin{align}
+  \nu^2(R_L) = 1. 
+\end{align}
+Finally, \emph{Q-barriers} occur at radii $R_{Qb}$ where
+\begin{align}
+  Q^2(R_{Qb})\left[1-\nu^2(R_{Qb})\right] = 1.  
+\end{align}
+According to Eq. \ref{wavenumber}, purely wave-like solutions with
+real $k$ are only possible when $Q^2(1-\nu^2)\leq1$.
 
-%Having chosen the wave frequency and set the wavenumber according to
-%the above,
+% We use Eq. \ref{wavenumber} to set trailing long waves as
+% perturbations, i.e. $k/k_T = -2Q^{-2}\left[1 - \sqrt{1 -
+%     Q^2(1-\nu^2)}\right]$.  
+% We only apply the perturbations where $Q^{-2}(1-\nu^2)<1$.
+
+% The amplitude of the surface density and velocity perturbations are
+% related by
+% \begin{align}
+%   &\left(\omega - m\Omega\right)\delta\Sigma + k\Sigma\delta v_R= 0,\\
+%   &\frac{\kappa^2}{2\Omega} \delta v_R + \ii (\omega - m\Omega)\delta
+%   v_\phi=0. 
+% \end{align}
+% We set $\delta\Sigma = 10^{-3}\Sigma$ and $(\delta v_R, \delta
+% v_\phi)$ accordingly.
+
+% %Having chosen the wave frequency and set the wavenumber according to
+% %the above,
+
+\section{Fiducial simulations}
+Here we describe one numerical simulation from each code to give an
+overall picture of disc evolution. In the next section, we analyse 2D
+simulations in more detail and explore parameter space.
+
+In these simulations we subject the disc to random perturbations in
+cylindrical velocity,
+\begin{align}
+  \delta v_R = \frac{\delta}{M}c_s(R)\sum_{m=1}^M\cos{m\phi},
+\end{align}
+where $\delta\in[-10^{-3},10^{-3}]$ is chosen randomly but does not
+depend on $\phi$, and we set $M=10$.
+
+\subsection{FARGO 2D}
+We consider a disc size $[R_\mathrm{min}, R_\mathrm{max}] =
+[0.4,10]R_0$ and resolution $N_R\times N_\phi = 1024\times 2048$ or
+about $16$ grids per $H$. We adopt $\epsilon_g=10^{-4}H$ for the 
+self-gravity softening length\footnote{In 2D self-gravity, $\epsilon_g$ also
+approximates for the vertical disc thickness. A more appropriate
+value would be $\sim H$ \citep{muller12}. However, because
+$\epsilon_g\propto R$ is needed in FARGO, the Poisson kernel
+(Eq. \ref{2d_grav}) is no longer symmetric. We choose a
+small  
+$\epsilon_g$ in favour of angular momentum conservation, keeping in
+mind that the strength of self-gravity will be over-estimated.}.
+
+\subsection{ZEUS-MP 3D}
+
+\subsection{PLUTO 3D}

### 2014-10-21 paper/model.tex
minor update

@@
 $\epsilon=0.1$, $\delta R_d=5$, $h=0.05$ and
-$Q_\mathrm{out}=2$. Initially there is no vertical or radial velocity
+$Q_\mathrm{out}=2$. An example of the initial surface density and the
+Toomre parameter (see\ref{wkb}) is shown in Fig. \ref{initial_surf}. 
+Initially there is no vertical or radial velocity
 ($v_R = v_r = v_\theta = 0$). The azimuthal velocity is initialized to
@@
 and similarly in 2D. The angular velocity is $\Omega = v_\phi/R$.
+
+\begin{figure}
+  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.7cm 0cm
+  0cm]{figures/compare_profiles_dens000} 
+  \includegraphics[width=\linewidth]{figures/compare_profiles_Q000}
+  \caption{Fiducial profiles of the surface density (top) and Toomre
+    parameter (bottom) used in this work.\label{initial_surf}}
+\end{figure}
 
@@
   \Phi_{d,z=0}(R,\phi) = - \int
-  \frac{G\Sigma(R^\prime,\phi^\prime)}{\sqrt{R^2+R^{\prime 2} -
+  \frac{G\Sigma(R^\prime,\phi^\prime)R^\prime dR^\prime d\phi^\prime}{\sqrt{R^2+R^{\prime 2} -
       2RR^\prime\cos{(\phi - \phi^\prime)} + \epsilon_g^2}} 
@@
 \begin{align}
-  \Delta \Sigma = \frac{\Sigma - \Sigma_{00}}{\Sigma_{00}}, 
+  \Delta \Sigma = \frac{\Sigma - \Sigma_{0}}{\Sigma_{0}}, 
 \end{align}
-where $\Sigma_{00} = \Sigma_0(t=0)$, while an individual mode as
+while an individual mode as
 \begin{align}
   \Delta\Sigma_m = \frac{2}{\Sigma_{00}} \real\left[\Sigma_m \exp{(\ii
-      m\phi)}\right].
+      m\phi)}\right]
 	\end{align}
+where $\Sigma_{00} = \Sigma_0(t=0)$.
 
@@
 
-\subsection{Linear density waves}
+\subsection{Linear density waves}\label{wkb}
 We will also use linear density wave theory to interpret

### 2014-10-22 paper/model.tex
minor updates

@@
 $B(R)$ represents a surface density boost between
-$R\in[R_{d1},R_{d2}]$ by a factor $\epsilon^{-1}$ ($\epsilon < 1$),
+$R\in[R_{d1},R_{d2}]$ by a factor $\epsilon^{-1}>1$,
 and $\delta R_d$ is the transition width between the bump and the
@@
 % %the above,
-
-\section{Fiducial simulations}
-Here we describe one numerical simulation from each code to give an
-overall picture of disc evolution. In the next section, we analyse 2D
-simulations in more detail and explore parameter space.
-
-In these simulations we subject the disc to random perturbations in
-cylindrical velocity,
-\begin{align}
-  \delta v_R = \frac{\delta}{M}c_s(R)\sum_{m=1}^M\cos{m\phi},
-\end{align}
-where $\delta\in[-10^{-3},10^{-3}]$ is chosen randomly but does not
-depend on $\phi$, and we set $M=10$.
-
-\subsection{FARGO 2D}
-We consider a disc size $[R_\mathrm{min}, R_\mathrm{max}] =
-[0.4,10]R_0$ and resolution $N_R\times N_\phi = 1024\times 2048$ or
-about $16$ grids per $H$. We adopt $\epsilon_g=10^{-4}H$ for the 
-self-gravity softening length\footnote{In 2D self-gravity, $\epsilon_g$ also
-approximates for the vertical disc thickness. A more appropriate
-value would be $\sim H$ \citep{muller12}. However, because
-$\epsilon_g\propto R$ is needed in FARGO, the Poisson kernel
-(Eq. \ref{2d_grav}) is no longer symmetric. We choose a
-small  
-$\epsilon_g$ in favour of angular momentum conservation, keeping in
-mind that the strength of self-gravity will be over-estimated.}.
-
-\subsection{ZEUS-MP 3D}
-
-\subsection{PLUTO 3D}

### 2014-10-22 paper/paper.tex
minor updates

@@
 \begin{abstract}
-
+  Theoretical models of circumstellar or protoplanetary discs indicate
+  the possible existence of `dead zones' --- regions in the disc
+  where mass accretion is greatly reduced because of inefficient
+  magneto-hydrodynamic turbulence that would otherwise enable 
+  accretion. It has been suggested that mass built-up in a dead zone
+  may lead to gravitational instability. However, the type of
+  gravitational instability relevant in this context has not been
+  clarified. In this work, we use numerical simulations to explore
+  gravitational instability in astrophysical discs in which only a 
+  sub-portion of it is self-gravitating, taken as a toy model for
+  protoplanetary disc containing a dead zone. We conduct numerical hydrodynamic
+  experiments in 2D using the FARGO code and in 3D with the ZEUS-MP
+  and PLUTO codes. In all cases we observe the development of a spiral
+  mode with azimuthal wavenumber $m=1$ inside the dead
+  zone. Surprisingly, the co-rotation radius of the spiral is well
+  outside the dead zone, resulting in a nearly stationary one-arm spiral
+  pattern in the dead zone. Our simulations demonstrate the potential
+  importance global gravitational interaction between a
+  self-gravitating dead zone and the exterior, non-self-gravitating
+  parts of the disc.  
 \end{abstract}
@@
 \input{model}
-%\input{results}
+\input{results}
 %\input{discussion}

### 2014-10-22 paper/results.tex
minor updates

@@
-\section{Growth of non-axisymmetric modes without the influence of the
-  planet}\label{linear1}
-In this section, the planet is introduced at $t=20P_0$ and 
-its potential switched on over $10P_0$. At $t=30P_0$ we switch off the
-planet potential and azimuthally average the surface density, energy
-and velocity fields. At this point the planet has carved a partial
-gap. We then perturb the surface density for $r>r_p$ and continue to 
-evolve the disc. We impose  sinusoidal perturbations with 
-azimuthal wavenumbers $m\in[1,10]$ and random amplitudes within $\pm 0.01$
- times the local surface density.% {\bf what are PERTMIN and PERTMAX values?}  
-This procedure allows us to analyse the growth of 
-non-axisymmetric modes associated with the gap, but without
-complications from non-axisymmetry arising directly from disc-planet
-interaction.
+\section{Preliminary experiments}
+Here we describe one numerical simulation from each code to give a
+qualitative picture of disc evolution. In the next section, we analyse
+2D simulations in more detail and explore parameter space.
 
-Note that these `planet-off' simulations are not linear stability
-calculations because since the cooling term in our energy equation
-restores the initial temperature profile corresponding to $H/r=0.05$,
-rather than the heated gap edge. Linear stability calculations and
-adiabatic cases will be further discussed in
-section~\ref{adiabatic_section}.
+In these simulations we subject the disc to random perturbations in
+cylindrical velocity,
+\begin{align}\label{randpert}
+  \frac{\delta v_R}{c_s} = \frac{\delta}{M}\times T(R) \sum_{m=1}^M\cos{m\phi},
+\end{align}
+where $\delta\in[-10^{-3},10^{-3}]$ is chosen randomly but does not
+depend on $\phi$, and 
+\begin{align}
+  T(R) =
+  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}_d}{\Delta
+          R_d}\right)^2\right]}, 
+\end{align}
+where $\overline{R}_d = (R_{d1}+R_{d2})/2$ and $\Delta R_d =
+(R_{d2}-R_{d1})/2$. 
+\subsection{FARGO 2D}\label{fargo_fiducial}
+We consider a disc size $[R_\mathrm{min}, R_\mathrm{max}] =
+[0.4,10]R_0$ and resolution $N_R\times N_\phi = 512\times 1024$ or
+about $8$ grids per $H$. We adopt $\epsilon_g=10^{-4}H$ for the 
+self-gravity softening length\footnote{In 2D self-gravity, $\epsilon_g$ also
+  approximates for the vertical disc thickness, so a more appropriate
+  value would be $\epsilon_g\sim H$ \citep{muller12}. However, because
+  $\epsilon_g\propto R$ is needed in FARGO, the Poisson kernel
+  (Eq. \ref{2d_grav}) is no longer symmetric in $(R,R^\prime)$. We
+  choose a small  
+  $\epsilon_g$ in favour of angular momentum conservation, keeping in
+  mind that the strength of self-gravity will be over-estimated.}, and 
+set $M=10$ for random perturbations (Eq. \ref{randpert}).
 
-Simulations here employ a resolution of $(N_r,N_{\phi})=(1024,2048)$
-with an open boundry condition at the outer disk edge $r_\mathrm{out}=25r_\mathrm{in}$. We consider 
-%for a time up to $200P_0$ allowing high resolution and detection of high
-%$m$ structures.
-cases of $\tilde{\beta}=0.1,1,10$ corresponding to fast, moderate,
-and slowly cooled discs. % {\bf boundary condition for these sims?}
-
-\subsection{Gap structure}
-%Since stability of the gap is of interest 
-We first examine the gap structure formed by planet-disc
-interaction as a function of the cooling time. The azimuthally-averaged 
-gap profiles are shown in Fig. \ref{intial1D} for varying
-$\tilde\beta$. Gaps formed with lower $\tilde\beta$ (faster cooling)
-are deeper with steeper gradients at the gap edges. Faster cooling rates also 
-increase and decrease the surface density maxima and 
-minima increase, respectively. However, a clean gap does not form
-in this short time period.
-
-%For all cases the gap width is $\approx 5r_{hill}$
-%which is set by torque balances in the disk \citep{crida06}.
-
-Larger $\tilde\beta$ values result in higher disc aspect ratios $h=H/r$,
-i.e. higher temperatures. Heating mostly occur at the gap edges
-due to planet-induced spiral shocks. Decreasing the cooling time
-implies that this heat is retained in the disc. In the inviscid limit the gap
-opening condition is $r_h\gtrsim H$ or $q\gtrsim 3h^3$
-\citep{crida06}. 
-% which is a value directly correlated to temperature in our model. 
-%For a planet to carve a gap in the surrounding material it has been
-%shown that $q>H^3$ is needed \citep{crida06}. 
-This indicates that for hotter discs (higher
-$h$), it becomes more difficult for a planet of fixed $q$ to open a
-gap. This explains the shallower gaps in surface density when
-$\tilde{\beta}$ is increased.
-
-%This is seen in our simulations as changes in $\tilde{\beta}$
-%result in changes of $h$ by less than $20\%$ yet gap depth almost
-%doubles.
-
-The important consequence of a heated gap edge is that the
-generalized vortensity profiles become smoother with increasing
-cooling times, with the extrema becoming less pronounced. Previous locally
-isothermal disc-planet simulations show the RWI associated with PV
-minima \citep{li05,lin10}. We can therefore expect the RWI to be associated with
-minima in the generalized vortensity (corresponding to local surface
-density maxima) in the non-isothermal case. Because the extrema are
-less sharp, the RWI is expected to be weaker and the gap to be more
-stable with longer cooling times.
-
-%Charactereristicly
-%for such disk-planet Rossby vortices the generalized vortensity
-%minima correspond to density maxima in the disk. 
-%Since generalized vortensity extrema are also
-% correspondingly less pronounced and density gradients become smaller
-% we expect that the stability of gaps to increase with larger
-% $\tilde{\beta}$.
+Snapshots from the disc evolution is shown in Fig. \ref{fargo_2d} in
+terms of the non-axisymmetric surface density.
 
 \begin{figure}
-  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
-    2cm 0cm 0cm]{figures/compare_sigma}
-  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
-    2cm 0cm 1cm]{figures/compare_aspectratio}
-\includegraphics[width=\linewidth,clip=true,trim=0.5cm
-    0.5cm 0cm 1cm]{figures/compare_gvortensity}
-  \caption{Gap profiles at $t=30P_0$ for the intial partial gap opened
-    before instability emerges for fast (solid), moderate
-    (dashed), and slow cooling (dashed-dot). The relative surface density
-    pertibation (top), disc aspectratio (middle) and generalized
-    vortensity pertibation (bottom) are shown. \label{intial1D}}  
+  \includegraphics[scale=0.27]{figures/polarxy_dens050}\includegraphics[scale=0.27,clip=true,trim=2.26cm 
+  0cm 0cm
+  0cm]{figures/polarxy_dens100}\includegraphics[scale=0.27,clip=true,trim=2.26cm
+  0cm 0cm 0cm]{figures/polarxy_dens125} 
+  \caption{Fiducial FARGO 2D simulation (\S\ref{fargo_fiducial}),
+    showing the evolution of the non-axisymmetric surface density
+    $\Delta\Sigma$. Note that the disc extends to $R=10R_0$, but only
+    the inner part is shown for clarity. \label{fargo_2d}} 
 \end{figure}
 
-%\begin{tabularx}{0.4\textwidth}{l*{10}{R}} \toprule
-%  \multicolumn{11}{c}{$\tilde{\beta}=0.1$} \\ \midrule
-%  m                    & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10  \\ 
- % $\gamma10^2/\Omega(r_o)$ & 6.56 & 6.82 & 6.73 & 5.78 & 6.00 & 6.38 & 5.97 & 5.62 & 4.61 & 3.36   \\ \bottomrule
-%\end{tabularx}
-
-%\begin{tabularx}{0.4\textwidth}{l*{5}{R}} \toprule
-%  \multicolumn{6}{c}{$\tilde{\beta}=1.0$} \\ \midrule
-%  m                    & 1 & 2 & 3 & 4 & 5  \\ 
-%  $\gamma10^2/\Omega(r_o)$ & 1.27 & 1.28 & 1.35 & 1.01 & 0.61   \\ \bottomrule
-%\end{tabularx}
-
-\begin{table}
-  \centering
-  \caption{Dominant mode and growth rates for
-    $\tilde{\beta}=0.1,1.0,10.0$ (fast, moderate, and slow cooling)
-    values during `planet-off' simulations \label{modetable}} 
-  \hfill
-  \begin{minipage}{0.3\linewidth}
-    \begin{tabularx}{\textwidth}{l R} 
-      \multicolumn{2}{c}{$\tilde{\beta}=0.1$} \\ 
-      \toprule
-      $m$ & $10^2\gamma/\Omega(r_p)$ \\
-      \midrule
-      6 & 7.3 \\
-      7 & 7.8 \\
-      8 & 7.9 \\
-      9 & 7.9 \\
-      10 & 6.8 \\ 
-      \bottomrule
-    \end{tabularx}
-  \end{minipage}
-  \hfill
-  \begin{minipage}{0.3\linewidth}
-    \begin{tabularx}{\textwidth}{l R} 
-      \multicolumn{2}{c}{$\tilde{\beta}=1.0$} \\ 
-      \toprule
-      $m$ & $10^2\gamma/\Omega(r_p)$ \\
-      \midrule
-      3 & 2.0 \\
-      4 & 2.2 \\
-      5 & 2.3 \\
-      6 & 1.6 \\
-      7 & 1.1 \\ 
-      \bottomrule
-    \end{tabularx}
-  \end{minipage}
-  \hfill
-  \begin{minipage}{0.3\linewidth}
-    \begin{tabularx}{\textwidth}{l R} 
-      \multicolumn{2}{c}{$\tilde{\beta}=10.0$} \\ 
-      \toprule
-      $m$ & $10^2\gamma/\Omega(r_p)$ \\
-      \midrule
-      1 & 1.1 \\
-      2 & 1.6 \\
-      3 & 1.7 \\
-      4 & 1.2 \\
-      5 & 0.1 \\ 
-      \bottomrule
-    \end{tabularx}
-  \end{minipage}
-  \hfill
-\end{table}
-
-{\bf
-\subsection{Axisymmetric stability}
-Check that the gap profiles are stable against axisymmetric
-stability. Can use analytical criteria given in section 3.1 of Li et
-al (2000, ApJ, 533, 1023). No plot needed, just statements. 
-}
-
-The intial planet-disk interaction form bumps and grooves in the gap profiles
- which can potentially be unstable due to axisymmetric instabilities. The
- generalized local axisymmetric stability condition is the Solberg-Hoiland
- criterion,
-\begin{equation}
- \kappa^2+N^2 \geq 0 $ where $ N^2=\frac{1}{\Sigma} \frac{\partial P}{\partial r} \left(\frac{1}{\Sigma} \frac{\partial \Sigma}{\partial r}-\frac{1}{\gamma P} \frac{\partial P}{\partial r}  \right)
-\end{equation}
-where $N$ is the Brunt-V\"ais\"al\"a frequency for the disk due to entropy
- variation which is non-zero for non-adiabatic models such as being considered.
- At  $t=30P_0$ the outer gap edge of $r=2.5r_h$,  where the RWI is excited the
- criterion value, is found to be approximately $0.002$ for all $\tilde\beta$
- which is the local minimum in the gap region which indicates local
- axisymmetric stability of gap. The Solberg-Hoiland criteria is similary
- satisfied for the entire disk and for all times within the simulation length
- with little change of values. Thus for all cases of $\tilde\beta$ the
- planet-induced gaps formed are stable to axisymmertic instabilities.
-
-\subsection{Non-axisymmetric instability}\label{linear}
-%{\bf present `planet-off' simulations. one `fourier mode v.s. time'
-%  plot to show growth of linear instability. do an adiabatic case (or
-%  extremely long cooling time) to see the effect of heated gap edge
-%  (i.e. temp doesn't go back to t=0 value too quickly)
-%  compare growth rate and dominant
-%  m as function of beta (table). 2D figs to contrast (also used to
-%  show it's the minimum in generalized pv that goes unstable.    
-%  result: increasing cooling time makes the gap
-%  more stable, and favors lower m. note: the `basic state' should be
-%  the system at t=30 after azimuthal average. linear results should
-%  have small perturbations. 
-%}
-
-%After the azimuthal averaging, the small pertibation excites rapidly
-%growing instabilities by the RWI.
-
-In all cases we observe growth of non-axisymmetric structures after the
-planet potential is switched off and the system subject to random
-perturbations. An example is shown in Fig. \ref{linearmodes} for 
-$\tilde{\beta}=10$. We charcterize these
-modes with an azimuthal wavenumber $m$ and growth rate $\gamma(m)$ as defined by
-Eq.~\ref{fouriertransform}---\ref{growth} with radial range of $[2r_h,5r_h]$. Table \ref{modetable}
-lists the growth rates for the three cooling
-times, measured during linear growth.
-
 \begin{figure}
-  \includegraphics[width=\linewidth,clip=true,trim=1.2cm
-    0cm 0cm 0cm]{figures/linear_stability}
-  \caption{Evolution of azimuthal Fourier modes of disc surface
-    density, non-dimenionlized by the axisymmetric background keplerian mode
- $\Sigma_0(t=0)$ for the
-    `planet-off' simulations with $\tilde{\beta}=10$. Colours correspond
-    to different $m$ values. The $m=3$ component is the fastest growing
-    mode during linear growth and has a corresponding
-    $\gamma=0.017\Omega(r_p)$.\label{linearmodes}}
-% {\bf is the $\Sigma_0$ at t=0?} }  
+  \includegraphics[width=\linewidth]{figures/nonaxi_evol_ang_fargo}
+  \caption{Evolution of angular momentum components in the FARGO
+    fiducial simulation (Fig. \ref{fargo_2d}). The perturbation
+    relative to $t=0$ is shown in units of the initial total angular
+    momentum $J_\mathrm{ref}$.\label{fargo_2d_angmom}} 
 \end{figure}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{ZEUS-MP 3D}
 
-Table \ref{modetable} show that as
-$\tilde{\beta}$ was increased from $ 0.1\rightarrow10$ the dominant
-azimuthal Fourier mode decreased from $ m=9\rightarrow3$ and the
-respective growth rate decreased from $ \gamma/\Omega(r_p)=0.079
-\rightarrow 0.017$. However, despite two orders of magnitude increase in the
-cooling time, the instability remains dynamical with growth time
-$\lesssim 10P_0$. Snapshots of the instability in $r-\phi$ plane for
-the different $\tilde\beta$ are shown in Fig \ref{2Dlinear}.
-
-These `planet-off' simulations show that gap edges become more stable with
-longer cooling times. This is expected because larger $\tilde{\beta}$
-result in hotter gap profiles at $t=30P_0$ with less pronounced
-generalized vortensity minima. Stabilization with increased
-cooling time is therefore due to a smoother basic state for the
-instability, as it is more difficult for the planet to open a gap if
-the disc is allowed to heat up.
-
-% which result in hotter gaps since the heating
-%due
-%This is as expected from the
-%intial gap profiles, because 
-% and the gap forming criteria discussed in the
-%previous section.
-
-\begin{figure}
-  \centering
-  \subfigure{
-    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity50}
-  }
-\hfill
-  \subfigure{
-    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity67}
-  }
-\hfill
-  \subfigure{
-    \includegraphics[width=0.3\linewidth]{figures/analysis_gvortensity75}
-  }  
-  \caption{Generalized vortensity perturbation (relative to $t=0$) for
-    cases of $\tilde{\beta}=0.1,1,10$ (left,middle,right) during
-    the growth of non-axisymmetric modes. The planet potential has
-    been switched off.  The number of vortices
-    decrease as $\tilde{\beta}$ increases. Note that snapshots are
-    taken later because it takes longer for the vortices to grow and
-    become visible with increasing cooling time. 
-    %    correspondingly increase to capture the longer growth phases of
-    %    the instabilities. 
-    \label{2Dlinear} {\bf I'm curious to know if plots clearer if
-      perturbations are measured relative to t=30? remember t=30 is
-      our reference state for instability. But we'll probably keep
-      these for consistency (later non-linear plots are all measure
-      with respect to t=0).}} 
-\end{figure}
-
-\subsubsection{Nearly adiabatic discs}
-\label{adiabatic_section}
-%{\bf a case with extrmely long cooling time (or adiabatic run) in
-%  order to look at effect of heated gap edge.  prelim result: not
-%  important, at t=30, gap edge only heats to $H/r\simeq0.06$ even for
-%  purely adiabatic disk. change in $h$ not important for linear
-%  perturbations (but important for setting up the basic state). 
-%}
-The above `planet-off' simulations are not formally linear
-stability calculations, because the cooling time is shorter
-than the instability growth time, 
-%
-%as we do not capture the effects of a heated
-%disk on the instability growth. 
-$t_c<\gamma^{-1}$.  
-Thus the disk cools back to its intial value of $h=0.05$
-before or during the instability growth, implying we do not have a
-steady basic state to formulate a standard linear stability
-problem. %This means the disc temperture of $h=0.05$ in the the above cases
-
-In order to perform a proper linear stability analysis and capture the
-effect of a heated gap edge during instability growth, we ran a simulation  with
-$\tilde{\beta}=100$, corresponding to an almost adiabtaic disc.  
-%To compare our 'planet-off' analysis with a true
-%linear stability analysis 
-In this simulation the cooling rate is slow enough that the gap 
-temperature profile (e.g. middle panel of Fig. \ref{intial1D}) changes
-only marginally over the instability growth timescale. %heat is
-                                %retained/maintained.
-
-We find very similar gap profiles and mode growth rates for
-$\tilde{\beta}=100$ as with $\tilde{\beta}=10$. The disc heats up to
-values $h\simeq0.06$ in the nearly adiabatic case. This is close to
-the original temperature of $h=0.05$, so linear growth rates are not expected
-to change significantly %according to the linear theory presented in
-\citep{li00}.
-
-According to \cite{li00}, increasing $h$ increases linear growth rates
-of the RWI because it is pressure-driven. However, in the case 
-of disc-planet interaction, increasing $h$ has a stabilizing effect
-through the gap profile because it results in smoother gap
-edges. The fact that we observe smaller growth rates as $h$ is
-increased indicate that for planetary gaps, the importance of $h$ on
-the \emph{linear} RWI is through the setting the gap profile (i.e. basic
-state for the instability).
-
-%results in shallower gaps
-%which is expected to be
-
-%the difference in growth rates between this and the
-%original disc temperature of $h=0.05$ changes almost insignificantly
-%as shown by linear theory \citep{li00}. 
-%Although linear
-%theory predicts higher growth rates with increasing $h$
-
-%This indicate that a
-%continuously hot gap edge is not as important for development of
-%vortices as the heating effects on the formation of the intial gap
-%state.
-
-{\bf
-If there's time, do low resolution planet off simulations but keep the planet on
-until $t=40P_0$ or $50P_0$. Look at stability properties when the gaps
-are generally deeper (including the nearly adiabatic case). Although
-with the planet instability may already appear by $t=40P_0$, the
-azimtuhal average performed prior to perturbations will erase it. 
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Non-linear evolution of
-  gap-edge vortices with finite cooling time}\label{nonlinear} 
-%{\bf main fig: vortex amplitude v.s. time for diff beta. enough data
-%  to plot vortex lifetime v.s. beta? table: 
-%  averaged quantities over quasi-steady state: aspect-ratio (to
-%  compare with fu at al), rossby number (vortex strength
-%  v.s. cooling?), maybe alpha visc. vortex size: visible difference?  
-%  only inviscid cases. describe evolution of one case. main
-%  conclusion: longer vortex lifetime with increasing cooling time (up
-%  to some optimal timescale). vortex death: induced-shock and/or
-%  smoothing the gap edge. describe simulation setup, resolution?
-%  should mention that results consistent with lower-resolution prelim
-%  runs. maybe torques? 
-%}
-
-We now consider long term simulations of gap-edge vortices for
-$\tilde{\beta}=0.1,0.5,1,5,10$. The planet potential is kept on
-throughout.  
-% up to a total of
-%$2000P_0$. 
-%Planets were left free to interact dynamiclly with the disk
-%and vortices after $t=30P_0$ as apposed to previous section.  
-We employ a grid with $(N_r,N_{\phi})=(512,1024)$ 
-in order for these simulations to be computationally feasible. We also use a larger
-disc with $r_{\mathrm{out}}=45r_{\mathrm{in}}$ to minimize boundary
-effects on vortex evolution. We apply an open boundary at
-$r_\mathrm{out}$.
-
-Preliminary simulations with same numerical scheme but lower resolution of
- $(N_r,N_{\phi})=(256,512)$ where also done for the above $\tilde\beta$ values
-with the inclusions of $\tilde{\beta}=0.25,0.75,2.5,7.5$ as well.
-Results and vortex behaviour are convergent with
-higher resolution simulations as discussed in this section.
- {\bf mention lower resolution preliminary runs -
-  similar results to those presented here (?). want to demonstrate convergence}
-
-%An open boundry
-%condition was used in the outer edge of
-% allowing significant room for
-%linblad resonances of the vortices.
-
-\subsection{Generic evolution}
-
-The linear growth of the RWI and vortex-formation is followed by 
-vortex merging. Merging timescales are independent of $\tilde\beta$ value,
- and by $70P_o$ only one vortex remains as the growth of these modes are now
- supported by the continuous driving of the planet.
-%i.e. only the $m=1$ mode remains. 
-{\bf is vortex merging affected by $\tilde{\beta}$? e.g. timescale
-  taken to reach one vortex as a function of cooling}
-The evolution of the amplitude of the $m=1$ surface density mode
- over the radial range of $[2r_h,10r_h]$ is shown in Fig~\ref{lifetimeplot}
- for different $\tilde\beta$.
-% {\bf should state (also for planet-off runs)
- % what the radial range is used for averaging the Fourier amplitudes}
-
-\begin{figure}
-  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
-    0cm 0cm 1cm]{figures/longterm_stability}
-  \caption{Log plot of the $m=1$ mode non-dimenionlized by the $m=0$
-    background mode for longterm non-linear simulations corrsponding
-    to vortex lifetime. The longest living mode correpsonds to
-    $\tilde\beta=5.0$ (the red line). \label{lifetimeplot}} 
-\end{figure}
-
-In all cases the system remains in a quasi-steady state for
-$\gtrsim800P_0$ with a single vortex circulating 
-the outer gap edge at the local Keplerian  
-frequency. Fig. \ref{Vortex2D} shows a typical plot of the relative 
-surface density perturbation in this state. During this stage, the 
-vortex grows in intensity. As the vortex forms it has a
- value of $Ro\approx-0.1$ for all $\tilde\beta$
- indicating anti-cyclonic
-motion. As the $m=1$ amplitude grows the relative spin of the vortex
-continuously grows along with it, reaching values of
-$Ro\approx-0.4$ characteristicly for all cooling rates. Relative surface
- density pertibations at the center of vortices can reach values of 7 times the
- intial density for all cases of $\tilde\beta$ in this quasi-steady state and
- can be found as high as 9 times for the $\tilde\beta=1.0$ case.
- The gap profiles apply a surface density boost in the vortex region as the
- vortices are located at the gap edge maxima
- but the values associated with this boost have relative density pertibations
- of under 1 and thus are negligibe compared to the max amplitudes of the vortices.
- Vortex growth is attributed to
-the continuous generation of vorticity by the planet-disc interactions. 
-This is supported by the evidence that in long term simulations where the
-planet is removed such as in section \ref{nonlinearplanetoff} the amplitude of
- the vortices are not seen to contiunously grow without the planet influence.
-
-{\bf for what cooling are these values quoted for? do
-  these values depend on cooling?}
- {\bf is the evidence for this the fact that growth is not observed in
-  planet-off simulations?}
-
-{\bf note that there
-  are two origins for the surface density boost relative to $t=0$: 1) gap
-  formation, which increases the surface density at the location of
-  the instability (since they happen at surface density max., and 2)
-  vortex instability). what is the surface density boost relative to
-  the gap profile?}
-
-Interestingly, in most cases we observe a sudden drop in 
-the $m=1$ amplitude within the simulation time, which occurs over
-timescales of $50P_0$, i.e. much shorter than the duration of the quasi-steady
-state. {\bf is this dissipation timescale independent of cooling? No but will wait for simulation to finish before commenting}.  
-%Eventually
-%the $m=1$ amplitudes are also characterized by 
-After these characteristic drops, the instability is never seen to
-reform at such large overdensities and surface density bumps are never larger
-than that due to the planetary wakes. We designate the time of these
-amplitude drops as the lifetime 
-of the quasi-steady vortex.
-The instability is not seen to emerge again after the intial vortex dissipation
- due to the smoothening out of the gap edge during dissipation by the
- transfer of density from the once tightly held vortex region to the surrounding
- disk and gap.  
-{\bf question: does instability
-  actually happen again? if it does, but just weaker, that may be
-  because the first vortex has smoothed out the gap edge, so
-  instabilities following that would be weaker.}
-
-\begin{figure}
-  \includegraphics[width=\linewidth,height=\linewidth]{figures/vortex2D}
-  \caption{Relative surface density perturbation for the
-    $\tilde\beta=0.1$ case during quasi-steady state with a single
-    vortex. The plot for other values of the cooling time $\tilde{\beta}$ are
-    similar.% {\bf is this plot truncated? in the text it says the
-      %domain goes to $r=45$?}
-    %The large scale non-axisyemetric $m=1$ overdensity can be
-    %seen. 
-    \label{Vortex2D}} 
-\end{figure}
-
-\subsection{Non-linear instability without planet influence} \label{nonlinearplanetoff}
-Non-linear observations were also made for the `planet-off' simulations.
-After the linear growth phase of the vortices, vortex merging takes hold but
- now on timescales of up to $150P_0$ as the vortices grow as linear
- instabilities discussed previously. The vortex merging time is depedendent on
- the growth rates of the modes and how long they take to saturate. Similar
- quasi-steady vortices form but the long-term behaviour is significantly
- different without the planet as seen in Fig.\ref{planetofflifetimeplot}.
- The quasi-steady vortices formed have significantly lower density pertibations
- and now slowly decrease in amplitute in time as apposed to previously
- where their growth was supported by planet interactions. The vortices simply
- dissipate viscously and no longer undergo the quick dissipation that
- characterized the dissipation as earlier. The dissipation rate is dependent
- on the cooling rate with viscous dissipation decreasing with larger
-$\tilde\beta$.
-
-\begin{figure}
-  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
-    0cm 0cm 1cm]{figures/longterm_planetoff}
-  \caption{Similar plot as Fig. \ref{lifetimeplot} but with `planet-off' simulations. Vortices dissipate viscously without planet driving influence} \label{planetofflifetimeplot}
-\end{figure}
-
-{\bf
-\subsection{Additional analysis on vortex dissipation}
-Conjecture: vortex death due to vortex shocking the system and
-dissipating its energy. discuss one case. here we want to explain or
-suggest why vortices die} 
-%Rossby values have dynamics that are coupled with the lifetime and
-%$m=1$ amplitude of the vortices.
-
-As the vortex dies off as described above its associated Rossby number
-of quickly approaches $0$. Gradients of surface density are seen to
-spike up in wake-like features around the vortex during dissipation. 
-{\bf show
-  this. e.g. vortex before, during and after death. maybe show
-  snapshots of surface density perturbation (vortex) compared with
-  $\nabla\Sigma/\Sigma$ (vortex shocks). demonstrate vortex shape
-  changes and shocks develop prior ot death. state whether this is
-  generic for all coolings}.  
-In addition, we also observed large increases in the Mach number near
-the vortex during dissipation, which indicate susceptibility to
-shocks. {\bf show mach number plot, or quote values. relevant mach
-  number is fluid flow speed with respect to 
-  vortex bulk motion around the disc. does the typical mach number
-  near the vortex just before its death depend on cooling?} 
-Thus, the sudden drop in vortex amplitude may  
-be a result of the development of associated spiral shocks dissipating
-its energy \citep[cf.][who 
-  suggested damping due to planet-induced shocks]{fu14}.
-
-{\bf note that fu et al suggested vortex damping by the planet
-  shocks. here, we are suggesting vortex dissipates energy by shocking
-  the disk. our evidence is that we observe vortex growth before it
-  dissipates, and spiral shocks associated with the vortex always
-  develop before vortex death. once it starts shocking the disc it
-  loses energy. if the planet-induced shocks were responsible, death
-  should be more gradual. can we think of other evidences that it's
-  not the planet shock that's responsible?                    
-}
-
-Vortex growth during the quasi-steady state also acts to smooth out
-the gap edge. 
-%Of note is also that the large overdensity the vortex
-%creates during it's quasi-steady state distorts the background
-%disk. 
-{\bf show this (smoothening of gap edge due to vortex growth),
-  e.g. general vortensity evolution or surface density 
-  evolution}. 
-This is because the vortex has an associated viscosity {\bf quote a
-  typical value of alpha viscosity associated with vortex}. This acts
-against gap-opening by the planet, so that the condition for the RWI
-becomes less favourable as the vortex grows. {\bf self-limiting
-  behaviour?}
-
-\subsection{Vortex lifetimes as a function of cooling rate}
-
-{\bf describe dependency on
-cooling. Do we have enough data to have a `lifetime v.s. beta'
-  plot? here we want to explain why vortex lives longer with longer
-  cooling, but very long cooling results in shorter lifetimes. 
-  Consider the following quantities as a function of cooling:
-  initial strength of the single vortex; growth rate of single vortex
-  strength from its formation to just before death; ampiltude of
-  vortex just before death; disc temperature in the vortex region.
-
-  we expect the initial $m=1$ vortex (after merging) to be weaker with
-  increasing cooling time because the instability is weaker. (see
-  meheut et al 2013, mnras, 430, 1988 who show that the initial vortex
-  amplitude after linear growth is correlated with the linear growth
-  rate).
-
-  suppose the vortex shocks when it reaches a certain amplitude (which
-  probably increases with longer cooling rate because it's more
-  difficult to shock a hotter disc). as we
-  increase the cooling time, the initial $m=1$ vortex is weaker, it
-  also grows more slowly because gap-opening is opposed by a hotter
-  disc. the above implies that it takes longer for the vortex reach
-  the shock-inducing amplitude and dissipate.
-
-  but the above is not consistent with non-monotonic behaviour, which
-  predicts the $\tilde{\beta}=10$ vortex to survive longest. this is
-  because we expect the initial vortex to be weakest (check) which
-  means it should take the longest to grow and shock. check
-  whether there's anything special about this case. one possibility is
-  that vortex growth during the quasi-steady state is actually faster
-  because of the raised temperature (li et al 2000 show increased
-  temperature favours RWI - possibiliy evidenced here by a stronger
-  vortex before death than other cases)
-
-  perhaps have a case of $\tilde{\beta}=7.5$ as further evidence of
-  non-monotonic behaviour
-}
-
-Fig. \ref{lifetimeplot} shows that increasing $\tilde{\beta}$
-increases the lifetime of the vortex up to a critical 
-value $\tilde{\beta}=5.0$ for which the vortex lasted beyond the
-simulation time of $2000P_0$. This is consistent with vortex
-dissipation when it induce shocks. Our `planet-off' simulations
-indicate weaker vortices are formed initially with increased cooling
-times. After merging, the growth of the single vortex, mediated by disc-planet
-interaction, is expected to be slower with increased cooling time
-because gap-opening becomes more difficult in a hotter
-disc. Furthermore, the increased sound-speed suggest the vortex should
-reach larger amplitudes in order to induce shocks. 
-These expectations
-imply that it takes longer for the vortex to induce shocks and
-dissipate, and hence longer lifetimes with increasing
-$\tilde{\beta}$.
-
-However, Fig. \ref{lifetimeplot} shows that increasing the cooling
-time further to $\tilde{\beta}=10$ actually results in a
-\emph{shorter} vortex lifetime. Notice for $\tilde{\beta}=10$ the vortex growth 
-during the quasi-steady state is faster than for
-$\tilde{\beta}=5$. This may be attributed to the higher temperature
-reached in the former case ($h\simeq 0.08$). {\bf temperature for
-  $\tilde{\beta}=5$?}, since \cite{li00} showed that the RWI grows
-faster with increased disc temperature. At long cooling times, this
-effect may dominate over the weaker gap-opening effect at higher $h$.
-
-Thus, unlike our `planet-off' simulations we find that there is
-a cooling time $t_c$ which maximises the vortex lifetime. In the long term
- 'planet-off' simulations we find that the vortices dissipate very gradually and visciously. {\bf to make
-this comparison we need to discuss vortex lifetimes for planet-off
-simulations as well} 
-This
-non-monotonic behaviour was also reported by \cite{fu14}, who
-performed isothermal disc-planet simluations with different fixed
-aspectratios.   
-%Cooling rates with
-%$\tilde\beta>5.0$ showed lifetimes that were well below the max
-%lifetime. 
-By analyzing the region
-around the vortex {\bf define this region} we find the
-$\tilde\beta=5.0$ case, with the longest vortex lifetime, has
-$h\approx0.06$. This value of $h$ was also found to maximize vortex
-lifetimes in the isothermal simulations of \cite{fu14}.
-
-%be the optimum lifetime for long-term simulations of
-%isothermal disks with varying intial temperature profiles $h$.
-
-%Competing effects of gap stability with $t_{\mathrm{coo}l}$ result in
-%the the non-montonic form of the lifetime. As the $\tilde\beta$
-%increases so does the stability of the gap edge as seen in
-%section~\ref{linear}, however as $\tilde\beta$ increases so does the
-%$c_{\mathrm{iso}}$ which is known from theory to effect the growth of
-%instability \citep{li00}. Thus increasing $\tilde\beta$ increases the
-%sound speed $c_{\mathrm{iso}}$ and likewise the gap instability up to
-%a certain point in which the effects of the more stable gap edge
-%becomes a significant effect.
-
-%The correlation of growth rate and sound
-%speed is non-negligible as in previous `planet-off' case as the disk
-%now can reach values of $h\simeq0.08$ for very slow cooling and the
-%importance of the effect can be imagined to accumulate over the now
-%considerably longer timescale. These effects dictate when then the
-%vortex reaches the critical value in which it shocks the system that
-%subsequently causes disspation and hence the lifetime fo the vortex.
+\subsection{PLUTO 3D}

### 2014-10-22 paper/model.tex
minor update

@@
 \end{align}
-where the sound-speed is given by $c_s=H\Omega_k$, where $H=hR$ is the
+where the sound-speed is given by $c_s=H\Omega_k$, $H=hR$ is the
 disc scale-height with constant aspect-ratio $h$ chosen below, and

### 2014-10-22 paper/paper.tex
minor update

@@
   accretion. It has been suggested that mass built-up in a dead zone
-  may lead to gravitational instability. However, the type of
-  gravitational instability relevant in this context has not been
+  may lead to gravitational instability. However, what type of
+  gravitational instability is relevant in this context has not been
   clarified. In this work, we use numerical simulations to explore
   gravitational instability in astrophysical discs in which only a 
-  sub-portion of it is self-gravitating, taken as a toy model for
-  protoplanetary disc containing a dead zone. We conduct numerical hydrodynamic
+  sub-portion of it is self-gravitating, taken as a toy model of a 
+  protoplanetary disc containing a massive dead zone. We conduct
+  numerical hydrodynamic 
   experiments in 2D using the FARGO code and in 3D with the ZEUS-MP
@@
   outside the dead zone, resulting in a nearly stationary one-arm spiral
-  pattern in the dead zone. Our simulations demonstrate the potential
-  importance global gravitational interaction between a
+  pattern in the dead zone. Our simulations demonstrate the 
+  importance of global gravitational interaction between a 
   self-gravitating dead zone and the exterior, non-self-gravitating

### 2014-10-22 paper/results.tex
minor update

@@
 \end{align}
-where $\delta\in[-10^{-3},10^{-3}]$ is chosen randomly but does not
-depend on $\phi$, and 
+where $\delta$ is an amplitude and 
 \begin{align}
@@
 We consider a disc size $[R_\mathrm{min}, R_\mathrm{max}] =
-[0.4,10]R_0$ and resolution $N_R\times N_\phi = 512\times 1024$ or
-about $8$ grids per $H$. We adopt $\epsilon_g=10^{-4}H$ for the 
+[0.4,10]R_0$. This gives a total disc mass $M_{d}=0.086M_*$, and the
+mass within $R\in[R_{d1},R_{d2}]$ is $0.049M_*$. 
+We use a resolution of $N_R\times N_\phi = 512\times 1024$ or
+about $8$ grids per $H$, and adopt $\epsilon_g=10^{-4}H$ for the 
 self-gravity softening length\footnote{In 2D self-gravity, $\epsilon_g$ also
@@
   $\epsilon_g$ in favour of angular momentum conservation, keeping in
-  mind that the strength of self-gravity will be over-estimated.}, and 
-set $M=10$ for random perturbations (Eq. \ref{randpert}).
+  mind that the strength of self-gravity will be over-estimated.}.
+Perturbations are set initially with $M=10$ and
+$\delta\in[-10^{-3},10^{-3}]$ chosen randomly but does not depend
+on $\phi$.
 
-Snapshots from the disc evolution is shown in Fig. \ref{fargo_2d} in
-terms of the non-axisymmetric surface density.
+%set $M=10$ for random perturbations (Eq. \ref{randpert}).
+
+Snapshots from the disc evolution are shown in Fig. \ref{fargo_2d} in
+terms of the non-axisymmetric surface density. At early times
+$t\lesssim100P_0$ the disc is dominated by low-amplitude high-$m$
+perturbations centred in the dead zone. Eventually a $m=1$ spiral 
+grows and dominates inside the dead zone.
+
+Fig. \ref{fargo_2d_angmom} compares the disc angular momenta
+evolution, most of which is contained in the $m=0$ and $m=1$ modes. 
+The $m=1$ spiral has an associated negative angular
+momentum. Its growth is correlated with an increase in the axisymmetric
+component of angular momentum, such that $\Delta J_0 \sim - \Delta
+J_1$. Note that FARGO does not explicitly conserve angular
+momentum. In addition, there may be an angular momentum flux due to
+self-gravity across disc boundaries since our domain size is finite. 
+Nevertheless, we find the total angular momentum is conserved to 
+$\Delta J = O(10^{-7}J)$, and is much smaller in magnitude than the change in the
+angular momenta components, $|\Delta J_m|$. 
+Fig. \ref{fargo_2d_angmom} then indicates a transfer of 
+angular momentum from the $m=1$ spiral to the background disc.
 
@@
 \subsection{ZEUS-MP 3D}
+The 3D disc has radial size 
+$[r_\mathrm{min},r_\mathrm{max}]=[0.4,10]R_0$ and vertical extent
+$n_H=2$ scale-heights. The resolution is $N_r\times N_\theta\times
+N_\phi=256\times32\times256$. Because of the much reduced resolution
+compared to 2D, we use a smooth perturbation by setting
+$\delta = 10^{-3}$ and $M=1$. This corresponds to a single $m=1$
+spiral in the dead zone, which was found to dominate the 2D
+simulation.
+
+Our 3D discs are initialised in approximate equilibrium only, so we
+first evolve the disc without perturbations using  
+$(\lmax,\mmax)=(32,0)$ up to $t=10P_0$, during which 
+meridional velocities are damped out. We then restart the simulation
+with the above perturbation and $(\lmax,\mmax)=(32,32)$.
 

### 2014-10-23 paper/paper.tex
minor updates

@@
 
-\title[One-arm spirals in self-gravitating discs]{One-arm spirals in
+\title[Spirals in self-gravitating dead zones]{Global spirals in
   self-gravitating dead zones in protoplanetary discs}
@@
   sub-portion of it is self-gravitating, taken as a toy model of a 
-  protoplanetary disc containing a massive dead zone. We conduct
+  protoplanetary disc containing a finite-size massive dead zone. We conduct
   numerical hydrodynamic 
@@
   mode with azimuthal wavenumber $m=1$ inside the dead
-  zone. Surprisingly, the co-rotation radius of the spiral is well
+  zone, with co-rotation radius well
   outside the dead zone, resulting in a nearly stationary one-arm spiral
@@
   self-gravitating dead zone and the exterior, non-self-gravitating
-  parts of the disc.  
+  parts of the disc. This non-local effect, due to disc structure, 
+  exists even when the total disc mass is below the threshold for
+  global effects found in
+  previous studies. 
 \end{abstract}

### 2014-10-23 paper/results.tex
minor updates

@@
 Here we describe one numerical simulation from each code to give a
-qualitative picture of disc evolution. In the next section, we analyse
-2D simulations in more detail and explore parameter space.
+qualitative picture of disc evolution. 
+%In the next section, we analyse
+%2D simulations in more detail and explore parameter space with
+%improved resolution runs.
 
@@
 (R_{d2}-R_{d1})/2$. 
-\subsection{FARGO 2D}\label{fargo_fiducial}
+\subsection{2D run}\label{fargo_fiducial}
 We consider a disc size $[R_\mathrm{min}, R_\mathrm{max}] =
@@
 mass within $R\in[R_{d1},R_{d2}]$ is $0.049M_*$. 
-We use a resolution of $N_R\times N_\phi = 512\times 1024$ or
-about $8$ grids per $H$, and adopt $\epsilon_g=10^{-4}H$ for the 
+We use a resolution of $N_R\times N_\phi = 1024\times 2048$ or
+about $16$ grids per $H$, and adopt $\epsilon_g=10^{-4}H$ for the 
 self-gravity softening length\footnote{In 2D self-gravity, $\epsilon_g$ also
@@
 Nevertheless, we find the total angular momentum is conserved to 
-$\Delta J = O(10^{-7}J)$, and is much smaller in magnitude than the change in the
-angular momenta components, $|\Delta J_m|$. 
+$|\Delta J/J|= O(10^{-6})$, and is much smaller in magnitude than the change in the
+angular momenta components, $|\Delta J_{0,1}/J|> O(10^{-5})$. 
 Fig. \ref{fargo_2d_angmom} then indicates a transfer of 
@@
   0cm 0cm 0cm]{figures/polarxy_dens125} 
-  \caption{Fiducial FARGO 2D simulation (\S\ref{fargo_fiducial}),
+  \caption{Preliminary FARGO 2D simulation (\S\ref{fargo_fiducial}),
     showing the evolution of the non-axisymmetric surface density
@@
   \caption{Evolution of angular momentum components in the FARGO
-    fiducial simulation (Fig. \ref{fargo_2d}). The perturbation
+     simulation show in Fig. \ref{fargo_2d}. The perturbation
     relative to $t=0$ is shown in units of the initial total angular
@@
 
-\subsection{ZEUS-MP 3D}
+\subsection{3D runs}
 The 3D disc has radial size 
@@
 
-\subsection{PLUTO 3D}
+Snapshots at the end of the simulations are shown in
+Fig. \ref{3d_prelim} for both ZEUS-MP and PLUTO. We again
+find the dead zone eventually dominated by an $m=1$ spiral. Note that
+the amplitude of the spiral is smaller than the 2D case above,
+possibily due to lower resolution and/or 3D self-gravity being weaker
+than that in 2D.
+
+\begin{figure}
+  \begin{center}
+    \subfigure[ZEUS-MP]{
+      \includegraphics[scale=0.33]{figures/polarxy_dens015_zeus}
+    }
+    \subfigure[PLUTO]{
+      \includegraphics[scale=0.33]{figures/polarxy_dens015_zeus}
+    }
+  \end{center}
+  \caption{Preliminary 3D simulations using the (a) ZEUS-MP and (b)
+    PLUTO. The non-axisymmetric midplane density at the end of the run
+    is shown. Here $\psi \equiv \pi/2 - \theta$ is the angular height
+    from the midplane.\label{3d_prelim}}   
+\end{figure}
+
+Fig. \ref{3d_prelim_angmom} shows the angular momentum evolution in
+the 3D runs. ZEUS-MP does not conserve angular momentum very well, but
+the variation $|\Delta J/J|< O(10^{-5})$ is not significant compared
+to the individual components $|\Delta J_{0,1}/J|\sim
+6\times10^{-5}$. We found high-$m$ structures to develop near the
+inner boundary, which may have contributed to this error. It
+is clear, however, that the growth of negative angluar momentum in
+$m=1$ is accompanied by an increase in the background angular
+momentum.
+
+\begin{figure}
+  \includegraphics[scale=0.41,clip=true,trim=0cm 1.75cm 0cm 0cm]{figures/nonaxi_evol_ang_zeus}
+  \includegraphics[scale=0.41]{figures/nonaxi_evol_ang_zeus}
+  \caption{Evolution of angular momentum components in the 3D 
+    simulations show in Fig. \ref{3d_prelim}. The perturbation
+    relative to $t=10P_0$ is shown in units of the initial total
+    angular momentum $J_\mathrm{ref}$.\label{3d_prelim_angmom}} 
+\end{figure}
+
+\section{Global one-arm spirals in structured discs}
+
+\subsection{Properties of the $m=1$ dead zone spiral}
+
+%\subsection{Three-dimensional structure}
+
+%\section{Additional simulations}
+%survery

### 2014-10-23 paper/results.tex
new figs

@@
     \subfigure[PLUTO]{
-      \includegraphics[scale=0.33]{figures/polarxy_dens015_zeus}
+      \includegraphics[scale=0.33]{figures/pdisk_020}
     }
@@
 \begin{figure}
-  \includegraphics[scale=0.41,clip=true,trim=0cm 1.75cm 0cm 0cm]{figures/nonaxi_evol_ang_zeus}
-  \includegraphics[scale=0.41]{figures/nonaxi_evol_ang_zeus}
+%scale=0.41
+  \includegraphics[scale=.41,clip=true,trim=0cm 1cm 0cm 0cm]{figures/nonaxi_evol_ang_zeus}
+  \includegraphics[scale=.41]{figures/nonaxi_evol_ang_pluto}
   \caption{Evolution of angular momentum components in the 3D 
@@
     angular momentum $J_\mathrm{ref}$.\label{3d_prelim_angmom}} 
+\end{figure}
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{figures/pdiskxy_020_m1}
+  \caption{Cartesian plot of the $m=1$ midplane density in the PLUTO
+    simulation shown in Fig. \ref{3d_prelim}.\label{pluto_cart}}   
 \end{figure}

### 2014-10-30 paper/model.tex
minor updates
update

@@
 $Q_\mathrm{out}=2$. An example of the initial surface density and the
-Toomre parameter (see\ref{wkb}) is shown in Fig. \ref{initial_surf}. 
+Toomre parameter (see\S\ref{wkb}) is shown in Fig. \ref{initial_surf}. 
 Initially there is no vertical or radial velocity
@@
 
-The non-axisymmetric surface density can be visualised with
-\begin{align}
-  \Delta \Sigma = \frac{\Sigma - \Sigma_{0}}{\Sigma_{0}}, 
-\end{align}
-while an individual mode as
+% The non-axisymmetric surface density can be visualised with
+% \begin{align}
+%   \Delta \Sigma = \frac{\Sigma - \Sigma_{0}}{\Sigma_{0}}, 
+% \end{align}
+% while an individual mode as
+% \begin{align}
+%   \Delta\Sigma_m = \frac{2}{\Sigma_{00}} \real\left[\Sigma_m \exp{(\ii
+%       m\phi)}\right]
+% 	\end{align}
+% where $\Sigma_{00} = \Sigma_0(t=0)$.
+
+% The global mode amplitude is 
+% \begin{align}
+%   A_m = \frac{|C_m|}{M_{d0}}
+% \end{align}
+% where $M_{d0}$ is the initial disc mass and 
+% \begin{align}
+%   C_m = 2\pi\int_{R_1}^{R_2} \Sigma_m RdR, 
+% \end{align}
+% with integration limits chosen later.
+
+The time evolution of the $m^\mathrm{th}$ mode can be characterized by
+assuming $\Sigma_m\propto\exp{(\ii \sigma t)}$, where
+$\sigma = \omega - \ii\gamma$ is the complex frequency, $\omega$ is
+the real frequency and $\gamma$ is the growth rate. In linear theory
+$\sigma$ is a constant eigenvalue. 
+To estimate $\sigma$ from simulation data, we first construct the real
+surface density 
 \begin{align}
@@
 \end{align}
-where $\Sigma_{00} = \Sigma_0(t=0)$.
-
-The global mode amplitude is 
+where $\Sigma_{00} = \Sigma_0(t=0)$. We then locate the maximum,
 \begin{align}
-  A_m = \frac{|C_m|}{M_{d0}}
+  \mathrm{max}(\Delta\Sigma_m) \equiv
+  \Delta\Sigma_m(R_\mathrm{peak},\phi_\mathrm{peak}). 
 \end{align}
-where $M_{d0}$ is the initial disc mass and 
-\begin{align}
-  C_m = 2\pi\int_{R_1}^{R_2} \Sigma_m RdR, 
-\end{align}
-with integration limits chosen later.
-
-The time evolution of the $m^\mathrm{th}$ mode can be characterized by
-assuming 
-$\Sigma_m\propto\exp{(\ii \sigma t)}$, where
-$\sigma = \omega - \ii\gamma$ is the complex frequency, $\omega$ is
-the real frequency and $\gamma$ is the growth rate.
 We estimate the growth rate as
 \begin{align}
-  \gamma = \frac{d\ln{A_m}}{dt},
+  \gamma = \frac{d}{dt}\ln{\left[\mathrm{max}(\Delta\Sigma_m)\right]}, 
 \end{align}
@@
 \begin{align}
-  \Omega_p = \frac{d\phi_\mathrm{max}}{dt},
+  \Omega_p = \frac{d\phi_\mathrm{peak}}{dt}. 
 \end{align}
-where $\phi_\mathrm{max}$ is the azimuthal co-ordinate of
-$\mathrm{max}(\Delta\Sigma_m)$. 
+The total non-axisymmetric surface density can be visualised with
+\begin{align}
+  \Delta\Sigma = \frac{\Sigma - \Sigma_0}{\Sigma_0}. 
+\end{align}
+
 %\begin{align}

### 2014-10-30 paper/results.tex
update

--- 2014-10-22 paper/results.tex
@@
-\section{Preliminary experiments}
-Here we describe one numerical simulation from each code to give a
-qualitative picture of disc evolution. In the next section, we analyse
-2D simulations in more detail and explore parameter space.
+\section{Results for two-dimensional discs}
+We first present results from FARGO simulations. The 2D disc spans
+$[R_\mathrm{min}, R_\mathrm{max}] = [0.4,10]R_0$. This gives a total
+disc mass $M_{d}=0.086M_*$. The mass within
+$R\in[R_\mathrm{min},R_{d1}]$ is $0.017M_*$, that within
+$R\in[R_{d1},R_{d2}]$ is $0.049M_*$, and that within
+$R\in[R_{d2},R_\mathrm{max}]$ is $0.021M_*$. We use a resolution of
+$N_R\times N_\phi = 1024\times 2048$, or about $16$ grids per $H$, and
+adopt $\epsilon_g=10^{-4}H$ for the   
+self-gravity softening length\footnote{In 2D self-gravity, $\epsilon_g$ also
+  approximates for the vertical disc thickness, so a more appropriate
+  value would be $\epsilon_g\sim H$ \citep{muller12}. However, because
+  $\epsilon_g\propto R$ is needed in FARGO, the Poisson kernel
+  (Eq. \ref{2d_grav}) is no longer symmetric in $(R,R^\prime)$. We
+  choose a small  
+  $\epsilon_g$ in favour of angular momentum conservation, keeping in
+  mind that the strength of self-gravity will be over-estimated.}.
 
-In these simulations we subject the disc to random perturbations in
-cylindrical velocity,
+\subsection{Reference run with random perturbations}
+In this fiducial simulation we subject the disc to initial perturbations in
+cylindrical radial velocity, 
 \begin{align}\label{randpert}
@@
 where $\overline{R}_d = (R_{d1}+R_{d2})/2$ and $\Delta R_d =
-(R_{d2}-R_{d1})/2$. 
-\subsection{FARGO 2D}\label{fargo_fiducial}
-We consider a disc size $[R_\mathrm{min}, R_\mathrm{max}] =
-[0.4,10]R_0$. This gives a total disc mass $M_{d}=0.086M_*$, and the
-mass within $R\in[R_{d1},R_{d2}]$ is $0.049M_*$. 
-We use a resolution of $N_R\times N_\phi = 512\times 1024$ or
-about $8$ grids per $H$, and adopt $\epsilon_g=10^{-4}H$ for the 
-self-gravity softening length\footnote{In 2D self-gravity, $\epsilon_g$ also
-  approximates for the vertical disc thickness, so a more appropriate
-  value would be $\epsilon_g\sim H$ \citep{muller12}. However, because
-  $\epsilon_g\propto R$ is needed in FARGO, the Poisson kernel
-  (Eq. \ref{2d_grav}) is no longer symmetric in $(R,R^\prime)$. We
-  choose a small  
-  $\epsilon_g$ in favour of angular momentum conservation, keeping in
-  mind that the strength of self-gravity will be over-estimated.}.
-Perturbations are set initially with $M=10$ and
-$\delta\in[-10^{-3},10^{-3}]$ chosen randomly but does not depend
-on $\phi$.
+(R_{d2}-R_{d1})/2$. We use $M=10$ and choose
+$\delta\in[-10^{-3},10^{-3}]$ randomly but independent of $\phi$.
 
-%set $M=10$ for random perturbations (Eq. \ref{randpert}).
+Fig. \ref{fargo_modeamp} plots evolution of the maximum
+non-axisymmetric surface density amplitudes in $R\in[R_{d1},R_{d2}]$
+for $m\in[1,10]$. Snapshots from the simulation are shown in
+Fig. \ref{fargo_2d}. 
+At early times $t\lesssim100P_0$ the dead zone is
+dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
+growth initially and saturate (or decays) after $t=40P_0$. Notice the
+low $m\leq 2$ modes decay initially, but grows between $t\in[20,40]P_0$,
+possibily due to saturation of the high-$m$ modes  
+\citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
+again after $t=70P_0$, and eventually dominates the dead zone.
 
-Snapshots from the disc evolution are shown in Fig. \ref{fargo_2d} in
-terms of the non-axisymmetric surface density. At early times
-$t\lesssim100P_0$ the disc is dominated by low-amplitude high-$m$
-perturbations centred in the dead zone. Eventually a $m=1$ spiral 
-grows and dominates inside the dead zone.
-
-Fig. \ref{fargo_2d_angmom} compares the disc angular momenta
-evolution, most of which is contained in the $m=0$ and $m=1$ modes. 
-The $m=1$ spiral has an associated negative angular
-momentum. Its growth is correlated with an increase in the axisymmetric
-component of angular momentum, such that $\Delta J_0 \sim - \Delta
-J_1$. Note that FARGO does not explicitly conserve angular
-momentum. In addition, there may be an angular momentum flux due to
-self-gravity across disc boundaries since our domain size is finite. 
-Nevertheless, we find the total angular momentum is conserved to 
-$\Delta J = O(10^{-7}J)$, and is much smaller in magnitude than the change in the
-angular momenta components, $|\Delta J_m|$. 
-Fig. \ref{fargo_2d_angmom} then indicates a transfer of 
-angular momentum from the $m=1$ spiral to the background disc.
+\begin{figure}
+  \includegraphics[width=\linewidth]{figures/nonaxi_evol_DZ_fargo}
+  \caption{Evolution of maxima in non-axisymmetric surface density 
+    in the fiducial FARGO simulation.\label{fargo_modeamp}} 
+\end{figure}
 
@@
   0cm 0cm
-  0cm]{figures/polarxy_dens100}\includegraphics[scale=0.27,clip=true,trim=2.26cm
-  0cm 0cm 0cm]{figures/polarxy_dens125} 
-  \caption{Fiducial FARGO 2D simulation (\S\ref{fargo_fiducial}),
-    showing the evolution of the non-axisymmetric surface density
-    $\Delta\Sigma$. Note that the disc extends to $R=10R_0$, but only
-    the inner part is shown for clarity. \label{fargo_2d}} 
+  0cm]{figures/polarxy_dens110}\includegraphics[scale=0.27,clip=true,trim=2.26cm
+  0cm 0cm 0cm]{figures/polarxy_dens130} 
+  \caption{Visualisation of the FARGO 2D simulation. The total  
+    non-axisymmetric surface density
+    $\Delta\Sigma$ is shown. \label{fargo_2d}} 
 \end{figure}
 
+Fig. \ref{2d_angmom} shows the disc angular momenta
+evolution. Only the $m=0,\,1$ components are 
+plotted since they are dominant. The $m=1$ structure has
+an associated negative angular momentum.   
+Its growth is compensated by an increase in the axisymmetric
+component of angular momentum, such that $\Delta J_0 + \Delta
+J_1 \sim 0$. Note that FARGO does not conserve angular momentum
+exactly. However, we find the total angular momentum varies by 
+$|\Delta J/J|= O(10^{-6})$, and is much smaller in magnitude than the
+change in the angular momenta components, $|\Delta J_{0,1}/J|>
+O(10^{-5})$. Fig. \ref{2d_angmom} then suggest that angular momentum
+is transferred from the one-arm spiral to the background disc.
+
 \begin{figure}
-  \includegraphics[width=\linewidth]{figures/nonaxi_evol_ang_fargo}
-  \caption{Evolution of angular momentum components in the FARGO
-    fiducial simulation (Fig. \ref{fargo_2d}). The perturbation
-    relative to $t=0$ is shown in units of the initial total angular
-    momentum $J_\mathrm{ref}$.\label{fargo_2d_angmom}} 
+  % scale=0.41
+  \includegraphics[scale=.41]{figures/nonaxi_evol_ang_fargo}
+  \caption{Evolution of angular momentum components in the fiducial
+    FARGO simulation. The perturbation
+    relative to $t=0$ in 2D is shown in units of the
+    initial total angular momentum $J_\mathrm{ref}$.\label{2d_angmom}} 
 \end{figure}
 
-\subsection{ZEUS-MP 3D}
-The 3D disc has radial size 
-$[r_\mathrm{min},r_\mathrm{max}]=[0.4,10]R_0$ and vertical extent
-$n_H=2$ scale-heights. The resolution is $N_r\times N_\theta\times
-N_\phi=256\times32\times256$. Because of the much reduced resolution
-compared to 2D, we use a smooth perturbation by setting
-$\delta = 10^{-3}$ and $M=1$. This corresponds to a single $m=1$
-spiral in the dead zone, which was found to dominate the 2D
-simulation.
+\subsection{Non-linear initial perturbation}
+In order to focus on the one-arm spiral that emerges in the above
+simulation, we here analyse a run with only an initial $m=1$
+perturbation in the dead zone. Specifically, we set
+\begin{align}
+  &\Sigma \to \Sigma_0(R) + \delta\Sigma_0(R) + 2\Sigma_1(R)\cos{\phi},\\
+  &v_\phi  \to v_{\phi 0}(R) + 2v_{\phi 1}(R)\cos{\phi},
+\end{align}
+for $R\in[R_{d1},R_{d2}]$, where
+\begin{align}
+  &\delta\Sigma_0 =
+  -\delta^2\times\Sigma_0(R_{d1})\frac{R_{d1}}{R}\sin^3{\left[\frac{2\pi\left(R-R_{d1}\right)}{R_{d2}-R_{d1}}\right]},\\ 
+  &\Sigma_1 =
+  \delta\times\Sigma_0\sin{\left[\frac{\pi\left(R-R_{d1}\right)}{R_{d2}-R_{d1}}\right]},  
+\end{align}
+with $\delta = 10^{-3}$, and
+\begin{align}
+  v_{\phi 1} = -\frac{v_{\phi 0}\delta\Sigma_0}{2\Sigma_1}. 
+\end{align}
+This perturbs the dead zone structure while conserving total mass and
+angular momentum. The $m=1$ perturbation has an associated negative
+angular momentum (as observed in the previous simulation) at the
+expense of increasing the angular momentum associated with the
+background disc. %local exchange for simplicity 
+%actually doesn't matter if we set plus or minus
+Note that the perturbation to the axisymmetric surface density is
+$O(\delta^2)$, whereas that in $m=1$ is $O(\delta)$. Since $Q>1$
+everywhere in the disc, there is no risk of axisymmetric gravitational
+instability.
 
-Our 3D discs are initialised in approximate equilibrium only, so we
-first evolve the disc without perturbations using  
-$(\lmax,\mmax)=(32,0)$ up to $t=10P_0$, during which 
-meridional velocities are damped out. We then restart the simulation
-with the above perturbation and $(\lmax,\mmax)=(32,32)$.
+While the above form of initial perturbations is artificial, we find 
+the spiral that forms have similar properties (see below) to that found in the
+previous simulation. However, the above perturbations have the
+advantage of producing cleaner results without complications from
+higher $m$ modes, and the required simulation time for the one-arm
+spiral to emerge is significantly reduced.
 
-\subsection{PLUTO 3D}
+Fig. shows the evolution of angular momenta.

### 2014-10-30 paper/results_3d.tex
update

\section{Three-dimensional simulations}
The 3D disc has radial size 
$[r_\mathrm{min},r_\mathrm{max}]=[0.4,10]R_0$ and vertical extent
$n_H=2$ scale-heights. The resolution is $N_r\times N_\theta\times
N_\phi=256\times32\times256$. Because of the much reduced resolution
compared to 2D, we use a smooth perturbation by setting
$\delta = 10^{-3}$ and $M=1$. This corresponds to a single $m=1$
spiral in the dead zone, which was found to dominate the 2D
simulation.

Our 3D discs are initialised in approximate equilibrium only, so we
first evolve the disc without perturbations using  
$(\lmax,\mmax)=(32,0)$ up to $t=10P_0$, during which 
meridional velocities are damped out. We then restart the simulation
with the above perturbation and $(\lmax,\mmax)=(32,32)$.

Snapshots at the end of the simulations are shown in
Fig. \ref{3d_prelim} for both ZEUS-MP and PLUTO. We again find the dead zone eventually
dominated by an $m=1$ spiral. A Cartesian visualisation of the
midplane density perturbation associated with the $m=1$ mode is shown in Fig. \ref{pluto_cart}.
The two codes give qualitatively similar results, but a longer PLUTO simulation was
needed to achieve similar mode amplitudes than ZEUS-MP. 
The amplitude of the spiral is smaller than the 2D case above,
possibily due to lower resolution and/or 3D self-gravity being weaker
than that in 2D.

\begin{figure}
  \begin{center}
    \subfigure[ZEUS-MP]{
      \includegraphics[scale=0.33]{figures/polarxy_dens015_zeus}
    }
    \subfigure[PLUTO]{
      \includegraphics[scale=0.33]{figures/pdisk_020}
    }
  \end{center}
  \caption{Preliminary 3D simulations using the (a) ZEUS-MP and (b)
    PLUTO. The non-axisymmetric midplane density at the end of the run
    is shown. Here $\psi \equiv \pi/2 - \theta$ is the angular height
    from the midplane.\label{3d_prelim}}   
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/pdiskxy_020_m1}
  \caption{Cartesian visualisation the $m=1$ midplane density in the PLUTO
    simulation shown in Fig. \ref{3d_prelim}.\label{pluto_cart}}   
\end{figure}

%\subsection{Angular momentum evolution}

Fig. \ref{3d_angmom} shows the angular momentum evolution in the 3D
runs. ZEUS-MP does not conserve angular momentum very well, but
the variation $|\Delta J/J|< O(10^{-5})$ is not significant compared
to the individual components $|\Delta J_{0,1}/J|\sim 
6\times10^{-5}$. The PLUTO run reaches similar values of
$|J_{0,1}|$, but achieves better conervation, with $|\Delta
J/J|=O(10^{-8})$.

%Note that FARGO does not explicitly conserve angular
%momentum. In addition, there may be an angular momentum flux due to
%self-gravity across disc boundaries since our domain size is finite. 
%Fig. \ref{prelim_angmom} shows the angular momentum evolution in
%the 3D runs.

\begin{figure}
%scale=0.41
  \includegraphics[scale=.41,clip=true,trim=0cm 1cm 0cm 0cm]{figures/nonaxi_evol_ang_zeus}
  \includegraphics[scale=.41]{figures/nonaxi_evol_ang_pluto}
  \caption{Evolution of angular momentum components in the fiducial 3D 
    simulations. The perturbation
    relative to $t=100P_0$ is shown in units of the
    initial total angular momentum $J_\mathrm{ref}$.\label{3d_angmom}} 
\end{figure}

% \section{One-arm spirals in structured discs}
% Here we examine a 2D FARGO simulation in more detail. The setup is the
% same as that in \S\ref{fargo_fiducial}, except we use a resolution of
% $N_R\times N_\phi = 2048\times4096$ and initialise the disc with a
% single $m=1$ perturbation.

% \subsection{Properties of the $m=1$ dead zone spiral}
% % growth rate, co-rotation radius, pattern speed, propagation diagram

### 2014-10-30 paper/results.tex
updated with fargo exchange simulations

@@
   % scale=0.41
-  \includegraphics[scale=.41]{figures/nonaxi_evol_ang_fargo}
+  \includegraphics[width=\linewidth]{figures/nonaxi_evol_ang_fargo}
   \caption{Evolution of angular momentum components in the fiducial
@@
 
-\subsection{Non-linear initial perturbation}
+\subsection{Non-linear initial perturbation}\label{fargo_nonlin}
 In order to focus on the one-arm spiral that emerges in the above
@@
 Fig. shows the evolution of angular momenta.
+
+\begin{figure*}
+  % scale=0.41
+  \includegraphics[scale=0.4,clip=true,trim=0cm 0cm 1.73cm
+  0cm]{figures/polarxy_fargo_ex_100}\includegraphics[scale=0.4,clip=true,trim=2.24cm
+  0cm 1.73cm 0cm]{figures/polarxy_fargo_ex_150}\includegraphics[scale=0.4,clip=true,trim=2.24cm
+  0cm 1.73cm 0cm]{figures/polarxy_fargo_ex_200}\includegraphics[scale=0.4,clip=true,trim=2.24cm
+  0cm 1.73cm 0cm]{figures/polarxy_fargo_ex_250}\includegraphics[scale=0.4,clip=true,trim=2.24cm
+  0cm 0.0cm 0cm]{figures/polarxy_fargo_ex_300}
+  \caption{Evolutoin of the $m=1$ surface density in the FARGO
+    simulation with the special $m=1$ initial perturbations
+    described in \S\ref{fargo_nonlin}.\label{2d_angmom_ex_evol}} 
+\end{figure*}
+
+\begin{figure}
+  % scale=0.41
+  \includegraphics[width=\linewidth]{figures/nonaxi_evol_ang_fargo_ex}
+  \caption{Evolution of angular momentum components in the 
+    FARGO simulation with the special $m=1$ initial perturbations
+    described in \S\ref{fargo_nonlin}. 
+    The perturbation
+    relative to $t=0$ in 2D is shown in units of the
+    initial total angular momentum $J_\mathrm{ref}$.\label{2d_angmom_ex}} 
+\end{figure}

### 2014-10-31 paper/model.tex
major change in paper - new idea - ang mom exchange with background because of temp gradient

--- 2014-10-22 paper/model.tex
@@
 
-The non-axisymmetric surface density can be visualised with
-\begin{align}
-  \Delta \Sigma = \frac{\Sigma - \Sigma_{0}}{\Sigma_{0}}, 
-\end{align}
-while an individual mode as
+The normalised surface density associated with $m$ is
 \begin{align}
@@
 \end{align}
-where $\Sigma_{00} = \Sigma_0(t=0)$.
+where $\Sigma_{00} = \Sigma_0(t=0)$. The time evolution of the
+$m^\mathrm{th}$ mode can be characterized by 
+assuming $\Sigma_m\propto\exp{(-\ii \sigma t)}$, where
+$\sigma = \omega + \ii\gamma$ is the complex frequency, $\omega$ is
+the real frequency and $\gamma$ is the growth rate. The total
+non-axisymmetric surface density is 
+\begin{align}
+  \Delta\Sigma = \frac{\Sigma - \Sigma_0}{\Sigma_0}. 
+\end{align}
 
-The global mode amplitude is 
-\begin{align}
-  A_m = \frac{|C_m|}{M_{d0}}
-\end{align}
-where $M_{d0}$ is the initial disc mass and 
-\begin{align}
-  C_m = 2\pi\int_{R_1}^{R_2} \Sigma_m RdR, 
-\end{align}
-with integration limits chosen later.
-
-The time evolution of the $m^\mathrm{th}$ mode can be characterized by
-assuming 
-$\Sigma_m\propto\exp{(\ii \sigma t)}$, where
-$\sigma = \omega - \ii\gamma$ is the complex frequency, $\omega$ is
-the real frequency and $\gamma$ is the growth rate.
-We estimate the growth rate as
-\begin{align}
-  \gamma = \frac{d\ln{A_m}}{dt},
-\end{align}
-and we estimate the real frequency by calculating the pattern speed
-$\Omega_p = \omega/m$,
-\begin{align}
-  \Omega_p = \frac{d\phi_\mathrm{max}}{dt},
-\end{align}
-where $\phi_\mathrm{max}$ is the azimuthal co-ordinate of
-$\mathrm{max}(\Delta\Sigma_m)$. 
-%\begin{align}
-%  \mathrm{max}\left\{2 \real\left[ \Sigma_m \exp{(\ii m
-%        \phi)}\right]\right\},
-
-%other methods possible too
-
-\subsection{Angular momentum decomposition}
+\subsubsection{Angular momentum decomposition}
 The total disc angular momentum in $R\in[R_1,R_2]$ is
@@
 \end{align}
-where $^*$ denotes complex conjugation, and we idenfity $J_m$ as the angular momentum 
-associated with the $m^\mathrm{th}$ mode, and $j_m$ the associated
-angular momentum per unit length.
+where $^*$ denotes complex conjugation, and we idenfity $J_m$ as the 
+angular momentum associated with disc structure with
+$m^\mathrm{th}$-fold symmetry, and $j_m$ the associated angular
+momentum per unit length. We will use $J_m$ to monitor angular momentum 
+conservation for the disc as a whole.
 
 \subsection{Linear density waves}\label{wkb}
-We will also use linear density wave theory to interpret
-simulation data. Here, we follow conventions in \cite{shu91} and quote
-results therein. The dispersion relation for tightly-wound density
-waves of the form 
-$\exp{[\ii(\omega t - m \phi + kR)]}$ in a razor-thin
-isothermal disc is 
+In addition to the above diagnostics for disc evolution, we
+will also use linear density wave theory to interpret simulation
+results.
+
+\subsubsection{Global angular momentum conservation for linear
+  perturbations} 
+It can be shown that linear perturbations with $\phi$-dependence in the form
+$\exp{\ii m\phi }$ satisfies angular momentum conservation
+in the form 
+\begin{align} 
+  \frac{\p \jlin}{\p t} + \nabla\cdot\bm{F} = \tbg, 
+\end{align}
+where
+\begin{align}
+  \jlin \equiv
+  -\frac{m\Sigma}{2}\imag\left(\bm{\xi}^*\cdot\frac{\p\bm{\xi}}{\p
+      t} + \Omega \hat{\bm{k}}\cdot\bm{\xi}\times\bm{\xi}^* + \ii
+    m \Omega |\bm{\xi}|^2\right) 
+\end{align}
+is the angular momentum density, $\bm{\xi}$ is the Lagragian
+displacement, and $\bm{F}$ is the angular momentum flux
+consisting of a Reynolds stress and a gravitational stress. Explicit
+expressions for $\bm{\xi}$ and $\bm{F}$ can be found in e.g.,
+\cite{papaloizou85} and \cite{lin11} respectively. In the above
+expressions, $\Sigma$ and $\Omega$ correspond to the unperturbed disc.
+
+The term $\tbg$ is 
+\begin{align}
+  \tbg \equiv
+  -\frac{m}{2}\imag\left(\Sigma_m\xi_R^\prime\frac{dc_s^2}{dR}\right), 
+\end{align}
+and arises because we have adopted a locally isothermal equation of
+state. In a barotropic fluid, such as a strictly isothermal disc,
+$\tbg$ vanishes and the angular momentum associated with
+the perturbation is conserved, provided boundary fluxes vanish.
+
+However, as noted in \cite{lin11}, if there is an imposed
+temperature gradient, as is the disc models considered in this paper,
+then $\tbg\neq 0$ in general, which corresponds to a torque
+exerted by the background disc on the perturbation.
+
+We remark that the perturbed angular momentum density $\jlin$ has been
+defined to satisfy a conservation law. This differs from the 
+empirical angular momentum decomposition above.
+
+%each m satisfies its own equation
+
+\subsubsection{Local results}
+\cite{shu91} gives the dispersion relation for tightly-wound density
+waves of the form $\exp{[\ii(\omega t - m \phi + kR)]}$ in a razor-thin
+isothermal disc as  
 \begin{align}\label{dispersion}
@@
 real $k$ are only possible when $Q^2(1-\nu^2)\leq1$.
-
-% We use Eq. \ref{wavenumber} to set trailing long waves as
-% perturbations, i.e. $k/k_T = -2Q^{-2}\left[1 - \sqrt{1 -
-%     Q^2(1-\nu^2)}\right]$.  
-% We only apply the perturbations where $Q^{-2}(1-\nu^2)<1$.
-
-% The amplitude of the surface density and velocity perturbations are
-% related by
-% \begin{align}
-%   &\left(\omega - m\Omega\right)\delta\Sigma + k\Sigma\delta v_R= 0,\\
-%   &\frac{\kappa^2}{2\Omega} \delta v_R + \ii (\omega - m\Omega)\delta
-%   v_\phi=0. 
-% \end{align}
-% We set $\delta\Sigma = 10^{-3}\Sigma$ and $(\delta v_R, \delta
-% v_\phi)$ accordingly.
-
-% %Having chosen the wave frequency and set the wavenumber according to
-% %the above,

### 2014-10-31 paper/results.tex
major change in paper - new idea - ang mom exchange with background because of temp gradient

--- 2014-10-30 paper/results.tex
@@
 \end{figure}
-
-\subsection{Non-linear initial perturbation}
-In order to focus on the one-arm spiral that emerges in the above
-simulation, we here analyse a run with only an initial $m=1$
-perturbation in the dead zone. Specifically, we set
-\begin{align}
-  &\Sigma \to \Sigma_0(R) + \delta\Sigma_0(R) + 2\Sigma_1(R)\cos{\phi},\\
-  &v_\phi  \to v_{\phi 0}(R) + 2v_{\phi 1}(R)\cos{\phi},
-\end{align}
-for $R\in[R_{d1},R_{d2}]$, where
-\begin{align}
-  &\delta\Sigma_0 =
-  -\delta^2\times\Sigma_0(R_{d1})\frac{R_{d1}}{R}\sin^3{\left[\frac{2\pi\left(R-R_{d1}\right)}{R_{d2}-R_{d1}}\right]},\\ 
-  &\Sigma_1 =
-  \delta\times\Sigma_0\sin{\left[\frac{\pi\left(R-R_{d1}\right)}{R_{d2}-R_{d1}}\right]},  
-\end{align}
-with $\delta = 10^{-3}$, and
-\begin{align}
-  v_{\phi 1} = -\frac{v_{\phi 0}\delta\Sigma_0}{2\Sigma_1}. 
-\end{align}
-This perturbs the dead zone structure while conserving total mass and
-angular momentum. The $m=1$ perturbation has an associated negative
-angular momentum (as observed in the previous simulation) at the
-expense of increasing the angular momentum associated with the
-background disc. %local exchange for simplicity 
-%actually doesn't matter if we set plus or minus
-Note that the perturbation to the axisymmetric surface density is
-$O(\delta^2)$, whereas that in $m=1$ is $O(\delta)$. Since $Q>1$
-everywhere in the disc, there is no risk of axisymmetric gravitational
-instability.
-
-While the above form of initial perturbations is artificial, we find 
-the spiral that forms have similar properties (see below) to that found in the
-previous simulation. However, the above perturbations have the
-advantage of producing cleaner results without complications from
-higher $m$ modes, and the required simulation time for the one-arm
-spiral to emerge is significantly reduced.
-
-Fig. shows the evolution of angular momenta.

### 2014-10-31 paper/results_exchange.tex
major change in paper - new idea - ang mom exchange with background because of temp gradient

\subsection{Non-linear initial perturbation}\label{fargo_nonlin}
In order to focus on the one-arm spiral that emerges in the above
simulation, we here analyse a run with only an initial $m=1$
perturbation in the dead zone. Specifically, we set
\begin{align}
  &\Sigma \to \Sigma_0(R) + \delta\Sigma_0(R) + 2\Sigma_1(R)\cos{\phi},\\
  &v_\phi  \to v_{\phi 0}(R) + 2v_{\phi 1}(R)\cos{\phi},
\end{align}
for $R\in[R_{d1},R_{d2}]$, where
\begin{align}
  &\delta\Sigma_0 =
  -\delta^2\times\Sigma_0(R_{d1})\frac{R_{d1}}{R}\sin^3{\left[\frac{2\pi\left(R-R_{d1}\right)}{R_{d2}-R_{d1}}\right]},\\ 
  &\Sigma_1 =
  \delta\times\Sigma_0\sin{\left[\frac{\pi\left(R-R_{d1}\right)}{R_{d2}-R_{d1}}\right]},  
\end{align}
with $\delta = 10^{-3}$, and
\begin{align}
  v_{\phi 1} = -\frac{v_{\phi 0}\delta\Sigma_0}{2\Sigma_1}. 
\end{align}
This perturbs the dead zone structure while conserving total mass and
angular momentum. The $m=1$ perturbation has an associated negative
angular momentum (as observed in the previous simulation) at the
expense of increasing the angular momentum associated with the
background disc. %local exchange for simplicity 
%actually doesn't matter if we set plus or minus
Note that the perturbation to the axisymmetric surface density is
$O(\delta^2)$, whereas that in $m=1$ is $O(\delta)$. Since $Q>1$
everywhere in the disc, there is no risk of axisymmetric gravitational
instability.

While the above form of initial perturbations is artificial, we find 
the spiral that forms have similar properties (see below) to that found in the
previous simulation, but the above procedure has the 
advantage of producing cleaner results without complications from
higher $m$ modes, and the required simulation time for the one-arm
spiral to emerge is reduced.

Fig. \ref{2d_angmom_ex} shows the evolution of angular momenta. For
$t\lesssim 15P_0$ there is angular momentum exchange between the
background and the imposed perturbation, but no overall growth takes
place. After $t\gtrsim15P_0$, however, the angular momenta stops 
oscillating and grows monotonically in magnitude.

\begin{figure}
  % scale=0.41
  \includegraphics[width=\linewidth]{figures/nonaxi_evol_ang_fargo_ex}
  \caption{Evolution of angular momentum components in the 
    FARGO simulation with the special $m=1$ initial perturbations
    described in \S\ref{fargo_nonlin}. 
    The perturbed angular momentum
    relative to $t=0$ is shown in units of the initial total angular
    momentum $J_\mathrm{ref}=J(t=0)$.\label{2d_angmom_ex}}  
\end{figure}

Fig. \ref{2d_angmom_ex_evol} show snapshots of the $m=1$
surface density at $t=10P_0$, when angular momenta oscillate; 
$t=15P_0$, when angular momenta stops oscillating; and
$t\in[20,30]P_0$, during the growth phase.

\begin{figure*}
  % scale=0.41
  \includegraphics[scale=0.4,clip=true,trim=0cm 0cm 1.73cm
  0cm]{figures/polarxy_fargo_ex_100}\includegraphics[scale=0.4,clip=true,trim=2.24cm
  0cm 1.73cm 0cm]{figures/polarxy_fargo_ex_150}\includegraphics[scale=0.4,clip=true,trim=2.24cm
  0cm 1.73cm 0cm]{figures/polarxy_fargo_ex_200}\includegraphics[scale=0.4,clip=true,trim=2.24cm
  0cm 1.73cm 0cm]{figures/polarxy_fargo_ex_250}\includegraphics[scale=0.4,clip=true,trim=2.24cm
  0cm 0.0cm 0cm]{figures/polarxy_fargo_ex_300}
  \caption{Evolutoin of the $m=1$ surface density in the FARGO
    simulation with the special $m=1$ initial perturbations
    described in \S\ref{fargo_nonlin}. 
    \label{2d_angmom_ex_evol}} 
\end{figure*}

### 2014-11-01 paper/model.tex
major updates

@@
 where $G$ is the gravitational constant,  and the disc potential
-$\Phi_d$. We further assume a locally 
+$\Phi_d$. We assume a locally 
 isothermal equation of state 
@@
 \end{align}
-where the sound-speed is given by $c_s=H\Omega_k$, $H=hR$ is the
-disc scale-height with constant aspect-ratio $h$ chosen below, and
-$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency.
+where the sound-speed is given by 
+\begin{align}
+  c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
+\end{align}
+where $h$ is the disc aspect-ratio $R=R_0$, 
+$\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
+$q$ is the imposed temperature gradient. A strictly isothermal disc
+corresponds to $q=0$, and a disc with constant aspect-ratio
+corresponds to $q=1$.
 
-We have purposefully adopted the simplest model that was found to
-develop one-arm spirals in discs with radial structure. The indirect
-potential associated with the non-inertial reference frame is
+% We have purposefully adopted the simplest model that was found to
+% develop one-arm spirals in discs with radial structure. 
+The indirect potential associated with the non-inertial reference frame is 
 neglected to avoid complications arising from the motion of the
 central star. For the adopted disc masses $M_d\lesssim 0.1 M_*$, this
-effect is not expected to be significant. The locally isothermal
-equation of state allows us to apply standard linear density wave
-theory for results interpretation. 
+effect is not expected to be significant.
+
+% The locally isothermal
+% equation of state allows us to apply standard linear density wave
+% theory for results interpretation.
+
 %This allows us to consider
@@
 
-\section{Diagnostics}
+\subsection{Diagnostics}
 
-\subsection{Evolution of non-axisymmetirc modes}
+\subsubsection{Evolution of non-axisymmetirc modes}
 The disc evolution is quantified using mode amplitudes and angular
@@
 \begin{align}
-  f(R,\phi) &= \sum_{m=-\infty}^{\infty}f_m(R)\exp{\ii m \phi} \notag\\
-  &= f_0 + 2 \real\left[\sum_{m=1}^\infty f_m \exp{(\ii
+  f(R,\phi,t) &= \sum_{m=-\infty}^{\infty}f_m(R,t)\exp{\ii m \phi} \notag\\
+  &= f_0(R) + 2 \real\left[\sum_{m=1}^\infty f_m \exp{(\ii
       m\phi)}\right], 
@@
 
-\subsection{Linear density waves}\label{wkb}
-In addition to the above diagnostics for disc evolution, we
-will also use linear density wave theory to interpret simulation
-results.
+\section{Linear density waves}\label{wkb}
+In this paper we take a fully numerical approach using non-linear
+hydrodynamic simulations. However, it will be useful to employ
+existing linear density wave theory to interpret the simulations. 
+We list here the main definitions and results from 2D linear theory,
+taken from \cite{lin11} and \cite{shu91}.
 
-\subsubsection{Global angular momentum conservation for linear
+In a linear analysis, one assumes a steady axisymmetric background state, such as
+that defined in \S\ref{model}, which is then perturbed such that
+\begin{align}  
+  \Sigma \to \Sigma(R) + \delta\Sigma_m(R)\exp{\left[\ii\left(-\sigma t +
+        m\phi\right)\right]}, 
+\end{align}
+and similarly for other variables. Since the background is
+axisymmetric, we identify $\delta\Sigma_m(R)\exp{\left(-\ii\sigma
+    t\right)} = \Sigma_m$ as defined for diagnostics. 
+The linearised mass and momentum equations are
+\begin{align}
+  &-\ii\sbar \delta\Sigma_m = -\frac{1}{R}\frac{d}{dR}\left(R\Sigma\delta
+    v_{Rm}\right) - \frac{\ii m \Sigma}{R}\delta v_{\phi m}, \\
+  &-\ii\sbar\delta v_{Rm} - 2\Omega \delta v_{\phi m} = -
+  c_s^2(R)\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right) - \frac{d}{dR}\delta\Phi_m,\\
+  & -\ii\sbar\delta v_{\phi m} + \frac{\kappa^2}{2\Omega}\delta v_{Rm} =
+  -\frac{\ii m }{R}\left(c_s^2\frac{\delta\Sigma_m}{\Sigma} + \delta\Phi_m\right),
+\end{align}
+where $\sbar = \sigma - m\Omega$ and $\kappa^2 =
+R^{-3}\p_R(R^4\Omega^2)$ is the square of the epicycle frequency. A
+locally isothermal equation of state has been assumed.    
+The linearised Poisson equation is 
+\begin{align}
+  \nabla^2\delta\Phi_m = 4\pi G \delta\Sigma_m \delta(z). 
+\end{align}
+
+The linearised equations can be combined to a single
+integro-differential eigenvalue problem. We list below some general
+properties of the linear perturbations that will be useful for
+understanding our simulations. We defer a full numerical
+exploration of the linear problem to a future study.
+
+\subsection{Global angular momentum conservation for linear
   perturbations} 
 It can be shown that linear perturbations with $\phi$-dependence in the form
-$\exp{\ii m\phi }$ satisfies angular momentum conservation
+$\exp{(\ii m\phi)}$ satisfies angular momentum conservation
 in the form 
@@
 where
-\begin{align}
+\begin{align}\label{lin_ang_mom_cons}
   \jlin \equiv
@@
 expressions for $\bm{\xi}$ and $\bm{F}$ can be found in e.g.,
-\cite{papaloizou85} and \cite{lin11} respectively. In the above
-expressions, $\Sigma$ and $\Omega$ correspond to the unperturbed disc.
+\cite{papaloizou85} and \cite{lin11} respectively.
 
 The term $\tbg$ is 
-\begin{align}
+\begin{align}\label{baroclinic_torque}
   \tbg \equiv
-  -\frac{m}{2}\imag\left(\Sigma_m\xi_R^\prime\frac{dc_s^2}{dR}\right), 
+  -\frac{m}{2}\imag\left(\Sigma_m\xi_R^*\frac{dc_s^2}{dR}\right), 
 \end{align}
@@
 
-\subsubsection{Local results}
-\cite{shu91} gives the dispersion relation for tightly-wound density
-waves of the form $\exp{[\ii(\omega t - m \phi + kR)]}$ in a razor-thin
-isothermal disc as  
+\subsection{Local results}
+%The linearised equations correspond to an integro-differential
+%equation eigenvalue problem. 
+In the local approximation, perturbations are assumed to vary rapidly
+relative to any background gradients. \cite{shu91} gives the
+dispersion relation for tightly-wound density 
+waves of the form $\exp{[\ii(-\sigma t + m \phi + kR)]}$ in a razor-thin
+disc as  
 \begin{align}\label{dispersion}
-  (\omega - m\Omega)^2 = \kappa^2 + k^2c_s^2 - 2\pi G \Sigma |k|, 
+  (\sigma - m\Omega)^2 = \kappa^2 + k^2c_s^2 - 2\pi G \Sigma |k|, 
 \end{align}
-where $\kappa^2 = R^{-3}\p_R(R^4\Omega^2)$ is the square of the epicycle frequency, 
-and $k$ is a real wavenumber such that $|kR|\gg1$.
+and $k$ is a real wavenumber such that $|kR|\gg1$. Note that in this
+approximation, only axisymmetric perturbations with $m=0$ can be
+unstable (having an imaginary $\sigma$).
 
-Given the pattern speed $\Omega_p$,
-Eq. \ref{dispersion} can be solved for $|k|$, 
+Given the pattern speed $\Omega_p$, Eq. \ref{dispersion} can be solved
+for $|k|$,   
 \begin{align}\label{wavenumber}
-  \frac{|k|}{k_T} = \frac{2}{Q^2}\left[1 \pm \sqrt{1 -
+  |k| = \frac{\pi G \Sigma}{c_s^2}\left[1 \pm \sqrt{1 -
       Q^2(1-\nu^2)}\right], 
@@
 where 
-\begin{align}
-  k_T \equiv \frac{\kappa^2}{2\pi G \Sigma}
-\end{align}
-is a characteristic wavenumber, 
+% \begin{align}
+%   k_T \equiv \frac{\kappa^2}{2\pi G \Sigma}
+% \end{align}
+% is a characteristic wavenumber, 
 \begin{align}
@@
 real $k$ are only possible when $Q^2(1-\nu^2)\leq1$.
+
+A detailed discussion of the properties of density waves in the local
+approximation can be found in \cite{shu91}. A important result is that
+waves interior to co-rotation ($R<R_c$) have negative angular momentum, while
+waves outside co-rotation ($R>R_c$) have positive angular
+momentum. The local approximation becomes less good for gobal modes,
+but will nevertheless be useful in results interpretation.

### 2014-11-01 paper/paper.tex
update
major change in paper - new idea - ang mom exchange with background because of temp gradient
major updates

--- 2014-10-22 paper/paper.tex
@@
 \newcommand{\tcool}{t_\mathrm{cool}}
+\newcommand{\jlin}{j_\mathrm{lin}}
+\newcommand{\jlintot}{J_\mathrm{lin}}
+\newcommand{\tbg}{T_\mathrm{BG}}
 \DeclareMathOperator{\erf}{erf}
@@
 
-\title[One-arm spirals in self-gravitating discs]{One-arm spirals in
-  self-gravitating dead zones in protoplanetary discs}
+\title[One-arm spirals]{One-arm spirals in 
+  locally isothermal, radially structured self-gravitating discs}
 
@@
 \begin{abstract} 
-  Theoretical models of circumstellar or protoplanetary discs indicate
-  the possible existence of `dead zones' --- regions in the disc
-  where mass accretion is greatly reduced because of inefficient
-  magneto-hydrodynamic turbulence that would otherwise enable 
-  accretion. It has been suggested that mass built-up in a dead zone
-  may lead to gravitational instability. However, the type of
-  gravitational instability relevant in this context has not been
-  clarified. In this work, we use numerical simulations to explore
-  gravitational instability in astrophysical discs in which only a 
-  sub-portion of it is self-gravitating, taken as a toy model for
-  protoplanetary disc containing a dead zone. We conduct numerical hydrodynamic
-  experiments in 2D using the FARGO code and in 3D with the ZEUS-MP
-  and PLUTO codes. In all cases we observe the development of a spiral
-  mode with azimuthal wavenumber $m=1$ inside the dead
-  zone. Surprisingly, the co-rotation radius of the spiral is well
-  outside the dead zone, resulting in a nearly stationary one-arm spiral
-  pattern in the dead zone. Our simulations demonstrate the potential
-  importance global gravitational interaction between a
-  self-gravitating dead zone and the exterior, non-self-gravitating
-  parts of the disc.  
+  We present numerical simulations of gravitational instability in
+  astrophysical discs with an an imposed temperature profile, and in
+  which only a sub-portion of it is self-gravitating. We conduct
+  numerical hydrodynamic experiments in 2D using the FARGO code and in
+  3D with the ZEUS-MP and PLUTO codes. In such systems we observe the
+  development of a 
+  spiral mode with azimuthal wavenumber $m=1$ with co-rotation radius well
+  outside the bulk of the spiral arm, resulting in a nearly stationary
+  one-arm spiral pattern in the self-gravitating region. This one-arm
+  spiral does not require the movement of 
+  the central star, as found in previous studies. Instead, we find it
+  to be associated with a structured disc, as well as an imposed
+  temperature gradient. We apply results
+  from existing linear theory and attribute the growth of this spiral
+  to the non-conservation of angular momentum for linear perturbations
+  in discs with an imposed temperature gradient. This enables angular
+  momentum exchange between the spiral and the background disc.
+
+  % or confirm in 3D
+  %, taken as a toy model of a 
+  %protoplanetary disc containing a finite-size massive dead zone.
+   % Theoretical models of circumstellar or protoplanetary discs indicate
+  % the possible existence of `dead zones' --- regions in the disc
+  % where mass accretion is greatly reduced because of inefficient
+  % magneto-hydrodynamic turbulence that would otherwise enable 
+  % accretion. It has been suggested that mass built-up in a dead zone
+  % may lead to gravitational instability. However, what type of
+  % gravitational instability is relevant in this context has not been
+  % clarified.
+
+  %
+  % Our simulations demonstrate the 
+  % importance of global gravitational interaction between a 
+  % self-gravitating dead zone and the exterior, non-self-gravitating
+  % parts of the disc. This non-local effect, due to disc structure, 
+  % exists even when the total disc mass is below the threshold for
+  % global effects found in
+  % previous studies. 
 \end{abstract}
@@
 \input{results}
+%\input{results_3d}
 %\input{discussion}

### 2014-11-01 paper/results.tex
major updates

@@
 \begin{align}\label{randpert}
-  \frac{\delta v_R}{c_s} = \frac{\delta}{M}\times T(R) \sum_{m=1}^M\cos{m\phi},
+  v_R \to v_R+ c_s\frac{\delta}{M}
+  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}_d}{\Delta 
+          R_d}\right)^2\right]}\sum_{m=1}^M\cos{m\phi},
 \end{align}
-where $\delta$ is an amplitude and 
-\begin{align}
-  T(R) =
-  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}_d}{\Delta
-          R_d}\right)^2\right]}, 
-\end{align}
-where $\overline{R}_d = (R_{d1}+R_{d2})/2$ and $\Delta R_d =
-(R_{d2}-R_{d1})/2$. We use $M=10$ and choose
+where $\delta$ is an amplitude, $\overline{R}_d = (R_{d1}+R_{d2})/2$
+and $\Delta R_d = (R_{d2}-R_{d1})/2$. We use $M=10$ and choose 
 $\delta\in[-10^{-3},10^{-3}]$ randomly but independent of $\phi$.
@@
 Fig. \ref{fargo_2d}. 
-At early times $t\lesssim100P_0$ the dead zone is
+At early times $t\lesssim100P_0$ the disc 
 dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
@@
 \citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
-again after $t=70P_0$, and eventually dominates the dead zone.
+again after $t=70P_0$, and eventually dominates the annulus.
 
@@
 \end{figure}
+
+\subsection{Trapping of the low-frequency $m=1$ spiral}
+In order to focus on the $m=1$ spiral that eventually emerges, here 
+we examine a simulation initialised with only $m=1$
+perturbations, by setting $M=1$ in Eq. \ref{randpert}. The disc
+evolution is largely similar to the previous case, but the final spiral is
+more co-herent, facilitating the analysis below. 
+Fig. \ref{2d_fargo_viz} shows the $m=1$ surface density
+distribution at the end of this simulation.
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{figures/polarxy2_dens120_fargo.ps}
+  \caption{Cartesian visualisation of the $m=1$ surface density
+    structure in the FARGO simulation initialised with only $m=1$
+    perturbations. 
+    \label{2d_fargo_viz}} 
+\end{figure}
+
+By examining the disc evolution in $t\in[80,120]P_0$, we find the
+one-arm spiral in Fig. \ref{2d_fargo_viz} has a 
+co-rotation radius and growth rate of
+\begin{align*}
+  &R_c \simeq 4.4R_0,\\
+  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_k(R_c). 
+\end{align*}
+The spiral can be considered as low frequency because $|\omega|
+\lesssim 0.3\Omega$ (recall $\omega = 
+m\Omega_p$) in the region it is promininent $R\in[R_{d1},R_{d2}]$. The
+growth rate is also slow relative to the local rotation. This gives a
+nearly stationary spiral pattern.
+
+%Since most of
+%the perturbation lies at $R<R_c$, this mode has negative
+%angular momentum according to local linear theory.
+
+Next, we write $\Sigma_1 = 
+|\Sigma_1|\exp{(\ii kR)}$, where $k$ is real, and assume the amplitude
+$|\Sigma_1|$ varies slowly compared to the complex phase. This is the
+main assumption in local theory. We plot in
+Fig. \ref{fargo_wavenumber} the normalise wavenumber associated with
+the $m=1$ spiral. We find that
+\begin{align*}
+  kR \sim \frac{\pi G \Sigma}{c_s^2}R \sim \frac{1}{hQ}, 
+\end{align*}
+where we used $Q\sim c_s\Omega/\pi G \Sigma$ and $R\Omega/c_s\sim
+h^{-1}$. Since $Q=O(1)$ and $h\ll 1$, $|kR|\gg 1$ and we 
+can apply results from local theory self-consistently. Since $k>0$,
+the spiral is trailing, as seen in Fig. \ref{2d_fargo_viz}.
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{figures/m1_analysis_kr120_fargo.ps}
+  \caption{Normalised radial wavenumber of the $m=1$ spiral in
+    Fig. \ref{2d_fargo_viz}.\label{fargo_wavenumber}} 
+\end{figure}
+
+Using the estimated value of $R_c$, we plot in
+Fig. \ref{fargo_qbarrier} the quantity $\nu^2 - 1 + Q^{-2}$, which is
+required to be positive in local theory for purely wave-like
+solutions to the dispersion relation (Eq. \ref{dispersion}) when the
+mode frequency is given. The figure shows two $Q$-barriers located in
+the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$, which is indeed the
+region where the $m=1$ spiral develops. This means that the one-arm
+spiral is trapped. There is one outer Lindblad resonance at $R_L\simeq
+7.2R_0$. Thus, waves may be launched in $R\gtrsim 7.2R_0$ by the main 
+spiral disturbance in the inner disc. %this would be a de-stabilising
+% effect.
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{figures/m1_analysis_Qbar_fargo.ps} 
+  \caption{Dimensionless mode frequency $\nu$ for the $m=1$ spiral in
+    Fig. \ref{2d_fargo_viz}. The disperion relation for local density
+    waves, Eq. \ref{dispersion}, permits purely wave-like solutions in 
+    regions where $\nu^2 - 1 + Q^{-2}>0$. 
+    \label{fargo_qbarrier}} 
+\end{figure}
+
+\subsection{Angular momentum exchange with the background disc} 
+The wavenumber $k = \pi G\Sigma/c_s^2$ corresponds to the
+wavelength most susceptible to gravitational instability for
+\emph{axisymmetric} perturbations. Physically, then, it is not
+surprising to find perturbations of this wavenumber in a
+self-gravitating disc where $Q\sim 1$. However, according the
+local dispersion relation, $m=1$ perturbations are formally stable for 
+real $k$.
+
+In order to track down the origin for the (slow) growth of the
+$m=1$ spiral, we invoke angular momentum conservation for linear
+perturbations, Eq. \ref{lin_ang_mom_cons}. Assuming fluxes are
+negligible at the disc boundaries, we can integrate
+Eq. \ref{lin_ang_mom_cons} to give
+\begin{align}\label{baroclinic_torque_int}
+  \frac{d}{dt}\underbrace{\int_{R_\mathrm{min}}^{R_\mathrm{max}}\jlin
+    2\pi R dR}_{\jlintot} 
+  =\int_{R\mathrm{min}}^{R\mathrm{max}}T_\mathrm{BG} 2\pi R dR, 
+\end{align}
+where we recall $T_{BG}$ is the torque density related to the imposed
+sound-speed profile (Eq. \ref{baroclinic_torque}). We have explicitly
+computed both sides of Eq. \ref{baroclinic_torque_int} using
+simulation data, and compare them in Fig. \ref{fargo_angmom_ex}.
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{figures/m1_analysis_ang_fargo.ps} 
+  \caption{
+    \label{fargo_angmom_ex}} 
+\end{figure}
+
+%ang mom decreases slightly more rapidly than torque exchange provides
+%- probably non-linear effects/shocks

### 2014-11-02 paper/discussion.tex
added analytical discussion

@@
-\section{Additional results analysis}\label{additional}
-In this section we examine some secondary quantities derived
-from the hydrodynamic simulations above. To keep this discussion
-concise, we will use selected simulations from above for illustration.
+\section{Discussion}
 
-\subsection{Three-dimensionality}
-A simple measure of three-dimensionality of the flow is to compare vertical 
-to horizontal motion. Since we are interested in non-axisymmetric
-perturbations to the gap edge, we first Fourier transform the
-meridional momentum densities
+\subsection{Angular momentum density of  low-frequency $m=1$ modes}
+From Eq. \ref{lin_ang_mom_cons} and assuming a time-dependence of the
+form $\exp{(-\ii \sigma t)}$ with $\real{\sigma} = \omega$,  
+the angular momentum density associated with linear waves is
+\begin{align}
+  \jlin = \frac{m\Sigma}{2}\left[\left(\omega -
+      m\Omega\right)|\bm{\xi}|^2 + \ii\Omega\left(\xi_R\xi_\phi^* -
+      \xi_R\xi_\phi^*\right)\right].  
+\end{align}
+For a low-frequency mode, $|\omega|\ll m\Omega$. Then neglecting the
+term $\omega|\bm{\xi}|^2$ and setting $m=1$,
+\begin{align}
+  \jlin &\simeq \frac{\Sigma\Omega}{2}\left[-|\bm{\xi}|^2 + \ii\left(\xi_R\xi_\phi^* -
+      \xi_R\xi_\phi^*\right)\right]\notag\\
+  & = -\frac{\Sigma\Omega}{2}|\xi_R + \ii\xi_\phi|^2\notag\\
+  & \leq 0.
+\end{align}
+Thus, low-frequency $m=1$ modes are generally associated with negative
+angular momentum. If the mode loses (positive) angular momentum
+to the background, then we can expect instability. We demonstrate
+this below.
 
+\subsection{Unstable interaction between low-frequency $m=1$ spirals
+  and the background disc due to an imposed temperature gradient}
+The $m=1$ spiral that emerges in our simulations have low frequency
+and grows slowly, and its spatial form is consistent local theory
+(\S\ref{fargo_m1}). To demonstrate possible instability, we
+first construct a neutral low-frequency mode using local theory, then
+show that the torque density acting on this mode due to the background
+temperature gradient is negative, which enforces the mode.
+
+We begin by recalling the expression for the torque density due to a
+locally isothermal equation of state,
 \begin{align}
-  (v_{Rm}, v_{zm} ) \equiv\int_0^{2\pi}\rho\times(u_R,
-  u_z)\exp{(-\ii m\phi)}d\phi. 
+  T_\mathrm{BG} &= -\frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\Sigma_m \left(\frac{\delta
+        v_{Rm}}{-\ii\sbar}\right)^*\right] \notag\\
+  & = \frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\ii \Sigma_m \left(\frac{\delta
+        v_{Rm}}{\sbar}\right)^*\right]
 \end{align}
-We define the three dimensionality as $\Theta_m(z/H)$, where 
+where we replaced the radial Lagrangian displacement by the Eulerian
+radial velocity perturbation in Eq. \ref{baroclinic_torque}. 
+The linearised momentum equations give
 \begin{align}
-  \Theta_m^2 \equiv \frac{\avg{|v_{zm}|^2}}{\avg{|v_{zm}|^2}
-    +\avg{|v_{Rm}|^2}},  
+  D\delta v_{Rm} =& 
+  \ii\sbar\left[c_s^2\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right)
+    + \frac{d}{dR}\delta\Phi_m\right] \notag\\ 
+  &- \frac{2\ii
+    m\Omega}{R}\left(c_s^2\frac{\delta\Sigma_m}{\Sigma} +
+    \delta\Phi_m\right),
 \end{align}
-and $\avg{\cdot}$ denotes a radial average. 
-Admittedly, this is a crude measure, and exact values of $\Theta_m$
-varies somewhat with details of the average. However, we have
-experimented with different averaging domains and 
-found the features described below are robust.
+where $D\equiv \kappa^2 - \sbar^2$.
 
-The top panel in Fig. \ref{compare_vprofiles_3d008} shows $\Theta_m$
-for Cases 1--3 at $t=40P_0$. The radial average is taken over
-$r-r_p\in[3,7]r_h$. These are all vortex modes (see
-Fig. \ref{vortex8_polar_dens} and
-Fig. \ref{vortex4_vortex10_overall}). The flow becomes increasingly
-three-dimensional away from the midplane but 
-$\Theta_m = O(10^{-1})$ is small. In an averaged sense the flow is
-mostly horizontal. At the end of the  
-simulation for Case 1, an azimuthally extended vortex dominates the
-flow, for which we measured $\Theta_1\sim
-0.2$---0.3. Thus, although vertical motion can become an appreciable
-fraction of horizontal motion, the former never dominates.
+We now invoke local theory by setting $d/dR \to \ii k$ where $k$ is
+real, and using the local solution to the Poisson equation
+\begin{align}
+  \delta\Phi_m = -\frac{2\pi G}{|k|}\delta\Sigma_m 
+\end{align}
+\citep{shu91}. Then the radial velocity perturbation becomes
+\begin{align}
+  \delta v_{Rm} =
+  \frac{1}{D}\frac{\delta\Sigma_m}{\Sigma}\left(\frac{2\pi G
+      \Sigma}{|k|} - c_s^2\right)\left(k\sbar + \frac{2\ii
+      m\Omega}{R}\right),
+\end{align}
+so that 
+\begin{align}
+  \ii\delta\Sigma_m \left(\frac{\delta v_{Rm}}{\sbar}\right)^* =
+  \frac{1}{D^*}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
+      \Sigma}{|k|} - c_s^2\right)\left(\ii k  + \frac{2
+      m\Omega}{R\sbar^*}\right). 
+\end{align}
+For a neutral mode, $\sbar$, and therfore $D$ is real. Then, after
+taking the imaginary part of the above expression, the torque
+density is 
+\begin{align}
+  T_\mathrm{BG} &=
+  \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{D}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
+      \Sigma}{|k|} - c_s^2\right)\right] \notag\\
+&\simeq \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{\left(\kappa^2 -
+        m^2\Omega^2\right)}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi
+      G \Sigma}{|k|} - c_s^2\right)\right]. 
+\end{align}
+where the second line follows from the low-frequency approximation
+$|\sigma|\ll m\Omega$.
 
-$\Theta_m$ for Cases 4---7 are shown in in bottom panel of
-Fig. \ref{compare_vprofiles_3d008}. The radial average is performed
-over $r-r_p\in[2,6]r_h$ 
-because the global spirals in Cases 6---7 significantly protrude the
-gap edge. The snapshot is taken at $t=50P_0$ for Cases 4---5, at
-$t=40P_0$ for Case 6 and at $t=30P_0$, so that the
-vortices and spirals have comparable over-densities at the
-gap edge. It also reflects the fact that spiral modes are more
-unstable than vortex modes and develop earlier  
-\citep{lin11b}. $\Theta_m \sim 0.2$ is again not particularly
-large, but the spiral modes are distinctly more three-dimensional than
-vortex modes. This is likely due to additional vertical
-acceleration provided by the strong self-gravity in those cases.
-
-%
-% can't do detail comparision between cases because instability have
-% different growth rates. different 'stages' in instability
-
-% Case 6 show 3 edge disturbances but this is likely to be a mixture
-% of spiral and vortex modes, rather than the m=3 spiral mode. This is
-% because 
-% we clearly see m=2 spiral in Case 7 (Q_0=1.5), and to see higher m spiral
-% modes one should increase the strength of self-gravity (e.g. lin11b
-% observe m=3 spirals with Q_0=1.2) 
-%
-
-\begin{figure}
-  \centering
-  \includegraphics[scale=.425,clip=true,trim=0.2cm 1.7cm 0cm
-    0cm]{figures/compare_vprofiles_3d008}  
-  \includegraphics[scale=.425,clip=true,trim=0.2cm 0cm 0cm
-    0.2cm]{figures/compare_vprofiles_3d010}  
-  \caption{Three-dimensionality of the non-axisymmetric
-    flow near the outer gap edge. Top: Cases 1---3 (vortex modes). 
-    Bottom: Cases 4---5 (vortex modes) and Cases 6---7 (spiral
-    modes). The azimuthal wavenumber $m$ is chosen to match the number
-    of vortices or large-scale spirals observed. 
-    \label{compare_vprofiles_3d008}}
-\end{figure}
-
-\subsection{Vortensity field}\label{vortensity}
-A fundamental distinction between the linear vortex and edge mode
-instability is their association with local vortensity minimum and
-maximum, respectively. In this section we compare vortensity fields of 
-discs with vortex modes (Case 2) and edge modes (Case 7). 
-%We use Case 2 
-%and Case 7 as examples, respectively. 
-More specifically, we examine the relative
-perturbation to the vertical component of vortensity,  
-\begin{align}
-  \Delta\eta_z\equiv \frac{\eta_z - \eta_z(t=0)}{\eta_z(t=0)},
-\end{align}
-where
-\begin{align}
-  \eta_z \equiv \frac{\bm{\hat{z}}\cdot\nabla\times\bm{u}}{\rho}. 
-\end{align}
-
-Fig. \ref{vortex4_vortex1_vortxy} compares $\Delta\eta_z$ in the
-midplane when vortices and spirals develop. For planetary gaps, 
-vortensity maxima and minima are both located near the gap edges with 
-characteristic separation of the local scale-height. The vortensity
-ring at the inner gap edge ($r-r_p\simeq - 2r_h$) remain
-well-defined. The vortex instability is associated with the local
-vortensity minimum near the outer gap edge --- seen as localised closed
-contour lines centred about $r - r_p \sim 4r_h$. The vortensity ring at
-$r-r_p\sim +2r_h$ becomes distorted as a \emph{consequence} of
-large-scale vortex formation just exterior to it. By contrast, the
-edge-spiral mode is associated with the local vortensity
-maximum. Their development inherently disrupts the vortensity
-rings. This is seen in the right panel as the outer ring is broken up.
-
-\begin{figure}
-  \centering
-  \includegraphics[scale=.425,clip=true,trim=0cm 0cm 1.9cm 
-    0.9cm]{figures/vortex4_vortxy_008}\includegraphics[scale=.425,clip=true,trim=2.3cm    
-    0.0cm 0cm 
-    0.9cm]{figures/vortex1_vortxy_006}
-  \caption{Relative perturbation to the vertical component of
-    vortensity at the midplane in a disc with the vortex instability
-    (left, Case 2 at $t=40P_0$) and the 
-    spiral instability (right, Case 7 at $t=30P_0$). Negative
-    perturbations in the region $r-r_p\in[2,6]r_h$ are outlined by white lines. 
-    Dotted horizontal lines indicate azimuthal cuts taken in
-    Fig. \ref{vortex4_vortex1_vortRZ}.   
-    \label{vortex4_vortex1_vortxy}}
-\end{figure}
-
-The vertical structures also differ. Fig. \ref{vortex4_vortex1_vortRZ}
-compares $\Delta\eta_z$ at azimuths coinciding with 
-a vortex or the edge disturbance of the spiral mode. Both
-instabilities involve $\Delta\eta_z<0$. It is clear that 
-the spiral mode has stronger vertical dependence. Its 
-region of $\Delta\eta_z<0$ becomes thinner away from
-the midplane. In the vortex case this region remains 
-about the same width and  $\Delta\eta_z$ is approximately uniform
-within it. 
-%This is qualitatively
-%consistent with increased (downards) vertical motion. 
-While $\mathrm{min}(\Delta\eta_z)$ is of comparable
-magnitude, the vortensity ring at $r-r_p=2r_h$is much weaker and
-thinner in the spiral case ($\Delta\eta_z$ being a factor $\sim 4$
-smaller than the vortex case).
-
-\begin{figure}
-  \centering
-  \includegraphics[scale=.47,clip=true,trim=0cm 1.32cm .0cm
-    .6cm]{figures/vortex4_vortRZ_008}\\\includegraphics[scale=.47,clip=true,trim=0cm 
-    0.cm .0cm
-    .6cm]{figures/vortex1_vortRZ_006}
-  \caption{Relative vertical vortensity perturbation
-    associated with the vortex instability (top) and spiral
-    instability (bottom). The slices are taken at azimuths shown by
-    white dotted lines in Fig. \ref{vortex4_vortex1_vortxy}. Negative
-    perturbations in the region $r-r_p\in[2,5.5]r_h$ are outlined by
-    white lines. \label{vortex4_vortex1_vortRZ}} 
-\end{figure}
-
-\subsection{Disc-planet torques}
-The presence of non-axisymmetric disturbances at gap edges is
-expected to significantly affect disc-planet torques. It has been 
-confirmed in 2D simulations that both vortex and spiral modes lead to
-oscillatory torques of either sign \citep{li05,lin11b}. 
-%Limited 
-%numerical resolution the current 3D calculations prevent accurate torque measurements, 
-It this section we measure the disc-on-planet torques in several of the above simulations
-to confirm the main features found in 2D.
-
-We calculate the specific torque acting on the planet due to a mass
-element as 
-\begin{align}
-d\bm{T}(\bm{r}) &=
-\frac{\bm{r}_p\times\bm{r}G\rho(\bm{r})d^3\bm{r}}{d_p^3}f(\bm{r},\bm{r}_p),\\
-f(\bm{r},\bm{r}_p)&\equiv 1
--\exp{\left(-\frac{1}{2}\left|\frac{\bm{r}-\bm{r}_p}{\epsilon_c
-      r_h}\right|^2\right)}.  
-\end{align} 
-The tapering function $f$ reduce contributions from close to the
-planet, thereby reducing numerical artifacts arising from this region
-because of the diverging potential and limited resolution. We set  
-the parameter $\epsilon_c=1$ so that tapering does not significantly
-reduce contributions from the instabilities, since they develop
-at $\gtrsim 2r_h$ away from the planet's orbital radius.
-
-Fig. \ref{torque3} shows the disc-on-planet torques in Case 3 and Case 5, which develop
-the $m=5$ and $m=6$ vortex modes, respectively. These plots are qualitatively similar to
-2D simulations \citep[e.g.][]{li05}. The torques oscillate on orbital 
-time-scales and its instantaneous values can be of either sign. However, 
-upon averaging over the simulation we find the total torques are negative in both cases. This
-means inwards migration is still favoured. 
-%The planet is expected to interact
-
-We extended Case 5 to $t=135P_0$ and find the vortices 
-have similar over-densities as at $t=50P_0$. However, 
-Fig. \ref{torque3} show the torque oscillation amplitudes decrease
-towards the end of Case 5 compared to $t\in[40,80]P_0$.  
-At $t=50P_0$ the vortices are located in $r-r_p\in[3.5,5.5]r_h$ but by
-$t=135P_0$ they are located in $r-r_p\in[4,6]r_h$. 
-Given that $t\in[40,80]P_0$ is only $20P_0$ to $50P_0$ after the planet  
-potential has been fully introduced, gap-formation is probably 
-ongoing during this time. We expect torque amplitudes to be larger
-during  gap-formation since the vortices lie closer to the planet. 
-%On the
-%other hand, gap-widening in the presence of vortices was also seen in  
-%2D simulations on long timescales \citep{lin11a}  
-%So there may also be
-%contributions from radial redistribution of material due to the
-%vortices. 
-% This might be
-%brought about by the of the density bump at the outer gap edge
-%\citep{meheut12b}
-
-%The original
-%density bump at the outer gap edge ($r-r_p\simeq 4r_h$) shifts outward
-%between these times. 
-%This is unlike in the fixed-orbit 2D simulation by \cite{li05}, where
-%the oscillation amplitude does not decrease over several hundred
-%orbits. 
-%The different setups (especially self-gravity and numerical resolution) 
-%prevent a detail comparison 
-roclinic instability
-
-\begin{figure}
-  \centering
-  \includegraphics[scale=.425,clip=true,trim=0.2cm 1.cm 0cm
-    0cm]{figures/vortex10_torque3_tqex}  
-  \includegraphics[scale=.425,clip=true,trim=0.2cm 0cm 0cm 0.2cm]{figures/vortex2b_torque3_tqex} 
-  \caption{Instantaneous disc-on-planet torques in simulations where the vortex mode develops. 
-    Top: Case 3. Bottom: Case 5. Note that Case 5 has been extended to $t=135P_0$. \label{torque3}}
-\end{figure}
-
-Next we examine disc-planet torques in the presence of the spiral
-modes. The top panel of Fig. \ref{torque3_spiral} shows the
-instantaneous disc-on-planet torques. Contributions from the
-inner disc ($r < r_p$) and outer disc ($r > r_p$) are plotted
-separately for comparison with Fig. 18b in \cite{lin11b}, which is 
-similar to the present plot. Large oscillations in the outer torque
-due to edge mode spirals cause the total torque to be positive or
-negative at a given instant.
-
-Unlike the vortex modes, Fig. \ref{torque3_spiral} shows that spiral
-modes can lead to a positive running-time averaged torques (bottom panel). 
-The average torques become more positive with time after spiral modes develop, 
-and with increasing self-gravity (which increases the instability strength). 
-%The torque is more positive with
-%increasing strength of the spiral instability.
-This was also observed in high-resolution 2D simulations in
-\cite{lin11b}. There it was suggested that the creation of large   
-`voids' in between spiral arms decreases the time-averaged density in  
-the planet-induced wakes, thereby reducing associated torque
-magnitudes. Since the outer planetary wake normally provide a negative
-torque, the spiral modes make the total torque more positive. 
-The similarity between 2D and 3D results indicate that
-outwards migration induced by spiral modes, which was seen in 2D by
-\cite{lin11b,lin12b}, will also operate in 3D.
-
-%It was later shown that the spiral instability can supply material to
-%go on horseshoe turns ahead of the planet, which also provide a
-%positive torque \citep{lin12b}. This is because the spiral modes 
-%are associated with local vortensity maximum which are located just
-%inside the gap edge in the co-orbital region.
-
-%This is because the
-%spirals are associated with local vortensity maximum, which is just
-%inside the outer gap edge. Thus over-densities can be supplied by the
-%instability to go on horseshoe orbits
-
-%Since spiral modes are associated with disturbances at the local
-%vortensity maximum just inside the outer gap edge, its influence on
-%disc-planet torques is expected to be more si
-
-\begin{figure}
-  \centering
-  \includegraphics[scale=.425,clip=true,trim=0.2cm 1.825cm 0cm
-    0cm]{figures/vortex1_torque2_tqex}  
-  \includegraphics[scale=.425,clip=true,trim=0.2cm 0cm 0cm 0.2cm]{figures/torque3_tqex} 
-  \caption{Disc-on-planet torques in the presence of spiral modes
-    associated with the outer gap edge. Top: total torque (solid),
-    torque from the inner disc (dotted) and from the outer disc
-    (dashed) in Case 7. Bottom: time-averaged torques in Case 6
-    (solid) and Case 7 (dotted).
-\label{torque3_spiral}}
-\end{figure}
+To determine the sign of $T_\mathrm{BG}$ for low frequency one-arm spirals,
+we examine each factor separately below for $m=1$:
+\begin{itemize}
+\item We observe trailing spirals in our simulations, so $k>0$. 
+\item The factor $\kappa^2 - \Omega^2>0$ for our self-gravitating disc
+  models in the region where the spiral appears. 
+\item We measure $|k|\sim \pi G\Sigma/c_s^2$, so the last factor in
+  the square brackets is positive. Note that this factor is positive
+  provided $|k| < 2/HQ$ or wavelengths $\lambda > \pi Q H$.  
+\end{itemize}
+These considerations imply the factor in the square brackets is
+positive. Then, since $dc_s^2/dR < 0$, $T_\mathrm{BG}<0$ so the background
+disc applies a negative torque on the low-frequency $m=1$ spiral,
+which has negative angular momentum. This suggests the spiral
+amplitude will grow.

### 2014-11-02 paper/results.tex
added analytical discussion

@@
 
-\subsection{Trapping of the low-frequency $m=1$ spiral}
+\subsection{Trapping the low-frequency $m=1$ spiral}\label{fargo_m1}
 In order to focus on the $m=1$ spiral that eventually emerges, here 

### 2014-11-02 paper/discussion.tex
added variable q discussion

@@
 density is 
-\begin{align}
+\begin{align}\label{tbg_explicit}
   T_\mathrm{BG} &=
   \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{D}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
-      \Sigma}{|k|} - c_s^2\right)\right] \notag\\
-&\simeq \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{\left(\kappa^2 -
-        m^2\Omega^2\right)}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi
-      G \Sigma}{|k|} - c_s^2\right)\right]. 
+      \Sigma}{|k|} - c_s^2\right)\right].
+% &\simeq \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{\left(\kappa^2 -
+%         m^2\Omega^2\right)}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi
+%       G \Sigma}{|k|} - c_s^2\right)\right]. 
 \end{align}
-where the second line follows from the low-frequency approximation
-$|\sigma|\ll m\Omega$.
+%where the second line follows from the low-frequency approximation
+%$|\sigma|\ll m\Omega$. 
+% To determine the sign of $T_\mathrm{BG}$ for low frequency one-arm spirals,
+% we examine each factor separately below for $m=1$:
+% \begin{itemize}
+% \item We observe trailing spirals in our simulations, so $k>0$. 
+% \item The factor $\kappa^2 - \Omega^2>0$ for our self-gravitating disc
+%   models in the region where the spiral appears. 
+% \item We measure $|k|\sim \pi G\Sigma/c_s^2$, so the last factor in
+%   the square brackets is positive. Note that this factor is positive
+%   provided $|k| < 2/HQ$ or wavelengths $\lambda > \pi Q H$.  
+% \end{itemize}
+We now use the local dispersion relation,
+Eq. \ref{dispersion}, to make the replacement $D = 2\pi G \Sigma |k| -
+k^2c_s^2$ in the above expression. This gives
+\begin{align}\label{tbg_simple}
+  T_\mathrm{BG} =
+  \frac{m}{2}\frac{dc_s^2}{dR}\frac{1}{k}\frac{|\delta\Sigma_m|^2}{\Sigma}.  
+\end{align} 
+This torque density is negative for trailing waves in discs with
+temperature decreasing outwards ($k>0$ and $dc_s^2/dR<0$,
+respectively). Note that this conclusion does not rely on the
+low-frequency approximation and holds for any $m$.
 
-To determine the sign of $T_\mathrm{BG}$ for low frequency one-arm spirals,
-we examine each factor separately below for $m=1$:
-\begin{itemize}
-\item We observe trailing spirals in our simulations, so $k>0$. 
-\item The factor $\kappa^2 - \Omega^2>0$ for our self-gravitating disc
-  models in the region where the spiral appears. 
-\item We measure $|k|\sim \pi G\Sigma/c_s^2$, so the last factor in
-  the square brackets is positive. Note that this factor is positive
-  provided $|k| < 2/HQ$ or wavelengths $\lambda > \pi Q H$.  
-\end{itemize}
-These considerations imply the factor in the square brackets is
-positive. Then, since $dc_s^2/dR < 0$, $T_\mathrm{BG}<0$ so the background
-disc applies a negative torque on the low-frequency $m=1$ spiral,
-which has negative angular momentum. This suggests the spiral
-amplitude will grow.
+However, if the linear disturbance under consideration \emph{is} a
+low-frequency $m=1$ 
+trailing spiral wave, then it has negative angular
+momentum. If $dc_s^2/dR<0$, as is typical for circumstellar or
+protoplanetary discs, then $T_\mathrm{BG}<0$, and the background disc
+applies a negative torque on the disturbance, which further decreases
+its angular momentum. This suggests the spiral amplitude will grow.
+
+%torque effective for large scales or sharp gradients
+
+%Then, since $dc_s^2/dR < 0$, $T_\mathrm{BG}<0$ so the background
+%disc applies a negative torque on the low-frequency $m=1$ spiral,
+%which has negative angular momentum. This suggests the spiral
+%amplitude will grow.
+
+\subsection{Role of disc structure}

### 2014-11-02 paper/model.tex
added analytical discussion
added variable q discussion

@@
 where $G$ is the gravitational constant,  and the disc potential
-$\Phi_d$. We assume a locally 
+$\Phi_d$. We impose a locally 
 isothermal equation of state 
@@
 where the sound-speed is given by 
-\begin{align}
+\begin{align}\label{sound-speed}
   c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
@@
 %We consider a smooth disc which contains a surface density bump
-%between $R\in[R_{d1},R_{d2}]$, mimicking mass built-up in a dead
+%between $R\in[R_{1},R_{2}]$, mimicking mass built-up in a dead
 %zone. 
@@
   \Sigma(R) = \Sigma_\mathrm{ref} \left(\frac{R}{R_0}\right)^{-s}\times B(R;
-  R_{d1}, R_{d2}, \epsilon, \delta R_d), 
+  R_{1}, R_{2}, \epsilon, \delta R_d), 
 \end{align}
@@
 surface density scale chosen by specifying $Q_\mathrm{out}$,
-the Keplerian Toomre parameter at $R=R_{d2}$,
+the Keplerian Toomre parameter at $R=R_{2}$,
 \begin{align}
   Q_\mathrm{out} = \left.\frac{c_s\Omega_k}{\pi G
-    \Sigma}\right|_{R=R_{d2}}. 
+    \Sigma}\right|_{R=R_{2}}. 
 \end{align}
@@
 $B(R)$ represents a surface density boost between
-$R\in[R_{d1},R_{d2}]$ by a factor $\epsilon^{-1}>1$,
+$R\in[R_{1},R_{2}]$ by a factor $\epsilon^{-1}>1$,
 and $\delta R_d$ is the transition width between the bump and the
@@
   &f_1(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 +
-    \tanh\left(\frac{R-R_{d1}}{\Delta_1}\right)\right]  + \epsilon,\\
+    \tanh\left(\frac{R-R_{1}}{\Delta_1}\right)\right]  + \epsilon,\\
   &f_2(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 -
-    \tanh\left(\frac{R-R_{d2}}{\Delta_2}\right)\right]  + \epsilon,
+    \tanh\left(\frac{R-R_{2}}{\Delta_2}\right)\right]  + \epsilon,
 \end{align}
-where $\Delta_1 = \delta R_d \times H(R_{d1})$ and similarly for
-$R_{d2}$. The bump function mimicks mass built-up in a dead zone.
+where $\Delta_1 = \delta R_d \times H(R_{1})$ and similarly for
+$R_{2}$. The bump function mimicks mass built-up in a dead zone.
 
@@
 
-Our fiducial parameter values are: $s=2$, $R_{d1}=R_0$, $R_{d2}=2R_0$,
+Our fiducial parameter values are: $s=2$, $R_{1}=R_0$, $R_{2}=2R_0$,
 $\epsilon=0.1$, $\delta R_d=5$, $h=0.05$ and

### 2014-11-02 paper/results.tex
added variable q discussion

@@
-\section{Results for two-dimensional discs}
+\section{Results}
 We first present results from FARGO simulations. The 2D disc spans
@@
 disc mass $M_{d}=0.086M_*$. The mass within
-$R\in[R_\mathrm{min},R_{d1}]$ is $0.017M_*$, that within
-$R\in[R_{d1},R_{d2}]$ is $0.049M_*$, and that within
-$R\in[R_{d2},R_\mathrm{max}]$ is $0.021M_*$. We use a resolution of
+$R\in[R_\mathrm{min},R_{1}]$ is $0.017M_*$, that within
+$R\in[R_{1},R_{2}]$ is $0.049M_*$, and that within
+$R\in[R_{2},R_\mathrm{max}]$ is $0.021M_*$. We use a resolution of
 $N_R\times N_\phi = 1024\times 2048$, or about $16$ grids per $H$, and
@@
 \end{align}
-where $\delta$ is an amplitude, $\overline{R}_d = (R_{d1}+R_{d2})/2$
-and $\Delta R_d = (R_{d2}-R_{d1})/2$. We use $M=10$ and choose 
+where $\delta$ is an amplitude, $\overline{R}_d = (R_{1}+R_{2})/2$
+and $\Delta R_d = (R_{2}-R_{1})/2$. We use $M=10$ and choose 
 $\delta\in[-10^{-3},10^{-3}]$ randomly but independent of $\phi$.
@@
 Fig. \ref{fargo_modeamp} plots evolution of the maximum
-non-axisymmetric surface density amplitudes in $R\in[R_{d1},R_{d2}]$
+non-axisymmetric surface density amplitudes in $R\in[R_{1},R_{2}]$
 for $m\in[1,10]$. Snapshots from the simulation are shown in
@@
 \lesssim 0.3\Omega$ (recall $\omega = 
-m\Omega_p$) in the region it is promininent $R\in[R_{d1},R_{d2}]$. The
+m\Omega_p$) in the region it is promininent $R\in[R_{1},R_{2}]$. The
 growth rate is also slow relative to the local rotation. This gives a
@@
 computed both sides of Eq. \ref{baroclinic_torque_int} using
-simulation data, and compare them in Fig. \ref{fargo_angmom_ex}.
+simulation data, and compare them in Fig. \ref{fargo_angmom_ex}. There
+is a good match between the two torques, especially at early times
+$t\lesssim110P_0$. The match is less good later on, when the spiral
+amplitude is no longer small ($\Delta\Sigma_1\sim 0.2$ at $t=110P_0$
+and $\Delta\Sigma_1\sim 0.4$ by $t=120P_0$) and linear theory becomes
+less valid. Fig. \ref{fargo_angmom_ex} confirms that angular momentum
+is lost from the wave to the disc.
+
+%wave losses at boundaries
 
@@
   \includegraphics[width=\linewidth]{figures/m1_analysis_ang_fargo.ps} 
-  \caption{
+  \caption{Rate of change of the $m=1$ wave angular momentum as defined by
+    Eq. \ref{baroclinic_torque_int} (solid) compared to the torque
+    exerted on the wave associated with the background temperature
+    gradient (dotted). 
     \label{fargo_angmom_ex}} 
@@
 %- probably non-linear effects/shocks
+
+\subsection{Dependence on the imposed temperature gradient} 
+At this point we hypothesize the growth of the $m=1$ spiral mode is
+associated with the imposed temperature gradient. The torque
+responsible for its growth, Eq. \ref{baroclinic_torque}, is
+proportional to $q$, the power-law index used to set the sound-speed
+profile (Eq. \ref{sound-speed}). To test this, we carried out
+simulations with $q\in[0,1]$ using only an $m=1$ initial 
+perturbation. However, to keep the Toomre profiles similar, we adjust
+the surface density power-law index such that $s = (3+q)/2$ for each
+case presented below. %should not be important, local effects
+
+Fig. \ref{fargo_varq} compares the $m=1$ spiral amplitudes as a
+function of $q$. We indeed observe slower growth with decreasing
+$q$. Although the figure indicates growth for the strictly isothermal
+disc ($q=0$), we did not observe a coherent one-arm spiral upon
+inspection of the $m=1$ surface density field. The growth in this case
+may be associated with higher $m$ modes. %"non-linear effects"
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_fargo_varq.ps}   
+  \caption{Evolution of the $m=1$ spiral amplitude as a function of
+    the imposed sound-speed gradient $q$. The maximum value of the
+    $m=1$ surface density in $R\in[R_1,R_2]$ is shown. 
+    \label{fargo_varq}} 
+\end{figure}
+
+We plot growth rates of the $m=1$ spiral as a function of $q$ in 
+Fig. \ref{fargo_varq_growth}. The correlation is well-fitted by a
+linear relation
+\begin{align*}
+  \gamma = 0.014 q \Omega_k{R_0}. 
+\end{align*}
+This confirms that the temperature gradient is responsible for the
+development of the one-arm spiral observed in our simulations. 
+\begin{figure}
+  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varq.ps}    
+  \caption{Growth rates of the $m=1$ spiral mode as a function of the
+    imposed sound-speed gradient $q$ (asterisks). A linear fit is also
+    plotted (dotted line). 
+    \label{fargo_varq_growth}} 
+\end{figure}
+
+% only cases where a coherent spiral is seen is plotted

### 2014-11-02 paper/discussion.tex
updated varq figs
minor updates

@@
 \section{Discussion}
+Having confirmed the $m=1$ spiral instability through numerical
+simulations, we now use results from the simulations combined with
+linear theory to construct an explanation for their growth. For
+simplicitly, we return to 2D discs.
 
@@
 
-\subsection{Role of disc structure}
+\subsection{Role of self-gravity and disc structure} 
+The background torque density $T_{BG}$ does not depend on
+self-gravity. 
+%and by definition, disc structure plays no role in the local description of linear modes 
+However, our numerical simulations show that the $m=1$ spiral is
+confined between two $Q$-barriers, where real solutions to the local
+dispersion relation is possible (Eq. \ref{wavenumber}), as a result of the adopted disc
+structure. Thus, the role of disc structure and self-gravity is to
+allow a neutral mode to be set up and confined, which is then
+destabilised through the background torque density.
+
+%talk about additional simulations at high Q
+%spiral mode as external potential to wider disc (lp11)

### 2014-11-02 paper/model.tex
minor updates

@@
 % develop one-arm spirals in discs with radial structure. 
-The indirect potential associated with the non-inertial reference frame is 
+For the simulations presented in this paper, the indirect potential
+associated with the non-inertial reference frame is purposefully 
 neglected to avoid complications arising from the motion of the
 central star. For the adopted disc masses $M_d\lesssim 0.1 M_*$, this
-effect is not expected to be significant.
+effect is not expected to be significant \citep{adams89}. Indeed,
+simulations in the early stages of this project included the indirect
+potential, and yield similar results.
+
+%ang mom cons
 
@@
 The 2D disc equations are obtained by replacing $\rho$ with the surface
-mass density $\Sigma$ and $p$ with the vertically-integrated pressure
-$\Pi$ in the continuity and momenta equations, and the 2D fluid
+mass density $\Sigma$, $p$ becomes the vertically-integrated pressure, 
+%$\Pi$ in the continuity and momenta equations, 
+and the 2D fluid
 velocity $\bm{v}=(v_R,v_\phi)$ is evaluated at the midplane, as are

### 2014-11-02 paper/paper.tex
added analytical discussion
minor updates

@@
   the central star, as found in previous studies. Instead, we find it
-  to be associated with a structured disc, as well as an imposed
-  temperature gradient. We apply results
+  to be associated with an imposed
+  temperature gradient and the radial disc structure. 
+  We apply results
   from existing linear theory and attribute the growth of this spiral
   to the non-conservation of angular momentum for linear perturbations
-  in discs with an imposed temperature gradient. This enables angular
+  in discs with a forced temperature gradient. This enables angular
   momentum exchange between the spiral and the background disc.
@@
 \input{results}
-%\input{results_3d}
-%\input{discussion}
+\input{results_3d}
+\input{discussion}
 %\input{summary}

### 2014-11-02 paper/results.tex
minor updates

@@
 
-\subsection{Reference run with random perturbations}
-In this fiducial simulation we subject the disc to initial perturbations in
+In these simulations we subject the disc to initial perturbations in 
 cylindrical radial velocity, 
@@
   v_R \to v_R+ c_s\frac{\delta}{M}
-  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}_d}{\Delta 
-          R_d}\right)^2\right]}\sum_{m=1}^M\cos{m\phi},
+  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}}{\Delta 
+        R}\right)^2\right]}\sum_{m=1}^M\cos{m\phi},
 \end{align}
-where $\delta$ is an amplitude, $\overline{R}_d = (R_{1}+R_{2})/2$
-and $\Delta R_d = (R_{2}-R_{1})/2$. We use $M=10$ and choose 
-$\delta\in[-10^{-3},10^{-3}]$ randomly but independent of $\phi$.
+where the amplitude $\delta\in[-10^{-3},10^{-3}]$ is set randomly but
+independent of $\phi$, $\overline{R} = (R_{1}+R_{2})/2$
+and $\Delta R = (R_{2}-R_{1})/2$.
 
+\subsection{Reference run with random perturbations}
+To obtain a picture of the overall disc evolution, here we consider a
+fiducial run initialised with $M=10$ in Eq. \ref{randpert}. 
+%In this fiducial simulation we subject the disc to initial perturbations in
+%cylindrical radial velocity, 
+%\begin{align}\label{randpert}
+%  v_R \to v_R+ c_s\frac{\delta}{M}
+%  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}_d}{\Delta 
+%          R_d}\right)^2\right]}\sum_{m=1}^M\cos{m\phi},
+%\end{align}
+%where $\delta$ is an amplitude, $\overline{R}_d = (R_{1}+R_{2})/2$
+%and $\Delta R_d = (R_{2}-R_{1})/2$. We use $M=10$ and choose 
+%$\delta\in[-10^{-3},10^{-3}]$ randomly but independent of $\phi$.   
 Fig. \ref{fargo_modeamp} plots evolution of the maximum
@@
 \begin{align*}
-  \gamma = 0.014 q \Omega_k{R_0}. 
+  \gamma \simeq 0.015 q \Omega_k(R_0). 
 \end{align*}

### 2014-11-02 paper/results_3d.tex
minor updates

@@
 \section{Three-dimensional simulations}
+In this section we briefly review results from 3D simulations carried
+out using the ZEUS-MP and PLUTO codes. The main purpose is to confirm
+the physical reality of the
+
 The 3D disc has radial size 

### 2014-11-04 paper/discussion.tex
minor updates

--- 2014-11-02 paper/discussion.tex
@@
 show that the torque density acting on this mode due to the background
-temperature gradient is negative, which enforces the mode.
+temperature gradient is negative, which enforces the mode. % Note that
+% we will not be assuming a self-gravitating disc until the end of this
+% section.
 
@@
 
-%torque effective for large scales or sharp gradients
+Using $\jlin$ and $T_\mathrm{BG}$, we can obtain a rough estimate of
+the growth rate $\gamma$ of linear perturbations as  
+\begin{align}
+  2\gamma \sim \frac{T_\mathrm{BG}}{\jlin},
+\end{align}
+where the factor of two accounts for the fact that the angular momentum
+density is quadradtic in the linear perturbations. Inserting the above
+expressions for $\jlin$ and $T_\mathrm{BG}$ for $m=1$ gives
+\begin{align}
+  2\gamma \sim
+  -\frac{dc_s^2}{dR}\frac{1}{k\Sigma^2\Omega}\frac{|\delta\Sigma_1|^2}{|\xi_R
+    + \ii\xi_\phi|^2}.
+\end{align}
+Now, using $\delta\Sigma_m/\Sigma = -\ii k \xi_R - \ii m \xi_\phi/R$
+in the local approximation, we have
+\begin{align}
+  2\gamma \sim -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R +
+    \xi_\phi/kR|^2}{|\xi_R + \ii \xi_\phi|^2}
+  \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
+\end{align}
+where we used $dc_s^2/dR <0$, $k>0$ and $|kR|\gg 1$ to obtain the
+inequality. For $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ adopted in our disc
+models,
+\begin{align}
+  2\gamma \lesssim q\frac{c_s^2}{R}\frac{k}{\Omega}. 
+\end{align}
+Now we consider perturbations in a self-gravitating disc with characteristic
+wavenumber $k \sim \pi G \Sigma/c_s^2 \sim \Omega/Q c_s$ (as seen in
+our simulations). Then
+\begin{align}
+  \gamma \lesssim \frac{qh}{2Q}\Omega,
+\end{align}
+where we used $c_s/R\sim h\Omega$. For our fiducial disc models with
+$q=1$, $Q=O(1)$, $h=0.05$, this gives $\gamma/\Omega = O(10^{-2})$,
+consistent with simulation results.
 
-%Then, since $dc_s^2/dR < 0$, $T_\mathrm{BG}<0$ so the background
-%disc applies a negative torque on the low-frequency $m=1$ spiral,
-%which has negative angular momentum. This suggests the spiral
-%amplitude will grow.
+Of course, we have been assuming a neutral mode to obtain expressions
+for $\jlin$ and $T_\mathrm{BG}$ in the first place, so the above can only be
+considered as an order-of-magnitude estimate. Nevertheless, the
+expected growth rate is in reasonable agreement with numerical
+results.
 
-\subsection{Role of disc structure}
+% \subsection{Role of self-gravity and disc structure} 
+% The background torque density $T_{BG}$ does not depend on
+% self-gravity. 
+% %and by definition, disc structure plays no role in the local description of linear modes 
+% However, our numerical simulations show that the $m=1$ spiral is
+% confined between two $Q$-barriers, where real solutions to the local
+% dispersion relation is possible (Eq. \ref{wavenumber}), as a result of the adopted disc
+% structure. Thus, the role of disc structure and self-gravity is to
+% allow a neutral mode to be set up and confined, which is then
+% destabilised through the background torque density.
+
+%talk about additional simulations at high Q
+%spiral mode as external potential to wider disc (lp11)

### 2014-11-04 paper/model.tex
minor updates

@@
 \begin{align}\label{sound-speed}
-  c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
+%  c_s(R) = hR_0\Omega_k(R_0)\left(\frac{R}{R_0}\right)^{-q/2}, 
+  c_s(R) = c_{s0}\left(\frac{R}{R_0}\right)^{-q/2}
 \end{align}
-where $h$ is the disc aspect-ratio $R=R_0$, 
+where $c_{s0} = h R_0\Omega_k(R_0)$ and 
+$h$ is the disc aspect-ratio and $R=R_0$, 
 $\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and

### 2014-11-04 paper/results_3d.tex
minor updates

@@
 \section{Three-dimensional simulations}
-In this section we briefly review results from 3D simulations carried
-out using the ZEUS-MP and PLUTO codes. The main purpose is to confirm
-the physical reality of the
+In this section we review results from 3D simulations carried
+out using the ZEUS-MP and PLUTO codes.
 
@@
 Fig. \ref{3d_prelim} for both ZEUS-MP and PLUTO. We again find the dead zone eventually
-dominated by an $m=1$ spiral. A Cartesian visualisation of the
-midplane density perturbation associated with the $m=1$ mode is shown in Fig. \ref{pluto_cart}.
+dominated by an $m=1$ spiral.%  A Cartesian visualisation of the
+% midplane density perturbation associated with the $m=1$ mode is shown in Fig. \ref{pluto_cart}.
 The two codes give qualitatively similar results, but a longer PLUTO simulation was
@@
 
-\begin{figure}
-  \includegraphics[width=\linewidth]{figures/pdiskxy_020_m1}
-  \caption{Cartesian visualisation the $m=1$ midplane density in the PLUTO
-    simulation shown in Fig. \ref{3d_prelim}.\label{pluto_cart}}   
-\end{figure}
+% \begin{figure}
+%   \includegraphics[width=\linewidth]{figures/pdiskxy_020_m1}
+%   \caption{Cartesian visualisation the $m=1$ midplane density in the PLUTO
+%     simulation shown in Fig. \ref{3d_prelim}.\label{pluto_cart}}   
+% \end{figure}
 

### 2014-11-05 paper/appendix.tex
appendix and minor updates

@@
-\section{Modification to vertical structure by self-gravity}\label{vertsg_mod}
-We describe a simple procedure to set up the vertical structure of a locally
-isothermal, self-gravitating disc. We imagine setting up a non-self-gravitating disc,
-then slowly switch on the vertical force due to self-gravity. We expect the midplane
-density to increase at the expense of gas density higher in the atmosphere. It is assumed
-that the temperature profile remains unchanged.
+\section{The background torque density in a three-dimensional disc with
+  a fixed temperature profile}\label{tbg_deriv}
+We give a brief derivation of the angular momentum exchange between
+linear perturbations and the background disc. We consider a
+three-dimensional disc in which the equilibrium pressure and density
+are related by 
+\begin{align}\label{iso_cond}
+  p = c_s^2(R,z)f(\rho),
+\end{align} 
+where $f(\rho)$ is an arbitrary function of $\rho$ with dimensions of
+mass per unit volume, and $c_s$ is a prescribed function of
+position with dimensions of velocity squared. The equilibrium disc
+satisfies 
+\begin{align}
+  R\Omega^2(R,z) &= \frac{1}{\rho}\frac{\p p}{d R} +
+  \frac{\p\Phi_\mathrm{tot}}{\p R}, \\
+  0 &= \frac{1}{\rho}\frac{\p p}{\p z} + \frac{\p\Phi_\mathrm{tot}}{\p
+    z}. 
+\end{align}
+Note that, in general, the equilibrium rotation $\Omega$ depends on
+$R$ and $z$.
 
-Vertical hydrostatic equilibrium between
-gas pressure, stellar gravity and self-gravity reads
+We begin with the linearised equation of motion in terms of the
+Lagragian displacement $\bm{\xi}$ as given by \cite{lin93b} but with an
+additional potential perturbation, 
+\begin{align}\label{lagragian_pert}
+  &\frac{D^2\bm{\xi}}{Dt^2} +
+  2\Omega\hat{\bm{z}}\times\frac{D\bm{\xi}}{Dt}  \notag \\ &= -
+  \frac{\nabla \delta p }{\rho} + \frac{\delta\rho}{\rho^2}\nabla\rho  
+  -\nabla\delta\Phi_d - R
+  \hat{\bm{R}}\left(\bm{\xi}\cdot\nabla\Omega^2\right) \notag \\
+  & = -\nabla\left(\frac{\delta p}{\rho} + \delta \Phi_d\right) -
+  \frac{\delta p}{\rho}\frac{\nabla\rho}{\rho} +
+  \frac{\delta\rho}{\rho}\frac{\nabla p}{\rho} - - R
+  \hat{\bm{R}}\left(\bm{\xi}\cdot\nabla\Omega^2\right),
+\end{align}
+where $D/Dt \equiv \p_t + \ii m \Omega$ for perturbations with
+azimuthal dependence in the form $\exp\left(\ii m \phi\right)$, and  
+the $\delta$ quantities denote Eulerian perturbations.
+
+As explained in \cite{lin11b}, a conservation law for the angular
+momentum of the perturbation may be obtained by taking the dot product
+between Eq. \ref{lagragian_pert} and $(-m/2)\rho\bm{\xi}^*$, then
+taking the imaginary part afterwards. The left hand side bcomes the
+angluar momentum density. The first term on the right hand side (RHS)
+becomes 
+\begin{align}\label{angflux1}
+  &-\frac{m}{2}\imag\left[-\rho\bm{\xi}^*\cdot\nabla\left(\frac{\delta p}{\rho} + \delta
+    \Phi_d\right)\right] \notag\\ 
+&= \frac{m}{2}\imag\left\{\nabla\cdot\left[\rho\bm{\xi}^*\left(\frac{\delta p}{\rho} + \delta
+    \Phi_d \right) + \frac{1}{4\pi
+    G}\delta\Phi_d\nabla\delta\Phi_d^*\right]\right\} \notag\\
+&+ \frac{m}{2}\imag\left(\delta\rho^*\frac{\delta p}{\rho}\right),
+\end{align}
+where $\delta\rho = - \nabla\cdot\left(\rho\bm{\xi}\right)$ and
+$\nabla^2\delta\Phi_d = 4\pi G \delta \rho$ have been used. The first
+term on the RHS of Eq. \ref{angflux1} is (minus) the 
+angular momentum flux. The second term on RHS of Eq. \ref{angflux1}, 
+together with the remaining terms on the RHS of 
+Eq. \ref{lagragian_pert} constitutes the background torque. That is, 
+\begin{align}\label{tbg_3d}
+  T_\mathrm{BG} = \frac{m}{2}\imag\left[
+    \frac{\delta p}{\rho} \Delta\rho^* -
+%\left(\delta\rho^* + \bm{\xi}^*\cdot\nabla\rho\right) - 
+    \frac{\delta\rho}{\rho}\bm{\xi}^*\cdot\nabla p  
+    + \rho \xi_R^*\xi_z\frac{\p\left(R\Omega^2\right)}{\p z}\right],
+\end{align}
+where $\Delta\rho = \delta\rho + \bm{\xi}\cdot\nabla\rho$ is the
+Lagragian density perturbation.
+
+So far we have not invoked an energy equation. For 
+adiabatic perturbations $T_\mathrm{BG}$ is zero
+\citep{lin93b}. However, if we impose the equilibrium relation
+Eq. \ref{iso_cond} to hold in the perturbed state, then
 \begin{align}
-c_\mathrm{iso}^2(R)\frac{\p\ln{\rho}}{\p z} = -\frac{\p\Phi_*}{\p z} - \frac{\p\Phi}{\p z}. 
+  \delta p = c_s^2(R,z) f^\prime \delta\rho,
 \end{align}
-Assuming a smooth radial density profile, we use the plane-parallel atmosphere
-approximation for the disc potential, i.e.
+where $f^\prime = df/d\rho$. Inserting this into Eq. \ref{tbg_3d}, we
+obtain
+\begin{align}\label{tbg_3d_2}
+  T_\mathrm{BG} = -\frac{m}{2}\frac{p}{\rho
+    c_s^2}\imag\left[\delta\rho\bm{\xi}^*\cdot\nabla c_s^2 +
+    \xi_R^*\xi_z \left(\frac{\p\rho}{\p z}\frac{\p c_s^2}{\p R} -
+      \frac{\p\rho}{\p R}\frac{\p c_s^2}{\p z}\right)\right],
+\end{align}
+where the equilibrium equations were used. 
+At this point setting $\xi_z=0$ gives $T_\mathrm{BG}$ for
+perturbations with no vertical motion, 
 \begin{align}
-\frac{\p^2\Phi}{\p z^2} = 4\pi G \rho. 
+  T_\mathrm{BG,2D} = -\frac{m}{2}\frac{p}{\rho
+    c_s^2}\imag\left(\delta\rho \xi_R^*  \p_R c_s^2 \right),
 \end{align}
-Next, we write the density field as
-\begin{align}
-\rho(R,z) = \rho_N(R, z) \times \beta(z; R)
+and is equivalent to the 2D
+expression, Eq. \ref{baroclinic_torque}, with $\delta\rho $ replaced
+by $\delta \Sigma$ and $p=c_s^2\rho$. % The 2D expression is convenient
+% for razor-thin discs or perturbations without vertical motion because
+% one can measure $\delta\rho$ (or $\delta \Sigma)$ directly.
+
+In fact, we can simplify Eq. \ref{tbg_3d_2} in the general case by
+using $\delta\rho = - \rho\nabla\cdot\bm{\xi} -
+\bm{\xi}\cdot\nabla\rho$, giving
+\begin{align}\label{tbg_general}
+  T_\mathrm{BG} = \frac{m}{2}\frac{p}{\rho
+    c_s^2}\imag\left[\rho\left(\nabla\cdot\bm{\xi}\right)\bm{\xi}^*\cdot\nabla
+  c_s^2\right].  
 \end{align} 
-where $\rho_N$ is the density field corresponding to the non-self-gravitating disc:
-\begin{align}
-&c_\mathrm{iso}^2(R)\frac{\p\ln{\rho_N}}{\p z} = -\frac{\p\Phi_*}{\p z},\\
-&\rho_N = \frac{\Sigma(R)}{\sqrt{2\pi}H(R)}\exp{(-z^2/2H^2)} \notag\\
-&\phantom{\rho_N}\equiv \rho_{N0}(R)\exp{(-z^2/2H^2)},
-\end{align}
-where $\rho_{N0}=\rho_N(R,z=0)$ is the midplane density. The explicit
-expression for $\rho_N$ above assumes a thin disc. The function
-$\beta$ describes the modification to the local density in order to be
-consistent with self-gravity. By construction, its governing equation
-is 
-\begin{align}
-c_\mathrm{iso}^2(R)\frac{\p^2\ln{\beta}}{\p z^2} = -4\pi G\rho_N\beta. 
-\end{align}
-Let 
-\begin{align}
-& \chi  = \ln\beta - z^2/2H^2 ,\\
-& \xi= \left(\frac{4\pi G \rho_{N0} }{c_\mathrm{iso}^2}\right)^{1/2} z,
-\end{align}
-then the governing equation can be written in dimensionless form
-\begin{align}
-&\frac{\p^2 \chi}{\p\xi^2} = -K -\exp{\chi}\label{vertsg_eqn},\\
-& K \equiv \frac{c_\mathrm{iso}^2}{4\pi G \rho_{N0} H^2}. \notag
-\end{align}
-Note that $K(R)$ is proportional to the local Keplerian Toomre
-parameter. Eq. \ref{vertsg_eqn} can be further reduced to a first
-order differential equation, but this is unnecessary because we
-pursue a numerical solution at the end. Appropriate 
-boundary conditions are
-\begin{align}
-&\chi(z=0)  =  \ln\beta_0, \\
-&\left.\frac{\p \chi}{\p \xi} \right|_{z=0} = 0.
-\end{align}
-$\beta_0(R)$ is the midplane density enhancement. To determine its value, we impose
-the surface density before and after modification by self-gravity to
-remain the same. Then we require 
-\begin{align}\label{beta0_eqn}
-&F(\beta_0) \equiv \sqrt{\frac{2K}{\pi}}\int_0^{ n /\sqrt{K} } 
-\exp{\chi(\xi;\beta_0)}d\xi - \erf{\left(\frac{n}{\sqrt{2}}\right)} \notag\\
-&\phantom{F(\beta_0)} = 0,  
-\end{align}
-where $n$ is the number of scale-heights of the non-self-gravitating
-disc we originally considered. At a given cylindrical radius $R$, we
-solve Eq. \ref{beta0_eqn} using Newton-Raphson iteration. Each
-iteration involves integrating the governing ODE for $\chi$
-(Eq. \ref{vertsg_eqn}). At the end of the iteration, we have $\beta(z;
-R)$ and the midplane enhancement $\beta_0(R)$.
+For a barotropic fluid $p=p(\rho)$, the function $c_s^2$ can be
+taken as constant (Eq. \ref{iso_cond}) for which $T_\mathrm{BG}$
+vanishes. When there is a forced temperature gradient,
+Eq. \ref{tbg_general} indicates there is angular momentum exchange
+between compressible perturbations ($\nabla\cdot\bm{\xi}\neq0$) and
+the background disc if there is motion along the temperature
+gradient ($\bm{\xi}\cdot\nabla c_s^2 \neq 0$).
 
-We comment that the procedure outlined above can be extended to
-polytropic discs. In this case, there is an additional unknown --- 
-the new disc thickness after adjustment by self-gravity and an
-additional constraint --- the density should vanish at the new disc
-surface.
+% total effect needs to be integrated
+
+% For the perturbations without vertical motion or those in razor-thin
+% discs, the expression Eq. \ref{}
+
+%eq baroclinic torque is more practical in 2d case because \delta
+%sigma can be measured direction from simulations (xi needs to be
+%calculated from velocity field, knowing the eigenfrequency)

### 2014-11-05 paper/discussion.tex
appendix and minor updates

@@
 \begin{align}
-  T_\mathrm{BG} &= -\frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\Sigma_m \left(\frac{\delta
+  T_\mathrm{BG} &= -\frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\delta\Sigma_m \left(\frac{\delta
         v_{Rm}}{-\ii\sbar}\right)^*\right] \notag\\
-  & = \frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\ii \Sigma_m \left(\frac{\delta
+  & = \frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\ii \delta\Sigma_m \left(\frac{\delta
         v_{Rm}}{\sbar}\right)^*\right]
@@
 models,
-\begin{align}
+\begin{align}\label{theoretical_rate}
   2\gamma \lesssim q\frac{c_s^2}{R}\frac{k}{\Omega}. 
@@
 
-% \subsection{Role of self-gravity and disc structure} 
-% The background torque density $T_{BG}$ does not depend on
-% self-gravity. 
-% %and by definition, disc structure plays no role in the local description of linear modes 
-% However, our numerical simulations show that the $m=1$ spiral is
-% confined between two $Q$-barriers, where real solutions to the local
-% dispersion relation is possible (Eq. \ref{wavenumber}), as a result of the adopted disc
-% structure. Thus, the role of disc structure and self-gravity is to
-% allow a neutral mode to be set up and confined, which is then
-% destabilised through the background torque density.
+\subsection{Role of self-gravity and disc structure} 
+We note that the background torque density $T_\mathrm{BG}$ does not depend on  
+self-gravity, and we only considered a self-gravitating perturbation to
+obtain an appropriate wavenumber $k$ needed to evaluate
+Eq. \ref{theoretical_rate}. Thus, self-gravity is not directly
+responsible for instability. However, 
+our numerical simulations show that the $m=1$ spiral is
+confined between two $Q$-barriers, where real solutions to the local
+dispersion relation is possible (Eq. \ref{wavenumber}), as a result of the adopted disc
+structure. Thus, the role of disc structure and self-gravity is to
+allow a neutral mode to be set up and confined, which is then
+destabilised through the background torque density.
 
@@
 %spiral mode as external potential to wider disc (lp11)
+%interaction across co-rotation

### 2014-11-05 paper/model.tex
appendix and minor updates

@@
 $\tbg$ vanishes and the angular momentum associated with
-the perturbation is conserved, provided boundary fluxes vanish.
+the perturbation is conserved, provided that there is no net angular
+momeumtum flux. We briefly outline the derivation of $T_\mathrm{BG}$
+in Appendix 
+\ref{tbg_deriv}.
 

### 2014-11-05 paper/paper.tex
appendix and minor updates

@@
 \begin{abstract} 
-  We present numerical simulations of gravitational instability in
-  astrophysical discs with an an imposed temperature profile, and in
-  which only a sub-portion of it is self-gravitating. We conduct
-  numerical hydrodynamic experiments in 2D using the FARGO code and in
-  3D with the ZEUS-MP and PLUTO codes. In such systems we observe the
-  development of a 
-  spiral mode with azimuthal wavenumber $m=1$ with co-rotation radius well
+  We present numerical simulations of astrophysical discs with an 
+  imposed temperature profile and in which a sub-portion of it is
+  self-gravitating. This might represent, for example, mass  
+  built-up in a circumstellar or protoplanetary
+  disc at the outer edge of a `dead zone', in which there
+  is inefficient magneto-hydrodynamic turbulence to enable accretion. 
+  We conduct numerical gravito-hydrodynamic experiments in 2D using
+  the FARGO code and in 3D with the ZEUS-MP and PLUTO codes. In such
+  systems we observe the development of a 
+  spiral mode with azimuthal wavenumber $m=1$ with co-rotation radius 
   outside the bulk of the spiral arm, resulting in a nearly stationary 
@@
   the central star, as found in previous studies. Instead, we find it 
-  to be associated with an imposed
+  to be associated with the imposed 
   temperature gradient and the radial disc structure. 
-  We apply results
-  from existing linear theory and attribute the growth of this spiral
-  to the non-conservation of angular momentum for linear perturbations
-  in discs with a forced temperature gradient. This enables angular
-  momentum exchange between the spiral and the background disc.
+  We apply results from existing linear theory and attribute the
+  growth of this spiral to the non-conservation of angular momentum
+  for linear perturbations in discs with a forced temperature
+  gradient. This enables angular momentum exchange between the spiral
+  and the background disc, and may lead to instability if the  
+  interaction increases the amplitude of the linear perturbations.
 
@@
   % clarified. 
-
   %
@@
 \appendix
-%\input{appendix}
+\input{appendix}
 %\input{appendix2}

### 2014-11-05 paper/results_3d.tex
appendix and minor updates

@@
 \section{Three-dimensional simulations}
-In this section we review results from 3D simulations carried
-out using the ZEUS-MP and PLUTO codes.
+In this section we briefly review results from 3D simulations carried 
+out using the ZEUS-MP and PLUTO codes. The main purpose is to check
+verify the above results with different numerical codes, and validate 
+the 2D approximation.
 
@@
 compared to 2D, we use a smooth perturbation by setting
-$\delta = 10^{-3}$ and $M=1$. This corresponds to a single $m=1$
-spiral in the dead zone, which was found to dominate the 2D
-simulation.
+$\delta = 10^{-3}$ and $M=1$ in Eq. \ref{randpert}. This corresponds
+to a single $m=1$ spiral in the dead zone, which was found to dominate
+the 2D simulation.
 
@@
 first evolve the disc without perturbations using  
-$(\lmax,\mmax)=(32,0)$ up to $t=10P_0$, during which 
+$(\lmax,\mmax)=(32,0)$ up to $t=100P_0$, during which 
 meridional velocities are damped out. We then restart the simulation
@@
 Fig. \ref{3d_prelim} for both ZEUS-MP and PLUTO. We again find the dead zone eventually
-dominated by an $m=1$ spiral.%  A Cartesian visualisation of the
+dominated by an $m=1$ spiral. %A Cartesian visualisation of the
 % midplane density perturbation associated with the $m=1$ mode is shown in Fig. \ref{pluto_cart}.
 The two codes give qualitatively similar results, but a longer PLUTO simulation was
-needed to achieve similar mode amplitudes than ZEUS-MP. 
-The amplitude of the spiral is smaller than the 2D case above,
-possibily due to lower resolution and/or 3D self-gravity being weaker
-than that in 2D.
+needed to achieve similar mode amplitudes in ZEUS-MP.
+
+%The amplitude of the spiral is smaller than the 2D case above,
+%possibily due to lower resolution and/or 3D self-gravity being weaker
+%than that in 2D.
 

### 2014-11-05 paper/_region_.tex
minor updates

--- 2014-11-05 paper/paper.tex
@@
-\documentclass[useAMS,usenatbib]{mn2e}
+\message{ !name(paper.tex)}\documentclass[useAMS,usenatbib]{mn2e}
 \usepackage{epsfig,rotate,graphicx}
@@
 
+\message{ !name(paper.tex) !offset(-3) }
+
 \maketitle
@@
   for linear perturbations in discs with a forced temperature
-  gradient. This enables angular momentum exchange between the spiral
-  and the background disc, and may lead to instability if the  
-  interaction increases the amplitude of the linear perturbations.
+  gradient. This may lead to an unstable exchange of angular momentum between the spiral 
+  and the background disc.
 
@@
 \end{document}
+
+\message{ !name(paper.tex) !offset(-143) }

### 2014-11-05 paper/discussion.tex
minor updates

@@
 dispersion relation is possible (Eq. \ref{wavenumber}), as a result of the adopted disc
-structure. Thus, the role of disc structure and self-gravity is to
+structure. Thus, the main role of disc structure and self-gravity is to
 allow a neutral mode to be set up and confined, which is then
@@
 %interaction across co-rotation
+
+\subsubsection{The confined spiral as an external potential}
+One possible effect of self-gravity is to allow the one-arm spiral 
+to act as an external potential for the wider disc. This effect is
+analogous to disc-satellite interaction
+\citep{goldreich79}, and may lead to instability 
+if the angular momentum associated with the spiral has the opposite 
+sign to that of the density waves it induces in the exterior disc
+\citep{lin11a}.
+
+Considering the spiral as a external potential (c.f. an embedded
+satellite), we expect it to  exert a torque on the disc at Lindblad
+and co-rotation resonances.
+
+%speculations - long term balance. forced temp gradient drives
+%instability (disc gains ang mom, spiral loses ang mom). but if spiral
+%shocks and dissiple, it dumps negative angular momentum (disc gains
+%negative ang mom). balance between gain from forced temp gradient and
+%loss due to dissiplatoin

### 2014-11-05 paper/results.tex
minor updates

@@
 
-\subsection{Dependence on the imposed temperature gradient} 
+\subsection{Dependence on the imposed temperature profile} 
 At this point we hypothesize the growth of the $m=1$ spiral mode is
@@
 % only cases where a coherent spiral is seen is plotted
+
+We also performed a series of simulations with aspect-ratio
+$h\in[0.03,0.07]$ with fixed $q=1$. This also affects the temperature
+gradient since $c_s \propto h$. However, with other parameters equal
+to that in the fiducial simulation,  varying $h$ also changes the disc
+mass. For $h\in[0.03,0.07]$ the total disc mass ranges from
+$M_d=0.052M_*$ to $M_d=0.12M_*$ and the 
+mass within $R\in[R_1,R_2]$  ranges from $0.033M_*$ to $0.062M_*$.
+
+Fig. \ref{fargo_varh_growth} shows the growth rates of the $m=1$
+spiral in $R\in[R_1,R_2]$ as a function of $h$. Growth rates increases
+with $h$, roughly as 
+\begin{align}
+  \gamma \simeq 0.10h\Omega_k(R_0).  
+\end{align}
+However, the linear fit is less good than for variable $q$ above. This
+may be due to the change in the total disc mass when $h$ changes. We
+find no qualitative difference between the spiral pattern that
+emerges.
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varh.ps}    
+  \caption{Growth rates of the $m=1$ spiral mode as a function of the
+    disc aspect-ratio $h$ (asterisks). A linear fit is also
+    plotted (dotted line). 
+    \label{fargo_varh_growth}} 
+\end{figure}

### 2014-11-05 paper/results_3d.tex
minor updates

@@
 In this section we briefly review results from 3D simulations carried 
-out using the ZEUS-MP and PLUTO codes. The main purpose is to check
-verify the above results with different numerical codes, and validate 
+out using the ZEUS-MP and PLUTO. The main purpose is to check 
+the above results with different numerical codes, and validate  
 the 2D approximation.

### 2014-11-05 paper/model.tex
backup

@@
   \Sigma(R) = \Sigma_\mathrm{ref} \left(\frac{R}{R_0}\right)^{-s}\times B(R;
-  R_{1}, R_{2}, \epsilon, \delta R_d), 
+  R_{1}, R_{2}, \epsilon, \delta R), 
 \end{align}
@@
 $R\in[R_{1},R_{2}]$ by a factor $\epsilon^{-1}>1$,
-and $\delta R_d$ is the transition width between the bump and the
+and $\delta R$ is the transition width between the bump and the
 smooth disc. We choose 
@@
 \end{align}
-where $\Delta_1 = \delta R_d \times H(R_{1})$ and similarly for
+where $\Delta_1 = \delta R \times H(R_{1})$ and similarly for
 $R_{2}$. The bump function mimicks mass built-up in a dead zone.
@@
 Our fiducial parameter values are: $s=2$, $R_{1}=R_0$, $R_{2}=2R_0$,
-$\epsilon=0.1$, $\delta R_d=5$, $h=0.05$ and
+$\epsilon=0.1$, $\delta R=5$, $h=0.05$ and
 $Q_\mathrm{out}=2$. An example of the initial surface density and the

### 2014-11-06 paper/appendix.tex
update

@@
 \begin{align}\label{iso_cond}
-  p = c_s^2(R,z)f(\rho),
+  p = c_s^2(R,z)F(\rho),
 \end{align} 
-where $f(\rho)$ is an arbitrary function of $\rho$ with dimensions of
+where $F(\rho)$ is an arbitrary function of $\rho$ with dimensions of
 mass per unit volume, and $c_s$ is a prescribed function of
@@
 \begin{align}
-  \delta p = c_s^2(R,z) f^\prime \delta\rho,
+  \delta p = c_s^2(R,z) F^\prime \delta\rho,
 \end{align}
-where $f^\prime = df/d\rho$. Inserting this into Eq. \ref{tbg_3d}, we
+where $F^\prime = dF/d\rho$. Inserting this into Eq. \ref{tbg_3d}, we
 obtain

### 2014-11-06 paper/discussion.tex
backup
update

@@
 Of course, we have been assuming a neutral mode to obtain expressions
-for $\jlin$ and $T_\mathrm{BG}$ in the first place, so the above can only be
-considered as an order-of-magnitude estimate. Nevertheless, the
-expected growth rate is in reasonable agreement with numerical
-results.
+for $\jlin$ and $T_\mathrm{BG}$ in the first place, so
+Eq. \ref{theoretical_rate} is only an order-of-magnitude
+constraint. Nevertheless, the xpected growth rate is in reasonable
+agreement with numerical results.
+
+%also total torque is integral -> may get cancelations
 
@@
 sign to that of the density waves it induces in the exterior disc
-\citep{lin11a}.
+\citep{lin11b}.
 
@@
 
+%required disc structure (for confined SG spiral)
+
+\subsection{Speculations}
 %speculations - long term balance. forced temp gradient drives
@@
 %loss due to dissiplatoin
+
+%other source of m=1 disturbance -> eccentric modes (adams, PJ06,
+%hopkins, tremaine)

### 2014-11-06 paper/model.tex
update

@@
 \section{Governing equations and numerical models}\label{model} 
-The system of interest is a self-gravitating, inviscid fluid disc
-orbting a central star of mass $M_*$. We will mainly consider two-dimensional
+We consider a self-gravitating, inviscid fluid disc of mass $M_d$ 
+orbting a central star of mass $M_*$. We will mainly examine two-dimensional
 (2D, or razor-thin) discs in favour of numerical 
 resolution, but have also carried out some three-dimensional (3D) disc
-simulations. Hence, for generality we describe the system in
-3D. We use both cylindrical $(R,\phi,z)$ and spherical polar
+simulations. Thus, for generality we describe the system in
+3D, using both cylindrical $(R,\phi,z)$ and spherical polar
 coordinates $(r,\theta,\phi)$ centred on the star. 
@@
 where $G$ is the gravitational constant,  and the disc potential
-$\Phi_d$. We impose a locally 
-isothermal equation of state 
+$\Phi_d$. We impose a locally isothermal equation of state 
 \begin{align}
@@
 where $c_{s0} = h R_0\Omega_k(R_0)$ and 
-$h$ is the disc aspect-ratio and $R=R_0$, 
+$h$ is the disc aspect-ratio at $R=R_0$, 
 $\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
-$q$ is the imposed temperature gradient. A strictly isothermal disc
-corresponds to $q=0$, and a disc with constant aspect-ratio
-corresponds to $q=1$.
+$q$ is the imposed temperature gradient (since, for an ideal gas the
+temperature is proportional to $c_s^2$). The vertical disc scale-height is
+defined by $H=c_s/\Omega_k$. Thus, a strictly isothermal disc with
+$q=0$ has $H\propto R^{3/2}$, and $q=1$ corresponds to a 
+a disc with constant $H/R$.
 
-% We have purposefully adopted the simplest model that was found to
-% develop one-arm spirals in discs with radial structure. 
 For the simulations presented in this paper, the indirect potential
@@
 central star. For the adopted disc masses $M_d\lesssim 0.1 M_*$, this
-effect is not expected to be significant \citep{adams89}. Indeed,
+effect is not expected to be significant \citep{adams89,shu90}. Indeed,
 simulations in the early stages of this project included the indirect
@@
 
-%ang mom cons
-
+% ang mom cons
 % The locally isothermal
@@
 % theory for results interpretation. 
-
-%This allows us to consider
+% This allows us to consider
 %an isolated disc.
@@
 mass density $\Sigma$, $p$ becomes the vertically-integrated pressure, 
-%$\Pi$ in the continuity and momenta equations, 
 and the 2D fluid
@@
 \subsection{Disc model and initial conditions}
-%We consider a smooth disc which contains a surface density bump
-%between $R\in[R_{1},R_{2}]$, mimicking mass built-up in a dead
-%zone. 
-We adopt a modified power-law disc with surface
-density profile 
+For the initial disc profile we adopt a modified power-law disc with
+surface density given by
 \begin{align}
@@
 \end{align}
-
 The bump function
@@
 where $\Delta_1 = \delta R \times H(R_{1})$ and similarly for
-$R_{2}$. The bump function mimicks mass built-up in a dead zone.
+$R_{2}$. The bump function mimicks mass built-up at some location in 
+the disc.
 
@@
 where $Z(R,z)$ describes vertical stratification. In practice, we
-numerically solve for $Z(R,z)$ by neglecting radial self-gravity
+numerically solve for $Z(R,z)$ by neglecting the radial self-gravity
 force compared to vertical self-gravity, which reduces the equations
@@
 $Q_\mathrm{out}=2$. An example of the initial surface density and the
-Toomre parameter (see\S\ref{wkb}) is shown in Fig. \ref{initial_surf}. 
-Initially there is no vertical or radial velocity
+Toomre parameter (see\S\ref{wkb}) is shown in
+Fig. \ref{initial_surf}. The transition from the self-gravitating and
+non-self-gravitating portions of the disc occur smoothly across
+$\sim10H$. Initially there is no vertical or radial velocity
 ($v_R = v_r = v_\theta = 0$). The azimuthal velocity is initialized to
@@
 \end{align}
-and similarly in 2D. The angular velocity is $\Omega = v_\phi/R$.
+and similarly in 2D. The angular velocity is $\Omega \equiv v_\phi/R$.
 
@@
 \section{Numerical methods}
-%We will mostly employ 2D simulations in favour of resolution, but have
-%also run some cases in 3D. We describe here the numerical setup used
-%in each code. 
-%introduce notation, terminlogy for each code
-
 We adopt computational units such that $G=M_*=R_0=1$. Time is measured
@@
 \subsection{FARGO}
-Our primary code is FARGO with self-gravity \citep{baruteau08}. This
+Our primary tool is FARGO with self-gravity \citep{baruteau08}. This
 is a popular, simple finite-difference code for 2D discs. `FARGO' refers
 to its azimuthal transport algorithm, which circumvents time-step
-contraint set by the inner disc boundary. 
-The 2D Poisson equation is solved in integral form 
-\begin{align}\label{2d_grav}
-  \Phi_{d,z=0}(R,\phi) = - \int
-  \frac{G\Sigma(R^\prime,\phi^\prime)R^\prime dR^\prime d\phi^\prime}{\sqrt{R^2+R^{\prime 2} -
-      2RR^\prime\cos{(\phi - \phi^\prime)} + \epsilon_g^2}} 
-\end{align}
-using Fast Fouier Transform (FFT), where $\epsilon_g$ is a softening
-length to prevent a numerical singularity and the integration is taken
-over the disc. The FFT approach requires
-$\epsilon_g\propto R$ \citep{baruteau08}.
+contraint set by the inner disc boundary \citep{masset00a,masset00b}.
 
@@
 
-%poisson kernal integration, softening 
-%where used?
-%BC
+The 2D Poisson equation is solved in integral form, 
+\begin{align}\label{2d_grav}
+  &\Phi_{d,z=0}(R,\phi) \notag \\
+  &=-\int_{R_\mathrm{min}}^{R_\mathrm{max}} \int_0^{2\pi}
+  \frac{G\Sigma(R^\prime,\phi^\prime)R^\prime dR^\prime d\phi^\prime}{\sqrt{R^2+R^{\prime 2} -
+      2RR^\prime\cos{(\phi - \phi^\prime)} + \epsilon_g^2}}, 
+\end{align}
+using Fast Fouier Transform (FFT), where $\epsilon_g$ is a softening
+length to prevent a numerical singularity. The FFT approach requires
+$\epsilon_g\propto R$ \citep{baruteau08}. In FARGO, $\epsilon_g$ is
+set to a fraction of $hR$.
+
+% poisson kernal integration, softening 
+% where used?
+% BC
 \subsection{ZEUS-MP}
@@
 $\phi\in[0,2\pi]$. The vertical domain is chosen to cover $n_H$
-scale-heights, i.e. $\tan{(\pi/2 - \theta_\mathrm{min})}/h=n_H$. 
+scale-heights at $R=R_0$, i.e. $\tan{(\pi/2 - \theta_\mathrm{min})}/h=n_H$. 
 The grid is logarithmically spaced in radius and uniformly spaced in the angular
@@
 ZEUS-MP solves the 3D Poisson equation using a conjugate gradient
-method. To suppliment boundary conditions for the linear solver, we
-expand the boundary potential using in spherical harmonics $Y_{lm}$ 
+method. To supply boundary conditions to the linear solver, we
+expand the boundary potential in spherical harmonics $Y_{lm}$ 
 as described in \cite{boss80}. The expansion is truncated at
-$\lmax,\mmax$. This code was used in \cite{lin12b} for
+$(l,m)=(\lmax,\mmax)$. This code was used in \cite{lin12b} for
 self-gravitating disc-planet simulations.
@@
 \subsubsection{Angular momentum decomposition}
-The total disc angular momentum in $R\in[R_1,R_2]$ is
+The total disc angular momentum is
 \begin{align}
-  J &= \int_{R_1}^{R_2}\int_0^{2\pi} \Sigma Rv_\phi RdRd\phi \notag\\
-  &= 2\pi\int_{R_1}^{R_2}R\Sigma_0 v_{\phi0} R dR \notag\\ 
+  J &= \int_{R_\mathrm{min}}^{R_\mathrm{max}}\int_0^{2\pi} \Sigma Rv_\phi RdRd\phi \notag\\
+  &= 2\pi\int_{R_\mathrm{min}}^{R_\mathrm{max}} R\Sigma_0 v_{\phi0} R dR \notag\\ 
   &\phantom{=}+
-  \sum_{m=1}^\infty2\pi\int_{R_1}^{R_2} 2R\real\left[\Sigma_m v_{\phi
-      m}^*\right] RdR, \notag\\
-  &= \sum_{m=0}^\infty \int_{R_1}^{R_2}  j_m(R) dR = \sum_{m=0}^\infty J_m. 
+  \sum_{m=1}^\infty2\pi\int_{R_\mathrm{min}}^{R_\mathrm{max}} 2R\real\left[\Sigma_m v_{\phi
+      m}^*\right] RdR 
+  %&= \sum_{m=0}^\infty \int _{R_\mathrm{min}}^{R_\mathrm{max}}  j_m(R)
+  %dR 
+  = \sum_{m=0}^\infty J_m. 
 \end{align}
-where $^*$ denotes complex conjugation, and we idenfity $J_m$ as the 
-angular momentum associated with disc structure with
-$m^\mathrm{th}$-fold symmetry, and $j_m$ the associated angular
-momentum per unit length. We will use $J_m$ to monitor angular momentum 
-conservation for the disc as a whole.
+where $^*$ denotes complex conjugation. We shall refer to $J_m$ as the
+$m^\mathrm{th}$ component of the total angular momentum, and use it to
+monitor angular momentum conservation in the numerical simulations. 
+It is important to distinguish this empirical definition from the
+angular momentum of linear perturbations, defined through a
+conservation law, which is given in \S\ref{wkb}.
+
+%and we idenfity $J_m$ as the  
+%angular momentum associated with disc structure with
+%$m^\mathrm{th}$-fold symmetry, and $j_m$ the associated angular
+%momentum per unit length. We will use $J_m$ to monitor angular momentum 
+%conservation for the disc as a whole.
+
+\subsection{Three-dimensionality}
+In 3D simulations we measure the importance of vertical motion with
+$\Theta$, where 
+\begin{align}
+  \Theta^2 \equiv \frac{\avg{v_z^2}}{\avg{v_R^2}+\avg{v_\phi^2}}, 
+\end{align}
+and $\avg{v_z^2}$ denotes the density-weighted average,
+\begin{align}
+  \avg{v_z^2} \equiv\frac{
+    \int_{R_1}^{R_2}\int_{\theta_\mathrm{min}}^{\pi/2} \int_{0}^{2\pi}
+    \rho v_z^2 dV}{
+    \int_{R_1}^{R_2}\int_{\theta_\mathrm{min}}^{\pi/2} \int_{0}^{2\pi}
+    \rho dV
+  },
+\end{align}
+and similarly for the horizontal velocities. The radial range of
+integration is taken over $r\in[R_1,R_2]$ since this is where
+we find the perturbations to be confined.
 
@@
 The linearised equations can be combined to a single
-integro-differential eigenvalue problem. We list below some general
-properties of the linear perturbations that will be useful for
-understanding our simulations. We defer a full numerical
-exploration of the linear problem to a future study.
+integro-differential equation eigenvalue problem. We defer a full
+exploration of the linear problem to a future study. Here, we list some
+general  properties of the linear perturbations that will be useful
+for analysing our simulation results.
 
@@
 \end{align}
-where
+\citep[e.g.][]{narayan87,ryu92,lin93} where
 \begin{align}\label{lin_ang_mom_cons}
@@
 consisting of a Reynolds stress and a gravitational stress. Explicit
-expressions for $\bm{\xi}$ and $\bm{F}$ can be found in e.g.,
-\cite{papaloizou85} and \cite{lin11} respectively.
+expressions for $\bm{\xi}$ can be found in, e.g. \cite{papaloizou85},
+and $\bm{F}$ can be read off Eq. \ref{angflux1}.  
+% \cite{papaloizou85} and \cite{lin11} respectively, and
 
-The term $\tbg$ is 
+% lin papaloizou kley 1993, ryu/goodman, narayan
+
+The background torque density $\tbg$ is 
 \begin{align}\label{baroclinic_torque}
@@
 the perturbation is conserved, provided that there is no net angular
-momeumtum flux. We briefly outline the derivation of $T_\mathrm{BG}$
-in Appendix 
+momeumtum flux.
+
+However, as noted in \cite{lin11}, if there is an imposed
+temperature gradient, as in the disc models we consider,
+then $\tbg\neq 0$ in general, which corresponds to a torque
+exerted by the background disc on the perturbation. We 
+outline the derivation of $T_\mathrm{BG}$ in Appendix 
 \ref{tbg_deriv}.
 
-However, as noted in \cite{lin11}, if there is an imposed
-temperature gradient, as is the disc models considered in this paper,
-then $\tbg\neq 0$ in general, which corresponds to a torque
-exerted by the background disc on the perturbation.
-
-We remark that the perturbed angular momentum density $\jlin$ has been
-defined to satisfy a conservation law. This differs from the 
-empirical angular momentum decomposition above.
-
-%each m satisfies its own equation
-
 \subsection{Local results}
-%The linearised equations correspond to an integro-differential
-%equation eigenvalue problem. 
 In the local approximation, perturbations are assumed to vary rapidly
@@
 
-Given the pattern speed $\Omega_p$, Eq. \ref{dispersion} can be solved
+Given the real frequency $\omega$ or pattern speed $\Omega_p\equiv
+\omega/m$, Eq. \ref{dispersion} can be solved 
 for $|k|$, 
@@
 is a dimensionless frequency. The upper (lower) signs in
-Eq. \ref{wavenumber} correspond to short (long) waves.
+Eq. \ref{wavenumber} correspond to short (long) waves, and $k>0$
+($k<0$) correspond to trailing (leading) spirals.
 
@@
 According to Eq. \ref{wavenumber}, purely wave-like solutions with
-real $k$ are only possible when $Q^2(1-\nu^2)\leq1$.
+real $k$ are only possible where $Q^2(1-\nu^2)\leq1$.
 
 A detailed discussion of the properties of density waves in the local
-approximation can be found in \cite{shu91}. A important result is that
+approximation can be found in \cite{shu91}. An important result is that
 waves interior to co-rotation ($R<R_c$) have negative angular momentum, while
@@
 momentum. The local approximation becomes less good for gobal modes,
-but will nevertheless be useful in results interpretation.
+but will nevertheless be useful for results interpretation.

### 2014-11-06 paper/results.tex
update

@@
 $N_R\times N_\phi = 1024\times 2048$, or about $16$ grids per $H$, and
-adopt $\epsilon_g=10^{-4}H$ for the   
+adopt $\epsilon_g=10^{-4}hR$ for the   
 self-gravity softening length\footnote{In 2D self-gravity, $\epsilon_g$ also
@@
 
-In these simulations we subject the disc to initial perturbations in 
+In these simulations the disc is subject to initial perturbations in 
 cylindrical radial velocity, 
@@
 \subsection{Reference run with random perturbations}
-To obtain a picture of the overall disc evolution, here we consider a
+To obtain a picture of the overall disc evolution, we describe a 
 fiducial run initialised with $M=10$ in Eq. \ref{randpert}. 
-%In this fiducial simulation we subject the disc to initial perturbations in
-%cylindrical radial velocity, 
-%\begin{align}\label{randpert}
-%  v_R \to v_R+ c_s\frac{\delta}{M}
-%  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}_d}{\Delta 
-%          R_d}\right)^2\right]}\sum_{m=1}^M\cos{m\phi},
-%\end{align}
-%where $\delta$ is an amplitude, $\overline{R}_d = (R_{1}+R_{2})/2$
-%and $\Delta R_d = (R_{2}-R_{1})/2$. We use $M=10$ and choose 
-%$\delta\in[-10^{-3},10^{-3}]$ randomly but independent of $\phi$.   
 Fig. \ref{fargo_modeamp} plots evolution of the maximum 
@@
   \caption{Evolution of maxima in non-axisymmetric surface density 
-    in the fiducial FARGO simulation.\label{fargo_modeamp}} 
+    in the fiducial FARGO simulation initialised with perturbations
+    with $m\in[1,10]$.\label{fargo_modeamp}} 
 \end{figure}
@@
   0cm 0cm 0cm]{figures/polarxy_dens130} 
-  \caption{Visualisation of the FARGO 2D simulation. The total  
+  \caption{Visualisation of the fiducial FARGO 2D simulation in
+    Fig. \ref{fargo_modeamp}. The total  
     non-axisymmetric surface density
@@
 
-Fig. \ref{2d_angmom} shows the disc angular momenta
-evolution. Only the $m=0,\,1$ components are 
+Fig. \ref{2d_angmom} shows the evolution of disc angular momentum
+components. Only the $m=0,\,1$ components are 
 plotted since they are dominant. The $m=1$ structure has
@@
   \caption{Evolution of angular momentum components in the fiducial
-    FARGO simulation. The perturbation
-    relative to $t=0$ in 2D is shown in units of the
+    FARGO simulation in Fig. \ref{fargo_modeamp}---\ref{fargo_2d}. The
+    perturbation relative to $t=0$ in 2D is shown in units of the 
     initial total angular momentum $J_\mathrm{ref}$.\label{2d_angmom}} 
@@
 evolution is largely similar to the previous case, but the final spiral is
-more co-herent, facilitating the analysis below. 
+more coherent, which facilitates the analysis below.
+
 Fig. \ref{2d_fargo_viz} shows the $m=1$ surface density
-distribution at the end of this simulation.
+at the end of this simulation. For this one-arm spiral
+we measure a co-rotation radius and growth rate of 
+\begin{align*}
+  &R_c \simeq 4.4R_0,\\
+  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_k(R_c). 
+\end{align*}
+The spiral can be considered as low frequency because its 
+pattern speed $\Omega_p\lesssim 0.3\Omega$ in $R\in[R_{1},R_{2}]$,
+where it is  most prominent. The growth rate $\gamma$ is also slow
+relative to the local rotation. Thus, the spiral 
+pattern is nearly stationary.
 
@@
 
-By examining the disc evolution in $t\in[80,120]P_0$, we find the
-one-arm spiral in Fig. \ref{2d_fargo_viz} has a 
-co-rotation radius and growth rate of
-\begin{align*}
-  &R_c \simeq 4.4R_0,\\
-  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_k(R_c). 
-\end{align*}
-The spiral can be considered as low frequency because $|\omega|
-\lesssim 0.3\Omega$ (recall $\omega = 
-m\Omega_p$) in the region it is promininent $R\in[R_{1},R_{2}]$. The
-growth rate is also slow relative to the local rotation. This gives a
-nearly stationary spiral pattern.
-
-%Since most of
-%the perturbation lies at $R<R_c$, this mode has negative
-%angular momentum according to local linear theory.
-
 Next, we write $\Sigma_1 = 
@@
 main assumption in local theory. We plot in
-Fig. \ref{fargo_wavenumber} the normalise wavenumber associated with
-the $m=1$ spiral. We find that
+Fig. \ref{fargo_wavenumber} the normalised radial wavenumber associated with
+the $m=1$ spiral. We find 
 \begin{align*}
@@
 where we used $Q\sim c_s\Omega/\pi G \Sigma$ and $R\Omega/c_s\sim
-h^{-1}$. Since $Q=O(1)$ and $h\ll 1$, $|kR|\gg 1$ and we 
-can apply results from local theory self-consistently. Since $k>0$,
-the spiral is trailing, as seen in Fig. \ref{2d_fargo_viz}.
+h^{-1}$. Since $Q=O(1)$ and $h\ll 1$ imply $|kR|\gg 1$, we 
+can apply results from local theory self-consistently. Recall $k>0$
+means the spiral is trailing, as seen in Fig. \ref{2d_fargo_viz}.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%
+
 Using the estimated value of $R_c$, we plot in
@@
 the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$, which is indeed the
-region where the $m=1$ spiral develops. This means that the one-arm
+region where the $m=1$ spiral develops. This suggests that the one-arm
 spiral is trapped. There is one outer Lindblad resonance at $R_L\simeq
 7.2R_0$. Thus, waves may be launched in $R\gtrsim 7.2R_0$ by the main 
-spiral disturbance in the inner disc. %this would be a de-stabilising
+spiral disturbance in the inner disc \citep{lin11b}. 
+%this would be a de-stabilising
 % effect.
@@
   \caption{Dimensionless mode frequency $\nu$ for the $m=1$ spiral in
-    Fig. \ref{2d_fargo_viz}. The disperion relation for local density
-    waves, Eq. \ref{dispersion}, permits purely wave-like solutions in 
-    regions where $\nu^2 - 1 + Q^{-2}>0$. 
+    Fig. \ref{2d_fargo_viz}. For a given real mode frequency, the
+    dispersion relation for local density waves, Eq. \ref{dispersion},
+    permits purely wave-like solutions in regions where $\nu^2 - 1 +
+    Q^{-2}>0$.    
     \label{fargo_qbarrier}} 
@@
 \begin{align*}
-  \gamma \simeq 0.015 q \Omega_k(R_0). 
+  \gamma \simeq \left[0.015 q - 7.9\times10^{-4}\right] \Omega_k(R_0). 
 \end{align*}
@@
 \begin{align}
-  \gamma \simeq 0.10h\Omega_k(R_0).  
+  \gamma \simeq \left[0.10h + 8.3\times10^{-3}\right]\Omega_k(R_0).  
 \end{align}

### 2014-11-06 paper/model.tex
minor edits

@@
 
-\subsection{Three-dimensionality}
+\subsubsection{Three-dimensionality}
 In 3D simulations we measure the importance of vertical motion with
@@
 \end{align}
-and $\avg{v_z^2}$ denotes the density-weighted average,
+and $\avg{\cdot}$ denotes the density-weighted average, e.g., 
 \begin{align}
@@
 \end{align}
-and similarly for other variables. Since the background is
+and similarly for other variables. We take $m>0$. Since the background is
 axisymmetric, we identify $\delta\Sigma_m(R)\exp{\left(-\ii\sigma
@@
 
-The linearised equations can be combined to a single
+The linearised equations can be combined into a single
 integro-differential equation eigenvalue problem. We defer a full
@@
 general  properties of the linear perturbations that will be useful
-for analysing our simulation results.
+for describing simulation results.
 
@@
 \end{align}
-\citep[e.g.][]{narayan87,ryu92,lin93} where
+\citep[e.g.][]{narayan87,ryu92,lin93b} where
 \begin{align}\label{lin_ang_mom_cons}
@@
 
-Given the real frequency $\omega$ or pattern speed $\Omega_p\equiv
+Considering a neutral mode, given the real frequency $\omega$ or
+pattern speed $\Omega_p\equiv 
 \omega/m$, Eq. \ref{dispersion} can be solved 
@@
 \end{align}
-is a dimensionless frequency. The upper (lower) signs in
-Eq. \ref{wavenumber} correspond to short (long) waves, and $k>0$
-($k<0$) correspond to trailing (leading) spirals.
+is a dimensionless frequency. In
+Eq. \ref{wavenumber}, the upper (lower) sign correspond to short
+(long) waves, and $k>0$ ($k<0$) correspond to trailing (leading)
+spirals.
 

### 2014-11-06 paper/paper.tex
appendix and minor updates
update
minor edits

@@
 \newcommand{\sbar}{\bar{\sigma}}
-\newcommand{\avg}[1]{\langle #1 \rangle_\phi}
+\newcommand{\avg}[1]{\langle #1 \rangle}
 \newcommand{\bu}{\bm{u}}
@@
   the FARGO code and in 3D with the ZEUS-MP and PLUTO codes. In such
-  systems we observe the development of a 
+  discs we observe the development of a 
   spiral mode with azimuthal wavenumber $m=1$ with co-rotation radius 
   outside the bulk of the spiral arm, resulting in a nearly stationary 
-  one-arm spiral pattern in the self-gravitating region. This one-arm 
+  one-arm spiral pattern in the self-gravitating region of the
+  disc. The development of this one-arm 
   spiral does not require the movement of  
@@
   growth of this spiral to the non-conservation of angular momentum
-  for linear perturbations in discs with a forced temperature
-  gradient. This enables angular momentum exchange between the spiral
-  and the background disc, and may lead to instability if the  
-  interaction increases the amplitude of the linear perturbations.
+  for non-axisymmetric linear perturbations in discs with a forced temperature
+  gradient. Such interaction may lead to an unstable exchange of angular momentum between the spiral 
+  and the background disc. However, because the exchange mechanism does
+  not explicitly depend on self-gravity, we speculate it may also
+  enhance low-frequency one-arm spirals in non-self-gravitating
+  discs.
 

### 2014-11-06 paper/results.tex
minor edits

@@
 \subsection{Trapping the low-frequency $m=1$ spiral}\label{fargo_m1}
-In order to focus on the $m=1$ spiral that eventually emerges, here 
-we examine a simulation initialised with only $m=1$
-perturbations, by setting $M=1$ in Eq. \ref{randpert}. The disc
+In order to focus on the $m=1$ spiral that eventually emerges, from
+now on we examine simulations initialised with
+$M=1$ in Eq. \ref{randpert}. The disc 
 evolution is largely similar to the previous case, but the final spiral is
@@
 where it is  most prominent. The growth rate $\gamma$ is also slow
-relative to the local rotation. Thus, the spiral 
+relative to the local rotation. (Although the characteristic growth
+time $\gamma^{-1} \simeq 10P_0$.) Thus, the spiral 
 pattern is nearly stationary.
@@
 $|\Sigma_1|$ varies slowly compared to the complex phase. This is the
-main assumption in local theory. We plot in
-Fig. \ref{fargo_wavenumber} the normalised radial wavenumber associated with
-the $m=1$ spiral. We find 
+main assumption in local theory. We measure $k$ numerically and plot
+its normalised value in Fig. \ref{fargo_wavenumber}. We find 
 \begin{align*}
@@
 can apply results from local theory self-consistently. Recall $k>0$
-means the spiral is trailing, as seen in Fig. \ref{2d_fargo_viz}.
+corresponds to a trailing spiral, as seen in Fig. \ref{2d_fargo_viz}.
 
@@
 spiral is trapped. There is one outer Lindblad resonance at $R_L\simeq
-7.2R_0$. Thus, waves may be launched in $R\gtrsim 7.2R_0$ by the main 
+7.2R_0$. Thus, waves may be launched in $R\gtrsim 7.2R_0$ by the  
 spiral disturbance in the inner disc \citep{lin11b}. 
 %this would be a de-stabilising
-% effect.
+% effect.  we discuss this in discussions?
 
@@
 \end{align}
-where we recall $T_{BG}$ is the torque density related to the imposed
+where we recall $T_\mathrm{BG}$ is the torque density due to the imposed
 sound-speed profile (Eq. \ref{baroclinic_torque}). We have explicitly
@@
 and $\Delta\Sigma_1\sim 0.4$ by $t=120P_0$) and linear theory becomes
-less valid. Fig. \ref{fargo_angmom_ex} confirms that angular momentum
-is lost from the wave to the disc.
+less valid. Fig. \ref{fargo_angmom_ex} confirms that the $m=1$ spiral
+wave experiences a negative torque that further reduces its angular
+momentum. This is consistent with angular momentum component
+measurements (Fig. \ref {2d_angmom}).
 
@@
 inspection of the $m=1$ surface density field. The growth in this case
-may be associated with higher $m$ modes. %"non-linear effects"
+may be associated with higher $m$ modes.  
+% show result of isothermal calc?
 
@@
 We plot growth rates of the $m=1$ spiral as a function of $q$ in 
-Fig. \ref{fargo_varq_growth}. The correlation is well-fitted by a
+Fig. \ref{fargo_varq_growth}. The correlation can be fitted with a
 linear relation
@@
 \end{figure}
+%maybe extend the low q simulations to get better rates?
 
-% only cases where a coherent spiral is seen is plotted
-
-We also performed a series of simulations with aspect-ratio
-$h\in[0.03,0.07]$ with fixed $q=1$. This also affects the temperature
+We also performed a series of simulations with variable aspect-ratio
+$h\in[0.03,0.07]$ but fixed $q=1$. This also affects the temperature
 gradient since $c_s \propto h$. However, with other parameters equal
@@
 mass within $R\in[R_1,R_2]$  ranges from $0.033M_*$ to $0.062M_*$. 
+% similar Q profiles
 
@@
 \end{align}
-However, the linear fit is less good than for variable $q$ above. This
+However, a linear fit is less good than for variable $q$ above. This
 may be due to the change in the total disc mass when $h$ changes. We

### 2014-11-06 paper/results_3d.tex
update
minor edits

@@
 \section{Three-dimensional simulations}
-In this section we briefly review results from 3D simulations carried 
-out using the ZEUS-MP and PLUTO. The main purpose is to check 
+In this section we briefly review 3D simulations carried 
+out using ZEUS-MP and PLUTO. The main purpose is to check 
 the above results with different numerical codes, and validate  
@@
 $n_H=2$ scale-heights. The resolution is $N_r\times N_\theta\times
-N_\phi=256\times32\times256$. Because of the much reduced resolution
+N_\phi=256\times32\times256$. This corresponds to about $4$ cells per
+$H$in radius. 
+Because of the much reduced resolution 
 compared to 2D, we use a smooth perturbation by setting
 $\delta = 10^{-3}$ and $M=1$ in Eq. \ref{randpert}. This corresponds
-to a single $m=1$ spiral in the dead zone, which was found to dominate
+to a single $m=1$ disturbance in $R\in[R_1,R_2]$, which was found to dominate
 the 2D simulation.
@@
 
-Snapshots at the end of the simulations are shown in
-Fig. \ref{3d_prelim} for both ZEUS-MP and PLUTO. We again find the dead zone eventually
-dominated by an $m=1$ spiral. %A Cartesian visualisation of the
-% midplane density perturbation associated with the $m=1$ mode is shown in Fig. \ref{pluto_cart}.
-The two codes give qualitatively similar results, but a longer PLUTO simulation was
-needed to achieve similar mode amplitudes in ZEUS-MP.
+Fig. \ref{3d_ampmax} plots the amplitude of the $m=1$ spirals measured
+in the ZEUS-MP and PLUTO runs. We also show results from strictly
+isothermal discs ($q=0$), which clearly display no growth. %why no high m
+                                %modes as in 2D for iso disc?
+                                %resolution?
+Results from ZEUS-MP appears to be off-set from that obtained using
+PLUTO. We suspect this may be due to larger numerical noise in the
+ZEUS-MP code. Thus the $m=1$ spiral grows to a larger amplitude by the
+end of the run. However, we measure similar growth rates,
+\begin{align*}
+&\gamma \simeq 0.0073\Omega_k(R_0) \quad\quad \mathrm{PLUTO},\\
+&\gamma \simeq 0.0075\Omega_k(R_0) \quad\quad \mathrm{ZEUS-MP}.
+\end{align*}
+Both are smaller than the 2D simulations above. This may be due to the
+low resolutions adopted. 
+%growth rate does not depend on pert amp
 
-%The amplitude of the spiral is smaller than the 2D case above,
-%possibily due to lower resolution and/or 3D self-gravity being weaker
-%than that in 2D.
+\begin{figure}
+  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ampmax3d.ps}
+  \caption{Evolution of the maximum $m=1$ density in  $r\in[R_1,R_2]$
+    measured in the 3D simulations. Both discs with a imposed
+    temperature gradient ($q=1$) and a strictly isothermal disc
+    ($q=0$) are shown. 
+    \label{3d_ampmax}}   
+\end{figure}
+
+Visualisations of the 3D simluations are shown in Fig. \ref{3d_prelim} for
+both ZEUS-MP and PLUTO. The snapshots are chosen for when the spiral
+amplitudes are similar. The agreement between the two codes, as well
+as with the 2D simulations, is satisfactory.
+
+%but a longer PLUTO simulation was 
+%needed to achieve similar mode amplitudes in ZEUS-MP.
 
@@
   \end{center}
-  \caption{Preliminary 3D simulations using the (a) ZEUS-MP and (b)
-    PLUTO. The non-axisymmetric midplane density at the end of the run
-    is shown. Here $\psi \equiv \pi/2 - \theta$ is the angular height
+  \caption{Three-dimensional simulations using the (a) ZEUS-MP and (b)
+    PLUTO. The non-axisymmetric midplane density is shown. Here $\psi
+    \equiv \pi/2 - \theta$ is the angular height 
     from the midplane.\label{3d_prelim}}   
@@
 
-% \begin{figure}
-%   \includegraphics[width=\linewidth]{figures/pdiskxy_020_m1}
-%   \caption{Cartesian visualisation the $m=1$ midplane density in the PLUTO
-%     simulation shown in Fig. \ref{3d_prelim}.\label{pluto_cart}}   
-% \end{figure}
+\subsection{Vertical motion}
 
-%\subsection{Angular momentum evolution}
+\subsection{Angular momentum evolution}
 
@@
 J/J|=O(10^{-8})$.
-
-%Note that FARGO does not explicitly conserve angular
-%momentum. In addition, there may be an angular momentum flux due to
-%self-gravity across disc boundaries since our domain size is finite. 
-%Fig. \ref{prelim_angmom} shows the angular momentum evolution in
-%the 3D runs.
 
@@
 \end{figure}
-
-% \section{One-arm spirals in structured discs}
-% Here we examine a 2D FARGO simulation in more detail. The setup is the
-% same as that in \S\ref{fargo_fiducial}, except we use a resolution of
-% $N_R\times N_\phi = 2048\times4096$ and initialise the disc with a
-% single $m=1$ perturbation.
-
-% \subsection{Properties of the $m=1$ dead zone spiral}
-% % growth rate, co-rotation radius, pattern speed, propagation diagram

### 2014-11-06 paper/summary.tex
minor edits

\section{Summary and conclusions}\label{summary}

\subsection{Spectulations and future work}

### 2014-11-06 paper/paper.tex
minor edits

@@
   for non-axisymmetric linear perturbations in discs with a forced temperature
-  gradient. Such interaction may lead to an unstable exchange of angular momentum between the spiral 
-  and the background disc. However, because the exchange mechanism does
+  gradient. This may lead to an unstable exchange of angular momentum between the spiral 
+  and the background disc. However, because the interaction does
   not explicitly depend on self-gravity, we speculate it may also

### 2014-11-07 paper/discussion.tex
minor edits
major updates

@@
 sign to that of the density waves it induces in the exterior disc
-\citep{lin11b}.
+\citep{lin11b}. To estimate this effect, we port basic results from
+disc-planet theory \citep[see, e.g.][and references
+therein]{papaloizou07}.
 
-Considering the spiral as a external potential (c.f. an embedded
-satellite), we expect it to  exert a torque on the disc at Lindblad
-and co-rotation resonances.
+Let us regard the one-arm spiral disturbance located in the inner disc as an  
+external potential of the form $\Phi_\mathrm{ext}(R)\cos{\left(\phi -
+    \Omega_pt\right)}$. We expect
+it to exert a torque on the exterior disc at Lindblad  
+and co-rotation resonances. At the outer Lindblad resonance (OLR), this
+torque is 
+\begin{align}
+  \Gamma_L =
+  \frac{\pi^2\Sigma_L}{3\Omega_L\Omega_p}
+\left[\left.R_L\frac{d\Phi_\mathrm{ext}}{dR}\right|_L + 2\left(1 - \frac{\Omega_p}{\Omega_L}\right)\Phi_\mathrm{ext}\right]^2,
+\end{align}   
+where a Keperian disc has been assumed and subscript $L$ denotes
+evaluation at $R=R_L$. (The inner Lindblad resonance does not exist
+for the pattern speeds observed in our simulations.)
+
+To order-of-magnitude, we reasonably take
+\begin{align}
+  \Phi_\mathrm{ext} = -\frac{GM_\mathrm{ring}}{|R - \bar{R}|}, 
+\end{align}
+for the potential associated with a disturbance located at
+$R=\bar{R}$, where $M_\mathrm{ring}$ is the mass in $R\in[R_1,R_2]$.
+If we also associate its angular momentum with 
+$J_\mathrm{ext}  = M_\mathrm{ring}\bar{R}^2\Omega_p$, we can obtain a
+rate $\gamma_L$ by evaluating $\Gamma_L/J_\mathrm{ext}$. Then 
+\begin{align}
+  \frac{\gamma_L}{\Omega_p} \sim &\frac{\pi h}{3
+    Q_L}\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_L}{\bar{R}}\right)^{-1/2}\left(\frac{R_c}{\bar{R}}\right)^{5/2}\notag\\
+  &\times \left\{ \frac{R_L}{R_L - \bar{R}} - 2\left[1 -
+      \left(\frac{R_c}{R_L}\right)^{-3/2}\right]\right\}^2\frac{R_c^2}{\left(R_L
+    - \bar{R}\right)^2}. 
+\end{align}
+Inserting $h=0.05$, $Q_L=10$, $M_\mathrm{ring} = 0.05M_*$,
+$R_L=7.2R_0$, $R_c=4.4R_0$ and $\bar{R}=1.5R_0$ from our fiducial
+FARGO simulation, we get
+\begin{align}
+  \gamma_L \sim 0.012\Omega_p. 
+\end{align}
+This represents an upper limit on the effect of the OLR on the spiral
+disturbance, and is much smaller than the growth rate measured in the
+simulation ($\gamma\sim0.1\Omega_p$). We conclude that for our models,
+the OLR has negligible effect on the one-arm spiral.
+
+For the co-rotation torque, we use the result
+\begin{align}
+  \Gamma_C = \left.
+    \pi^2\Phi_\mathrm{ext}^2\left(\frac{d\Omega}{dR}\right)^{-1}\frac{d}{dR}\left(\frac{2\Sigma}{\Omega}\right)\right|_{C}
+\end{align}
+for a Keplerian disc, where subscript $C$ denotes evaluation at
+co-rotation radius $R=R_c$. For a power-law surface density profile
+$\Sigma\propto R^{-s}$ we have
+\begin{align}
+  \Gamma_C = \frac{4}{3}\frac{\pi h}{Q_C} \left(s -
+    \frac{3}{2}\right)\left(\frac{M_\mathrm{ring}}{M_*}\right)\frac{R_c^4}{\bar{R}^2\left(R_c
+      - \bar{R}\right)^2}.   
+\end{align}
+Using the above parameter values with $s=2$ and $Q_c=10$, we obtain a
+rate
+\begin{align}
+  \frac{\gamma_C}{\Omega_p} \sim 0.010\Omega_p,
+\end{align}
+which is again small compared to the measured growth rates.
+
+%expressions for linear response, but have massive "planet" -> but CR
+%and LR far from where spiral is
 
@@
 
-\subsection{Speculations}
+%\subsection{Speculations}
 %speculations - long term balance. forced temp gradient drives

### 2014-11-07 paper/model.tex
major updates

@@
 $\Theta$, where 
-\begin{align}
+\begin{align}\label{theta}
   \Theta^2 \equiv \frac{\avg{v_z^2}}{\avg{v_R^2}+\avg{v_\phi^2}}, 
@@
 We list here the main definitions and results from 2D linear theory,
-taken from \cite{lin11} and \cite{shu91}.
+taken from \cite{lin11b} and \cite{shu91}.
 
@@
 
-However, as noted in \cite{lin11}, if there is an imposed
+However, as noted in \cite{lin11b}, if there is an imposed
 temperature gradient, as in the disc models we consider,

### 2014-11-07 paper/paper.tex
major updates

@@
 \newcommand{\tbg}{T_\mathrm{BG}}
+\newcommand{\varA}[1]{{\operatorname{#1}}}
 \DeclareMathOperator{\erf}{erf}
@@
   not explicitly depend on self-gravity, we speculate it may also
-  enhance low-frequency one-arm spirals in non-self-gravitating
+  affect low-frequency one-arm spirals in non-self-gravitating
   discs.

### 2014-11-07 paper/results.tex
major updates

@@
 solutions to the dispersion relation (Eq. \ref{dispersion}) when the
-mode frequency is given. The figure shows two $Q$-barriers located in
+mode frequency is given. %Note that, with the above measured wavenumber $k$, 
+%Eq. \ref{dispersion} requires $\nu^2 - 1 + Q^{-2}\sim0$ and the lower
+%sign (long waves) to be taken. 
+Fig. \ref{fargo_qbarrier} shows two $Q$-barriers located in
 the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$, which is indeed the
 region where the $m=1$ spiral develops. This suggests that the one-arm
-spiral is trapped. There is one outer Lindblad resonance at $R_L\simeq
+spiral is trapped. Note in this region, $\nu^2 - 1 +
+Q^{-2}\simeq 0.1\ll 1$, as required for consistency with
+Eq. \ref{dispersion} (with the lower sign taken, i.e. long waves).  
+There is one outer Lindblad resonance at $R_L\simeq
 7.2R_0$. Thus, waves may be launched in $R\gtrsim 7.2R_0$ by the  

### 2014-11-07 paper/results_3d.tex
major updates

@@
 &\gamma \simeq 0.0073\Omega_k(R_0) \quad\quad \mathrm{PLUTO},\\
-&\gamma \simeq 0.0075\Omega_k(R_0) \quad\quad \mathrm{ZEUS-MP}.
+&\gamma \simeq 0.0075\Omega_k(R_0) \quad\quad \varA{ZEUS-MP}.
 \end{align*}
@@
 
-Visualisations of the 3D simluations are shown in Fig. \ref{3d_prelim} for
-both ZEUS-MP and PLUTO. The snapshots are chosen for when the spiral
-amplitudes are similar. The agreement between the two codes, as well
-as with the 2D simulations, is satisfactory.
-
-%but a longer PLUTO simulation was 
-%needed to achieve similar mode amplitudes in ZEUS-MP.
+Visualisations of the 3D simluations are shown in
+Fig. \ref{3d_prelim} for the disc midplane and near the upper disc
+boundary. The snapshots are chosen when the spiral 
+amplitudes reached in the two codes are similar. Although the
+non-axisymmetric density has similar amplitudes at both heights, a
+coherent spiral is less evident away from the disc midplane. 
+The agreement between the two codes, as well as with the 2D
+simulations, is satisfactory.
 
@@
     \subfigure[ZEUS-MP]{
-      \includegraphics[scale=0.33]{figures/polarxy_dens015_zeus}
+      \includegraphics[scale=0.33]{figures/polarxy_dens015_z0}
+      \includegraphics[scale=0.33,clip=true,trim=2.24cm 0cm 0cm
+      0cm]{figures/polarxy_dens015_zmax}  
     }
     \subfigure[PLUTO]{
-      \includegraphics[scale=0.33]{figures/pdisk_020}
+      \includegraphics[scale=0.33]{figures/pdisk_021_z0}
+      \includegraphics[scale=0.33,clip=true,trim=2.24cm 0cm 0cm
+      0cm]{figures/pdisk_021_zmax}
     }
   \end{center}
-  \caption{Three-dimensional simulations using the (a) ZEUS-MP and (b)
-    PLUTO. The non-axisymmetric midplane density is shown. Here $\psi
+  \caption{Three-dimensional simulations using the ZEUS-MP (top) and 
+    PLUTO (bottom) codes. The non-axisymmetric density at the midplane (left) and
+    two scale-heights above the midplane (right) is shown. Here $\psi
     \equiv \pi/2 - \theta$ is the angular height  
@@
 \subsection{Vertical motion}
+In Fig. \ref{3d_vert} we measure three-dimensionality using the
+parameter $\Theta$ as defined in Eq. \ref{theta}. Vertical velocities
+do increase with the growth of the one-arm spiral. Larger amplitudes
+are reached in ZEUS because of larger numerical noise. 
+However, even 
+in this case, $\Theta \lesssim 10^{-2}$. This implies $|v_z/c_s|\lesssim
+0.2$ for $h=0.05$. Thus, vertical motions are sub-sonic and negligible
+compared to horizontal motions.
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_vert.ps}
+  \caption{Evolution of three-dimensionality $\Theta$, as defined by
+    Eq. \ref{theta}, which corresponds to the
+    average vertical kinetic energy density, scaled by 
+    the average horizontal kinetic energy.\label{3d_vert}} 
+\end{figure}
 
@@
   \includegraphics[scale=.41]{figures/nonaxi_evol_ang_pluto}
-  \caption{Evolution of angular momentum components in the fiducial 3D 
+  \caption{Evolution of angular momentum components in the 3D 
     simulations. The perturbation 
-    relative to $t=100P_0$ is shown in units of the
-    initial total angular momentum $J_\mathrm{ref}$.\label{3d_angmom}} 
+    relative to $t=100P_0$, during the growth of the one-arm spiral,
+    is shown in units of the initial total angular momentum
+    $J_\mathrm{ref}$.\label{3d_angmom}}  
 \end{figure}

### 2014-11-07 paper/results_3d.tex
update

@@
 \subsection{Vertical motion}
-In Fig. \ref{3d_vert} we measure three-dimensionality using the
-parameter $\Theta$ as defined in Eq. \ref{theta}. Vertical velocities
-do increase with the growth of the one-arm spiral. Larger amplitudes
-are reached in ZEUS because of larger numerical noise. 
-However, even 
-in this case, $\Theta \lesssim 10^{-2}$. This implies $|v_z/c_s|\lesssim
+We measure three-dimensionality using the parameter $\Theta$ as
+defined in Eq. \ref{theta}.  Vertical velocities
+do increase with the growth of the one-arm spiral, 
+%Larger amplitudes
+%are reached in ZEUS because of larger numerical noise. 
+but we find  $\Theta \lesssim 10^{-2}$. This implies $|v_z/c_s|\lesssim
 0.2$ for $h=0.05$. Thus, vertical motions are sub-sonic and negligible
-compared to horizontal motions.
+compared to horizontal motions. This supports the 2D approximation.
 
-\begin{figure}
-  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_vert.ps}
-  \caption{Evolution of three-dimensionality $\Theta$, as defined by
-    Eq. \ref{theta}, which corresponds to the
-    average vertical kinetic energy density, scaled by 
-    the average horizontal kinetic energy.\label{3d_vert}} 
-\end{figure}
-
+%\begin{figure}
+%  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_vert.ps}
+% \caption{Evolution of three-dimensionality $\Theta$, as defined by
+%    Eq. \ref{theta}, which corresponds to the
+%   average vertical kinetic energy density, scaled by 
+%    the average horizontal kinetic energy.\label{3d_vert}} 
+%\end{figure}   
 \subsection{Angular momentum evolution}

### 2014-11-08 paper/discussion.tex
update

@@
 
-\subsection{Role of self-gravity and disc structure} 
-We note that the background torque density $T_\mathrm{BG}$ does not depend on  
-self-gravity, and we only considered a self-gravitating perturbation to
+\subsection{The role of self-gravity and disc structure} 
+The background torque density $T_\mathrm{BG}$ does not depend on  
+self-gravity explicitly. In the above discussion, we only considered a self-gravitating perturbation to 
 obtain an appropriate wavenumber $k$ needed to evaluate
-Eq. \ref{theoretical_rate}. Thus, self-gravity is not directly
+Eq. \ref{theoretical_rate}. Self-gravity is not directly
 responsible for instability. However, 
@@
 dispersion relation is possible (Eq. \ref{wavenumber}), as a result of the adopted disc
-structure. Thus, the main role of disc structure and self-gravity is to
+structure. Thus, in our disc models the main role of disc structure and self-gravity is to 
 allow a neutral mode to be set up and confined, which is then
 destabilised through the background torque density.
+
+In order to confine an $m=1$ spiral between two $Q$-barriers, we
+should have $Q^2(1-\nu^2)=1$ at two radii. Assuming 
+Keperlian rotation and a slow pattern speed $\Omega_p\ll\Omega$, this amounts to
+\begin{align}
+  \left(\frac{R_{Qb}}{R_c}\right)^{-3/2} = 2Q^2(R_{Qb}). 
+\end{align}
+Then it may be possible to have two $Q$-barriers when the $Q$ profile
+of the disc rises more rapidly (decays more slowly) than $R^{-3/2}$
+for decreasing (increasing) $R$. One possibility is a gap
+opened by a giant planet in a constant-$Q$ disc. In that case $Q$
+rises rapidly towards the gap centre since it is a region of low
+surface density.  %multi spirals associated with one planet
 
@@
 
-\subsubsection{The confined spiral as an external potential}
+\subsection{The confined spiral as an external potential}
 One possible effect of self-gravity is to allow the one-arm spiral 
@@
     \Omega_pt\right)}$. We expect
-it to exert a torque on the exterior disc at Lindblad  
+it to exert a torque on the exterior disc at the Lindblad  
 and co-rotation resonances. At the outer Lindblad resonance (OLR), this
@@
 where a Keperian disc has been assumed and subscript $L$ denotes
-evaluation at $R=R_L$. (The inner Lindblad resonance does not exist
-for the pattern speeds observed in our simulations.)
+evaluation at the OLR, $R=R_L$. (The inner Lindblad resonance does not
+exist for the pattern speeds observed in our simulations.)
 
@@
 \begin{align}
-  \Gamma_C = \frac{4}{3}\frac{\pi h}{Q_C} \left(s -
+  \frac{\Gamma_C}{J_\mathrm{ext}\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_C} \left(s -
     \frac{3}{2}\right)\left(\frac{M_\mathrm{ring}}{M_*}\right)\frac{R_c^4}{\bar{R}^2\left(R_c
@@
 \begin{align}
-  \frac{\gamma_C}{\Omega_p} \sim 0.010\Omega_p,
+  \gamma_C\sim 0.010\Omega_p,
 \end{align}
 which is again small compared to the measured growth rates.
+
+%not important for our models, but may be important if outer disc is
+%self-gravitating too
 

### 2014-11-08 paper/paper.tex
update

@@
 \newcommand{\apjl}{ApJL}
+\newcommand{\pasj}{PASJ}
 \newcommand{\araa}{ARAA}
@@
   discs. 
-
-  % or confirm in 3D
-  %, taken as a toy model of a 
-  %protoplanetary disc containing a finite-size massive dead zone.
-   % Theoretical models of circumstellar or protoplanetary discs indicate
-  % the possible existence of `dead zones' --- regions in the disc
-  % where mass accretion is greatly reduced because of inefficient
-  % magneto-hydrodynamic turbulence that would otherwise enable 
-  % accretion. It has been suggested that mass built-up in a dead zone
-  % may lead to gravitational instability. However, what type of
-  % gravitational instability is relevant in this context has not been
-  % clarified. 
-  %
-  % Our simulations demonstrate the 
-  % importance of global gravitational interaction between a 
-  % self-gravitating dead zone and the exterior, non-self-gravitating
-  % parts of the disc. This non-local effect, due to disc structure, 
-  % exists even when the total disc mass is below the threshold for
-  % global effects found in
-  % previous studies. 
 \end{abstract}
@@
 \begin{keywords}
-  planetary systems: formation --- planetary systems:
-  protoplanetary discs
+  accretion, accretion discs, hydrodynamics, instabilities, methods: numerical, protoplanetary discs 
 \end{keywords}
@@
 \section{Introduction}\label{intro}
+% Protoplanetary discs are expected to contain complex structures
+% \citep{armitage11}, which directly impact planet formation theories.  
+One of the exciting developments in the study of circumstellar 
+discs is the direct observation of large-scale, non-axisymmmetric
+structures within them. These
+include lopsided dust distributions 
+\citep{marel13,fukagawa13,casassus13,isella13,perez14,follette14} and
+spiral arms 
+\citep{muto12,grady13,christiaens14,avenhaus14,boccaletti14}.
+
+% planet is popular answer for spiral structure
+An attractive explanation for asymmetries in circumstellar discs is to
+invoke disc-planet interaction. In particular, spiral structures
+naturally arise from the gravitational interaction between a
+planet and the gaseous protoplanetary disc it is embedded in
+\citep[see, e.g.][for a recent review]{baruteau13b}. Thus, the
+presence of spiral arms in circumstellar discs could be signposts of 
+planet formation.
+
+% but also natural in GI. when/where does it happen? 
+Spiral arms are also naturally associated with gravitational
+instability in differentially rotating discs.
+
+%original motivation for this work (DZGI)
+
+%specifically m=1 (history, conditions)
 
@@
 \input{discussion}
-%\input{summary}
+\input{summary}
 
@@
 \input{appendix}
-%\input{appendix2}
 

### 2014-11-09 paper/paper.tex
udpated

@@
 
-% but also natural in GI. when/where does it happen? 
-Spiral arms are also naturally associated with gravitational
-instability in differentially rotating discs.
+% but also natural in GI.
+Spiral arms are also naturally associated with global gravitational 
+instability (GI) in differentially rotating discs
+\citep{laughlin96b,laughlin98,lodato05,forgan11}. Large-scale spiral
+arms may provide significant angular momentum transport to enable
+accretion \citep{papaloizou91,balbus99,lodato04}, and spiral
+structures are potentially observable with the Atacama Large
+Millimeter/sub-millimeter Array \citep{cossins10,dipierror14}.
+
+% when/where does it happen? formation stage/dead zones (special for ppd)
+GI can be expected in the earliest stages of
+circumstellar disc formation \citep{inutsuka10,tsukamoto13}. %also
+% outer parts
+In the case of protoplanetary discs, which have complex structure
+\citep{armitage10}, GI may also occur in `dead
+zones'. Indeed, the original motivation for our study was to clarify
+exactly what type of GI is relevant to dead zones?
 

### 2014-11-09 paper/paper.tex
updated intro

--- 2014-11-08 paper/paper.tex
@@
 
-% but also natural in GI. when/where does it happen? 
-Spiral arms are also naturally associated with gravitational
-instability in differentially rotating discs.
+% but also natural in GI.
+Spiral arms are also naturally associated with global gravitational 
+instability (GI) of differentially rotating discs
+\citep{goldreich65,laughlin96b,laughlin98,lodato05,forgan11}. Large-scale
+spiral arms may provide significant angular momentum transport to enable
+accretion \citep{lynden-bell72, papaloizou91,balbus99,lodato04}, and
+spiral structures are potentially observable with the Atacama Large 
+Millimeter/sub-millimeter Array \citep{cossins10,dipierror14}.
 
-%original motivation for this work (DZGI)
+% when/where does it happen? formation stage/dead zones (special for ppd)
+GI can be expected in the earliest stages of 
+circumstellar disc formation \citep{kratter10b,inutsuka10,tsukamoto13}. %also
+% outer parts  
+In the case of protoplanetary discs, which have complex structure
+\citep{armitage10}, GI may also occur in `dead
+zones' --- sub-portions of the disc where the magneto-rotational
+instability is ineffective for angular momentum transport
+\citep{gammie96,turner08,landry13} and therefore has reduced mass
+accretion. Dead zones may become
+self-gravitating with sufficient mass built-up
+\citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13,bae14}.
 
-%specifically m=1 (history, conditions)
+% original motivation for this work (DZGI) 
+Indeed, the original goal for this project was to clarify
+what type of GI is relevant to astrophysical discs in which only part 
+of it is self-gravitating, such as a protoplanetary disc containing a 
+massive dead zone. However, our seemingly simple initial disc models
+consistently produced one-armed spiral structures, despite not 
+satisfying the usual requirements to generate them
+(see below). We therefore dedicate this study to verify their origin.
+
+%brief history of m=1 
+%stellar wobble from indiana group 
+Single-arm, or eccentric modes in
+
+%this work  mechanism is not strictly sg
+
+%this paper is organised as follows
 

### 2014-11-09 paper/discussion.tex
intro draft completed

@@
-\section{Discussion}
+\section{Discussion}\label{discussions}
 Having confirmed the $m=1$ spiral instability through numerical

### 2014-11-09 paper/paper.tex
intro draft completed

@@
 \newcommand{\apj}{ApJ}
+\newcommand{\aj}{AJ}
 \newcommand{\aap}{A\&A}
+\newcommand{\apss}{ApSS}
 \newcommand{\apjl}{ApJL}
@@
 % \citep{armitage11}, which directly impact planet formation theories.  
-One of the exciting developments in the study of circumstellar 
+An exciting development in the study of circumstellar 
 discs is the direct observation of large-scale, non-axisymmmetric
@@
 % planet is popular answer for spiral structure
-An attractive explanation for asymmetries in circumstellar discs is to
+The attractive explanation for asymmetries in circumstellar discs is to
 invoke disc-planet interaction. In particular, spiral structures
@@
 % but also natural in GI.
-Spiral arms are also naturally associated with global gravitational 
+However, spiral arms are also naturally associated with global gravitational 
 instability (GI) of differentially rotating discs
-\citep{goldreich65,laughlin96b,laughlin98,lodato05,forgan11}. Large-scale
+\citep{goldreich65,laughlin96b,laughlin98,nelson98,lodato05,forgan11}. Large-scale
 spiral arms may provide significant angular momentum transport to enable
 accretion \citep{lynden-bell72, papaloizou91,balbus99,lodato04}, and
-spiral structures are potentially observable with the Atacama Large 
+spiral structures due to GI are potentially observable with the Atacama Large 
 Millimeter/sub-millimeter Array \citep{cossins10,dipierror14}.
@@
 % original motivation for this work (DZGI) 
-Indeed, the original goal for this project was to clarify
-what type of GI is relevant to astrophysical discs in which only part 
-of it is self-gravitating, such as a protoplanetary disc containing a 
-massive dead zone. However, our seemingly simple initial disc models
-consistently produced one-armed spiral structures, despite not 
+Given the wide range of GIs that may occur in  gaseous 
+astrophysical discs \citep{papaloizou89,christo92,christo93,hadley11,hadley14}, it would of
+theoretical interest to clarify what type of GI is relevant to 
+partially self-gravitating discs such as a protoplanetary disc containing a 
+massive dead zone. This was indeed the original goal of this project. 
+However, our seemingly simple initial disc models
+consistently produced \emph{one-armed} spiral structures, despite not 
 satisfying the usual requirements to generate them 
@@
 %stellar wobble from indiana group 
-Single-arm, or eccentric modes in
+Single-arm spirals, or eccentric modes, corresponding to perturbations 
+with azimuthal wavenumber $m=1$, have received interest in
+the context of protoplanetary discs because of their global nature
+\citep{adams89,heemskerk92,laughlin96,tremaine01,papaloizou02,hopkins10}. Their development is 
+often associated with the unique property of $m=1$ perturbations to
+move the central star \citep{shu90}, which may have observable
+consequeces \citep{michael10}. 
+%these are self-gravitating instabilities
 
-%this work  mechanism is not strictly sg
+The aforementioned analytical studies have mostly adopted barotropic, smooth disc models. 
+In this work, we identify a mechanism to generate one-arm
+spiral patterns that is related to a non-barotropic equation of
+state and disc structure. Specifically, we find that if the disc temperature or
+sound-speed is forced to be a prescribed function of position, then 
+one-arm spiral modes grows inside a cavity that results from the
+adopted disc structure.
 
-%this paper is organised as follows
+Our study is based on direct hydrodynamic simulations using three
+different grid-based codes, both in two-dimensions (2D) and
+three-dimensions (3D). We also apply results from previous linear
+theory to explain their growth. We find that, when the disc
+temperature is prescribed, the usual statement for the conservation of
+angular momentum for linear perturbations acquires a source term
+proportional to the temperature gradient which may act to destabilise
+$m=1$ perturbations. Although we work with self-gravitating discs, the
+source term itself does not depend explicitly on disc self-gravity. Thus, the
+destabilisation effect we describe may also operate in
+non-self-gravitating discs.
+
+This paper is organised as follows. In \S\ref{model} we list the
+governing equations and describe the disc equilibria under
+consideration. \S\ref{methods} describes the hydrodynamic codes
+employed in this work and several diagnostic measures. We briefly
+review basic results from linear theory in \S\ref{wkb} to aid our
+interpretation of numerical simulations.  Our main results are presented in 
+\S\ref{results2d} for 2D simulations and \S\ref{results3d} for 3D
+simulations, and we further analyse them in \S\ref{discussions}. 
+We summarise in \S\ref{summary}, including a discussion on
+spectulations for future work.
 

### 2014-11-10 paper/results_3d.tex
intro draft completed
begun summary, updated 3d figs

@@
-\section{Three-dimensional simulations}
+\section{Three-dimensional simulations}\label{results3d}
 In this section we briefly review 3D simulations carried 
@@
 &\gamma \simeq 0.0073\Omega_k(R_0) \quad\quad \mathrm{PLUTO},\\
-&\gamma \simeq 0.0075\Omega_k(R_0) \quad\quad \varA{ZEUS-MP}.
+&\gamma \simeq 0.0084\Omega_k(R_0) \quad\quad \varA{ZEUS-MP}.
 \end{align*}
@@
     \subfigure[PLUTO]{
-      \includegraphics[scale=0.33]{figures/pdisk_021_z0}
+      \includegraphics[scale=0.33]{figures/pdisk_020_z0}
       \includegraphics[scale=0.33,clip=true,trim=2.24cm 0cm 0cm
-      0cm]{figures/pdisk_021_zmax}
+      0cm]{figures/pdisk_020_zmax}
     }

### 2014-11-10 paper/summary.tex
begun summary, updated 3d figs

@@
 \section{Summary and conclusions}\label{summary}
+In this paper, we carried out direct hydrodynamic simulations of 
+radially structured, self-gravitating locally isothermal discs. 
+We find such systems can be unstable to perturbations with azimuthal
+wavenumber $m=1$, which lead to the development of one-arm 
+trailing spirals that persist for at least $O(10^2)$ orbits. The
+spiral pattern speed is smaller than the local disc rotation, making
+it a low-frequency mode that is almost stationary.
+
+According to previous studies \citep{adams89}, our disc models
+should not support $m=1$ linear spiral instabilities. This is because
+our discs are not very massive ($M_d\lesssim 0.1M_*$) and, more
+importantly, we have neglected the motion of the central star. The 
+latter was shown to be important for one-arm spiral modes to grow  
+\citep{shu90}.
+
+In order to explain our numerical results, we appealed to  
+angular momentum conservation within linear theory. We find that a  
+forced temperature gradient introduces a torque on linear
+perturbations. This torque
+
+%works in 2D/3D
+%forced gradient feedback 
+%neutral mode destabilized
 
 \subsection{Spectulations and future work}
+
+%numerical simulations to include cooling/heating
+%non-sg sims, Be stars, eccentric discs
+%linear theory: explicit calculations

### 2014-11-10 paper/summary.tex
update

--- 2014-11-06 paper/summary.tex
@@
 \section{Summary and conclusions}\label{summary}
+In this paper, we carried out direct hydrodynamic simulations of 
+radially structured, self-gravitating locally isothermal discs. 
+We find such systems can be unstable to low-frequency perturbations
+with azimuthal wavenumber $m=1$, which lead to the development of one-arm 
+trailing spirals that persist for at least $O(10^2)$ orbits. The 
+spiral pattern speed is smaller than the local disc rotation, making
+it almost stationary.
+
+According to previous studies \citep{adams89}, our disc models
+should not support $m=1$ linear spiral instabilities. This is because
+our discs are not sufficiently massive ($M_d\lesssim 0.1M_*$) and, more
+importantly, we have surpressed the motion of the central star. The 
+latter can induce $m=1$ instabilities in massive discs  \citep{shu90}.
+
+In order to explain our numerical results, we invoked   
+angular momentum conservation within linear theory. A    
+forced temperature gradient introduces a torque on linear
+perturbations.  We showed that when the linear perturbation is a
+low-frequency $m=1$ mode,
+
+%works in 2D/3D
+%forced gradient feedback 
+%neutral mode destabilized
 
 \subsection{Spectulations and future work}
+
+%numerical simulations to include cooling/heating
+%non-sg sims, Be stars, eccentric discs
+%linear theory: explicit calculations

### 2014-11-10 paper/summary.tex
very rough first draft

--- 2014-11-06 paper/summary.tex
@@
 \section{Summary and conclusions}\label{summary}
+In this paper, we have carried out direct numerical hydrodynamic
+simulations of radially structured, self-gravitating locally
+isothermal discs.     
+We find such systems can be unstable to low-frequency perturbations
+with azimuthal wavenumber $m=1$, which lead to the development of one-arm 
+trailing spirals that persist for at least $O(10^2)$ orbits. The 
+spiral pattern speed is smaller than the local disc rotation, making
+it almost stationary. The growth rates are $O(10^{-2}\Omega)$ which
+gives a characteristic growth time of $O(10P_0)$. 
+%While this is slow
+%compared to the local rotation, it is dynamical at co-rotation.
 
+We used three independent numerical codes --- FARGO in 2D, ZEUS-MP and
+PLUTO in 3D --- to verify the growth of  
+one-arm spirals in our disc models is due to the imposed temperature
+profile and disc structure. The spiral is confined between two
+$Q$-barriers owing to a surface density bump. Growth rates increased
+with the magnitude of the imposed temperature gradient, and one-armed
+spirals did not develop in strictly isothermal simulations.  
+We find no significant vertical 
+motion in 3D simulations, which indicates the instability is
+two-dimensional.
+
+We applied angular momentum conservation within linear theory to
+explain our numerical results. A  
+forced temperature gradient introduces a torque on linear 
+perturbations. We called this the background torque because it
+represents an exchange of angular momentum between the background disc
+and the perturbations. In the local approximation, we showed that this
+background torque is negative for trailing spirals, which enforces  
+low-frequency $m=1$ modes because they are associated with negative
+angular momentum. We therefore interpret the one-arm spiral instability
+observed in our simulations as a neutral, tightly-wound $m=1$ mode
+being destabilised by the temperature gradient.
+
+%works in 2D/3D
+%forced gradient feedback 
+%neutral mode destabilized      
 \subsection{Spectulations and future work}
+%balancing heating and cooling 
+In the deeply non-linear regime, the one-arm spiral may 
+shock and dissipate, depositing negative angular momentum onto
+the background disc. The spiral amplitude would saturate by gaining
+positive angular momentum. However, if the temperature gradient is
+forced, it may be possible to achieve a balance between the gain of negative
+angular momentum through the forced temperature gradient, and the gain
+of positive angular momentum through shock dissipation. In order to
+study this possibility, improved numerical models are needed to ensure 
+total angular momentum conservation in long-term simulations.
+
+%numerical simulations to include cooling/heating
+A locally isothermal disc represents the limit of infinitely short
+cooling (and heating) timescales, so the disc temperature instantly
+returns to its inital value. %so we don't go to deeply non-linear
+                             %regime 
+This assumption can be relaxed by including an energy
+equation with thermal relaxation. Preliminary FARGO simulations
+indicate a relaxation timescale $t_c < 0.1\Omega_k^{-1}$ is needed for
+the one-arm spiral to develop. This will be presented in a follow-up
+study. %prelim sims show they are long lasting
+
+%non-sg sims, Be stars, eccentric discs
+We stress that destabilisation through the temperature gradient
+does not explicitly depend on self-gravity. In principle, this effect
+will destabilise low-frequency $m=1$ trailing spirals regardless of
+its origin. In our models, self-gravity coupled with the disc
+structure permits a neutral one-arm mode, which is then destabilised. 
+It would be of interest to explore other origins of the neutral
+one-arm spiral. One possibility is in discs around classic Be stars
+\citep{rivinius13}, for which several authors have demonstrated the
+existence of one-arm oscillations \citep[see e.g.][and references
+therein]{okasaki97,papaloizou06c,ogilvie08}. These authors employ
+strictly isothermal disc models, so it would be interesting to explore
+the effect of a temperature gradient on the stability of these
+oscillations.
+
+%linear theory: explicit calculations 
+%For a self-consistent description of the instability, one needs to
+%explicitly solve for linear stability problem.

### 2014-11-11 paper/model.tex
intro draft completed
updating 3d write up

@@
 where $c_{s0} = h R_0\Omega_k(R_0)$ and 
-$h$ is the disc aspect-ratio at $R=R_0$, 
+$h$ is the disc aspect-ratio at the reference radius $R=R_0$, 
 $\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
@@
 \end{align}
-where $R_0$ is a reference cylindrical radius and $s$ is the power-law
-index describing the smooth disc, and $\Sigma_\mathrm{ref}$ is a
+where $s$ is the power-law index describing the smooth disc and
+$\Sigma_\mathrm{ref}$ is a 
 surface density scale chosen by specifying $Q_\mathrm{out}$,
@@
 
-\section{Numerical methods}
+\section{Numerical methods}\label{methods}
 We adopt computational units such that $G=M_*=R_0=1$. Time is measured

### 2014-11-11 paper/results_3d.tex
updating 3d write up

@@
 \section{Three-dimensional simulations}\label{results3d}
-In this section we briefly review 3D simulations carried 
-out using ZEUS-MP and PLUTO. The main purpose is to check 
+In this section we review 3D simulations carried 
+out using ZEUS-MP and PLUTO. The main purpose is to verify 
 the above results with different numerical codes, and validate  
@@
 $[r_\mathrm{min},r_\mathrm{max}]=[0.4,10]R_0$ and vertical extent  
-$n_H=2$ scale-heights. The resolution is $N_r\times N_\theta\times
+$n_H=2$ scale-heights at $R=R_0$. The resolution is $N_r\times N_\theta\times
 N_\phi=256\times32\times256$. This corresponds to about $4$ cells per
-$H$in radius. 
-Because of the much reduced resolution 
+$H$in radius. Because of the much reduced resolution 
 compared to 2D, we use a smooth perturbation by setting
@@
 
-Our 3D discs are initialised in approximate equilibrium only, so we
+The 3D discs are initialised in approximate equilibrium only, so we
 first evolve the disc without perturbations using  
-$(\lmax,\mmax)=(32,0)$ up to $t=100P_0$, during which 
+$(\lmax,\mmax)=(32,0)$ up to $t=10P_0$, during which 
 meridional velocities are damped out. We then restart the simulation
-with the above perturbation and $(\lmax,\mmax)=(32,32)$.
+with the above perturbation and $(\lmax,\mmax)=(32,32)$ and damp
+merdional velocities near the radial boundaries.
 
@@
                                 %resolution?
-Results from ZEUS-MP appears to be off-set from that obtained using
+The mode amplitude from ZEUS-MP appears to be off-set from that obtained using
 PLUTO. We suspect this may be due to larger numerical noise in the
@@
 Both are smaller than the 2D simulations above. This may be due to the
-low resolutions adopted. 
+low resolutions adopted and or the small softening length employed in
+2D.  
 %growth rate does not depend on pert amp
@@
 Fig. \ref{3d_prelim} for the disc midplane and near the upper disc
-boundary. The snapshots are chosen when the spiral 
-amplitudes reached in the two codes are similar. Although the
-non-axisymmetric density has similar amplitudes at both heights, a
-coherent spiral is less evident away from the disc midplane. 
-The agreement between the two codes, as well as with the 2D
-simulations, is satisfactory.
+boundary. The snapshots are chosen when the spirals in the two codes
+have reached similar amplitudes. %The agreement between the codes,
+% as well as with the 2D simulations, is satisfactory.    
+Away from the disc midplane, a coherent spiral arm is
+not prominent, but the disturbance is radially more global, extending
+up to $R\sim 4R_0$. %implications?
 
@@
     \subfigure[ZEUS-MP]{
-      \includegraphics[scale=0.33]{figures/polarxy_dens015_z0}
-      \includegraphics[scale=0.33,clip=true,trim=2.24cm 0cm 0cm
+      \includegraphics[scale=0.382]{figures/polarxy_dens015_z0}
+      \includegraphics[scale=0.382,clip=true,trim=2.24cm 0cm 0cm
       0cm]{figures/polarxy_dens015_zmax}  
@@
     \subfigure[PLUTO]{
-      \includegraphics[scale=0.33]{figures/pdisk_020_z0}
-      \includegraphics[scale=0.33,clip=true,trim=2.24cm 0cm 0cm
+      \includegraphics[scale=0.382]{figures/pdisk_020_z0}
+      \includegraphics[scale=0.382,clip=true,trim=2.24cm 0cm 0cm
       0cm]{figures/pdisk_020_zmax}
@@
 
-\subsection{Vertical motion}
+\subsection{Vertical structure}
+Fig. \ref{3d_rz} shows the vertical structure of the one-arm
+spiral in the PLUTO run. The main spiral is confined to $z < H$.
+However, outside $R>2R_0$ we observe spirals
+
+\begin{figure}
+  \includegraphics[scale=0.47,clip=true,trim=0cm 0.79cm 0cm 0cm]{figures/pdisk_rz_020_sg.ps}
+  \includegraphics[scale=0.47,clip=true,trim=0cm 0cm 0cm 0.64cm]{figures/pdisk_rz_020_nsg.ps}
+  \caption{Non-axisymmetric density in the meridional plane from the
+    PLUTO simulation. The slices are taken at the azimuth of   
+    $\mathrm{max}[\Delta\rho(r,\pi/2,\phi)]$. Arrows correspond to the vector 
+    $(v_r/R_0,-v_\theta/rh\sin^2{\theta})$. The top (bottom) panel corresponds
+    to the inner (outer) portions of the disc. 
+    \label{3d_rz}} 
+\end{figure}
+
 We measure three-dimensionality using the parameter $\Theta$ as
@@
 %\end{figure}
+
 \subsection{Angular momentum evolution}
@@
 runs. ZEUS-MP does not conserve angular momentum very well, but
-the variation $|\Delta J/J|< O(10^{-5})$ is not significant compared
+the variation $|\Delta J/J|< O(10^{-6})$ is not significant compared
 to the individual components $|\Delta J_{0,1}/J|\sim 
-6\times10^{-5}$. The PLUTO run reaches similar values of
+5\times10^{-5}$. The PLUTO run reaches similar values of
 $|J_{0,1}|$, but achieves better conervation, with $|\Delta

### 2014-11-11 paper/discussion.tex
very rough first draft
updated external torque estimates using goldreich

@@
 simplicitly, we return to 2D discs.
+
+% According to previous studies \citep{adams89}, our disc models
+% should not support $m=1$ linear spiral instabilities. This is because
+% our discs are not sufficiently massive ($M_d\lesssim 0.1M_*$) and, more
+% importantly, we have surpressed the motion of the central star. The 
+% latter can induce $m=1$ instabilities in massive discs  \citep{shu90}.
 
@@
 sign to that of the density waves it induces in the exterior disc
-\citep{lin11b}. To estimate this effect, we port basic results from
+\citep{lin11b}. To estimate the magnitude of this effect, we port basic results from
 disc-planet theory \citep[see, e.g.][and references
@@
 
-Let us regard the one-arm spiral disturbance located in the inner disc as an  
+Let us treat the one-arm spiral disturbance located in the inner disc as an  
 external potential of the form $\Phi_\mathrm{ext}(R)\cos{\left(\phi -
-    \Omega_pt\right)}$. We expect
-it to exert a torque on the exterior disc at the Lindblad  
+    \Omega_pt\right)}$. Let $M_\mathrm{ring}$ denote the
+mass contained within $R\in[R_1,R_2]$ of the disc, where the spiral is
+confined in the numerical simulations. We take 
+\begin{align} 
+  \Phi_\mathrm{ext}
+  =-\frac{GM_\mathrm{ring}}{\bar{R}}b^{1}_{1/2}(\beta),   
+\end{align}
+where $b_{n}^m(\beta)$ is the Laplace coefficient, $\beta = R/\bar{R}$ 
+\citep{goldreich79} and $\bar{R} = (R_1+R_2)/2$ is the
+approximate location of the spiral.
+
+%and here we are interested in the
+%potential it produces at $R>\bar{R}$  
+%This potential is associated with a mass
+%$M_\mathrm{ring}$ contained in   is the mass in $R\in[R_1,R_2]$.
+
+%For an order-of-magnitude exercise, we simply take 
+%\begin{align}
+ % \Phi_\mathrm{ext} = -\frac{GM_\mathrm{ring}}{|R - \bar{R}|}, 
+%\end{align}
+%for the potential associated with a disturbance located at
+%$R=\bar{R}$, where $M_\mathrm{ring}$
+
+We expect the external potential to exert a torque at the Lindblad  
 and co-rotation resonances. At the outer Lindblad resonance (OLR), this
@@
 
-To order-of-magnitude, we reasonably take
-\begin{align}
-  \Phi_\mathrm{ext} = -\frac{GM_\mathrm{ring}}{|R - \bar{R}|}, 
-\end{align}
-for the potential associated with a disturbance located at
-$R=\bar{R}$, where $M_\mathrm{ring}$ is the mass in $R\in[R_1,R_2]$.
 If we also associate its angular momentum with 
 $J_\mathrm{ext}  = M_\mathrm{ring}\bar{R}^2\Omega_p$, we can obtain a
-rate $\gamma_L$ by evaluating $\Gamma_L/J_\mathrm{ext}$. Then 
+rate $\gamma_L=\Gamma_L/J_\mathrm{ext}$. Then 
 \begin{align}
-  \frac{\gamma_L}{\Omega_p} \sim &\frac{\pi h}{3
-    Q_L}\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_L}{\bar{R}}\right)^{-1/2}\left(\frac{R_c}{\bar{R}}\right)^{5/2}\notag\\
-  &\times \left\{ \frac{R_L}{R_L - \bar{R}} - 2\left[1 -
-      \left(\frac{R_c}{R_L}\right)^{-3/2}\right]\right\}^2\frac{R_c^2}{\left(R_L
-    - \bar{R}\right)^2}. 
+  \frac{\gamma_L}{\Omega_p} = &\frac{\pi
+    h}{3Q_L}\left(\frac{M_p}{M_*}\right)\left(\frac{R_L}{\bar{R}}\right)\left(\frac{R_c}{\bar{R}}\right)^3\left(\frac{R_L}{R_c}\right)^{-3/2}\notag\\
+  &\times \left\{\frac{R_L}{\bar{R}}\left.\frac{db_{1/2}^1}{d\beta}\right|_L + 2\left[1 -
+      \left(\frac{R_c}{R_L}\right)^{-3/2}\right]b_{1/2}^1(\beta_L)\right\}^2.
 \end{align}
+
+%\begin{align}
+ % \frac{\gamma_L}{\Omega_p} \sim &\frac{\pi h}{3
+  %  Q_L}\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_L}{\bar{R}}\right)^{-1/2}\left(\frac{R_c}{\bar{R}}\right)^{5/2}\notag\\
+ % &\times \left\{ \frac{R_L}{R_L - \bar{R}} - 2\left[1 -
+  %    \left(\frac{R_c}{R_L}\right)^{-3/2}\right]\right\}^2\frac{R_c^2}{\left(R_L
+   % - \bar{R}\right)^2}. 
+%\end{align}
+
 Inserting $h=0.05$, $Q_L=10$, $M_\mathrm{ring} = 0.05M_*$,
@@
 FARGO simulation, we get
+
 \begin{align}
-  \gamma_L \sim 0.012\Omega_p. 
+  \gamma_L \sim 0.0005\Omega_p. 
 \end{align}
-This represents an upper limit on the effect of the OLR on the spiral
-disturbance, and is much smaller than the growth rate measured in the
-simulation ($\gamma\sim0.1\Omega_p$). We conclude that for our models,
-the OLR has negligible effect on the one-arm spiral.
+
+%\begin{align}
+ % \gamma_L \sim 0.012\Omega_p. 
+%\end{align}
+
+%This represents an upper limit on the effect of the OLR on the spiral
+%disturbance, and is much smaller than the growth rate measured in the
+%simulation ($\gamma\sim0.1\Omega_p$). We conclude that for our models,
+%the OLR has negligible effect on the one-arm spiral.
 
@@
 \begin{align}
-  \frac{\Gamma_C}{J_\mathrm{ext}\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_C} \left(s -
-    \frac{3}{2}\right)\left(\frac{M_\mathrm{ring}}{M_*}\right)\frac{R_c^4}{\bar{R}^2\left(R_c
-      - \bar{R}\right)^2}.   
+\frac{\gamma_c}{\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_c}
+\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_c}{\bar{R}}\right)^4\left(s-\frac{3}{2}\right) \left[b_{1/2}^1(\beta_c)\right]^2 
 \end{align}
+
+%\begin{align}
+%  \frac{\Gamma_C}{J_\mathrm{ext}\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_C} \left(s -
+ %   \frac{3}{2}\right)\left(\frac{M_\mathrm{ring}}{M_*}\right)\frac{R_c^4}{\bar{R}^2\left(R_c
+ %     - \bar{R}\right)^2}.   
+%\end{align}
+
 Using the above parameter values with $s=2$ and $Q_c=10$, we obtain a
@@
 \begin{align}
-  \gamma_C\sim 0.010\Omega_p,
+  \gamma_c\sim 0.0006\Omega_p,
 \end{align}
+
+%\begin{align}
+%  \gamma_C\sim 0.010\Omega_p,
+%\end{align}
 which is again small compared to the measured growth rates.

### 2014-11-11 paper/paper.tex
very rough first draft
updated external torque estimates using goldreich

@@
 \newcommand{\aap}{A\&A}
+\newcommand{\aapr}{A\&ARv}
 \newcommand{\apss}{ApSS}
@@
 spiral arms 
-\citep{muto12,grady13,christiaens14,avenhaus14,boccaletti14}.
+\citep{hashimoto11,muto12,boccaletti14,grady13,christiaens14,avenhaus14}.
 
 % planet is popular answer for spiral structure
-The attractive explanation for asymmetries in circumstellar discs is to
-invoke disc-planet interaction. In particular, spiral structures
+The attractive explanation for asymmetries in circumstellar discs is
+disc-planet interaction. In particular, spiral structures  
 naturally arise from the gravitational interaction between a
@@
 Given the wide range of GIs that may occur in  gaseous 
-astrophysical discs \citep{papaloizou89,christo92,christo93,hadley11,hadley14}, it would of
+astrophysical discs
+\citep{papaloizou89,christo92,christo93,hadley11,hadley14}, it is of 
 theoretical interest to clarify what type of GI is relevant to 
-partially self-gravitating discs such as a protoplanetary disc containing a 
-massive dead zone. This was indeed the original goal of this project. 
-However, our seemingly simple initial disc models
-consistently produced \emph{one-armed} spiral structures, despite not 
+partially self-gravitating discs, such as a protoplanetary disc containing a 
+massive dead zone.  %This was indeed the original goal of this project. 
+This work presents a first step towards this goal. We begin with
+the simplest disc models which, surprisingly, still produced
+unexpected results. Specifically, our seemingly simple initial disc models
+consistently developed \emph{one-armed} spiral structures, despite not 
 satisfying the usual requirements to generate them 
-(see below). We therefore dedicate this study to verify their origin.
+(see below).
+
+%through this work we identify a new method to generate one arm
+%spirals 
+%We therefore dedicate this study to verify their origin.
 
@@
 
-The aforementioned analytical studies have mostly adopted barotropic, smooth disc models. 
+The aforementioned analytical studies have adopted barotropic, smooth disc models. 
 In this work, we identify a mechanism to generate one-arm
 spiral patterns that is related to a non-barotropic equation of
-state and disc structure. Specifically, we find that if the disc temperature or
-sound-speed is forced to be a prescribed function of position, then 
-one-arm spiral modes grows inside a cavity that results from the
+state and disc structure. We find that if the disc temperature or
+sound-speed is a prescribed function of position, then 
+one-arm spiral modes can grow inside a cavity that results from the
 adopted disc structure.
@@
 angular momentum for linear perturbations acquires a source term
-proportional to the temperature gradient which may act to destabilise
+proportional to the temperature gradient, which may act to destabilise
 $m=1$ perturbations. Although we work with self-gravitating discs, the

### 2014-11-11 paper/results_3d.tex
updated external torque estimates using goldreich

@@
 Fig. \ref{3d_ampmax} plots the amplitude of the $m=1$ spirals measured
-in the ZEUS-MP and PLUTO runs. We also show results from strictly
-isothermal discs ($q=0$), which clearly display no growth. %why no high m
+in the ZEUS-MP and PLUTO runs. 
+%why no high m
                                 %modes as in 2D for iso disc?
                                 %resolution?
-The mode amplitude from ZEUS-MP appears to be off-set from that obtained using
-PLUTO. We suspect this may be due to larger numerical noise in the
-ZEUS-MP code. Thus the $m=1$ spiral grows to a larger amplitude by the
-end of the run. However, we measure similar growth rates,
+Results from ZEUS-MP appears to be off-set from that obtained using
+PLUTO.  
+%We suspect this is due to larger numerical noise in the
+%ZEUS-MP code, 
+In ZEUS-MP we observe disturbances developed near
+the inner boundary, which is probably responsible for the initial growth
+for $t<70P_0$. This growth is likely a numerical artifact and effectively seeds
+the ZEUS-UP simulation with a larger initial perturbation.  However,
+once the coherent $m=1$ spiral begins to grow ($t\gtrsim 100P_0$), 
+we measure similar growth rates in both codes: 
 \begin{align*}
@@
 \end{align*}
-Both are smaller than the 2D simulations above. This may be due to the
-low resolutions adopted and or the small softening length employed in
-2D.  
+Both are smaller than the 2D simulations. This may be due to the
+low resolutions adopted and/or the small softening length employed in
+2D.  Fig. \ref{3d_ampmax} also show results from strictly isothermal
+simulations, which display no growth compared to cases with a 
+temperature gradient.
+
+% We also show results from strictly
+% isothermal discs ($q=0$). 
 %growth rate does not depend on pert amp
@@
   \caption{Evolution of the maximum $m=1$ density in  $r\in[R_1,R_2]$
-    measured in the 3D simulations. Both discs with a imposed
+    measured in the 3D simulations. Results from discs with a imposed 
     temperature gradient ($q=1$) and a strictly isothermal disc
@@
 boundary. The snapshots are chosen when the spirals in the two codes
-have reached similar amplitudes. %The agreement between the codes,
-% as well as with the 2D simulations, is satisfactory.    
+have reached similar amplitudes. The agreement between the codes,
+as well as with the 2D simulations, is satisfactory.    
 Away from the disc midplane, a coherent spiral arm is
 not prominent, but the disturbance is radially more global, extending
-up to $R\sim 4R_0$. %implications?
+up to $R\sim 4R_0$. 
+%implications?
 
@@
 Fig. \ref{3d_rz} shows the vertical structure of the one-arm
-spiral in the PLUTO run. The main spiral is confined to $z < H$.
-However, outside $R>2R_0$ we observe spirals
+spiral in the PLUTO run. The main spiral is confined to $z < H$. Thus,
+a 2D disc model, representing dynamics near the disc midplane, 
+should be sufficient to study the instability. Interestingly, however, for $R>2R_0$ the
+spirals  increase in amplitude away from the midplane. It remains
+small  in our disc mode ($|\Delta\rho| < 0.1$) l, but could become 
+significant with a larger vertical domain. This shows that 
+the one-arm spiral can have a global influence.
 
@@
 
-We measure three-dimensionality using the parameter $\Theta$ as
-defined in Eq. \ref{theta}.  Vertical velocities
-do increase with the growth of the one-arm spiral, 
+Although Fig. \ref{3d_rz} appears to show significant vertical motion,
+the three-dimensionality parameter $\Theta < 10^{-2}$
+(Eq. \ref{theta}), implying the kinetic energy associated with
+vertical motion is insignificant compared to that in horizontal
+motion. We find  $|v_z/c_s|\lesssim
+0.2$ for $h=0.05$. Thus, vertical motions are sub-sonic.  
+%This again supports the 2D approximation, altthou
+
+%Vertical velocities
+%do increase with the growth of the one-arm spiral, 
 %Larger amplitudes
 %are reached in ZEUS because of larger numerical noise. 
-but we find  $\Theta \lesssim 10^{-2}$. This implies $|v_z/c_s|\lesssim
-0.2$ for $h=0.05$. Thus, vertical motions are sub-sonic and negligible
-compared to horizontal motions. This supports the 2D approximation.
+%but we find  $\Theta \lesssim 10^{-2}$.
 
@@
 \subsection{Angular momentum evolution} 
-
 Fig. \ref{3d_angmom} shows the angular momentum evolution in the 3D
-runs. ZEUS-MP does not conserve angular momentum very well, but
-the variation $|\Delta J/J|< O(10^{-6})$ is not significant compared
+runs. These plots are again similar to the 2D runs, i.e. angular
+momentum lost by $J_1$ is gained by $J_0$. ZEUS-MP does not conserve
+angular momentum very well, but the variation $|\Delta J/J|<
+O(10^{-6})$ is small compared 
 to the individual components $|\Delta J_{0,1}/J|\sim 
@@
     simulations. The perturbation 
-    relative to $t=100P_0$, during the growth of the one-arm spiral,
+    relative to $t=10P_0$, during the growth of the one-arm spiral,
     is shown in units of the initial total angular momentum

### 2014-11-11 paper/summary.tex
updated external torque estimates using goldreich

@@
 observed in our simulations as a neutral, tightly-wound $m=1$ mode
-being destabilised by the temperature gradient.
+being destabilised by the imposed temperature gradient.
 

### 2014-11-12 paper/discussion.tex
minor updates

@@
 One possible effect of self-gravity is to allow the one-arm spiral  
-to act as an external potential for the wider disc. This effect is
+to act as an external potential for the wider disc. This is
 analogous to disc-satellite interaction 
@@
 sign to that of the density waves it induces in the exterior disc
-\citep{lin11b}. To estimate the magnitude of this effect, we port basic results from
-disc-planet theory \citep[see, e.g.][and references
+\citep{lin11b}. Here, we estimate the magnitude of this effect using
+basic results from disc-planet theory \citep[see, e.g.][and references 
 therein]{papaloizou07}.
 
-Let us treat the one-arm spiral disturbance located in the inner disc as an  
+Let us treat the one-arm spiral confined in $R\in[R_1,R_2]$ as an  
 external potential of the form $\Phi_\mathrm{ext}(R)\cos{\left(\phi -
-    \Omega_pt\right)}$. Let $M_\mathrm{ring}$ denote the
-mass contained within $R\in[R_1,R_2]$ of the disc, where the spiral is
-confined in the numerical simulations. We take 
+    \Omega_pt\right)}$. We take 
 \begin{align} 
   \Phi_\mathrm{ext}
-  =-\frac{GM_\mathrm{ring}}{\bar{R}}b^{1}_{1/2}(\beta),   
+  =-\frac{GM_\mathrm{ring}}{\overline{R}}b^{1}_{1/2}(\beta),   
 \end{align}
-where $b_{n}^m(\beta)$ is the Laplace coefficient, $\beta = R/\bar{R}$ 
-\citep{goldreich79} and $\bar{R} = (R_1+R_2)/2$ is the
-approximate location of the spiral.
+where $M_\mathrm{ring}$ is the disc mass contained within
+$R\in[R_1,R_2]$, $\overline{R} = (R_1+R_2)/2$ is the approximate radial
+location of the spiral, $b_{n}^m(\beta)$ is the Laplace coefficient
+and $\beta = R/\overline{R}$. This form of
+$\Phi_\mathrm{ext}$ is the $m=1$ component of the gravitational
+potential of an external satellite on a circular orbit
+\citep{goldreich79}.
 
 %and here we are interested in the
-%potential it produces at $R>\bar{R}$  
+%potential it produces at $R>\overline{R}$  
 %This potential is associated with a mass
@@
 %\begin{align}
- % \Phi_\mathrm{ext} = -\frac{GM_\mathrm{ring}}{|R - \bar{R}|}, 
+ % \Phi_\mathrm{ext} = -\frac{GM_\mathrm{ring}}{|R - \overline{R}|}, 
 %\end{align}
 %for the potential associated with a disturbance located at
-%$R=\bar{R}$, where $M_\mathrm{ring}$
+%$R=\overline{R}$, where $M_\mathrm{ring}$
 
-We expect the external potential to exert a torque at the Lindblad  
-and co-rotation resonances. At the outer Lindblad resonance (OLR), this
-torque is 
+We expect the external potential to exert a torque on the disc at the
+Lindblad and co-rotation resonances. At the outer Lindblad resonance
+(OLR), this torque is 
 \begin{align}
@@
 
-If we also associate its angular momentum with 
-$J_\mathrm{ext}  = M_\mathrm{ring}\bar{R}^2\Omega_p$, we can obtain a
+If we associate the external potential with angular momentum 
+$J_\mathrm{ext}  = M_\mathrm{ring}\overline{R}^2\Omega_p$, we can obtain a
 rate $\gamma_L=\Gamma_L/J_\mathrm{ext}$. Then 
@@
   \frac{\gamma_L}{\Omega_p} = &\frac{\pi
-    h}{3Q_L}\left(\frac{M_p}{M_*}\right)\left(\frac{R_L}{\bar{R}}\right)\left(\frac{R_c}{\bar{R}}\right)^3\left(\frac{R_L}{R_c}\right)^{-3/2}\notag\\
-  &\times \left\{\frac{R_L}{\bar{R}}\left.\frac{db_{1/2}^1}{d\beta}\right|_L + 2\left[1 -
+    h}{3Q_L}\left(\frac{M_p}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)\left(\frac{R_c}{\overline{R}}\right)^3\left(\frac{R_L}{R_c}\right)^{-3/2}\notag\\
+  &\times \left\{\frac{R_L}{\overline{R}}\left.\frac{db_{1/2}^1}{d\beta}\right|_L + 2\left[1 -
       \left(\frac{R_c}{R_L}\right)^{-3/2}\right]b_{1/2}^1(\beta_L)\right\}^2.
 \end{align}
-
 %\begin{align}
  % \frac{\gamma_L}{\Omega_p} \sim &\frac{\pi h}{3
-  %  Q_L}\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_L}{\bar{R}}\right)^{-1/2}\left(\frac{R_c}{\bar{R}}\right)^{5/2}\notag\\
- % &\times \left\{ \frac{R_L}{R_L - \bar{R}} - 2\left[1 -
+  %  Q_L}\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)^{-1/2}\left(\frac{R_c}{\overline{R}}\right)^{5/2}\notag\\
+ % &\times \left\{ \frac{R_L}{R_L - \overline{R}} - 2\left[1 -
   %    \left(\frac{R_c}{R_L}\right)^{-3/2}\right]\right\}^2\frac{R_c^2}{\left(R_L
-   % - \bar{R}\right)^2}. 
+   % - \overline{R}\right)^2}. 
 %\end{align}
-
 Inserting $h=0.05$, $Q_L=10$, $M_\mathrm{ring} = 0.05M_*$,
-$R_L=7.2R_0$, $R_c=4.4R_0$ and $\bar{R}=1.5R_0$ from our fiducial
+$R_L=7.2R_0$, $R_c=4.4R_0$ and $\overline{R}=1.5R_0$ from our fiducial
 FARGO simulation, we get
-
 \begin{align}
-  \gamma_L \sim 0.0005\Omega_p. 
+  \gamma_L \sim 5\times10^{-4}\Omega_p. 
 \end{align}
-
 %\begin{align}
@@
 %\end{align}
-
 %This represents an upper limit on the effect of the OLR on the spiral
@@
 %the OLR has negligible effect on the one-arm spiral.  
-
 For the co-rotation torque, we use the result
 \begin{align}
-  \Gamma_C = \left.
-    \pi^2\Phi_\mathrm{ext}^2\left(\frac{d\Omega}{dR}\right)^{-1}\frac{d}{dR}\left(\frac{2\Sigma}{\Omega}\right)\right|_{C}
+  \Gamma_c = \left.
+    \pi^2\Phi_\mathrm{ext}^2\left(\frac{d\Omega}{dR}\right)^{-1}\frac{d}{dR}\left(\frac{2\Sigma}{\Omega}\right)\right|_{c}   
 \end{align}
-for a Keplerian disc, where subscript $C$ denotes evaluation at
+for a Keplerian disc, where subscript $c$ denotes evaluation at
 co-rotation radius $R=R_c$. For a power-law surface density profile
@@
 \frac{\gamma_c}{\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_c}
-\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_c}{\bar{R}}\right)^4\left(s-\frac{3}{2}\right) \left[b_{1/2}^1(\beta_c)\right]^2 
+\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_c}{\overline{R}}\right)^4\left(s-\frac{3}{2}\right)
+\left[b_{1/2}^1(\beta_c)\right]^2   
 \end{align}
-
 %\begin{align}
 %  \frac{\Gamma_C}{J_\mathrm{ext}\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_C} \left(s -
- %   \frac{3}{2}\right)\left(\frac{M_\mathrm{ring}}{M_*}\right)\frac{R_c^4}{\bar{R}^2\left(R_c
- %     - \bar{R}\right)^2}.   
+ %   \frac{3}{2}\right)\left(\frac{M_\mathrm{ring}}{M_*}\right)\frac{R_c^4}{\overline{R}^2\left(R_c
+ %     - \overline{R}\right)^2}.   
 %\end{align}
-
 Using the above parameter values with $s=2$ and $Q_c=10$, we obtain a
@@
 \begin{align}
-  \gamma_c\sim 0.0006\Omega_p,
+  \gamma_c\sim 6\times 10^{-4}\Omega_p. 
 \end{align}
+
+These estimates are much smaller than that due to
+the imposed temperature gradient as measured in the simulations
+$\gamma\sim 0.1\Omega_p$). We conclude that for our disc models, the
+Lindblad and co-rotation resonances have negligible effect on the
+growth of the spiral arm in the inner disc. We confirmed this with
+additional FARGO simulations with a reduced radial domain size,
+setting $R_\mathrm{max}=3R_0$, which still developed the one-arm
+spiral.
+
+% both are positive -> destabilizing in principle
 
@@
 %\end{align}
-which is again small compared to the measured growth rates.
+%which is again small compared to the measured growth rates.
 

### 2014-11-12 paper/model.tex
minor updates

@@
 We consider a self-gravitating, inviscid fluid disc of mass $M_d$ 
-orbting a central star of mass $M_*$. We will mainly examine two-dimensional
-(2D, or razor-thin) discs in favour of numerical 
-resolution, but have also carried out some three-dimensional (3D) disc
+orbting a central star of mass $M_*$. We will mainly examine 2D
+(or razor-thin) discs in favour of numerical 
+resolution, but have also carried out some 3D disc
 simulations. Thus, for generality we describe the system in
@@
 \end{align}
-where $\rho$ is the mass density, $\bm{v}=(v_r,v_\theta,v_\phi)$ the
-fluid velocity, $p$ is the pressure and the total potential
+where $\rho$ is the mass density, $\bm{v}$ is the 
+fluid velocity (the angular velocity being $\Omega \equiv v_\phi/R$), 
+$p$ is the pressure and the total potential
 $\Phi_\mathrm{tot} = \Phi_* + \Phi_d$ consists of that from the
@@
 \end{align}
-where the sound-speed is given by 
+where the sound-speed $c_s$ is given by 
 \begin{align}\label{sound-speed}
@@
 central star. For the adopted disc masses $M_d\lesssim 0.1 M_*$, this
-effect is not expected to be significant \citep{adams89,shu90}. Indeed,
-simulations in the early stages of this project included the indirect
-potential, and yield similar results.
+effect is not expected to be significant
+\citep{adams89,shu90}. Indeed, simulations including the indirect
+potential, carried out in the early stages of this project, produced
+similar results.
 
+%isolated system 
 % ang mom cons
@@
 mass density $\Sigma$, $p$ becomes the vertically-integrated pressure, 
-and the 2D fluid
-velocity $\bm{v}=(v_R,v_\phi)$ is evaluated at the midplane, as are
-the forces from the potential. In the Poisson equation, $\rho$ is
+and the 2D fluid velocity $\bm{v}$ is evaluated at the midplane, as are
+the forces in the momentum equations. In the Poisson equation, $\rho$ is
 replaced by $\Sigma\delta(z)$, where $\delta(z)$ is the
@@
 \end{align}
-where $\Delta_1 = \delta R \times H(R_{1})$ and similarly for
-$R_{2}$. The bump function mimicks mass built-up at some location in 
-the disc.
+where $\Delta_{1,2} = \delta R \times H(R_{1,2})$. The bump function
+mimicks mass built-up in $R\in[R_1,R_2]$.
 
@@
 $Q_\mathrm{out}=2$. An example of the initial surface density and the
-Toomre parameter (see\S\ref{wkb}) is shown in
-Fig. \ref{initial_surf}. The transition from the self-gravitating and
+Toomre $Q$ parameter (see\S\ref{wkb}) is shown in
+Fig. \ref{initial_surf}. Since $Q>1$, the disc is stable to local
+axisymmetric perturbations \citep{toomre64}. %min Q is 1.3
+The transition between self-gravitating and 
 non-self-gravitating portions of the disc occur smoothly across
@@
 \end{align}
-and similarly in 2D. The angular velocity is $\Omega \equiv v_\phi/R$.
+and similarly in 2D.
 
@@
 \subsection{FARGO}
-Our primary tool is FARGO with self-gravity \citep{baruteau08}. This
+Our primary code is FARGO with self-gravity \citep{baruteau08}. This
 is a popular, simple finite-difference code for 2D discs. `FARGO' refers
@@
 divided into $(N_R,N_\phi)$ grids, logarithimically spaced in radius and
-uniform spaced in azimuth. At radial boundaries we set the
+uniformly spaced in azimuth. At radial boundaries we set the
 hydrodynamic variables to their initial values.
@@
 
-The normalised surface density associated with $m$ is
+The normalised surface density with azimuthal wavenumber $m$ is
 \begin{align}
@@
 assuming $\Sigma_m\propto\exp{(-\ii \sigma t)}$, where
-$\sigma = \omega + \ii\gamma$ is the complex frequency, $\omega$ is
+$\sigma = \omega + \ii\gamma$ is a complex frequency, $\omega$ being 
 the real frequency and $\gamma$ is the growth rate. The total
@@
 \end{align}
-where $^*$ denotes complex conjugation. We shall refer to $J_m$ as the
+where $^*$ denotes complex conjugation. We will refer to $J_m$ as the
 $m^\mathrm{th}$ component of the total angular momentum, and use it to
@@
 \end{align}
-and similarly for the horizontal velocities. The radial range of
+and similarly for the horizontal velocities. Thus $\Theta$ is the
+ratio of the average kinetic energy associated with vertical motion to
+that in horizontal motion. The radial range of
 integration is taken over $r\in[R_1,R_2]$ since this is where
@@
 hydrodynamic simulations. However, it will be useful to employ
-existing linear density wave theory to interpret the simulations.
+linear density wave theory to interpret the simulations.
 We list here the main definitions and results from 2D linear theory,
@@
 the perturbation is conserved, provided that there is no net angular
-momeumtum flux.
+momeumtum flux and the perturbations are regular.
 
@@
 In the local approximation, perturbations are assumed to vary rapidly
-relative to any background gradients. \cite{shu91} gives the
+relative to any background gradients. The
 dispersion relation for tightly-wound density 
 waves of the form $\exp{[\ii(-\sigma t + m \phi + kR)]}$ in a razor-thin
-disc as  
+disc is  
 \begin{align}\label{dispersion}
@@
 \end{align}
-and $k$ is a real wavenumber such that $|kR|\gg1$. Note that in this
-approximation, only axisymmetric perturbations with $m=0$ can be
-unstable (having an imaginary $\sigma$).
+where $k$ is a real wavenumber such that $|kR|\gg1$ \citep{shu91}. 
+Note that in this approximation, only axisymmetric perturbations with
+$m=0$ can be unstable (having an imaginary $\sigma$).
 
-Considering a neutral mode, given the real frequency $\omega$ or
-pattern speed $\Omega_p\equiv 
-\omega/m$, Eq. \ref{dispersion} can be solved 
+Given the real frequency $\omega$ or pattern speed $\Omega_p\equiv  
+\omega/m$ of a neutral mode 
+, Eq. \ref{dispersion} can be solved 
 for $|k|$, 
@@
 
-A detailed discussion of the properties of density waves in the local
-approximation can be found in \cite{shu91}. An important result is that
+A detailed discussion of the properties of local density waves 
+is given in \cite{shu91}. An important result is that
 waves interior to co-rotation ($R<R_c$) have negative angular momentum, while

### 2014-11-12 paper/paper.tex
minor updates

@@
   discs we observe the development of a 
-  spiral mode with azimuthal wavenumber $m=1$ with co-rotation radius 
+  spiral mode with azimuthal wavenumber $m=1$ and co-rotation radius 
   outside the bulk of the spiral arm, resulting in a nearly stationary 
@@
   the central star, as found in previous studies. Instead, we find it 
-  to be associated with the imposed 
+  is associated with the imposed 
   temperature gradient and the radial disc structure. 
-  We apply results from existing linear theory and attribute the
+  We apply results from linear theory and attribute the
   growth of this spiral to the non-conservation of angular momentum
@@
   and the background disc. However, because the interaction does
-  not explicitly depend on self-gravity, we speculate it may also
+  not explicitly depend on self-gravity, we suggest it may also
   affect low-frequency one-arm spirals in non-self-gravitating
@@
 % but also natural in GI.
-However, spiral arms are also naturally associated with global gravitational 
+However, spiral arms are also associated with global gravitational 
 instability (GI) of differentially rotating discs
 \citep{goldreich65,laughlin96b,laughlin98,nelson98,lodato05,forgan11}. Large-scale
-spiral arms may provide significant angular momentum transport to enable
-accretion \citep{lynden-bell72, papaloizou91,balbus99,lodato04}, and
+spiral arms can provide significant angular momentum transport
+neccessary for accretion \citep{lynden-bell72, papaloizou91,balbus99,lodato04}, and
 spiral structures due to GI are potentially observable with the Atacama Large 
@@
 GI can be expected in the earliest stages of 
-circumstellar disc formation \citep{kratter10b,inutsuka10,tsukamoto13}. %also
+circumstellar disc formation
+\citep{kratter10b,inutsuka10,tsukamoto13}, and may be possible in the
+outer parts of the disc \citep{rafikov05,matzner05,kimura12}.  
 % outer parts  
 In the case of protoplanetary discs, which have complex structure
-\citep{armitage10}, GI may also occur in `dead
+\citep{armitage10}, GI can also occur in `dead
 zones' --- sub-portions of the disc where the magneto-rotational
@@
 \citep{gammie96,turner08,landry13} and therefore has reduced mass
-accretion. Dead zones may become
+accretion. Dead zones may then become
 self-gravitating with sufficient mass built-up
@@
 % original motivation for this work (DZGI) 
-Given the wide range of GIs that may occur in  gaseous 
+A wide range of GIs can occur in  gaseous 
 astrophysical discs 
-\citep{papaloizou89,christo92,christo93,hadley11,hadley14}, it is of 
-theoretical interest to clarify what type of GI is relevant to 
+\citep{papaloizou89,christo92,christo93,hadley11,hadley14}. It is then
+of theoretical importance to clarify what type of GI is relevant to 
 partially self-gravitating discs, such as a protoplanetary disc containing a 
 massive dead zone.  %This was indeed the original goal of this project. 
-This work presents a first step towards this goal. We begin with
-the simplest disc models which, surprisingly, still produced
-unexpected results. Specifically, our seemingly simple initial disc models
-consistently developed \emph{one-armed} spiral structures, despite not 
-satisfying the usual requirements to generate them 
-(see below).
+This work presents a first step towards this goal. Specifically, we 
+study a class of seemingly simple disc models which were found to 
+consistently produce \emph{one-armed} spiral structures, despite not 
+satisfying the usual requirements to generate them.
 
@@
 the context of protoplanetary discs because of their global nature
-\citep{adams89,heemskerk92,laughlin96,tremaine01,papaloizou02,hopkins10}. Their development is 
-often associated with the unique property of $m=1$ perturbations to
-move the central star \citep{shu90}, which may have observable
-consequeces \citep{michael10}. 
+\citep{adams89,heemskerk92,laughlin96,tremaine01,papaloizou02,hopkins10}. 
+In the `SLING' mechanism proposed by \cite{shu90}, the $m=1$ spiral
+instability is attributed to motion of the central star induced by the
+one-arm perturbation and requires a massive disc \citep[the former may
+have observable consequences, ][]{michael10}. However, our numerical
+models do not have these features.
+
 %these are self-gravitating instabilities
@@
 different grid-based codes, both in two-dimensions (2D) and
-three-dimensions (3D). We also apply results from previous linear
+three-dimensions (3D). We also apply results from linear
 theory to explain their growth. We find that, when the disc
@@
 proportional to the temperature gradient, which may act to destabilise
-$m=1$ perturbations. Although we work with self-gravitating discs, the
-source term itself does not depend explicitly on disc self-gravity. Thus, the
-destabilisation effect we describe may also operate in
+$m=1$ perturbations. Although we work with self-gravitating discs,
+this source term does not depend explicitly on disc self-gravity. Thus, the
+destabilisation effect we describe should also be applicable to
 non-self-gravitating discs.
@@
 simulations, and we further analyse them in \S\ref{discussions}. 
-We summarise in \S\ref{summary}, including a discussion on
+We summarise in \S\ref{summary}, with a discussion on
 spectulations for future work.

### 2014-11-12 paper/results.tex
update
intro draft completed
minor updates

@@
-\section{Results}
+\section{Results}\label{results2d}
 We first present results from FARGO simulations. The 2D disc spans
@@
 low $m\leq 2$ modes decay initially, but grows between $t\in[20,40]P_0$,
-possibily due to saturation of the high-$m$ modes  
+possibily due to non-linear interaction of the high-$m$ modes  
 \citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
@@
   \includegraphics[width=\linewidth]{figures/nonaxi_evol_DZ_fargo}
-  \caption{Evolution of maxima in non-axisymmetric surface density 
-    in the fiducial FARGO simulation initialised with perturbations
+  \caption{Evolution of non-axisymmetric surface density maxima 
+    in the FARGO simulation initialised with perturbations
     with $m\in[1,10]$.\label{fargo_modeamp}} 
@@
   0cm 0cm 0cm]{figures/polarxy_dens130} 
-  \caption{Visualisation of the fiducial FARGO 2D simulation in
+  \caption{Visualisation of the FARGO 2D simulation in
     Fig. \ref{fargo_modeamp}. The total  
@@
 exactly. However, we find the total angular momentum varies by 
-$|\Delta J/J|= O(10^{-6})$, and is much smaller in magnitude than the
+$|\Delta J/J|= O(10^{-6})$, and is much smaller than the
 change in the angular momenta components, $|\Delta J_{0,1}/J|>
@@
   \includegraphics[width=\linewidth]{figures/nonaxi_evol_ang_fargo}
-  \caption{Evolution of angular momentum components in the fiducial
+  \caption{Evolution of angular momentum components in the 
     FARGO simulation in Fig. \ref{fargo_modeamp}---\ref{fargo_2d}. The
@@
 In order to focus on the $m=1$ spiral that eventually emerges, from
-now on we examine simulations initialised with
+now on we examine simulations initialised 
 $M=1$ in Eq. \ref{randpert}. The disc 
-evolution is largely similar to the previous case, but the final spiral is
+evolution is largely similar to the previous case, but the one-armed spiral is
 more coherent, which facilitates the analysis below.
@@
 Fig. \ref{2d_fargo_viz} shows the $m=1$ surface density
-at the end of this simulation. For this one-arm spiral
-we measure a co-rotation radius and growth rate of 
+at the end of this simulation. For this one-armed spiral
+we measure a co-rotation radius and growth rate  
 \begin{align*}
   &R_c \simeq 4.4R_0,\\
-  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_k(R_c). 
+  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_p. 
 \end{align*}
@@
 $|\Sigma_1|$ varies slowly compared to the complex phase. This is the
-main assumption in local theory. We measure $k$ numerically and plot
+main assumption in local theory. We measured $k$ numerically and plot
 its normalised value in Fig. \ref{fargo_wavenumber}. We find 
@@
 h^{-1}$. Since $Q=O(1)$ and $h\ll 1$ imply $|kR|\gg 1$, we 
-can apply results from local theory self-consistently. Recall $k>0$
-corresponds to a trailing spiral, as seen in Fig. \ref{2d_fargo_viz}.
+can apply results from local theory
+self-consistently. Fig. \ref{2d_fargo_viz} shows the $m=1$ spiral is
+trailing, which is consistent with $k>0$.
 
@@
 Q^{-2}\simeq 0.1\ll 1$, as required for consistency with
-Eq. \ref{dispersion} (with the lower sign taken, i.e. long waves).  
+Eq. \ref{dispersion} with the lower sign taken (i.e. long waves).  
 There is one outer Lindblad resonance at $R_L\simeq
-7.2R_0$. Thus, waves may be launched in $R\gtrsim 7.2R_0$ by the  
-spiral disturbance in the inner disc \citep{lin11b}. 
-%this would be a de-stabilising
-% effect.  we discuss this in discussions?
+7.2R_0$. Thus, acoustic waves may be launched in $R\gtrsim 7.2R_0$ by
+the spiral disturbance in the inner disc \citep{lin11b}. 
+%implications: effect on dynamics in outer disc
 
@@
 perturbation. However, to keep the Toomre profiles similar, we adjust
-the surface density power-law index such that $s = (3+q)/2$ for each
-case presented below. %should not be important, local effects
+the surface density power-law index such that $s = (3+q)/2$.
 

### 2014-11-12 paper/results.tex
minor updates

@@
 The spiral can be considered as low frequency because its 
-pattern speed $\Omega_p\lesssim 0.3\Omega$ in $R\in[R_{1},R_{2}]$,
-where it is  most prominent. The growth rate $\gamma$ is also slow
+pattern speed $\Omega_p \simeq 0.1\Omega_k(R_0)\lesssim 0.3\Omega$ in
+$R\in[R_{1},R_{2}]$ (where it has the largest amplitude). The growth rate $\gamma$ is also slow
 relative to the local rotation. (Although the characteristic growth
@@
 Fig. \ref{fargo_qbarrier} shows two $Q$-barriers located in
-the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$, which is indeed the
-region where the $m=1$ spiral develops. This suggests that the one-arm
+the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$; the bounded region is indeed the
+where the $m=1$ spiral develops. This suggests that the one-arm 
 spiral is trapped. Note in this region, $\nu^2 - 1 + 

### 2014-11-12 paper/summary.tex
minor updates

@@
 It would be of interest to explore other origins of the neutral
-one-arm spiral. One possibility is in discs around classic Be stars
-\citep{rivinius13}, for which several authors have demonstrated the
-existence of one-arm oscillations \citep[see e.g.][and references
-therein]{okasaki97,papaloizou06c,ogilvie08}. These authors employ
+one-arm spiral. One possibility is in Be star discs
+\citep{rivinius13}, for which 
+one-arm oscillations may explain long-timescale variations in their
+emission lines \citep[see e.g.][and references
+therein]{okasaki97,papaloizou06c,ogilvie08}. These studies employ
 strictly isothermal disc models, so it would be interesting to explore

### 2014-11-13 paper/discussion.tex
update

@@
 \end{align}
-Thus, low-frequency $m=1$ modes are generally associated with negative
+Thus, low-frequency $m=1$ modes generally have negative
 angular momentum. If the mode loses (positive) angular momentum
@@
   and the background disc due to an imposed temperature gradient}
-The $m=1$ spiral that emerges in our simulations have low frequency
-and grows slowly, and its spatial form is consistent local theory
-(\S\ref{fargo_m1}). To demonstrate possible instability, we
-first construct a neutral low-frequency mode using local theory, then
-show that the torque density acting on this mode due to the background
-temperature gradient is negative, which enforces the mode. % Note that
+The $m=1$ spiral that emerges in our simulations have low frequency,
+grows slowly, and its spatial form is consistent local theory  
+(\S\ref{fargo_m1}). Then to a first approximation, we can regard it 
+as a neutral, low-frequency $m=1$ mode in satifying the 
+dispersion relation Eq. \ref{dispersion}.  
+To demonstrate possible instability, %we 
+%first construct a neutral low-frequency mode using local theory, 
+we show that the torque density acting on such a mode due to the background
+temperature gradient is negative, which enforces the mode, because it
+has negative angular mometum.  
+% Note that
 % we will not be assuming a self-gravitating disc until the end of this
@@
 \end{align}
-For a neutral mode, $\sbar$, and therfore $D$ is real. Then, after
+For a neutral mode, $\sbar$, and therefore $D$ is real. Then, after
 taking the imaginary part of the above expression, the torque
@@
 where we used $c_s/R\sim h\Omega$. For our fiducial disc models with
-$q=1$, $Q=O(1)$, $h=0.05$, this gives $\gamma/\Omega = O(10^{-2})$,
+$q=1$, $Q=O(1)$, $h=0.05$, this gives $\gamma/\Omega < O(10^{-2})$,
 consistent with simulation results.
 
-Of course, we have been assuming a neutral mode to obtain expressions
+Of course, we assumed a neutral mode to obtain expressions
 for $\jlin$ and $T_\mathrm{BG}$ in the first place, so
 Eq. \ref{theoretical_rate} is only an order-of-magnitude
-constraint. Nevertheless, the xpected growth rate is in reasonable
+constraint. Nevertheless, the expected growth rate is in reasonable
 agreement with numerical results.
@@
 \subsection{The role of self-gravity and disc structure} 
+In the above discussion, we only considered a self-gravitating
+perturbation to obtain an appropriate wavenumber $k$ needed to
+evaluate Eq. \ref{theoretical_rate}. 
 The background torque density $T_\mathrm{BG}$ does not depend on  
-self-gravity explicitly. In the above discussion, we only considered a self-gravitating perturbation to 
-obtain an appropriate wavenumber $k$ needed to evaluate
-Eq. \ref{theoretical_rate}. Self-gravity is not directly
+self-gravity explicitly, so self-gravity is not directly
 responsible for instability. However, 
@@
 the imposed temperature gradient as measured in the simulations
-$\gamma\sim 0.1\Omega_p$). We conclude that for our disc models, the
+($\gamma\sim 0.1\Omega_p$). We conclude that for our disc models, the
 Lindblad and co-rotation resonances have negligible effect on the

### 2014-11-13 paper/results.tex
update

@@
 \subsection{Angular momentum exchange with the background disc}  
-The wavenumber $k = \pi G\Sigma/c_s^2$ corresponds to the
-wavelength most susceptible to gravitational instability for
-\emph{axisymmetric} perturbations. Physically, then, it is not
+The wavenumber $k = \pi G\Sigma/c_s^2$ minimises the shifted frequency
+$\sbar$ in the dispersion relation, Eq. \ref{dispersion}.  
+%\emph{axisymmetric} perturbations. 
+Physically, then, it is not
 surprising to find perturbations of this wavenumber in a
 self-gravitating disc where $Q\sim 1$. However, according the
-local dispersion relation, $m=1$ perturbations are formally stable for 
-real $k$.
+local dispersion relation, low-frequency $m=1$ perturbations are
+formally stable for real $k$.
 
@@
 where we recall $T_\mathrm{BG}$ is the torque density due to the imposed
-sound-speed profile (Eq. \ref{baroclinic_torque}). We have explicitly
-computed both sides of Eq. \ref{baroclinic_torque_int} using
+sound-speed profile (Eq. \ref{baroclinic_torque}). We explicitly 
+compute both sides of Eq. \ref{baroclinic_torque_int} using
 simulation data, and compare them in Fig. \ref{fargo_angmom_ex}. There
 is a good match between the two torques, especially at early times
-$t\lesssim110P_0$. The match is less good later on, when the spiral
+$t\lesssim110P_0$. The average discrepency is $\simeq 5\%$. 
+The match is less good later on, when the spiral
 amplitude is no longer small ($\Delta\Sigma_1\sim 0.2$ at $t=110P_0$
@@
 
+% similar results if only intergrating between R_1 and R_2
 %wave losses at boundaries
@@
 \subsection{Dependence on the imposed temperature profile} 
-At this point we hypothesize the growth of the $m=1$ spiral mode is
+We now hypothesize the growth of the $m=1$ spiral mode is
 associated with the imposed temperature gradient. The torque
@@
 simulations with $q\in[0,1]$ using only an $m=1$ initial 
-perturbation. However, to keep the Toomre profiles similar, we adjust
+perturbation. However, to keep the Toomre $Q$ profiles similar, we adjust
 the surface density power-law index such that $s = (3+q)/2$.
@@
 inspection of the $m=1$ surface density field. The growth in this case
-may be associated with higher $m$ modes.  
+may be associated with higher $m$ modes, which dominated the
+simulation.    
 % show result of isothermal calc?
@@
 \end{align*}
+% difficult to measure very small growth rates
 This confirms that the temperature gradient is responsible for the
-development of the one-arm spiral observed in our simulations. 
+development of the one-arm spirals observed in our simulations. 
 \begin{figure}
@@
 We also performed a series of simulations with variable aspect-ratio
-$h\in[0.03,0.07]$ but fixed $q=1$. This also affects the temperature
+$h\in[0.03,0.07]$ but fixed $q=1$. This affects the magnitude of the temperature
 gradient since $c_s \propto h$. However, with other parameters equal
@@
 \end{align}
-However, a linear fit is less good than for variable $q$ above. This
+However, a linear fit is less good than for variable $q$ cases above. This
 may be due to the change in the total disc mass when $h$ changes. We

### 2014-11-13 paper/results_3d.tex
update

@@
 N_\phi=256\times32\times256$. This corresponds to about $4$ cells per
-$H$in radius. Because of the much reduced resolution 
+$H$ in radius. Because of the much reduced resolution 
 compared to 2D, we use a smooth perturbation by setting
@@
                                 %resolution? 
-Results from ZEUS-MP appears to be off-set from that obtained using
-PLUTO.  
-%We suspect this is due to larger numerical noise in the
-%ZEUS-MP code, 
 In ZEUS-MP we observe disturbances developed near
 the inner boundary, which is probably responsible for the initial growth
-for $t<70P_0$. This growth is likely a numerical artifact and effectively seeds
-the ZEUS-UP simulation with a larger initial perturbation.  However,
+for $t<70P_0$. This is likely a numerical artifact and effectively seeds
+the ZEUS-UP simulation with a larger initial perturbation. Results
+from ZEUS-MP are therefore off-set from PLUTO. However,
 once the coherent $m=1$ spiral begins to grow ($t\gtrsim 100P_0$), 
@@
 \end{align*}
-Both are smaller than the 2D simulations. This may be due to the
+Both are smaller than the 2D simulations and is probably due to the
 low resolutions adopted and/or the small softening length employed in
@@
 simulations, which display no growth compared to cases with a 
-temperature gradient.
+temperature gradient. This confirms the temperature gradient effect is
+the same in 3D.
 
@@
 Fig. \ref{3d_prelim} for the disc midplane and near the upper disc
-boundary. The snapshots are chosen when the spirals in the two codes
+boundary. The snapshots are chosen when the one-armed spirals in the two codes
 have reached similar amplitudes. The agreement between the codes,
-as well as with the 2D simulations, is satisfactory.    
-Away from the disc midplane, a coherent spiral arm is
-not prominent, but the disturbance is radially more global, extending
-up to $R\sim 4R_0$. 
+as well as with the 2D simulations, is satisfactory. Away from the
+disc midplane, a coherent spiral arm is not prominent, but the
+disturbance is radially more global, extending up to $R\sim 4R_0$.   
 %implications?
@@
 spirals  increase in amplitude away from the midplane. It remains
-small  in our disc mode ($|\Delta\rho| < 0.1$) l, but could become 
-significant with a larger vertical domain. This shows that 
-the one-arm spiral can have a global influence.
+small in our disc model ($|\Delta\rho| < 0.1$), but could become 
+significant with a larger vertical domain. The figure shows that 
+the one-arm spiral can have a global influence away from the
+midplane of the self-gravitating annulus.
 
@@
 
-Although Fig. \ref{3d_rz} appears to show significant vertical motion,
+Although Fig. \ref{3d_rz} appears to display significant vertical motion,
 the three-dimensionality parameter $\Theta < 10^{-2}$
@@
 motion. We find  $|v_z/c_s|\lesssim 
-0.2$ for $h=0.05$. Thus, vertical motions are sub-sonic.  
+0.2$ for $h=0.05$. Thus, vertical motions are sub-sonic and negligible
+compared to horizontal motions, which supports a 2D approximation.
+
+%negligible vertical motion
+%unstratified approxi good?
+
 %This again supports the 2D approximation, altthou
@@
 
-\subsection{Angular momentum evolution} 
+\subsection{Angular momentum conservation} 
 Fig. \ref{3d_angmom} shows the angular momentum evolution in the 3D
-runs. These plots are again similar to the 2D runs, i.e. angular
-momentum lost by $J_1$ is gained by $J_0$. ZEUS-MP does not conserve
-angular momentum very well, but the variation $|\Delta J/J|<
-O(10^{-6})$ is small compared 
+runs. ZEUS-MP does not conserve angular momentum very well, but the
+variation $|\Delta J/J|< O(10^{-6})$ is small compared 
 to the individual components $|\Delta J_{0,1}/J|\sim 
 5\times10^{-5}$. The PLUTO run reaches similar values of
-$|J_{0,1}|$, but achieves better conervation, with $|\Delta
-J/J|=O(10^{-8})$.
+$|J_{0,1}|$, but achieves better conservation, with $|\Delta
+J/J|=O(10^{-8})$. These plots are again similar to the 2D simulations,
+i.e. angular momentum lost by $J_1$ is gained by $J_0$. This confirms 
+that the interaction between $J_1$ and $J_0$ operates in 3D and 2D
+similarly.
 

### 2014-11-15 paper/appendix.tex
simplified discussion

@@
 %calculated from velocity field, knowing the eigenfrequency)
+
+\section{Relation between horizontal Lagragian displacements for
+  local, low frequency $m=1$ disturbances}\label{horizontal_displacements}
+Here, we aim to relate the horizontal Lagrangian displacements $\xi_R$
+and $\xi_\phi$ in the local approximation. Using the local solution to
+the Poisson equation 
+\begin{align}
+  \delta \Phi_m = -\frac{2\pi G}{|k|} \delta\Sigma_m 
+\end{align}
+\citep{shu91}, the linearised azimuthal equation of motion becomes 
+\begin{align} 
+  - \ii\sbar \delta v_{\phi m}  + \frac{\kappa^2}{2\Omega}\delta v_{Rm} = -\frac{\ii
+    m}{R\Sigma}\left(c_s^2 - \frac{2\pi G
+      \Sigma}{|k|}\right)\delta\Sigma_m. 
+\end{align}
+Next, we replace the surface density perturbation
+$\delta \Sigma_m = -\ii k \Sigma \xi_R$, and use the expressions
+\begin{align}
+  &\delta v_{Rm} = -\ii\sbar\xi_R,\\
+  &\delta v_{\phi m} = -\ii\sbar\xi_\phi - \frac{\ii R
+    \p_R\Omega}{\sbar} \delta v_{Rm}
+\end{align}
+\citep{papaloizou85} to obtain 
+\begin{align}
+  -\sbar^2\xi_\phi - 2\ii\sbar\Omega \xi_R =
+  \frac{m}{kR}\left(\kappa^2 - \sbar^2\right)\xi_R, 
+\end{align}
+where the dispersion relation Eq. \ref{dispersion} was used. 
+In the local approximation, $|kR|\gg m$ by assumption.%  Furthermore,  
+% Also, for low frequency $m=1$ modes $\sbar \simeq -\Omega$, so the
+% quantity $|\kappa^2 - \sbar^2| \ll \Omega^2$ in a nearly Keplerian disc
+% ($\kappa\sim \Omega$). 
+ Hence the RHS of this equation can be neglected. Then 
+\begin{align}
+  \xi_\phi \simeq -\frac{2\ii\Omega}{\sbar}\xi_R.
+\end{align}
+For low-frequency $m=1$ modes we have $\sbar \simeq -\Omega$, so
+$\xi_\phi\simeq 2\ii\xi_R$, as used in the main text.

### 2014-11-15 paper/discussion.tex
simplified discussion

@@
   and the background disc due to an imposed temperature gradient}
-The $m=1$ spiral that emerges in our simulations have low frequency,
-grows slowly, and its spatial form is consistent local theory  
+The $m=1$ spiral that emerges in our simulations have low frequency
+($\Omega_p < \Omega$),
+grows slowly ($\gamma\ll\Omega_p$), and its spatial form is consistent local theory  
 (\S\ref{fargo_m1}). Then to a first approximation, we can regard it 
 as a neutral, low-frequency $m=1$ mode in satifying the 
-dispersion relation Eq. \ref{dispersion}.  
+dispersion relation Eq. \ref{dispersion} in local theory.  %attribute
+                                %growth to temp gradient 
 To demonstrate possible instability, %we 
-%first construct a neutral low-frequency mode using local theory, 
+% first construct a neutral low-frequency mode using local theory, 
 we show that the torque density acting on such a mode due to the background
@@
 locally isothermal equation of state,
+\begin{align*}
+  T_\mathrm{BG} = - \frac{m}{2}\frac{dc_s^2}{dR}\imag\left(\delta\Sigma_m\xi_R^*\right).   
+\end{align*}
+The Eulerian surface density perturbation is given by
+\begin{align}\label{den_pert}
+  \delta\Sigma_m = -\nabla\cdot\left(\Sigma\bm{\xi}\right) 
+  = -\frac{1}{R}\frac{d}{dR}\left(R\Sigma \xi_R\right) - \frac{\ii m}{R}\Sigma\xi_\phi.  
+\end{align}
+We now invoke local theory by setting $d/dR \to \ii k$ where $k$ is
+real, and assume $|kR|\gg m$ so that the second term on the right hand
+side of Eq. \ref{den_pert} is negligible. Then    
 \begin{align}
-  T_\mathrm{BG} &= -\frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\delta\Sigma_m \left(\frac{\delta
-        v_{Rm}}{-\ii\sbar}\right)^*\right] \notag\\
-  & = \frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\ii \delta\Sigma_m \left(\frac{\delta
-        v_{Rm}}{\sbar}\right)^*\right]
+  \delta\Sigma _m  \simeq -\ii k \Sigma \xi_R,
 \end{align}
-where we replaced the radial Lagrangian displacement by the Eulerian
-radial velocity perturbation in Eq. \ref{baroclinic_torque}. 
-The linearised momentum equations give
+which gives
 \begin{align}
-  D\delta v_{Rm} =& 
-  \ii\sbar\left[c_s^2\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right)
-    + \frac{d}{dR}\delta\Phi_m\right] \notag\\ 
-  &- \frac{2\ii
-    m\Omega}{R}\left(c_s^2\frac{\delta\Sigma_m}{\Sigma} +
-    \delta\Phi_m\right),
+  T_\mathrm{BG} = \frac{m}{2}\frac{dc_s^2}{dR}k\Sigma |\xi_R|^2. 
 \end{align}
-where $D\equiv \kappa^2 - \sbar^2$.
-
-We now invoke local theory by setting $d/dR \to \ii k$ where $k$ is
-real, and using the local solution to the Poisson equation
-\begin{align}
-  \delta\Phi_m = -\frac{2\pi G}{|k|}\delta\Sigma_m 
-\end{align}
-\citep{shu91}. Then the radial velocity perturbation becomes
-\begin{align}
-  \delta v_{Rm} =
-  \frac{1}{D}\frac{\delta\Sigma_m}{\Sigma}\left(\frac{2\pi G
-      \Sigma}{|k|} - c_s^2\right)\left(k\sbar + \frac{2\ii
-      m\Omega}{R}\right),
-\end{align}
-so that 
-\begin{align}
-  \ii\delta\Sigma_m \left(\frac{\delta v_{Rm}}{\sbar}\right)^* =
-  \frac{1}{D^*}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
-      \Sigma}{|k|} - c_s^2\right)\left(\ii k  + \frac{2
-      m\Omega}{R\sbar^*}\right). 
-\end{align}
-For a neutral mode, $\sbar$, and therefore $D$ is real. Then, after
-taking the imaginary part of the above expression, the torque
-density is 
-\begin{align}\label{tbg_explicit}
-  T_\mathrm{BG} &=
-  \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{D}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
-      \Sigma}{|k|} - c_s^2\right)\right].
-% &\simeq \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{\left(\kappa^2 -
-%         m^2\Omega^2\right)}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi
-%       G \Sigma}{|k|} - c_s^2\right)\right]. 
-\end{align}
-%where the second line follows from the low-frequency approximation
-%$|\sigma|\ll m\Omega$. 
-% To determine the sign of $T_\mathrm{BG}$ for low frequency one-arm spirals,
-% we examine each factor separately below for $m=1$:
-% \begin{itemize}
-% \item We observe trailing spirals in our simulations, so $k>0$. 
-% \item The factor $\kappa^2 - \Omega^2>0$ for our self-gravitating disc
-%   models in the region where the spiral appears. 
-% \item We measure $|k|\sim \pi G\Sigma/c_s^2$, so the last factor in
-%   the square brackets is positive. Note that this factor is positive
-%   provided $|k| < 2/HQ$ or wavelengths $\lambda > \pi Q H$.  
-% \end{itemize}
-We now use the local dispersion relation,
-Eq. \ref{dispersion}, to make the replacement $D = 2\pi G \Sigma |k| -
-k^2c_s^2$ in the above expression. This gives
-\begin{align}\label{tbg_simple}
-  T_\mathrm{BG} =
-  \frac{m}{2}\frac{dc_s^2}{dR}\frac{1}{k}\frac{|\delta\Sigma_m|^2}{\Sigma}.  
-\end{align} 
+% \begin{align}
+%   T_\mathrm{BG} &= -\frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\delta\Sigma_m \left(\frac{\delta
+%         v_{Rm}}{-\ii\sbar}\right)^*\right] \notag\\
+%   & = \frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\ii \delta\Sigma_m \left(\frac{\delta
+%         v_{Rm}}{\sbar}\right)^*\right]
+% \end{align}
+% where we replaced the radial Lagrangian displacement by the Eulerian
+% radial velocity perturbation in Eq. \ref{baroclinic_torque}. 
+% The linearised momentum equations give
+% \begin{align}
+%   D\delta v_{Rm} =& 
+%   \ii\sbar\left[c_s^2\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right)
+%     + \frac{d}{dR}\delta\Phi_m\right] \notag\\ 
+%   &- \frac{2\ii
+%     m\Omega}{R}\left(c_s^2\frac{\delta\Sigma_m}{\Sigma} +
+%     \delta\Phi_m\right),
+% \end{align}
+% where $D\equiv \kappa^2 - \sbar^2$. 
+% We now invoke local theory by setting $d/dR \to \ii k$ where $k$ is
+% real, and using the local solution to the Poisson equation
+% \begin{align}
+%   \delta\Phi_m = -\frac{2\pi G}{|k|}\delta\Sigma_m 
+% \end{align}
+% \citep{shu91}. Then the radial velocity perturbation becomes
+% \begin{align}
+%   \delta v_{Rm} =
+%   \frac{1}{D}\frac{\delta\Sigma_m}{\Sigma}\left(\frac{2\pi G
+%       \Sigma}{|k|} - c_s^2\right)\left(k\sbar + \frac{2\ii
+%       m\Omega}{R}\right),
+% \end{align}
+% so that 
+% \begin{align}
+%   \ii\delta\Sigma_m \left(\frac{\delta v_{Rm}}{\sbar}\right)^* =
+%   \frac{1}{D^*}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
+%       \Sigma}{|k|} - c_s^2\right)\left(\ii k  + \frac{2
+%       m\Omega}{R\sbar^*}\right). 
+% \end{align}
+% For a neutral mode, $\sbar$, and therefore $D$ is real. Then, after
+% taking the imaginary part of the above expression, the torque
+% density is 
+% \begin{align}\label{tbg_explicit}
+%   T_\mathrm{BG} &=
+%   \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{D}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
+%       \Sigma}{|k|} - c_s^2\right)\right].
+% % &\simeq \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{\left(\kappa^2 -
+% %         m^2\Omega^2\right)}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi
+% %       G \Sigma}{|k|} - c_s^2\right)\right]. 
+% \end{align}
+% We now use the local dispersion relation,
+% Eq. \ref{dispersion}, to make the replacement $D = 2\pi G \Sigma |k| -
+% k^2c_s^2$ in the above expression. This gives
+% \begin{align}\label{tbg_simple}
+%   T_\mathrm{BG} =
+%   \frac{m}{2}\frac{dc_s^2}{dR}\frac{1}{k}\frac{|\delta\Sigma_m|^2}{\Sigma}.  
+% \end{align}  
 This torque density is negative for trailing waves in discs with
-temperature decreasing outwards ($k>0$ and $dc_s^2/dR<0$,
+a fixed temperature profile decreasing outwards ($k>0$ and $dc_s^2/dR<0$,
 respectively). Note that this conclusion does not rely on the
@@
 However, if the linear disturbance under consideration \emph{is} a
-low-frequency $m=1$ 
-trailing spiral wave, then it has negative angular
-momentum. If $dc_s^2/dR<0$, as is typical for circumstellar or
-protoplanetary discs, then $T_\mathrm{BG}<0$, and the background disc
-applies a negative torque on the disturbance, which further decreases
-its angular momentum. This suggests the spiral amplitude will grow.
+low-frequency $m=1$ mode, then it has negative angular
+momentum. If it is a trailing spiral and $dc_s^2/dR<0$, as is typical
+for circumstellar or protoplanetary discs, then $T_\mathrm{BG}<0$, and
+the background disc applies a negative torque on the disturbance,
+which further decreases its angular momentum. This suggests the spiral
+amplitude will grow.
 
 Using $\jlin$ and $T_\mathrm{BG}$, we can obtain a rough estimate of
-the growth rate $\gamma$ of linear perturbations as  
+the growth rate $\gamma$ of linear perturbations due to the background
+torque as  
 \begin{align}
@@
   2\gamma \sim
-  -\frac{dc_s^2}{dR}\frac{1}{k\Sigma^2\Omega}\frac{|\delta\Sigma_1|^2}{|\xi_R
-    + \ii\xi_\phi|^2}.
+  -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R|^2}{|\xi_R + \ii
+    \xi_\phi|^2}. 
+%  \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
 \end{align}
-Now, using $\delta\Sigma_m/\Sigma = -\ii k \xi_R - \ii m \xi_\phi/R$
-in the local approximation, we have
+In Appendix \ref{horizontal_displacements}, we argue $\xi_\phi \simeq
+2\ii \xi_R$ for the modes of interest. Then 
+%To further simplify this expression, we need to relate $\xi_R$ and
+%$\xi_\phi$. 
+% \begin{align}
+%   2\gamma \sim
+%   -\frac{dc_s^2}{dR}\frac{1}{k\Sigma^2\Omega}\frac{|\delta\Sigma_1|^2}{|\xi_R
+%     + \ii\xi_\phi|^2}.
+% \end{align}
+% Now, using $\delta\Sigma_m/\Sigma = -\ii k \xi_R - \ii m \xi_\phi/R$
+% in the local approximation, we have
+% \begin{align}
+%   2\gamma \sim -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R +
+%     \xi_\phi/kR|^2}{|\xi_R + \ii \xi_\phi|^2}
+%   \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
+% \end{align}
+% where we used $dc_s^2/dR <0$, $k>0$ and $|kR|\gg 1$ to obtain the
+% inequality. 
+%where we used $dc_s^2/dR <0$ and $k>0$ to obtain the inequality. 
+for the temperature profiles $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ as
+adopted in our disc models,  
 \begin{align}
-  2\gamma \sim -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R +
-    \xi_\phi/kR|^2}{|\xi_R + \ii \xi_\phi|^2}
-  \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
+  2\gamma \sim q\frac{c_s^2}{R}\frac{k}{\Omega}. 
 \end{align}
-where we used $dc_s^2/dR <0$, $k>0$ and $|kR|\gg 1$ to obtain the
-inequality. For $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ adopted in our disc
-models,
+For perturbations in a self-gravitating disc with characteristic
+wavenumber $k \sim \pi G \Sigma/c_s^2 \sim \Omega/Q c_s$, as seen in
+our simulations, we have 
 \begin{align}\label{theoretical_rate}
-  2\gamma \lesssim q\frac{c_s^2}{R}\frac{k}{\Omega}. 
-\end{align}
-Now we consider perturbations in a self-gravitating disc with characteristic
-wavenumber $k \sim \pi G \Sigma/c_s^2 \sim \Omega/Q c_s$ (as seen in
-our simulations). Then
-\begin{align}
-  \gamma \lesssim \frac{qh}{2Q}\Omega,
+  \gamma \sim \frac{qh}{2Q}\Omega,
 \end{align}
 where we used $c_s/R\sim h\Omega$. For our fiducial disc models with
-$q=1$, $Q=O(1)$, $h=0.05$, this gives $\gamma/\Omega < O(10^{-2})$,
+$q=1$, $Q=O(1)$, $h=0.05$, this gives $\gamma/\Omega \sim O(10^{-2})$,
 consistent with simulation results.
 
-Of course, we assumed a neutral mode to obtain expressions
-for $\jlin$ and $T_\mathrm{BG}$ in the first place, so
-Eq. \ref{theoretical_rate} is only an order-of-magnitude
-constraint. Nevertheless, the expected growth rate is in reasonable
-agreement with numerical results.
+%Of course, we assumed a neutral mode to obtain expressions
+%for $\jlin$ and $T_\mathrm{BG}$ in the first place, 
+Note that by using local theory we have implicitly
+assumed a neutral mode because the dispersion relation,
+Eq. \ref{dispersion}, states that non-axisymmetric modes are
+stable for real $k$. Eq. \ref{theoretical_rate} should be interpreted
+as an order-of-magnitude estimate of the perturbative effect of a
+temperature gradient on local, low-frequency $m=1$ modes.
+
+%neutral modes are not actually neutral)
+%gamma << omega_p << omega 
+%so
+%Eq. \ref{theoretical_rate} is only an order-of-magnitude
+%constraint. Nevertheless, the expected growth rate is in reasonable
+%agreement with numerical results.
 

### 2014-11-15 paper/summary.tex
update
simplified discussion

@@
 %non-sg sims, Be stars, eccentric discs
-We stress that destabilisation through the temperature gradient
-does not explicitly depend on self-gravity. In principle, this effect
-will destabilise low-frequency $m=1$ trailing spirals regardless of
-its origin. In our models, self-gravity coupled with the disc
+% We stress that destabilisation through the temperature gradient
+% does not explicitly depend on self-gravity. In principle, this effect
+% will destabilise low-frequency $m=1$ trailing spirals regardless of
+% its origin.
+In our models and interpretation, self-gravity coupled with the disc 
 structure permits a neutral one-arm mode, which is then destabilised.   
-It would be of interest to explore other origins of the neutral
+It would be interesting to explore other origins of the neutral
 one-arm spiral. One possibility is in Be star discs
@@
 
+%vertical shear instability?
+
 %linear theory: explicit calculations 

### 2014-11-15 paper/appendix.tex
minor updates

@@
 \section{Relation between horizontal Lagragian displacements for
-  local, low frequency $m=1$ disturbances}\label{horizontal_displacements}
+  local, low frequency \lowercase{$m=1$} disturbances}\label{horizontal_displacements}
 Here, we aim to relate the horizontal Lagrangian displacements $\xi_R$

### 2014-11-15 paper/discussion.tex
minor updates

@@
       m\Omega\right)|\bm{\xi}|^2 + \ii\Omega\left(\xi_R\xi_\phi^* -
-      \xi_R\xi_\phi^*\right)\right].  
+      \xi_R^*\xi_\phi\right)\right].  
 \end{align}
@@
 \subsection{The confined spiral as an external potential}
-One possible effect of self-gravity is to allow the one-arm spiral  
+One possible effect of self-gravity is to allow the one-armed spiral  
 to act as an external potential for the wider disc. This is
@@
 
-Let us treat the one-arm spiral confined in $R\in[R_1,R_2]$ as an  
+Let us treat the one-armed spiral confined in $R\in[R_1,R_2]$ as an  
 external potential of the form $\Phi_\mathrm{ext}(R)\cos{\left(\phi -
@@
 %simulation ($\gamma\sim0.1\Omega_p$). We conclude that for our models,
-%the OLR has negligible effect on the one-arm spiral.  
+%the OLR has negligible effect on the one-armed spiral.  
 For the co-rotation torque, we use the result
@@
 additional FARGO simulations with a reduced radial domain size,
-setting $R_\mathrm{max}=3R_0$, which still developed the one-arm
+setting $R_\mathrm{max}=3R_0$, which still developed the one-armed
 spiral.

### 2014-11-15 paper/model.tex
simplified discussion
minor updates

@@
 resolution, but have also carried out some 3D disc
-simulations. Thus, for generality we describe the system in
+simulations. Hence, for generality we describe the system in
 3D, using both cylindrical $(R,\phi,z)$ and spherical polar
@@
 effect is not expected to be significant
-\citep{adams89,shu90}. Indeed, simulations including the indirect
+\citep{adams89,shu90,michael10}. Indeed, simulations including the indirect
 potential, carried out in the early stages of this project, produced
@@
 
-In 3D, we assume vertical hydrostatic balance
+%In 3D, we assume vertical hydrostatic balance
+The 3D disc structure is obtained by assuming vertical hydrostatic
+balance  
 \begin{align}
@@
 coordinates. We assume reflectional symmetry across the midplane, and
-apply relfective boundary conditions at radial boundaries and the
+apply reflective boundary conditions at radial boundaries and the
 upper disc boundary.
@@
 is the angular momentum density, $\bm{\xi}$ is the Lagragian
-displacement, and $\bm{F}$ is the angular momentum flux
-consisting of a Reynolds stress and a gravitational stress. Explicit
-expressions for $\bm{\xi}$ can be found in, e.g. \cite{papaloizou85},
-and $\bm{F}$ can be read off Eq. \ref{angflux1}.  
+displacement, and $\bm{F}$ is the vertically-integrated angular
+momentum flux consisting of a Reynolds stress and a gravitational
+stress \citep{lin11b}. Explicit expressions for $\bm{\xi}$ can be found in,
+e.g. \cite{papaloizou85}. 
+%and $\bm{F}$ can be read off Eq. \ref{angflux1}.
+
 % \cite{papaloizou85} and \cite{lin11} respectively, and
@@
   \tbg \equiv
-  -\frac{m}{2}\imag\left(\Sigma_m\xi_R^*\frac{dc_s^2}{dR}\right), 
+  -\frac{m}{2}\imag\left(\delta\Sigma_m\xi_R^*\frac{dc_s^2}{dR}\right), 
 \end{align}
@@
 state. In a barotropic fluid, such as a strictly isothermal disc,
-$\tbg$ vanishes and the angular momentum associated with
+$\tbg$ vanishes and the total angular momentum associated with
 the perturbation is conserved, provided that there is no net angular
-momeumtum flux and the perturbations are regular.
+momeumtum flux.% and the perturbations are regular.
 

### 2014-11-15 paper/paper.tex
minor updates

@@
 
-\title[One-arm spirals]{One-arm spirals in 
+\title[One-armed spirals]{One-armed spirals in 
   locally isothermal, radially structured self-gravitating discs}
@@
   outside the bulk of the spiral arm, resulting in a nearly stationary 
-  one-arm spiral pattern in the self-gravitating region of the
-  disc. The development of this one-arm 
+  one-armed spiral pattern in the self-gravitating region of the
+  disc. The development of this one-armed 
   spiral does not require the movement of  
@@
   not explicitly depend on self-gravity, we suggest it may also
-  affect low-frequency one-arm spirals in non-self-gravitating
+  affect low-frequency one-armed spirals in non-self-gravitating
   discs. 
@@
 massive dead zone.  %This was indeed the original goal of this project. 
-This work presents a first step towards this goal. Specifically, we 
+This work presents a first step towards this goal. Specifically, here we 
 study a class of seemingly simple disc models which were found to 
@@
 instability is attributed to motion of the central star induced by the
-one-arm perturbation and requires a massive disc \citep[the former may
+one-armed perturbation and requires a massive disc \citep[the former may
 have observable consequences, ][]{michael10}. However, our numerical
@@
 The aforementioned analytical studies have adopted barotropic, smooth disc models. 
-In this work, we identify a mechanism to generate one-arm
+In this work, we identify a mechanism to generate one-armed
 spiral patterns that is related to a non-barotropic equation of
@@
 sound-speed is a prescribed function of position, then 
-one-arm spiral modes can grow inside a cavity that results from the
+one-armed spiral modes can grow inside a cavity that results from the
 adopted disc structure.
@@
 different grid-based codes, both in two-dimensions (2D) and
-three-dimensions (3D). We also apply results from linear
-theory to explain their growth. We find that, when the disc
+three-dimensions (3D). We also apply linear
+theory to interpret simulation results. We find that, when the disc
 temperature is prescribed, the usual statement for the conservation of

### 2014-11-15 paper/results.tex
minor updates

@@
 O(10^{-5})$. Fig. \ref{2d_angmom} then suggest that angular momentum
-is transferred from the one-arm spiral to the background disc.
+is transferred from the one-armed spiral to the background disc.
 
@@
 Fig. \ref{2d_fargo_viz} shows the $m=1$ surface density
-at the end of this simulation. For this one-armed spiral
+at the end of this simulation. For this one-armeded spiral
 we measure a co-rotation radius and growth rate  
@@
 pattern speed $\Omega_p \simeq 0.1\Omega_k(R_0)\lesssim 0.3\Omega$ in
-$R\in[R_{1},R_{2}]$ (where it has the largest amplitude). The growth rate $\gamma$ is also slow
-relative to the local rotation. (Although the characteristic growth
-time $\gamma^{-1} \simeq 10P_0$.) Thus, the spiral 
-pattern is nearly stationary.
-
+$R\in[R_{1},R_{2}]$ (where it has the largest amplitude). Thus, the spiral 
+pattern appears nearly stationary. The growth rate $\gamma$ is also slow
+relative to the local rotation, although the characteristic growth
+time $\gamma^{-1} \simeq 10P_0$ is not very long.  
 \begin{figure}
@@
 the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$; the bounded region is indeed the
-where the $m=1$ spiral develops. This suggests that the one-arm 
+where the $m=1$ spiral develops. This suggests that the one-armed 
 spiral is trapped. Note in this region, $\nu^2 - 1 + 
-Q^{-2}\simeq 0.1\ll 1$, as required for consistency with
-Eq. \ref{dispersion} with the lower sign taken (i.e. long waves).  
+Q^{-2}\simeq 0.1\ll 1$, which is necessary for consistency with 
+the measured wavenumber $k$ and Eq. \ref{wavenumber}.  
+%Eq. \ref{dispersion} with the lower sign taken (i.e. long waves).  
 There is one outer Lindblad resonance at $R_L\simeq
@@
 self-gravitating disc where $Q\sim 1$. However, according the
-local dispersion relation, low-frequency $m=1$ perturbations are
+local dispersion relation, $m=1$ perturbations are
 formally stable for real $k$.
@@
 $q$. Although the figure indicates growth for the strictly isothermal
-disc ($q=0$), we did not observe a coherent one-arm spiral upon
+disc ($q=0$), we did not observe a coherent one-armed spiral upon
 inspection of the $m=1$ surface density field. The growth in this case
@@
 This confirms that the temperature gradient is responsible for the
-development of the one-arm spirals observed in our simulations. 
+development of the one-armed spirals observed in our simulations. 
 \begin{figure}

### 2014-11-15 paper/results_3d.tex
minor updates

@@
 Fig. \ref{3d_prelim} for the disc midplane and near the upper disc
-boundary. The snapshots are chosen when the one-armed spirals in the two codes
+boundary. The snapshots are chosen when the one-armeded spirals in the two codes
 have reached similar amplitudes. The agreement between the codes,
@@
 \subsection{Vertical structure}
-Fig. \ref{3d_rz} shows the vertical structure of the one-arm
+Fig. \ref{3d_rz} shows the vertical structure of the one-armed
 spiral in the PLUTO run. The main spiral is confined to $z < H$. Thus,
@@
 significant with a larger vertical domain. The figure shows that 
-the one-arm spiral can have a global influence away from the
+the one-armed spiral can have a global influence away from the
 midplane of the self-gravitating annulus.
@@
 %Vertical velocities
-%do increase with the growth of the one-arm spiral, 
+%do increase with the growth of the one-armed spiral, 
 %Larger amplitudes
@@
     simulations. The perturbation 
-    relative to $t=10P_0$, during the growth of the one-arm spiral,
+    relative to $t=10P_0$, during the growth of the one-armed spiral,
     is shown in units of the initial total angular momentum

### 2014-11-15 paper/summary.tex
minor updates

@@
 We find such systems can be unstable to low-frequency perturbations
-with azimuthal wavenumber $m=1$, which lead to the development of one-arm 
+with azimuthal wavenumber $m=1$, which lead to the development of one-armed 
 trailing spirals that persist for at least $O(10^2)$ orbits. The 
@@
 PLUTO in 3D --- to verify the growth of  
-one-arm spirals in our disc models is due to the imposed temperature
+one-armed spirals in our disc models is due to the imposed temperature
 profile and disc structure. The spiral is confined between two
 $Q$-barriers owing to a surface density bump. Growth rates increased
-with the magnitude of the imposed temperature gradient, and one-armed
+with the magnitude of the imposed temperature gradient, and one-armeded
 spirals did not develop in strictly isothermal simulations.  
@@
 low-frequency $m=1$ modes because they are associated with negative
-angular momentum. We therefore interpret the one-arm spiral instability
+angular momentum. We therefore interpret the one-armed spiral instability
 observed in our simulations as a neutral, tightly-wound $m=1$ mode
@@
 %balancing heating and cooling 
-In the deeply non-linear regime, the one-arm spiral may 
+In the deeply non-linear regime, the one-armed spiral may 
 shock and dissipate, depositing negative angular momentum onto
@@
 indicate a relaxation timescale $t_c < 0.1\Omega_k^{-1}$ is needed for
-the one-arm spiral to develop. This will be presented in a follow-up
+the one-armed spiral to develop. This will be presented in a follow-up
 study. %prelim sims show they are long lasting
@@
 In our models and interpretation, self-gravity coupled with the disc 
-structure permits a neutral one-arm mode, which is then destabilised.   
-It would be interesting to explore other origins of the neutral
-one-arm spiral. One possibility is in Be star discs
+structure permits a neutral one-armed mode, which is then
+destabilised. Other origins of a neutral
+one-armed spiral should be investigated. One possibility is in Be star discs
 \citep{rivinius13}, for which 
-one-arm oscillations may explain long-timescale variations in their
+one-armed oscillations may explain long-timescale variations in their
 emission lines \citep[see e.g.][and references

### 2014-11-16 paper/results_3d.tex
updating 3d results

@@
     \subfigure[ZEUS-MP]{
-      \includegraphics[scale=0.382]{figures/polarxy_dens015_z0}
-      \includegraphics[scale=0.382,clip=true,trim=2.24cm 0cm 0cm
-      0cm]{figures/polarxy_dens015_zmax}  
+      \includegraphics[scale=0.305,clip=true,trim=0cm 0cm 0cm 0cm]{figures/polarxy2_dens015_z0}  
+      \includegraphics[scale=0.305,clip=true,trim=0.8cm 0cm 0cm
+      0cm]{figures/polarxy2_dens015_zmax}  
     }
     \subfigure[PLUTO]{
-      \includegraphics[scale=0.382]{figures/pdisk_020_z0}
-      \includegraphics[scale=0.382,clip=true,trim=2.24cm 0cm 0cm
-      0cm]{figures/pdisk_020_zmax}
+      \includegraphics[scale=0.305,clip=true,trim=0cm 0cm 0cm 0cm]{figures/pdiskxy_023_z0}
+      \includegraphics[scale=0.305,clip=true,trim=0.8cm 0cm 0cm
+      0cm]{figures/pdiskxy_023_zmax}
     }
@@
   \caption{Three-dimensional simulations using the ZEUS-MP (top) and 
-    PLUTO (bottom) codes. The non-axisymmetric density at the midplane (left) and
+    PLUTO (bottom) codes. The $m=1$ density component $\Delta\rho_1$ at the midplane (left) and
     two scale-heights above the midplane (right) is shown. Here $\psi
     \equiv \pi/2 - \theta$ is the angular height  
-    from the midplane.\label{3d_prelim}}   
+    from the midplane. \label{3d_prelim}}   
 \end{figure}
@@
 \begin{figure}
-  \includegraphics[scale=0.47,clip=true,trim=0cm 0.79cm 0cm 0cm]{figures/pdisk_rz_020_sg.ps}
-  \includegraphics[scale=0.47,clip=true,trim=0cm 0cm 0cm 0.64cm]{figures/pdisk_rz_020_nsg.ps}
-  \caption{Non-axisymmetric density in the meridional plane from the
+  \includegraphics[scale=0.47,clip=true,trim=0cm 0.79cm 0cm
+  0cm]{figures/pdisk_rz_023_sg.ps}  
+  \includegraphics[scale=0.47,clip=true,trim=0cm 0cm 0cm
+  0.64cm]{figures/pdisk_rz_023_nsg.ps}  
+  \caption{Vertical distribution of the $m=1$ density component in the meridional plane from the 
     PLUTO simulation. The slices are taken at the azimuth of   
-    $\mathrm{max}[\Delta\rho(r,\pi/2,\phi)]$. Arrows correspond to the vector 
+    $\mathrm{max}[\Delta\rho_1(r,\pi/2,\phi)]$. Arrows correspond to the vector 
     $(v_r/R_0,-v_\theta/rh\sin^2{\theta})$. The top (bottom) panel corresponds

### 2014-11-16 paper/results_3d.tex
updating 3d stuff

@@
 $n_H=2$ scale-heights at $R=R_0$. The resolution is $N_r\times N_\theta\times
-N_\phi=256\times32\times256$. This corresponds to about $4$ cells per
-$H$ in radius. Because of the much reduced resolution 
+N_\phi=256\times32\times512$. This corresponds to about $4$ cells per
+$H$ in radius. Because of the reduced resolution 
 compared to 2D, we use a smooth perturbation by setting
 $\delta = 10^{-3}$ and $M=1$ in Eq. \ref{randpert}. This corresponds
-to a single $m=1$ disturbance in $R\in[R_1,R_2]$, which was found to dominate
-the 2D simulation.
+to a single $m=1$ disturbance in $R\in[R_1,R_2]$, which
+eventually dominates in 2D simulations.
 
@@
 meridional velocities are damped out. We then restart the simulation
-with the above perturbation and $(\lmax,\mmax)=(32,32)$ and damp
+with the above perturbation and $(\lmax,\mmax)=(32,32)$, and damp
 merdional velocities near the radial boundaries.
@@
                                 %resolution? 
-In ZEUS-MP we observe disturbances developed near
+In the ZEUS-MP run we observe disturbances developed near
 the inner boundary, which is probably responsible for the initial growth
@@
 the ZEUS-UP simulation with a larger initial perturbation. Results
-from ZEUS-MP are therefore off-set from PLUTO. However,
+from ZEUS-MP are therefore off-set from PLUTO by $\sim50P_0$ However,
 once the coherent $m=1$ spiral begins to grow ($t\gtrsim 100P_0$), 
@@
 \end{align*}
-Both are smaller than the 2D simulations and is probably due to the
+Both are smaller than the 2D simulations and due to the
 low resolutions adopted and/or the small softening length employed in
@@
   \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ampmax3d.ps}
-  \caption{Evolution of the maximum $m=1$ density in  $r\in[R_1,R_2]$
+  \caption{Evolution of the maximum $m=1$ density component in  $r\in[R_1,R_2]$
     measured in the 3D simulations. Results from discs with a imposed 

### 2014-11-17 paper/results_3d.tex
minor updates to tex

@@
 $n_H=2$ scale-heights at $R=R_0$. The resolution is $N_r\times N_\theta\times
-N_\phi=256\times32\times512$. This corresponds to about $4$ cells per
-$H$ in radius. Because of the reduced resolution 
+N_\phi=256\times32\times512$, or about $4$ cells per
+$H$. Because of the reduced resolution 
 compared to 2D, we use a smooth perturbation by setting
@@
 with the above perturbation and $(\lmax,\mmax)=(32,32)$, and damp
-merdional velocities near the radial boundaries.
+meridional velocities near the radial boundaries.
 
-Fig. \ref{3d_ampmax} plots the amplitude of the $m=1$ spirals measured
-in the ZEUS-MP and PLUTO runs.  
+Fig. \ref{3d_ampmax} plots the evolution of the $m=1$ spiral amplitudes measured
+in the ZEUS-MP and PLUTO runs. We also ran simulations
+with a strictly isothermal equation of state ($q=0$), which display no
+growth compared to that with a temperature gradient.  This confirms
+the temperature gradient effect is the same in 3D. 
 %why no high m
@@
 the ZEUS-UP simulation with a larger initial perturbation. Results
-from ZEUS-MP are therefore off-set from PLUTO by $\sim50P_0$ However,
+from ZEUS-MP are therefore off-set from PLUTO by $\sim50P_0$. However,
 once the coherent $m=1$ spiral begins to grow ($t\gtrsim 100P_0$), 
@@
 \end{align*}
-Both are smaller than the 2D simulations and due to the
-low resolutions adopted and/or the small softening length employed in
-2D.  Fig. \ref{3d_ampmax} also show results from strictly isothermal
-simulations, which display no growth compared to cases with a 
-temperature gradient. This confirms the temperature gradient effect is
-the same in 3D.
+Both are smaller than the 2D simulations, because of to the
+low resolutions adopted in 3D and/or the small softening length employed in
+2D. 
+%Fig. \ref{3d_ampmax} also show results from strictly isothermal
+%simulations, which display no growth compared to cases with a 
+%temperature gradient.
 
@@
 Fig. \ref{3d_prelim} for the disc midplane and near the upper disc
-boundary. The snapshots are chosen when the one-armeded spirals in the two codes
-have reached similar amplitudes. The agreement between the codes,
-as well as with the 2D simulations, is satisfactory. Away from the
-disc midplane, a coherent spiral arm is not prominent, but the
-disturbance is radially more global, extending up to $R\sim 4R_0$.   
+boundary. The snapshots are chosen when the one-armed spirals in the two codes
+have reached comparable amplitudes. The agreement between the codes,
+as well as with the 2D simulations, is satisfactory. The largest
+spiral amplitude is found in $R\in[R_1,R_2]$ independent of
+height. However, notice the spiral pattern extends into the
+non-self-gravitating outer disc ($R>R_2$) at $z\sim 2H$, i.e. 
+the disturbance becomes more global away from the midplane.
+
+%Away from the
+%disc midplane, a coherent spiral arm is not prominent, but the
+%disturbance is radially more global, extending up to $R\sim 4R_0$.
+
 %implications?

### 2014-11-19 paper/model.tex
minor updates

@@
 We consider a self-gravitating, inviscid fluid disc of mass $M_d$ 
-orbting a central star of mass $M_*$. We will mainly examine 2D
+orbiting a central star of mass $M_*$. We will mainly examine 2D
 (or razor-thin) discs in favour of numerical 

### 2014-11-19 paper/paper.tex
minor updates

@@
   the central star, as found in previous studies. Instead, we find it
-  is associated with the imposed 
+  to be associated with the imposed 
   temperature gradient and the radial disc structure. 
   We apply results from linear theory and attribute the
-  growth of this spiral to the non-conservation of angular momentum
+  growth of this spiral mode to the non-conservation of angular momentum
   for non-axisymmetric linear perturbations in discs with a forced temperature
-  gradient. This may lead to an unstable exchange of angular momentum between the spiral 
+  gradient. This may lead to an unstable exchange of angular momentum
+  between the spiral mode 
   and the background disc. However, because the interaction does
@@
 spiral arms can provide significant angular momentum transport
-neccessary for accretion \citep{lynden-bell72, papaloizou91,balbus99,lodato04}, and
+neccessary for mass accretion \citep{lynden-bell72, papaloizou91,balbus99,lodato04}, and
 spiral structures due to GI are potentially observable with the Atacama Large 
@@
 instability is ineffective for angular momentum transport
-\citep{gammie96,turner08,landry13} and therefore has reduced mass
-accretion. Dead zones may then become
+\citep{gammie96,turner08,landry13} and therefore has a reduced mass
+accretion rate. Dead zones may then become
 self-gravitating with sufficient mass built-up

### 2014-11-20 paper/appendix.tex
minor updates
spell checked

--- 2014-11-15 paper/appendix.tex
@@
 We begin with the linearised equation of motion in terms of the
-Lagragian displacement $\bm{\xi}$ as given by \cite{lin93b} but with an
+Lagrangian displacement $\bm{\xi}$ as given by \cite{lin93b} but with an
 additional potential perturbation, 
@@
   \frac{\delta p}{\rho}\frac{\nabla\rho}{\rho} +
-  \frac{\delta\rho}{\rho}\frac{\nabla p}{\rho} - - R
+  \frac{\delta\rho}{\rho}\frac{\nabla p}{\rho} -  R
   \hat{\bm{R}}\left(\bm{\xi}\cdot\nabla\Omega^2\right),
@@
 between Eq. \ref{lagragian_pert} and $(-m/2)\rho\bm{\xi}^*$, then
-taking the imaginary part afterwards. The left hand side bcomes the
-angluar momentum density. The first term on the right hand side (RHS)
+taking the imaginary part afterwards. The left hand side becomes the
+angular momentum density. The first term on the right hand side (RHS)
 becomes 
@@
 where $\delta\rho = - \nabla\cdot\left(\rho\bm{\xi}\right)$ and
-$\nabla^2\delta\Phi_d = 4\pi G \delta \rho$ have been used. The first
-term on the RHS of Eq. \ref{angflux1} is (minus) the 
+$\nabla^2\delta\Phi_d = 4\pi G \delta \rho$ have been used. The 
+terms in square brackets on the RHS of Eq. \ref{angflux1} is (minus) the 
 angular momentum flux. The second term on RHS of Eq. \ref{angflux1}, 
@@
 where $\Delta\rho = \delta\rho + \bm{\xi}\cdot\nabla\rho$ is the
-Lagragian density perturbation.
+Lagrangian density perturbation.
 
@@
 
-\section{Relation between horizontal Lagragian displacements for
-  local, low frequency $m=1$ disturbances}\label{horizontal_displacements}
+\section{Relation between horizontal Lagrangian displacements for
+  local, low frequency \lowercase{$m=1$} disturbances}\label{horizontal_displacements}
 Here, we aim to relate the horizontal Lagrangian displacements $\xi_R$

### 2014-11-20 paper/discussion.tex
spell checked

@@
 linear theory to construct an explanation for their growth. For
-simplicitly, we return to 2D discs.
+simplicity, we return to 2D discs.
 
@@
 (\S\ref{fargo_m1}). Then to a first approximation, we can regard it 
-as a neutral, low-frequency $m=1$ mode in satifying the 
+as a neutral, low-frequency $m=1$ mode in satisfying the 
 dispersion relation Eq. \ref{dispersion} in local theory.  %attribute
@@
 temperature gradient is negative, which enforces the mode, because it
-has negative angular mometum.  
+has negative angular momentum.  
 % Note that
@@
 where the factor of two accounts for the fact that the angular momentum
-density is quadradtic in the linear perturbations. Inserting the above
+density is quadratic in the linear perturbations. Inserting the above
 expressions for $\jlin$ and $T_\mathrm{BG}$ for $m=1$ gives
@@
 should have $Q^2(1-\nu^2)=1$ at two radii. Assuming 
-Keperlian rotation and a slow pattern speed $\Omega_p\ll\Omega$, this amounts to
+Keplerian rotation and a slow pattern speed $\Omega_p\ll\Omega$, this amounts to
 \begin{align}
@@
 \end{align}   
-where a Keperian disc has been assumed and subscript $L$ denotes
+where a Keplerian disc has been assumed and subscript $L$ denotes
 evaluation at the OLR, $R=R_L$. (The inner Lindblad resonance does not

### 2014-11-20 paper/model.tex
spell checked

@@
 where $\Delta_{1,2} = \delta R \times H(R_{1,2})$. The bump function
-mimicks mass built-up in $R\in[R_1,R_2]$.
+mimics mass built-up in $R\in[R_1,R_2]$.
 
@@
 to its azimuthal transport algorithm, which circumvents time-step
-contraint set by the inner disc boundary \citep{masset00a,masset00b}.
+constraint set by the inner disc boundary \citep{masset00a,masset00b}.
 
@@
 $R\in[R_\mathrm{min},R_\mathrm{max}],\,\phi\in[0,2\pi]$ and is 
-divided into $(N_R,N_\phi)$ grids, logarithimically spaced in radius and
+divided into $(N_R,N_\phi)$ grids, logarithmically spaced in radius and
 uniformly spaced in azimuth. At radial boundaries we set the
@@
 \end{align}
-using Fast Fouier Transform (FFT), where $\epsilon_g$ is a softening
+using Fast Fourier Transform (FFT), where $\epsilon_g$ is a softening
 length to prevent a numerical singularity. The FFT approach requires
@@
 setup is the same as that adopted in ZEUS-MP above. We configure the
-code similarly to that used in \cite{lin14}: piecewise linear
+code similarly to that used in \cite{lin14}: piece-wise linear
 reconstruction, a Roe solver and second order Runge-Kutta time
@@
 where $\sbar = \sigma - m\Omega$ and $\kappa^2 =
-R^{-3}\p_R(R^4\Omega^2)$ is the square of the epicycle frequency. A
+R^{-3}\p_R(R^4\Omega^2)$ is the square of the epicyclic frequency. A
 locally isothermal equation of state has been assumed.    
@@
 \end{align}
-is the angular momentum density, $\bm{\xi}$ is the Lagragian
+is the angular momentum density, $\bm{\xi}$ is the Lagrangian
 displacement, and $\bm{F}$ is the vertically-integrated angular
@@
 the perturbation is conserved, provided that there is no net angular
-momeumtum flux.% and the perturbations are regular.
+momentum flux.% and the perturbations are regular.
 
@@
 waves outside co-rotation ($R>R_c$) have positive angular
-momentum. The local approximation becomes less good for gobal modes,
+momentum. The local approximation becomes less good for global modes,
 but will nevertheless be useful for results interpretation.

### 2014-11-20 paper/paper.tex
spell checked

@@
 spiral arms can provide significant angular momentum transport
-neccessary for mass accretion \citep{lynden-bell72, papaloizou91,balbus99,lodato04}, and
+necessary for mass accretion \citep{lynden-bell72, papaloizou91,balbus99,lodato04}, and
 spiral structures due to GI are potentially observable with the Atacama Large 
@@
 We summarise in \S\ref{summary}, with a discussion on
-spectulations for future work.
+speculations for future work.
 

### 2014-11-20 paper/results.tex
spell checked

@@
 low $m\leq 2$ modes decay initially, but grows between $t\in[20,40]P_0$,
-possibily due to non-linear interaction of the high-$m$ modes  
+possibly due to non-linear interaction of the high-$m$ modes  
 \citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
@@
 Fig. \ref{2d_fargo_viz} shows the $m=1$ surface density
-at the end of this simulation. For this one-armeded spiral
+at the end of this simulation. For this one-armed spiral
 we measure a co-rotation radius and growth rate  
@@
 is a good match between the two torques, especially at early times
-$t\lesssim110P_0$. The average discrepency is $\simeq 5\%$. 
+$t\lesssim110P_0$. The average discrepancy is $\simeq 5\%$. 
 The match is less good later on, when the spiral

### 2014-11-20 paper/results_3d.tex
spell checked

@@
 $\delta = 10^{-3}$ and $M=1$ in Eq. \ref{randpert}. This corresponds
-to a single $m=1$ disturbance in $R\in[R_1,R_2]$, which
-eventually dominates in 2D simulations.
+to a single $m=1$ disturbance in $R\in[R_1,R_2]$.
+%, which
+%eventually dominates in 2D simulations.
 
@@
                                 %resolution?
-In the ZEUS-MP run we observe disturbances developed near
-the inner boundary, which is probably responsible for the initial growth
-for $t<70P_0$. This is likely a numerical artifact and effectively seeds
-the ZEUS-UP simulation with a larger initial perturbation. Results
+
+In the ZEUS-MP run, we observed high-$m$ disturbances developed near 
+the inner boundary initially, which is likely responsible for the
+growth seen at $t<50P_0$. This is a numerical artifact and effectively
+seeds the simulation with a larger perturbation. Results
 from ZEUS-MP are therefore off-set from PLUTO by $\sim50P_0$. However,
@@
   &\gamma \simeq 0.0073\Omega_k(R_0) \quad\quad \mathrm{PLUTO},\\
-  &\gamma \simeq 0.0084\Omega_k(R_0) \quad\quad \varA{ZEUS-MP}.
+  &\gamma \simeq 0.0085\Omega_k(R_0) \quad\quad \varA{ZEUS-MP}.
 \end{align*}
-Both are smaller than the 2D simulations, because of to the
-low resolutions adopted in 3D and/or the small softening length employed in
-2D. 
-%Fig. \ref{3d_ampmax} also show results from strictly isothermal
-%simulations, which display no growth compared to cases with a 
-%temperature gradient.
-
-% We also show results from strictly
-% isothermal discs ($q=0$). 
-%growth rate does not depend on pert amp
+Both are smaller than the 2D simulations, because of the 
+lower resolutions adopted in 3D and/or the small softening length
+employed in 2D. 
+%Rc = 8.6R_0
 
@@
   \caption{Evolution of the maximum $m=1$ density component in  $r\in[R_1,R_2]$
-    measured in the 3D simulations. Results from discs with a imposed 
+    in the 3D simulations. Results from discs with a 
     temperature gradient ($q=1$) and a strictly isothermal disc
@@
 
-Visualisations of the 3D simluations are shown in   
+Visualisations of the 3D simulations are shown in   
 Fig. \ref{3d_prelim} for the disc midplane and near the upper disc
@@
 have reached comparable amplitudes. The agreement between the codes,
-as well as with the 2D simulations, is satisfactory. The largest
-spiral amplitude is found in $R\in[R_1,R_2]$ independent of
-height. However, notice the spiral pattern extends into the
-non-self-gravitating outer disc ($R>R_2$) at $z\sim 2H$, i.e. 
+as well as with the 2D simulations (Fig. \ref{2d_fargo_viz}), is
+satisfactory. The largest spiral amplitude is found in the
+self-gravitating region $R\in[R_1,R_2]$, independent of
+height. However, notice the spiral pattern extends into 
+the non-self-gravitating outer disc ($R>R_2$) at $z\sim 2H$, i.e.    
 the disturbance becomes more global away from the midplane.
-
-%Away from the
-%disc midplane, a coherent spiral arm is not prominent, but the
-%disturbance is radially more global, extending up to $R\sim 4R_0$.
-
-%implications?
 
@@
   \caption{Three-dimensional simulations using the ZEUS-MP (top) and 
-    PLUTO (bottom) codes. The $m=1$ density component $\Delta\rho_1$ at the midplane (left) and
+    PLUTO (bottom) codes. The $m=1$ density component $\Delta\rho_1$
+    at the midplane (left) and approximately 
     two scale-heights above the midplane (right) is shown. Here $\psi
@@
 Fig. \ref{3d_rz} shows the vertical structure of the one-armed 
-spiral in the PLUTO run. The main spiral is confined to $z < H$. Thus,
-a 2D disc model, representing dynamics near the disc midplane, 
-should be sufficient to study the instability. Interestingly, however, for $R>2R_0$ the
-spirals  increase in amplitude away from the midplane. It remains
-small in our disc model ($|\Delta\rho| < 0.1$), but could become 
-significant with a larger vertical domain. The figure shows that 
-the one-armed spiral can have a global influence away from the
-midplane of the self-gravitating annulus.
+spiral in the PLUTO run. The spiral is vertically confined to $z
+\lesssim H$ at $R\sim R_0$ (the self-gravitating region). Thus, a 2D
+disc model, representing dynamics near the disc midplane, is  
+sufficient capture the instability. However, for $R>2R_0$ the spiral amplitude
+increases away from the midplane. It remains  
+small in our disc model ($|\Delta\rho_1| \lesssim 0.1$), but could become 
+significant with a larger vertical domain. This means that 3D 
+simulations are necessary to study the effect of the one-armed spiral
+on the exterior, non-self-gravitating disc.
 
@@
   0.64cm]{figures/pdisk_rz_023_nsg.ps}  
-  \caption{Vertical distribution of the $m=1$ density component in the meridional plane from the 
+  \caption{The $m=1$ density component in the meridional plane in the 
     PLUTO simulation. The slices are taken at the azimuth of   
-    $\mathrm{max}[\Delta\rho_1(r,\pi/2,\phi)]$. Arrows correspond to the vector 
+    $\mathrm{max}[\Delta\rho_1(r,\pi/2,\phi)]$. Arrows represent the vector 
     $(v_r/R_0,-v_\theta/rh\sin^2{\theta})$. The top (bottom) panel corresponds
@@
 Although Fig. \ref{3d_rz} appears to display significant vertical motion,
-the three-dimensionality parameter $\Theta < 10^{-2}$
-(Eq. \ref{theta}), implying the kinetic energy associated with
-vertical motion is insignificant compared to that in horizontal
-motion. We find  $|v_z/c_s|\lesssim 
-0.2$ for $h=0.05$. Thus, vertical motions are sub-sonic and negligible
-compared to horizontal motions, which supports a 2D approximation.
-
-%negligible vertical motion
-%unstratified approxi good?
-
-%This again supports the 2D approximation, altthou
-
-%Vertical velocities
-%do increase with the growth of the one-armed spiral, 
-%Larger amplitudes
-%are reached in ZEUS because of larger numerical noise. 
-%but we find  $\Theta \lesssim 10^{-2}$.
-
-%\begin{figure}
-%  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_vert.ps}
-% \caption{Evolution of three-dimensionality $\Theta$, as defined by
-%    Eq. \ref{theta}, which corresponds to the
-%   average vertical kinetic energy density, scaled by 
-%    the average horizontal kinetic energy.\label{3d_vert}} 
-%\end{figure}
+we measure the three-dimensionality parameter $\Theta < 10^{-2}$
+(Eq. \ref{theta}), so vertical motion is insignificant compared to
+horizontal motion. This supports a 2D approximation. On the other
+hand, we find  $|v_z/c_s|\lesssim 0.2$ which, although sub-sonic, is
+not very small.
 
@@
 Fig. \ref{3d_angmom} shows the angular momentum evolution in the 3D 
-runs. ZEUS-MP does not conserve angular momentum very well, but the
-variation $|\Delta J/J|< O(10^{-6})$ is small compared 
-to the individual components $|\Delta J_{0,1}/J|\sim 
-5\times10^{-5}$. The PLUTO run reaches similar values of
-$|J_{0,1}|$, but achieves better conservation, with $|\Delta
-J/J|=O(10^{-8})$. These plots are again similar to the 2D simulations,
-i.e. angular momentum lost by $J_1$ is gained by $J_0$. This confirms 
-that the interaction between $J_1$ and $J_0$ operates in 3D and 2D
-similarly.
+runs during the linear growth of the one-armed spiral. Because the
+ZEUS-MP simulation is off-set from PLUTO, the time interval for the
+plot was chosen such that the change in the angular momentum
+components are comparable in the two codes.
+
+ZEUS-MP does not conserve angular momentum very well, but the
+variation in total angular momentum $|\Delta J/J|< O(10^{-6})$ is
+small compared to the individual components $|\Delta J_{0,1}/J|\sim
+10^{-4}$. %it worsens with time 
+PLUTO reaches similar values of $|\Delta J_{0,1}|$, but achieves better
+conservation, with $|\Delta J/J|=O(10^{-8})$. These plots are again
+similar to the 2D simulations, i.e. angular momentum lost by $J_1$ is
+gained by $J_0$. This confirms that the interaction between $J_1$ and
+$J_0$ operates in 3D and 2D similarly.
 

### 2014-11-20 paper/summary.tex
minor updates to tex
spell checked

@@
 $Q$-barriers owing to a surface density bump. Growth rates increased
-with the magnitude of the imposed temperature gradient, and one-armeded
+with the magnitude of the imposed temperature gradient, and one-armed
 spirals did not develop in strictly isothermal simulations.  
@@
 %neutral mode destabilized      
-\subsection{Spectulations and future work}
+\subsection{Speculations and future work}
+%accretion??
 %balancing heating and cooling 
@@
 cooling (and heating) timescales, so the disc temperature instantly
-returns to its inital value. %so we don't go to deeply non-linear
+returns to its initial value. %so we don't go to deeply non-linear
                              %regime 

### 2014-11-20 paper/appendix.tex
first draft done, waiting on timeres results

@@
 vanishes. When there is a forced temperature gradient,
-Eq. \ref{tbg_general} indicates there is angular momentum exchange
-between compressible perturbations ($\nabla\cdot\bm{\xi}\neq0$) and
-the background disc if there is motion along the temperature
-gradient ($\bm{\xi}\cdot\nabla c_s^2 \neq 0$).
+Eq. \ref{tbg_general} indicates a torque is applied to compressible
+perturbations ($\nabla\cdot\bm{\xi}\neq0$) if there is motion along
+the temperature gradient ($\bm{\xi}\cdot\nabla c_s^2 \neq 0$).
 

### 2014-11-20 paper/discussion.tex
first draft done, waiting on timeres results

@@
 analogous to disc-satellite interaction 
-\citep{goldreich79}, and may lead to instability 
-if the angular momentum associated with the spiral has the opposite 
-sign to that of the density waves it induces in the exterior disc
-\citep{lin11b}. Here, we estimate the magnitude of this effect using
-basic results from disc-planet theory \citep[see, e.g.][and references 
-therein]{papaloizou07}.
+\citep{goldreich79}. Here, we estimate the magnitude of this effect
+using basic results from disc-planet theory \citep[see, e.g.][and
+references therein]{papaloizou07}.
 
@@
 
-%and here we are interested in the
-%potential it produces at $R>\overline{R}$  
-%This potential is associated with a mass
-%$M_\mathrm{ring}$ contained in   is the mass in $R\in[R_1,R_2]$.
-
 %For an order-of-magnitude exercise, we simply take 
@@
   \frac{\pi^2\Sigma_L}{3\Omega_L\Omega_p}
-  \left[\left.R_L\frac{d\Phi_\mathrm{ext}}{dR}\right|_L + 2\left(1 - \frac{\Omega_p}{\Omega_L}\right)\Phi_\mathrm{ext}\right]^2,
+  \left[\left.R_L\frac{d\Phi_\mathrm{ext}}{dR}\right|_L + 2\left(1 -
+      \frac{\Omega_p}{\Omega_L}\right)\Phi_\mathrm{ext}\right]^2, 
 \end{align}   
@@
   \frac{\gamma_L}{\Omega_p} = &\frac{\pi
-    h}{3Q_L}\left(\frac{M_p}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)\left(\frac{R_c}{\overline{R}}\right)^3\left(\frac{R_L}{R_c}\right)^{-3/2}\notag\\
-  &\times \left\{\frac{R_L}{\overline{R}}\left.\frac{db_{1/2}^1}{d\beta}\right|_L + 2\left[1 -
+    h}{3Q_L}\left(\frac{M_p}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)\left(\frac{R_c}{\overline{R}}\right) 
+  ^3\left(\frac{R_L}{R_c}\right)^{-3/2}\notag\\ 
+  &\times
+  \left\{\frac{R_L}{\overline{R}}\left.\frac{db_{1/2}^1}{d\beta}\right|_L
+    + 2\left[1 - 
       \left(\frac{R_c}{R_L}\right)^{-3/2}\right]b_{1/2}^1(\beta_L)\right\}^2. 
@@
 
-% both are positive -> destabilizing in principle
+The torque exerted on the disc at the OLR by an
+external potential is positive, while that at co-rotation 
+depends on the gradient of potential vorticity there
+\citep{goldreich79}. For our disc models with surface density
+$\Sigma\propto R^{-2}$ in the outer disc, this co-rotation torque is
+positive. Interpreting the one-armed spiral as an external potential
+for the outer disc, this means that the spiral loses
+angular momentum by launching density waves with positive angular momentum at
+the OLR, and by applying a positive co-rotation torque on the disc. In
+principle, this interaction is destabilising because the one-armed
+spiral has negative angular momentum \citep{lin11b}. As estimated
+above, this effect is small in our simulations, but could be
+important for other parameter regimes.
 
-%\begin{align}
-%  \gamma_C\sim 0.010\Omega_p,
-%\end{align}
-%which is again small compared to the measured growth rates.
-
-%not important for our models, but may be important if outer disc is
-%self-gravitating too
-
-%expressions for linear response, but have massive "planet" -> but CR
-%and LR far from where spiral is
-
-%required disc structure (for confined SG spiral)
-
-%\subsection{Speculations}
-%speculations - long term balance. forced temp gradient drives
-%instability (disc gains ang mom, spiral loses ang mom). but if spiral
-%shocks and dissiple, it dumps negative angular momentum (disc gains
-%negative ang mom). balance between gain from forced temp gradient and
-%loss due to dissiplatoin
-
-%other source of m=1 disturbance -> eccentric modes (adams, PJ06,
-%hopkins, tremaine)
+%also transport ang mom outwards

### 2014-11-20 paper/results.tex
first draft done, waiting on timeres results

@@
 \begin{figure}
-  \includegraphics[width=\linewidth]{figures/polarxy2_dens120_fargo.ps}
+  \includegraphics[width=\linewidth]{figures/polarxy2_dens120_fargo}
   \caption{Cartesian visualisation of the $m=1$ surface density
@@
 \begin{figure}
-  \includegraphics[width=\linewidth]{figures/m1_analysis_kr120_fargo.ps}
+  \includegraphics[width=\linewidth]{figures/m1_analysis_kr120_fargo}
   \caption{Normalised radial wavenumber of the $m=1$ spiral in 
@@
 \begin{figure}
-  \includegraphics[width=\linewidth]{figures/m1_analysis_Qbar_fargo.ps} 
+  \includegraphics[width=\linewidth]{figures/m1_analysis_Qbar_fargo} 
   \caption{Dimensionless mode frequency $\nu$ for the $m=1$ spiral in
@@
 \begin{figure}
-  \includegraphics[width=\linewidth]{figures/m1_analysis_ang_fargo.ps} 
+  \includegraphics[width=\linewidth]{figures/m1_analysis_ang_fargo} 
   \caption{Rate of change of the $m=1$ wave angular momentum as defined by
@@
 \begin{figure}
-  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_fargo_varq.ps}   
+  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_fargo_varq}   
   \caption{Evolution of the $m=1$ spiral amplitude as a function of
@@
 \begin{figure}
-  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varq.ps}    
+  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varq}    
   \caption{Growth rates of the $m=1$ spiral mode as a function of the
@@
 \begin{figure}
-  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varh.ps}    
+  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varh}    
   \caption{Growth rates of the $m=1$ spiral mode as a function of the

### 2014-11-20 paper/results_3d.tex
first draft done, waiting on timeres results

@@
 \begin{figure}
-  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ampmax3d.ps}
+  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ampmax3d}
   \caption{Evolution of the maximum $m=1$ density component in  $r\in[R_1,R_2]$
@@
   \includegraphics[scale=0.47,clip=true,trim=0cm 0.79cm 0cm
-  0cm]{figures/pdisk_rz_023_sg.ps}  
+  0cm]{figures/pdisk_rz_023_sg}  
   \includegraphics[scale=0.47,clip=true,trim=0cm 0cm 0cm
-  0.64cm]{figures/pdisk_rz_023_nsg.ps}  
+  0.64cm]{figures/pdisk_rz_023_nsg}  
   \caption{The $m=1$ density component in the meridional plane in the 

### 2014-11-20 paper/summary.tex
first draft done, waiting on timeres results

@@
 trailing spirals that persist for at least $O(10^2)$ orbits. The 
-spiral pattern speed is smaller than the local disc rotation, making
-it almost stationary. The growth rates are $O(10^{-2}\Omega)$ which
-gives a characteristic growth time of $O(10P_0)$. 
+spiral pattern speed is smaller than the local disc rotation, and
+growth rates are $O(10^{-2}\Omega)$ which gives a characteristic
+growth time of $O(10)$ orbits.   
 %While this is slow
@@
 spirals did not develop in strictly isothermal simulations.  
-We find no significant vertical 
-motion in 3D simulations, which indicates the instability is
-two-dimensional.
+We find the instability behaves similarly in 2D and 3D. However, in 3D
+the spiral disturbance can become radially global away from 
+the midplane.
 
 We applied angular momentum conservation within linear theory to
-explain our numerical results. A  
+explain our numerical results. We find a   
 forced temperature gradient introduces a torque on linear 
@@
 represents an exchange of angular momentum between the background disc
-and the perturbations. In the local approximation, we showed that this
-background torque is negative for trailing spirals, which enforces  
+and the perturbations. This offers a previously unexplored pathway to
+instability.
+
+In the local approximation, we showed that this background torque is
+negative for trailing spirals in discs with a fixed temperature or
+sound-speed profile that decrease outwards. This torque enforces  
 low-frequency $m=1$ modes because they are associated with negative
 angular momentum. We therefore interpret the one-armed spiral instability
-observed in our simulations as a neutral, tightly-wound $m=1$ mode
-being destabilised by the imposed temperature gradient.
+observed in our simulations as an initially neutral, tightly-wound
+$m=1$ mode being destabilised by the imposed temperature gradient.
 
-%works in 2D/3D
-%forced gradient feedback 
-%neutral mode destabilized      
 \subsection{Speculations and future work}
-%accretion??
-%balancing heating and cooling 
+%numerical simulations to include cooling/heating
+A locally isothermal equation of state represents the limit of
+infinitely short cooling (and heating) timescales, so the disc
+temperature instantly returns to its initial value when perturbed.  
+This assumption can be relaxed by including an energy equation with
+thermal relaxation. Preliminary FARGO simulations
+indicate a thermal relaxation timescale $t_c < 0.1\Omega_k^{-1}$ is
+needed for the one-armed spiral to develop. However, this value may
+depend on disc parameters, and will be explored in a follow-up  
+study.
+
+% prelim sims show they are long lasting 
+% balancing heating and cooling 
+% should not get fragmentation (CR away from spiral) 
 In the deeply non-linear regime, the one-armed spiral may 
-shock and dissipate, depositing negative angular momentum onto
+shock and deposit negative angular momentum onto
 the background disc. The spiral amplitude would saturate by gaining
 positive angular momentum. However, if the temperature gradient is
-forced, it may be possible to achieve a balance between the gain of negative
-angular momentum through the forced temperature gradient, and the gain
-of positive angular momentum through shock dissipation. In order to
-study this possibility, improved numerical models are needed to ensure 
-total angular momentum conservation in long-term simulations.
+maintained, it may be possible to achieve a balance between the gain
+of negative angular momentum through the temperature gradient, and the
+gain of positive angular momentum through shock dissipation. We remark  
+that fragmentation is unlikely because the co-rotation radius 
+is outside the bulk of the spiral arm, in the non-self-gravitating
+portion of the disc \citep{durisen08}. In order
+to study these possibilities, improved numerical models are needed to
+ensure total angular momentum conservation on timescales much longer
+than that considered in this paper.
 
-%numerical simulations to include cooling/heating
-A locally isothermal disc represents the limit of infinitely short
-cooling (and heating) timescales, so the disc temperature instantly
-returns to its initial value. %so we don't go to deeply non-linear
-                             %regime 
-This assumption can be relaxed by including an energy
-equation with thermal relaxation. Preliminary FARGO simulations
-indicate a relaxation timescale $t_c < 0.1\Omega_k^{-1}$ is needed for
-the one-armed spiral to develop. This will be presented in a follow-up
-study. %prelim sims show they are long lasting
-
-%non-sg sims, Be stars, eccentric discs
+% non-sg sims, Be stars, eccentric discs
 % We stress that destabilisation through the temperature gradient
@@
 % its origin.
+
 In our models and interpretation, self-gravity coupled with the disc 
 structure permits a neutral one-armed mode, which is then
-destabilised. Other origins of a neutral
+destabilised by the temperature gradient. Other origins of the neutral
 one-armed spiral should be investigated. One possibility is in Be star discs

### 2014-11-20 paper/paper.tex
minor updates

@@
   not explicitly depend on self-gravity, we suggest it may also
-  affect low-frequency one-armed spirals in non-self-gravitating
+  affect low-frequency one-armed oscillations in non-self-gravitating
   discs. 

### 2014-11-26 paper/discussion.tex
minor updates
second draft

@@
 
+The dominant disturbance in our simulations is an one-armed, $m=1$
+trailing spiral, and its growth is associated with the imposed temperature
+gradient. As discussed in \S\ref{global_cons}, in a locally isothermal disc this imposed
+temperature gradient generally results in a torque being exerted on
+linear perturbations. An instability is therefore possible if the sign
+of this torque is the same as that of the angular momentum associated
+with the linear perturbation.
+
+%associated angular mometum density is the same as that
+%of this
 % According to previous studies \citep{adams89}, our disc models
@@
   \jlin &\simeq \frac{\Sigma\Omega}{2}\left[-|\bm{\xi}|^2 + \ii\left(\xi_R\xi_\phi^* -
-      \xi_R\xi_\phi^*\right)\right]\notag\\
+      \xi_R^*\xi_\phi\right)\right]\notag\\
   & = -\frac{\Sigma\Omega}{2}|\xi_R + \ii\xi_\phi|^2\notag\\
@@
 to the background, then we can expect instability. We demonstrate
-this below.
+below how this is possible through a forced temperature
+gradient.
 
@@
 real, and assume $|kR|\gg m$ so that the second term on the right hand
-side of Eq. \ref{den_pert} is negligible. Then    
+side of Eq. \ref{den_pert} can be neglected. Then    
 \begin{align}
@@
 
-Using $\jlin$ and $T_\mathrm{BG}$, we can obtain a rough estimate of
+Using $\jlin$ and $T_\mathrm{BG}$, we can estimate 
 the growth rate $\gamma$ of linear perturbations due to the background
@@
 $q=1$, $Q=O(1)$, $h=0.05$, this gives $\gamma/\Omega \sim O(10^{-2})$,
-consistent with simulation results.
+consistent with simulation results. Specifically,
+Eq. \ref{theoretical_rate} indicates the growth rate should increase
+linearly with $q$ and $h$, as seen in the simulations
+(Fig. \ref{fargo_varq_growth}---\ref{fargo_varh_growth}).
 
+%Note that changing $q$ affects the magntiude of the background
+%torque through its explicit depdence on the temperature gradient. By
+%definition, the global temperature gradient does not affect modes in
+%local theory. 
+%however, changing $h$ may also modify the mode properties (since
+%temperature appears in local theory) in addition
+%to the magnitude of the temperature gradient. Thus, we expect the depedence of
+%the background torque on $q$ to be better described by
+%Eq. \{theoretical_rate}
+
+%by definition, q does not affect local neutral modes
 %Of course, we assumed a neutral mode to obtain expressions
 %for $\jlin$ and $T_\mathrm{BG}$ in the first place, 
-Note that by using local theory we have implicitly
-assumed a neutral mode because the dispersion relation,
-Eq. \ref{dispersion}, states that non-axisymmetric modes are
-stable for real $k$. Eq. \ref{theoretical_rate} should be interpreted
-as an order-of-magnitude estimate of the perturbative effect of a
-temperature gradient on local, low-frequency $m=1$ modes.
+%Recall that in local theory non-axisymmetric modes are stable for real
+%$k$  
+%Eq. \ref{theoretical_rate} should be interpreted
+%as an order-of-magnitude estimate of the perturbative effect of a
+%temperature gradient on local, low-frequency $m=1$ modes.
 
@@
 \subsection{The role of self-gravity and disc structure} 
-In the above discussion, we only considered a self-gravitating
+It is important to note that in the above discussion, we only considered a self-gravitating
 perturbation to obtain an appropriate wavenumber $k$ needed to
@@
 self-gravity explicitly, so self-gravity is not directly
-responsible for instability. However, 
+responsible for instability.
+
+However, 
 our numerical simulations show that the $m=1$ spiral is

### 2014-11-26 paper/model.tex
minor updates
second draft

@@
 $q=0$ has $H\propto R^{3/2}$, and $q=1$ corresponds to a 
-a disc with constant $H/R$.
+ disc with constant $H/R$.
 
@@
 
-\section{Numerical methods}\label{methods}
-We adopt computational units such that $G=M_*=R_0=1$. Time is measured
-in the Keplerian orbital period at the reference radius, $P_0\equiv 
-2\pi/\Omega_k(R_0)$.
-
-\subsection{FARGO}
-Our primary code is FARGO with self-gravity \citep{baruteau08}. This
-is a popular, simple finite-difference code for 2D discs. `FARGO' refers
-to its azimuthal transport algorithm, which circumvents time-step
-constraint set by the inner disc boundary \citep{masset00a,masset00b}.
-
-The 2D disc occupies
-$R\in[R_\mathrm{min},R_\mathrm{max}],\,\phi\in[0,2\pi]$ and is 
-divided into $(N_R,N_\phi)$ grids, logarithmically spaced in radius and
-uniformly spaced in azimuth. At radial boundaries we set the
-hydrodynamic variables to their initial values.
-
-The 2D Poisson equation is solved in integral form, 
-\begin{align}\label{2d_grav}
-  &\Phi_{d,z=0}(R,\phi) \notag \\
-  &=-\int_{R_\mathrm{min}}^{R_\mathrm{max}} \int_0^{2\pi}
-  \frac{G\Sigma(R^\prime,\phi^\prime)R^\prime dR^\prime d\phi^\prime}{\sqrt{R^2+R^{\prime 2} -
-      2RR^\prime\cos{(\phi - \phi^\prime)} + \epsilon_g^2}}, 
-\end{align}
-using Fast Fourier Transform (FFT), where $\epsilon_g$ is a softening
-length to prevent a numerical singularity. The FFT approach requires
-$\epsilon_g\propto R$ \citep{baruteau08}. In FARGO, $\epsilon_g$ is
-set to a fraction of $hR$.
-
-% poisson kernal integration, softening 
-% where used?
-% BC
-\subsection{ZEUS-MP}
-ZEUS-MP  is a general-purpose finite difference
-code \citep{hayes06}. We use the code in 3D spherical geometry, covering
-$r\in[r_\mathrm{min},r_\mathrm{max}]$, $\theta\in[\theta_\mathrm{min},\pi/2]$,
-$\phi\in[0,2\pi]$. The vertical domain is chosen to cover $n_H$
-scale-heights at $R=R_0$, i.e. $\tan{(\pi/2 - \theta_\mathrm{min})}/h=n_H$. 
-The grid is logarithmically spaced in radius and uniformly spaced in the angular
-coordinates. We assume reflectional symmetry across the midplane, and
-apply reflective boundary conditions at radial boundaries and the
-upper disc boundary.
-
-ZEUS-MP solves the 3D Poisson equation using a conjugate gradient
-method. To supply boundary conditions to the linear solver, we
-expand the boundary potential in spherical harmonics $Y_{lm}$ 
-as described in \cite{boss80}. The expansion is truncated at
-$(l,m)=(\lmax,\mmax)$. This code was used in \cite{lin12b} for
-self-gravitating disc-planet simulations.
-
-%used in lin 2012 
-\subsection{PLUTO} 
-PLUTO is a general-purpose Godunov code \citep{mignone07}. The grid
-setup is the same as that adopted in ZEUS-MP above. We configure the
-code similarly to that used in \cite{lin14}: piece-wise linear
-reconstruction, a Roe solver and second order Runge-Kutta time
-integration. We also enable the FARGO algorithm for azimuthal
-transport.
-
-We solve the 3D Poisson equation throughout the domain using spherical
-harmonic expansion \citep{boss80}, as used for the boundary potential
-in ZEUS-MP. This version of PLUTO was used in \cite{lin14b} for
-self-gravitating disc-planet simulations, producing similar results to
-that of ZEUS-MP and FARGO.
-
-%used in conference proceeding
-
-\subsection{Diagnostics}
-
-\subsubsection{Evolution of non-axisymmetirc modes}
-The disc evolution is quantified using mode amplitudes and angular
-momenta as follows. We list the 2D definitions with obvious 3D generalisations. 
-A hydrodynamic variable $f$ (e.g. $\Sigma$) is written as 
-\begin{align}
-  f(R,\phi,t) &= \sum_{m=-\infty}^{\infty}f_m(R,t)\exp{\ii m \phi} \notag\\
-  &= f_0(R) + 2 \real\left[\sum_{m=1}^\infty f_m \exp{(\ii
-      m\phi)}\right], 
-\end{align}
-where the $f_m$ may be obtained from Fourier transform in $\phi$.
-
-The normalised surface density with azimuthal wavenumber $m$ is
-\begin{align}
-  \Delta\Sigma_m = \frac{2}{\Sigma_{00}} \real\left[\Sigma_m \exp{(\ii
-      m\phi)}\right]
-\end{align}
-where $\Sigma_{00} = \Sigma_0(t=0)$. The time evolution of the
-$m^\mathrm{th}$ mode can be characterized by 
-assuming $\Sigma_m\propto\exp{(-\ii \sigma t)}$, where
-$\sigma = \omega + \ii\gamma$ is a complex frequency, $\omega$ being 
-the real frequency and $\gamma$ is the growth rate. The total
-non-axisymmetric surface density is 
-\begin{align}
-  \Delta\Sigma = \frac{\Sigma - \Sigma_0}{\Sigma_0}. 
-\end{align}
-
-\subsubsection{Angular momentum decomposition}
-The total disc angular momentum is
-\begin{align}
-  J &= \int_{R_\mathrm{min}}^{R_\mathrm{max}}\int_0^{2\pi} \Sigma Rv_\phi RdRd\phi \notag\\
-  &= 2\pi\int_{R_\mathrm{min}}^{R_\mathrm{max}} R\Sigma_0 v_{\phi0} R dR \notag\\ 
-  &\phantom{=}+
-  \sum_{m=1}^\infty2\pi\int_{R_\mathrm{min}}^{R_\mathrm{max}} 2R\real\left[\Sigma_m v_{\phi
-      m}^*\right] RdR 
-  %&= \sum_{m=0}^\infty \int _{R_\mathrm{min}}^{R_\mathrm{max}}  j_m(R)
-  %dR 
-  = \sum_{m=0}^\infty J_m. 
-\end{align}
-where $^*$ denotes complex conjugation. We will refer to $J_m$ as the
-$m^\mathrm{th}$ component of the total angular momentum, and use it to
-monitor angular momentum conservation in the numerical simulations. 
-It is important to distinguish this empirical definition from the
-angular momentum of linear perturbations, defined through a
-conservation law, which is given in \S\ref{wkb}.
-
-%and we idenfity $J_m$ as the  
-%angular momentum associated with disc structure with
-%$m^\mathrm{th}$-fold symmetry, and $j_m$ the associated angular
-%momentum per unit length. We will use $J_m$ to monitor angular momentum 
-%conservation for the disc as a whole.
-
-\subsubsection{Three-dimensionality}
-In 3D simulations we measure the importance of vertical motion with
-$\Theta$, where 
-\begin{align}\label{theta}
-  \Theta^2 \equiv \frac{\avg{v_z^2}}{\avg{v_R^2}+\avg{v_\phi^2}}, 
-\end{align}
-and $\avg{\cdot}$ denotes the density-weighted average, e.g., 
-\begin{align}
-  \avg{v_z^2} \equiv\frac{
-    \int_{R_1}^{R_2}\int_{\theta_\mathrm{min}}^{\pi/2} \int_{0}^{2\pi}
-    \rho v_z^2 dV}{
-    \int_{R_1}^{R_2}\int_{\theta_\mathrm{min}}^{\pi/2} \int_{0}^{2\pi}
-    \rho dV
-  },
-\end{align}
-and similarly for the horizontal velocities. Thus $\Theta$ is the
-ratio of the average kinetic energy associated with vertical motion to
-that in horizontal motion. The radial range of
-integration is taken over $r\in[R_1,R_2]$ since this is where
-we find the perturbations to be confined.
-
 \section{Linear density waves}\label{wkb}
-In this paper we take a fully numerical approach using non-linear
-hydrodynamic simulations. However, it will be useful to employ
-linear density wave theory to interpret the simulations.
-We list here the main definitions and results from 2D linear theory,
-taken from \cite{lin11b} and \cite{shu91}.
+In this paper we will mainly use non-linear hydrodynamic simulations to investigate the
+stability of disc equilibra described 
+above. However, it is useful to  interpret simulation results
+using linear density theory. We list here the main definitions and
+results from 2D linear theory, taken from \cite{lin11b} and
+\cite{shu91}.   In particular, we highlight a key feature of
+locally isothermal discs that enables angular momentum exchange
+between linear perturbations and the background disc.
 
@@
 \end{align}
-and similarly for other variables. We take $m>0$. Since the background is
-axisymmetric, we identify $\delta\Sigma_m(R)\exp{\left(-\ii\sigma
-    t\right)} = \Sigma_m$ as defined for diagnostics. 
+and similarly for other variables, where $\sigma=\omega+\ii\gamma$ is
+a complex frequency with $\omega$ being the real frequency, 
+$\gamma$ the growth rate, and $m$ is an integer. Here, we take $m>0$ without
+loss of generality.  
+%Since the background is
+%axisymmetric, we identify $\delta\Sigma_m(R)\exp{\left(-\ii\sigma
+ %   t\right)} = \Sigma_m$ as defined for diagnostics.
+
 The linearised mass and momentum equations are
@@
   &-\ii\sbar\delta v_{Rm} - 2\Omega \delta v_{\phi m} = -
-  c_s^2(R)\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right) - \frac{d}{dR}\delta\Phi_m,\\
+  c_s^2(R)\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right) - \frac{d}{dR}\delta\Phi_m,\label{radmom}\\
   & -\ii\sbar\delta v_{\phi m} + \frac{\kappa^2}{2\Omega}\delta v_{Rm} =
@@
 R^{-3}\p_R(R^4\Omega^2)$ is the square of the epicyclic frequency. A
-locally isothermal equation of state has been assumed.    
-The linearised Poisson equation is 
+locally isothermal equation of state has been assumed in
+Eq. \ref{radmom}.  The linearised Poisson equation is 
 \begin{align}
@@
 
-The linearised equations can be combined into a single
+These linearised equations can be combined into a single
 integro-differential equation eigenvalue problem. We defer a full
 exploration of the linear problem to a future study. Here, we list some
-general  properties of the linear perturbations that will be useful
+general  properties of the linear perturbations that will be relevant
 for describing simulation results.
@@
 \subsection{Global angular momentum conservation for linear
-  perturbations} 
+  perturbations} \label{global_cons}
 It can be shown that linear perturbations with $\phi$-dependence in the form
@@
 in the form 
-\begin{align} 
+\begin{align}\label{angcons}
   \frac{\p \jlin}{\p t} + \nabla\cdot\bm{F} = \tbg, 
@@
 \end{align}
-is the angular momentum density, $\bm{\xi}$ is the Lagrangian
-displacement, and $\bm{F}$ is the vertically-integrated angular
-momentum flux consisting of a Reynolds stress and a gravitational
-stress \citep{lin11b}. Explicit expressions for $\bm{\xi}$ can be found in,
-e.g. \cite{papaloizou85}. 
-%and $\bm{F}$ can be read off Eq. \ref{angflux1}.
+is the angular momentum density of the linear disturbance (which may
+be positive or negative), where $\bm{\xi}$ is the Lagrangian
+displacement and $^*$ denotes complex conjugation, and $\bm{F}$ is the
+vertically-integrated angular momentum flux consisting of a Reynolds
+stress and a gravitational stress \citep{lin11b}. Explicit expressions
+for $\bm{\xi}$ can be found in, e.g. \cite{papaloizou85}.
 
-% \cite{papaloizou85} and \cite{lin11} respectively, and
-
-% lin papaloizou kley 1993, ryu/goodman, narayan
-
-The background torque density $\tbg$ is 
+In Eq. \ref{angcons}, the background torque density $\tbg$ is 
 \begin{align}\label{baroclinic_torque}
@@
 the perturbation is conserved, provided that there is no net angular
-momentum flux.% and the perturbations are regular.
+momentum flux. % and the perturbations are regular.
 
@@
 temperature gradient, as in the disc models we consider,
-then $\tbg\neq 0$ in general, which corresponds to a torque
-exerted by the background disc on the perturbation. We 
-outline the derivation of $T_\mathrm{BG}$ in Appendix 
-\ref{tbg_deriv}.
+then $\tbg\neq 0$ in general, which corresponds to a local torque
+exerted by the background disc on the perturbation. We outline the
+derivation of $T_\mathrm{BG}$ in Appendix \ref{tbg_deriv}.
+
+The important consequence of the background torque is the possibilty
+of instability if $\tbg\jlin>0$. That is, if $\jlin$ is positive
+(negative) and the background torque $\tbg$ is positive (negative),
+then the local angular momentum density of the linear disturbance
+further increases (decreases). This implies the amplitude of the disturbance
+may grow by exchanging angular momentum with the background
+disc. In our disc models with $c_s^2\propto R^{-q}$, we expect the
+growth rate of such an instability to scale with $q$.
 
@@
 where $k$ is a real wavenumber such that $|kR|\gg1$ \citep{shu91}. 
-Note that in this approximation, only axisymmetric perturbations with
-$m=0$ can be unstable (having an imaginary $\sigma$).
+Note that in the local approximation, only axisymmetric perturbations
+($m=0$) can be unstable. % (having an imaginary $\sigma$).
 
@@
 but will nevertheless be useful for results interpretation.
+
+\section{Numerical methods}\label{methods}
+We adopt computational units such that
+$G=M_*=R_0=1$. Time is measured  in the Keplerian orbital period at
+the reference radius, $P_0\equiv  2\pi/\Omega_k(R_0)$.
+
+\subsection{FARGO}
+Our primary code is FARGO with self-gravity \citep{baruteau08}. This
+is a popular, simple finite-difference code for 2D discs. `FARGO' refers
+to its azimuthal transport algorithm, which removes the mean azimuthal
+velocity of the disc, thereby permit larger timesteps than that would otherwise be allowed
+by the usual Courant condition based on the full azimuthal
+velocity \citep{masset00a,masset00b}.
+
+The 2D disc occupies
+$R\in[R_\mathrm{min},R_\mathrm{max}],\,\phi\in[0,2\pi]$ and is  
+divided into $(N_R,N_\phi)$ grids, logarithmically spaced in radius and
+uniformly spaced in azimuth. At radial boundaries we set the
+hydrodynamic variables to their initial values.
+
+The 2D Poisson equation is solved in integral form, 
+\begin{align}\label{2d_grav}
+  &\Phi_{d,z=0}(R,\phi) \notag \\
+  &=-\int_{R_\mathrm{min}}^{R_\mathrm{max}} \int_0^{2\pi}
+  \frac{G\Sigma(R^\prime,\phi^\prime)R^\prime dR^\prime d\phi^\prime}{\sqrt{R^2+R^{\prime 2} -
+      2RR^\prime\cos{(\phi - \phi^\prime)} + \epsilon_g^2}}, 
+\end{align}
+using Fast Fourier Transform (FFT), where $\epsilon_g$ is a softening
+length to prevent a numerical singularity. The FFT approach requires
+$\epsilon_g\propto R$ \citep{baruteau08}. In FARGO, $\epsilon_g$ is
+set to a fraction of $hR$.
+
+% poisson kernal integration, softening 
+% where used?
+% BC
+\subsection{ZEUS-MP}
+ZEUS-MP  is a general-purpose finite difference
+code \citep{hayes06}. We use the code in 3D spherical geometry, covering
+$r\in[r_\mathrm{min},r_\mathrm{max}]$, $\theta\in[\theta_\mathrm{min},\pi/2]$,
+$\phi\in[0,2\pi]$. The vertical domain is chosen to cover $n_H$
+scale-heights at $R=R_0$, i.e. $\tan{(\pi/2 - \theta_\mathrm{min})}/h=n_H$. 
+The grid is logarithmically spaced in radius and uniformly spaced in the angular
+coordinates. We assume symmetry across the midplane, and
+apply reflective boundary conditions at radial boundaries and the
+upper disc boundary.
+
+ZEUS-MP solves the 3D Poisson equation using a conjugate gradient
+method. To supply boundary conditions to the linear solver, we
+expand the boundary potential in spherical harmonics $Y_{lm}$ 
+as described in \cite{boss80}. The expansion is truncated at
+$(l,m)=(\lmax,\mmax)$. This code was used in \cite{lin12b} for
+self-gravitating disc-planet simulations.
+
+%used in lin 2012 
+\subsection{PLUTO} 
+PLUTO is a general-purpose Godunov code \citep{mignone07}. The grid
+setup is the same as that adopted in ZEUS-MP above. We configure the
+code similarly to that used in \cite{lin14}: piece-wise linear
+reconstruction, a Roe solver and second order Runge-Kutta time
+integration. We also enable the FARGO algorithm for azimuthal
+transport.
+
+We solve the 3D Poisson equation throughout the domain using spherical
+harmonic expansion \citep{boss80}, as used for the boundary potential
+in ZEUS-MP. This version of PLUTO was used in \cite{lin14b} for
+self-gravitating disc-planet simulations, producing similar results to
+that of ZEUS-MP and FARGO.
+
+%used in conference proceeding
+
+\subsection{Diagnostics}
+
+\subsubsection{Evolution of non-axisymmetirc modes}
+The disc evolution is quantified using mode amplitudes and angular
+momenta as follows. We list the 2D definitions with obvious 3D generalisations. 
+A hydrodynamic variable $f$ (e.g. $\Sigma$) is written as 
+\begin{align}
+  f(R,\phi,t) &= \sum_{m=-\infty}^{\infty}f_m(R,t)\exp{\ii m \phi} \notag\\
+  &= f_0 + 2 \real\left[\sum_{m=1}^\infty f_m \exp{(\ii
+      m\phi)}\right], 
+\end{align}
+where the $f_m$ may be obtained from Fourier transform in $\phi$.
+
+The normalised surface density with azimuthal wavenumber $m$ is
+\begin{align}
+  \Delta\Sigma_m = \frac{2}{\Sigma_{00}} \real\left[\Sigma_m \exp{(\ii
+      m\phi)}\right]
+\end{align}
+where $\Sigma_{00} = \Sigma_0(t=0)$. The time evolution of the
+$m^\mathrm{th}$ mode can be characterized by 
+assuming $\Sigma_m\propto\exp{(-\ii \sigma t)}$ as in linear
+theory. 
+%where
+%$\sigma = \omega + \ii\gamma$ is a complex frequency, $\omega$ being 
+%the real frequency and $\gamma$ is the growth rate. 
+The total non-axisymmetric surface density is 
+\begin{align}
+  \Delta\Sigma = \frac{\Sigma - \Sigma_0}{\Sigma_0}. 
+\end{align}
+
+\subsubsection{Angular momentum decomposition}
+The total disc angular momentum is
+\begin{align}
+  J &= \int_{R_\mathrm{min}}^{R_\mathrm{max}}\int_0^{2\pi} \Sigma Rv_\phi RdRd\phi \notag\\
+  &= 2\pi\int_{R_\mathrm{min}}^{R_\mathrm{max}} R\Sigma_0 v_{\phi0} R dR \notag\\ 
+  &\phantom{=}+
+  \sum_{m=1}^\infty2\pi\int_{R_\mathrm{min}}^{R_\mathrm{max}} 2R\real\left[\Sigma_m v_{\phi
+      m}^*\right] RdR 
+  %&= \sum_{m=0}^\infty \int _{R_\mathrm{min}}^{R_\mathrm{max}}  j_m(R)
+  %dR 
+  = \sum_{m=0}^\infty J_m. 
+\end{align}
+We will refer to $J_m$ as the
+$m^\mathrm{th}$ component of the total angular momentum, and use it to
+monitor numerical angular momentum conservation in the simulations. 
+It is important to distinguish this empirical definition from the
+angular momentum of linear perturbations given in \S\ref{wkb}, which
+is defined through a conservation law.
+
+%and we idenfity $J_m$ as the  
+%angular momentum associated with disc structure with
+%$m^\mathrm{th}$-fold symmetry, and $j_m$ the associated angular
+%momentum per unit length. We will use $J_m$ to monitor angular momentum 
+%conservation for the disc as a whole.
+
+\subsubsection{Three-dimensionality}
+In 3D simulations we measure the importance of vertical motion with
+$\Theta$, where 
+\begin{align}\label{theta}
+  \Theta^2 \equiv \frac{\avg{v_z^2}}{\avg{v_R^2}+\avg{v_\phi^2}}, 
+\end{align}
+and $\avg{\cdot}$ denotes the density-weighted average, e.g., 
+\begin{align}
+  \avg{v_z^2} \equiv\frac{
+    \int_{R_1}^{R_2}\int_{\theta_\mathrm{min}}^{\pi/2} \int_{0}^{2\pi}
+    \rho v_z^2 dV}{
+    \int_{R_1}^{R_2}\int_{\theta_\mathrm{min}}^{\pi/2} \int_{0}^{2\pi}
+    \rho dV
+  },
+\end{align}
+and similarly for the horizontal velocities. Thus $\Theta$ is the
+ratio of the average kinetic energy associated with vertical motion to
+that in horizontal motion. The radial range of
+integration is taken over $r\in[R_1,R_2]$ since this is where
+we find the perturbations to be confined.

### 2014-11-26 paper/results.tex
second draft

@@
 
-\subsection{Reference run with random perturbations}
+\subsection{Reference run}
 To obtain a picture of the overall disc evolution, we describe a 
@@
 
-\subsection{Trapping the low-frequency $m=1$ spiral}\label{fargo_m1}
-In order to focus on the $m=1$ spiral that eventually emerges, from
-now on we examine simulations initialised 
-$M=1$ in Eq. \ref{randpert}. The disc 
-evolution is largely similar to the previous case, but the one-armed spiral is
-more coherent, which facilitates the analysis below.
+\subsection{Properties of the $m=1$ spiral and its growth}\label{fargo_m1}
+In order to characterise the $m=1$ spiral that eventually
+dominates, here we examine a simulation initialised with 
+$M=1$ in Eq. \ref{randpert}. The one-armed spiral that emerges is more
+coherent than the above simulation, which facilitates the analysis
+below.
 
-Fig. \ref{2d_fargo_viz} shows the $m=1$ surface density
-at the end of this simulation. For this one-armed spiral
-we measure a co-rotation radius and growth rate  
+Fig. \ref{2d_fargo_viz} shows a snapshot of the $m=1$ surface
+density. By measuring the $m=1$ surface density amplitude and its
+pattern speed, we obtain a co-rotation radius and growth rate 
 \begin{align*}
@@
 \end{align*}
-The spiral can be considered as low frequency because its 
+This one-armed spiral can be considered as low frequency because its 
 pattern speed $\Omega_p \simeq 0.1\Omega_k(R_0)\lesssim 0.3\Omega$ in
@@
 $|\Sigma_1|$ varies slowly compared to the complex phase. This is the
-main assumption in local theory. We measured $k$ numerically and plot
+main assumption in local theory. We calculate $k$ numerically and plot
 its normalised value in Fig. \ref{fargo_wavenumber}. We find 
@@
 
+%snapshot shows it's tightly wound
 \begin{figure}
@@
 Fig. \ref{fargo_qbarrier} shows two $Q$-barriers located in
-the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$; the bounded region is indeed the
+the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$; the bounded region is indeed 
 where the $m=1$ spiral develops. This suggests that the one-armed 
@@
 
-\subsection{Angular momentum exchange with the background disc}  
+\subsubsection{Angular momentum exchange with the background disc}  
 The wavenumber $k = \pi G\Sigma/c_s^2$ minimises the shifted frequency
 $\sbar$ in the dispersion relation, Eq. \ref{dispersion}.  
-%\emph{axisymmetric} perturbations. 
-Physically, then, it is not
-surprising to find perturbations of this wavenumber in a
+Physically, then, it is not surprising to 
+find perturbations of this wavenumber in a
 self-gravitating disc where $Q\sim 1$. However, according the
 local dispersion relation, $m=1$ perturbations are
-formally stable for real $k$.
+formally stable for real $k$ and $Q>1$.
+
+%most of the perturbation is inside co-rotation 
+%so resonant interaction unlikely for instability
 
@@
 and $\Delta\Sigma_1\sim 0.4$ by $t=120P_0$) and linear theory becomes
-less valid. Fig. \ref{fargo_angmom_ex} confirms that the $m=1$ spiral
+less applicable. Fig. \ref{fargo_angmom_ex} confirms that the $m=1$ spiral
 wave experiences a negative torque that further reduces its angular
@@
 \subsection{Dependence on the imposed temperature profile}
-We now hypothesize the growth of the $m=1$ spiral mode is
-associated with the imposed temperature gradient. The torque
-responsible for its growth, Eq. \ref{baroclinic_torque}, is
-proportional to $q$, the power-law index used to set the sound-speed
-profile (Eq. \ref{sound-speed}). To test this, we carried out
-simulations with $q\in[0,1]$ using only an $m=1$ initial 
-perturbation. However, to keep the Toomre $Q$ profiles similar, we adjust
-the surface density power-law index such that $s = (3+q)/2$.
+The simulations above suggest the growth of the $m=1$ spiral mode
+can be attributed to the background torque applied on linear pertubrations
+as a result of an imposed temperature gradient, as discussed in
+\S\ref{global_cons}. According to Eq. \ref{baroclinic_torque}, this
+torque is proportional to $q$, the power-law index used to set the
+sound-speed profile (Eq. \ref{sound-speed}) in our simulations. 
+We confirm this through a series of simulations with $q\in[0,1]$, 
+initialised with $m=1$ perturbations. However, to keep the Toomre $Q$
+profiles similar, we adjust the surface density power-law index such
+that $s = (3+q)/2$.
 
@@
 inspection of the $m=1$ surface density field. The growth in this case
-may be associated with higher $m$ modes, which dominated the
+may be associated with high-$m$ modes, which dominated the
 simulation.
+
+% difficult to measure  
 % show result of isothermal calc?
@@
 \end{figure}
-
 We plot growth rates of the $m=1$ spiral as a function of $q$ in 
@@
 % difficult to measure very small growth rates
-This confirms that the temperature gradient is responsible for the
+This shows that the temperature gradient is responsible for the
 development of the one-armed spirals observed in our simulations.
+
 \begin{figure}
@@
 with $h$, roughly as  
-\begin{align}
+\begin{align*}
   \gamma \simeq \left[0.10h + 8.3\times10^{-3}\right]\Omega_k(R_0).  
-\end{align}
+\end{align*}
 However, a linear fit is less good than for variable $q$ cases above. This

### 2014-11-26 paper/results_3d.tex
second draft

@@
 
+%more global than 2D?
+
 Visualisations of the 3D simulations are shown in   
@@
 boundary. The snapshots are chosen when the one-armed spirals in the two codes
-have reached comparable amplitudes. The agreement between the codes,
-as well as with the 2D simulations (Fig. \ref{2d_fargo_viz}), is
-satisfactory. The largest spiral amplitude is found in the
+have reached comparable amplitudes. %The agreement between the codes,
+%as well as with the 2D simulations (Fig. \ref{2d_fargo_viz}), is
+%satisfactory. 
+Both codes show similar one-armed patterns at either height, and the
+midplane snapshot is similar to the 2D simulation
+(Fig. \ref{fargo_2d}). 
+The largest spiral amplitude is found in the
 self-gravitating region $R\in[R_1,R_2]$, independent of
@@
 the disturbance becomes more global away from the midplane.  
-
+%2d only adequate for self-gravitaing region 
 \begin{figure}
@@
 we measure the three-dimensionality parameter $\Theta < 10^{-2}$
-(Eq. \ref{theta}), so vertical motion is insignificant compared to
-horizontal motion. This supports a 2D approximation. On the other
-hand, we find  $|v_z/c_s|\lesssim 0.2$ which, although sub-sonic, is
-not very small.
+(Eq. \ref{theta}), so vertical motions are insignificant compared to
+horizontal motions. This supports a 2D approximation. On the other
+hand, we find  $\mathrm{max}|v_z/c_s|\sim 0.2$ which, although
+sub-sonic, is not very small.  
+%may increase with height
 

### 2014-11-26 paper/appendix.tex
final draft

@@
 azimuthal dependence in the form $\exp\left(\ii m \phi\right)$, and  
-the $\delta$ quantities denote Eulerian perturbations.
+the $\delta$ quantities denote Eulerian perturbations. % Note that we
+% have written the perturbed pressure force such that the state
+%the associated 
+%angular momentum flux (see below) has the same form as that in 
+%barotropic or adiabatic discs \citep{lin93b}.
 
@@
 So far we have not invoked an energy equation. For 
-adiabatic perturbations $T_\mathrm{BG}$ is zero
-\citep{lin93b}. However, if we impose the equilibrium relation
+adiabatic perturbations $T_\mathrm{BG}$ is zero, and we recover the
+same statement of angular momentum conservation as in \cite{lin93b} but 
+modified by self-gravity in the fluxes.
+
+However, if we impose the equilibrium relation
 Eq. \ref{iso_cond} to hold in the perturbed state, then
@@
 where the dispersion relation Eq. \ref{dispersion} was used. 
-In the local approximation, $|kR|\gg m$ by assumption.%  Furthermore,  
+In the local approximation, $|kR|\gg m$ by assumption. %  Furthermore,  
 % Also, for low frequency $m=1$ modes $\sbar \simeq -\Omega$, so the

### 2014-11-26 paper/discussion.tex
final draft

@@
 Thus, low-frequency $m=1$ modes generally have negative
-angular momentum. If the mode loses (positive) angular momentum 
+angular momentum. These are the modes observed in our simulations. 
+If the mode loses (positive) angular momentum 
 to the background, then we can expect instability. We demonstrate
@@
 The $m=1$ spiral that emerges in our simulations have low frequency
-($\Omega_p < \Omega$),
-grows slowly ($\gamma\ll\Omega_p$), and its spatial form is consistent local theory  
-(\S\ref{fargo_m1}). Then to a first approximation, we can regard it 
-as a neutral, low-frequency $m=1$ mode in satisfying the 
-dispersion relation Eq. \ref{dispersion} in local theory.  %attribute
-                                %growth to temp gradient 
-To demonstrate possible instability, %we 
-% first construct a neutral low-frequency mode using local theory, 
-we show that the torque density acting on such a mode due to the background
-temperature gradient is negative, which enforces the mode, because it
-has negative angular momentum.  
+($\Omega_p < \Omega$), grows slowly ($\gamma\ll\Omega_p$) and its
+spatial form is consistent local theory (\S\ref{fargo_m1}). Then to a
+first approximation, we can regard it   
+as a mode within the framework of local theory, satisfying the
+% temperature gradient ignored in local theory
+dispersion relation Eq. \ref{dispersion}. To demonstrate possible
+instability, we show that the torque density acting on such a mode due
+to the background temperature gradient is negative, which enforces the
+mode, because it has negative angular momentum.
+
 % Note that
@@
 momentum. If it is a trailing spiral and $dc_s^2/dR<0$, as is typical
-for circumstellar or protoplanetary discs, then $T_\mathrm{BG}<0$, and
+in astrophysical discs, then $T_\mathrm{BG}<0$, and
 the background disc applies a negative torque on the disturbance,
@@
 
-Using $\jlin$ and $T_\mathrm{BG}$, we can estimate 
-the growth rate $\gamma$ of linear perturbations due to the background
+Using $\jlin$ and $T_\mathrm{BG}$, we can estimate the 
+growth rate $\gamma$ of linear perturbations due to the background
 torque as  
@@
 expressions for $\jlin$ and $T_\mathrm{BG}$ for $m=1$ gives
-\begin{align}
+\begin{align}\label{theoretical_rate0}
   2\gamma \sim
@@
 (Fig. \ref{fargo_varq_growth}---\ref{fargo_varh_growth}).     
+%the latter assumes  k=pi\Sigma G /c_s^2 is always the most favoured
 
-%Note that changing $q$ affects the magntiude of the background
-%torque through its explicit depdence on the temperature gradient. By
+% The temperature gradient affects the growth rate through the
+% explicit depdence of the background torque on $q$. The growth 
+% rate is also dependent on the underlying disturbance being
+% considered (i.e. through $k$ and $\bm{\xi}$ in
+% Eq. \ref{theoretical_rate0}). % Of course, the temperature gradient
+% % plays no role on local modes. However, other disc parameters such as
+% % $c_s^2$ may affect the mode that develops. 
+% The dependence on $h$ in 
+% Eq. \ref{theoretical_rate} assumes $k=\pi\Sigma G/c_s^2$ is always the
+% most favourable wavenumber for low-frequency modes.
+
+%By 
 %definition, the global temperature gradient does not affect modes in
@@
 rises rapidly towards the gap centre since it is a region of low
-surface density.  %multi spirals associated with one planet
+surface density. The outer gap edge is then a potential site for the
+growth of a low-frequency one-armed spiral through the background
+torque.
 
@@
 If we associate the external potential with angular momentum 
-$J_\mathrm{ext}  = M_\mathrm{ring}\overline{R}^2\Omega_p$, we can obtain a
-rate $\gamma_L=\Gamma_L/J_\mathrm{ext}$. Then 
+$J_\mathrm{ext}  = M_\mathrm{ring}\overline{R}^2\Omega_p$, we can
+calculate a growth rate $\gamma_L=\Gamma_L/J_\mathrm{ext}$. Then 
 \begin{align}
@@
 ($\gamma\sim 0.1\Omega_p$). We conclude that for our disc models, the
-Lindblad and co-rotation resonances have negligible effect on the
-growth of the spiral arm in the inner disc. We confirmed this with
+Lindblad and co-rotation resonances have negligible effects on the
+growth of the one-armed spiral in the inner disc. We confirmed this with
 additional FARGO simulations with a reduced radial domain size,
@@
 angular momentum by launching density waves with positive angular momentum at
-the OLR, and by applying a positive co-rotation torque on the disc. In
-principle, this interaction is destabilising because the one-armed
-spiral has negative angular momentum \citep{lin11b}. As estimated
-above, this effect is small in our simulations, but could be
-important for other parameter regimes.
+the OLR, and by applying a positive co-rotation torque on the disc. 
+In principle, this interaction is destabilising because the
+one-armed spiral has negative angular momentum
+\citep{lin11b}. This effect is small in our simulations, but it  
+could be significant in other parameter regimes. %Furthermore, this
+%non-local interaction may be important for the long-term evolution of
+%the outer disc, since it gains angular momentum from the
 

### 2014-11-26 paper/model.tex
final draft

@@
 In this paper we will mainly use non-linear hydrodynamic simulations to investigate the
-stability of disc equilibra described 
+stability of disc equilibria described 
 above. However, it is useful to  interpret simulation results
@@
 
-The important consequence of the background torque is the possibilty
+The important consequence of the background torque is the possibility
 of instability if $\tbg\jlin>0$. That is, if $\jlin$ is positive
@@
 
-\subsubsection{Evolution of non-axisymmetirc modes}
+\subsubsection{Evolution of non-axisymmetric modes}
 The disc evolution is quantified using mode amplitudes and angular

### 2014-11-26 paper/paper.tex
final draft

@@
 An exciting development in the study of circumstellar 
-discs is the direct observation of large-scale, non-axisymmmetric
+discs is the direct observation of large-scale, non-axisymmetric
 structures within them. These
@@
 governing equations and describe the disc equilibria under
-consideration. \S\ref{methods} describes the hydrodynamic codes
-employed in this work and several diagnostic measures. We briefly
-review basic results from linear theory in \S\ref{wkb} to aid our
-interpretation of numerical simulations.  Our main results are presented in 
+consideration. We briefly review basic results from linear theory in
+\S\ref{wkb} to aid our interpretation of numerical simulations. 
+\S\ref{methods} describes the hydrodynamic codes
+employed in this work and several diagnostic measures. Our main results are presented in 
 \S\ref{results2d} for 2D simulations and \S\ref{results3d} for 3D

### 2014-11-26 paper/results.tex
final draft

@@
 The simulations above suggest the growth of the $m=1$ spiral mode
-can be attributed to the background torque applied on linear pertubrations
+can be attributed to the background torque applied on linear perturbations
 as a result of an imposed temperature gradient, as discussed in

### 2014-11-26 paper/summary.tex
final draft

@@
 We find such systems can be unstable to low-frequency perturbations
-with azimuthal wavenumber $m=1$, which lead to the development of one-armed 
-trailing spirals that persist for at least $O(10^2)$ orbits. The 
-spiral pattern speed is smaller than the local disc rotation, and
+with azimuthal wavenumber $m=1$, which leads to the development of an one-armed 
+trailing spiral that persist for at least $O(10^2)$ orbits. The spiral
+pattern speed is smaller than the local disc rotation and
 growth rates are $O(10^{-2}\Omega)$ which gives a characteristic
@@
 We used three independent numerical codes --- FARGO in 2D, ZEUS-MP and
-PLUTO in 3D --- to verify the growth of  
+PLUTO in 3D --- to show that the growth of  
 one-armed spirals in our disc models is due to the imposed temperature 
-profile and disc structure. The spiral is confined between two
-$Q$-barriers owing to a surface density bump. Growth rates increased
+profile and disc structure. Growth rates increased 
 with the magnitude of the imposed temperature gradient, and one-armed
-spirals did not develop in strictly isothermal simulations.  
+spirals did not develop in strictly isothermal simulations. The spiral
+is confined between two $Q$-barriers owing to a surface density bump. 
 We find the instability behaves similarly in 2D and 3D. However, in 3D
@@
 maintained, it may be possible to achieve a balance between the gain
-of negative angular momentum through the temperature gradient, and the
+of negative angular momentum through the background torque, and the
 gain of positive angular momentum through shock dissipation. We remark  
@@
 %explicitly solve for linear stability problem.
+
+%theoretical investigation on effect on other m modes

### 2015-01-01 paper/discussion.tex
major restructuring and rewrite in response to report

@@
 
-\subsection{Angular momentum density of  low-frequency $m=1$ modes}
-From Eq. \ref{lin_ang_mom_cons} and assuming a time-dependence of the
-form $\exp{(-\ii \sigma t)}$ with $\real{\sigma} = \omega$,  
-the angular momentum density associated with linear waves is
-\begin{align}
-  \jlin = \frac{m\Sigma}{2}\left[\left(\omega -
-      m\Omega\right)|\bm{\xi}|^2 + \ii\Omega\left(\xi_R\xi_\phi^* -
-      \xi_R^*\xi_\phi\right)\right].  
-\end{align}
-For a low-frequency mode, $|\omega|\ll m\Omega$. Then neglecting the
-term $\omega|\bm{\xi}|^2$ and setting $m=1$,
-\begin{align}
-  \jlin &\simeq \frac{\Sigma\Omega}{2}\left[-|\bm{\xi}|^2 + \ii\left(\xi_R\xi_\phi^* -
-      \xi_R^*\xi_\phi\right)\right]\notag\\
-  & = -\frac{\Sigma\Omega}{2}|\xi_R + \ii\xi_\phi|^2\notag\\
-  & \leq 0.
-\end{align}
-Thus, low-frequency $m=1$ modes generally have negative
-angular momentum. These are the modes observed in our simulations. 
-If the mode loses (positive) angular momentum 
-to the background, then we can expect instability. We demonstrate
-below how this is possible through a forced temperature
-gradient.
-
-\subsection{Unstable interaction between low-frequency $m=1$ spirals 
-  and the background disc due to an imposed temperature gradient}
-The $m=1$ spiral that emerges in our simulations have low frequency
-($\Omega_p < \Omega$), grows slowly ($\gamma\ll\Omega_p$) and its
-spatial form is consistent local theory (\S\ref{fargo_m1}). Then to a
-first approximation, we can regard it   
-as a mode within the framework of local theory, satisfying the
-% temperature gradient ignored in local theory
-dispersion relation Eq. \ref{dispersion}. To demonstrate possible
-instability, we show that the torque density acting on such a mode due
-to the background temperature gradient is negative, which enforces the
-mode, because it has negative angular momentum.
-
-% Note that
-% we will not be assuming a self-gravitating disc until the end of this
-% section.
-
-We begin by recalling the expression for the torque density due to a
-locally isothermal equation of state,
-\begin{align*}
-  T_\mathrm{BG} = - \frac{m}{2}\frac{dc_s^2}{dR}\imag\left(\delta\Sigma_m\xi_R^*\right).   
-\end{align*}
-The Eulerian surface density perturbation is given by
-\begin{align}\label{den_pert}
-  \delta\Sigma_m = -\nabla\cdot\left(\Sigma\bm{\xi}\right) 
-  = -\frac{1}{R}\frac{d}{dR}\left(R\Sigma \xi_R\right) - \frac{\ii m}{R}\Sigma\xi_\phi.  
-\end{align}
-We now invoke local theory by setting $d/dR \to \ii k$ where $k$ is
-real, and assume $|kR|\gg m$ so that the second term on the right hand
-side of Eq. \ref{den_pert} can be neglected. Then    
-\begin{align}
-  \delta\Sigma _m  \simeq -\ii k \Sigma \xi_R,
-\end{align}
-which gives
-\begin{align}
-  T_\mathrm{BG} = \frac{m}{2}\frac{dc_s^2}{dR}k\Sigma |\xi_R|^2. 
-\end{align}
-% \begin{align}
-%   T_\mathrm{BG} &= -\frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\delta\Sigma_m \left(\frac{\delta
-%         v_{Rm}}{-\ii\sbar}\right)^*\right] \notag\\
-%   & = \frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\ii \delta\Sigma_m \left(\frac{\delta
-%         v_{Rm}}{\sbar}\right)^*\right]
-% \end{align}
-% where we replaced the radial Lagrangian displacement by the Eulerian
-% radial velocity perturbation in Eq. \ref{baroclinic_torque}. 
-% The linearised momentum equations give
-% \begin{align}
-%   D\delta v_{Rm} =& 
-%   \ii\sbar\left[c_s^2\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right)
-%     + \frac{d}{dR}\delta\Phi_m\right] \notag\\ 
-%   &- \frac{2\ii
-%     m\Omega}{R}\left(c_s^2\frac{\delta\Sigma_m}{\Sigma} +
-%     \delta\Phi_m\right),
-% \end{align}
-% where $D\equiv \kappa^2 - \sbar^2$. 
-% We now invoke local theory by setting $d/dR \to \ii k$ where $k$ is
-% real, and using the local solution to the Poisson equation
-% \begin{align}
-%   \delta\Phi_m = -\frac{2\pi G}{|k|}\delta\Sigma_m 
-% \end{align}
-% \citep{shu91}. Then the radial velocity perturbation becomes
-% \begin{align}
-%   \delta v_{Rm} =
-%   \frac{1}{D}\frac{\delta\Sigma_m}{\Sigma}\left(\frac{2\pi G
-%       \Sigma}{|k|} - c_s^2\right)\left(k\sbar + \frac{2\ii
-%       m\Omega}{R}\right),
-% \end{align}
-% so that 
-% \begin{align}
-%   \ii\delta\Sigma_m \left(\frac{\delta v_{Rm}}{\sbar}\right)^* =
-%   \frac{1}{D^*}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
-%       \Sigma}{|k|} - c_s^2\right)\left(\ii k  + \frac{2
-%       m\Omega}{R\sbar^*}\right). 
-% \end{align}
-% For a neutral mode, $\sbar$, and therefore $D$ is real. Then, after
-% taking the imaginary part of the above expression, the torque
-% density is 
-% \begin{align}\label{tbg_explicit}
-%   T_\mathrm{BG} &=
-%   \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{D}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
-%       \Sigma}{|k|} - c_s^2\right)\right].
-% % &\simeq \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{\left(\kappa^2 -
-% %         m^2\Omega^2\right)}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi
-% %       G \Sigma}{|k|} - c_s^2\right)\right]. 
-% \end{align}
-% We now use the local dispersion relation,
-% Eq. \ref{dispersion}, to make the replacement $D = 2\pi G \Sigma |k| -
-% k^2c_s^2$ in the above expression. This gives
-% \begin{align}\label{tbg_simple}
-%   T_\mathrm{BG} =
-%   \frac{m}{2}\frac{dc_s^2}{dR}\frac{1}{k}\frac{|\delta\Sigma_m|^2}{\Sigma}.  
-% \end{align}  
-This torque density is negative for trailing waves in discs with
-a fixed temperature profile decreasing outwards ($k>0$ and $dc_s^2/dR<0$,
-respectively). Note that this conclusion does not rely on the
-low-frequency approximation and holds for any $m$.
-
-However, if the linear disturbance under consideration \emph{is} a
-low-frequency $m=1$ mode, then it has negative angular
-momentum. If it is a trailing spiral and $dc_s^2/dR<0$, as is typical
-in astrophysical discs, then $T_\mathrm{BG}<0$, and
-the background disc applies a negative torque on the disturbance,
-which further decreases its angular momentum. This suggests the spiral
-amplitude will grow.
-
-Using $\jlin$ and $T_\mathrm{BG}$, we can estimate the 
-growth rate $\gamma$ of linear perturbations due to the background
-torque as  
-\begin{align}
-  2\gamma \sim \frac{T_\mathrm{BG}}{\jlin},
-\end{align}
-where the factor of two accounts for the fact that the angular momentum
-density is quadratic in the linear perturbations. Inserting the above
-expressions for $\jlin$ and $T_\mathrm{BG}$ for $m=1$ gives
-\begin{align}\label{theoretical_rate0}
-  2\gamma \sim
-  -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R|^2}{|\xi_R + \ii
-    \xi_\phi|^2}. 
-%  \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
-\end{align}
-In Appendix \ref{horizontal_displacements}, we argue $\xi_\phi \simeq
-2\ii \xi_R$ for the modes of interest. Then 
-%To further simplify this expression, we need to relate $\xi_R$ and
-%$\xi_\phi$. 
-% \begin{align}
-%   2\gamma \sim
-%   -\frac{dc_s^2}{dR}\frac{1}{k\Sigma^2\Omega}\frac{|\delta\Sigma_1|^2}{|\xi_R
-%     + \ii\xi_\phi|^2}.
-% \end{align}
-% Now, using $\delta\Sigma_m/\Sigma = -\ii k \xi_R - \ii m \xi_\phi/R$
-% in the local approximation, we have
-% \begin{align}
-%   2\gamma \sim -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R +
-%     \xi_\phi/kR|^2}{|\xi_R + \ii \xi_\phi|^2}
-%   \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
-% \end{align}
-% where we used $dc_s^2/dR <0$, $k>0$ and $|kR|\gg 1$ to obtain the
-% inequality. 
-%where we used $dc_s^2/dR <0$ and $k>0$ to obtain the inequality. 
-for the temperature profiles $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ as
-adopted in our disc models,  
-\begin{align}
-  2\gamma \sim q\frac{c_s^2}{R}\frac{k}{\Omega}. 
-\end{align}
-For perturbations in a self-gravitating disc with characteristic
-wavenumber $k \sim \pi G \Sigma/c_s^2 \sim \Omega/Q c_s$, as seen in
-our simulations, we have 
-\begin{align}\label{theoretical_rate}
-  \gamma \sim \frac{qh}{2Q}\Omega,
-\end{align}
-where we used $c_s/R\sim h\Omega$. For our fiducial disc models with
-$q=1$, $Q=O(1)$, $h=0.05$, this gives $\gamma/\Omega \sim O(10^{-2})$,
-consistent with simulation results. Specifically,
-Eq. \ref{theoretical_rate} indicates the growth rate should increase
-linearly with $q$ and $h$, as seen in the simulations
-(Fig. \ref{fargo_varq_growth}---\ref{fargo_varh_growth}).     
 %the latter assumes  k=pi\Sigma G /c_s^2 is always the most favoured

### 2015-01-01 paper/model.tex
major restructuring and rewrite in response to report

@@
-\section{Governing equations and numerical models}\label{model} 
-We consider a self-gravitating, inviscid fluid disc of mass $M_d$ 
+\section{Governing equations}\label{model} 
+We consider an inviscid fluid disc of mass $M_d$ 
 orbiting a central star of mass $M_*$. We will mainly examine 2D
@@
 delta function.
+
+\section{Linear density waves}\label{wkb}
+% In this paper we will mainly use non-linear hydrodynamic simulations to investigate the
+% stability of disc equilibria described 
+% above. However, it is useful to  interpret simulation results
+% using linear density theory. We list here the main definitions and
+% results from 2D linear theory, taken from \cite{lin11b} and
+% \cite{shu91}.  
+{\bf We highlight a key feature of locally isothermal discs that
+  enables angular momentum exchange between small disturbances and the
+  background disc through an imposed radial temperature gradient. This
+  conclusion results from the consideration of angular momentum
+  conservation within the framework of linear perturbation theory. 
+  For simplicity, we consider 2D discs, although generalisation to 3D is straight
+  forward.  
+}
+
+In a linear analysis, one assumes a steady axisymmetric background state, % such as
+% that defined in \S\ref{model},
+which is then perturbed such that
+\begin{align}  
+  \Sigma \to \Sigma(R) + \delta\Sigma_m(R)\exp{\left[\ii\left(-\sigma t +
+        m\phi\right)\right]}, 
+\end{align}
+and similarly for other variables, where $\sigma=\omega+\ii\gamma$ is
+a complex frequency with $\omega$ being the real frequency, 
+$\gamma$ the growth rate, and $m$ is an integer. Here, we take $m>0$ without
+loss of generality.  
+%Since the background is
+%axisymmetric, we identify $\delta\Sigma_m(R)\exp{\left(-\ii\sigma
+ %   t\right)} = \Sigma_m$ as defined for diagnostics.
+
+The linearised mass and momentum equations are
+\begin{align}
+  &-\ii\sbar \delta\Sigma_m = -\frac{1}{R}\frac{d}{dR}\left(R\Sigma\delta
+    v_{Rm}\right) - \frac{\ii m \Sigma}{R}\delta v_{\phi m}, \\
+  &-\ii\sbar\delta v_{Rm} - 2\Omega \delta v_{\phi m} = -
+  c_s^2(R)\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right) - \frac{d}{dR}\delta\Phi_m,\label{radmom}\\
+  & -\ii\sbar\delta v_{\phi m} + \frac{\kappa^2}{2\Omega}\delta v_{Rm} =
+  -\frac{\ii m }{R}\left(c_s^2\frac{\delta\Sigma_m}{\Sigma} + \delta\Phi_m\right),
+\end{align}
+where $\sbar = \sigma - m\Omega$ and $\kappa^2 =
+R^{-3}\p_R(R^4\Omega^2)$ is the square of the epicyclic frequency. A
+locally isothermal equation of state has been assumed in
+Eq. \ref{radmom}.  The linearised Poisson equation is 
+\begin{align}
+  \nabla^2\delta\Phi_m = 4\pi G \delta\Sigma_m \delta(z). 
+\end{align}
+
+These linearised equations can be combined into a single
+integro-differential equation eigenvalue problem. We defer a full
+{\bf numerical} exploration of the linear problem to a future study. Here, we {\bf
+  discuss} some
+general properties of the linear perturbations. % that will be relevant
+% for describing simulation results.
+
+\subsection{Global angular momentum conservation for linear
+  perturbations} \label{global_cons}
+It can be shown that linear perturbations with $\phi$-dependence in the form
+$\exp{(\ii m\phi)}$ satisfies angular momentum conservation
+in the form 
+\begin{align}\label{angcons}
+  \frac{\p \jlin}{\p t} + \nabla\cdot\bm{F} = \tbg, 
+\end{align}
+\citep[e.g.][]{narayan87,ryu92,lin93b} where
+\begin{align}\label{lin_ang_mom_cons}
+  \jlin \equiv
+  -\frac{m\Sigma}{2}\imag\left(\bm{\xi}^*\cdot\frac{\p\bm{\xi}}{\p
+      t} + \Omega \hat{\bm{k}}\cdot\bm{\xi}\times\bm{\xi}^* + \ii
+    m \Omega |\bm{\xi}|^2\right) 
+\end{align}
+is the angular momentum density of the linear disturbance (which may
+be positive or negative), $\bm{\xi}$ is the Lagrangian
+displacement and $^*$ denotes complex conjugation, and $\bm{F}$ is the
+vertically-integrated angular momentum flux consisting of a Reynolds
+stress and a gravitational stress \citep{lin11b}. Explicit expressions
+for $\bm{\xi}$ can be found in, e.g. \cite{papaloizou85}.
+
+In Eq. \ref{angcons}, the background torque density $\tbg$ is 
+\begin{align}\label{baroclinic_torque}
+  \tbg \equiv
+  -\frac{m}{2}\imag\left(\delta\Sigma_m\xi_R^*\frac{dc_s^2}{dR}\right), 
+\end{align}
+and arises because we have adopted a locally isothermal equation of
+state. {\bf  We outline the
+derivation of $T_\mathrm{BG}$ in Appendix \ref{tbg_deriv}.}
+
+In a barotropic fluid, such as a strictly isothermal disc,
+$\tbg$ vanishes and the total angular momentum associated with
+the perturbation is conserved, provided that there is no net angular
+momentum flux. However, as noted in \cite{lin11b}, if there is an imposed
+temperature gradient, as in the disc models we consider,
+then $\tbg\neq 0$ in general, which corresponds to a local torque
+exerted by the background disc on the perturbation.
+
+The important consequence of the background torque is the possibility
+of instability if $\tbg\jlin>0$. That is, if $\jlin$ is positive
+(negative) and the background torque $\tbg$ is {\bf also} positive (negative),
+then the local angular momentum density of the linear disturbance
+{\bf will} further increase (decrease) {\bf with time}. This implies the amplitude of the disturbance
+may grow by exchanging angular momentum with the background
+disc.%  In our disc models with $c_s^2\propto R^{-q}$, we expect the
+% growth rate of such an instability to scale with $q$.
+
+{\bf 
+  It is simplest to demonstrate instability by determining  
+  the sign of the angular momentum density 
+  $\jlin$ and the background torque $T_\mathrm{BG}$ for perturbations
+  in the local approximation, which we review next. 
+  % depends on the
+  % type of perturbation being considered. 
+  We then show that low-frequency perturbations ($|\omega|\ll\Omega$)
+  are unstable when there a negative radial 
+  temperature gradient is imposed.  
+}
+
+\subsection{Angular momentum density of  non-axisymmetirc low-frequency modes}
+{\bf We now specialise to low-frequency modes and first determine the sign of its
+  angular momentum density.} From Eq. \ref{lin_ang_mom_cons} and assuming a time-dependence of the
+form $\exp{(-\ii \sigma t)}$ with $\real{\sigma} = \omega$,  
+the angular momentum density associated with linear waves is
+\begin{align}
+  \jlin = \frac{m\Sigma}{2}\left[\left(\omega -
+      m\Omega\right)|\bm{\xi}|^2 + \ii\Omega\left(\xi_R\xi_\phi^* -
+      \xi_R^*\xi_\phi\right)\right].  
+\end{align}
+For a low-frequency mode, $|\omega|\ll m\Omega$. Then neglecting the
+term $\omega|\bm{\xi}|^2$, {\bf we find}
+\begin{align}
+  \jlin &\simeq \frac{m\Sigma\Omega}{2}\left[-m|\bm{\xi}|^2 + \ii\left(\xi_R\xi_\phi^* -
+      \xi_R^*\xi_\phi\right)\right]\notag\\
+  & = -\frac{m\Sigma\Omega}{2}\left[ (m-1)|\bm{\xi}|^2 +  |\xi_R + \ii\xi_\phi|^2\right].
+%  & \leq 0.
+\end{align}
+
+Thus, low-frequency non-axisymmetric modes ($m\geq1$)
+
+Thus, low-frequency $m=1$ modes generally have negative
+angular momentum. These are the modes observed in our simulations. 
+If the mode loses (positive) angular momentum 
+to the background, then we can expect instability. We demonstrate
+below how this is possible through a forced temperature
+gradient.
+
+%and setting $m=1$,
+%\begin{align}
+ % \jlin &\simeq \frac{\Sigma\Omega}{2}\left[-|\bm{\xi}|^2 + \ii\left(\xi_R\xi_\phi^* -
+  %    \xi_R^*\xi_\phi\right)\right]\notag\\
+%  & = -\frac{\Sigma\Omega}{2}|\xi_R + \ii\xi_\phi|^2\notag\\
+ % & \leq 0.
+%\end{align}
+
+\subsection{Local results}
+In the local approximation, perturbations are assumed to vary rapidly
+relative to any background gradients. The
+dispersion relation for tightly-wound density 
+waves of the form $\exp{[\ii(-\sigma t + m \phi + kR)]}$ in a razor-thin
+disc is  
+\begin{align}\label{dispersion}
+  (\sigma - m\Omega)^2 = \kappa^2 + k^2c_s^2 - 2\pi G \Sigma |k|, 
+\end{align}
+where $k$ is a real wavenumber such that $|kR|\gg1$ \citep{shu91}. 
+Note that in the local approximation, only axisymmetric perturbations
+($m=0$) can be unstable. % (having an imaginary $\sigma$).
+
+Given the real frequency $\omega$ or pattern speed $\Omega_p\equiv  
+\omega/m$ of a neutral mode 
+, Eq. \ref{dispersion} can be solved 
+for $|k|$, 
+\begin{align}\label{wavenumber}
+%  |k| = \frac{\pi G \Sigma}{c_s^2}\left[1 \pm \sqrt{1 -
+ %     Q^2(1-\nu^2)}\right], 
+|k| = k_c\left[1 \pm \sqrt{1 -
+     Q^2(1-\nu^2)}\right], 
+\end{align}
+where 
+ \begin{align}
+   k_c \equiv \frac{\pi G \Sigma}{c_s^2}
+ \end{align}
+ is a characteristic wavenumber, 
+\begin{align}
+  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}
+\end{align}
+is the usual Toomre parameter, and
+\begin{align}
+  \nu \equiv \frac{(\omega - m\Omega)}{\kappa}
+\end{align}
+is a dimensionless frequency. In
+Eq. \ref{wavenumber}, the upper (lower) sign correspond to short
+(long) waves, and $k>0$ ($k<0$) correspond to trailing (leading)
+spirals.
+
+The \emph{co-rotation radius} $R_c$ is where the pattern speed matches
+the fluid rotation,
+\begin{align}
+  \Omega(R_c) = \Omega_p.
+\end{align}
+\emph{Lindblad resonances} $R_L$ occurs where
+\begin{align}
+  \nu^2(R_L) = 1. 
+\end{align}
+Finally, \emph{Q-barriers} occur at radii $R_{Qb}$ where
+\begin{align}
+  Q^2(R_{Qb})\left[1-\nu^2(R_{Qb})\right] = 1.  
+\end{align}
+According to Eq. \ref{wavenumber}, purely wave-like solutions with
+real $k$ are only possible where $Q^2(1-\nu^2)\leq1$.
+
+A detailed discussion of the properties of local density waves 
+is given in \cite{shu91}. An important result is that
+waves interior to co-rotation ($R<R_c$) have negative angular momentum, while
+waves outside co-rotation ($R>R_c$) have positive angular
+momentum. The local approximation becomes less good for global modes,
+but will nevertheless be useful for results interpretation.
+
+\subsection{Unstable interaction between low-frequency modes 
+  and the background disc due to an imposed temperature gradient}
+%The $m=1$ spiral that emerges in our simulations have low frequency
+%($\Omega_p < \Omega$), grows slowly ($\gamma\ll\Omega_p$) and its
+%spatial form is consistent local theory (\S\ref{fargo_m1}). 
+%{\bf
+%  We first regard a low-frequency, $m=1$ disturbance 
+ % as a mode within the framework of local theory, satisfying the
+ % % temperature gradient ignored in local theory
+ % dispersion relation Eq. \ref{dispersion}. }
+%To demonstrate possible
+%instability, 
+{\bf 
+Here we show that the torque density acting on a low-frequency, $m=1$
+local mode due to the background temperature gradient can be negative, which 
+would enforce the mode, because it has negative angular momentum.  
+}
+
+% Note that
+% we will not be assuming a self-gravitating disc until the end of this
+% section. 
+%We begin by recalling the expression for the torque density due to a
+%locally isothermal equation of state,
+%\begin{align*}
+ % T_\mathrm{BG} = - \frac{m}{2}\frac{dc_s^2}{dR}\imag\left(\delta\Sigma_m\xi_R^*\right).   
+%\end{align*}
+The Eulerian surface density perturbation is given by
+\begin{align}\label{den_pert}
+  \delta\Sigma_m = -\nabla\cdot\left(\Sigma\bm{\xi}\right) 
+  = -\frac{1}{R}\frac{d}{dR}\left(R\Sigma \xi_R\right) - \frac{\ii m}{R}\Sigma\xi_\phi.  
+\end{align}
+We invoke local theory by setting $d/dR \to \ii k$ where $k$ is
+real, and assume $|kR|\gg m$ so that the second term on the right hand
+side of Eq. \ref{den_pert} can be neglected. Then    
+\begin{align}
+  \delta\Sigma _m  \simeq -\ii k \Sigma \xi_R.
+\end{align}
+{\bf Inserting this into the expression for the background torque,
+  Eq. \ref{baroclinic_torque}, we find}
+\begin{align}
+  T_\mathrm{BG} = \frac{m}{2}\frac{dc_s^2}{dR}k\Sigma |\xi_R|^2. 
+\end{align}
+
+This torque density is negative for trailing waves in discs with
+a fixed temperature profile decreasing outwards ($k>0$ and $dc_s^2/dR<0$,
+respectively). Note that this conclusion does not rely on the
+low-frequency approximation and holds for any $m$.
+
+However, if the linear disturbance under consideration \emph{is} a
+low-frequency $m=1$ mode, then it has negative angular
+momentum. {\bf Then} if it is a trailing spiral and $dc_s^2/dR<0$, as is typical
+in astrophysical discs, then $T_\mathrm{BG}<0$, and
+the background disc applies a negative torque on the disturbance,
+which further decreases its angular momentum. This suggests the {\bf
+  mode} 
+amplitude will grow.
+
+Using $\jlin$ and $T_\mathrm{BG}$, we can estimate the 
+growth rate $\gamma$ of linear perturbations due to the background
+torque as  
+\begin{align}
+  2\gamma \sim \frac{T_\mathrm{BG}}{\jlin},
+\end{align}
+where the factor of two accounts for the fact that the angular momentum
+density is quadratic in the linear perturbations. Inserting the above
+expressions for $\jlin$ and $T_\mathrm{BG}$ for $m=1$ gives
+\begin{align}\label{theoretical_rate0}
+  2\gamma \sim
+  -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R|^2}{|\xi_R + \ii
+    \xi_\phi|^2}. 
+%  \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
+\end{align}
+In Appendix \ref{horizontal_displacements}, we argue $\xi_\phi \simeq
+2\ii \xi_R$ for the modes of interest. Then 
+%To further simplify this expression, we need to relate $\xi_R$ and
+%$\xi_\phi$. 
+% \begin{align}
+%   2\gamma \sim
+%   -\frac{dc_s^2}{dR}\frac{1}{k\Sigma^2\Omega}\frac{|\delta\Sigma_1|^2}{|\xi_R
+%     + \ii\xi_\phi|^2}.
+% \end{align}
+% Now, using $\delta\Sigma_m/\Sigma = -\ii k \xi_R - \ii m \xi_\phi/R$
+% in the local approximation, we have
+% \begin{align}
+%   2\gamma \sim -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R +
+%     \xi_\phi/kR|^2}{|\xi_R + \ii \xi_\phi|^2}
+%   \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
+% \end{align}
+% where we used $dc_s^2/dR <0$, $k>0$ and $|kR|\gg 1$ to obtain the
+% inequality. 
+%where we used $dc_s^2/dR <0$ and $k>0$ to obtain the inequality. 
+for the temperature profiles $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ as
+adopted in our disc models,  
+\begin{align}
+  2\gamma \sim q\frac{c_s^2}{R}\frac{k}{\Omega} \sim q h
+  \left(kH\right)\Omega.  
+\end{align}
+
+{\bf It is important to note that this growth is
+due to the temperature gradient and does not depend explicitly on
+whether or not the disc is self-gravitating, i.e. destabilisation is
+not due to self-gravity.}
+
+For perturbations in a self-gravitating disc with characteristic
+wavenumber $k \sim \pi G \Sigma/c_s^2 \sim \Omega/Q c_s$, as seen in
+our simulations, we have 
+\begin{align}\label{theoretical_rate}
+  \gamma \sim \frac{qh}{2Q}\Omega,
+\end{align}
+where we used $c_s/R\sim h\Omega$.
+
+%For our fiducial disc models with
+%$q=1$, $Q=O(1)$, $h=0.05$, this gives $\gamma/\Omega \sim O(10^{-2})$,
+%consistent with simulation results. Specifically,
+%Eq. \ref{theoretical_rate} indicates the growth rate should increase
+%linearly with $q$ and $h$, as seen in the simulations
+%(Fig. \ref{fargo_varq_growth}---\ref{fargo_varh_growth}).
+
+\section{Numerical simulations}\label{methods}
+%above discussion is generic for low-frequency non-axisymmetric modes,
+%but illustrate with a concrete example
+We adopt computational units such that
+$G=M_*=R_0=1$. Time is measured  in the Keplerian orbital period at
+the reference radius, $P_0\equiv  2\pi/\Omega_k(R_0)$.
 
@@
 \end{figure}
-
-\section{Linear density waves}\label{wkb}
-In this paper we will mainly use non-linear hydrodynamic simulations to investigate the
-stability of disc equilibria described 
-above. However, it is useful to  interpret simulation results
-using linear density theory. We list here the main definitions and
-results from 2D linear theory, taken from \cite{lin11b} and
-\cite{shu91}.   In particular, we highlight a key feature of
-locally isothermal discs that enables angular momentum exchange
-between linear perturbations and the background disc.
-
-In a linear analysis, one assumes a steady axisymmetric background state, such as
-that defined in \S\ref{model}, which is then perturbed such that
-\begin{align}  
-  \Sigma \to \Sigma(R) + \delta\Sigma_m(R)\exp{\left[\ii\left(-\sigma t +
-        m\phi\right)\right]}, 
-\end{align}
-and similarly for other variables, where $\sigma=\omega+\ii\gamma$ is
-a complex frequency with $\omega$ being the real frequency, 
-$\gamma$ the growth rate, and $m$ is an integer. Here, we take $m>0$ without
-loss of generality.  
-%Since the background is
-%axisymmetric, we identify $\delta\Sigma_m(R)\exp{\left(-\ii\sigma
- %   t\right)} = \Sigma_m$ as defined for diagnostics.
-
-The linearised mass and momentum equations are
-\begin{align}
-  &-\ii\sbar \delta\Sigma_m = -\frac{1}{R}\frac{d}{dR}\left(R\Sigma\delta
-    v_{Rm}\right) - \frac{\ii m \Sigma}{R}\delta v_{\phi m}, \\
-  &-\ii\sbar\delta v_{Rm} - 2\Omega \delta v_{\phi m} = -
-  c_s^2(R)\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right) - \frac{d}{dR}\delta\Phi_m,\label{radmom}\\
-  & -\ii\sbar\delta v_{\phi m} + \frac{\kappa^2}{2\Omega}\delta v_{Rm} =
-  -\frac{\ii m }{R}\left(c_s^2\frac{\delta\Sigma_m}{\Sigma} + \delta\Phi_m\right),
-\end{align}
-where $\sbar = \sigma - m\Omega$ and $\kappa^2 =
-R^{-3}\p_R(R^4\Omega^2)$ is the square of the epicyclic frequency. A
-locally isothermal equation of state has been assumed in
-Eq. \ref{radmom}.  The linearised Poisson equation is 
-\begin{align}
-  \nabla^2\delta\Phi_m = 4\pi G \delta\Sigma_m \delta(z). 
-\end{align}
-
-These linearised equations can be combined into a single
-integro-differential equation eigenvalue problem. We defer a full
-exploration of the linear problem to a future study. Here, we list some
-general  properties of the linear perturbations that will be relevant
-for describing simulation results.
-
-\subsection{Global angular momentum conservation for linear
-  perturbations} \label{global_cons}
-It can be shown that linear perturbations with $\phi$-dependence in the form
-$\exp{(\ii m\phi)}$ satisfies angular momentum conservation
-in the form 
-\begin{align}\label{angcons}
-  \frac{\p \jlin}{\p t} + \nabla\cdot\bm{F} = \tbg, 
-\end{align}
-\citep[e.g.][]{narayan87,ryu92,lin93b} where
-\begin{align}\label{lin_ang_mom_cons}
-  \jlin \equiv
-  -\frac{m\Sigma}{2}\imag\left(\bm{\xi}^*\cdot\frac{\p\bm{\xi}}{\p
-      t} + \Omega \hat{\bm{k}}\cdot\bm{\xi}\times\bm{\xi}^* + \ii
-    m \Omega |\bm{\xi}|^2\right) 
-\end{align}
-is the angular momentum density of the linear disturbance (which may
-be positive or negative), where $\bm{\xi}$ is the Lagrangian
-displacement and $^*$ denotes complex conjugation, and $\bm{F}$ is the
-vertically-integrated angular momentum flux consisting of a Reynolds
-stress and a gravitational stress \citep{lin11b}. Explicit expressions
-for $\bm{\xi}$ can be found in, e.g. \cite{papaloizou85}.
-
-In Eq. \ref{angcons}, the background torque density $\tbg$ is 
-\begin{align}\label{baroclinic_torque}
-  \tbg \equiv
-  -\frac{m}{2}\imag\left(\delta\Sigma_m\xi_R^*\frac{dc_s^2}{dR}\right), 
-\end{align}
-and arises because we have adopted a locally isothermal equation of
-state. In a barotropic fluid, such as a strictly isothermal disc,
-$\tbg$ vanishes and the total angular momentum associated with
-the perturbation is conserved, provided that there is no net angular
-momentum flux. % and the perturbations are regular.
-
-However, as noted in \cite{lin11b}, if there is an imposed
-temperature gradient, as in the disc models we consider,
-then $\tbg\neq 0$ in general, which corresponds to a local torque
-exerted by the background disc on the perturbation. We outline the
-derivation of $T_\mathrm{BG}$ in Appendix \ref{tbg_deriv}.
-
-The important consequence of the background torque is the possibility
-of instability if $\tbg\jlin>0$. That is, if $\jlin$ is positive
-(negative) and the background torque $\tbg$ is positive (negative),
-then the local angular momentum density of the linear disturbance
-further increases (decreases). This implies the amplitude of the disturbance
-may grow by exchanging angular momentum with the background
-disc. In our disc models with $c_s^2\propto R^{-q}$, we expect the
-growth rate of such an instability to scale with $q$.
-
-\subsection{Local results}
-In the local approximation, perturbations are assumed to vary rapidly
-relative to any background gradients. The
-dispersion relation for tightly-wound density 
-waves of the form $\exp{[\ii(-\sigma t + m \phi + kR)]}$ in a razor-thin
-disc is  
-\begin{align}\label{dispersion}
-  (\sigma - m\Omega)^2 = \kappa^2 + k^2c_s^2 - 2\pi G \Sigma |k|, 
-\end{align}
-where $k$ is a real wavenumber such that $|kR|\gg1$ \citep{shu91}. 
-Note that in the local approximation, only axisymmetric perturbations
-($m=0$) can be unstable. % (having an imaginary $\sigma$).
-
-Given the real frequency $\omega$ or pattern speed $\Omega_p\equiv  
-\omega/m$ of a neutral mode 
-, Eq. \ref{dispersion} can be solved 
-for $|k|$, 
-\begin{align}\label{wavenumber}
-  |k| = \frac{\pi G \Sigma}{c_s^2}\left[1 \pm \sqrt{1 -
-      Q^2(1-\nu^2)}\right], 
-\end{align}
-where 
-% \begin{align}
-%   k_T \equiv \frac{\kappa^2}{2\pi G \Sigma}
-% \end{align}
-% is a characteristic wavenumber, 
-\begin{align}
-  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}
-\end{align}
-is the usual Toomre parameter, and
-\begin{align}
-  \nu \equiv \frac{(\omega - m\Omega)}{\kappa}
-\end{align}
-is a dimensionless frequency. In
-Eq. \ref{wavenumber}, the upper (lower) sign correspond to short
-(long) waves, and $k>0$ ($k<0$) correspond to trailing (leading)
-spirals.
-
-The \emph{co-rotation radius} $R_c$ is where the pattern speed matches
-the fluid rotation,
-\begin{align}
-  \Omega(R_c) = \Omega_p.
-\end{align}
-\emph{Lindblad resonances} $R_L$ occurs where
-\begin{align}
-  \nu^2(R_L) = 1. 
-\end{align}
-Finally, \emph{Q-barriers} occur at radii $R_{Qb}$ where
-\begin{align}
-  Q^2(R_{Qb})\left[1-\nu^2(R_{Qb})\right] = 1.  
-\end{align}
-According to Eq. \ref{wavenumber}, purely wave-like solutions with
-real $k$ are only possible where $Q^2(1-\nu^2)\leq1$.
-
-A detailed discussion of the properties of local density waves 
-is given in \cite{shu91}. An important result is that
-waves interior to co-rotation ($R<R_c$) have negative angular momentum, while
-waves outside co-rotation ($R>R_c$) have positive angular
-momentum. The local approximation becomes less good for global modes,
-but will nevertheless be useful for results interpretation.
-
-\section{Numerical methods}\label{methods}
-We adopt computational units such that
-$G=M_*=R_0=1$. Time is measured  in the Keplerian orbital period at
-the reference radius, $P_0\equiv  2\pi/\Omega_k(R_0)$.
 
@@
 we find the perturbations to be confined.
+
+% \begin{align}
+%   T_\mathrm{BG} &= -\frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\delta\Sigma_m \left(\frac{\delta
+%         v_{Rm}}{-\ii\sbar}\right)^*\right] \notag\\
+%   & = \frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\ii \delta\Sigma_m \left(\frac{\delta
+%         v_{Rm}}{\sbar}\right)^*\right]
+% \end{align}
+% where we replaced the radial Lagrangian displacement by the Eulerian
+% radial velocity perturbation in Eq. \ref{baroclinic_torque}. 
+% The linearised momentum equations give
+% \begin{align}
+%   D\delta v_{Rm} =& 
+%   \ii\sbar\left[c_s^2\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right)
+%     + \frac{d}{dR}\delta\Phi_m\right] \notag\\ 
+%   &- \frac{2\ii
+%     m\Omega}{R}\left(c_s^2\frac{\delta\Sigma_m}{\Sigma} +
+%     \delta\Phi_m\right),
+% \end{align}
+% where $D\equiv \kappa^2 - \sbar^2$. 
+% We now invoke local theory by setting $d/dR \to \ii k$ where $k$ is
+% real, and using the local solution to the Poisson equation
+% \begin{align}
+%   \delta\Phi_m = -\frac{2\pi G}{|k|}\delta\Sigma_m 
+% \end{align}
+% \citep{shu91}. Then the radial velocity perturbation becomes
+% \begin{align}
+%   \delta v_{Rm} =
+%   \frac{1}{D}\frac{\delta\Sigma_m}{\Sigma}\left(\frac{2\pi G
+%       \Sigma}{|k|} - c_s^2\right)\left(k\sbar + \frac{2\ii
+%       m\Omega}{R}\right),
+% \end{align}
+% so that 
+% \begin{align}
+%   \ii\delta\Sigma_m \left(\frac{\delta v_{Rm}}{\sbar}\right)^* =
+%   \frac{1}{D^*}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
+%       \Sigma}{|k|} - c_s^2\right)\left(\ii k  + \frac{2
+%       m\Omega}{R\sbar^*}\right). 
+% \end{align}
+% For a neutral mode, $\sbar$, and therefore $D$ is real. Then, after
+% taking the imaginary part of the above expression, the torque
+% density is 
+% \begin{align}\label{tbg_explicit}
+%   T_\mathrm{BG} &=
+%   \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{D}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
+%       \Sigma}{|k|} - c_s^2\right)\right].
+% % &\simeq \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{\left(\kappa^2 -
+% %         m^2\Omega^2\right)}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi
+% %       G \Sigma}{|k|} - c_s^2\right)\right]. 
+% \end{align}
+% We now use the local dispersion relation,
+% Eq. \ref{dispersion}, to make the replacement $D = 2\pi G \Sigma |k| -
+% k^2c_s^2$ in the above expression. This gives
+% \begin{align}\label{tbg_simple}
+%   T_\mathrm{BG} =
+%   \frac{m}{2}\frac{dc_s^2}{dR}\frac{1}{k}\frac{|\delta\Sigma_m|^2}{\Sigma}.  
+% \end{align}

### 2015-01-01 paper/paper.tex
major restructuring and rewrite in response to report

@@
 \begin{abstract} 
-  We present numerical simulations of astrophysical discs with an 
-  imposed temperature profile and in which a sub-portion of it is
-  self-gravitating. This might represent, for example, mass  
-  built-up in a circumstellar or protoplanetary
-  disc at the outer edge of a `dead zone', in which there
-  is inefficient magneto-hydrodynamic turbulence to enable accretion. 
-  We conduct numerical gravito-hydrodynamic experiments in 2D using
-  the FARGO code and in 3D with the ZEUS-MP and PLUTO codes. In such
-  discs we observe the development of a 
-  spiral mode with azimuthal wavenumber $m=1$ and co-rotation radius 
-  outside the bulk of the spiral arm, resulting in a nearly stationary 
-  one-armed spiral pattern in the self-gravitating region of the
-  disc. The development of this one-armed 
-  spiral does not require the movement of  
-  the central star, as found in previous studies. Instead, we find it
-  to be associated with the imposed 
-  temperature gradient and the radial disc structure. 
-  We apply results from linear theory and attribute the
-  growth of this spiral mode to the non-conservation of angular momentum
-  for non-axisymmetric linear perturbations in discs with a forced temperature
-  gradient. This may lead to an unstable exchange of angular momentum
-  between the spiral mode 
-  and the background disc. However, because the interaction does
-  not explicitly depend on self-gravity, we suggest it may also
+  {\bf We describe a new mechanism that leads to the destabilisation of
+  one-armed spiral or eccentric modes in astrophysical discs with 
+  an imposed radial temperature gradient.  We use linear density wave
+  theory 
+  to show that non-axisymmetric perturbations generally do not conserve
+  their angular momentum in the presence of a forced temperature
+  gradient. This implies an exchange of angular momentum
+  between linear perturbations and the background disc. In particular, 
+  when the disturbance is a low-frequency, one-armed
+  trailing spiral and the disc temperature decreases outwards, this 
+  interaction is unstable and leads to the growth of the spiral mode.  
+  We demonstrate this phenomenon through numerical hydrodynamic
+  simulations of locally isothermal discs in 2D using the FARGO
+  code and in 3D with  the ZEUS-MP and PLUTO codes.  We consider
+  radially structured discs with a self-gravitating sub-portion which remains
+  stable in the absence of a temperature gradient. However,
+  when a temperature gradient is imposed we observe exponential growth of an one-armed  
+  spiral mode (azimuthal wavenumber $m=1$) with co-rotation radius 
+  outside the bulk of the spiral arm, resulting in a low-frequency  
+  spiral pattern in the self-gravitating region of the disc.  The development of this one-armed 
+ spiral does not require the movement of the central star, as found in previous studies. 
+ Because destabilisation by the imposed temperature gradient does 
+ not explicitly depend on self-gravity, we suggest this mechanism may also
  affect low-frequency one-armed oscillations in non-self-gravitating
  discs. 
+}
 \end{abstract}
+
+%This might represent, for example, mass  
+ % built-up in a circumstellar or protoplanetary
+ % disc at the outer edge of a `dead zone', in which there
+  %is inefficient magneto-hydrodynamic turbulence to enable accretion.
 

### 2015-01-02 paper/appendix.tex
major restructuring and rewrite in response to report

@@
 \section{Relation between horizontal Lagrangian displacements for
-  local, low frequency \lowercase{$m=1$} disturbances}\label{horizontal_displacements}
+  local, low frequency disturbances}\label{horizontal_displacements}
 Here, we aim to relate the horizontal Lagrangian displacements $\xi_R$
@@
 \end{align}
-For low-frequency $m=1$ modes we have $\sbar \simeq -\Omega$, so
-$\xi_\phi\simeq 2\ii\xi_R$, as used in the main text.
+{\bf 
+For low-frequency modes we have $\sbar \simeq -m\Omega$, so
+$\xi_\phi\simeq 2\ii\xi_R/m$, as used in the main text.  
+}

### 2015-01-02 paper/discussion.tex
major restructuring and rewrite in response to report

@@
+%to discuss - fixed central star
+%application to ppd
+%do smooth disc results carry over to structured disks?
+
 \section{Discussion}\label{discussions}
-Having confirmed the $m=1$ spiral instability through numerical
-simulations, we now use results from the simulations combined with
-linear theory to construct an explanation for their growth. For
-simplicity, we return to 2D discs.
 
-The dominant disturbance in our simulations is an one-armed, $m=1$
-trailing spiral, and its growth is associated with the imposed temperature
-gradient. As discussed in \S\ref{global_cons}, in a locally isothermal disc this imposed
-temperature gradient generally results in a torque being exerted on
-linear perturbations. An instability is therefore possible if the sign
-of this torque is the same as that of the angular momentum associated
-with the linear perturbation.
+% Having confirmed the $m=1$ spiral instability through numerical
+% simulations, we now use results from the simulations combined with
+% linear theory to construct an explanation for their growth. For
+% simplicity, we return to 2D discs.
+
+% The dominant disturbance in our simulations is an one-armed, $m=1$
+% trailing spiral, and its growth is associated with the imposed temperature
+% gradient. As discussed in \S\ref{global_cons}, in a locally isothermal disc this imposed
+% temperature gradient generally results in a torque being exerted on
+% linear perturbations. An instability is therefore possible if the sign
+% of this torque is the same as that of the angular momentum associated
+% with the linear perturbation.
 
@@
 %also transport ang mom outwards
+
+{\bf
+  \subsection{Relevance to protoplanetary discs}
+
+}

### 2015-01-02 paper/model.tex
major restructuring and rewrite in response to report

@@
 {\bf 
-  It is simplest to demonstrate instability by determining  
-  the sign of the angular momentum density 
-  $\jlin$ and the background torque $T_\mathrm{BG}$ for perturbations
-  in the local approximation, which we review next. 
-  % depends on the
-  % type of perturbation being considered. 
-  We then show that low-frequency perturbations ($|\omega|\ll\Omega$)
-  are unstable when there a negative radial 
+  We demonstrate instability for low-frequency
+  ($\omega\ll m\Omega$) perturbations by explicitly calculating its 
+  angular momentum density 
+  $\jlin$ and the background torque $T_\mathrm{BG}$, and show that
+  they are unstable when a negative radial 
   temperature gradient is imposed.  
@@
 \subsection{Angular momentum density of  non-axisymmetirc low-frequency modes}
-{\bf We now specialise to low-frequency modes and first determine the sign of its
-  angular momentum density.} From Eq. \ref{lin_ang_mom_cons} and assuming a time-dependence of the
+From Eq. \ref{lin_ang_mom_cons} and assuming a time-dependence of the
 form $\exp{(-\ii \sigma t)}$ with $\real{\sigma} = \omega$,  
@@
       m\Omega\right)|\bm{\xi}|^2 + \ii\Omega\left(\xi_R\xi_\phi^* -
-      \xi_R^*\xi_\phi\right)\right].  
+      \xi_R^*\xi_\phi\right)\right].\label{ang_mom_def}  
 \end{align}
@@
 \end{align}
-
-Thus, low-frequency non-axisymmetric modes ($m\geq1$)
-
-Thus, low-frequency $m=1$ modes generally have negative
-angular momentum. These are the modes observed in our simulations. 
-If the mode loses (positive) angular momentum 
-to the background, then we can expect instability. We demonstrate
+{\bf
+Thus, non-axisymmetric ($m\geq1$) low-frequency modes generally have 
+negative angular momentum. If the mode loses (positive) angular momentum 
+to the background, then we can expect instability. We show 
 below how this is possible through a forced temperature
-gradient.
+gradient via the background torque. It is simplest to calculate
+$T_\mathrm{BG}$ in the local approximation, which we review first.  
+}
 
@@
 
-\subsection{Local results}
+\subsection{Local results}\label{local_approx}
 In the local approximation, perturbations are assumed to vary rapidly
@@
 where $k$ is a real wavenumber such that $|kR|\gg1$ \citep{shu91}. 
-Note that in the local approximation, only axisymmetric perturbations
+Note that in the {\bf strictly} local approximation, {\bf where all
+  global effects are neglected}, only axisymmetric perturbations
 ($m=0$) can be unstable. % (having an imaginary $\sigma$).
@@
 A detailed discussion of the properties of local density waves 
-is given in \cite{shu91}. An important result is that
+is given in \cite{shu91}. An important result, {\bf which holds for
+  waves of all frequencies}, is that
 waves interior to co-rotation ($R<R_c$) have negative angular momentum, while
 waves outside co-rotation ($R>R_c$) have positive angular
-momentum. The local approximation becomes less good for global modes,
-but will nevertheless be useful for results interpretation.
+momentum. 
+%The local approximation becomes less good for global modes,
+%but will nevertheless be useful for results interpretation.
 
@@
 {\bf 
-Here we show that the torque density acting on a low-frequency, $m=1$
+Here we show that the torque density acting on a low-frequency, 
 local mode due to the background temperature gradient can be negative, which 
@@
 However, if the linear disturbance under consideration \emph{is} a
-low-frequency $m=1$ mode, then it has negative angular
-momentum. {\bf Then} if it is a trailing spiral and $dc_s^2/dR<0$, as is typical
+low-frequency mode, then it has negative angular
+momentum. If it is trailing and $dc_s^2/dR<0$, as is typical
 in astrophysical discs, then $T_\mathrm{BG}<0$, and
@@
 density is quadratic in the linear perturbations. Inserting the above
-expressions for $\jlin$ and $T_\mathrm{BG}$ for $m=1$ gives
+expressions for $\jlin$ and $T_\mathrm{BG}$ for gives
 \begin{align}\label{theoretical_rate0}
   2\gamma \sim
-  -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R|^2}{|\xi_R + \ii
-    \xi_\phi|^2}. 
-%  \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
+  -\frac{dc_s^2}{dR}
+  \frac{k}{\Omega}\frac{|\xi_R|^2}{\left[(m-1)|\bm{\xi}|^2 + |\xi_R +
+      \ii\xi_\phi|^2 \right]} 
+  = -\frac{dc_s^2}{dR}
+  \frac{k}{m\Omega},
+%  -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R|^2}{|\xi_R + \ii
+%   \xi_\phi|^2}. 
 \end{align}
-In Appendix \ref{horizontal_displacements}, we argue $\xi_\phi \simeq
-2\ii \xi_R$ for the modes of interest. Then 
+{\bf where the second equality uses $\xi_\phi \simeq
+2\ii \xi_R/m$ in the local approximation, as shown in  Appendix
+\ref{horizontal_displacements}.} 
+ % {\bf
+% This holds for local modes of all frequencies and can be shown
+% explicitly by evaluating Eq. \ref{ang_mom_def} with 
+% \begin{align}
+%   \xi_\phi = -\frac{2\ii\Omega}{\sbar}\xi_R, 
+% \end{align}
+% consistent with the local approximation as shown in Appendix
+% \ref{horizontal_displacements}. 
+% }
+Then 
 %To further simplify this expression, we need to relate $\xi_R$ and
@@
 \begin{align}
-  2\gamma \sim q\frac{c_s^2}{R}\frac{k}{\Omega} \sim q h
-  \left(kH\right)\Omega.  
+  2\gamma \sim q\frac{c_s^2}{R}\frac{k}{m\Omega} \sim q h
+  \left(\frac{kH}{m}\right)\Omega,\label{theoretical_rate}
 \end{align}
-
-{\bf It is important to note that this growth is
-due to the temperature gradient and does not depend explicitly on
+where we used $c_s\sim H\Omega\sim hR\Omega$. 
+{\bf It is important to note that, although our description of the local
+  approximation accounts for disc self-gravity, mode growth due 
+  to the fixed temperature gradient does not depend explicitly on
   whether or not the disc is self-gravitating, i.e. destabilisation is
@@
 
-For perturbations in a self-gravitating disc with characteristic
-wavenumber $k \sim \pi G \Sigma/c_s^2 \sim \Omega/Q c_s$, as seen in
-our simulations, we have 
-\begin{align}\label{theoretical_rate}
-  \gamma \sim \frac{qh}{2Q}\Omega,
-\end{align}
-where we used $c_s/R\sim h\Omega$.
+% For perturbations in a self-gravitating disc with characteristic
+% wavenumber $k \sim \pi G \Sigma/c_s^2 \sim \Omega/Q c_s$, as seen in
+% our simulations, we have 
+% \begin{align}\label{theoretical_rate}
+%   \gamma \sim \frac{qh}{2Q}\Omega,
+% \end{align}
+% where we used $c_s/R\sim h\Omega$.
 
@@
 %but illustrate with a concrete example
-We adopt computational units such that
-$G=M_*=R_0=1$. Time is measured  in the Keplerian orbital period at
-the reference radius, $P_0\equiv  2\pi/\Omega_k(R_0)$.
+{\bf
+  We illustrate the destabilising effect of a fixed temperature
+  gradient using numerical simulations. The above discussion is
+  generic for low-frequency non-axisymmetric modes, but for
+  simulations we will consider specific examples.
+
+  There are two components to the destabilising mechanism.  First
+  the disc should support low-frequency modes, which is then
+  destabilised by an imposed temperature gradient. The latter is
+  straight forward to implement by adopting a locally isothermal
+  equation of state as described in \S\ref{model}.  For the former, we
+  consider discs with a  
+  radially-structured Toomre $Q$ profile. We use local theory to show that such discs support
+  low-frequency trapped one-armed ($m=1$) modes. 
+  This is convenient because Eq. \ref{theoretical_rate} indicates that
+  modes with small $m$ are
+  most favourable for destabilisation by an imposed temperature
+  gradient.  
+}
 
@@
 \end{align}
-where $\Delta_{1,2} = \delta R \times H(R_{1,2})$. The bump function
-mimics mass built-up in $R\in[R_1,R_2]$.
+where $\Delta_{1,2} = \delta R \times H(R_{1,2})$.
+
+% {\bf The bump function mimics mass built-up in $R\in[R_1,R_2]$.
+
+% }
+
+% In the case of protoplanetary discs, which have complex structure
+% \citep{armitage10}, GI can also occur in `dead
+% zones' --- sub-portions of the disc where the magneto-rotational
+% instability is ineffective for angular momentum transport
+% \citep{gammie96,turner08,landry13} and therefore has a reduced mass
+% accretion rate. Dead zones may then become
+% self-gravitating with sufficient mass built-up
+% \citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13,bae14}.
+
+%This might represent, for example, mass  
+ % built-up in a circumstellar or protoplanetary
+ % disc at the outer edge of a `dead zone', in which there
+  %is inefficient magneto-hydrodynamic turbulence to enable accretion.
 
@@
 $Q_\mathrm{out}=2$. An example of the initial surface density and the
-Toomre $Q$ parameter (see\S\ref{wkb}) is shown in
+Toomre $Q$ parameter is shown in
 Fig. \ref{initial_surf}. Since $Q>1$, the disc is stable to local
@@
 
-\subsection{FARGO}
+\subsection{Codes}
+{\bf We use three independent grid-based codes to simulate the aforementioned
+system.} We adopt computational units such that
+$G=M_*=R_0=1$. Time is measured  in the Keplerian orbital period at
+the reference radius, $P_0\equiv  2\pi/\Omega_k(R_0)$.
+
+\subsubsection{FARGO}
 Our primary code is FARGO with self-gravity \citep{baruteau08}. This
@@
 % BC
-\subsection{ZEUS-MP}
+\subsubsection{ZEUS-MP}
 ZEUS-MP  is a general-purpose finite difference
@@
 %used in lin 2012 
-\subsection{PLUTO} 
+\subsubsection{PLUTO} 
 PLUTO is a general-purpose Godunov code \citep{mignone07}. The grid

### 2015-01-02 paper/paper.tex
major restructuring and rewrite in response to report

@@
   {\bf We describe a new mechanism that leads to the destabilisation of
-  one-armed spiral or eccentric modes in astrophysical discs with 
+ non-axisymmetric waves in astrophysical discs with 
   an imposed radial temperature gradient.  We use linear density wave
@@
   between linear perturbations and the background disc. In particular, 
-  when the disturbance is a low-frequency, one-armed
-  trailing spiral and the disc temperature decreases outwards, this 
-  interaction is unstable and leads to the growth of the spiral mode.  
+  when the disturbance is a low-frequency
+  trailing wave and the disc temperature decreases outwards, this 
+  interaction is unstable and leads to the growth of the wave.  
   We demonstrate this phenomenon through numerical hydrodynamic
@@
   spiral mode (azimuthal wavenumber $m=1$) with co-rotation radius 
-  outside the bulk of the spiral arm, resulting in a low-frequency  
-  spiral pattern in the self-gravitating region of the disc.  The development of this one-armed 
+  outside the bulk of the spiral arm, resulting in a nearly-stationary
+  one-armed spiral pattern.  The development of this one-armed 
  spiral does not require the movement of the central star, as found in previous studies. 
@@
 outer parts of the disc \citep{rafikov05,matzner05,kimura12}.
-% outer parts  
-In the case of protoplanetary discs, which have complex structure
-\citep{armitage10}, GI can also occur in `dead
-zones' --- sub-portions of the disc where the magneto-rotational
-instability is ineffective for angular momentum transport
-\citep{gammie96,turner08,landry13} and therefore has a reduced mass
-accretion rate. Dead zones may then become
-self-gravitating with sufficient mass built-up
-\citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13,bae14}.
+
+% In the case of protoplanetary discs, which have complex structure
+% \citep{armitage10}, GI can also occur in `dead
+% zones' --- sub-portions of the disc where the magneto-rotational
+% instability is ineffective for angular momentum transport
+% \citep{gammie96,turner08,landry13} and therefore has a reduced mass
+% accretion rate. Dead zones may then become
+% self-gravitating with sufficient mass built-up
+% \citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13,bae14}.
 
 % original motivation for this work (DZGI)
-A wide range of GIs can occur in  gaseous 
-astrophysical discs 
-\citep{papaloizou89,christo92,christo93,hadley11,hadley14}. It is then
-of theoretical importance to clarify what type of GI is relevant to 
-partially self-gravitating discs, such as a protoplanetary disc containing a 
-massive dead zone.  %This was indeed the original goal of this project. 
-This work presents a first step towards this goal. Specifically, here we 
-study a class of seemingly simple disc models which were found to 
-consistently produce \emph{one-armed} spiral structures, despite not 
-satisfying the usual requirements to generate them.
+
+%A wide range of GIs can occur in  gaseous 
+%astrophysical discs 
+%\citep{papaloizou89,christo92,christo93,hadley11,hadley14}.
+
+% It is then
+% of theoretical importance to clarify what type of GI is relevant to 
+% partially self-gravitating discs, such as a protoplanetary disc containing a 
+% massive dead zone. 
+% This work presents a first step towards this goal. Specifically, here we 
+% study a class of seemingly simple disc models which were found to 
+% consistently produce \emph{one-armed} spiral structures, despite not 
+% satisfying the usual requirements to generate them.
 

### 2015-01-02 paper/results.tex
major restructuring and rewrite in response to report

@@
 Fig. \ref{fargo_2d}. 
-At early times $t\lesssim100P_0$ the disc 
+At early times $t\lesssim100P_0$ the disc {\bf is } 
 dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
@@
 plotted since they are dominant. The $m=1$ structure has
-an associated negative angular momentum.  
+an associated negative angular momentum, {\bf which indicates it is a
+  low-frequency mode.}   
 Its growth is compensated by an increase in the axisymmetric
@@
 
-\subsection{Properties of the $m=1$ spiral and its growth}\label{fargo_m1}
-In order to characterise the $m=1$ spiral that eventually
-dominates, here we examine a simulation initialised with 
-$M=1$ in Eq. \ref{randpert}. The one-armed spiral that emerges is more
-coherent than the above simulation, which facilitates the analysis
-below.
+\subsection{Dependence on the imposed temperature profile}
+{\bf We show that the growth of the $m=1$ spiral is
+  associated with the imposed temperature gradient  by performing a 
+  series of simulations with $q\in[0,1]$,  However, to keep the Toomre $Q$
+  profiles similar, we adjust the surface density power-law index such
+  that $s = (3+q)/2$.  For clarity these simulations are initialised
+  with $m=1$ perturbations only.  
+}
 
-Fig. \ref{2d_fargo_viz} shows a snapshot of the $m=1$ surface
-density. By measuring the $m=1$ surface density amplitude and its
-pattern speed, we obtain a co-rotation radius and growth rate 
-\begin{align*}
-  &R_c \simeq 4.4R_0,\\
-  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_p. 
-\end{align*}
-This one-armed spiral can be considered as low frequency because its 
-pattern speed $\Omega_p \simeq 0.1\Omega_k(R_0)\lesssim 0.3\Omega$ in
-$R\in[R_{1},R_{2}]$ (where it has the largest amplitude). Thus, the spiral 
-pattern appears nearly stationary. The growth rate $\gamma$ is also slow
-relative to the local rotation, although the characteristic growth
-time $\gamma^{-1} \simeq 10P_0$ is not very long.  
-\begin{figure}
-  \includegraphics[width=\linewidth]{figures/polarxy2_dens120_fargo}
-  \caption{Cartesian visualisation of the $m=1$ surface density
-    structure in the FARGO simulation initialised with only $m=1$
-    perturbations. 
-    \label{2d_fargo_viz}} 
-\end{figure}
-
-Next, we write $\Sigma_1 = 
-|\Sigma_1|\exp{(\ii kR)}$, where $k$ is real, and assume the amplitude
-$|\Sigma_1|$ varies slowly compared to the complex phase. This is the
-main assumption in local theory. We calculate $k$ numerically and plot
-its normalised value in Fig. \ref{fargo_wavenumber}. We find 
-\begin{align*}
-  kR \sim \frac{\pi G \Sigma}{c_s^2}R \sim \frac{1}{hQ}, 
-\end{align*}
-where we used $Q\sim c_s\Omega/\pi G \Sigma$ and $R\Omega/c_s\sim
-h^{-1}$. Since $Q=O(1)$ and $h\ll 1$ imply $|kR|\gg 1$, we 
-can apply results from local theory
-self-consistently. Fig. \ref{2d_fargo_viz} shows the $m=1$ spiral is
-trailing, which is consistent with $k>0$.
-
-%snapshot shows it's tightly wound
-\begin{figure}
-  \includegraphics[width=\linewidth]{figures/m1_analysis_kr120_fargo}
-  \caption{Normalised radial wavenumber of the $m=1$ spiral in 
-    Fig. \ref{2d_fargo_viz}.\label{fargo_wavenumber}} 
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%
-
-Using the estimated value of $R_c$, we plot in
-Fig. \ref{fargo_qbarrier} the quantity $\nu^2 - 1 + Q^{-2}$, which is
-required to be positive in local theory for purely wave-like
-solutions to the dispersion relation (Eq. \ref{dispersion}) when the
-mode frequency is given. %Note that, with the above measured wavenumber $k$, 
-%Eq. \ref{dispersion} requires $\nu^2 - 1 + Q^{-2}\sim0$ and the lower
-%sign (long waves) to be taken. 
-Fig. \ref{fargo_qbarrier} shows two $Q$-barriers located in
-the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$; the bounded region is indeed 
-where the $m=1$ spiral develops. This suggests that the one-armed 
-spiral is trapped. Note in this region, $\nu^2 - 1 + 
-Q^{-2}\simeq 0.1\ll 1$, which is necessary for consistency with 
-the measured wavenumber $k$ and Eq. \ref{wavenumber}.  
-%Eq. \ref{dispersion} with the lower sign taken (i.e. long waves).  
-There is one outer Lindblad resonance at $R_L\simeq
-7.2R_0$. Thus, acoustic waves may be launched in $R\gtrsim 7.2R_0$ by
-the spiral disturbance in the inner disc \citep{lin11b}. 
-%implications: effect on dynamics in outer disc
-
-\begin{figure}
-  \includegraphics[width=\linewidth]{figures/m1_analysis_Qbar_fargo} 
-  \caption{Dimensionless mode frequency $\nu$ for the $m=1$ spiral in
-    Fig. \ref{2d_fargo_viz}. For a given real mode frequency, the
-    dispersion relation for local density waves, Eq. \ref{dispersion},
-    permits purely wave-like solutions in regions where $\nu^2 - 1 +
-    Q^{-2}>0$.    
-    \label{fargo_qbarrier}} 
-\end{figure}
-
-\subsubsection{Angular momentum exchange with the background disc}  
-The wavenumber $k = \pi G\Sigma/c_s^2$ minimises the shifted frequency
-$\sbar$ in the dispersion relation, Eq. \ref{dispersion}.  
-Physically, then, it is not surprising to 
-find perturbations of this wavenumber in a
-self-gravitating disc where $Q\sim 1$. However, according the
-local dispersion relation, $m=1$ perturbations are
-formally stable for real $k$ and $Q>1$.
-
-%most of the perturbation is inside co-rotation 
-%so resonant interaction unlikely for instability
-
-In order to track down the origin for the (slow) growth of the
-$m=1$ spiral, we invoke angular momentum conservation for linear
-perturbations, Eq. \ref{lin_ang_mom_cons}. Assuming fluxes are
-negligible at the disc boundaries, we can integrate
-Eq. \ref{lin_ang_mom_cons} to give
-\begin{align}\label{baroclinic_torque_int}
-  \frac{d}{dt}\underbrace{\int_{R_\mathrm{min}}^{R_\mathrm{max}}\jlin
-    2\pi R dR}_{\jlintot} 
-  =\int_{R\mathrm{min}}^{R\mathrm{max}}T_\mathrm{BG} 2\pi R dR, 
-\end{align}
-where we recall $T_\mathrm{BG}$ is the torque density due to the imposed
-sound-speed profile (Eq. \ref{baroclinic_torque}). We explicitly 
-compute both sides of Eq. \ref{baroclinic_torque_int} using
-simulation data, and compare them in Fig. \ref{fargo_angmom_ex}. There
-is a good match between the two torques, especially at early times
-$t\lesssim110P_0$. The average discrepancy is $\simeq 5\%$. 
-The match is less good later on, when the spiral
-amplitude is no longer small ($\Delta\Sigma_1\sim 0.2$ at $t=110P_0$
-and $\Delta\Sigma_1\sim 0.4$ by $t=120P_0$) and linear theory becomes
-less applicable. Fig. \ref{fargo_angmom_ex} confirms that the $m=1$ spiral
-wave experiences a negative torque that further reduces its angular
-momentum. This is consistent with angular momentum component
-measurements (Fig. \ref {2d_angmom}).
-
-% similar results if only intergrating between R_1 and R_2
-%wave losses at boundaries
-
-\begin{figure}
-  \includegraphics[width=\linewidth]{figures/m1_analysis_ang_fargo} 
-  \caption{Rate of change of the $m=1$ wave angular momentum as defined by
-    Eq. \ref{baroclinic_torque_int} (solid) compared to the torque
-    exerted on the wave associated with the background temperature
-    gradient (dotted). 
-    \label{fargo_angmom_ex}} 
-\end{figure}
-
-%ang mom decreases slightly more rapidly than torque exchange provides
-%- probably non-linear effects/shocks
-
-\subsection{Dependence on the imposed temperature profile}
-The simulations above suggest the growth of the $m=1$ spiral mode
-can be attributed to the background torque applied on linear perturbations
-as a result of an imposed temperature gradient, as discussed in
-\S\ref{global_cons}. According to Eq. \ref{baroclinic_torque}, this
-torque is proportional to $q$, the power-law index used to set the
-sound-speed profile (Eq. \ref{sound-speed}) in our simulations. 
-We confirm this through a series of simulations with $q\in[0,1]$, 
-initialised with $m=1$ perturbations. However, to keep the Toomre $Q$
-profiles similar, we adjust the surface density power-law index such
-that $s = (3+q)/2$.
+% The simulations above suggest the growth of the $m=1$ spiral mode
+% can be attributed to the background torque applied on linear perturbations
+% as a result of an imposed temperature gradient, as discussed in
+% \S\ref{global_cons}. 
+% According to Eq. \ref{baroclinic_torque}, this
+% torque is proportional to $q$, the power-law index used to set the
+% sound-speed profile (Eq. \ref{sound-speed}) in our simulations.
 
@@
 simulation.   
-
 % difficult to measure  
 % show result of isothermal calc?
-
 \begin{figure}
@@
   \caption{Evolution of the $m=1$ spiral amplitude as a function of
-    the imposed sound-speed gradient $q$. The maximum value of the
+    the {\bf negative of the imposed temperature gradient} $q$. The maximum value of the
     $m=1$ surface density in $R\in[R_1,R_2]$ is shown. 
@@
 \end{figure}
-We plot growth rates of the $m=1$ spiral as a function of $q$ in 
+
+We plot growth rates of the $m=1$ mode as a function of $q$ in 
 Fig. \ref{fargo_varq_growth}. The correlation can be fitted with a 
@@
 % difficult to measure very small growth rates 
-This shows that the temperature gradient is responsible for the
+{\bf As the background torque is proportional to $q$
+  (Eq. \ref{theoretical_rate})}, 
+this {\bf indicates} that the temperature gradient is responsible for the
 development of the one-armed spirals observed in our simulations.
@@
 \end{figure}
+
+\subsection{Properties of the $m=1$ spiral and its growth}\label{fargo_m1}
+{\bf Here we analyse the $q=1$ case in Fig. \ref{fargo_varq_growth} in
+  more detail.  
+}
+%In order to characterise the $m=1$ spiral that eventually
+%dominates, here we examine a simulation initialised with 
+%$M=1$ in Eq. \ref{randpert}. The one-armed spiral that emerges is more
+%coherent than the above simulation, which facilitates the analysis
+%below.   
+Fig. \ref{2d_fargo_viz} shows a snapshot of the $m=1$ surface
+density {\bf of this run}. 
+By measuring the $m=1$ surface density amplitude and its
+pattern speed, we obtain a co-rotation radius and growth rate 
+\begin{align*}
+  &R_c \simeq 4.4R_0,\\
+  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_p. 
+\end{align*}
+This one-armed spiral can be considered as low frequency because its 
+pattern speed $\Omega_p \simeq 0.1\Omega_k(R_0)\lesssim 0.3\Omega$ in
+$R\in[R_{1},R_{2}]$ (where it has the largest amplitude). Thus, the spiral 
+pattern appears nearly stationary. The growth rate $\gamma$ is also slow
+relative to the local rotation, although the characteristic growth
+time $\gamma^{-1} \simeq 10P_0$ is not very long.  
+\begin{figure}
+  \includegraphics[width=\linewidth]{figures/polarxy2_dens120_fargo}
+  \caption{Cartesian visualisation of the $m=1$ surface density
+    structure in the FARGO simulation initialised with only $m=1$
+    perturbations. 
+    \label{2d_fargo_viz}} 
+\end{figure}
+
+Next, we write $\Sigma_1 = 
+|\Sigma_1|\exp{(\ii kR)}$, where $k$ is real, and assume the amplitude
+$|\Sigma_1|$ varies slowly compared to the complex phase. This is the
+main assumption in local theory. We calculate $k$ numerically and plot
+its normalised value in Fig. \ref{fargo_wavenumber}. We find 
+\begin{align*}
+  kR \sim \frac{\pi G \Sigma}{c_s^2}R \sim \frac{1}{hQ}, 
+\end{align*}
+where we used $Q\sim c_s\Omega/\pi G \Sigma$ and $R\Omega/c_s\sim
+h^{-1}$. Since $Q=O(1)$ and $h\ll 1$ imply $|kR|\gg 1$, we 
+can apply {\bf results from local theory
+  (\S\ref{local_approx})}. {\bf Note also that $k\simeq k_c$. }
+Fig. \ref{2d_fargo_viz} shows the $m=1$
+spiral is trailing, consistent with $k>0$.
+
+%snapshot shows it's tightly wound
+\begin{figure}
+  \includegraphics[width=\linewidth]{figures/m1_analysis_kr120_fargo}
+  \caption{Normalised radial wavenumber of the $m=1$ spiral in 
+    Fig. \ref{2d_fargo_viz}.\label{fargo_wavenumber}} 
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%
+
+Using the estimated value of $R_c$, we plot in
+Fig. \ref{fargo_qbarrier} the quantity $\nu^2 - 1 + Q^{-2}$, which is
+required to be positive in local theory for purely wave-like
+solutions to the dispersion relation (Eq. \ref{dispersion}) when the
+mode frequency is given. %Note that, with the above measured wavenumber $k$, 
+%Eq. \ref{dispersion} requires $\nu^2 - 1 + Q^{-2}\sim0$ and the lower
+%sign (long waves) to be taken. 
+Fig. \ref{fargo_qbarrier} shows two $Q$-barriers located in
+the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$; the bounded region is indeed 
+where the $m=1$ spiral develops. This shows that the one-armed 
+spiral is trapped. Note in this region, $\nu^2 - 1 + 
+Q^{-2}\simeq 0.1\ll 1$, which is necessary for consistency with 
+the measured wavenumber $k$ and Eq. \ref{wavenumber}.  
+%Eq. \ref{dispersion} with the lower sign taken (i.e. long waves).  
+There is one outer Lindblad resonance at $R_L\simeq
+7.2R_0$. Thus, acoustic waves may be launched in $R\gtrsim 7.2R_0$ by
+the spiral disturbance in the inner disc \citep{lin11b}. 
+%implications: effect on dynamics in outer disc
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{figures/m1_analysis_Qbar_fargo} 
+  \caption{Dimensionless mode frequency $\nu$ for the $m=1$ spiral in
+    Fig. \ref{2d_fargo_viz}. For a given real mode frequency, the
+    dispersion relation for local density waves, Eq. \ref{dispersion},
+    permits purely wave-like solutions in regions where $\nu^2 - 1 +
+    Q^{-2}>0$.    
+    \label{fargo_qbarrier}} 
+\end{figure}
+
+{\bf
+  We can estimate the expected growth rate of the $m=1$ mode due to the
+  temperature gradient. Setting $k = k_c$ and $m=1$ into
+  Eq. \ref{theoretical_rate} gives
+  \begin{align}\label{theoretical_rate1}
+    \gamma \sim \frac{qh}{2Q}\Omega. 
+  \end{align}
+  Inserting $q=1$, $h=0.05$ and $Q\simeq 1.5$  gives
+  $\gamma \simeq 0.017\Omega$, consistent with numerical
+  results.  
+}
+
+\subsubsection{Angular momentum exchange with the background disc}  
+% The wavenumber $k = \pi G\Sigma/c_s^2$ minimises the shifted frequency
+% $\sbar$ in the dispersion relation, Eq. \ref{dispersion}.  
+% Physically, then, it is not surprising to 
+% find perturbations of this wavenumber in a
+% self-gravitating disc where $Q\sim 1$. However, according the
+% local dispersion relation, $m=1$ perturbations are
+% formally stable for real $k$ and $Q>1$.   
+%most of the perturbation is inside co-rotation 
+%so resonant interaction unlikely for instability
+
+{\bf We explicitly show that the growth of the $m=1$ spiral is due to
+  the forced temperature gradient via the background torque described
+  in \S\ref{wkb}. We integrate the statement for angular momentum 
+  conservation for linear perturbations, Eq. \ref{lin_ang_mom_cons},
+  assuming boundary fluxes are negligible, to obtain
+}
+% we invoke angular momentum conservation for linear
+% perturbations, Eq. \ref{lin_ang_mom_cons}. 
+% Assuming fluxes are
+% negligible at the disc boundaries, we can integrate
+% Eq. \ref{lin_ang_mom_cons} to give
+\begin{align}\label{baroclinic_torque_int}
+  \frac{d}{dt}\underbrace{\int_{R_\mathrm{min}}^{R_\mathrm{max}}\jlin
+    2\pi R dR}_{\jlintot} 
+  =\int_{R\mathrm{min}}^{R\mathrm{max}}T_\mathrm{BG} 2\pi R dR, 
+\end{align}
+where we recall $T_\mathrm{BG}$ is the torque density associated with the imposed
+sound-speed profile (Eq. \ref{baroclinic_torque}). We  
+compute both sides of Eq. \ref{baroclinic_torque_int} using
+simulation data, and compare them in Fig. \ref{fargo_angmom_ex}. There
+is a good match between the two torques, especially at early times
+$t\lesssim110P_0$. The average discrepancy is $\simeq 5\%$. 
+The match is less good later on, when the spiral
+amplitude is no longer small ($\Delta\Sigma_1\sim 0.2$ at $t=110P_0$
+and $\Delta\Sigma_1\sim 0.4$ by $t=120P_0$) and linear theory becomes
+less applicable. Fig. \ref{fargo_angmom_ex} confirms that the $m=1$ spiral
+wave experiences a negative torque that further reduces its angular
+momentum, {\bf leading to its growth.} 
+This is consistent with angular momentum component
+measurements (Fig. \ref {2d_angmom}).
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{figures/m1_analysis_ang_fargo} 
+  \caption{Rate of change of the $m=1$ wave angular momentum as defined by
+    Eq. \ref{baroclinic_torque_int} (solid) compared to the torque
+    exerted on the wave associated with the background temperature
+    gradient (dotted). 
+    \label{fargo_angmom_ex}} 
+\end{figure}

### 2015-01-02 paper/results_3d.tex
major restructuring and rewrite in response to report

@@
 \end{align*}
-Both are smaller than the 2D simulations, because of the 
-lower resolutions adopted in 3D and/or the small softening length
-employed in 2D. 
+Both are somewhat smaller than the 2D simulations. {\bf This is
+  possibily} because of the lower resolutions adopted in 3D {\bf
+  and/or because the effective Toomre parameter is larger in 3D which,
+  from Eq. \ref{theoretical_rate1}, is stabilising.}
+
+% and/or the small softening length
+% employed in 2D. 
 %Rc = 8.6R_0

### 2015-01-07 paper/appendix.tex
major revision in response to report

@@
 }
+
+\section{The confined spiral as an external potential}\label{disc-planet}
+Let us treat the one-armed spiral confined in $R\in[R_1,R_2]$ as an  
+external potential of the form $\Phi_\mathrm{ext}(R)\cos{\left(\phi -
+    \Omega_pt\right)}$. We take 
+\begin{align} 
+  \Phi_\mathrm{ext}
+  =-\frac{GM_\mathrm{ring}}{\overline{R}}b^{1}_{1/2}(\beta),   
+\end{align}
+where $M_\mathrm{ring}$ is the disc mass contained within
+$R\in[R_1,R_2]$, $\overline{R} = (R_1+R_2)/2$ is the approximate radial
+location of the spiral, $b_{n}^m(\beta)$ is the Laplace coefficient
+and $\beta = R/\overline{R}$. This form of
+$\Phi_\mathrm{ext}$ is the $m=1$ component of the gravitational
+potential of an external satellite on a circular orbit
+\citep{goldreich79}.
+
+%For an order-of-magnitude exercise, we simply take 
+%\begin{align}
+ % \Phi_\mathrm{ext} = -\frac{GM_\mathrm{ring}}{|R - \overline{R}|}, 
+%\end{align}
+%for the potential associated with a disturbance located at
+%$R=\overline{R}$, where $M_\mathrm{ring}$
+
+We expect the external potential to exert a torque on the disc at the
+Lindblad and co-rotation resonances. At the outer Lindblad resonance
+(OLR), this torque is 
+\begin{align}
+  \Gamma_L =
+  \frac{\pi^2\Sigma_L}{3\Omega_L\Omega_p}
+  \left[\left.R_L\frac{d\Phi_\mathrm{ext}}{dR}\right|_L + 2\left(1 -
+      \frac{\Omega_p}{\Omega_L}\right)\Phi_\mathrm{ext}\right]^2, 
+\end{align}   
+where a Keplerian disc has been assumed and subscript $L$ denotes
+evaluation at the OLR, $R=R_L$. (The inner Lindblad resonance does not
+exist for the pattern speeds observed in our simulations.)
+
+If we associate the external potential with angular momentum 
+$J_\mathrm{ext}  = M_\mathrm{ring}\overline{R}^2\Omega_p$, we can
+calculate a growth rate $\gamma_L=\Gamma_L/J_\mathrm{ext}$. Then 
+\begin{align}
+  \frac{\gamma_L}{\Omega_p} = &\frac{\pi
+    h}{3Q_L}\left(\frac{M_p}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)\left(\frac{R_c}{\overline{R}}\right) 
+  ^3\left(\frac{R_L}{R_c}\right)^{-3/2}\notag\\ 
+  &\times
+  \left\{\frac{R_L}{\overline{R}}\left.\frac{db_{1/2}^1}{d\beta}\right|_L
+    + 2\left[1 - 
+      \left(\frac{R_c}{R_L}\right)^{-3/2}\right]b_{1/2}^1(\beta_L)\right\}^2. 
+\end{align}
+%\begin{align}
+ % \frac{\gamma_L}{\Omega_p} \sim &\frac{\pi h}{3
+  %  Q_L}\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)^{-1/2}\left(\frac{R_c}{\overline{R}}\right)^{5/2}\notag\\
+ % &\times \left\{ \frac{R_L}{R_L - \overline{R}} - 2\left[1 -
+  %    \left(\frac{R_c}{R_L}\right)^{-3/2}\right]\right\}^2\frac{R_c^2}{\left(R_L
+   % - \overline{R}\right)^2}. 
+%\end{align}
+Inserting $h=0.05$, $Q_L=10$, $M_\mathrm{ring} = 0.05M_*$,
+$R_L=7.2R_0$, $R_c=4.4R_0$ and $\overline{R}=1.5R_0$ from our fiducial
+FARGO simulation, we get
+\begin{align}
+  \gamma_L \sim 5\times10^{-4}\Omega_p. 
+\end{align}
+%\begin{align}
+ % \gamma_L \sim 0.012\Omega_p. 
+%\end{align}
+%This represents an upper limit on the effect of the OLR on the spiral
+%disturbance, and is much smaller than the growth rate measured in the
+%simulation ($\gamma\sim0.1\Omega_p$). We conclude that for our models,
+%the OLR has negligible effect on the one-armed spiral.  
+For the co-rotation torque, we use the result
+\begin{align}
+  \Gamma_c = \left.
+    \pi^2\Phi_\mathrm{ext}^2\left(\frac{d\Omega}{dR}\right)^{-1}\frac{d}{dR}\left(\frac{2\Sigma}{\Omega}\right)\right|_{c}     
+\end{align}
+for a Keplerian disc, where subscript $c$ denotes evaluation at
+co-rotation radius $R=R_c$. For a power-law surface density profile
+$\Sigma\propto R^{-s}$ we have
+\begin{align}
+\frac{\gamma_c}{\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_c}
+\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_c}{\overline{R}}\right)^4\left(s-\frac{3}{2}\right) 
+\left[b_{1/2}^1(\beta_c)\right]^2    
+\end{align}
+%\begin{align}
+%  \frac{\Gamma_C}{J_\mathrm{ext}\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_C} \left(s -
+ %   \frac{3}{2}\right)\left(\frac{M_\mathrm{ring}}{M_*}\right)\frac{R_c^4}{\overline{R}^2\left(R_c
+ %     - \overline{R}\right)^2}.   
+%\end{align}
+Using the above parameter values with $s=2$ and $Q_c=10$, we obtain a
+rate
+\begin{align}
+  \gamma_c\sim 6\times 10^{-4}\Omega_p. 
+\end{align}

### 2015-01-07 paper/discussion.tex
major revision in response to report

@@
 \section{Discussion}\label{discussions} 
+   % specific to our numerical 
+   % simulations%  and may or may not be relevant to generic low-frequency
+   %   % modes destabilised by a forced temperature gradient.  
+   %
+
+{\bf
+  \subsection{Applicability to circumstellar discs}
+
+  \subsubsection{Thermodynamic requirements}
+
+  % talk about additional simulations at high Q
+
+  %issues with fragmentation 
+ % requirement for fast cooling  
+  % fast cooling fragments 
+  % stabilised by irradiation
+  % background torque and fragmentation may not be mutually exclusive
+  % (speculations) 
+  % these studies do no account for possible global disk structure 
+  %Clarke (2009), Rafikov (2005), and Rice & Armitage (2009)
+  %no fragmentation in our case because disc is maintained at slightly
+  %higher values of Q. no fragmentation of one arm spiral because
+  %co-rotation radius isn't on it 
+  %if irradiation 
+  %one-armed low frequency spiral probably will not fragment because
+  %co-rotation is far away
+
+  A locally isothermal equation of state represents the limit
+  of infinitely short cooling (and heating) timescales, so the
+  disc temperature instantly returns to its initial value when
+  perturbed.  A short cooling timescale $t_c$ can occur in the outer
+  parts of protoplanetary discs 
+  \citep{rafikov05,clarke09,rice09,cossins10}.  %rafikov07 (convection)
+  However, if a disc with 
+  $Q\simeq 1$ is cooled (towards zero temperature) on a timescale
+  $t_c\ll\Omega_k^{-1}$, it will fragment
+  \citep{gammie01,paarkedooper12}.
+
+  %internal heating?
+  To avoid fragmentation, the disc must be heated to
+  maintain $Q$ above the threshold for fragmentation \citep[about
+  $Q\lesssim 1.4$ for isothermal discs][]{durisen07}. This may be
+  possible in the outer parts of protoplanetary discs due to heating
+  by irradiation \citep{rafikov07,rice11}.
+
+  % Numerical simulations performed by
+  % \cite{rice11} indicate external irradiation can stablise
+  % the disc against fragmentation.
+
+  %It is important to note, however,
+  %that these studies do not consider the possible effect of global
+  %structures present in our numerical simulations.
+
+  %chiang and goldreich
+  The background torque may therefore exist in the outer
+  parts of protoplanetary discs that are strongly irradiated, such
+  that the disc temperature is set externally with a non-zero radial
+  gradient.  Of course, if external irradiation sets a strictly
+  isothermal outer disc, then the background torque vanishes.
+
+%where does the structure come from? (mass built up, planet gap)
+
+  \subsubsection{Radial disc structure}
+  In our numerical simulations the 
+  $m=1$ spiral is confined between two $Q$-barriers, where real solutions to the local
+  dispersion relation is possible (Eq. \ref{wavenumber}). The
+  existence of such a cavity is a result of the adopted disc
+  structure. Thus, in our disc models the main role of disc structure
+  and self-gravity is to 
+  allow a local $m=1$ mode to be set up, which is then 
+  destabilised by the background torque density.
+
+  In order to confine an $m=1$ spiral between two $Q$-barriers, we
+  should have $Q^2(1-\nu^2)=1$ at two radii. Assuming 
+  Keplerian rotation and a slow pattern speed $\Omega_p\ll\Omega$, this amounts to
+  \begin{align}\label{qb_cond}
+    \left(\frac{R_{Qb}}{R_c}\right)^{-3/2} = 2Q^2(R_{Qb}). 
+  \end{align}
+  Then it may be possible to have two $Q$-barriers when the $Q$ profile
+  of the disc rises more rapidly (decays more slowly) than $R^{-3/2}$
+  for decreasing (increasing) $R$.
+
+  %in fall/mass loading? 
+  The surface density bump in our initial condition might represent,
+  for example, mass built-up in a 
+`dead zone' in a protoplanetary
+  disc, where the magneto-rotational 
+  instability is ineffective for angular momentum transport
+  \citep{gammie96,turner08,landry13} and therefore has a reduced mass
+  accretion rate. The dead zone may become 
+  self-gravitating with sufficient mass built-up
+  \citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13,bae14}. 
+  However, the thermodynamic conditions in a dead zone 
+  %internal heating and cooling can balance such that a global radial
+  %temperature gradient is maintained - then background torque may
+  %exist.
+
+  %particularly simple functional representation of the bump. a
+  %realistic bump may be different
+
+  Another possibility is a gap opened by a giant planet. In that case $Q$
+  rises rapidly towards the gap centre since it is a region of low
+  surface density, which may satisfy Eq. \ref{qb_cond}. Then the inner
+  edge of our bump
+  function mimics the outer gap edge. The outer gap edge is then a
+  possible site for the growth of a low-frequency one-armed spiral
+  through the background torque. 
+ %or gap opened by a fragment
+
+}
+
+% The simulations themselves consider an isothermal or locally isothermal equation of state only.
+Gammie (2001) and other work since have shown that the evolution of self-gravitating discs depends both on the stability criterion and on the balance between heating and cooling.
+If the cooling rate is fast, it is likely to fragment, otherwise it will settle into a quasi-steady, self-gravitating state.
+
+% When considering isothermal simulations, the likely evolution is one of two things.
+It's either initially so unstable that the disc quickly fragments, or it's initially stable enough that any initial perturbations simply drives the system towards being even more stable and the instability quickly weakens through rapid mass transfer (see the work by Laughlin & Bodenheimer 1994/1996).
+There is no real way to drive a long-lived quasi-steady, non-fragmenting, self-gravitating state because this is only possible if the system can remain quasi-stable though balancing heating and cooling (see, for example, Lodato & Rice 2004).
+
+% In fact, in the conclusions it is suggested that the conditions used in the simulation would require a cooling < 0.1 \Omega^{-1} which, based on the work of Gammie 2001, would quickly lead to fragmentation.
+The reason it doesn't in the work considered here is that the isothermal assumption assumes that there is some equivalently fast heating mechanism but, in the work here, this is not provided by the instability itself.
+If it were, the simulations would become very quickly unstable.
+Also, if one looks at the work of Clarke (2009), Rafikov (2005), and Rice & Armitage (2009) the only region of a realistic disc where such rapid cooling is possible is in the outer disc, but such regions are either very gravitationally unstable and fragment, or quickly stablised by an external heating source.
+I guess it is possible that there could be a scenario where this exactly produces the conditions assumed here, but that would seem rather fine-tuned and it would be more likely that an external heating source would simply stabilise the disc (unless the disc were extremely massive and could remain unstable even at reasonably high disc temperatures).
+
+{\bf
+  \subsection{Motion of the central star}
+
+  \subsection{Role of Lindblad and co-rotation torques}
+  One possible effect of self-gravity is to allow the one-armed spiral  
+  to act as an external potential for the exterior disc. This is
+  analogous to disc-satellite interaction 
+  \citep{goldreich79}, where the embedded satellite exerts a torque on
+  the disc at
+
+In Appendix \ref{disc-planet} we estimate the magnitude of this effect
+  using basic results from disc-planet theory \citep[see, e.g.][and
+  references therein]{papaloizou07}.
+
+One possible effect of self-gravity is to allow the one-armed spiral  
+to act as an external potential for the wider disc. This is
+analogous to disc-satellite interaction 
+\citep{goldreich79}. Here, we estimate the magnitude of this effect
+using basic results from disc-planet theory \citep[see, e.g.][and
+references therein]{papaloizou07}.
+
+}
+
+%Furthermore, this
+%non-local interaction may be important for the long-term evolution of
+%the outer disc, since it gains angular momentum from the
+
+%also transport ang mom outwards
 
@@
 %also total torque is integral -> may get cancelations
-
-\subsection{The role of self-gravity and disc structure} 
-It is important to note that in the above discussion, we only considered a self-gravitating
-perturbation to obtain an appropriate wavenumber $k$ needed to
-evaluate Eq. \ref{theoretical_rate}. 
-The background torque density $T_\mathrm{BG}$ does not depend on  
-self-gravity explicitly, so self-gravity is not directly
-responsible for instability.
-
-However, 
-our numerical simulations show that the $m=1$ spiral is
-confined between two $Q$-barriers, where real solutions to the local
-dispersion relation is possible (Eq. \ref{wavenumber}), as a result of the adopted disc
-structure. Thus, in our disc models the main role of disc structure and self-gravity is to 
-allow a neutral mode to be set up and confined, which is then
-destabilised through the background torque density.
-
-In order to confine an $m=1$ spiral between two $Q$-barriers, we
-should have $Q^2(1-\nu^2)=1$ at two radii. Assuming 
-Keplerian rotation and a slow pattern speed $\Omega_p\ll\Omega$, this amounts to
-\begin{align}
-  \left(\frac{R_{Qb}}{R_c}\right)^{-3/2} = 2Q^2(R_{Qb}). 
-\end{align}
-Then it may be possible to have two $Q$-barriers when the $Q$ profile
-of the disc rises more rapidly (decays more slowly) than $R^{-3/2}$
-for decreasing (increasing) $R$. One possibility is a gap
-opened by a giant planet in a constant-$Q$ disc. In that case $Q$
-rises rapidly towards the gap centre since it is a region of low
-surface density. The outer gap edge is then a potential site for the
-growth of a low-frequency one-armed spiral through the background
-torque.
-
-%talk about additional simulations at high Q
-%spiral mode as external potential to wider disc (lp11)
-%interaction across co-rotation
-
-\subsection{The confined spiral as an external potential}
-One possible effect of self-gravity is to allow the one-armed spiral  
-to act as an external potential for the wider disc. This is
-analogous to disc-satellite interaction 
-\citep{goldreich79}. Here, we estimate the magnitude of this effect
-using basic results from disc-planet theory \citep[see, e.g.][and
-references therein]{papaloizou07}.
-
-Let us treat the one-armed spiral confined in $R\in[R_1,R_2]$ as an  
-external potential of the form $\Phi_\mathrm{ext}(R)\cos{\left(\phi -
-    \Omega_pt\right)}$. We take 
-\begin{align} 
-  \Phi_\mathrm{ext}
-  =-\frac{GM_\mathrm{ring}}{\overline{R}}b^{1}_{1/2}(\beta),   
-\end{align}
-where $M_\mathrm{ring}$ is the disc mass contained within
-$R\in[R_1,R_2]$, $\overline{R} = (R_1+R_2)/2$ is the approximate radial
-location of the spiral, $b_{n}^m(\beta)$ is the Laplace coefficient
-and $\beta = R/\overline{R}$. This form of
-$\Phi_\mathrm{ext}$ is the $m=1$ component of the gravitational
-potential of an external satellite on a circular orbit
-\citep{goldreich79}.
-
-%For an order-of-magnitude exercise, we simply take 
-%\begin{align}
- % \Phi_\mathrm{ext} = -\frac{GM_\mathrm{ring}}{|R - \overline{R}|}, 
-%\end{align}
-%for the potential associated with a disturbance located at
-%$R=\overline{R}$, where $M_\mathrm{ring}$
-
-We expect the external potential to exert a torque on the disc at the
-Lindblad and co-rotation resonances. At the outer Lindblad resonance
-(OLR), this torque is 
-\begin{align}
-  \Gamma_L =
-  \frac{\pi^2\Sigma_L}{3\Omega_L\Omega_p}
-  \left[\left.R_L\frac{d\Phi_\mathrm{ext}}{dR}\right|_L + 2\left(1 -
-      \frac{\Omega_p}{\Omega_L}\right)\Phi_\mathrm{ext}\right]^2, 
-\end{align}   
-where a Keplerian disc has been assumed and subscript $L$ denotes
-evaluation at the OLR, $R=R_L$. (The inner Lindblad resonance does not
-exist for the pattern speeds observed in our simulations.)
-
-If we associate the external potential with angular momentum 
-$J_\mathrm{ext}  = M_\mathrm{ring}\overline{R}^2\Omega_p$, we can
-calculate a growth rate $\gamma_L=\Gamma_L/J_\mathrm{ext}$. Then 
-\begin{align}
-  \frac{\gamma_L}{\Omega_p} = &\frac{\pi
-    h}{3Q_L}\left(\frac{M_p}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)\left(\frac{R_c}{\overline{R}}\right) 
-  ^3\left(\frac{R_L}{R_c}\right)^{-3/2}\notag\\ 
-  &\times
-  \left\{\frac{R_L}{\overline{R}}\left.\frac{db_{1/2}^1}{d\beta}\right|_L
-    + 2\left[1 - 
-      \left(\frac{R_c}{R_L}\right)^{-3/2}\right]b_{1/2}^1(\beta_L)\right\}^2. 
-\end{align}
-%\begin{align}
- % \frac{\gamma_L}{\Omega_p} \sim &\frac{\pi h}{3
-  %  Q_L}\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)^{-1/2}\left(\frac{R_c}{\overline{R}}\right)^{5/2}\notag\\
- % &\times \left\{ \frac{R_L}{R_L - \overline{R}} - 2\left[1 -
-  %    \left(\frac{R_c}{R_L}\right)^{-3/2}\right]\right\}^2\frac{R_c^2}{\left(R_L
-   % - \overline{R}\right)^2}. 
-%\end{align}
-Inserting $h=0.05$, $Q_L=10$, $M_\mathrm{ring} = 0.05M_*$,
-$R_L=7.2R_0$, $R_c=4.4R_0$ and $\overline{R}=1.5R_0$ from our fiducial
-FARGO simulation, we get
-\begin{align}
-  \gamma_L \sim 5\times10^{-4}\Omega_p. 
-\end{align}
-%\begin{align}
- % \gamma_L \sim 0.012\Omega_p. 
-%\end{align}
-%This represents an upper limit on the effect of the OLR on the spiral
-%disturbance, and is much smaller than the growth rate measured in the
-%simulation ($\gamma\sim0.1\Omega_p$). We conclude that for our models,
-%the OLR has negligible effect on the one-armed spiral.  
-For the co-rotation torque, we use the result
-\begin{align}
-  \Gamma_c = \left.
-    \pi^2\Phi_\mathrm{ext}^2\left(\frac{d\Omega}{dR}\right)^{-1}\frac{d}{dR}\left(\frac{2\Sigma}{\Omega}\right)\right|_{c}     
-\end{align}
-for a Keplerian disc, where subscript $c$ denotes evaluation at
-co-rotation radius $R=R_c$. For a power-law surface density profile
-$\Sigma\propto R^{-s}$ we have
-\begin{align}
-\frac{\gamma_c}{\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_c}
-\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_c}{\overline{R}}\right)^4\left(s-\frac{3}{2}\right) 
-\left[b_{1/2}^1(\beta_c)\right]^2    
-\end{align}
-%\begin{align}
-%  \frac{\Gamma_C}{J_\mathrm{ext}\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_C} \left(s -
- %   \frac{3}{2}\right)\left(\frac{M_\mathrm{ring}}{M_*}\right)\frac{R_c^4}{\overline{R}^2\left(R_c
- %     - \overline{R}\right)^2}.   
-%\end{align}
-Using the above parameter values with $s=2$ and $Q_c=10$, we obtain a
-rate
-\begin{align}
-  \gamma_c\sim 6\times 10^{-4}\Omega_p. 
-\end{align}
-
-These estimates are much smaller than that due to
-the imposed temperature gradient as measured in the simulations
-($\gamma\sim 0.1\Omega_p$). We conclude that for our disc models, the
-Lindblad and co-rotation resonances have negligible effects on the
-growth of the one-armed spiral in the inner disc. We confirmed this with
-additional FARGO simulations with a reduced radial domain size,
-setting $R_\mathrm{max}=3R_0$, which still developed the one-armed
-spiral.
-
-The torque exerted on the disc at the OLR by an
-external potential is positive, while that at co-rotation 
-depends on the gradient of potential vorticity there
-\citep{goldreich79}. For our disc models with surface density
-$\Sigma\propto R^{-2}$ in the outer disc, this co-rotation torque is
-positive. Interpreting the one-armed spiral as an external potential
-for the outer disc, this means that the spiral loses
-angular momentum by launching density waves with positive angular momentum at
-the OLR, and by applying a positive co-rotation torque on the disc. 
-In principle, this interaction is destabilising because the
-one-armed spiral has negative angular momentum
-\citep{lin11b}. This effect is small in our simulations, but it  
-could be significant in other parameter regimes. %Furthermore, this
-%non-local interaction may be important for the long-term evolution of
-%the outer disc, since it gains angular momentum from the
-
-%also transport ang mom outwards
-
-{\bf
-  \subsection{Relevance to protoplanetary discs}
-
-}

### 2015-01-07 paper/paper.tex
major revision in response to report

@@
  Because destabilisation by the imposed temperature gradient does 
- not explicitly depend on self-gravity, we suggest this mechanism may also
+ not explicitly require disc self-gravity, we suggest this mechanism may also
  affect low-frequency one-armed oscillations in non-self-gravitating

### 2015-01-08 paper/discussion.tex
updating...

@@
 \section{Discussion}\label{discussions} 
-   % specific to our numerical 
-   % simulations%  and may or may not be relevant to generic low-frequency
-   %   % modes destabilised by a forced temperature gradient.  
-   %
+{\bf
+  We discuss below several issues related to self-gravitating
+  discs in the context of our numerical simulations. However, it is
+  important to keep in mind that, strictly speaking, 
+  the growth of the one-armed spiral in our simulations is
+  \emph{not} a gravitational instability in the sense that
+  destabilisation is through the background torque associated with a
+  forced temperature gradient, and not by self-gravitational torques.     
+}
 
@@
   \subsubsection{Thermodynamic requirements}  
-
   % talk about additional simulations at high Q
-
   %issues with fragmentation 
@@
   disc temperature instantly returns to its initial value when 
-  perturbed.  A short cooling timescale $t_c$ can occur in the outer
+  perturbed. The background torque is generally non-zero if the 
+  resulting temperature profile has a non-zero radial gradient.
+
+  A short cooling timescale $t_c$ can occur in the outer
   parts of protoplanetary discs 
-  \citep{rafikov05,clarke09,rice09,cossins10}.  %rafikov07 (convection)
-  However, if a disc with 
-  $Q\simeq 1$ is cooled (towards zero temperature) on a timescale
-  $t_c\ll\Omega_k^{-1}$, it will fragment
-  \citep{gammie01,paarkedooper12}.
+  \citep{rafikov05,clarke09,rice09,cossins10b}.  %rafikov07 (convection)
+  However, if a disc with $Q\simeq 1$ is cooled (towards zero
+  temperature) on a timescale $t_c\lesssim\Omega_k^{-1}$, it will
+  fragment following gravitational instability
+  \citep{gammie01,rice05,paardekooper12}.
 
-  %internal heating?
-  To avoid fragmentation, the disc must be heated to
-  maintain $Q$ above the threshold for fragmentation \citep[about
-  $Q\lesssim 1.4$ for isothermal discs][]{durisen07}. This may be
-  possible in the outer parts of protoplanetary discs due to heating
-  by irradiation \citep{rafikov07,rice11}.
-
-  % Numerical simulations performed by
-  % \cite{rice11} indicate external irradiation can stablise
-  % the disc against fragmentation.
-
-  %It is important to note, however,
-  %that these studies do not consider the possible effect of global
-  %structures present in our numerical simulations.
+  Fragmentation can be avoided if the disc is heated to maintain 
+  $Q>Q_c$, the threshold for fragmentation \citep[$Q_c\simeq
+  1.4$ for isothermal discs,][]{mayer04}. This may be 
+  possible in the outer parts of protoplanetary discs due to
+  stellar irradiation \citep{rafikov07}. Sufficiently strong
+  external irradiation is expected to suppress the linear gravitational
+  instabilities altogether \citep{rice11}.
 
@@
 
-%where does the structure come from? (mass built up, planet gap)
+  % We note, however, that there is no obvious reason why the background
+  % torque and gravitational instability/fragmentation cannot coexist.
+
+  % It is important to note, however,
+  % that these studies do not consider the possible effect of global
+  % structures present in our numerical simulations.
+
+  % When $Q\simeq 1$ and the cooling timescale
+  % $t_c\gtrsim\Omega_k^{-1}$, the disc reaches a
+  % quasi-steady in which cooling is balanced by heating due to
+  % gravito-turbulence \citep{lodato04}. This would, in general, produce
+  % a global radial temperature gradient.
+
+  % There
+  % is no real way to drive a long-lived quasi-steady, non-fragmenting,
+  % self-gravitating state because this is only possible if the system
+  % can remain quasi-stable though balancing heating and cooling (see,
+  % for example, Lodato & Rice 2004).
+
+  % where does the structure come from? (mass built up, planet gap, infall)
 
   \subsubsection{Radial disc structure}
-  In our numerical simulations the 
+  In our simulations the 
   $m=1$ spiral is confined between two $Q$-barriers, where real solutions to the local
@@
   %in fall/mass loading? 
-  The surface density bump in our initial condition might represent,
-  for example, mass built-up in a 
+  The surface density bump adopted in our numerical simulations might
+  represent 
   `dead zone' in a protoplanetary
@@
   \citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13,bae14}.
-  However, the thermodynamic conditions in a dead zone 
-  %internal heating and cooling can balance such that a global radial
-  %temperature gradient is maintained - then background torque may
-  %exist.
 
@@
   rises rapidly towards the gap centre since it is a region of low
-  surface density, which may satisfy Eq. \ref{qb_cond}. Then the inner
+  surface density, which can satisfy Eq. \ref{qb_cond}. Then the inner
   edge of our bump
@@
 
-% The simulations themselves consider an isothermal or locally isothermal equation of state only.
-Gammie (2001) and other work since have shown that the evolution of self-gravitating discs depends both on the stability criterion and on the balance between heating and cooling.
-If the cooling rate is fast, it is likely to fragment, otherwise it will settle into a quasi-steady, self-gravitating state.
+% The simulations themselves consider an isothermal or locally
+% isothermal equation of state only.  Gammie (2001) and other work
+% since have shown that the evolution of self-gravitating discs
+% depends both on the stability criterion and on the balance between
+% heating and cooling.  If the cooling rate is fast, it is likely to
+% fragment, otherwise it will settle into a quasi-steady,
+% self-gravitating state.
 
-% When considering isothermal simulations, the likely evolution is one of two things.
-It's either initially so unstable that the disc quickly fragments, or it's initially stable enough that any initial perturbations simply drives the system towards being even more stable and the instability quickly weakens through rapid mass transfer (see the work by Laughlin & Bodenheimer 1994/1996).
-There is no real way to drive a long-lived quasi-steady, non-fragmenting, self-gravitating state because this is only possible if the system can remain quasi-stable though balancing heating and cooling (see, for example, Lodato & Rice 2004).
+% When considering isothermal simulations, the likely evolution is one
+% of two things.  It's either initially so unstable that the disc
+% quickly fragments, or it's initially stable enough that any initial
+% perturbations simply drives the system towards being even more
+% stable and the instability quickly weakens through rapid mass
+% transfer (see the work by Laughlin & Bodenheimer 1994/1996).  There
+% is no real way to drive a long-lived quasi-steady, non-fragmenting,
+% self-gravitating state because this is only possible if the system
+% can remain quasi-stable though balancing heating and cooling (see,
+% for example, Lodato & Rice 2004).
 
-% In fact, in the conclusions it is suggested that the conditions used in the simulation would require a cooling < 0.1 \Omega^{-1} which, based on the work of Gammie 2001, would quickly lead to fragmentation.
-The reason it doesn't in the work considered here is that the isothermal assumption assumes that there is some equivalently fast heating mechanism but, in the work here, this is not provided by the instability itself.
-If it were, the simulations would become very quickly unstable.
-Also, if one looks at the work of Clarke (2009), Rafikov (2005), and Rice & Armitage (2009) the only region of a realistic disc where such rapid cooling is possible is in the outer disc, but such regions are either very gravitationally unstable and fragment, or quickly stablised by an external heating source.
-I guess it is possible that there could be a scenario where this exactly produces the conditions assumed here, but that would seem rather fine-tuned and it would be more likely that an external heating source would simply stabilise the disc (unless the disc were extremely massive and could remain unstable even at reasonably high disc temperatures).
+% In fact, in the conclusions it is suggested that the conditions used
+% in the simulation would require a cooling < 0.1 \Omega^{-1} which,
+% based on the work of Gammie 2001, would quickly lead to
+% fragmentation.  The reason it doesn't in the work considered here is
+% that the isothermal assumption assumes that there is some
+% equivalently fast heating mechanism but, in the work here, this is
+% not provided by the instability itself.  If it were, the simulations
+% would become very quickly unstable.  Also, if one looks at the work
+% of Clarke (2009), Rafikov (2005), and Rice & Armitage (2009) the
+% only region of a realistic disc where such rapid cooling is possible
+% is in the outer disc, but such regions are either very
+% gravitationally unstable and fragment, or quickly stablised by an
+% external heating source.  I guess it is possible that there could be
+% a scenario where this exactly produces the conditions assumed here,
+% but that would seem rather fine-tuned and it would be more likely
+% that an external heating source would simply stabilise the disc
+% (unless the disc were extremely massive and could remain unstable
+% even at reasonably high disc temperatures).
 
@@
 In Appendix \ref{disc-planet} we estimate the magnitude of this effect
-  using basic results from disc-planet theory \citep[see, e.g.][and
-  references therein]{papaloizou07}.
-
-One possible effect of self-gravity is to allow the one-armed spiral  
-to act as an external potential for the wider disc. This is
-analogous to disc-satellite interaction 
-\citep{goldreich79}. Here, we estimate the magnitude of this effect
   using basic results from disc-planet theory \citep[see, e.g.][and

### 2015-01-08 paper/discussion.tex
updating...

@@
   possible in the outer parts of protoplanetary discs due to
-  stellar irradiation \citep{rafikov07}. Sufficiently strong
+  stellar irradiation \citep{rafikov09,zhu12}. Sufficiently strong
   external irradiation is expected to suppress the linear gravitational
@@
   %chiang and goldreich 
-  The background torque may therefore exist in the outer
+  The background torque may thus exist in the outer
   parts of protoplanetary discs that are strongly irradiated, such
@@
   gradient. Of course, if external irradiation sets a strictly
-  isothermal outer disc, then the background torque vanishes.
+  isothermal outer disc \citep[e.g.][]{boley09}, then the background
+  torque vanishes.
 
@@
   dispersion relation is possible (Eq. \ref{wavenumber}). The
-  existence of such a cavity is a result of the adopted disc
-  structure. Thus, in our disc models the main role of disc structure
-  and self-gravity is to 
-  allow a local $m=1$ mode to be set up, which is then 
-  destabilised by the background torque density.
+  existence of such a cavity is a result of the adopted surface
+  density bump (Eq. \ref{sig_bump}).  
+  Thus, in our disc models the main role of disc structure 
+  and self-gravity is to allow a local $m=1$ mode to be set up, which is then 
+  destabilised by the background torque.
 
-  In order to confine an $m=1$ spiral between two $Q$-barriers, we
+  In order to confine an $m=1$ mode between two $Q$-barriers, we
   should have $Q^2(1-\nu^2)=1$ at two radii. Assuming 
-  Keplerian rotation and a slow pattern speed $\Omega_p\ll\Omega$, this amounts to
+  Keplerian rotation and a slow pattern speed $\Omega_p\ll\Omega$,
+  this amounts to 
   \begin{align}\label{qb_cond}
@@
   \end{align}
-  Then it may be possible to have two $Q$-barriers when the $Q$ profile
-  of the disc rises more rapidly (decays more slowly) than $R^{-3/2}$
+  Then two $Q$-barriers may exist when the $Q$ profile
+  rises more rapidly (decays more slowly) than $R^{-3/2}$
   for decreasing (increasing) $R$.
@@
   The surface density bump adopted in our numerical simulations might
-  represent 
-  `dead zone' in a protoplanetary
-  disc, where the magneto-rotational 
+  represent, for example, mass built-up in a 
+  `dead zone' in protoplanetary 
+  discs, where the magneto-rotational 
   instability is ineffective for angular momentum transport
   \citep{gammie96,turner08,landry13} and therefore has a reduced mass
-  accretion rate. The dead zone may become 
+  accretion rate. The dead zone becomes 
   self-gravitating with sufficient mass built-up 
-  \citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13,bae14}.
+  \citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13}.
+
+%  There is some attempt to do so when discussing dead zones, but I
+%  doubt that the conditions there are actually suitable as such a
+%  region would either be stable, or cool very slowly and be
+%  susceptible to the growth of long-lived, high-m-mode spiral
+%  structure.
 
@@
   \subsection{Motion of the central star} 
+  For the simulations presented in this paper, the indirect potential
+associated with the non-inertial reference frame is purposefully 
+neglected to avoid complications arising from the motion of the
+central star. For the adopted disc masses $M_d\lesssim 0.1 M_*$, this
+effect is not expected to be significant
+\citep{adams89,shu90,michael10}. Indeed, simulations including the indirect
+potential, carried out in the early stages of this project, produced
+similar results.
 

### 2015-01-08 paper/model.tex
updating...

@@
  disc with constant $H/R$.
-
-For the simulations presented in this paper, the indirect potential
-associated with the non-inertial reference frame is purposefully 
-neglected to avoid complications arising from the motion of the
-central star. For the adopted disc masses $M_d\lesssim 0.1 M_*$, this
-effect is not expected to be significant
-\citep{adams89,shu90,michael10}. Indeed, simulations including the indirect
-potential, carried out in the early stages of this project, produced
-similar results.
 
@@
 smooth disc. We choose 
-\begin{align}
+\begin{align}\label{sig_bump}
   &B(R) = f_1(R)\times f_2(R),\\
@@
 where $\Delta_{1,2} = \delta R \times H(R_{1,2})$.
-
-% {\bf The bump function mimics mass built-up in $R\in[R_1,R_2]$.
-
-% }
-
-% In the case of protoplanetary discs, which have complex structure
-% \citep{armitage10}, GI can also occur in `dead
-% zones' --- sub-portions of the disc where the magneto-rotational
-% instability is ineffective for angular momentum transport
-% \citep{gammie96,turner08,landry13} and therefore has a reduced mass
-% accretion rate. Dead zones may then become
-% self-gravitating with sufficient mass built-up
-% \citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13,bae14}.
-
-%This might represent, for example, mass  
- % built-up in a circumstellar or protoplanetary
- % disc at the outer edge of a `dead zone', in which there
-  %is inefficient magneto-hydrodynamic turbulence to enable accretion.
 

### 2015-01-09 paper/appendix.tex
very rough first revised version

@@
 \end{align}
+These estimates are much smaller than that due to the
+imposed temperature gradient as measured in the simulations
+($\gamma\sim∼0.1\Omega_p$). We conclude that for our disc models, 
+the Lindblad and co-rotation resonances have negligible effects
+on the growth of the one-armed spiral in the inner disc.
+
+The torque exerted on the disc at the OLR by an external potential is
+positive, while that at co-rotation depends 
+on the gradient of potential vorticity there \citep{goldreich79}. For
+our disc models with surface density $\Sigma\propto R^{-2}$ in the
+outer disc, this co-rotation torque is positive. 
+%Interpreting the one-armed spiral as an external po-
+%tential for the outer disc, 
+This means that the one-armed spiral loses
+angular momentum by launching density waves with positive angular
+momentum at the OLR, and by applying a 
+positive co-rotation torque on the disc. In principle, this
+interaction is destabilising because the one-armed spiral has 
+negative angular momentum \citep{lin11b}. This
+effect is small in our simulations, but it could be significant
+in other parameter regimes.

### 2015-01-09 paper/discussion.tex
very rough first revised version

@@
   forced temperature gradient, and not by self-gravitational torques.     
+}
+
+{\bf
+  \subsection{Motion of the central star} 
+  In our models we have purposefully neglected the indirect potential
+  associated with a non-inertial reference frame to avoid
+  complications arising from the motion of the central star.  
+  It has been established that such motion, in response to an $m=1$
+  disturbance in the disc, can render the disturbance
+  unstable in a barotropic disc (so the background torque is
+  ineffective).
+
+  However, for the disc masses in our models, $M_d\lesssim 
+  0.1 M_*$, this effect is not expected to be significant
+  \citep{adams89,shu90,michael10}. Indeed, simulations including the
+  indirect potential, carried out in the early stages of this project,
+  produced similar results.
+
+  \subsection{Role of Lindblad and co-rotation torques}
+  One effect of self-gravity is to allow the one-armed spiral,
+  confined to $R\sim R_0$, to act as an external potential for the
+  exterior disc in $R\gtrsim R_0$. This is 
+  analogous to disc-satellite interaction 
+  \citep{goldreich79}, where the embedded satellite exerts a torque on
+  the disc at Lindblad and co-rotation resonances.
+
+  In Appendix \ref{disc-planet} we estimate the magnitude of this effect
+  using basic results from disc-planet theory \citep[see, e.g.][and
+  references therein]{papaloizou07}. There, we find that the angular
+  momentum exchange between the one-armed spiral and the exterior disc
+  is insignificant compared to the background torque.
+
+  We confirmed this with additional FARGO simulations that exclude the
+  co-rotation and outer Lindblad resonances (OLR) by reducing radial domain
+  size, setting $R_\mathrm{max} = 3 R_0$. This model still developed
+  the one-armed spiral. 
 }
@@
   dispersion relation is possible (Eq. \ref{wavenumber}). The
-  existence of such a cavity is a result of the adopted surface
+  existence of such a cavity is a result of the initial surface
   density bump (Eq. \ref{sig_bump}).  
@@
 
-  %in fall/mass loading? 
-  The surface density bump adopted in our numerical simulations might
-  represent, for example, mass built-up in a 
-  `dead zone' in protoplanetary 
-  discs, where the magneto-rotational 
+  A surface density bump can develop in 
+  `dead zones' of protoplanetary 
+  discs, where there is reduced mass accretion because the magneto-rotational 
   instability is ineffective for angular momentum transport 
-  \citep{gammie96,turner08,landry13} and therefore has a reduced mass
-  accretion rate. The dead zone becomes 
+  \citep{gammie96,turner08,landry13}. The dead zone becomes 
   self-gravitating with sufficient mass built-up 
   \citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13}.
+
+  However, conditions in a dead zone may not be suitable for 
+  sustaining a background torque because it may not cool/heat fast enough
+  to maintain a fixed temperature profile. Recently, 
+  \cite{bae14} presented sophisticated numerical models of
+  dead zones, including a range of heating and 
+  cooling processes, which show that dead zones developed large-scale
+  (genuine) gravitational instabilities with multiple spiral
+  arms. Although this does not prove absence of the background torque,
+  it is probably insignificant compared to gravitational torques.
 
@@
 
-  %particularly simple functional representation of the bump. a
-  %realistic bump may be different
-
-  Another possibility is a gap opened by a giant planet. In that case $Q$
+  Another possibility is a gap opened by an embedded planet. In that case $Q$
   rises rapidly towards the gap centre since it is a region of low
-  surface density, which can satisfy Eq. \ref{qb_cond}. Then the inner
-  edge of our bump
-  function mimics the outer gap edge. The outer gap edge is then a
-  possible site for the growth of a low-frequency one-armed spiral
-  through the background torque. 
- %or gap opened by a fragment
-
+  surface density. This can satisfy Eq. \ref{qb_cond}. Then the inner
+  edge of our bump function mimics the outer gap edge. The outer gap
+  edge is then a potential site for the growth of a low-frequency
+  one-armed spiral through the background torque. However, the
+  locally isothermal requirement limits this process to the
+  outer disc.   
 }
@@
 % even at reasonably high disc temperatures).
-
-{\bf
-  \subsection{Motion of the central star} 
-  For the simulations presented in this paper, the indirect potential
-associated with the non-inertial reference frame is purposefully 
-neglected to avoid complications arising from the motion of the
-central star. For the adopted disc masses $M_d\lesssim 0.1 M_*$, this
-effect is not expected to be significant
-\citep{adams89,shu90,michael10}. Indeed, simulations including the indirect
-potential, carried out in the early stages of this project, produced
-similar results.
-
-  \subsection{Role of Lindblad and co-rotation torques}
-  One possible effect of self-gravity is to allow the one-armed spiral  
-  to act as an external potential for the exterior disc. This is
-  analogous to disc-satellite interaction 
-  \citep{goldreich79}, where the embedded satellite exerts a torque on
-  the disc at
-
-In Appendix \ref{disc-planet} we estimate the magnitude of this effect
-  using basic results from disc-planet theory \citep[see, e.g.][and
-  references therein]{papaloizou07}.
-
-}
 

### 2015-01-09 paper/paper.tex
updating...
very rough first revised version

@@
  non-axisymmetric waves in astrophysical discs with 
-  an imposed radial temperature gradient.  We use linear density wave
-  theory 
+ an imposed radial temperature gradient. This might apply, for
+ example, to the outer parts of protoplanetary discs. We use linear
+ density wave theory 
  to show that non-axisymmetric perturbations generally do not conserve
@@
  code and in 3D with  the ZEUS-MP and PLUTO codes.  We consider
-  radially structured discs with a self-gravitating sub-portion which remains
+ radially structured discs with a self-gravitating region which remains
  stable in the absence of a temperature gradient. However,
@@
 % but also natural in GI.
-However, spiral arms are also associated with global gravitational 
-instability (GI) of differentially rotating discs
+Spiral arms are also {\bf characteristic of} global gravitational 
+instability (GI) {\bf in} differentially rotating discs
 \citep{goldreich65,laughlin96b,laughlin98,nelson98,lodato05,forgan11}. Large-scale 
 spiral arms can provide significant angular momentum transport
-necessary for mass accretion \citep{lynden-bell72, papaloizou91,balbus99,lodato04}, and
-spiral structures due to GI are potentially observable with the Atacama Large 
-Millimeter/sub-millimeter Array \citep{cossins10,dipierror14}.
-
-% when/where does it happen? formation stage/dead zones (special for ppd)
-GI can be expected in the earliest stages of 
-circumstellar disc formation
+necessary for mass accretion \citep{lynden-bell72,
+  papaloizou91,balbus99,lodato04}, and spiral structures due to GI are
+potentially observable with the Atacama Large
+Millimeter/sub-millimeter Array \citep{cossins10,dipierror14}. GI can
+be expected in the earliest stages of circumstellar disc formation 
 \citep{kratter10b,inutsuka10,tsukamoto13}, and may be possible in the
 outer parts of the disc \citep{rafikov05,matzner05,kimura12}.
+
+Single-arm spirals, or eccentric modes, corresponding to perturbations 
+with azimuthal wavenumber $m=1$, have received interest in
+the context of protoplanetary discs because of their global nature
+\citep{adams89,heemskerk92,laughlin96,tremaine01,papaloizou02,hopkins10}. 
+In the `SLING' mechanism proposed by \cite{shu90}, the $m=1$ {\bf
+  gravitational spiral} instability is attributed to motion of the
+central star induced by the one-armed perturbation and requires a
+massive disc \citep[the former may have observable consequences,
+][]{michael10}.
+
+% The aforementioned analytical studies have adopted
+% barotropic, smooth disc models.
+% In this work, we identify a mechanism to generate one-
+% armed spiral patterns that is related to a non-barotropic
+% equation of state and disc structure.
+% We find that if the disc temperature or sound-speed is a
+% prescribed function of position, then one-armed spiral modes
+% can grow inside a cavity that results from the adopted disc
+% structure.
+
+%The aforementioned analytical studies have adopted barotropic, smooth
+%disc models.
+
+{\bf
+  In this work we identify a new mechanism that leads to the growth of
+  one-armed spirals in astrophysical discs. We show
+  that when the disc temperature is prescribed (called locally isothermal
+  discs), the usual statement for the conservation of angular momentum
+  for linear perturbations acquires a source term proportional to the
+  temperature gradient. This permits angular momenum exchange between
+  linear perturbations and the background disc. This `background
+  torque' may act to destabilise low-frequency non-axisymmetric
+  trailing waves when the disc temperature decrease outwards.       
+}
+
+{\bf We employ direct hydrodynamic simulations using three different
+  grid-based codes to demonstrate how this `background torque' 
+  can lead to the growth of one-armed spirals in radially structured, 
+  self-gravitating discs. This is despite the fact that our disc
+  models do not meet the requirements for the `SLING'
+  mechanism. Although our numerical simulations consider
+  self-gravitating discs, this `background torque' is generic for
+  locally isothermal discs and its existence does not depend
+  explicitly on disc self-gravity. Thus, the destabilisation effect we
+  describe should also be applicable to non-self-gravitating discs.  
+}
+
+This paper is organised as follows. We describe the system of interest
+and list the governing equations in 
+\S\ref{model}. {\bf In \S\ref{wkb} we use linear theory  to show
+  how a fixed temperature profile 
+  can destabilise non-axisymmetric waves in discs.} \S\ref{methods} describes
+the {\bf numerical setup and} hydrodynamic codes {\bf we use to 
+  demonstrate the growth of one-armed spirals due to an imposed
+  temperature gradient.} Our {\bf
+  simulation} results are presented in  \S\ref{results2d} for two-
+dimensional (2D) {\bf
+  discs} and \S\ref{results3d} for three-dimensional (3D) {\bf discs}, and we further
+discuss them in \S\ref{discussions}. We summarise in \S\ref{summary} 
+with a discussion on speculations for future work.
+
+\input{model}
+\input{results}
+\input{results_3d}
+\input{discussion}
+\input{summary}
+
+{\bf
+\section*{Acknowledgments}
+This material is based upon work supported by the National Science Foundation under Grant No. 1228509.
+}
+\bibliographystyle{mn2e}
+\bibliography{ref}
+
+\appendix
+\input{appendix}
+
+\end{document}
 
@@
 %stellar wobble from indiana group
-Single-arm spirals, or eccentric modes, corresponding to perturbations 
-with azimuthal wavenumber $m=1$, have received interest in
-the context of protoplanetary discs because of their global nature
-\citep{adams89,heemskerk92,laughlin96,tremaine01,papaloizou02,hopkins10}. 
-In the `SLING' mechanism proposed by \cite{shu90}, the $m=1$ spiral
-instability is attributed to motion of the central star induced by the
-one-armed perturbation and requires a massive disc \citep[the former may
-have observable consequences, ][]{michael10}. However, our numerical
-models do not have these features.
-
-%these are self-gravitating instabilities
-
-The aforementioned analytical studies have adopted barotropic, smooth disc models. 
-In this work, we identify a mechanism to generate one-armed
-spiral patterns that is related to a non-barotropic equation of
-state and disc structure. We find that if the disc temperature or
-sound-speed is a prescribed function of position, then 
-one-armed spiral modes can grow inside a cavity that results from the
-adopted disc structure.
-
-Our study is based on direct hydrodynamic simulations using three
-different grid-based codes, both in two-dimensions (2D) and
-three-dimensions (3D). We also apply linear
-theory to interpret simulation results. We find that, when the disc
-temperature is prescribed, the usual statement for the conservation of
-angular momentum for linear perturbations acquires a source term
-proportional to the temperature gradient, which may act to destabilise
-$m=1$ perturbations. Although we work with self-gravitating discs,
-this source term does not depend explicitly on disc self-gravity. Thus, the
-destabilisation effect we describe should also be applicable to
-non-self-gravitating discs.
-
-This paper is organised as follows. In \S\ref{model} we list the
-governing equations and describe the disc equilibria under
-consideration. We briefly review basic results from linear theory in
-\S\ref{wkb} to aid our interpretation of numerical simulations. 
-\S\ref{methods} describes the hydrodynamic codes
-employed in this work and several diagnostic measures. Our main results are presented in 
-\S\ref{results2d} for 2D simulations and \S\ref{results3d} for 3D
-simulations, and we further analyse them in \S\ref{discussions}. 
-We summarise in \S\ref{summary}, with a discussion on
-speculations for future work.
-
-\input{model}
-\input{results}
-\input{results_3d}
-\input{discussion}
-\input{summary}
-
-%\section*{Acknowledgments}
-
-\bibliographystyle{mn2e}
-\bibliography{ref}
-
-\appendix
-\input{appendix}
-
-\end{document}

### 2015-01-09 paper/summary.tex
major revision in response to report
very rough first revised version

@@
+ % Strictly speaking, the growth of the one-armed spiral
+ %   observed in our simulations is \emph{not} a gravitational instability
+ %   because destabilisation is through the forced temperature gradient
+ %   and not disc self-gravity.  Nevertheless, we discuss below several issues 
+ %   related to self-gravitating discs.
+
 \section{Summary and conclusions}\label{summary}
-In this paper, we have carried out direct numerical hydrodynamic
-simulations of radially structured, self-gravitating locally
+{\bf
+  In this paper, we have described a destabilising  
+  effect of adopting a fixed temperature profile to model   
+  astrophysical discs. By applying angular momentum conservation
+  within linear theory, we showed that a forced temperature gradient 
+  introduces a torque on linear perturbations. We call this the
+  background torque because it represents an exchange of angular
+  momentum between the background disc and the perturbations. This 
+  offers a previously unexplored pathway to instability in locally
   isothermal discs.  
-We find such systems can be unstable to low-frequency perturbations
+}
+
+In the local approximation, we showed that this background torque is
+negative for {\bf non-axisymmetric} trailing waves in discs with a fixed temperature or
+sound-speed profile that decrease outwards. {\bf The background torque} enforces  
+low-frequency {\bf non-axisymmetric} modes because they are associated
+with negative angular momentum.
+
+{\bf 
+  We demonstrated the destabilising effect of the background torque by
+  carrying out direct numerical hydrodynamic simulations of 
+  locally isothermal discs with a self-gravitating surface density
+  bump. 
+}
+We find such systems {\bf are} unstable to low-frequency perturbations
 with azimuthal wavenumber $m=1$, which leads to the development of an one-armed 
@@
 growth time of $O(10)$ orbits.
-%While this is slow
-%compared to the local rotation, it is dynamical at co-rotation.
 
@@
 PLUTO in 3D --- to show that the growth of     
-one-armed spirals in our disc models is due to the imposed temperature 
-profile and disc structure. Growth rates increased 
+one-armed spirals in our disc model is due to the imposed
+temperature {\bf gradient: growth rates increased linearly}  
 with the magnitude of the imposed temperature gradient, and one-armed
-spirals did not develop in strictly isothermal simulations. The spiral
-is confined between two $Q$-barriers owing to a surface density bump. 
-We find the instability behaves similarly in 2D and 3D. However, in 3D
-the spiral disturbance can become radially global away from 
+spirals did not develop in strictly isothermal simulations. {\bf This 
+  one-armed spiral instability can be interpreted as an initially 
+  neutral, tightly-wound $m=1$ mode being destabilised by the 
+  background torque.} The spiral
+is mostly confined between two $Q$-barriers surface density bump. 
+We find the instability behaves similarly in 2D and 3D {\bf, but in 3D}
+the spiral disturbance becomes more radially global away from    
 the midplane.
 
-We applied angular momentum conservation within linear theory to
-explain our numerical results. We find a   
-forced temperature gradient introduces a torque on linear 
-perturbations. We called this the background torque because it
-represents an exchange of angular momentum between the background disc
-and the perturbations. This offers a previously unexplored pathway to
-instability.
+\subsection{Speculations and future work}
+% numerical simulations to include cooling/heating
+% A locally isothermal equation of state represents the limit of
+% infinitely short cooling (and heating) timescales, so the disc
+% temperature instantly returns to its initial value when perturbed.
 
-In the local approximation, we showed that this background torque is
-negative for trailing spirals in discs with a fixed temperature or
-sound-speed profile that decrease outwards. This torque enforces  
-low-frequency $m=1$ modes because they are associated with negative
-angular momentum. We therefore interpret the one-armed spiral instability
-observed in our simulations as an initially neutral, tightly-wound
-$m=1$ mode being destabilised by the imposed temperature gradient.
+{\bf There are several issues that remain to be addressed in future
+  works:}
 
-\subsection{Speculations and future work}
-%numerical simulations to include cooling/heating
-A locally isothermal equation of state represents the limit of
-infinitely short cooling (and heating) timescales, so the disc
-temperature instantly returns to its initial value when perturbed.  
-This assumption can be relaxed by including an energy equation with
-thermal relaxation. Preliminary FARGO simulations
-indicate a thermal relaxation timescale $t_c < 0.1\Omega_k^{-1}$ is
-needed for the one-armed spiral to develop. However, this value may
-depend on disc parameters, and will be explored in a follow-up  
-study.
+{\bf \emph{Thermal relaxation.} The locally isothermal assumption 
+} can be relaxed by including an energy equation with
+{\bf a source term that restores the disc temperature over a
+  characteristic timescale $t_\mathrm{relax}$}. Preliminary FARGO
+simulations indicate a thermal relaxation timescale $t_\mathrm{relax} <
+0.1\Omega_k^{-1}$ is needed for the one-armed spiral to
+develop. However, this value may {\bf be model-dependent} and will be
+explored in a follow-up study. {\bf For example,
+a longer $t_\mathrm{relax}$ may be permitted with larger temperature
+gradients.}
 
@@
 % should not get fragmentation (CR away from spiral)
-In the deeply non-linear regime, the one-armed spiral may 
+
+{\bf \emph{Non-linear evolution}.} In the deeply non-linear regime, the
+one-armed spiral may  
 shock and deposit negative angular momentum onto 
@@
 gain of positive angular momentum through shock dissipation. We remark  
-that fragmentation is unlikely because the co-rotation radius 
-is outside the bulk of the spiral arm, in the non-self-gravitating
+that fragmentation {\bf of a low-frequency spiral} is unlikely
+because the co-rotation radius is outside the bulk of the spiral arm, in the non-self-gravitating
 portion of the disc \citep{durisen08}. In order
@@
 
-In our models and interpretation, self-gravity coupled with the disc 
+{\bf The background torque in other contexts.}  
+{\bf In our simulations}, self-gravity coupled with the disc 
 structure permits a neutral one-armed mode, which is then
-destabilised by the temperature gradient. Other origins of the neutral
-one-armed spiral should be investigated. One possibility is in Be star discs
-\citep{rivinius13}, for which 
-one-armed oscillations may explain long-timescale variations in their
-emission lines \citep[see e.g.][and references
+destabilised by the {\bf fixed} temperature gradient.  
+Other origins of the neutral one-armed spiral should be
+investigated. One possibility is in Be star discs \citep{rivinius13},
+for which one-armed oscillations may explain long-timescale variations
+in their emission lines \citep[see e.g.][and references 
 therein]{okasaki97,papaloizou06c,ogilvie08}. These studies employ
@@
 %vertical shear instability?
-
 %linear theory: explicit calculations 
@@
 %explicitly solve for linear stability problem. 
-
 %theoretical investigation on effect on other m modes

### 2015-01-09 paper/discussion.tex
second rough draft of revised paper

@@
   complications arising from the motion of the central star.  
-  It has been established that such motion, in response to an $m=1$
-  disturbance in the disc, can render the disturbance
-  unstable in a barotropic disc (so the background torque is
-  ineffective).
-
-  However, for the disc masses in our models, $M_d\lesssim 
-  0.1 M_*$, this effect is not expected to be significant
-  \citep{adams89,shu90,michael10}. Indeed, simulations including the
+  Although it has been established that such motion can destabilise an
+  $m=1$ disturbance in the disc \citep{adams89,shu90,michael10}, 
+  the disc masses in our models ($M_d\lesssim 
+  0.1 M_*$) are not expected to be sufficiently massive for this effect to be
+  significant. Indeed, simulations including the
   indirect potential, carried out in the early stages of this project,
@@
   one-armed spiral through the background torque. However, the
-  locally isothermal requirement limits this process to the
-  outer disc.   
+  locally isothermal requirement would limit this process to the
+  outer disc irradiated by the central star, or that the temperature
+  profile about the gap edge is set by the planet luminosity.  
 }

### 2015-01-09 paper/model.tex
second rough draft of revised paper

@@
   conservation within the framework of linear perturbation theory. 
-  For simplicity, we consider 2D discs, although generalisation to 3D is straight
-  forward.  
+  For simplicity, in this section we consider 2D discs. 
 }
@@
 a complex frequency with $\omega$ being the real frequency, 
-$\gamma$ the growth rate, and $m$ is an integer. Here, we take $m>0$ without
+$\gamma$ the growth rate, and $m$ is an integer. We take $m>0$ without
 loss of generality.  
@@
 {\bf 
-  We demonstrate instability for low-frequency
-  ($\omega\ll m\Omega$) perturbations by explicitly calculating its 
+  We demonstrate instability for low-frequency modes 
+  ($|\omega|\ll m\Omega$) by explicitly showing its 
   angular momentum density 
-  $\jlin$ and the background torque $T_\mathrm{BG}$, and show that
-  they are unstable when a negative radial 
-  temperature gradient is imposed.  
+  $\jlin<0$, and the background torque $T_\mathrm{BG}<0$ for
+  appropriate perturbations and radial temperature gradients. 
+%, and show that
+%  low-frequency modes can be unstable when a radial 
+%  temperature gradient is imposed.  
 }
@@
 (long) waves, and $k>0$ ($k<0$) correspond to trailing (leading)
-spirals.
+waves.
 
@@
 
-This torque density is negative for trailing waves in discs with
-a fixed temperature profile decreasing outwards ($k>0$ and $dc_s^2/dR<0$,
-respectively). Note that this conclusion does not rely on the
-low-frequency approximation and holds for any $m$.
+This torque density is negative for trailing waves ($k>0$) in discs with
+a fixed temperature profile decreasing outwards ($dc_s^2/dR<0$). Note
+that this conclusion does not rely on the low-frequency approximation.
 
@@
 where we used $c_s\sim H\Omega\sim hR\Omega$. 
-{\bf It is important to note that, although our description of the local
+{\bf Note the absence of the Toomre parameter $Q$ in
+  Eq. \ref{theoretical_rate}. Although our description of the local 
   approximation accounts for disc self-gravity, mode growth due 

### 2015-01-09 paper/paper.tex
second rough draft of revised paper

@@
  Because destabilisation by the imposed temperature gradient does 
- not explicitly require disc self-gravity, we suggest this mechanism may also
+ not explicitly require disc self-gravity, this mechanism may also
  affect low-frequency one-armed oscillations in non-self-gravitating
- discs. 
+ discs, e.g in Be star discs. 
 }
@@
   self-gravitating discs, this `background torque' is generic for
-  locally isothermal discs and its existence does not depend
-  explicitly on disc self-gravity. Thus, the destabilisation effect we
+  locally isothermal discs and its existence does not require
+  disc self-gravity. Thus, the destabilisation effect we
   describe should also be applicable to non-self-gravitating discs.  
@@
 
-This paper is organised as follows. We describe the system of interest
-and list the governing equations in 
-\S\ref{model}. {\bf In \S\ref{wkb} we use linear theory  to show
+This paper is organised as follows. In \S\ref {model} we describe the system of interest
+and list the governing equations. {\bf In \S\ref{wkb} we use linear theory  to show
   how a fixed temperature profile 
@@
   temperature gradient.} Our {\bf
-  simulation} results are presented in  \S\ref{results2d} for two-
-dimensional (2D) {\bf
-  discs} and \S\ref{results3d} for three-dimensional (3D) {\bf discs}, and we further
+  simulation} results are presented in  \S\ref{results2d} for two-dimensional (2D) {\bf
+  discs} and in \S\ref{results3d} for three-dimensional (3D) {\bf discs}, and we further
 discuss them in \S\ref{discussions}. We summarise in \S\ref{summary} 
@@
 \section*{Acknowledgments}
-This material is based upon work supported by the National Science Foundation under Grant No. 1228509.
+I thank K. Kratter and A. Youdin for valuable discussions, and the
+anonymous referee for comments that significantly improved this paper. All
+computations were performed on the El Gato cluster at the University
+of Arizona. This material is based upon work supported by the National
+Science Foundation under Grant No. 1228509. 
 }

### 2015-01-09 paper/summary.tex
second rough draft of revised paper

@@
 negative for {\bf non-axisymmetric} trailing waves in discs with a fixed temperature or
-sound-speed profile that decrease outwards. {\bf The background torque} enforces  
+sound-speed profile that decrease outwards. {\bf A negative background torque} enforces  
 low-frequency {\bf non-axisymmetric} modes because they are associated
@@
   background torque.} The spiral
-is mostly confined between two $Q$-barriers surface density bump. 
-We find the instability behaves similarly in 2D and 3D {\bf, but in 3D}
+is mostly confined between two $Q$-barriers in the surface density bump. 
+We find the instability behaves similarly in 2D and 3D{\bf, but in 3D}
 the spiral disturbance becomes more radially global away from    
@@
 
+%interaction with turbulence
+
 % non-sg sims, Be stars, eccentric discs
@@
 
-{\bf The background torque in other contexts.}  
-{\bf In our simulations}, self-gravity coupled with the disc 
-structure permits a neutral one-armed mode, which is then
-destabilised by the {\bf fixed} temperature gradient.  
-Other origins of the neutral one-armed spiral should be
-investigated. One possibility is in Be star discs \citep{rivinius13},
+{\bf\emph{Other applications of the background torque.}}     
+{\bf The background torque is a generic feature in  
+  discs for which the temperature is set externally.  It may therefore
+  be relevant in other astrophysical contexts. }
+One possibility is in Be star discs \citep{rivinius13}, 
 for which one-armed oscillations may explain long-timescale variations
 in their emission lines \citep[see e.g.][and references 
-therein]{okasaki97,papaloizou06c,ogilvie08}. These studies employ
-strictly isothermal disc models, so it would be interesting to explore
-the effect of a temperature gradient on the stability of these
+therein]{okasaki97,papaloizou06c,ogilvie08}. {\bf These studies 
+invoke other mechanisms to produce \emph{neutral} one-armed
+oscillations (e.g. rotational deformation of the star), but consider 
+strictly isothermal discs.} It would be interesting to explore
+the effect of a radial temperature gradient on the stability of these
 oscillations.

### 2015-01-10 paper/discussion.tex
regular update (revising)

@@
   \end{align}
-  Then two $Q$-barriers may exist when the $Q$ profile
+  Then two $Q$-barriers may exist when the $Q^2$ profile
   rises more rapidly (decays more slowly) than $R^{-3/2}$

### 2015-01-10 paper/model.tex
regular update (revising)

@@
 \begin{align}
-  T_\mathrm{BG} = \frac{m}{2}\frac{dc_s^2}{dR}k\Sigma |\xi_R|^2. 
+  T_\mathrm{BG} = \frac{m}{2}\frac{dc_s^2}{dR}k\Sigma |\xi_R|^2. \label{baroclinic_torque1}
 \end{align}
@@
 \end{align}
-where we used $c_s\sim H\Omega\sim hR\Omega$. 
-{\bf Note the absence of the Toomre parameter $Q$ in
-  Eq. \ref{theoretical_rate}. Although our description of the local 
-  approximation accounts for disc self-gravity, mode growth due 
-  to the fixed temperature gradient does not depend explicitly on
-  whether or not the disc is self-gravitating, i.e. destabilisation is
-  not due to self-gravity.}
+where we used $c_s\sim H\Omega\sim hR\Omega$. {\bf
+  Eq. \ref{theoretical_rate} suggests that perturbations with small
+  radial length-scales ($kH\gtrsim m$) are most favourable for
+  destabilisation. Taking the local approximation is then
+  appropriate.}
+
+{\bf Note that the derivation of Eq. \ref{baroclinic_torque1} and 
+  Eq. \ref{theoretical_rate} do not 
+  require the disc to be self-gravitating. Thus, destabilisation by the
+  background torque is not directly associated with disc
+  self-gravity. However, in order to evaluate 
+  Eq. \ref{baroclinic_torque1} or Eq. \ref{theoretical_rate} in terms 
+  of disc parameters (as done in \S\ref{fargo_m1}), we need to insert a 
+  value of $k$, which may depend on disc    
+  self-gravity (e.g. from Eq. \ref{wavenumber}).  
+}
 

### 2015-01-10 paper/paper.tex
regular update (revising)

@@
  Because destabilisation by the imposed temperature gradient does 
- not explicitly require disc self-gravity, this mechanism may also
+ not explicitly require disc self-gravity, we suggest this mechanism may also
  affect low-frequency one-armed oscillations in non-self-gravitating
- discs, e.g in Be star discs. 
+ discs. 
 }

### 2015-01-13 paper/model.tex
still revising...

@@
 
-%isolated system 
-% ang mom cons
-% The locally isothermal
-% equation of state allows us to apply standard linear density wave
-% theory for results interpretation. 
-% This allows us to consider
-%an isolated disc.
-
 The 2D disc equations are obtained by replacing $\rho$ with the surface
@@
 \section{Linear density waves}\label{wkb}
-% In this paper we will mainly use non-linear hydrodynamic simulations to investigate the
-% stability of disc equilibria described 
-% above. However, it is useful to  interpret simulation results
-% using linear density theory. We list here the main definitions and
-% results from 2D linear theory, taken from \cite{lin11b} and
-% \cite{shu91}.  
-{\bf We highlight a key feature of locally isothermal discs that
+
+{\bf We describe a key feature of locally isothermal discs that
   enables angular momentum exchange between small disturbances and the
@@
 
-In a linear analysis, one assumes a steady axisymmetric background state, % such as
-% that defined in \S\ref{model},
+In a linear analysis, one assumes a steady axisymmetric background state, 
 which is then perturbed such that
@@
 loss of generality.
-%Since the background is
-%axisymmetric, we identify $\delta\Sigma_m(R)\exp{\left(-\ii\sigma
- %   t\right)} = \Sigma_m$ as defined for diagnostics.
 
@@
   discuss} some
-general properties of the linear perturbations. % that will be relevant
-% for describing simulation results.
+general properties of the linear perturbations.
 
@@
 and arises because we have adopted a locally isothermal equation of
-state. {\bf  We outline the
+state {\bf in the perturbed disk}. {\bf  We outline the
 derivation of $T_\mathrm{BG}$ in Appendix \ref{tbg_deriv}.}
@@
 may grow by exchanging angular momentum with the background
-disc.%  In our disc models with $c_s^2\propto R^{-q}$, we expect the
-% growth rate of such an instability to scale with $q$.
+disc.
 
@@
   appropriate perturbations and radial temperature gradients. 
-%, and show that
-%  low-frequency modes can be unstable when a radial 
-%  temperature gradient is imposed.  
 }
@@
 Given the real frequency $\omega$ or pattern speed $\Omega_p\equiv  
-\omega/m$ of a neutral mode 
+\omega/m$ of a {\bf non-axisymmetric} neutral mode 
 , Eq. \ref{dispersion} can be solved 
@@
 
-The \emph{co-rotation radius} $R_c$ is where the pattern speed matches
+{\bf At the \emph{co-rotation radius} $R_c$} the pattern speed matches
 the fluid rotation,
@@
 momentum.
-%The local approximation becomes less good for global modes,
-%but will nevertheless be useful for results interpretation.
 
@@
   and the background disc due to an imposed temperature gradient}
-%The $m=1$ spiral that emerges in our simulations have low frequency
-%($\Omega_p < \Omega$), grows slowly ($\gamma\ll\Omega_p$) and its
-%spatial form is consistent local theory (\S\ref{fargo_m1}). 
-%{\bf
-%  We first regard a low-frequency, $m=1$ disturbance 
- % as a mode within the framework of local theory, satisfying the
- % % temperature gradient ignored in local theory
- % dispersion relation Eq. \ref{dispersion}. }
-%To demonstrate possible
-%instability, 
 {\bf 
@@
 
-% Note that
-% we will not be assuming a self-gravitating disc until the end of this
-% section. 
-%We begin by recalling the expression for the torque density due to a
-%locally isothermal equation of state,
-%\begin{align*}
- % T_\mathrm{BG} = - \frac{m}{2}\frac{dc_s^2}{dR}\imag\left(\delta\Sigma_m\xi_R^*\right).   
-%\end{align*}
 The Eulerian surface density perturbation is given by
@@
 \ref{horizontal_displacements}.} 
- % {\bf
-% This holds for local modes of all frequencies and can be shown
-% explicitly by evaluating Eq. \ref{ang_mom_def} with 
-% \begin{align}
-%   \xi_\phi = -\frac{2\ii\Omega}{\sbar}\xi_R, 
-% \end{align}
-% consistent with the local approximation as shown in Appendix
-% \ref{horizontal_displacements}. 
-% }
-Then 
-%To further simplify this expression, we need to relate $\xi_R$ and
-%$\xi_\phi$. 
-% \begin{align}
-%   2\gamma \sim
-%   -\frac{dc_s^2}{dR}\frac{1}{k\Sigma^2\Omega}\frac{|\delta\Sigma_1|^2}{|\xi_R
-%     + \ii\xi_\phi|^2}.
-% \end{align}
-% Now, using $\delta\Sigma_m/\Sigma = -\ii k \xi_R - \ii m \xi_\phi/R$
-% in the local approximation, we have
-% \begin{align}
-%   2\gamma \sim -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R +
-%     \xi_\phi/kR|^2}{|\xi_R + \ii \xi_\phi|^2}
-%   \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
-% \end{align}
-% where we used $dc_s^2/dR <0$, $k>0$ and $|kR|\gg 1$ to obtain the
-% inequality. 
-%where we used $dc_s^2/dR <0$ and $k>0$ to obtain the inequality. 
-for the temperature profiles $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ as
+Then for the temperature profiles $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ as
 adopted in our disc models,  
@@
 
-% For perturbations in a self-gravitating disc with characteristic
-% wavenumber $k \sim \pi G \Sigma/c_s^2 \sim \Omega/Q c_s$, as seen in
-% our simulations, we have 
-% \begin{align}\label{theoretical_rate}
-%   \gamma \sim \frac{qh}{2Q}\Omega,
-% \end{align}
-% where we used $c_s/R\sim h\Omega$.
+\section{Numerical simulations}\label{methods}
 
-%For our fiducial disc models with
-%$q=1$, $Q=O(1)$, $h=0.05$, this gives $\gamma/\Omega \sim O(10^{-2})$,
-%consistent with simulation results. Specifically,
-%Eq. \ref{theoretical_rate} indicates the growth rate should increase
-%linearly with $q$ and $h$, as seen in the simulations
-%(Fig. \ref{fargo_varq_growth}---\ref{fargo_varh_growth}).
-
-\section{Numerical simulations}\label{methods}
-%above discussion is generic for low-frequency non-axisymmetric modes,
-%but illustrate with a concrete example
 {\bf
-  We illustrate the destabilising effect of a fixed temperature
+  We demonstrate the destabilising effect of a fixed temperature
   gradient using numerical simulations. The above discussion is
@@
   modes with small $m$ are
-  most favourable for destabilisation by an imposed temperature
+  more favourable for destabilisation by an imposed temperature
   gradient.  
@@
 
-%In 3D, we assume vertical hydrostatic balance
 The 3D disc structure is obtained by assuming vertical hydrostatic
@@
 \subsection{Codes}
-{\bf We use three independent grid-based codes to simulate the aforementioned
+{\bf We use three independent grid-based codes to simulate the above 
 system.} We adopt computational units such that

### 2015-01-13 paper/paper.tex
still revising...

@@
  stable in the absence of a temperature gradient. However,
- when a temperature gradient is imposed we observe exponential growth of an one-armed  
+ when a temperature gradient is imposed we observe exponential growth of a one-armed  
  spiral mode (azimuthal wavenumber $m=1$) with co-rotation radius 
@@
  spiral does not require the movement of the central star, as found in previous studies. 
- Because destabilisation by the imposed temperature gradient does 
+ Because destabilisation by a forced temperature gradient does 
  not explicitly require disc self-gravity, we suggest this mechanism may also
@@
 
-%This might represent, for example, mass  
- % built-up in a circumstellar or protoplanetary
- % disc at the outer edge of a `dead zone', in which there
-  %is inefficient magneto-hydrodynamic turbulence to enable accretion.
-
 \begin{keywords}
@@
 \section{Introduction}\label{intro}
-% Protoplanetary discs are expected to contain complex structures
-% \citep{armitage11}, which directly impact planet formation theories.  
 An exciting development in the study of circumstellar 
@@
 \citep{adams89,heemskerk92,laughlin96,tremaine01,papaloizou02,hopkins10}. 
-In the `SLING' mechanism proposed by \cite{shu90}, the $m=1$ {\bf
-  gravitational spiral} instability is attributed to motion of the
-central star induced by the one-armed perturbation and requires a
+In the `SLING' mechanism proposed by \cite{shu90}, a $m=1$ {\bf
+  gravitational} instability arises from the motion of the
+central star induced by the one-armed perturbation, and requires a
 massive disc \citep[the former may have observable consequences,
 ][]{michael10}.
-
-% The aforementioned analytical studies have adopted
-% barotropic, smooth disc models.
-% In this work, we identify a mechanism to generate one-
-% armed spiral patterns that is related to a non-barotropic
-% equation of state and disc structure.
-% We find that if the disc temperature or sound-speed is a
-% prescribed function of position, then one-armed spiral modes
-% can grow inside a cavity that results from the adopted disc
-% structure.
-
-%The aforementioned analytical studies have adopted barotropic, smooth
-%disc models.
 
@@
   that when the disc temperature is prescribed (called locally isothermal
-  discs), the usual statement for the conservation of angular momentum
+  discs), the usual statement of the conservation of angular momentum
   for linear perturbations acquires a source term proportional to the
@@
   torque' may act to destabilise low-frequency non-axisymmetric
-  trailing waves when the disc temperature decrease outwards.       
+  trailing waves when the disc temperature decreases outwards.       
 }
@@
   mechanism. Although our numerical simulations consider
-  self-gravitating discs, this `background torque' is generic for
+  self-gravitating discs, the `background torque' is generic for
   locally isothermal discs and its existence does not require
@@
 \end{document}
-
-% In the case of protoplanetary discs, which have complex structure
-% \citep{armitage10}, GI can also occur in `dead
-% zones' --- sub-portions of the disc where the magneto-rotational
-% instability is ineffective for angular momentum transport
-% \citep{gammie96,turner08,landry13} and therefore has a reduced mass
-% accretion rate. Dead zones may then become
-% self-gravitating with sufficient mass built-up
-% \citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13,bae14}.
-
-% original motivation for this work (DZGI)
-
-%A wide range of GIs can occur in  gaseous 
-%astrophysical discs 
-%\citep{papaloizou89,christo92,christo93,hadley11,hadley14}.
-
-% It is then
-% of theoretical importance to clarify what type of GI is relevant to 
-% partially self-gravitating discs, such as a protoplanetary disc containing a 
-% massive dead zone. 
-% This work presents a first step towards this goal. Specifically, here we 
-% study a class of seemingly simple disc models which were found to 
-% consistently produce \emph{one-armed} spiral structures, despite not 
-% satisfying the usual requirements to generate them.
-
-%through this work we identify a new method to generate one arm
-%spirals 
-%We therefore dedicate this study to verify their origin.
-
-%brief history of m=1 
-%stellar wobble from indiana group

### 2015-01-13 paper/discussion.tex
almost done with first revision

@@
-%to discuss - fixed central star
-%application to ppd
-%do smooth disc results carry over to structured disks?
-
 \section{Discussion}\label{discussions} 
@@
   discs in the context of our numerical simulations. However, it is
-  important to keep in mind that, strictly speaking, 
+  important to keep in mind that 
   the growth of the one-armed spiral in our simulations is
@@
   destabilisation is through the background torque associated with a
-  forced temperature gradient, and not by self-gravitational torques.     
+  forced temperature gradient, and not by self-gravitational torques\footnote{In fact, additional simulations with $Q_\mathrm{out}=4$ still
+  develops the one-armed spiral, but with a smaller growth rate.}. 
 }
@@
   One effect of self-gravity is to allow the one-armed spiral,
-  confined to $R\sim R_0$, to act as an external potential for the
-  exterior disc in $R\gtrsim R_0$. This is 
+  confined to $R\sim R_0$ in our models, to act as an external potential for the
+  exterior disc in $R>R_0$. This is 
   analogous to disc-satellite interaction 
@@
   co-rotation and outer Lindblad resonances (OLR) by reducing radial domain
-  size, setting $R_\mathrm{max} = 3 R_0$. This model still developed
+  size to $R_\mathrm{max} = 3 R_0$, which still developed 
   the one-armed spiral. 
+  %instability due to local exchange of angular momentum with the
+  %background disc 
+  %CR and LR do not play a role - extended outer disc not essential
+  %(but used here to avoid potential boundary effects) cannot exclude
+  %a priori 
 }
@@
 {\bf
-  \subsection{Applicability to circumstellar discs}
+  \subsection{Applicability to protoplanetary discs}
 
@@
 
-  A locally isothermal equation of state represents the limit
-  of infinitely short cooling (and heating) timescales, so the 
+  A locally isothermal equation of state represents the ideal limit
+  of infinitely short cooling and heating timescales, so the 
   disc temperature instantly returns to its initial value when 
@@
 
-  % There
-  % is no real way to drive a long-lived quasi-steady, non-fragmenting,
-  % self-gravitating state because this is only possible if the system
-  % can remain quasi-stable though balancing heating and cooling (see,
-  % for example, Lodato & Rice 2004).
-
   % where does the structure come from? (mass built up, planet gap, infall)
@@
   rises more rapidly (decays more slowly) than $R^{-3/2}$
-  for decreasing (increasing) $R$.
+  for decreasing (increasing) $R$. Note that 
+  Eq. \ref{qb_cond} does not necessarily require strong self-gravity
+  if $R_c$ is large.
 
@@
 
-%  There is some attempt to do so when discussing dead zones, but I
-%  doubt that the conditions there are actually suitable as such a
-%  region would either be stable, or cool very slowly and be
-%  susceptible to the growth of long-lived, high-m-mode spiral
-%  structure.
-
   Another possibility is a gap opened by an embedded planet. In that case $Q$
@@
 }
-
-% The simulations themselves consider an isothermal or locally
-% isothermal equation of state only.  Gammie (2001) and other work
-% since have shown that the evolution of self-gravitating discs
-% depends both on the stability criterion and on the balance between
-% heating and cooling.  If the cooling rate is fast, it is likely to
-% fragment, otherwise it will settle into a quasi-steady,
-% self-gravitating state.
-
-% When considering isothermal simulations, the likely evolution is one
-% of two things.  It's either initially so unstable that the disc
-% quickly fragments, or it's initially stable enough that any initial
-% perturbations simply drives the system towards being even more
-% stable and the instability quickly weakens through rapid mass
-% transfer (see the work by Laughlin & Bodenheimer 1994/1996).  There
-% is no real way to drive a long-lived quasi-steady, non-fragmenting,
-% self-gravitating state because this is only possible if the system
-% can remain quasi-stable though balancing heating and cooling (see,
-% for example, Lodato & Rice 2004).
-
-% In fact, in the conclusions it is suggested that the conditions used
-% in the simulation would require a cooling < 0.1 \Omega^{-1} which,
-% based on the work of Gammie 2001, would quickly lead to
-% fragmentation.  The reason it doesn't in the work considered here is
-% that the isothermal assumption assumes that there is some
-% equivalently fast heating mechanism but, in the work here, this is
-% not provided by the instability itself.  If it were, the simulations
-% would become very quickly unstable.  Also, if one looks at the work
-% of Clarke (2009), Rafikov (2005), and Rice & Armitage (2009) the
-% only region of a realistic disc where such rapid cooling is possible
-% is in the outer disc, but such regions are either very
-% gravitationally unstable and fragment, or quickly stablised by an
-% external heating source.  I guess it is possible that there could be
-% a scenario where this exactly produces the conditions assumed here,
-% but that would seem rather fine-tuned and it would be more likely
-% that an external heating source would simply stabilise the disc
-% (unless the disc were extremely massive and could remain unstable
-% even at reasonably high disc temperatures).
-
-%Furthermore, this
-%non-local interaction may be important for the long-term evolution of
-%the outer disc, since it gains angular momentum from the
-
-%also transport ang mom outwards
-
-% Having confirmed the $m=1$ spiral instability through numerical
-% simulations, we now use results from the simulations combined with
-% linear theory to construct an explanation for their growth. For
-% simplicity, we return to 2D discs.
-
-% The dominant disturbance in our simulations is an one-armed, $m=1$
-% trailing spiral, and its growth is associated with the imposed temperature
-% gradient. As discussed in \S\ref{global_cons}, in a locally isothermal disc this imposed
-% temperature gradient generally results in a torque being exerted on
-% linear perturbations. An instability is therefore possible if the sign
-% of this torque is the same as that of the angular momentum associated
-% with the linear perturbation.
-
-%associated angular mometum density is the same as that
-%of this
-% According to previous studies \citep{adams89}, our disc models
-% should not support $m=1$ linear spiral instabilities. This is because
-% our discs are not sufficiently massive ($M_d\lesssim 0.1M_*$) and, more
-% importantly, we have surpressed the motion of the central star. The 
-% latter can induce $m=1$ instabilities in massive discs  \citep{shu90}.
-
-%the latter assumes  k=pi\Sigma G /c_s^2 is always the most favoured
-
-% The temperature gradient affects the growth rate through the
-% explicit depdence of the background torque on $q$. The growth 
-% rate is also dependent on the underlying disturbance being
-% considered (i.e. through $k$ and $\bm{\xi}$ in
-% Eq. \ref{theoretical_rate0}). % Of course, the temperature gradient
-% % plays no role on local modes. However, other disc parameters such as
-% % $c_s^2$ may affect the mode that develops. 
-% The dependence on $h$ in 
-% Eq. \ref{theoretical_rate} assumes $k=\pi\Sigma G/c_s^2$ is always the
-% most favourable wavenumber for low-frequency modes.
-
-%By 
-%definition, the global temperature gradient does not affect modes in
-%local theory. 
-%however, changing $h$ may also modify the mode properties (since
-%temperature appears in local theory) in addition
-%to the magnitude of the temperature gradient. Thus, we expect the depedence of
-%the background torque on $q$ to be better described by
-%Eq. \{theoretical_rate}
-
-%by definition, q does not affect local neutral modes
-%Of course, we assumed a neutral mode to obtain expressions
-%for $\jlin$ and $T_\mathrm{BG}$ in the first place, 
-%Recall that in local theory non-axisymmetric modes are stable for real
-%$k$  
-%Eq. \ref{theoretical_rate} should be interpreted
-%as an order-of-magnitude estimate of the perturbative effect of a
-%temperature gradient on local, low-frequency $m=1$ modes.
-
-%neutral modes are not actually neutral)
-%gamma << omega_p << omega 
-%so
-%Eq. \ref{theoretical_rate} is only an order-of-magnitude
-%constraint. Nevertheless, the expected growth rate is in reasonable
-%agreement with numerical results.
-
-%also total torque is integral -> may get cancelations

### 2015-01-13 paper/model.tex
almost done with first revision

@@
 $\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
-$q$ is the imposed temperature gradient (since, for an ideal gas the
-temperature is proportional to $c_s^2$). The vertical disc scale-height is
+$q$ is the imposed temperature gradient since, for an ideal gas the
+temperature is proportional to $c_s^2$. {\bf For convenience we will
+  refer to $c_s^2$ as the disc temperature.} The vertical disc scale-height is
 defined by $H=c_s/\Omega_k$. Thus, a strictly isothermal disc with
@@
 momentum. If it is trailing and $dc_s^2/dR<0$, as is typical
-in astrophysical discs, then $T_\mathrm{BG}<0$, and
+in astrophysical discs, then $T_\mathrm{BG}<0$ and
 the background disc applies a negative torque on the disturbance,
@@
 
-  There are two components to the destabilising mechanism.  First
+  There are two parts to the destabilising mechanism: 
   the disc should support low-frequency modes, which is then
@@
   consider discs with a  
-  radially-structured Toomre $Q$ profile. We use local theory to show that such discs support
-  low-frequency trapped one-armed ($m=1$) modes. 
+  radially-structured Toomre $Q$ profile. We use local theory to show
+  that such discs can trap 
+  low-frequency one-armed ($m=1$) modes.  
   This is convenient because Eq. \ref{theoretical_rate} indicates that
   modes with small $m$ are
-  more favourable for destabilisation by an imposed temperature
-  gradient.  
+  more favourable for destabilisation. 
 }
@@
 
-% poisson kernal integration, softening 
-% where used?
-% BC
 \subsubsection{ZEUS-MP}
@@
 
-%used in lin 2012 
 \subsubsection{PLUTO} 
@@
 that of ZEUS-MP and FARGO.
-
-%used in conference proceeding
 
@@
 theory.
-%where
-%$\sigma = \omega + \ii\gamma$ is a complex frequency, $\omega$ being 
-%the real frequency and $\gamma$ is the growth rate. 
+
 The total non-axisymmetric surface density is 
@@
       m}^*\right] RdR 
-  %&= \sum_{m=0}^\infty \int _{R_\mathrm{min}}^{R_\mathrm{max}}  j_m(R)
-  %dR 
   = \sum_{m=0}^\infty J_m. 
@@
 is defined through a conservation law.
-
-%and we idenfity $J_m$ as the  
-%angular momentum associated with disc structure with
-%$m^\mathrm{th}$-fold symmetry, and $j_m$ the associated angular
-%momentum per unit length. We will use $J_m$ to monitor angular momentum 
-%conservation for the disc as a whole.
 
@@
 we find the perturbations to be confined.
-
-% \begin{align}
-%   T_\mathrm{BG} &= -\frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\delta\Sigma_m \left(\frac{\delta
-%         v_{Rm}}{-\ii\sbar}\right)^*\right] \notag\\
-%   & = \frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\ii \delta\Sigma_m \left(\frac{\delta
-%         v_{Rm}}{\sbar}\right)^*\right]
-% \end{align}
-% where we replaced the radial Lagrangian displacement by the Eulerian
-% radial velocity perturbation in Eq. \ref{baroclinic_torque}. 
-% The linearised momentum equations give
-% \begin{align}
-%   D\delta v_{Rm} =& 
-%   \ii\sbar\left[c_s^2\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right)
-%     + \frac{d}{dR}\delta\Phi_m\right] \notag\\ 
-%   &- \frac{2\ii
-%     m\Omega}{R}\left(c_s^2\frac{\delta\Sigma_m}{\Sigma} +
-%     \delta\Phi_m\right),
-% \end{align}
-% where $D\equiv \kappa^2 - \sbar^2$. 
-% We now invoke local theory by setting $d/dR \to \ii k$ where $k$ is
-% real, and using the local solution to the Poisson equation
-% \begin{align}
-%   \delta\Phi_m = -\frac{2\pi G}{|k|}\delta\Sigma_m 
-% \end{align}
-% \citep{shu91}. Then the radial velocity perturbation becomes
-% \begin{align}
-%   \delta v_{Rm} =
-%   \frac{1}{D}\frac{\delta\Sigma_m}{\Sigma}\left(\frac{2\pi G
-%       \Sigma}{|k|} - c_s^2\right)\left(k\sbar + \frac{2\ii
-%       m\Omega}{R}\right),
-% \end{align}
-% so that 
-% \begin{align}
-%   \ii\delta\Sigma_m \left(\frac{\delta v_{Rm}}{\sbar}\right)^* =
-%   \frac{1}{D^*}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
-%       \Sigma}{|k|} - c_s^2\right)\left(\ii k  + \frac{2
-%       m\Omega}{R\sbar^*}\right). 
-% \end{align}
-% For a neutral mode, $\sbar$, and therefore $D$ is real. Then, after
-% taking the imaginary part of the above expression, the torque
-% density is 
-% \begin{align}\label{tbg_explicit}
-%   T_\mathrm{BG} &=
-%   \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{D}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
-%       \Sigma}{|k|} - c_s^2\right)\right].
-% % &\simeq \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{\left(\kappa^2 -
-% %         m^2\Omega^2\right)}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi
-% %       G \Sigma}{|k|} - c_s^2\right)\right]. 
-% \end{align}
-% We now use the local dispersion relation,
-% Eq. \ref{dispersion}, to make the replacement $D = 2\pi G \Sigma |k| -
-% k^2c_s^2$ in the above expression. This gives
-% \begin{align}\label{tbg_simple}
-%   T_\mathrm{BG} =
-%   \frac{m}{2}\frac{dc_s^2}{dR}\frac{1}{k}\frac{|\delta\Sigma_m|^2}{\Sigma}.  
-% \end{align}

### 2015-01-13 paper/results.tex
almost done with first revision

@@
 
-\begin{figure}
-  \includegraphics[scale=0.27]{figures/polarxy_dens050}\includegraphics[scale=0.27,clip=true,trim=2.26cm 
+\begin{figure*}
+  \includegraphics[scale=0.55]{figures/polarxy_dens050}\includegraphics[scale=0.55,clip=true,trim=2.26cm 
   0cm 0cm 
-  0cm]{figures/polarxy_dens110}\includegraphics[scale=0.27,clip=true,trim=2.26cm
+  0cm]{figures/polarxy_dens110}\includegraphics[scale=0.55,clip=true,trim=2.26cm
   0cm 0cm 0cm]{figures/polarxy_dens130} 
@@
     $\Delta\Sigma$ is shown. \label{fargo_2d}} 
-\end{figure}
+\end{figure*}
 
@@
   associated with the imposed temperature gradient  by performing a 
-  series of simulations with $q\in[0,1]$,  However, to keep the Toomre $Q$
-  profiles similar, we adjust the surface density power-law index such
+  series of simulations with $q\in[0,1]$.  However, to maintain similar Toomre $Q$
+  profiles, we adjust the surface density power-law index such
   that $s = (3+q)/2$.  For clarity these simulations are initialised
@@
 }
-
-% The simulations above suggest the growth of the $m=1$ spiral mode
-% can be attributed to the background torque applied on linear perturbations
-% as a result of an imposed temperature gradient, as discussed in
-% \S\ref{global_cons}. 
-% According to Eq. \ref{baroclinic_torque}, this
-% torque is proportional to $q$, the power-law index used to set the
-% sound-speed profile (Eq. \ref{sound-speed}) in our simulations.
 
@@
 simulation.
-% difficult to measure  
-% show result of isothermal calc?
+
 \begin{figure}
@@
 \end{figure}
-%maybe extend the low q simulations to get better rates?
 
@@
 } 
-%In order to characterise the $m=1$ spiral that eventually
-%dominates, here we examine a simulation initialised with 
-%$M=1$ in Eq. \ref{randpert}. The one-armed spiral that emerges is more
-%coherent than the above simulation, which facilitates the analysis
-%below.   
 Fig. \ref{2d_fargo_viz} shows a snapshot of the $m=1$ surface
@@
 
-%snapshot shows it's tightly wound
 \begin{figure}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%
-
 Using the estimated value of $R_c$, we plot in
@@
 solutions to the dispersion relation (Eq. \ref{dispersion}) when the
-mode frequency is given. %Note that, with the above measured wavenumber $k$, 
-%Eq. \ref{dispersion} requires $\nu^2 - 1 + Q^{-2}\sim0$ and the lower
-%sign (long waves) to be taken. 
+mode frequency is given. 
 Fig. \ref{fargo_qbarrier} shows two $Q$-barriers located in
@@
 the measured wavenumber $k$ and Eq. \ref{wavenumber}.  
-%Eq. \ref{dispersion} with the lower sign taken (i.e. long waves).  
 There is one outer Lindblad resonance at $R_L\simeq
@@
 the spiral disturbance in the inner disc \citep{lin11b}.
-%implications: effect on dynamics in outer disc
 
@@
 \subsubsection{Angular momentum exchange with the background disc}  
-% The wavenumber $k = \pi G\Sigma/c_s^2$ minimises the shifted frequency
-% $\sbar$ in the dispersion relation, Eq. \ref{dispersion}.  
-% Physically, then, it is not surprising to 
-% find perturbations of this wavenumber in a
-% self-gravitating disc where $Q\sim 1$. However, according the
-% local dispersion relation, $m=1$ perturbations are
-% formally stable for real $k$ and $Q>1$.   
-%most of the perturbation is inside co-rotation 
-%so resonant interaction unlikely for instability
-
 {\bf We explicitly show that the growth of the $m=1$ spiral is due to
@@
 }
-% we invoke angular momentum conservation for linear
-% perturbations, Eq. \ref{lin_ang_mom_cons}. 
-% Assuming fluxes are
-% negligible at the disc boundaries, we can integrate
-% Eq. \ref{lin_ang_mom_cons} to give
+
 \begin{align}\label{baroclinic_torque_int}
@@
 The match is less good later on, when the spiral
-amplitude is no longer small ($\Delta\Sigma_1\sim 0.2$ at $t=110P_0$
-and $\Delta\Sigma_1\sim 0.4$ by $t=120P_0$) and linear theory becomes
+amplitude is no longer small ($\mathrm{max}\Delta\Sigma_1\sim 0.2$ at $t=110P_0$
+and $\mathrm{max}\Delta\Sigma_1\sim 0.4$ by $t=120P_0$) and linear theory becomes
 less applicable. Fig. \ref{fargo_angmom_ex} confirms that the $m=1$ spiral
-wave experiences a negative torque that further reduces its angular
-momentum, {\bf leading to its growth.} 
-This is consistent with angular momentum component
-measurements (Fig. \ref {2d_angmom}).
+wave experiences a negative torque that further reduces its {\bf (negative)} angular
+momentum, {\bf leading to its amplitude growth.} This is consistent
+with angular momentum component measurements (Fig. \ref {2d_angmom}).
 

### 2015-01-13 paper/results_3d.tex
almost done with first revision

@@
 to a single $m=1$ disturbance in $R\in[R_1,R_2]$.
-%, which
-%eventually dominates in 2D simulations.
 
@@
 the temperature gradient effect is the same in 3D.
-%why no high m
-                                %modes as in 2D for iso disc?
-                                %resolution?
 
@@
   possibily} because of the lower resolutions adopted in 3D {\bf
-  and/or because the effective Toomre parameter is larger in 3D which,
-  from Eq. \ref{theoretical_rate1}, is stabilising.}
+  and/or because the effective Toomre parameter is larger in 3D
+  \citep{mamat10} which, from Eq. \ref{theoretical_rate1}, is
+  stabilising.}
 
-% and/or the small softening length
-% employed in 2D. 
 %Rc = 8.6R_0
@@
 boundary. The snapshots are chosen when the one-armed spirals in the two codes
-have reached comparable amplitudes. %The agreement between the codes,
-%as well as with the 2D simulations (Fig. \ref{2d_fargo_viz}), is
-%satisfactory. 
+have reached comparable amplitudes. 
 Both codes show similar one-armed patterns at either height, and the
@@
 simulations are necessary to study the effect of the one-armed spiral
-on the exterior, non-self-gravitating disc.
+on the exterior disc.
 
@@
 sub-sonic, is not very small.
-%may increase with height
 

### 2015-01-13 paper/summary.tex
regular update (revising)
almost done with first revision

@@
 0.1\Omega_k^{-1}$ is needed for the one-armed spiral to
-develop. However, this value may {\bf be model-dependent} and will be
-explored in a follow-up study. {\bf For example,
+develop. However, this value {\bf is likely model-dependent. For example,
 a longer $t_\mathrm{relax}$ may be permitted with larger temperature
-gradients.}
+gradients. This issue, together with a parameter survey, will be
+considered in a follow-up study.}
 
@@
 gain of positive angular momentum through shock dissipation. We remark  
-that fragmentation {\bf of a low-frequency spiral} is unlikely
-because the co-rotation radius is outside the bulk of the spiral arm, in the non-self-gravitating
-portion of the disc \citep{durisen08}. In order
+that fragmentation is unlikely because the co-rotation radius is
+outside the bulk of the spiral arm \citep{durisen08}. In order
 to study these possibilities, improved numerical models are needed to
@@
 therein]{okasaki97,papaloizou06c,ogilvie08}. {\bf These studies 
-invoke other mechanisms to produce \emph{neutral} one-armed
+invoke alternative mechanisms to produce \emph{neutral} one-armed
 oscillations (e.g. rotational deformation of the star), but consider 

### 2015-01-13 paper/paper.tex
almost done with first revision

@@
 presence of spiral arms in circumstellar discs could be signposts of 
-planet formation.
+planet formation \citep{juhasz14}.
 
@@
   linear perturbations and the background disc. This `background
-  torque' may act to destabilise low-frequency non-axisymmetric
+  torque' can destabilise low-frequency non-axisymmetric
   trailing waves when the disc temperature decreases outwards.       
@@
 \section*{Acknowledgments}
-I thank K. Kratter and A. Youdin for valuable discussions, and the
+I thank K. Kratter, Y. Wu and A. Youdin for valuable discussions, and the
 anonymous referee for comments that significantly improved this paper. All

### 2015-01-13 paper/paper.tex
almost done with first revision

@@
   mechanism. Although our numerical simulations consider
-  self-gravitating discs, the `background torque' is generic for
+  self-gravitating discs, this `background torque' is generic for
   locally isothermal discs and its existence does not require

### 2015-01-14 paper/response.txt
writing response to referee

I thank the reviewer for a timely report and for providing important 
comments. I have significantly re-structured the 
paper to re-orientate the purpose of this work in order to address
the issues raised by the reviewer. The revisions are made in bold,
which I further explain below in chronological order. Response to the
reviewer's specific comments follow afterwards.

**********************************************************************

Abstract:

This paper demonstrates a potentially destabilising effect of adopting
a locally isothermal equation of state. The fact that  
linear disturbances do not generally conserve angular momentum in
locally isothermal discs has not been appreciated. I believe it is 
worth pointing out this property of locally isothermal disks given 
that such models are still widely used to model protoplanetary discs.

The theory is now presented first. The numerical simulations then serve to
illustrate an application of this destabilisation mechanism to make
one-armed spiral patterns grow in locally isothermal discs.

Introduction:

In accordance with the reviewer's concerns, I have removed references
to dead zones in this section. I deferred this issue
to the discussions section.

I now simply give a brief description of possible origins of spiral 
structure in circumstellar discs. I highlight previous works on m=1 or
eccentric modes, since the present simulations give one-armed
spirals. I then state that this paper presents a distinct way to grow
one-armed spirals by exploiting the non-conservation of angular
momentum of linear disturbances in locally isothermal disks.

The new introduction prevents the reader from confusing the results of  
this work to proper gravitational instabilities.

Section 2: Governing equations

The description of the specific disc models adopted in the numerical
simulations has been removed from this section in order to allow a
more general discussion of linear density waves in the next section.

Section 3: Linear density waves

I now describe the destabilising mechanism before presenting numerical
simulations. I also extended the description to general
non-axisymmetric disturbances.

At the end of Section 3.4, I emphasize the fact that destabilisation
by the background torque is not directly associated with self-gravity.

Section 4: Numerical simulations

I describe the specific disc models and numeical codes used to
demonstrate the background torque in action.

Section 5: Results

The results on variable temperature profiles (Section 5.2) now follows
the reference case (Section 5.1).

Section 7: Discussion

Here I address most of the reviewer's comments.

I reiterate the one-armed spiral instability is not 
self-gravitational. Our radially-structured disc models permits trapped 
one-armed modes because of disc self-gravity, but it does not grow by 
exerting gravitational torques on other parts of the disc.

I checked this by performing a new simulation with a lower disc mass 
such that Q>2.5 everywhere in the disc. Then classic
non-axisymmmetric gravitational instabilities should not occur. However,
this model still develops the one-armed spiral because of the forced
temperature gradient.

Section 7.1:

In fact, simulations performed in the early stages of this project did
include the indirect potential, but produced similar results. This
confirms that the indirect potential is not related to the formation
of the one-armed spiral patterns in our disc models.

Section 7.2:

The calculations for this section has been moved in
Appendix C. Here, I state the finding that co-rotation and Lindblad
resonances are not important for the one-armed spiral pattern in our
disc models. This supports the notion that the instability is not
gravitational.

Section 7.3:

Here I discuss possible relevance of our results to protoplanetary
discs.

I state that the destabilising mechanism --- the background
torque --- only exists in the presence of a fixed temperature
gradient. This in turn requires the system to return to its initial
temperature profile quickly when perturbed.

I suggest that outer parts of protoplanetary discs, where rapid
cooling may be possible and in which stellar irradiation can be
important, may be a potential site for the background torque to
operate.

For example, in radiation hydrodynamic simulations performed by
Boley (2009) the outer regions of protoplanetary discs (>100AU)
behaves isothermally because of external irradiation. In that work the
irradiation temperature is uniform so the background torque is
absent. However, it is conceivable that external irradiation can set a 
non-uniform disc temperature profile (Chiang & Goldreich 1997),
which may enable the background torque.

(The existence of the background torque and gravitational
instability/fragmentation are not necessarily mutually exclusive.)

The other requirement, more specific to forming trapped ne-armed
spirals, is the radial disc structure. Our disc models can trap
one-armed modes because of the rising Q-profile towards the inner disc
which, in turn, is a result of the adopted surface density bump.

The surface density bump may represent mass built-up at some radial
location in the disc. Protoplanetary disc dead zones are an obvious
example. However, as the reviewer points out, dead zones may not
support the background torque because it cools too slowly.

Actually, recent simulations performed by Bae et
al. (2014) show that dead zones develop large-scale gravitational
instabilities, as opposed to high-m, small-scale local
gravito-turbulence.

I suggest planetary gaps as another possibility. For example, if a
giant planet opens a gap in a constant-Q disc, then the trapping
condition Eq. 49, may give rise to two Q-barriers at the outer gap
edge. Note that this does not necessarly require strong self-gravity.

Section 8: Summaryand conclusions

The summary section has been re-ordered to reflect the new paper
structure.

Seciton 8.1:

************************************************************************

This is one of those papers that I find quite difficult to review.
There's nothing that I can find that is fundamentally wrong with the
paper.  It's easy to read.  The methods and assumptions are clearly
explained.  It's trying to address a topic that is clearly of interest
(how can we explain spiral structure seen in protostellar
discs). However, it is presenting something that I would regard as
largely unrealistic.   I'll explain my reasoning below.

The simulations themselves consider an isothermal or locally
isothermal equation of state only.  Gammie (2001) and other work since
have shown that the evolution of self-gravitating discs depends both
on the stability criterion and on the balance between heating and
cooling.  If the cooling rate is fast, it is likely to fragment,
otherwise it will settle into a quasi-steady, self-gravitating state.

When considering isothermal simulations, the likely evolution is one
of two things.  It's either initially so unstable that the disc
quickly fragments, or it's initially stable enough that any initial
perturbations simply drives the system towards being even more stable
and the instability quickly weakens through rapid mass transfer (see
the work by Laughlin & Bodenheimer 1994/1996).  There is no real way
to drive a long-lived quasi-steady, non-fragmenting, self-gravitating
state because this is only possible if the system can remain
quasi-stable though balancing heating and cooling (see, for example,
Lodato & Rice 2004).

In fact, in the conclusions it is suggested that the conditions used
in the simulation would require a cooling < 0.1 \Omega^{-1} which,
based on the work of Gammie 2001, would quickly lead to fragmentation.
The reason it doesn't in the work considered here is that the
isothermal assumption assumes that there is some equivalently fast
heating mechanism but, in the work here, this is not provided by the
instability itself.  If it were, the simulations would become very
quickly unstable.  Also, if one looks at the work of Clarke (2009),
Rafikov (2005), and Rice & Armitage (2009) the only region of a
realistic disc where such rapid cooling is possible is in the outer
disc, but such regions are either very gravitationally unstable and
fragment, or quickly stablised by an external heating source.  I guess
it is possible that there could be a scenario where this exactly
produces the conditions assumed here, but that would seem rather
fine-tuned and it would be more likely that an external heating source
would simply stabilise the disc (unless the disc were extremely
massive and could remain unstable even at reasonably high disc
temperatures).

When considering the reference run discussed on page 5, my suspicion
is that the behaviour can largely be explained as being a consequence
of the chosen numerical setup.  For example, the initial Q profile has
a minimum of about Q = 2, so the disc is unlikely to become very
unstable.  The simulation is seeded with an initial high-m
perturbation that decays quite quickly and is then replaced by a
long-lived m=1 spiral mode that continues to grow and in which the
growth rate depends on the imposed temperature profile.

It's my view that if you used a realistic equation of state that
included heating and cooling, the high-m mode instability would either
persist (as suggested by Lodato & Rice 2004) or the system would
simply heat up until it became stable against the growth of any
perturbation.  The disc mass is also quite low (0.086M*) and if one
were to estimate the disc temperature using some kind of realistic
assumption about the central star mass and disc outer radius, my
suspicion is that to have the Q profile used, the outer radius would
either have to be very small (a very compact disc) or the disc
temperature would be unrealistically low.

I have a further issue with the m=1 mode.  In these simulations the
central star is fixed.  As I understand it, the m=1 mode is not
axi-symmetric and hence it would seem to me that if the central star
is fixed, the presence of a dominant m=1 mode means that the
centre-of-mass is moving in the simulations.  In fact, it seems to me
that it is possible that the fixed central star is what is driving the
m=1 mode (i.e., the central star is unable to move to keep the centre
of mass fixed and this may have implications for the growth of
instabilities in the disc).  It would certainly be interesting to know
if you could get the same result as obtained here if the central star
was allowed to move.

As I said at the beginning, I don't have any fundamental problems with
how the paper is written, the problem it is trying to address, or how
the assumptions and setup is explained.  The main issue is that I
think it is sufficiently unrealistic that adding any kind of energy
equation (or even allowing the central star to move) would change the
evolution of these discs so much that what is presented here would not
be observed in reality.  That isn't, by itself, a reason not to
publish this paper, but it would seem that it would benefit from a
some more discussion about whether or not such a process could
actually operate and where it could operate.  There is some attempt to
do so when discussing dead zones, but I doubt that the conditions
there are actually suitable as such a region would either be stable,
or cool very slowly and be susceptible to the growth of long-lived,
high-m-mode spiral structure.

### 2015-01-14 paper/appendix.tex
writing response to referee

@@
 
-If we associate the external potential with angular momentum 
+If we associate the external potential with {\bf an} angular momentum {\bf magnitude of} 
 $J_\mathrm{ext}  = M_\mathrm{ring}\overline{R}^2\Omega_p$, we can
-calculate a growth rate $\gamma_L=\Gamma_L/J_\mathrm{ext}$. Then 
+calculate a rate {\bf of change of angular momentum} $\gamma_L=\Gamma_L/J_\mathrm{ext}$. Then  
 \begin{align}
@@
 \end{align}
-These estimates are much smaller than that due to the
-imposed temperature gradient as measured in the simulations
-($\gamma\sim∼0.1\Omega_p$). We conclude that for our disc models, 
-the Lindblad and co-rotation resonances have negligible effects
-on the growth of the one-armed spiral in the inner disc.
 
@@
 outer disc, this co-rotation torque is positive. 
-%Interpreting the one-armed spiral as an external po-
-%tential for the outer disc, 
 This means that the one-armed spiral loses
@@
 interaction is destabilising because the one-armed spiral has 
-negative angular momentum \citep{lin11b}. This
-effect is small in our simulations, but it could be significant
-in other parameter regimes.
+negative angular momentum \citep{lin11b}.
+
+{\bf However, the above estimates for $\gamma_L$ and $\gamma_c$} are much smaller than that due to the 
+imposed temperature gradient as measured in the simulations 
+($\gamma\sim∼0.1\Omega_p$). We conclude that for our disc models, 
+the Lindblad and co-rotation resonances have negligible effects
+on the growth of the one-armed spiral in the inner disc (but it could be {\bf important} 
+in other parameter regimes).

### 2015-01-14 paper/discussion.tex
writing response to referee

@@
   destabilisation is through the background torque associated with a
-  forced temperature gradient, and not by self-gravitational torques\footnote{In fact, additional simulations with $Q_\mathrm{out}=4$ still
+  forced temperature gradient, and not by self-gravitational torques\footnote{In fact, additional simulations with $Q_\mathrm{out}=4$ (giving $Q\gtrsim2.5$ throughout the disc) still 
   develops the one-armed spiral, but with a smaller growth rate.}. 

### 2015-01-14 paper/response.txt
writing response

@@
 comments. I have significantly re-structured the 
-paper to re-orientate the purpose of this work in order to address
+paper to re-orientate this work in order to address
 the issues raised by the reviewer. The revisions are made in bold,
@@
 
-This paper demonstrates a potentially destabilising effect of adopting
+This paper describes a potentially destabilising effect of adopting
 a locally isothermal equation of state. The fact that  
@@
 
-The theory is now presented first. The numerical simulations then serve to
-illustrate an application of this destabilisation mechanism to make
-one-armed spiral patterns grow in locally isothermal discs.
+The basic idea behind the background torque is now presented first. 
+The numerical simulations then serve to demonstrate
+this destabilisation mechanism to make one-armed spiral patterns 
+grow in locally isothermal discs.
 
@@
 The new introduction prevents the reader from confusing the results of  
-this work to proper gravitational instabilities.
+this work with proper gravitational instabilities.
 
@@
 I reiterate the one-armed spiral instability is not 
-self-gravitational. Our radially-structured disc models permits trapped 
+self-gravitational. The radially-structured disc models adopted here permits trapped 
 one-armed modes because of disc self-gravity, but it does not grow by 
-exerting gravitational torques on other parts of the disc.
+exerting gravitational torques on other parts of the disc (which 
+underpins classic gravitational instabilities).
 
@@
 this model still develops the one-armed spiral because of the forced
-temperature gradient.
+temperature gradient, but it grows more slowly.
 
@@
 In fact, simulations performed in the early stages of this project did
-include the indirect potential, but produced similar results. This
-confirms that the indirect potential is not related to the formation
-of the one-armed spiral patterns in our disc models.
+include the indirect potential (allowing the star to move), but these produced 
+similar results. This confirms that the indirect potential is not related 
+to the formation of the one-armed spiral patterns in our disc models.
 
@@
 
-Here I discuss possible relevance of our results to protoplanetary
+Here I discuss possible relevance of the results to protoplanetary
 discs.
@@
 absent. However, it is conceivable that external irradiation can set a 
-non-uniform disc temperature profile (Chiang & Goldreich 1997),
+non-uniform disc temperature profile, 
 which may enable the background torque.
@@
 
-The other requirement, more specific to forming trapped ne-armed
+The other requirement, more specific to trapping one-armed
 spirals, is the radial disc structure. Our disc models can trap
@@
 
-The surface density bump may represent mass built-up at some radial
+A surface density bump may represent mass built-up at some radial
 location in the disc. Protoplanetary disc dead zones are an obvious
@@
 
+At the end of this section I remind the reader that the background torque
+is a generic feature of locally isothermal discs. It may be relevant whenever 
+the disc temperature is set externally. Here, I mention another possible application in 
+Be star discs.
+
 ************************************************************************
@@
 otherwise it will settle into a quasi-steady, self-gravitating state.  
+************************************************************************
+The issue of fragmentation due to rapid cooling, as implied by assuming a 
+locally isothermal equation of state, is now mentioned in Section 7.3.1.
+
+*************************************************************************
 
@@
 Lodato & Rice 2004).  
+************************************************************************
+Our disc models have radial structure. It is not clear how to  
+apply results obtained the above studies that use smooth discs to the 
+present work.
+
+In a smooth disc non-axisymmetric gravitational instabilities can 
+operate when Q<1.5-1.7. This is satisfied in our reference disc model, 
+but this region is rather narrow, about 10 scale heights wide within the 
+surface density bump. It is not obvious that the high-m modes 
+observed early on in this simulation is a gravitational instability 
+(Fig. 3, right panel), because the radial disc structure may permit 
+other hydrodynamical shear instabilities (Lovelace et al. 1999).
+
+In any case, the high-m modes saturate at small amplitude (<0.01) and no 
+fragmentation is seen, even though the discs are (locally) isothermal. 
+Eventually a one-armed spiral grows, but as explained above, this is not 
+a gravitational instability in the usual sense.
+
+Furthermore, fragmentation of the one-armed spiral is unlikely. Durisen
 

### 2015-01-14 paper/model.tex
writing response to referee
writing response

@@
 and arises because we have adopted a locally isothermal equation of
-state {\bf in the perturbed disk}. {\bf  We outline the
+state {\bf in the perturbed disc}. {\bf  We outline the
 derivation of $T_\mathrm{BG}$ in Appendix \ref{tbg_deriv}.}
@@
 {\bf 
-Here we show that the torque density acting on a low-frequency, 
+Here we show that the torque density acting on a 
 local mode due to the background temperature gradient can be negative, which 
-would enforce the mode, because it has negative angular momentum.  
+would enforce low-frequency modes, because they have negative angular momentum.  
 }

### 2015-01-15 paper/response.txt
first revision complete, but still need spell check and proof read

@@
 
+fukagawa 13
+
 Section 8: Summaryand conclusions
@@
 is a generic feature of locally isothermal discs. It may be relevant whenever 
-the disc temperature is set externally. Here, I mention another possible application in 
-Be star discs.
+the disc temperature is set externally. Here, I mention another
+possible application in  Be star discs.
 
@@
 
-In a smooth disc non-axisymmetric gravitational instabilities can 
+In a smooth disc, non-axisymmetric gravitational instabilities can 
 operate when Q<1.5-1.7. This is satisfied in our reference disc model, 
@@
 surface density bump. It is not obvious that the high-m modes 
-observed early on in this simulation is a gravitational instability 
+observed early on in this simulation is a true gravitational instability 
 (Fig. 3, right panel), because the radial disc structure may permit 
@@
 In any case, the high-m modes saturate at small amplitude (<0.01) and no 
-fragmentation is seen, even though the discs are (locally) isothermal. 
-Eventually a one-armed spiral grows, but as explained above, this is not 
-a gravitational instability in the usual sense.
+fragmentation is seen (Fig. 2), even though the discs are (locally)
+isothermal and min(Q) ~ 1.3. Thus, our models concern 
+self-gravitating, but (effectively) gravitationally stable discs.  
+Eventually a one-armed spiral grows, but as explained above, this is
+not a gravitational instability in the usual sense.
 
-Furthermore, fragmentation of the one-armed spiral is unlikely. Durisen
+Durisen, Hartquist & Pickett (2008) and Rogers & Wadsley (2012) 
+find that fragmentation occurs the corotation radius. However, the
+one-armed spirals in our simulations have low-frequency, so its 
+corotation radius is not within the spiral arm itself. This makes
+fragmentation unlikely (the one-armed spiral does not grow because 
+of self-gravity anyway). 
+*************************************************************************
 
@@
 temperatures).  
+*************************************************************************
+The value 0.1/Omega is based on a limited number of preliminary
+simulations that are currently underway. The critical thermal
+relaxation time may depend on the disc parameters. I mention this in
+Section 8.1.
 
+I thank the reviewer for pointing out external heating. In Section
+7.3.1 I suggest that our disc models may be appropriate to describe
+the outer parts of protoplanetary discs that are subject to external
+irradiation such that they are gravitationally stable.
+
+Note that stabilisation by external irradiation against  
+self-gravitational instabilities does not necessarily imply
+stablisation against the one-armed spiral instability described in
+this paper.
+
+**************************************************************************
 When considering the reference run discussed on page 5, my suspicion
@@
 growth rate depends on the imposed temperature profile.  
+***************************************************************************
+The minimum Q in the reference run is between 1.3 and 1.4 which,
+according to previous studies, should result in  
+strong non-axisymmetric gravitational instabilities (Durisen et
+al. 2007, Helled et al. 2013). This does not occur here, despite the
+simulations being (effectively) isothermal, which should favour
+GI.
 
+Fig. 2 shows that the high-m modes do not decay over the simulation
+timescale, but it saturates at a small amplitude. Thus, if these
+high-m modes are GI, then GI is weak in our models despite having low
+Q.
+
+One possible origin of discrepency is the fact that our discs are
+radially structured, while the aforementioned studies employ global
+smooth discs, or local disc models where any effect of global radial 
+structure are absent. It is not obvious that results from smooth-disc
+studies can be directly applied to structured discs.
+
+It is not unreasonable to study radially-structured discs. Realistic
+discs may not be described by smooth power-laws, e.g. a disc 
+with an gap induced by a planet (Section 7.3.2).
+
+The numerical setup was chosen to effectively demonstrate the
+destabilising effect of the background torque (Section 5.2 and
+5.3). However, its occurence is not limited to the particular setup
+presented in Section 5 and 6.
+
+I have performed additional simulations with 
+higher Q and simulations with only an inner bump edge (by setting a
+large R2, see Eq. 33). These simulations still developed the one-armed
+spiral, but require longer runs so high-resolution/3D is not
+feasible. A full parameter survey is beyond the scope of this work,
+but is the subject of a follow-up study.
+
+***************************************************************************
 It's my view that if you used a realistic equation of state that
@@
 temperature would be unrealistically low.  
+***************************************************************************
+The high-m perturbations do in fact persist in the reference run 
+(Fig. 2), but it is overshadowed by the one-armed instability.
 
+I note that Lodato & Rice (2004) cool their discs to zero temperature
+which is balanced against shock-heating (as in many similar
+studies). The system we consider are closer to irradiated discs (Rice
+et al. 2011) in which the temperature is set externally (e.g. Boley
+2009), which may then give rise to the background torque (Section
+7.3.1).
+
+The properties of the disc as a whole, i.e. its total mass and outer
+radius, is not expected to be important as far as destabilisation by
+the background torque is concerned. These parameters do not appear in
+our description of this effect (Section 3). In Section 7.3.2 I suggest
+disc-planet interaction as one possible origin for the required
+structure in the Q-profile to trap m=1 modes.
+
+***************************************************************************
 I have a further issue with the m=1 mode.  In these simulations the
@@
 was allowed to move.  
+***************************************************************************
+The usual interpretation of m=1 gravitational instabilities is
+attributed to the motion of the central star (Adams, Ruden & Shu 1989,
+Shu et al. 1990, Heemskerk, Papaloizou, Savonije 1992). Moving the
+star with the disc requires a somewhat massive disc, which is not the
+case considered in this work.
 
+I have performed simulations including the indirect potential, and
+thus allowing the star to move, which gave very similar results to
+those presented, i.e. the one-armed mode still develops. This is
+mentioned in Section 7.1.
+
+***************************************************************************
 As I said at the beginning, I don't have any fundamental problems with
@@
 high-m-mode spiral structure. 
+****************************************************************************
+I have moved the issue of dead zones to Section 7.3.2. Dead
+zones may have the appropriate radial structure to trap one-armed
+modes. However, I state that, in agreement with the reviewer, dead
+zones may not have the suitable thermodynamic conditions to enable the 
+background torque.
+
+I suggest a planetary gap as another possibility. I  
+highlight the transition disc around HD 142527 which shows a gap structure
+at large radii (>100AU), for which observations estimate a low, 
+but possibily stable Toomre Q value against self-gravity. Thus, the disc model 
+considered in this work may not be too artificial.

### 2015-01-15 paper/discussion.tex
first revision complete, but still need spell check and proof read

@@
   parts of protoplanetary discs 
-  \citep{rafikov05,clarke09,rice09,cossins10b}.  %rafikov07 (convection)
+  \citep{rafikov05,clarke09,rice09,cossins10b,tsukamoto15}.  %rafikov07 (convection)
   However, if a disc with $Q\simeq 1$ is cooled (towards zero
@@
   possible in the outer parts of protoplanetary discs due to
-  stellar irradiation \citep{rafikov09,zhu12}. Sufficiently strong
+  stellar irradiation \citep{rafikov09,kratter11,zhu12}. Sufficiently strong
   external irradiation is expected to suppress the linear gravitational
@@
   The background torque may thus exist in the outer
-  parts of protoplanetary discs that are strongly irradiated, such
+  parts of protoplanetary discs that are irradiated, such
   that the disc temperature is set externally with a non-zero radial
@@
   profile about the gap edge is set by the planet luminosity.
+
+  Here, it is worth mentioning transition disc 
+  around HD 142527, the outer parts of which displays an $m=1$
+  asymmetry \citep{fukagawa13} and spiral arms 
+  \citep{christiaens14} just outside a disc gap. These authors estimate 
+  $Q\simeq$1---2 in the outer disc, implying self-gravity is
+  important, but the disc may remain gravitationally-stable
+  \citep{christiaens14}. This is a  
+  possible situtation that our disc models represent. 
 }

### 2015-01-15 paper/response.txt
first revision complete, but still need spell check and proof read

@@
 I state that the destabilising mechanism --- the background
-torque --- only exists in the presence of a fixed temperature
-gradient. This in turn requires the system to return to its initial
-temperature profile quickly when perturbed.
+torque --- exists in the presence of a fixed temperature
+gradient. (It only vanishes identically for strictly 
+adiabatic/barotropic discs, see Appendix A.) This in turn requires the system t
+o return to its initial temperature profile quickly when perturbed.
 
@@
 condition Eq. 49, may give rise to two Q-barriers at the outer gap
-edge. Note that this does not necessarly require strong self-gravity.
-
-fukagawa 13
+edge. Note that this does not necessarily require strong self-gravity.
 

### 2015-01-15 paper/response.txt
nearly done with first revision

@@
 
-I note that Lodato & Rice (2004) cool their discs to zero temperature
+Lodato & Rice (2004) cool their discs to zero temperature
 which is balanced against shock-heating (as in many similar
-studies). The system we consider are closer to irradiated discs (Rice
-et al. 2011) in which the temperature is set externally (e.g. Boley
-2009), which may then give rise to the background torque (Section
-7.3.1).
+studies). The systems we consider are closer to irradiated discs (Rice
+et al. 2011) in which the temperature is set externally 
+(e.g. Stamatellos &  Whitworth 2008, Boley 2009). I note that 
+such global simulations including heating (irradiation) and cooling 
+may not have been run long enough to see low-frequency modes 
+which grow slowly, and thus appear stable.
 

### 2015-01-15 paper/discussion.tex
nearly done with first revision

@@
   that the disc temperature is set externally with a non-zero radial
-  gradient. Of course, if external irradiation sets a strictly
+  gradient \citep[e.g.][]{stamatellos08}. Of course, if external irradiation sets a strictly
   isothermal outer disc \citep[e.g.][]{boley09}, then the background

### 2015-01-15 paper/response.txt
first revision done

@@
 I thank the reviewer for a timely report and for providing important 
-comments. I have significantly re-structured the 
+comments. I have re-structured the 
 paper to re-orientate this work in order to address
@@
 In accordance with the reviewer's concerns, I have removed references
-to dead zones in this section. I deferred this issue
+to dead zones in this section. I defer this issue
 to the discussions section.
 
-I now simply give a brief description of possible origins of spiral 
+I now simply give a brief overview of possible origins of spiral 
 structure in circumstellar discs. I highlight previous works on m=1 or
@@
 
-The description of the specific disc models adopted in the numerical
-simulations has been removed from this section in order to allow a
+I have removed the description of the specific disc models adopted in
+the numerical simulations from this section in order to allow a
 more general discussion of linear density waves in the next section.
@@
 
-I describe the specific disc models and numeical codes used to
+I describe the specific disc models and numerical codes used to
 demonstrate the background torque in action.
@@
 I reiterate the one-armed spiral instability is not 
-self-gravitational. The radially-structured disc models adopted here permits trapped 
+self-gravitational in the usual sense. 
+The radially-structured disc models adopted here permits trapped 
 one-armed modes because of disc self-gravity, but it does not grow by 
@@
 such that Q>2.5 everywhere in the disc. Then classic
-non-axisymmmetric gravitational instabilities should not occur. However,
+non-axisymmetric gravitational instabilities should not occur. However,
 this model still develops the one-armed spiral because of the forced
@@
 gradient. (It only vanishes identically for strictly 
-adiabatic/barotropic discs, see Appendix A.) This in turn requires the system t
-o return to its initial temperature profile quickly when perturbed.
+adiabatic/barotropic discs, see Appendix A.) This in turn requires the
+system to return to its initial temperature profile quickly when perturbed.
 
@@
 absent. However, it is conceivable that external irradiation can set a 
-non-uniform disc temperature profile, 
-which may enable the background torque.
+non-uniform disc temperature profile, which may enable the background
+torque.
 
@@
 
-Section 8: Summaryand conclusions
+Section 8: Summary and conclusions
 
@@
 
-Seciton 8.1:
+Section 8.1:
 
@@
 observed early on in this simulation is a true gravitational instability 
-(Fig. 3, right panel), because the radial disc structure may permit 
-other hydrodynamical shear instabilities (Lovelace et al. 1999).
+(Fig. 3, left panel), because the radial disc structure may permit 
+other hydrodynamic shear instabilities (Lovelace et al. 1999).
 
@@
 isothermal and min(Q) ~ 1.3. Thus, our models concern 
-self-gravitating, but (effectively) gravitationally stable discs.  
-Eventually a one-armed spiral grows, but as explained above, this is
-not a gravitational instability in the usual sense.
+self-gravitating, but (effectively) gravitationally stable discs
+(Stamatellos &  Whitworth 2008). Eventually a one-armed spiral grows,
+but as explained above, this is not a gravitational instability in the
+usual sense.
 
 Durisen, Hartquist & Pickett (2008) and Rogers & Wadsley (2012) 
-find that fragmentation occurs the corotation radius. However, the
+find that fragmentation occurs at the corotation radius. However, the
 one-armed spirals in our simulations have low-frequency, so its 
@@
 of self-gravity anyway).
+
 *************************************************************************
@@
 **************************************************************************
+
 When considering the reference run discussed on page 5, my suspicion
@@
 
-One possible origin of discrepency is the fact that our discs are
+One possible origin of discrepancy is the fact that our discs are
 radially structured, while the aforementioned studies employ global
-smooth discs, or local disc models where any effect of global radial 
-structure are absent. It is not obvious that results from smooth-disc
-studies can be directly applied to structured discs.
+smooth discs, or local models which excludes any effects of
+global radial disc structure. It is not obvious that results from
+smooth-disc studies can be directly applied to structured discs.
 
@@
 discs may not be described by smooth power-laws, e.g. a disc 
-with an gap induced by a planet (Section 7.3.2).
+with an gap induced by a planet (Section 7.3.2) and transition discs.
 
 The numerical setup was chosen to effectively demonstrate the
-destabilising effect of the background torque (Section 5.2 and
-5.3). However, its occurence is not limited to the particular setup
+destabilising effect of the fixed temperature gradient (Section 5.2 and
+5.3). However, its occurrence is not limited to the particular setup
 presented in Section 5 and 6.
@@
 radius, is not expected to be important as far as destabilisation by
-the background torque is concerned. These parameters do not appear in
+the temperature gradient is concerned. These parameters do not appear in
 our description of this effect (Section 3). In Section 7.3.2 I suggest
@@
 modes. However, I state that, in agreement with the reviewer, dead
-zones may not have the suitable thermodynamic conditions to enable the 
-background torque.
+zones may not have the suitable thermodynamic conditions for the instability.
 
@@
 at large radii (>100AU), for which observations estimate a low, 
-but possibily stable Toomre Q value against self-gravity. Thus, the disc model 
+but possibly stable Toomre Q value against self-gravity. Thus, the disc model 
 considered in this work may not be too artificial.

### 2015-01-15 paper/discussion.tex
first revision done

@@
   dispersion relation is possible (Eq. \ref{wavenumber}). The
-  existence of such a cavity is a result of the initial surface
+  existence of such a cavity results from the adopted initial surface
   density bump (Eq. \ref{sig_bump}).  
@@
   to maintain a fixed temperature profile. Recently, 
-  \cite{bae14} presented sophisticated numerical models of
-  dead zones, including a range of heating and 
+  \cite{bae14} presented numerical models of
+  dead zones including a range of heating and 
   cooling processes, which show that dead zones developed large-scale
@@
   locally isothermal requirement would limit this process to the
-  outer disc irradiated by the central star, or that the temperature
+  outer disc, or that the temperature 
   profile about the gap edge is set by the planet luminosity.
 
-  Here, it is worth mentioning transition disc 
+  Here, it is worth mentioning the transition disc 
   around HD 142527, the outer parts of which displays an $m=1$
@@
   \citep{christiaens14}. This is a  
-  possible situtation that our disc models represent. 
+  possible situation that our disc models represent. 
 }

### 2015-01-15 paper/model.tex
first revision done

@@
 of instability if $\tbg\jlin>0$. That is, if $\jlin$ is positive
-(negative) and the background torque $\tbg$ is {\bf also} positive (negative),
+(negative) and $\tbg$ is {\bf also} positive (negative),
 then the local angular momentum density of the linear disturbance
@@
 
-\subsection{Angular momentum density of  non-axisymmetirc low-frequency modes}
+\subsection{Angular momentum density of  non-axisymmetric low-frequency modes}
 From Eq. \ref{lin_ang_mom_cons} and assuming a time-dependence of the
@@
 {\bf where the second equality uses $\xi_\phi \simeq
-2\ii \xi_R/m$ in the local approximation, as shown in  Appendix
+2\ii \xi_R/m$ for low-frequency modes in the local approximation, as shown in  Appendix
 \ref{horizontal_displacements}.} 
@@
 to its azimuthal transport algorithm, which removes the mean azimuthal
-velocity of the disc, thereby permit larger timesteps than that would otherwise be allowed
+velocity of the disc, thereby permit larger time-steps than that would otherwise be allowed
 by the usual Courant condition based on the full azimuthal
@@
 assuming $\Sigma_m\propto\exp{(-\ii \sigma t)}$ as in linear
-theory.
-
-The total non-axisymmetric surface density is 
+theory. The total non-axisymmetric surface density is 
 \begin{align}

### 2015-01-15 paper/paper.tex
first revision done

@@
 include lopsided dust distributions 
-\citep{marel13,fukagawa13,casassus13,isella13,perez14,follette14} and
+\citep{marel13,fukagawa13,casassus13,isella13,perez14,follette14,plas14} and
 spiral arms 
@@
   for linear perturbations acquires a source term proportional to the
-  temperature gradient. This permits angular momenum exchange between
+  temperature gradient. This permits angular momentum exchange between
   linear perturbations and the background disc. This `background
@@
 discuss them in \S\ref{discussions}. We summarise in \S\ref{summary} 
-with a discussion on speculations for future work.
+with some speculations for future work.
 

### 2015-02-06 paper/appendix.tex
first revision done
paper accepted, minor corrections

@@
   2\Omega\hat{\bm{z}}\times\frac{D\bm{\xi}}{Dt}  \notag \\ &= -
-  \frac{\nabla \delta p }{\rho} + \frac{\delta\rho}{\rho^2}\nabla\rho  
+  \frac{\nabla \delta p }{\rho} + \frac{\delta\rho}{\rho^2}\nabla p  
   -\nabla\delta\Phi_d - R
@@
 taking the imaginary part afterwards. The left hand side becomes the
-angular momentum density. The first term on the right hand side (RHS)
+ rate of change of angular momentum density. The first term on the right hand side (RHS)
 becomes 
@@
 \end{align} 
-{\bf 
 For low-frequency modes we have $\sbar \simeq -m\Omega$, so
 $\xi_\phi\simeq 2\ii\xi_R/m$, as used in the main text.
-}
 
@@
 
-If we associate the external potential with {\bf an} angular momentum {\bf magnitude of} 
+If we associate the external potential with  an angular momentum magnitude of  
 $J_\mathrm{ext}  = M_\mathrm{ring}\overline{R}^2\Omega_p$, we can
-calculate a rate {\bf of change of angular momentum} $\gamma_L=\Gamma_L/J_\mathrm{ext}$. Then  
+calculate a rate of change of angular momentum $\gamma_L=\Gamma_L/J_\mathrm{ext}$. Then  
 \begin{align}
@@
 
-{\bf However, the above estimates for $\gamma_L$ and $\gamma_c$} are much smaller than that due to the 
+ However, the above estimates for $\gamma_L$ and $\gamma_c$ are much smaller than that due to the 
 imposed temperature gradient as measured in the simulations 
@@
 the Lindblad and co-rotation resonances have negligible effects
-on the growth of the one-armed spiral in the inner disc (but it could be {\bf important} 
+on the growth of the one-armed spiral in the inner disc (but it could be  important 
 in other parameter regimes).

### 2015-02-06 paper/model.tex
paper accepted, minor corrections

@@
 $\Omega_k=\sqrt{GM_*/R^3}$ is the midplane Keplerian frequency, and
-$q$ is the imposed temperature gradient since, for an ideal gas the
-temperature is proportional to $c_s^2$. {\bf For convenience we will
-  refer to $c_s^2$ as the disc temperature.} The vertical disc scale-height is
+$-q$ is the imposed temperature gradient since, for an ideal gas the
+temperature is proportional to $c_s^2$. For convenience we will
+  refer to $c_s^2$ as the disc temperature. The vertical disc scale-height is
 defined by $H=c_s/\Omega_k$. Thus, a strictly isothermal disc with
@@
 
-{\bf We describe a key feature of locally isothermal discs that
+ We describe a key feature of locally isothermal discs that
   enables angular momentum exchange between small disturbances and the
@@
   For simplicity, in this section we consider 2D discs.
-}
 
@@
 integro-differential equation eigenvalue problem. We defer a full
-{\bf numerical} exploration of the linear problem to a future study. Here, we {\bf
-  discuss} some
+ numerical exploration of the linear problem to a future study. Here, we 
+  discuss some
 general properties of the linear perturbations.
@@
 and arises because we have adopted a locally isothermal equation of
-state {\bf in the perturbed disc}. {\bf  We outline the
-derivation of $T_\mathrm{BG}$ in Appendix \ref{tbg_deriv}.}
+state in the perturbed disc.  We outline the
+derivation of $T_\mathrm{BG}$ in Appendix \ref{tbg_deriv}.
 
@@
 of instability if $\tbg\jlin>0$. That is, if $\jlin$ is positive
-(negative) and $\tbg$ is {\bf also} positive (negative),
+(negative) and $\tbg$ is also positive (negative),
 then the local angular momentum density of the linear disturbance
-{\bf will} further increase (decrease) {\bf with time}. This implies the amplitude of the disturbance
+ will further increase (decrease) with time. This implies the amplitude of the disturbance
 may grow by exchanging angular momentum with the background
@@
 
-{\bf 
   We demonstrate instability for low-frequency modes 
@@
   appropriate perturbations and radial temperature gradients.
-}
 
@@
 For a low-frequency mode, $|\omega|\ll m\Omega$. Then neglecting the
-term $\omega|\bm{\xi}|^2$, {\bf we find}
+term $\omega|\bm{\xi}|^2$,  we find 
 \begin{align}
@@
 \end{align}
-{\bf
+
 Thus, non-axisymmetric ($m\geq1$) low-frequency modes generally have 
@@
 $T_\mathrm{BG}$ in the local approximation, which we review first.
-}
 
@@
 where $k$ is a real wavenumber such that $|kR|\gg1$ \citep{shu91}. 
-Note that in the {\bf strictly} local approximation, {\bf where all
-  global effects are neglected}, only axisymmetric perturbations
+Note that in the strictly local approximation, where all
+  global effects are neglected, only axisymmetric perturbations
 ($m=0$) can be unstable. % (having an imaginary $\sigma$).
@@
 Given the real frequency $\omega$ or pattern speed $\Omega_p\equiv  
-\omega/m$ of a {\bf non-axisymmetric} neutral mode 
+\omega/m$ of a non-axisymmetric neutral mode 
 , Eq. \ref{dispersion} can be solved 
@@
 
-{\bf At the \emph{co-rotation radius} $R_c$} the pattern speed matches
+ At the \emph{co-rotation radius} $R_c$ the pattern speed matches
 the fluid rotation,
@@
 A detailed discussion of the properties of local density waves 
-is given in \cite{shu91}. An important result, {\bf which holds for
-  waves of all frequencies}, is that
+is given in \cite{shu91}. An important result, which holds for
+  waves of all frequencies, is that
 waves interior to co-rotation ($R<R_c$) have negative angular momentum, while
@@
   and the background disc due to an imposed temperature gradient}
-{\bf 
+
 Here we show that the torque density acting on a 
@@
 would enforce low-frequency modes, because they have negative angular momentum.
-}
 
@@
 \end{align}
-{\bf Inserting this into the expression for the background torque,
-  Eq. \ref{baroclinic_torque}, we find}
+ Inserting this into the expression for the background torque,
+  Eq. \ref{baroclinic_torque}, we find 
 \begin{align}
@@
 the background disc applies a negative torque on the disturbance,
-which further decreases its angular momentum. This suggests the {\bf
-  mode} 
+which further decreases its angular momentum. This suggests the 
+  mode  
 amplitude will grow.
@@
 \end{align}
-{\bf where the second equality uses $\xi_\phi \simeq
+ where the second equality uses $\xi_\phi \simeq
 2\ii \xi_R/m$ for low-frequency modes in the local approximation, as shown in  Appendix
-\ref{horizontal_displacements}.} 
+\ref{horizontal_displacements}. 
 Then for the temperature profiles $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ as
@@
 \end{align}
-where we used $c_s\sim H\Omega\sim hR\Omega$. {\bf
+where we used $c_s\sim H\Omega\sim hR\Omega$. 
   Eq. \ref{theoretical_rate} suggests that perturbations with small
@@
   destabilisation. Taking the local approximation is then
-  appropriate.}
+  appropriate.
 
-{\bf Note that the derivation of Eq. \ref{baroclinic_torque1} and 
+ Note that the derivation of Eq. \ref{baroclinic_torque1} and 
   Eq. \ref{theoretical_rate} do not 
@@
   self-gravity (e.g. from Eq. \ref{wavenumber}).
-}
 
@@
 
-{\bf
   We demonstrate the destabilising effect of a fixed temperature
@@
   more favourable for destabilisation.
-}
 
@@
 \subsection{Codes}
-{\bf We use three independent grid-based codes to simulate the above 
-system.} We adopt computational units such that
+We use three independent grid-based codes to simulate the above 
+system. We adopt computational units such that
 $G=M_*=R_0=1$. Time is measured  in the Keplerian orbital period at

### 2015-02-06 paper/paper.tex
paper accepted, minor corrections

@@
 \begin{abstract} 
-  {\bf We describe a new mechanism that leads to the destabilisation of
+   We describe a new mechanism that leads to the destabilisation of
  non-axisymmetric waves in astrophysical discs with 
@@
  discs.
-}
+
 \end{abstract}
@@
 presence of spiral arms in circumstellar discs could be signposts of 
-planet formation \citep{juhasz14}.
+planet formation \citep[but see][]{juhasz14}.
 
 % but also natural in GI.
-Spiral arms are also {\bf characteristic of} global gravitational 
-instability (GI) {\bf in} differentially rotating discs
+Spiral arms are also  characteristic of global gravitational 
+instability (GI)  in differentially rotating discs
 \citep{goldreich65,laughlin96b,laughlin98,nelson98,lodato05,forgan11}. Large-scale 
@@
 \citep{adams89,heemskerk92,laughlin96,tremaine01,papaloizou02,hopkins10}. 
-In the `SLING' mechanism proposed by \cite{shu90}, a $m=1$ {\bf
-  gravitational} instability arises from the motion of the
+In the `SLING' mechanism proposed by \cite{shu90}, an $m=1$ 
+  gravitational instability arises from the motion of the
 central star induced by the one-armed perturbation, and requires a
@@
 
-{\bf
   In this work we identify a new mechanism that leads to the growth of
@@
   trailing waves when the disc temperature decreases outwards.
-}
 
-{\bf We employ direct hydrodynamic simulations using three different
+ We employ direct hydrodynamic simulations using three different
   grid-based codes to demonstrate how this `background torque' 
@@
   describe should also be applicable to non-self-gravitating discs.
-}
 
 This paper is organised as follows. In \S\ref {model} we describe the system of interest
-and list the governing equations. {\bf In \S\ref{wkb} we use linear theory  to show
+and list the governing equations. In \S\ref{wkb} we use linear theory  to show
   how a fixed temperature profile 
-  can destabilise non-axisymmetric waves in discs.} \S\ref{methods} describes
-the {\bf numerical setup and} hydrodynamic codes {\bf we use to 
+  can destabilise non-axisymmetric waves in discs. \S\ref{methods} describes
+the numerical setup and hydrodynamic codes we use to 
   demonstrate the growth of one-armed spirals due to an imposed
-  temperature gradient.} Our {\bf
-  simulation} results are presented in  \S\ref{results2d} for two-dimensional (2D) {\bf
-  discs} and in \S\ref{results3d} for three-dimensional (3D) {\bf discs}, and we further
+  temperature gradient. Our
+  simulation results are presented in  \S\ref{results2d} for two-dimensional (2D) 
+  discs and in \S\ref{results3d} for three-dimensional (3D) discs, and we further
 discuss them in \S\ref{discussions}. We summarise in \S\ref{summary}  
@@
 
-{\bf
 \section*{Acknowledgments}
@@
 Science Foundation under Grant No. 1228509.
-}
+
 \bibliographystyle{mn2e}

### 2015-02-06 paper/results.tex
paper accepted, minor corrections

@@
 Fig. \ref{fargo_2d}. 
-At early times $t\lesssim100P_0$ the disc {\bf is } 
+At early times $t\lesssim100P_0$ the disc is  
 dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
@@
 plotted since they are dominant. The $m=1$ structure has
-an associated negative angular momentum, {\bf which indicates it is a
-  low-frequency mode.}   
+an associated negative angular momentum,  which indicates it is a
+  low-frequency mode. 
 Its growth is compensated by an increase in the axisymmetric
@@
 \subsection{Dependence on the imposed temperature profile}
-{\bf We show that the growth of the $m=1$ spiral is
+ We show that the growth of the $m=1$ spiral is
   associated with the imposed temperature gradient  by performing a 
@@
   with $m=1$ perturbations only.
-}
 
@@
   \caption{Evolution of the $m=1$ spiral amplitude as a function of
-    the {\bf negative of the imposed temperature gradient} $q$. The maximum value of the
+    the negative of the imposed temperature gradient $q$. The maximum value of the
     $m=1$ surface density in $R\in[R_1,R_2]$ is shown. 
@@
 % difficult to measure very small growth rates 
-{\bf As the background torque is proportional to $q$
-  (Eq. \ref{theoretical_rate})}, 
-this {\bf indicates} that the temperature gradient is responsible for the
+ As the background torque is proportional to $q$
+  (Eq. \ref{theoretical_rate}), 
+this indicates that the temperature gradient is responsible for the
 development of the one-armed spirals observed in our simulations.
@@
 \subsection{Properties of the $m=1$ spiral and its growth}\label{fargo_m1}
-{\bf Here we analyse the $q=1$ case in Fig. \ref{fargo_varq_growth} in
+ Here we analyse the $q=1$ case in Fig. \ref{fargo_varq_growth} in
   more detail.
-} 
+
 Fig. \ref{2d_fargo_viz} shows a snapshot of the $m=1$ surface
-density {\bf of this run}. 
+density of this run. 
 By measuring the $m=1$ surface density amplitude and its
@@
 h^{-1}$. Since $Q=O(1)$ and $h\ll 1$ imply $|kR|\gg 1$, we 
-can apply {\bf results from local theory
-  (\S\ref{local_approx})}. {\bf Note also that $k\simeq k_c$. }
+can apply results from local theory
+  (\S\ref{local_approx}). Note also that $k\simeq k_c$. 
 Fig. \ref{2d_fargo_viz} shows the $m=1$
@@
 
-{\bf
   We can estimate the expected growth rate of the $m=1$ mode due to the
@@
   results.
-}
 
 \subsubsection{Angular momentum exchange with the background disc}  
-{\bf We explicitly show that the growth of the $m=1$ spiral is due to
+ We explicitly show that the growth of the $m=1$ spiral is due to
   the forced temperature gradient via the background torque described
@@
   assuming boundary fluxes are negligible, to obtain
-}
 
@@
 less applicable. Fig. \ref{fargo_angmom_ex} confirms that the $m=1$ spiral
-wave experiences a negative torque that further reduces its {\bf (negative)} angular
-momentum, {\bf leading to its amplitude growth.} This is consistent
+wave experiences a negative torque that further reduces its (negative) angular
+momentum, leading to its amplitude growth. This is consistent
 with angular momentum component measurements (Fig. \ref {2d_angmom}).

### 2015-02-06 paper/summary.tex
first revision complete, but still need spell check and proof read
paper accepted, minor corrections

--- 2015-01-09 paper/summary.tex
@@
 \section{Summary and conclusions}\label{summary}
-{\bf
+
   In this paper, we have described a destabilising  
@@
   isothermal discs.
-}
 
 In the local approximation, we showed that this background torque is
-negative for {\bf non-axisymmetric} trailing waves in discs with a fixed temperature or
-sound-speed profile that decrease outwards. {\bf The background torque} enforces  
-low-frequency {\bf non-axisymmetric} modes because they are associated
+negative for non-axisymmetric trailing waves in discs with a fixed temperature or
+sound-speed profile that decrease outwards. A negative background torque enforces  
+low-frequency non-axisymmetric modes because they are associated
 with negative angular momentum.
 
-{\bf 
   We demonstrated the destabilising effect of the background torque by
@@
   bump. 
-}
-We find such systems {\bf are} unstable to low-frequency perturbations
+We find such systems are unstable to low-frequency perturbations
 with azimuthal wavenumber $m=1$, which leads to the development of an one-armed 
@@
 one-armed spirals in our disc model is due to the imposed
-temperature {\bf gradient: growth rates increased linearly}  
+temperature gradient: growth rates increased linearly   
 with the magnitude of the imposed temperature gradient, and one-armed
-spirals did not develop in strictly isothermal simulations. {\bf This 
+spirals did not develop in strictly isothermal simulations. This 
   one-armed spiral instability can be interpreted as an initially 
   neutral, tightly-wound $m=1$ mode being destabilised by the 
-  background torque.} The spiral
-is mostly confined between two $Q$-barriers surface density bump. 
-We find the instability behaves similarly in 2D and 3D {\bf, but in 3D}
+  background torque. The spiral
+is mostly confined between two $Q$-barriers in the surface density bump. 
+We find the instability behaves similarly in 2D and 3D, but in 3D
 the spiral disturbance becomes more radially global away from    
@@
 
-{\bf There are several issues that remain to be addressed in future
-  works:}
+ There are several issues that remain to be addressed in future
+  works:
 
-{\bf \emph{Thermal relaxation.} The locally isothermal assumption 
-} can be relaxed by including an energy equation with
-{\bf a source term that restores the disc temperature over a
-  characteristic timescale $t_\mathrm{relax}$}. Preliminary FARGO
+ \emph{Thermal relaxation.} The locally isothermal assumption 
+ can be relaxed by including an energy equation with
+ a source term that restores the disc temperature over a
+  characteristic timescale $t_\mathrm{relax}$. Preliminary FARGO
 simulations indicate a thermal relaxation timescale $t_\mathrm{relax} <
 0.1\Omega_k^{-1}$ is needed for the one-armed spiral to
-develop. However, this value may {\bf be model-dependent} and will be
-explored in a follow-up study. {\bf For example,
+develop. However, this value is likely model-dependent. For example,
 a longer $t_\mathrm{relax}$ may be permitted with larger temperature
-gradients.}
+gradients. This issue, together with a parameter survey, will be
+considered in a follow-up study.
 
@@
 
-{\bf \emph{Non-linear evolution}.} In the deeply non-linear regime, the
+ \emph{Non-linear evolution}. In the deeply non-linear regime, the
 one-armed spiral may  
@@
 gain of positive angular momentum through shock dissipation. We remark  
-that fragmentation {\bf of a low-frequency spiral} is unlikely
-because the co-rotation radius is outside the bulk of the spiral arm, in the non-self-gravitating
-portion of the disc \citep{durisen08}. In order
+that fragmentation is unlikely because the co-rotation radius is
+outside the bulk of the spiral arm \citep{durisen08,rogers12}. In order
 to study these possibilities, improved numerical models are needed to
@@
 than that considered in this paper.
+
+%interaction with turbulence
 
@@
 
-{\bf The background torque in other contexts.}  
-{\bf In our simulations}, self-gravity coupled with the disc 
-structure permits a neutral one-armed mode, which is then
-destabilised by the {\bf fixed} temperature gradient.  
-Other origins of the neutral one-armed spiral should be
-investigated. One possibility is in Be star discs \citep{rivinius13},
+\emph{Other applications of the background torque.}     
+ The background torque is a generic feature in  
+  discs for which the temperature is set externally.  It may therefore
+  be relevant in other astrophysical contexts. 
+One possibility is in Be star discs \citep{rivinius13}, 
 for which one-armed oscillations may explain long-timescale variations
 in their emission lines \citep[see e.g.][and references 
-therein]{okasaki97,papaloizou06c,ogilvie08}. These studies employ
-strictly isothermal disc models, so it would be interesting to explore
-the effect of a temperature gradient on the stability of these
+therein]{okasaki97,papaloizou06c,ogilvie08}. These studies 
+invoke alternative mechanisms to produce \emph{neutral} one-armed
+oscillations (e.g. rotational deformation of the star), but consider 
+strictly isothermal discs. It would be interesting to explore
+the effect of a radial temperature gradient on the stability of these
 oscillations.
